### 2014-05-20 Rnw/metadata_tables.Rnw
first commit; contains metadata_tables

\documentclass{article}

\begin{document}

\title{Metadata, tables, maps}
\maketitle

<<load, echo=FALSE, message=FALSE>>=
baad <- readRDS("data/baad.rds")$data

# library(smatr)
# library(magicaxis)
library(scales)
library(xtable)
source("R/analysis_utils.R")
source("R/plotting.R")
@

<<results='asis', echo=FALSE>>=

# Top 15 families
topfam <- names(sort(table(baad$family),T)[1:15])
datx <- droplevels(subset(baad, family %in% topfam))
datx$family <- with(datx, reorder(as.factor(family),family,function(x)1/length(x)))

famtab <- xtable(xtabs(~ family, data=datx), digits=0,
                 caption="Most represented families (number of observations).",
                 label="table:familytable")
print(famtab, caption.placement="top",table.placement="h")

# Top 15 species
topspec <- names(sort(table(baad$speciesMatched),T)[1:15])
datx <- droplevels(subset(baad, speciesMatched %in% topspec))
datx$Species <- with(datx, reorder(as.factor(speciesMatched),speciesMatched,
                                   function(x)1/length(x)))

spectab <- xtable(xtabs(~ Species, data=datx), digits=0,
                 caption="Most represented species (number of observations).",
                 label="table:spectable")
print(spectab, caption.placement="top",table.placement="h")

dat_pft <- within(baad, {
  pft[is.na(pft) | pft == ""] <- "None"
  growingCondition[is.na(growingCondition) | growingCondition == ""] <- "None"
  vegetation[is.na(vegetation) | vegetation == ""] <- "None"

  pft <- relevel(as.factor(pft), "None")
  growingCondition <- relevel(as.factor(growingCondition), "None")
  vegetation <- relevel(as.factor(vegetation), "None")
})

# Vegetation and plant functional type
vegpfttab <- xtable(addmargins(xtabs(~ vegetation + pft, data=dat_pft)), 
                    digits=0,
                    caption="Number of observations by family and plant functional type.",
                    label="table:vegpfttab")
print(vegpfttab, caption.placement="top",table.placement="h")

# growinCondition and plant functional type
gcpfttab <- xtable(addmargins(xtabs(~ growingCondition + pft, data=dat_pft)), 
                   digits=0,
                    caption="Number of observations by growing condition and plant functional type.",
                    label="table:gcpfttab")
print(gcpfttab, caption.placement="top",table.placement="h")

# Table of variables and nr of observations.
nr <- function(x)sum(!is.na(baad[,x]))
cfg <- read.csv("data/variabledefinitions.csv")

vars <- c("m.st","m.so","m.lf","a.lf","m.rt","h.t","c.d","d.bh", "ma.ilf","r.st")
nrdf <- data.frame(Variable=vars, nobs=unname(sapply(vars,nr)))

nrdf <- merge(nrdf, cfg[,c("Variable","label")])
nrdf <- nrdf[order(nrdf$nobs,decreasing=TRUE),]
nrdftab <- xtable(nrdf, caption = "Number of observations by selected variables.", 
                  label = "nrdftable")
print(nrdftab,caption.placement = "top", include.rownames= FALSE, table.placement = "h")
@

%,  out.width='1.3\\linewidth', out.height='1.3\\linewidth'>>=
<<echo=FALSE, fig.cap="Global distribution of datasets by vegetation type and sample size.",out.width='1.1\\linewidth'>>=  
# Colored by vegetation type
cols <- alpha(c("blue","brown","forestgreen","darkorange","dodgerblue2",
                "navajowhite4","lightgoldenrod3","lavender"),0.67)
vegs <- c("BorF","Sav","TempF","TempRF","TropRF","TropSF","Wo")
vegs_labels <- c("Boreal forest","Savannah","Temperate forest","Temperate rainforest","Tropical rainforest",
                 "Tropical seasonal forest","Woodland","Glasshouse")

drawWorldPlot(baad[baad$vegetation == vegs[1],], horlines=FALSE, sizebyn=TRUE, pchcol=cols[1])
for(i in 2:length(vegs)){
  drawWorldPlot(baad[baad$vegetation == vegs[i],], horlines=FALSE, sizebyn=TRUE, pchcol=cols[i], add=TRUE)
}
drawWorldPlot(baad[baad$growingCondition == "GH",], horlines=FALSE, sizebyn=TRUE, 
              pchcol=cols[8], add=TRUE)
par(xpd=NA)
legend(-160, -100, vegs_labels[1:4], fill=cols[1:4], bty='n')
legend(-50, -100, vegs_labels[5:8], fill=cols[5:8], bty='n')

@

\end{document}

### 2014-05-20 R/plotting.R

@@ -1,0 +1,438 @@
+suppressPackageStartupMessages(library(maps))
+suppressPackageStartupMessages(library(mapdata))
+suppressPackageStartupMessages(library(gdata))
+suppressPackageStartupMessages(library(smatr))
+
+
+to.pdf <- function(expr, filename, ..., verbose=TRUE) {
+  if ( verbose )
+    cat(sprintf("Creating %s\n", filename))
+  
+  if(!file.exists(dirname(filename)))
+    dir.create(dirname(filename), recursive=TRUE)
+  
+  pdf(filename, ...)
+  on.exit(dev.off())
+  eval.parent(substitute(expr))
+}
+
+makePlotPanel<-function(data, study, dir="report-per-study", col="grey", pdf=TRUE, quiet=FALSE){
+  if(!quiet)cat("This is how the study", study, "fits in the entire dataset distribution")
+  
+  #browser()
+  if(pdf){    
+    path<-paste0("output/", dir)
+    if(!file.exists(path))
+      dir.create(path)
+    pdf(file=paste0(path,"/", study,".pdf"))
+  }
+  
+  dat        <-  data[data$dataset==study,]
+  plot.vars  <-  mashrDetail("var.def")$Variable[mashrDetail("var.def")$Group=="tree"]
+  plot.vars  <-  plot.vars[plot.vars %in% c("growingCondition","status","light") == FALSE]
+  available  <-  dat[,names(dat) %in% plot.vars]   
+  available  <-  apply(available,2,as.numeric)
+  available  <-  available[,colSums(available, na.rm=TRUE) != 0]
+  
+  #set up a vector of colors, each species with different color
+  species  <-  as.numeric(as.factor(dat$species))
+  
+#  if(min(species)!=1){species  <-  species - (min(species)-1)}
+  
+  colorvec <-  niceColors(length(unique(species)))
+  colorvec <-  colorvec[match(species, unique(species))]
+  
+  par(mfrow=c(2,2))
+  z  <- 0
+  for(j in 1:(dim(available)[2]-1)){
+    for(k in (j+1):dim(available)[2]){
+      testNa  <-  available[,j]+available[,k]
+      if(length(testNa[!is.na(testNa)]) != 0){
+        z  <-  z + 1
+        if(z==5){par(mfrow=c(2,2)); z  <- 1}
[…]

### 2014-05-20 analysis.R

@@ -1,0 +1,14 @@
+
+
+source("R/analysis_utils.R")
+
+# Also need scales, xtable
+library(knitr)
+
+
+
+knitFromHere("Rnw/metadata_tables.Rnw")
+
+
+# BAAD
+baad <- readRDS("data/baad.rds")$data

### 2014-05-20 Rnw/stem_leaf_scaling.Rnw
stem_leaf_scaling compiles

\documentclass{article}

\begin{document}

\title{Scaling of leaf and above-ground woody biomass}
\maketitle

<<load, echo=FALSE, message=FALSE>>=
baad <- readRDS("data/baad.rds")$data
source("R/loadPackages.R")
source("R/analysis_utils.R")
source("R/plotting.R")
@

<<prep, echo=FALSE>>=
# Stem-leaf scaling

# Select datasets that have stem, leaf biomass, and height
x <- suppressMessages(studyWithVars(baad, c("m.st","m.lf","h.t")))
x$dataset_species <- paste0(x$dataset,"_",x$species)

# clean
x <- subset(x, m.lf > 0 & m.st > 0 & !is.na(m.lf) & !is.na(m.st) & !is.na(h.t))

# Keep only dataset_species with at least 20 observations
tab <- table(x$dataset_species)
x <- x[x$dataset_species %in% names(tab)[tab>19],]
@

The selected data has \Sexpr{nrow(x)} rows, and \Sexpr{length(unique(x$dataset_species))} unique dataset-species combinations.

<<echo=FALSE>>=
xtabs(~ pft + vegetation,data=x)
@

<<analysis, echo=FALSE>>=

# sma by dataset_species
sm_mlf <- sma(m.lf ~ h.t * dataset_species, log="xy", data=x)
sm_mst <- sma(m.st ~ h.t * dataset_species, log="xy", data=x)

# quadratic across all data

# sma coefficients
makedfr <- function(z){
  z$dataset_species <- rownames(z)
  as.data.frame(z)
}
p_mlf <- coef(sm_mlf)
names(p_mlf) <- c("b0_mlf","b1_mlf")
p_mst <- coef(sm_mst)
names(p_mst) <- c("b0_mst","b1_mst")
p <- merge(makedfr(p_mlf),makedfr(p_mst))

# Find ranges for h.t
r <- t(sapply(p$dataset_species, function(lev){
    range(x$h.t[x$dataset_species==lev], na.rm=TRUE)
}))
colnames(r) <- c("ht_min","ht_max")

# P is a dataframe with leaf, stem scaling, and ranges of h.t
P <- cbind(p,r)
rownames(P) <- NULL

# Add some metadata
dfr <- baad[,c("dataset","species","pft","vegetation","growingCondition")]
dfr$dataset_species <- with(dfr, paste0(dataset,"_",species))
dfr <- dfr[!duplicated(dfr$dataset_species),]
P <- merge(P, dfr)
P$pft <- as.factor(P$pft)
P$vegetation[is.na(P$vegetation)] <- "GH"
P$vegetation <- as.factor(P$vegetation)

# Plot : save predictions in a list
l <- lapply(1:nrow(P), function(i){
  h <- seq(P$ht_min[i], P$ht_max[i], length=101)
  a <- 10^(P$b0_mlf[i])
  b <- P$b1_mlf[i]
  c <- 10^(P$b0_mst[i])
  d <- P$b1_mst[i]
  wfrel <- a*h^b/(a*h^b + c*h^d)
  mst <- c*h^d
  mlf <- a*h^b
  return(data.frame(h=h,wfrel=wfrel,mlf=mlf,mst=mst))
})

@

\section{By plant functional type}

<<plotting, echo=FALSE>>=

# nonfield <- union(which(P$vegetation == "GH"), which(P$dataset == "Poorter1999"))
nonfield <- which(P$growingCondition %in% c("CG","GH"))
field <- setdiff(1:nrow(P), nonfield)

# Non-glasshouse
palette(alpha(c("blue","royalblue","red","darkorange"),0.4))
plot(1, xlim=c(-3,2), ylim=c(-3,0), type='n',
     xlab="Plant height (m)", ylab="Leaf mass / aboveground mass (-)",
     axes=FALSE, main="Field grown")
addlogaxes(10^-3, 10^2, 10^-3, 10^0, tck1=-0.03, tck2=-0.015)
for(i in field){
  points(log10(l[[i]]$h), log10(l[[i]]$wfrel), type='l',
         col=P$pft[i])
}
legend("bottomleft", c("Deciduous angiosperm","Deciduous gymnosperm",
                       "Evergreen angiosperm","Evergreen gymnosperm"),
       fill=alpha(palette(),0.9))
# Some flourishes for pdf in a presentation
# l_eve <- rbind.fill(l[P$pft %in% c("EA","EG")])
# l_dec <- rbind.fill(l[P$pft %in% c("DA","DG")])
# 
# library(plantecophys)
# l_eve$logwfrel <- log10(l_eve$wfrel)
# l_eve$logh <- log10(l_eve$h)
# l_dec$logwfrel <- log10(l_dec$wfrel)
# l_dec$logh <- log10(l_dec$h)
# gamplot("logh","logwfrel",dfr=l_eve,linecolor="red",add=TRUE)
# gamplot("logh","logwfrel",dfr=l_dec,linecolor="blue",add=TRUE)
# 
# dev.copy2pdf(file="analysis/output/figures/LMF_height_loglog.pdf")

palette(alpha(c("blue","royalblue","red","darkorange"),0.4))
plot(1, xlim=c(-3,2), ylim=c(-3,0), type='n',
     xlab="Plant height (m)", ylab="Leaf mass / aboveground mass (-)",
     axes=FALSE, main="Glasshouse and Common garden")
addlogaxes(10^-3, 10^2, 10^-3, 10^0, tck1=-0.03, tck2=-0.015)
for(i in nonfield){
  points(log10(l[[i]]$h), log10(l[[i]]$wfrel), type='l',
         col=P$pft[i])
}
legend("bottomleft", c("Deciduous angiosperm","Deciduous gymnosperm",
                       "Evergreen angiosperm","Evergreen gymnosperm"),
       fill=alpha(palette(),0.9))

# m.lf, m.st vs. h.t by pft.
plot(1, xlim=c(-3,2), ylim=c(-7,5), type='n',
     xlab="Plant height (m)", ylab="Leaf mass (kg)",
     axes=FALSE)
magaxis(side=1:2, unlog='xy',box=TRUE)
for(i in 1:length(l))points(log10(l[[i]]$h), log10(l[[i]]$mlf), type='l',
                            col=P$pft[i])
legend("topleft", c("Deciduous angiosperm","Deciduous gymnosperm",
                       "Evergreen angiosperm","Evergreen gymnosperm"),
       fill=palette(), inset=0.04)

plot(1, xlim=c(-3,2), ylim=c(-7,5), type='n',
     xlab="Plant height (m)", ylab="Above-ground woody mass (kg)",
     axes=FALSE)
magaxis(side=1:2, unlog='xy',box=TRUE)
for(i in 1:length(l))points(log10(l[[i]]$h), log10(l[[i]]$mst), type='l',
                            col=P$pft[i])
legend("topleft", c("Deciduous angiosperm","Deciduous gymnosperm",
                    "Evergreen angiosperm","Evergreen gymnosperm"),
       fill=palette(), inset=0.04)
@

\section{Tropical vs. temperate evergreens}

<<echo=FALSE>>=
ii <- which(P$pft == "EA" & P$vegetation %in% c("TempF","TempRF","TropRF","TropSF"))
l <- l[ii]
P <- P[ii,]

P$TempTrop <- as.factor(ifelse(P$vegetation %in% c("TempF","TempRF"), "Temp", "Trop"))
palette(c("blue","red"))

@

<<plotEA, echo=FALSE>>=

# log-log scale
plot(1, xlim=c(-3,2), ylim=c(0,1), type='n',
     xlab="Plant height (m)", ylab="Leaf mass / aboveground mass (-)",
     axes=FALSE)
addlogaxes(10^-3, 10^2, 10^-3, 10^0, tck1=-0.03, tck2=-0.015,addY=F)
axis(2)
for(i in 1:length(l)){
  points(log10(l[[i]]$h), l[[i]]$wfrel, type='l',
         col=P$TempTrop[i])
}
legend("bottomleft", c("Temperate evergreen angiosperm",
                       "Tropical evergreen angiosperm"), fill=palette())

plot(1, xlim=c(-3,2), ylim=c(-7,5), type='n',
     xlab="Plant height (m)", ylab="Leaf mass (kg)",
     axes=FALSE)
magaxis(side=1:2, unlog='xy',box=TRUE)
for(i in 1:length(l))points(log10(l[[i]]$h), log10(l[[i]]$mlf), type='l',
                            col=P$TempTrop[i])
legend("topleft", c("Temperate evergreen angiosperm",
                       "Tropical evergreen angiosperm"), fill=palette(),inset=0.03)

plot(1, xlim=c(-3,2), ylim=c(-7,5), type='n',
     xlab="Plant height (m)", ylab="Above-ground woody mass (kg)",
     axes=FALSE)
magaxis(side=1:2, unlog='xy',box=TRUE)
for(i in 1:length(l))points(log10(l[[i]]$h), log10(l[[i]]$mst), type='l',
                            col=P$TempTrop[i])
legend("topleft", c("Temperate evergreen angiosperm",
                       "Tropical evergreen angiosperm"), fill=palette(),inset=0.03)

@

\end{document}

### 2014-05-20 analysis.R

@@ -4,8 +4,16 @@ source("R/analysis_utils.R")
 source("R/loadPackages.R")
 source("R/plotting.R")
 
-
+# Meta data summaries, tables, world map
 knitFromHere("Rnw/metadata_tables.Rnw")
-  
+
+# stem-leaf scaling
+knitFromHere("Rnw/stem_leaf_scaling.Rnw")
+
+# pipe model plot from Daniel
+source("R/pipe.R")
+
+
+
 # BAAD
 baad <- readRDS("data/baad.rds")$data

### 2014-06-18 Rmd/allombypft.Rmd
added Rmd/allombypft.Rmd

---
title: "Allometry by dataset/species and pft"
output: html_document
---

```{r echo=FALSE}

source("load.R")

# baad2 : discard pot-grown plants
baad$Group <- paste(baad$dataset, baad$speciesMatched)
baad2 <- subset(baad, growingCondition %in% c("FW","PM","PU","FE"))
baad2$pft <- as.factor(baad2$pft)

# estimate basal diameter from dbh and h.t
dba <- with(baad2, d.bh * h.t / (h.t - 1.3))
dba[!is.finite(dba)] <- NA
dba[!(baad2$h.bh %in% c(1.3,1.37))] <- NA 
dba[baad2$h.t < 1.5] <- NA
baad2$d.ba[is.na(baad2$d.ba)] <- dba[is.na(baad2$d.ba)]

allombygroupandpftplot <- function(xvar, yvar, dataset, xlab=AxisLabels[[xvar]], ylab=AxisLabels[[yvar]], 
                                   slopecomp=1, n_min=20,legend=FALSE,
                                   pftcols=list(EA="red", EG="hotpink", DA="blue")){

  dataset$X <- dataset[,xvar]
  dataset$Y <- dataset[,yvar]

  smfitsEA <- sma(Y ~ X*Group, data=subset(dataset,pft=="EA"), log='xy', n_min=n_min, quiet=TRUE)
  smfitsEG <- sma(Y ~ X*Group, data=subset(dataset,pft=="EG"), log='xy', n_min=n_min, quiet=TRUE)
  smfitsDA <- sma(Y ~ X*Group, data=subset(dataset,pft=="DA"), log='xy', n_min=n_min, quiet=TRUE)
  smfits <- sma(Y ~ X*Group, data=dataset, log='xy', n_min=n_min, quiet=TRUE)

  plot(smfits, col="black", type='n', axes=FALSE,
       xlab=xlab,
       ylab=ylab,
       panel.first={
         for(z in -15:15)abline(z,slopecomp,col="grey")
       })
  magaxis(side=1:2, unlog=1:2)
  plot(smfitsEA, add=TRUE, type='l', col=pftcols$EA)
  plot(smfitsEG, add=TRUE, type='l', col=pftcols$EG)
  plot(smfitsDA, add=TRUE, type='l', col=pftcols$DA)

  if(legend)legend("topleft", c("Evergreen Angio.","Evergreen Gymno", "Deciduous Angio."), 
                   fill=unlist(pftcols), cex=0.9)  
return(invisible(smfits))
}

AxisLabels <- list(m.to = "Total plant mass (kg)",
                   m.so = "Above-ground plant mass (kg)",
                   m.lf = "Total leaf mass (kg)",
                   m.rt = "Total root mass (kg)",
                   m.rf = "Fine root mass (kg)",
                   d.bh = "Diameter at breast height (m)",
                   d.ba = "Diameter at base (m)",
                   a.stbh = expression("Stem area at breast height"~~(m^2)),
                   a.stba = expression("Stem area at base"~~(m^2)),
                   a.lf = expression("Total leaf area"~~(m^2)),
                   h.t = "Total plant height (m)",
                   h.c = "Height to crown base (m)",
                   c.d = "Crown length (m)",
                   d.cr = "Crown diameter (m)")

```

```{r echo=FALSE}

par(cex.lab=1.2, mar=c(5,5,2,2))
allombygroupandpftplot("m.to", "m.lf", baad2)
allombygroupandpftplot("m.to", "a.lf", baad2)

allombygroupandpftplot("m.so", "m.lf", baad2)
allombygroupandpftplot("m.so", "a.lf", baad2)

allombygroupandpftplot("a.stbh", "m.lf", baad2)
allombygroupandpftplot("a.stba", "m.lf", baad2)

allombygroupandpftplot("h.t", "m.so", baad2, slopecomp=2, legend=T)
allombygroupandpftplot("d.bh", "m.so", baad2, slopecomp=2, legend=T)
allombygroupandpftplot("d.ba", "m.so", baad2, slopecomp=2, legend=T)

allombygroupandpftplot("m.so", "m.rt", baad2, legend=T)
allombygroupandpftplot("m.lf", "m.rf", baad2, legend=T)

allombygroupandpftplot("c.d", "m.lf", baad2, legend=T)
allombygroupandpftplot("c.d", "d.cr", baad2, legend=T)

```

### 2014-06-18 Rmd/baadplotscollection.Rmd
Lots of updating; most plots now collected in Rmd/baadplotscollection.Rmd

---
title: "Collection of BAAD plots"
output: html_document
---

```{r echo=FALSE}

source("load.R")

```

Goals
1. Find systematic differences that need to be accounted for in models.
2. Previous focus on universal patterns, but have not been able to look at pft / biome effects.
3. How much of variance due to species, biome, pft?
4. Functional balance : plots of root mass vs. leaf area or leaf mass
5. Root-shoot ratios : aboveground biomass vs. root mass.

### Pipe-model relationships

This figure shows average calculated pipe-model relationships, either at breast height (using `a.stbh`) or base (`a.stba`). Pipe-model ratios are calculated by dataset/species combinations, error bars are 1 S.E., sample sizes above bars are number of dataset/species combinations.

Deciduous gymnosperm (*Larix*) was removed for clarity (only two datasets with a single species, and only breast height).

In the plots, A<sub>L</sub> is leaf area, A<sub>W</sub> is stem cross-sectional area, M<sub>F</sub> is leaf mass.

```{r echo=FALSE,message=FALSE}

# BH pipemodel relationships
datl <- studyWithVars(baad, c("a.stbh", "a.lf","m.lf"), returnwhat="list")
dat <- rbind_all(datl)

# Roth2007 massive outlier, remove here
dat <- subset(dat, dataset != "Roth2007")

# remove DG; there is only one species which makes it confusing
dat <- subset(dat, pft != "DG")

pipe_bh <- summarize(group_by(dat, dataset, species), 
           alf_stbh=mean(a.lf / a.stbh),
           mlf_stbh=mean(m.lf / a.stbh),
           sla = mean(a.lf / m.lf),
           pft=first(pft),
           MAP=mean(MAP),
           MAT=mean(MAT),
           growingCondition=first(growingCondition),
           vegetation=first(vegetation))

# Basal pipemodel relationships
datl <- studyWithVars(baad, c("a.stba", "a.lf","m.lf"), returnwhat="list")

dat$pft <- as.factor(dat$pft)
pipe_bh$pft <- as.factor(pipe_bh$pft)

pipe_ba <- summarize(group_by(dat, dataset, species), 
                     alf_stba=mean(a.lf / a.stba),
                     mlf_stba=mean(m.lf / a.stba),
                     pft=first(pft),
                     MAP=mean(MAP),
                     MAT=mean(MAT),
                     growingCondition=first(growingCondition),
                     vegetation=first(vegetation))

BARplot <- function(x,y,ymult=1.2,...){

  N <- table(x)
  tap <- tapply(y,x,mean,na.rm=T)
  Ylim <- c(0, max(tap)*ymult)
  b <- bargraph.CI(x,y,ylim=Ylim,err.width=0.04,...)

  d <- 0.3
  text(c(0.3, b$xvals+d), max(tap)*(1.1/1.2)*ymult, labels=c("N =", as.character(unname(N))))

  return(invisible(b))
}

Cols <- c("blue","red","darkorange")
palette(Cols)
# Cols <- c("blue","royalblue","red","darkorange")

par(mfrow=c(2,2), mar=c(5,5,2,2))
with(pipe_bh, BARplot(pft,alf_stbh, xlab="Plant functional type", 
                       ylab=expression(A[L]/A[W]~breast~height~~(m^2~m^-2)),
                       col=Cols))
with(pipe_bh, BARplot(pft,mlf_stbh, xlab="Plant functional type",
                       ylab=expression(M[F]/A[W]~breast~height~~(kg~m^-2)),
                       col=Cols
                       ))
with(pipe_ba, BARplot(pft,alf_stba, xlab="Plant functional type", 
                          ylab=expression(A[L]/A[W]~base~~(m^2~m^-2)),
                          col=Cols))
b <- with(pipe_ba, BARplot(pft,mlf_stba, xlab="Plant functional type",
                          ylab=expression(M[F]/A[W]~base~~(kg~m^-2)),
                          col=Cols,ymult=1.3
))

with(pipe_bh, plot(sla, alf_stbh, pch=19, col=pft, xlim=c(0,50),
                   xlab=expression(Specific~leaf~area~~(m^2~kg^-1)),
                   ylab=expression(A[L]/A[W]~breast~height~~(m^2~m^-2))))
legend("topright", levels(pipe_bh$pft), col=palette(), pch=19)

with(pipe_bh, plot(sla, mlf_stbh, pch=19, col=pft, xlim=c(0,50),
                   xlab=expression(Specific~leaf~area~~(m^2~kg^-1)),
                   ylab=expression(M[F]/A[W]~breast~height~~(m^2~m^-2))))
legend("topright", levels(pipe_bh$pft), col=palette(), pch=19)

```

### Allometry relationships by dataset and pft

The following figures use a subset of data, including only field-grown plants, and only species/dataset combinations that have at least 20 observations. Lines are `sma` fits.

```{r echo=FALSE}

# baad2 : discard pot-grown plants
baad$Group <- paste(baad$dataset, baad$speciesMatched)
baad2 <- subset(baad, growingCondition %in% c("FW","PM","PU","FE"))
baad2$pft <- as.factor(baad2$pft)

# estimate basal diameter from dbh and h.t
dba <- with(baad2, d.bh * h.t / (h.t - 1.3))
dba[!is.finite(dba)] <- NA
dba[!(baad2$h.bh %in% c(1.3,1.37))] <- NA 
dba[baad2$h.t < 1.5] <- NA
baad2$d.ba[is.na(baad2$d.ba)] <- dba[is.na(baad2$d.ba)]

allombygroupandpftplot <- function(xvar, yvar, dataset, xlab=AxisLabels[[xvar]], ylab=AxisLabels[[yvar]], 
                                   slopecomp=1, n_min=20,legend=FALSE,
                                   pftcols=list(EA="red", EG="hotpink", DA="blue")){

  dataset$X <- dataset[,xvar]
  dataset$Y <- dataset[,yvar]

  smfitsEA <- sma(Y ~ X*Group, data=subset(dataset,pft=="EA"), log='xy', n_min=n_min, quiet=TRUE)
  smfitsEG <- sma(Y ~ X*Group, data=subset(dataset,pft=="EG"), log='xy', n_min=n_min, quiet=TRUE)
  smfitsDA <- sma(Y ~ X*Group, data=subset(dataset,pft=="DA"), log='xy', n_min=n_min, quiet=TRUE)
  smfits <- sma(Y ~ X*Group, data=dataset, log='xy', n_min=n_min, quiet=TRUE)

  plot(smfits, col="black", type='n', axes=FALSE,
       xlab=xlab,
       ylab=ylab,
       panel.first={
         for(z in -15:15)abline(z,slopecomp,col="grey")
       })
  magaxis(side=1:2, unlog=1:2)
  plot(smfitsEA, add=TRUE, type='l', col=pftcols$EA)
  plot(smfitsEG, add=TRUE, type='l', col=pftcols$EG)
  plot(smfitsDA, add=TRUE, type='l', col=pftcols$DA)

  if(legend)legend("topleft", c("Evergreen Angio.","Evergreen Gymno", "Deciduous Angio."), 
                   fill=unlist(pftcols), cex=0.9)  
return(invisible(smfits))
}

AxisLabels <- list(m.to = "Total plant mass (kg)",
                   m.so = "Above-ground plant mass (kg)",
                   m.lf = "Total leaf mass (kg)",
                   m.rt = "Total root mass (kg)",
                   m.rf = "Fine root mass (kg)",
                   d.bh = "Diameter at breast height (m)",
                   d.ba = "Diameter at base (m)",
                   a.stbh = expression("Stem area at breast height"~~(m^2)),
                   a.stba = expression("Stem area at base"~~(m^2)),
                   a.lf = expression("Total leaf area"~~(m^2)),
                   h.t = "Total plant height (m)",
                   h.c = "Height to crown base (m)",
                   c.d = "Crown length (m)",
                   d.cr = "Crown diameter (m)")

```

```{r echo=FALSE, warning=FALSE}

par(cex.lab=1.2, mar=c(5,5,2,2))
allombygroupandpftplot("m.to", "m.lf", baad2, legend=T)
allombygroupandpftplot("m.to", "a.lf", baad2)

allombygroupandpftplot("m.so", "m.lf", baad2)
allombygroupandpftplot("m.so", "a.lf", baad2)

allombygroupandpftplot("a.stbh", "m.lf", baad2)
allombygroupandpftplot("a.stba", "m.lf", baad2)

allombygroupandpftplot("h.t", "m.so", baad2, slopecomp=2, legend=T)

allombygroupandpftplot("h.t", "m.lf", baad2, slopecomp=2, legend=T)
allombygroupandpftplot("h.t", "a.lf", baad2, slopecomp=2, legend=T)

allombygroupandpftplot("d.bh", "m.so", baad2, slopecomp=2, legend=T)
allombygroupandpftplot("d.ba", "m.so", baad2, slopecomp=2, legend=T)

allombygroupandpftplot("m.so", "m.rt", baad2, legend=T)
allombygroupandpftplot("m.lf", "m.rf", baad2, legend=T)

allombygroupandpftplot("c.d", "m.lf", baad2, legend=T)
allombygroupandpftplot("c.d", "d.cr", baad2, legend=T)

```

### Leaf mass ratio (LMR) by plant functional type

The following uses only dataset/species combinations that have at least 20 observations. Using `sma`, I fit leaf mass vs. plant height, and stem mass vs. plant height. From the coefficients of the fits we can calculate the size-dependent LMR relationship, which is not a log-log relationship (math will follow here).

We conclude that

1. Evergreens have much higher LMR at a given plant height than deciduous plants, but angiosperm/gymnosperm does not seem to matter (analysis will follow)
2. Field grown plants show similar relationships, a declining LMR with increasing plant height. Sheltered (glasshouse or common garden) plants show a wide range of relationships.

```{r echo=FALSE, message=FALSE}

# Select datasets that have stem, leaf biomass, and height
x <- studyWithVars(baad, c("m.st","m.lf","h.t"))
x$dataset_species <- paste0(x$dataset,"_",x$species)

# clean
x <- subset(x, m.lf > 0 & m.st > 0 & !is.na(m.lf) & !is.na(m.st) & !is.na(h.t))

# Keep only dataset_species with at least 20 observations
tab <- table(x$dataset_species)
x <- x[x$dataset_species %in% names(tab)[tab>19],]

sm_mlf <- sma(m.lf ~ h.t * dataset_species, log="xy", data=x)
sm_mst <- sma(m.st ~ h.t * dataset_species, log="xy", data=x)

# sma coefficients
makedfr <- function(z){
  z$dataset_species <- rownames(z)
  as.data.frame(z)
}
p_mlf <- coef(sm_mlf)
names(p_mlf) <- c("b0_mlf","b1_mlf")
p_mst <- coef(sm_mst)
names(p_mst) <- c("b0_mst","b1_mst")
p <- merge(makedfr(p_mlf),makedfr(p_mst))

# Find ranges for h.t
r <- t(sapply(p$dataset_species, function(lev){
    range(x$h.t[x$dataset_species==lev], na.rm=TRUE)
}))
colnames(r) <- c("ht_min","ht_max")

# P is a dataframe with leaf, stem scaling, and ranges of h.t
P <- cbind(p,r)
rownames(P) <- NULL

# Add some metadata
dfr <- baad[,c("dataset","species","pft","vegetation","growingCondition")]
dfr$dataset_species <- with(dfr, paste0(dataset,"_",species))
dfr <- dfr[!duplicated(dfr$dataset_species),]
P <- merge(P, dfr)
P$pft <- as.factor(P$pft)
P$vegetation[is.na(P$vegetation)] <- "GH"
P$vegetation <- as.factor(P$vegetation)

# Plot : save predictions in a list
l <- lapply(1:nrow(P), function(i){
  h <- seq(P$ht_min[i], P$ht_max[i], length=101)
  a <- 10^(P$b0_mlf[i])
  b <- P$b1_mlf[i]
  c <- 10^(P$b0_mst[i])
  d <- P$b1_mst[i]
  wfrel <- a*h^b/(a*h^b + c*h^d)
  mst <- c*h^d
  mlf <- a*h^b
  return(data.frame(h=h,wfrel=wfrel,mlf=mlf,mst=mst))
})

# nonfield <- union(which(P$vegetation == "GH"), which(P$dataset == "Poorter1999"))
nonfield <- which(P$growingCondition %in% c("CG","GH"))
field <- setdiff(1:nrow(P), nonfield)

# Non-glasshouse
palette(alpha(c("blue","royalblue","red","darkorange"),0.4))
plot(1, xlim=c(-3,2), ylim=c(-3,0), type='n',
     xlab="Plant height (m)", ylab="Leaf mass / aboveground mass (-)",
     axes=FALSE, main="Field grown")
addlogaxes(10^-3, 10^2, 10^-3, 10^0, tck1=-0.03, tck2=-0.015)
for(i in field){
  points(log10(l[[i]]$h), log10(l[[i]]$wfrel), type='l',
         col=P$pft[i])
}
legend("bottomleft", c("Deciduous angiosperm","Deciduous gymnosperm",
                       "Evergreen angiosperm","Evergreen gymnosperm"),
       fill=alpha(palette(),0.9))
# Some flourishes for pdf in a presentation
# l_eve <- rbind.fill(l[P$pft %in% c("EA","EG")])
# l_dec <- rbind.fill(l[P$pft %in% c("DA","DG")])
# 
# library(plantecophys)
# l_eve$logwfrel <- log10(l_eve$wfrel)
# l_eve$logh <- log10(l_eve$h)
# l_dec$logwfrel <- log10(l_dec$wfrel)
# l_dec$logh <- log10(l_dec$h)
# gamplot("logh","logwfrel",dfr=l_eve,linecolor="red",add=TRUE)
# gamplot("logh","logwfrel",dfr=l_dec,linecolor="blue",add=TRUE)
# 
# dev.copy2pdf(file="analysis/output/figures/LMF_height_loglog.pdf")

palette(alpha(c("blue","royalblue","red","darkorange"),0.4))
plot(1, xlim=c(-3,2), ylim=c(-3,0), type='n',
     xlab="Plant height (m)", ylab="Leaf mass / aboveground mass (-)",
     axes=FALSE, main="Glasshouse and Common garden")
addlogaxes(10^-3, 10^2, 10^-3, 10^0, tck1=-0.03, tck2=-0.015)
for(i in nonfield){
  points(log10(l[[i]]$h), log10(l[[i]]$wfrel), type='l',
         col=P$pft[i])
}
legend("bottomleft", c("Deciduous angiosperm","Deciduous gymnosperm",
                       "Evergreen angiosperm","Evergreen gymnosperm"),
       fill=alpha(palette(),0.9))
```

### MAT and MAP

Comparison between user-supplied and estimates from WorldClim of mean annual temperature (MAT) and mean annual precipitation (MAP).

```{r echo=FALSE}
par(cex.axis=0.7, xaxs="i", yaxs="i")
dat <- baad[!duplicated(paste(baad$latitude, baad$longitude)),]

par(mfrow=c(1,2))
with(dat, plot(MAP, map, xlab="MAP - WorldClim (mm)", ylab="MAP - User supplied (mm)", 
               pch=19, col=make.transparent("blue"),
               xlim=c(0,5200), ylim=c(0,5200)))
abline(0,1)
with(dat, plot(MAT, mat, xlab=expression("MAT - WorldClim"~(degree*C)), 
               ylab=expression("MAT - User supplied"~(degree*C)),
               xlim=c(0,32), ylim=c(0,32),
               pch=19, col=make.transparent("red")))
abline(0,1)
```

### Pipe model fits

```{r warning=FALSE, message=FALSE, echo=FALSE}
getAttributeByStudy <- function(data, var, studies){
  i <- !is.na(data[[var]])
  data[c("dataset", var)][!duplicated(paste( data[["dataset"]], data[[var]])),]
}

addIsoclines <- function(A= seq(-10,10),...){
  for(a in A) abline(a,1,lty="dotted",...)
}

myplot <- function(data, xvar,yvar,g,add=FALSE,col=niceColors(1),add11=TRUE,...){
  bivarPlotColorBy(data, xvar, yvar,g,col=make.transparent("grey", 0.4), 
                   pch=16,type='p',add=add,...)
  bivarPlotColorBy(data, xvar, yvar,group=g,col=col,lwd=1, type='l', add=TRUE)
}

myfigure <- function(data){

  #par(mfrow=c(2,4), oma = c(4,4,4,4))
  var <- "species"
  xlim <- c(10^-7, 1)
  ylim <- c(10^-5, 10^3)
  cols <- niceColors(3)

  f <- list()
  f[[1]] <- myplot(data,"a.stba", "a.lf", var,col=cols[1],xlim=xlim, 
                   ylim=ylim, main = "base")
  addIsoclines()
  f[[2]] <- myplot(data,"a.stbh", "a.lf", var,col=cols[2], xlim=xlim, 
                   ylim=ylim, main = "breast")
  addIsoclines()
  f[[3]] <- myplot(data,"a.stbc", "a.lf", var,col=cols[3],xlim=xlim, 
                   ylim=ylim, main = "crown base")
  addIsoclines()

  bivarPlotColorBy(data, "a.stba", "a.lf", var,type='l', col=cols[1], xlim=xlim, ylim=ylim, main = "all")
  bivarPlotColorBy(data, "a.stbh", "a.lf", var, type='l', col=cols[2], add=TRUE)
  bivarPlotColorBy(data, "a.stbc", "a.lf", var, type='l', col=cols[3], add=TRUE)
  addIsoclines()
  mtext("Total stem", side=4)

  f <- list()
  f[[1]] <- myplot(data,"a.ssba", "a.lf", var,col=cols[1],
                   xlim=xlim, ylim=ylim, main = "base")
  addIsoclines()
  f[[2]] <- myplot(data,"a.ssbh", "a.lf", var,col=cols[2],
                   xlim=xlim, ylim=ylim, main = "breast")
  addIsoclines()
  f[[3]] <- myplot(data,"a.ssbc", "a.lf", var,col=cols[3],
                   xlim=xlim, ylim=ylim, main = "crown base")
  addIsoclines()

  bivarPlotColorBy(data, "a.ssba", "a.lf", var, type='l', col=cols[1], xlim=xlim, ylim=ylim, main = "all")
  bivarPlotColorBy(data, "a.ssbh", "a.lf", var, type='l', col=cols[2], add=TRUE)
  bivarPlotColorBy(data, "a.ssbc", "a.lf", var, type='l', col=cols[3], add=TRUE)
  addIsoclines()
  mtext("Sapwood", side=4)
}
myfigure(baad)
```

### 2014-06-18 Rnw/metadata_tables.Rnw
Lots of updating; most plots now collected in Rmd/baadplotscollection.Rmd

@@
 
-%,  out.width='1.3\\linewidth', out.height='1.3\\linewidth'>>=
-<<echo=FALSE, fig.cap="Global distribution of datasets by vegetation type and sample size.",out.width='1.1\\linewidth'>>=  
-# Colored by vegetation type
-cols <- alpha(c("blue","brown","forestgreen","darkorange","dodgerblue2",
-                "navajowhite4","lightgoldenrod3","lavender"),0.67)
-vegs <- c("BorF","Sav","TempF","TempRF","TropRF","TropSF","Wo")
-vegs_labels <- c("Boreal forest","Savannah","Temperate forest","Temperate rainforest","Tropical rainforest",
-                 "Tropical seasonal forest","Woodland","Glasshouse")
-
-drawWorldPlot(baad[baad$vegetation == vegs[1],], horlines=FALSE, sizebyn=TRUE, pchcol=cols[1])
-for(i in 2:length(vegs)){
-  drawWorldPlot(baad[baad$vegetation == vegs[i],], horlines=FALSE, sizebyn=TRUE, pchcol=cols[i], add=TRUE)
-}
-drawWorldPlot(baad[baad$growingCondition == "GH",], horlines=FALSE, sizebyn=TRUE, 
-              pchcol=cols[8], add=TRUE)
-par(xpd=NA)
-legend(-160, -100, vegs_labels[1:4], fill=cols[1:4], bty='n')
-legend(-50, -100, vegs_labels[5:8], fill=cols[5:8], bty='n')
-
-@
-
 \end{document}

### 2014-06-18 R/WorldMap.R

@@ -1,0 +1,27 @@
+
+to.pdf({
+  
+  # Colored by vegetation type
+  cols <- alpha(c("blue","brown","forestgreen","darkorange","dodgerblue2",
+                    "navajowhite4","lightgoldenrod3","lavender"),0.67)
+  vegs <- c("BorF","Sav","TempF","TempRF","TropRF","TropSF","Wo")
+  vegs_labels <- c("Boreal forest","Savannah","Temperate forest","Temperate rainforest","Tropical rainforest",
+                   "Tropical seasonal forest","Woodland","Glasshouse / Common garden")
+  
+  drawWorldPlot(baad[baad$vegetation == vegs[1],], horlines=FALSE, sizebyn=TRUE, pchcol=cols[1])
+  for(i in 2:length(vegs)){
+    drawWorldPlot(baad[baad$vegetation == vegs[i],], horlines=FALSE, sizebyn=TRUE, pchcol=cols[i], add=TRUE)
+  }
+  drawWorldPlot(baad[baad$growingCondition == "GH",], horlines=FALSE, sizebyn=TRUE, 
+                pchcol=cols[8], add=TRUE)
+  par(xpd=NA)
+  legend(-160, -100, vegs_labels[1:4], fill=cols[1:4], bty='n')
+  legend(-40, -100, vegs_labels[5:8], fill=cols[5:8], bty='n')
+}, "output/pdf/WorldMap.pdf", width=10, height=8)
+
+
+
+
+
+
+

### 2014-06-18 R/analysis_utils.R

@@ -56,7 +56,7 @@ studyWithVars <- function(allom, hasVars, returnwhat=c("dataframe","list"),
     message("No datasets have values in all those columns.")
     return(NA)
   } else {
-    message("Your query returned ",sum(ihasvars)," studies.")
+    message("Your query returned ",sum(ihasvars)," studies with n=",sum(sapply(l[ihasvars],nrow))," observations.")
     if(returnwhat=="list"){
       if(complete.cases)warning("complete.cases=TRUE ignored when list is returned.")
       return(l[ihasvars])

### 2014-06-18 R/plotting.R

@@ -196,6 +196,8 @@ prepMapInfo<-function(data, study=NA){
   data  
 }  
 
+
+
 drawWorldPlot  <-  function(data, horlines=TRUE, sizebyn=FALSE, add=FALSE, 
                             pchcol="red", 
                             legend=TRUE){
@@ -215,7 +217,8 @@ drawWorldPlot  <-  function(data, horlines=TRUE, sizebyn=FALSE, add=FALSE,
   lat <- data$latitude[!duplicated(latlon)]
   lon <- data$longitude[!duplicated(latlon)]
   
-  j  <-  !is.na(lat) & !is.na(lon)  
+  j  <-  !is.na(lat) & !is.na(lon)
+  
   # Location only sometimes missing - but lat/lon can still be in dataset anyway.
   # & data$loc != "NA" | is.na(data$loc)
   
@@ -225,7 +228,7 @@ drawWorldPlot  <-  function(data, horlines=TRUE, sizebyn=FALSE, add=FALSE,
   } else {
     
     if(!sizebyn){
-      points(lon,lat, pch=19, col=pchcol, bg=pchcol, cex=0.6)
+      points(lon,lat, pch=19, col=alpha("blue",0.5), cex=1.6)
     } else {
       n <- table(latlon)
       symbols(lon,lat, circles=log10(n), inches=0.1, fg="black", bg=pchcol, add=TRUE)
@@ -246,6 +249,8 @@ drawWorldPlot  <-  function(data, horlines=TRUE, sizebyn=FALSE, add=FALSE,
   }
   
 }
+
+
 
 
 repMissingInfo  <-  function(data){

### 2014-06-18 load.R

@@ -10,3 +10,6 @@ baad <- readRDS("data/baad.rds")$data
 # Add Worldclim MAP and MAT
 # (usecache=FALSE depends on wordclim layers; only works on remko's machine)
 baad <- addWorldClimMAPMAT(baad, usecache=TRUE)
+
+
+cfg <- read.csv("data/variableDefinitions.csv", stringsAsFactors=FALSE)

### 2014-06-19 Rmd/baadplotscollection.Rmd
Added more figures, also Daniel's plots of all data. html compiles fine, many small details fixed.

@@
 ```{r echo=FALSE}
-
 source("load.R")
@@
 pipe_bh <- summarize(group_by(dat, dataset, species), 
-           alf_stbh=mean(a.lf / a.stbh),
-           mlf_stbh=mean(m.lf / a.stbh),
-           sla = mean(a.lf / m.lf),
+           alf_stbh=mean(a.lf / a.stbh, na.rm=TRUE),
+           mlf_stbh=mean(m.lf / a.stbh, na.rm=TRUE),
+           sla = mean(a.lf / m.lf, na.rm=TRUE),
            pft=first(pft),
@@
 datl <- studyWithVars(baad, c("a.stba", "a.lf","m.lf"), returnwhat="list")
-
-dat$pft <- as.factor(dat$pft)
-pipe_bh$pft <- as.factor(pipe_bh$pft)
+dat <- rbind_all(datl)
 
 pipe_ba <- summarize(group_by(dat, dataset, species), 
-                     alf_stba=mean(a.lf / a.stba),
-                     mlf_stba=mean(m.lf / a.stba),
+                     alf_stba=mean(a.lf / a.stba, na.rm=TRUE),
+                     mlf_stba=mean(m.lf / a.stba, na.rm=TRUE),
                      pft=first(pft),
@@
 
+dat$pft <- as.factor(dat$pft)
+pipe_bh$pft <- as.factor(pipe_bh$pft)
+pipe_ba$pft <- as.factor(pipe_ba$pft)
+
 BARplot <- function(x,y,ymult=1.2,...){
 
-  N <- table(x)
+  N <- table(x[!is.na(x)])
   tap <- tapply(y,x,mean,na.rm=T)
@@
   d <- 0.3
-  text(c(0.3, b$xvals+d), max(tap)*(1.1/1.2)*ymult, labels=c("N =", as.character(unname(N))))
+  text(c(0.3, b$xvals+d), max(tap)*(1.1/1.2)*ymult, 
+       labels=c("N =", as.character(unname(N))))
 
@@
                        ))
+
 with(pipe_ba, BARplot(pft,alf_stba, xlab="Plant functional type", 
@@
                           col=Cols))
-b <- with(pipe_ba, BARplot(pft,mlf_stba, xlab="Plant functional type",
+with(pipe_ba, BARplot(pft,mlf_stba, xlab="Plant functional type",
                           ylab=expression(M[F]/A[W]~base~~(kg~m^-2)),
@@
 ))
+
+par(mfrow=c(1,1))
 
@@
 
-allombygroupandpftplot <- function(xvar, yvar, dataset, xlab=AxisLabels[[xvar]], ylab=AxisLabels[[yvar]], 
+allombygroupandpftplot <- function(xvar, yvar, dataset, xlab=AxisLabels[[xvar]], 
+                                   ylab=AxisLabels[[yvar]], 
                                    slopecomp=1, n_min=20,legend=FALSE,
@@
   plot(smfitsDA, add=TRUE, type='l', col=pftcols$DA)
+
+  legend("bottomright", as.expression(bquote(Slope==.(slopecomp))),
+         bty="n", col="grey", cex=0.7)
 
@@
 par(cex.lab=1.2, mar=c(5,5,2,2))
-allombygroupandpftplot("m.to", "m.lf", baad2, legend=T)
-allombygroupandpftplot("m.to", "a.lf", baad2)
+allombygroupandpftplot("m.to", "m.lf", baad2, legend=T, slopecomp=3/4)
+allombygroupandpftplot("m.to", "a.lf", baad2, slopecomp=3/4)
 
-allombygroupandpftplot("m.so", "m.lf", baad2)
-allombygroupandpftplot("m.so", "a.lf", baad2)
+allombygroupandpftplot("m.so", "m.lf", baad2, slopecomp=3/4)
+allombygroupandpftplot("m.so", "a.lf", baad2, slopecomp=3/4)
 
-allombygroupandpftplot("a.stbh", "m.lf", baad2)
-allombygroupandpftplot("a.stba", "m.lf", baad2)
+allombygroupandpftplot("a.stbh", "m.lf", baad2, slopecomp=1)
+allombygroupandpftplot("a.stba", "m.lf", baad2, slopecomp=1)
+
+allombygroupandpftplot("d.bh", "h.t", baad2, slopecomp=1)
+allombygroupandpftplot("d.ba", "h.t", baad2, slopecomp=1)
 
@@
 
+allombygroupandpftplot("c.d", "m.lf", baad2, legend=T, slopecomp=1)
+allombygroupandpftplot("c.d", "d.cr", baad2, legend=T, slopecomp=1)
+
+```
+
+### Root-shoot ratios
+
+Functional balance assumption : does fine root mass scale with leaf mass?
+Total root-shoot ratio; would be nice if we find approximate isometry.
+
+```{r echo=FALSE, warning=FALSE, message=FALSE}
 allombygroupandpftplot("m.so", "m.rt", baad2, legend=T)
 allombygroupandpftplot("m.lf", "m.rf", baad2, legend=T)
-
-allombygroupandpftplot("c.d", "m.lf", baad2, legend=T)
-allombygroupandpftplot("c.d", "d.cr", baad2, legend=T)
-
 ```
@@
 
+### Tropical vs. temperate evergreens
+
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+ii <- which(P$pft == "EA" & P$vegetation %in% c("TempF","TempRF","TropRF","TropSF"))
+l <- l[ii]
+P <- P[ii,]
+
+P$TempTrop <- as.factor(ifelse(P$vegetation %in% c("TempF","TempRF"), "Temp", "Trop"))
+palette(c("blue","red"))
+
+# log-log scale
+plot(1, xlim=c(-3,2), ylim=c(0,1), type='n',
+     xlab="Plant height (m)", ylab="Leaf mass / aboveground mass (-)",
+     axes=FALSE)
+addlogaxes(10^-3, 10^2, 10^-3, 10^0, tck1=-0.03, tck2=-0.015,addY=F)
+axis(2)
+for(i in 1:length(l)){
+  points(log10(l[[i]]$h), l[[i]]$wfrel, type='l',
+         col=P$TempTrop[i])
+}
+legend("bottomleft", c("Temperate evergreen angiosperm",
+                       "Tropical evergreen angiosperm"), fill=palette())
+
+plot(1, xlim=c(-3,2), ylim=c(-7,5), type='n',
+     xlab="Plant height (m)", ylab="Leaf mass (kg)",
+     axes=FALSE)
+magaxis(side=1:2, unlog='xy',box=TRUE)
+for(i in 1:length(l))points(log10(l[[i]]$h), log10(l[[i]]$mlf), type='l',
+                            col=P$TempTrop[i])
+legend("topleft", c("Temperate evergreen angiosperm",
+                       "Tropical evergreen angiosperm"), fill=palette(),inset=0.03)
+
+plot(1, xlim=c(-3,2), ylim=c(-7,5), type='n',
+     xlab="Plant height (m)", ylab="Above-ground woody mass (kg)",
+     axes=FALSE)
+magaxis(side=1:2, unlog='xy',box=TRUE)
+for(i in 1:length(l))points(log10(l[[i]]$h), log10(l[[i]]$mst), type='l',
+                            col=P$TempTrop[i])
+legend("topleft", c("Temperate evergreen angiosperm",
+                       "Tropical evergreen angiosperm"), fill=palette(),inset=0.03)
+
+```
+
 ### MAT and MAP
@@
 ```
+
+### Plots of all data
+
+Something like this for first Figure in manuscript.
+
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+data <- baad
+
+## Fill sapwood area. Let basal=breast height where only data at breast height exists.
+# For small plants treat total stem cross section as sapwood
+data[["sap"]] <- data[["a.ssba"]]
+i <- is.na(data[["sap"]])
+data[["sap"]][i] <- data[["a.ssbh"]][i]
+i <- is.na(data[["sap"]]) & !is.na(data[["a.stba"]]) & !is.na(data[["h.t"]]) & data[["h.t"]] < 2
+data[["sap"]][i] <- data[["a.stba"]][i]
+
+addIsoclines <- function(A= seq(-10,10),b=1,...){
+  for(a in A) abline(a,b,lty="dotted",...)
+}
+
+add.sma.by.group <-function (data, xvar, yvar, smagroup, colorBygroup, col.table,nmin=10,...){
+  if(sum(!is.na(data[[yvar]]*data[[xvar]]))>nmin){
+		col=colour.by.category(data[[colorBygroup]][1],col.table)
+		fit <- sma(data[[yvar]]~data[[xvar]]*data[[smagroup]], log="xy",quiet=TRUE)
+		plot(fit, type='l',col=col,p.lines.transparent=0.1,add=TRUE,...)
+		invisible(fit)
+	}
+}
+
+make.color.table <- function(names){
+	col.table <- niceColors(length(names))
+	names(col.table) <- names
+	col.table
+}
+
+# make a plot, fits lines to each within each group, applyiong same colour within that group
+plot_with_sma_by_var_within_group <- function(data, xvar,yvar,group="vegetation",location=NULL, isoclines=NULL,...){
+
+	plot(data[[xvar]], data[[yvar]],  type= 'n', log="xy", las=1, yaxt="n", xaxt="n",...)
+	axis.log10(1)
+	axis.log10(2)
+
+	if(!is.null(isoclines))
+		addIsoclines(b=isoclines)
+
+	points(data[[xvar]], data[[yvar]],  type= 'p', col = make.transparent("grey", 0.4), pch=16)
+
+	col.table <- make.color.table(unique(data[[group]]))
+
+	fits <- dlply(data, group, add.sma.by.group, xvar=xvar, yvar=yvar, col.table=col.table,
+		smagroup="species",colorBygroup=group)
+
+	get.significant.groups <- function(fit, p=0.1){
+		if(is.null(fit)) return(FALSE)
+		any(fit$groupsummary$pval < p)
+	}
+
+	tmp <- laply(fits, get.significant.groups)
+	lab <- sort(names(tmp)[tmp])
+
+	if(!is.null(location))
+		legend(location, legend = tolower(lab), bty = "n", pch=16, col=colour.by.category(lab,col.table), cex=0.5)
+	fits
+}
+
+myfigure <- function(){
+# 	par(mfrow = c(1,3))
+	fits<-plot_with_sma_by_var_within_group(data,"a.lf", "h.t", "vegetation",
+                                          ylim=c(0.005, 120), xlim=c(1E-5, 1E3), 
+                                          ylab="Height (m)", 
+                                          xlab="",location="topleft", isoclines=0.5, main="a) Architecture")
+	legend("bottomright", legend = "Slope = 0.5", bty = "n", cex=1)
+
+	plot_with_sma_by_var_within_group(data,"a.lf", "sap", "vegetation",xlim=c(1E-5, 1E3), ylim=c(1E-8, 1E0), 
+                                    ylab=expression(paste("Sapwood area (",m^2,")")), 
+                                    xlab=expression(paste("Leaf area (",m^2,")")),location="topleft", 
+                                    isoclines=1, "b) Pipe Model")
+	legend("bottomright", legend = "Slope = 1", bty = "n", cex=1)
+
+	plot_with_sma_by_var_within_group(data,"a.lf", "m.rt", "vegetation",xlim=c(1E-5, 1E3), ylim=c(1E-8, 400), 
+                                    ylab="mass root (kg)", 
+                                    xlab="",location="topleft", isoclines=1, main="c) Roots")
+	legend("bottomright", legend = "Slope = 1", bty = "n", cex=1)
+
+}
+
+myfigure()
+```

### 2014-06-19 R/plot-utils.R

@@ -1,0 +1,130 @@
+
+blank_plot <- function(xlim, ylim,
+  xlab=NULL, ylab=NULL,line=4,
+  xtick=NULL,ytick=NULL,
+  xtick.lab=xtick,ytick.lab=tick,...){
+
+  plot(NA, xlim = xlim, ylim = ylim, xaxt='n', yaxt='n', ann=FALSE,...)
+
+  if(!is.null(xtick))
+   axis(1, at=xtick, labels=xtick.lab)
+
+ if(!is.null(ytick))
+   axis(2, at=ytick, labels=ytick.lab, las=1)
+
+ if(!is.null(xlab))
+  mtext(xlab, 1, line=line)
+
+ if(!is.null(ylab))
+  mtext(ylab, 2, line=line)
+
+}
+
+
+to.dev <- function(expr, dev, filename, ..., verbose=TRUE){
+  if(!file.exists(dirname(filename)))
+    dir.create(dirname(filename), recursive=TRUE)
+  if ( verbose )
+    cat(sprintf("Creating %s\n", filename))
+  dev(filename, ...)
+  on.exit(dev.off())
+  eval.parent(substitute(expr))
+}
+
+to.pdf <- function(expr, filename, ..., verbose=TRUE) {
+  if(!file.exists(dirname(filename)))
+    dir.create(dirname(filename), recursive=TRUE)
+  if ( verbose )
+    cat(sprintf("Creating %s\n", filename))
+  pdf(filename, ...)
+  on.exit(dev.off())
+  eval.parent(substitute(expr))
+}
+
+na.clean <-function(x){x[!is.na(x)]}
+
+#returns vector from lo to hi with multiplication steps of incr. Used for making ticks to a log-scaled axis
+seq.log <- function(from, to, base)
+{base^(log(from,base):log(to,base))}
+
+#returns vector of minor tick spacings approrpaite for log 10 scaled axis with major ticks given by 'major'.
+seq.log.minor<-function(major)
+{
+  temp<-NULL;
+  if(length(major) > 1)
+    for (i in 1:(length(major)-1))
+      temp<-c(temp, seq(major[i], major[i+1]-major[i], major[i]));
+  temp<-c(temp, major[length(major)]);
+  return (temp);
+}
+
+
+
+#returns up to 80 unique, nice colors, generated using http://tools.medialab.sciences-po.fr/iwanthue/
[…]

### 2014-06-19 R/plotting.R

@@ -68,12 +68,14 @@ getLabel <- function(xvar){
 
 highlightStudies <- function(data, xvar, yvar, group, studies, col = niceColors(length(studies)), type='b', pch=1){
   
-  bivarPlotColorBy(data, xvar, yvar, group =group, col=make.transparent("grey", 0.5), legend= FALSE, type = 'p', pch=pch)
+  bivarPlotColorBy(data, xvar, yvar, group =group, col=make.transparent("grey", 0.5), 
+                   legend= FALSE, type = 'p', pch=pch)
   
   for(i  in seq_len(length(studies))){
     ii <- data$dataset==studies[i]
     if(sum(ii) > 0)
-      bivarPlotColorBy(data[ii,], xvar, yvar, group = group, add = TRUE, legend = FALSE, col=col[i], type=type, pch=pch)  
+      bivarPlotColorBy(data[ii,], xvar, yvar, group = group, add = TRUE, 
+                       legend = FALSE, col=col[i], type=type, pch=pch)  
   }
   
   leg <- data.frame(group = studies, col = col, stringsAsFactors=FALSE)
@@ -99,6 +101,7 @@ bivarPlotColorBy <- function(data, xvar, yvar, group, type='b',
   
   i <- order(colorBy, decreasing = TRUE)
   
+  
   bivarPlot(data[i,], xvar, yvar, xlab =getLabel(xvar), 
             ylab = getLabel(yvar), col= colours[i], 
             add=add, type=type, pch=pch, ...)
@@ -116,13 +119,13 @@ bivarPlotColorBy <- function(data, xvar, yvar, group, type='b',
      fit <- add.sma(data, xvar, yvar, colorBy, 
                     col=out$col, from=from, to=to,add=TRUE,...)
   
-  invisible(list(colours=out, fit=fit))
+  return(invisible(list(colours=out, fit=fit)))
 }    
 
 add.sma <-function (data, xvar, yvar, colorBy, col, from=NA, to=NA,...){
 
   i <- findPositive(data, xvar, yvar)
-  fit <- sma(data[i,yvar]~data[i,xvar]*colorBy[i], log="xy")
+  fit <- sma(data[i,yvar]~data[i,xvar]*colorBy[i], log="xy", quiet=TRUE)
   plot(fit, type='l',col=col,p.lines.transparent=0.1,...)
   invisible(fit)
 } 

### 2014-06-19 analysis.R

@@ -1,4 +1,5 @@
 
+#file.copy("c:/repos/baad/output/baad.rds", "c:/repos/baadanalysis/data", overwrite=T)
 source("load.R")
 
 # Reports, scripts, etc.

### 2014-06-19 Rmd/baadplotscollection.Rmd
now works with new baad (where dataset was renamed to studyName)

@@
 # Roth2007 massive outlier, remove here
-dat <- subset(dat, dataset != "Roth2007")
+dat <- subset(dat, studyName != "Roth2007")
 
@@
 
-pipe_bh <- summarize(group_by(dat, dataset, species), 
+pipe_bh <- summarize(group_by(dat, studyName, species), 
            alf_stbh=mean(a.lf / a.stbh, na.rm=TRUE),
@@
 
-pipe_ba <- summarize(group_by(dat, dataset, species), 
+pipe_ba <- summarize(group_by(dat, studyName, species), 
                      alf_stba=mean(a.lf / a.stba, na.rm=TRUE),
@@
 # baad2 : discard pot-grown plants
-baad$Group <- paste(baad$dataset, baad$speciesMatched)
+baad$Group <- paste(baad$studyName, baad$speciesMatched)
 baad2 <- subset(baad, growingCondition %in% c("FW","PM","PU","FE"))
@@
 
-allombygroupandpftplot <- function(xvar, yvar, dataset, xlab=AxisLabels[[xvar]], 
+allombygroupandpftplot <- function(xvar, yvar, studyName, xlab=AxisLabels[[xvar]], 
                                    ylab=AxisLabels[[yvar]], 
@@
 
-  dataset$X <- dataset[,xvar]
-  dataset$Y <- dataset[,yvar]
+  studyName$X <- studyName[,xvar]
+  studyName$Y <- studyName[,yvar]
 
-  smfitsEA <- sma(Y ~ X*Group, data=subset(dataset,pft=="EA"), log='xy', n_min=n_min, quiet=TRUE)
-  smfitsEG <- sma(Y ~ X*Group, data=subset(dataset,pft=="EG"), log='xy', n_min=n_min, quiet=TRUE)
-  smfitsDA <- sma(Y ~ X*Group, data=subset(dataset,pft=="DA"), log='xy', n_min=n_min, quiet=TRUE)
-  smfits <- sma(Y ~ X*Group, data=dataset, log='xy', n_min=n_min, quiet=TRUE)
+  smfitsEA <- sma(Y ~ X*Group, data=subset(studyName,pft=="EA"), log='xy', n_min=n_min, quiet=TRUE)
+  smfitsEG <- sma(Y ~ X*Group, data=subset(studyName,pft=="EG"), log='xy', n_min=n_min, quiet=TRUE)
+  smfitsDA <- sma(Y ~ X*Group, data=subset(studyName,pft=="DA"), log='xy', n_min=n_min, quiet=TRUE)
+  smfits <- sma(Y ~ X*Group, data=studyName, log='xy', n_min=n_min, quiet=TRUE)
 
@@
 
-The following uses only dataset/species combinations that have at least 20 observations.
+The following uses only studyName/species combinations that have at least 20 observations.
 Using `sma`, I fit leaf mass vs.
@@
 x <- studyWithVars(baad, c("m.st","m.lf","h.t"))
-x$dataset_species <- paste0(x$dataset,"_",x$species)
+x$dataset_species <- paste0(x$studyName,"_",x$species)
 
@@
 # Add some metadata
-dfr <- baad[,c("dataset","species","pft","vegetation","growingCondition")]
-dfr$dataset_species <- with(dfr, paste0(dataset,"_",species))
+dfr <- baad[,c("studyName","species","pft","vegetation","growingCondition")]
+dfr$dataset_species <- with(dfr, paste0(studyName,"_",species))
 dfr <- dfr[!duplicated(dfr$dataset_species),]
@@
 
-# nonfield <- union(which(P$vegetation == "GH"), which(P$dataset == "Poorter1999"))
 nonfield <- which(P$growingCondition %in% c("CG","GH"))
@@
   i <- !is.na(data[[var]])
-  data[c("dataset", var)][!duplicated(paste( data[["dataset"]], data[[var]])),]
+  data[c("studyName", var)][!duplicated(paste( data[["studyName"]], data[[var]])),]
 }

### 2014-06-19 R/analysis_utils.R

@@ -48,7 +48,7 @@ studyWithVars <- function(allom, hasVars, returnwhat=c("dataframe","list"),
   r <- require(plyr)
   if(!r)stop("Install plyr package first.")
   returnwhat <- match.arg(returnwhat)
-  l <- split(allom, allom$dataset)
+  l <- split(allom, allom$studyName)
   l <- lapply(l, removeNAcols)
   ihasvars <- sapply(l, function(x) all(hasVars %in% names(x)))
     

### 2014-06-19 R/plotting.R

@@ -22,7 +22,7 @@ makePlotPanel<-function(data, study, dir="report-per-study", col="grey", pdf=TRU
     pdf(file=paste0(path,"/", study,".pdf"))
   }
   
-  dat        <-  data[data$dataset==study,]
+  dat        <-  data[data$studyName==study,]
   plot.vars  <-  mashrDetail("var.def")$Variable[mashrDetail("var.def")$Group=="tree"]
   plot.vars  <-  plot.vars[plot.vars %in% c("growingCondition","status","light") == FALSE]
   available  <-  dat[,names(dat) %in% plot.vars]   
@@ -72,7 +72,7 @@ highlightStudies <- function(data, xvar, yvar, group, studies, col = niceColors(
                    legend= FALSE, type = 'p', pch=pch)
   
   for(i  in seq_len(length(studies))){
-    ii <- data$dataset==studies[i]
+    ii <- data$studyName==studies[i]
     if(sum(ii) > 0)
       bivarPlotColorBy(data[ii,], xvar, yvar, group = group, add = TRUE, 
                        legend = FALSE, col=col[i], type=type, pch=pch)  
@@ -165,7 +165,7 @@ bivarPlot <- function(data, xvar, yvar, xlab=xvar, ylab=yvar,
 }  
 
 whichStudies <- function(alldata, var, value){
-  unique(alldata$data$dataset[alldata$data[[var]] == value])  
+  unique(alldata$data$studyName[alldata$data[[var]] == value])  
 }
 
 makePlot <-function(data, subset, xvar, yvar, xlab, ylab, main="", maincol=make.transparent("grey", 0.5), studycol = "red",pch=19){
@@ -183,12 +183,12 @@ makePlot <-function(data, subset, xvar, yvar, xlab, ylab, main="", maincol=make.
 prepMapInfo<-function(data, study=NA){
   
   if(!is.na(study))
-    data   <-  data[data$dataset %in% study,]
+    data   <-  data[data$studyName %in% study,]
   
   #Remove duplicate locations
-  keep <- c("dataset", "latitude", "longitude", "location") 
+  keep <- c("studyName", "latitude", "longitude", "location") 
   
-  data <- data[!duplicated(paste0(data$dataset,";", data$latitude,";", data$longitude, ";", data$location)), keep]
+  data <- data[!duplicated(paste0(data$studyName,";", data$latitude,";", data$longitude, ";", data$location)), keep]
   
[…]

### 2014-06-19 Rmd/baadplotscollection.Rmd
updated plots of pipe model ratios vs. SLA

@@
 ))
+```
 
+It seems that *across* PFT's, leaf area to stem cross-sectional area ('area pipe ratio') is more conserved than the leaf mass per unit cross-sectional area ('mass pipe ratio').
+Conifers have a much higher leaf mass supported, but end up with similar leaf area.
+
+Within PFT's, the area pipe model ratio is correlated with SLA, but not for the mass pipe ratio.
+
+```{r echo=FALSE, message=FALSE, warning=FALSE}
 par(mfrow=c(1,1))
+smfit <- sma(alf_stbh ~ sla*pft, log="xy", data=pipe_bh, quiet=TRUE)
 
-with(pipe_bh, plot(sla, alf_stbh, pch=19, col=pft, xlim=c(0,50),
+plot(smfit,  type='o', pch=19, axes=F,
      xlab=expression(Specific~leaf~area~~(m^2~kg^-1)),
-                   ylab=expression(A[L]/A[W]~breast~height~~(m^2~m^-2))))
-legend("topright", levels(pipe_bh$pft), col=palette(), pch=19)
+     ylab=expression(A[L]/A[W]~breast~height~~(m^2~m^-2)))
+magaxis(side=1:2, unlog=1:2)
+legend("topleft", levels(pipe_bh$pft), col=palette(), pch=19)
 
-with(pipe_bh, plot(sla, mlf_stbh, pch=19, col=pft, xlim=c(0,50),
+smfit2 <- sma(mlf_stbh ~ sla*pft, log="xy", data=pipe_bh, quiet=TRUE)
+plot(smfit2,  type='p', pch=19, axes=F,
      xlab=expression(Specific~leaf~area~~(m^2~kg^-1)),
-                   ylab=expression(M[F]/A[W]~breast~height~~(m^2~m^-2))))
-legend("topright", levels(pipe_bh$pft), col=palette(), pch=19)
-
+     ylab=expression(M[F]/A[W]~breast~height~~(m^2~m^-2)))
+magaxis(side=1:2, unlog=1:2)
+legend("bottomleft", levels(pipe_bh$pft), col=palette(), pch=19)
 ```
@@
 
-  smfitsEA <- sma(Y ~ X*Group, data=subset(studyName,pft=="EA"), log='xy', n_min=n_min, quiet=TRUE)
-  smfitsEG <- sma(Y ~ X*Group, data=subset(studyName,pft=="EG"), log='xy', n_min=n_min, quiet=TRUE)
-  smfitsDA <- sma(Y ~ X*Group, data=subset(studyName,pft=="DA"), log='xy', n_min=n_min, quiet=TRUE)
+  smfitsEA <- sma(Y ~ X*Group, data=subset(studyName,pft=="EA"), 
+                  log='xy', n_min=n_min, quiet=TRUE)
+  smfitsEG <- sma(Y ~ X*Group, data=subset(studyName,pft=="EG"), 
+                  log='xy', n_min=n_min, quiet=TRUE)
+  smfitsDA <- sma(Y ~ X*Group, data=subset(studyName,pft=="DA"), 
+                  log='xy', n_min=n_min, quiet=TRUE)
   smfits <- sma(Y ~ X*Group, data=studyName, log='xy', n_min=n_min, quiet=TRUE)
@@
   legend("bottomright", as.expression(bquote(Slope==.(slopecomp))),
-         bty="n", col="grey", cex=0.7)
+         col="grey", cex=0.7)
 
-  if(legend)legend("topleft", c("Evergreen Angio.","Evergreen Gymno", "Deciduous Angio."), 
-                   fill=unlist(pftcols), cex=0.9)  
+  if(legend)legend("topleft", 
+                   c("Evergreen Angio.","Evergreen Gymno", "Deciduous Angio."), 
+                   fill=unlist(pftcols), cex=0.9,
+                   , bg="white", box.col="white", inset=0.01)
 return(invisible(smfits))

### 2014-07-25 Rmd/baadplotscollection.Rmd
more analysis - very drafty!

@@
 
-allombygroupandpftplot <- function(xvar, yvar, studyName, xlab=AxisLabels[[xvar]], 
+# basal area times height
+baad2$a.stbh_h.t <- with(baad2, h.t * a.stbh)
+
+allombygroupandpftplot <- function(xvar, yvar, data, xlab=AxisLabels[[xvar]], 
                                    ylab=AxisLabels[[yvar]], 
@@
 
-  studyName$X <- studyName[,xvar]
-  studyName$Y <- studyName[,yvar]
+  data$X <- data[,xvar]
+  data$Y <- data[,yvar]
 
-  smfitsEA <- sma(Y ~ X*Group, data=subset(studyName,pft=="EA"), 
+  smfitsEA <- sma(Y ~ X*Group, data=subset(data,pft=="EA"), 
                   log='xy', n_min=n_min, quiet=TRUE)
-  smfitsEG <- sma(Y ~ X*Group, data=subset(studyName,pft=="EG"), 
+  smfitsEG <- sma(Y ~ X*Group, data=subset(data,pft=="EG"), 
                   log='xy', n_min=n_min, quiet=TRUE)
-  smfitsDA <- sma(Y ~ X*Group, data=subset(studyName,pft=="DA"), 
+
+  smfitsDA <- sma(Y ~ X*Group, data=subset(data,pft=="DA"), 
                   log='xy', n_min=n_min, quiet=TRUE)
-  smfits <- sma(Y ~ X*Group, data=studyName, log='xy', n_min=n_min, quiet=TRUE)
+
+  smfits <- sma(Y ~ X*Group, data=data, log='xy', n_min=n_min, quiet=TRUE)
 
@@
                    m.so = "Above-ground plant mass (kg)",
+                   m.st = "Above-ground wood mass (kg)",
                    m.lf = "Total leaf mass (kg)",
@@
                    c.d = "Crown length (m)",
-                   d.cr = "Crown diameter (m)")
+                   d.cr = "Crown diameter (m)",
+                   a.stbh_h.t = expression("Stem area * height"~(m^3)))
 
@@
 
+#allombygroupandpftplot("a.stbh_h.t", "m.st", baad2, slopecomp=1)
+
 allombygroupandpftplot("h.t", "m.so", baad2, slopecomp=2, legend=T)
@@
 allombygroupandpftplot("c.d", "d.cr", baad2, legend=T, slopecomp=1)
+
+allombygroupandpftplot("h.t", "m.st", baad2, legend=T, slopecomp=3)
+allombygroupandpftplot("a.stbh", "m.st", baad2, legend=T, slopecomp=2)
+allombygroupandpftplot("a.stbh", "h.t", baad2, legend=T, slopecomp=1/3)
+
+```
+
+```{r eval=FALSE}
+df <- studyWithVars(baad2, c("m.st","a.stbh_h.t"))
+x <- lapply(split(df,df$Group), function(x)try(lm(log10(m.st) ~ log10(a.stbh_h.t),data=x)))
+
+getR2 <- function(x){
+  if(inherits(x,"try-error"))
+    NA
+  else
+    summary(x)$r.squared
+}
 

### 2014-08-01 manuscript/draftpaper.tex
more on predBarPlot, beginning of tex manuscript with ideas

\documentclass[a4paper]{article}

\usepackage{amssymb,amsmath}
\usepackage{palatino}
\usepackage{framed}
\usepackage{parskip}
\usepackage{geometry}
\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}

\begin{document}

\title{Biomass allocation, plant functional types and leaf area vs. mass}

\author{Remko Duursma, Daniel Falster and BAAD data contributors}

\maketitle

%--------------------------------------------------------------------------------------------%

\section{Main text}

\begin{enumerate}
  \item Importance of biomass allocation in earth system models, feedback on fluxes of CO2, lifetime of C in the system.
  \item Lots of approaches to model biomass allocation, these models tend to settle on very simple algorithms.
  \item Little is known about differences between plant functional types, which are accounted for in many ways in these models already, especially via physiology etc. etc. 
  \item We now have a dataset that we can use to evaluate differences between plant functional types.
  \item Important constraints on allocation are a) leaf mass and leaf area fraction, b) root-shoot ratio and c) leaf mass and area per unit sapwood area.
  \item Many models allocate some percentage of NPP to each biomass pool, and then multiply this by SLA to get leaf area, which is the area relevant for fluxes of CO2, H2O and light interception.
  \item Because PFTs differ so much in SLA, and little is known about fNPP to each pool by pft, this would lead to \emph{certain pft differences}.
  \item Here we show that, in fact, it is far more likely that trees vary allocation to end up with a given leaf area / sapwood area ratio, or equivalently, a given leaf area ratio. These fractions, when expressed as mass fraction were far more variable between pft's.
  \item These results major implications evaluation and development of allocation models employed in earth system models.
\end{enumerate}

\section{Methods}

\end{document}

### 2014-08-05 manuscript/draftpaper.tex
notes in manuscript.tex

@@
 
+Friend et al.
+(2014) showed that 30\% of the uncertainty in global vegetation carbon simulated with seven vegetation models is due to the residence time of carbon in vegetation.
+This residence time depends on turnover rates of the various plant pools, and hence also on the distribution of biomass between these pools.
+For example, vegetation with a higher leaf mass fraction will have a lower overall residence time of carbon.
+
+De Kauwe et al.
+(2014) : "We have shown that allocation approaches that are constrained by biomass fractions (such as functional relationships) were more successful at capturing observed trends, and were generally more robust, than approaches based on allocation coefficients."
+
+Poorter et al.
+(2012) found a slightly larger LMF in gymnosperms, but the difference was small and could not split between deciduous and evergreen angiosperms due to a lack of data.
+They argue that "most conifers retain leaves for two or more years.
+Thus, the higher LMF may simply be the consequence of leaf longevity rather than higher assimilate partitioning per se.
+In this way, conifers are able to compensate for their lower physiological activity per gram of leaf." This raises the question whether LMF differs by angiosperms and gymnosperms, or as a consequence of leaf longevity (especially deciduous vs.
+evergreen).
+
+Serious lack of biomass data to constrain models.
+Studies like Wolf et al.
+(2011) resort to studying individual tree allometry based on dividing stand biomass by stand density.
+Wolf et al.
+(2010) did not find any differences in biomass fractions (root,stem,leaf) between angiosperms and gymnosperms (but again this was based on stand biomass data, but still this is strange!).
+
 \section{Methods}

### 2014-08-06 manuscript/draftpaper.tex
added better SLA - pipe model figures, and glopnet data

@@
 
+\section{Results}
+(this will not be separate in the manuscript, just temporary notes here)
+
+Figure (figure\_mlfastbh\_sla\_pft.pdf) shows that across plant functional types (pfts) (but not within pfts), the ratio of leaf mass to stem cross-sectional area is negatively correlated with specific leaf area.
+This means that pfts with larger SLA display less leaf mass per unit stem area (and hence per unit aboveground biomass), in a way so that leaf area per stem cross sectional area does not differ by pft (figure\_alfastbh\_sla\_pft.pdf).
+The convergent trait, independent of plant functional type, is then the ratio of leaf area to stem area.
+
+Though leaf area per stem area did not differ between pfts (figure\_alfastbh\_sla\_pft and other figure tba), within pfts it was correlated to sla.
+In other words, for a given pft, species with higher sla tended to have larger total leaf area at a given stem cross-sectional area.
+This was in a way so that leaf mass per unit stem area did not correlate to sla within a pft, so that within pfts the more convergent trait is in fact leaf mass per unit stem area.
+
+The implication for global vegetation models that use pft as the unit of parameterization is that specific leaf area (pft-specific) should be combined with a target leaf area to stem area ratio to give a leaf mass fraction that constrains the allocation of carbon.
+This is what LPJ does, but there the pipe model ratio is actually pft-specific (and nobody knows where it comes from; see sitch and also Hickler).
+
+Fixed allocation fraction models would get the distribution of biomass right if leaf turnover rate is exactly proportional to specific leaf area (because if an extra unit of leaf area is produced when SLA is increased, this could be offset exactly by a faster turnover rate).
+However glopnet says that turnover rate scales with more than SLA (exponent 1.3, but not a great fit).
+
 \section{Methods}

### 2014-08-06 R/figure_pipemodel_sla_pft.R

@@ -1,0 +1,50 @@
+predictit <- function(model, varname, dataset){
+  
+  m <- dataset[,c("Group","pft")]
+  m <- m[!duplicated(m),]
+  
+  newdat <- with(dataset, expand.grid(Group=rownames(ranef(model)$Group),
+                                      lh.t=log10(10)))
+  newdat <- merge(newdat, m, by="Group", all=FALSE)
+  y <- 10^predict(model, newdat, re.form=NULL)
+  newdat[,varname] <- y
+  return(newdat)
+}
+
+
+fit1 <- lmer(lalf_astbh ~ pft + lh.t + pft:lh.t + (lh.t|Group), data=dataset)
+fit1b <- lmer(lmlf_astbh ~ pft + lh.t + pft:lh.t + (lh.t|Group), data=dataset)
+fit2 <- lmer(log10(a.lf/m.lf) ~ pft + lh.t + pft:lh.t + (lh.t|Group), data=dataset)
+
+df1 <- predictit(fit1, "alf_astbh", dataset)
+df2 <- predictit(fit2, "sla", dataset)[,c("Group","sla")]
+df <- merge(df1, df2, by="Group")
+df3 <- predictit(fit1b, "mlf_astbh", dataset)[,c("Group","mlf_astbh")]
+df <- merge(df, df3, by="Group")
+
+smfit <- sma(alf_astbh ~ sla*pft, log="xy", data=df, quiet=TRUE)
+smfit2 <- sma(mlf_astbh ~ sla*pft, log="xy", data=df, quiet=TRUE)
+
+Cols <- c("blue","red","darkorange")
+palette(Cols)
+
+to.pdf({
+  plot(smfit,  type='o', pch=19, axes=F,
+       ylim=c(100,15000),xlim=c(1,50),
+       xlab=expression(Specific~leaf~area~~(m^2~kg^-1)),
+       ylab=expression(A[L]/A[W]~breast~height~~(m^2~m^-2)))
+  magaxis(side=1:2, unlog=1:2)
+  legend("topleft", levels(df$pft), col=palette(), pch=19, cex=0.8, bty='n')
+}, filename="output/figures/figure_alfastbh_sla_pft.pdf")
+
+
+to.pdf({
+  plot(smfit2,  type='p', pch=19, axes=F,
+       ylim=c(10, 1000),xlim=c(1,50),
+       xlab=expression(Specific~leaf~area~~(m^2~kg^-1)),
[…]

### 2014-08-06 R/lme4_analysis.R

@@ -132,6 +132,15 @@ fit7 <- lmer(la.lf ~ pft + lh.t + pft:lh.t + (1|Group), data=dataset)
 plotPredBarplot(fit7, ylab="a.lf")
 
 
+
+
+
+
+
+#-----------------------------------------------------------------------------------------------#
+
+
+
 #- boxplots of variation between species, instead of just fixed effects variances.
 #- hoping that this would still show obvious differences between groups, but with overlap.
 
@@ -143,9 +152,74 @@ newdat <- with(dataset, expand.grid(pft=unique(pft),
                                     Group=rownames(ranef(model)$Group),
                                     lh.t=log10(Hs)))
 
-P <- bootMer(fit5, nsim=25, FUN=function(.)predict(., newdat, re.form=~(1|Group)))
-sims <-  10^as.data.frame(P)
-newdat$ypred <- 10^predict(model, newdat, re.form=NULL)
+# P <- bootMer(fit5, nsim=25, FUN=function(.)predict(., newdat, re.form=~(1|Group)))
+# sims <-  10^as.data.frame(P)
+
+
+predictit <- function(model, varname, dataset){
+  
+  m <- dataset[,c("Group","pft")]
+  m <- m[!duplicated(m),]
+  
+  newdat <- with(dataset, expand.grid(Group=rownames(ranef(model)$Group),
+                                      lh.t=log10(10)))
+  newdat <- merge(newdat, m, by="Group", all=FALSE)
+  y <- 10^predict(model, newdat, re.form=NULL)
+  newdat[,varname] <- y
+return(newdat)
+}
+
+
+fit1 <- lmer(lalf_astbh ~ pft + lh.t + pft:lh.t + (lh.t|Group), data=dataset)
+fit1b <- lmer(lmlf_astbh ~ pft + lh.t + pft:lh.t + (lh.t|Group), data=dataset)
+fit2 <- lmer(log10(a.lf/m.lf) ~ pft + lh.t + pft:lh.t + (lh.t|Group), data=dataset)
+
+df1 <- predictit(fit1, "alf_astbh", dataset)
+df2 <- predictit(fit2, "sla", dataset)[,c("Group","sla")]
+df <- merge(df1, df2, by="Group")
+df3 <- predictit(fit1b, "mlf_astbh", dataset)[,c("Group","mlf_astbh")]
[…]

### 2014-08-07 manuscript/draftpaper.tex
more outline notes

@@
 
-\title{Biomass allocation, plant functional types and leaf area vs. mass}
+\title{Leaf to stem biomass partitioning is explained by specific leaf area}
 
@@
 \begin{enumerate}
-  \item Importance of biomass allocation in earth system models, feedback on fluxes of CO2, lifetime of C in the system.
-  \item Lots of approaches to model biomass allocation, these models tend to settle on very simple algorithms.
-  \item Little is known about differences between plant functional types, which are accounted for in many ways in these models already, especially via physiology etc. etc. 
-  \item We now have a dataset that we can use to evaluate differences between plant functional types.
-  \item Important constraints on allocation are a) leaf mass and leaf area fraction, b) root-shoot ratio and c) leaf mass and area per unit sapwood area.
-  \item Many models allocate some percentage of NPP to each biomass pool, and then multiply this by SLA to get leaf area, which is the area relevant for fluxes of CO2, H2O and light interception.
-  \item Because PFTs differ so much in SLA, and little is known about fNPP to each pool by pft, this would lead to \emph{certain pft differences}.
-  \item Here we show that, in fact, it is far more likely that trees vary allocation to end up with a given leaf area / sapwood area ratio, or equivalently, a given leaf area ratio.
-These fractions, when expressed as mass fraction were far more variable between pft's.
-  \item These results major implications evaluation and development of allocation models employed in earth system models.
+  \item Importance of biomass allocation in global vegetation models, feedback on fluxes of CO2, lifetime of C in the system. 
+  \item Close relationship to allometric scaling models, which provide an important constraint on the allocation of biomass into the various plant compartments (roots, leaves and stems).
+  \item Previous theoretical work on allometric scaling has predominantly focused on the universal nature of size-biomass relationships in plants.
+This side-steps the known difference in biomass fractions at a given plant size between plant functional types (Enquist, Poorter).
+  \item There is no shortage of theoretical developments in understanding biomass allocation (Franklin).
+Other theoretical work focusing on optimal partitioning theory is promising but is yet to lead to operational models for dynamic global vegetation models.
+  \item These DGVMs typically are parameterized by plant functional type (of woody vegetation types that we here allude to include deciduous, evergreen, angiosperm, gymnosperm), current knowledge of allometry differences between pfts is very weak.
+  \item Whatever the way forward in allocation, what is lacking is a solid empirical base that can be used to constrain any model of allocation, that includes woody plants of all sizes, of all major plant functional types and biomes.
+  \item Here we use such a database (Falster et al.) to evaluate whether important differences exist in biomass partitioning between four major woody plant functional types.
+  \item We find that specific leaf area is the major determinant of the partitioning of biomass between leaves and stems.
+After correcting for important size effects, the ratio of leaf mass to total aboveground biomass is varied in such a way to keep the ratio of leaf area to aboveground biomass invariable between plant functional types.
+  \item As a first approximation, which is useful for large scale DGVMs, models should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly.
+Specific leaf area then gives the ratio of leaf biomass to stem biomass.
+This follows because aboveground woody biomass scales closely with stem cross-sectional area and plant height, in a way that was found to not vary between pft's (*****).
+
 \end{enumerate}
+
+\section{Lit notes}
 
@@
 
-\section{Results}
+\section{Results and Discussion}
 (this will not be separate in the manuscript, just temporary notes here)

### 2014-08-08 manuscript/draftpaper.tex
more progress ....

@@
 
-\title{Leaf to stem biomass partitioning is explained by specific leaf area}
+\title{Specific leaf area explains difference in biomass partitioning between woody plant functional types}
 
@@
 
+\section{Analyses}
+
+A list of analyses to be included, either in the main text or as supporting information. We need to show all these bits to support the storyline.
+
+\begin{enumerate}
+  \item Leaf mass vs. aboveground biomass; line per species/studyName, colored by pft.
+  \item Leaf mass / aboveground biomass vs. plant height; line per species/studyName, colored by pft.
+  \item Barplot leaf mass / aboveground biomass at reference height with bootstrap CI, by pft (and vegetation type)
+  \item Barplot leaf area / aboveground biomass at reference height with bootstrap CI, by pft (and vegetation type)
+  \item Same ones but for leaf mass or leaf area vs. stem area (at BH or base).
+  \item Ratio mass / stem area vs. SLA, ratio area / stem area vs. SLA, as estimated from random effects of fitted lme model.
+  \item Root - shoot ratio, and barplots of pft/vegetation averages with bootstrap CI.
+  \item Histogram of estimated effects at a reference height, by plant functional type, showing the variation between species. 
+\end{enumerate}
+
 \section{Methods}
 
+\begin{itemize}
+  \item Removed deciduous gymnosperms, we have a few of those in BAAD (6 species / studyName combinations), but only one includes estimates of leaf area and mass.
+We did not attempt here to include SLA from another database (* although we can do this, we need to decide!).
+\end{itemize}
+
 \end{document}

### 2014-08-08 R/figure_predbarplots.R

@@ -1,0 +1,76 @@
+
+source("R/prepareDataset.R")
+
+plotPredBarplot <- function(model, ylab="", Hs=c(2,15), nboot=100,...){
+  
+  
+  newdat <- with(dataset, expand.grid(pft=unique(dataset$pft), 
+                                      lh.t=log10(Hs)))
+  P <- bootMer(model, nsim=nboot, FUN=function(.)predict(., newdat, re.form=NA))
+  sims <-  10^as.data.frame(P)
+  
+  newdat$ypred <- 10^predict(model, newdat, re.form=NA)
+  
+  m <- matrix(newdat$ypred, ncol=3, byrow=TRUE)
+  colnames(m) <- unique(dataset$pft)
+  
+  qu <- apply(sims, 2, quantile, probs=c(0.025,0.975))
+  
+  b <- barplot(m, beside=TRUE, ylim=c(0,max(qu[2,])),ylab=ylab,...)
+  f <- function(x)as.vector(matrix(x,ncol=3,byrow=T))
+  arrows(x0=as.vector(b),
+         y0=f(qu[1,]),
+         y1=f(qu[2,]),
+         angle=90,code=3,length=0.1)
+}
+
+
+fit5 <- lmer(lmlf_astbh ~ pft + lh.t + pft:lh.t + (1|Group), data=dataset)
+plotPredBarplot(fit5, ylab="m.lf / a.stbh")
+
+fit6 <- lmer(lalf_astbh ~ pft + lh.t + pft:lh.t + (1|Group), data=dataset)
+plotPredBarplot(fit6, ylab="a.lf / a.stbh")
+
+fit8 <- lmer(lmlf_mso ~ pft + lh.t + pft:lh.t + (1|Group), data=dataset)
+plotPredBarplot(fit8, ylab="m.lf / m.so")
+
+fit9 <- lmer(lalf_mso ~ pft + lh.t + pft:lh.t + (1|Group), data=dataset)
+plotPredBarplot(fit9, ylab="a.lf / m.so")
+
+fit10 <- lmer(lmrt_mso ~ pft + lh.t + pft:lh.t + (1|Group), data=dataset)
+plotPredBarplot(fit10, ylab="m.rt / m.so")
+
+
+
+#-------------------------------------------------------------------------------------------#
+
+plotPredBarplot2 <- function(model, ylab="", Hs=c(2,15), nboot=100,...){
+  
+  
[…]

### 2014-08-08 R/figure_pipemodel_sla_pft.R
more progress ....

@@ -1,26 +1,38 @@
-predictit <- function(model, varname, dataset){
+
+source("R/prepareDataset.R")
+
+
+# Predict from a fitted mixed effects model at some reference height (or multiple reference heights!).
+predictit <- function(model, varname, dataset, refheight=10){
   
+  # Group - pft combinations in the data.
   m <- dataset[,c("Group","pft")]
   m <- m[!duplicated(m),]
   
   newdat <- with(dataset, expand.grid(Group=rownames(ranef(model)$Group),
-                                      lh.t=log10(10)))
+                                      lh.t=log10(refheight)))
+  
+  # Avoid combinations of Group and pft that don't exist in the data
   newdat <- merge(newdat, m, by="Group", all=FALSE)
+  
+  # predict from fitted model and backtransform.
   y <- 10^predict(model, newdat, re.form=NULL)
+  
   newdat[,varname] <- y
   return(newdat)
 }
 
-
+# Three fits; 2 pipe model relationships and one sla model.
 fit1 <- lmer(lalf_astbh ~ pft + lh.t + pft:lh.t + (lh.t|Group), data=dataset)
 fit1b <- lmer(lmlf_astbh ~ pft + lh.t + pft:lh.t + (lh.t|Group), data=dataset)
 fit2 <- lmer(log10(a.lf/m.lf) ~ pft + lh.t + pft:lh.t + (lh.t|Group), data=dataset)
 
+# Make dataframe with sla, and mass or area-based pipe model relationships.
 df1 <- predictit(fit1, "alf_astbh", dataset)
 df2 <- predictit(fit2, "sla", dataset)[,c("Group","sla")]
-df <- merge(df1, df2, by="Group")
-df3 <- predictit(fit1b, "mlf_astbh", dataset)[,c("Group","mlf_astbh")]
-df <- merge(df, df3, by="Group")
+df <- merge(df1, df2, by="Group", all=TRUE)
+df3 <- predictit(fit1b, "mlf_astbh", dataset)[,c("Group","mlf_astbh", "pft")]
+df <- merge(df, df3, by=c("Group","pft"), all=TRUE)
 
 smfit <- sma(alf_astbh ~ sla*pft, log="xy", data=df, quiet=TRUE)
 smfit2 <- sma(mlf_astbh ~ sla*pft, log="xy", data=df, quiet=TRUE)
[…]

### 2014-08-08 R/prepareDataset.R

@@ -28,3 +28,25 @@ dataset$lmrt_mso <- with(dataset, log10(m.rt / m.so))
 # Colours
 pftcols <- list(EA="red", EG="hotpink", DA="blue", DG = "skyblue2")
 
+
+# Dataset with simplified vegetation types, tossing ones that don't easily fit in temperate/boreal/tropical classes.
+dataset2 <- droplevels(subset(dataset, vegetation %in% c("BorF","TempF","TempRF","TropRF","TropSF")))
+
+sw <- function(type){
+  switch(type,
+                               BorF = "boreal",
+                               TempF = "temperate",
+                               TempRF = "temperate",
+                               TropRF = "tropical",
+                               TropSF = "tropical"
+                               )
+}
+dataset2$bortemptrop <- as.vector(sapply(dataset2$vegetation, sw))
+
+
+
+
+
+
+
+

### 2014-09-05 Rmd/lmfwithgam.Rmd
Redid LMF, LAR, pipemodel analysis with a gam (similar results)

---
title: "LMR with gam"
author: "Remko Duursma"
date: "Friday, September 05, 2014"
output: html_document
---

```{r init, echo=FALSE}
library(knitr)
opts_knit$set(root.dir=normalizePath('../'))
opts_chunk$set(echo=FALSE, warning=FALSE, error=FALSE, message=FALSE)
```

```{r load}

library(mgcv)
library(plantecophys)
library(reshape2)

source("load.R")
source("R/prepareDataset.R")

```

```{r eval=FALSE}
# Make quick pdf of gam by pftlong.
pdf("msowithpftandgam.pdf")
for(i in 1:nlevels(dat$pftlong)){
  d <- subset(dat, pftlong == levels(pftlong)[i])
  with(d, plot(lh.t, lmso, pch=19, col="dimgrey",main=levels(pftlong)[i]))
  gamplot("lh.t", "lmso", dfr=d, add=TRUE)

}
dev.off()

pdf("slawithpftandgam.pdf")
for(i in 1:nlevels(dat$pftlong)){
  d <- subset(dat, pftlong == levels(pftlong)[i])
  if(all(is.na(d$lsla)))next
  with(d, plot(lh.t, lsla, pch=19, col="dimgrey",main=levels(pftlong)[i]))
  gamplot("lh.t", "lsla", dfr=d, add=TRUE)

}
dev.off()

```

```{r functions}

fitgam <- function(X,Y,dfr, k=-1, model=1){
  dfr$Y <- dfr[,Y]
  dfr$X <- dfr[,X]
  dfr <- droplevels(dfr)
  g<- gam(Y ~ s(X, k=k), data=dfr)

  return(g)
}
addpoly <- function(x,y1,y2,col=alpha("lightgrey",0.8),...){
  ii <- order(x)
  y1 <- y1[ii]
  y2 <- y2[ii]
  x <- x[ii]
  polygon(c(x,rev(x)), c(y1, rev(y2)), col=col, border=NA,...)
}

gamplotandpred <- function(dat, pftvar, yvar, Hs = c(1,10)){

  o <- par()
  dat$PFT <- dat[,pftvar]
  dat$Y <- dat[,yvar]
  dat <- droplevels(subset(dat, !is.na(lh.t) & !is.na(Y)))

  d <- split(dat, dat$PFT)

  hran <- lapply(d, function(x)range(x$lh.t, na.rm=TRUE))
  fits <- lapply(d, function(x)fitgam("lh.t","Y",x, k=4))
  pred <- lapply(fits, function(x)predict(x, newdata=data.frame(X=log10(Hs)), 
                                          se.fit=TRUE))

  palette(rainbow(nlevels(dat$PFT)))

  with(dat, plot(lh.t, Y, axes=FALSE, pch=16, col=PFT, ylab=yvar))
  magaxis(side=1:2, unlog=1:2)

  for(i in 1:length(d)){
    nd <- data.frame(X=seq(hran[[i]][1], hran[[i]][2], length=101))
    p <- predict(fits[[i]],nd,se.fit=TRUE)
    addpoly(nd$X, p$fit-2*p$se.fit, p$fit+2*p$se.fit, col=alpha("lightgrey",0.7))
    lines(nd$X, p$fit, col=palette()[i], lwd=2)
  }
  abline(v=log10(Hs), col="dimgrey", lty=5)
  legend("bottomleft", levels(dat$PFT), fill=palette(), cex=0.7)

  logY <- sapply(pred, "[[", "fit")
  se <- sapply(pred, "[[", "se.fit")
  Y_cil <- 10^(logY - 2*se)
  Y_ciu <- 10^(logY + 2*se)
  Y <- 10^logY

  par(mar=c(7,5,2,2), las=2)
  b <- barplot2(Y[1,], beside=T, plot.ci=TRUE, ci.l=Y_cil[1,], ci.u=Y_ciu[1,],
           main=paste0("Height = ",Hs[1],"m"), col=palette(), ylab=yvar)

  b <- barplot2(Y[2,], beside=T, plot.ci=TRUE, ci.l=Y_cil[2,], ci.u=Y_ciu[2,],
           main=paste0("Height = ",Hs[2],"m"), col=palette(), ylab=yvar)

  df1 <- as.data.frame(Y)
  df1$H <- Hs
  df2 <- as.data.frame(se)
  df2$H <- Hs
  df1 <- melt(df1, id="H", value.name=yvar, variable.name=pftvar)
  df2 <- melt(df2, id="H", value.name=paste0(yvar,"_SE"), variable.name=pftvar)  
  df <- merge(df1, df2)
  par(o)
return(invisible(df))
}

```

### Leaf mass fraction and leaf area ratio by pft x biome

#### LMF
```{r lmf}
gamplotandpred(dataset2, "pftlong", "lmlf_mso")
```

### LAR
```{r lar}
gamplotandpred(dataset2, "pftlong", "lalf_mso")
```

### Leaf mass fraction and leaf area ratio by pft

#### LMF
```{r lmfpft}
gamplotandpred(dataset2, "pft", "lmlf_mso")
```

#### LAR
```{r larpft}
gamplotandpred(dataset2, "pft", "lalf_mso")
```

### Pipe model ratios by pft x biome

#### Leaf mass pipe ratio
```{r pipelmfpftlong}
pipelongmlf <- gamplotandpred(dataset2, "pftlong", "lmlf_astbh", Hs=c(3,10))
```

#### Leaf area pipe ratio
```{r pipeamfpftlong}
pipelongalf <- gamplotandpred(dataset2, "pftlong", "lalf_astbh", Hs=c(3,10))
```

### Pipe model ratios by pft

#### Leaf mass pipe ratio
```{r pipemlfpft}
pipemlf <- gamplotandpred(dataset2, "pft", "lmlf_astbh", Hs=c(3,10))
```

#### Leaf area pipe ratio
```{r pipealfpft}
pipealf <- gamplotandpred(dataset2, "pft", "lalf_astbh", Hs=c(3,10))
```

```{r}
sla <- gamplotandpred(dataset2, "pft", "lsla", Hs=c(3,10))
```

```{r}
slalong <- gamplotandpred(dataset2, "pftlong", "lsla", Hs=c(3,10))
```

### SLA and pipe model ratios across pft x biome

```{r}

long <- merge(slalong, pipelongmlf)
long$lma <- 1/long$lsla
long <- merge(long, pipelongalf)

makeci <- function(dfr, v){

  y <- dfr[,v]
  se <- dfr[,paste0(v,"_SE")]

  yl <- 10^(log10(y) - 2*se)
  yu <- 10^(log10(y) + 2*se)
return(data.frame(uci=yu, lci=yl))
}

addse <- function(dfr, x, y, direction, transformx=NULL){

  cix <- makeci(dfr,x)
  ciy <- makeci(dfr,y)
  if(!is.null(transformx)){
    cix <- transformx(cix)
    dfr[,x] <- transformx(dfr[,x])
  }

  arrows(x0=cix$lci, x1=cix$uci, y0=dfr[,y], y1=dfr[,y],
         code=3, angle=90, length=0.025)
  arrows(x0=dfr[,x], x1=dfr[,x], y0=ciy$lci, y1=ciy$uci,
         code=3, angle=90, length=0.025)

}

with(long,  plot(lma, lmlf_astbh, pch=19, col=c("grey","black")[as.factor(H)],
                 ylim=c(0,1000), xlim=c(0,0.4), 
                 panel.first={
                    addse(long, "lsla", "lmlf_astbh", transformx=function(x)1/x)
                  }))

with(long,  plot(lma, lalf_astbh, pch=19, col=c("grey","black")[as.factor(H)],
                 ylim=c(0,4500), xlim=c(0,0.4), 
                 panel.first={
                    addse(long, "lsla", "lalf_astbh", transformx=function(x)1/x)
                  }))

```

```{r}

palette(rainbow(7))
with(dataset2, plot(log10(a.stbh), log10(a.lf), pch=19, col=pftlong))

# SLA problems!
pinus <- dataset2[grep("Pinus", dataset2$speciesMatched),]
pinus <- subset(pinus, !is.na(a.lf) & !is.na(m.lf))
with(pinus, tapply(a.lf/m.lf, Group, mean))

```

### 2014-09-10 Rmd/lmfwithgam_bypft.Rmd
split gam analysis by pft or pftlong (Rmd)

--- 2014-09-05 Rmd/lmfwithgam.Rmd
@@
 source("R/prepareDataset.R")
-
-```
-
-```{r eval=FALSE}
-# Make quick pdf of gam by pftlong.
-pdf("msowithpftandgam.pdf")
-for(i in 1:nlevels(dat$pftlong)){
-  d <- subset(dat, pftlong == levels(pftlong)[i])
-  with(d, plot(lh.t, lmso, pch=19, col="dimgrey",main=levels(pftlong)[i]))
-  gamplot("lh.t", "lmso", dfr=d, add=TRUE)
-
-}
-dev.off()
-
-pdf("slawithpftandgam.pdf")
-for(i in 1:nlevels(dat$pftlong)){
-  d <- subset(dat, pftlong == levels(pftlong)[i])
-  if(all(is.na(d$lsla)))next
-  with(d, plot(lh.t, lsla, pch=19, col="dimgrey",main=levels(pftlong)[i]))
-  gamplot("lh.t", "lsla", dfr=d, add=TRUE)
-
-}
-dev.off()
-
 ```
@@
 
-### Leaf mass fraction and leaf area ratio by pft x biome
-
-#### LMF
-```{r lmf}
-gamplotandpred(dataset2, "pftlong", "lmlf_mso")
-```
-
-### LAR
-```{r lar}
-gamplotandpred(dataset2, "pftlong", "lalf_mso")
-```
-
 ### Leaf mass fraction and leaf area ratio by pft
@@
 gamplotandpred(dataset2, "pft", "lalf_mso")
-```
-
-### Pipe model ratios by pft x biome
-
-#### Leaf mass pipe ratio
-```{r pipelmfpftlong}
-pipelongmlf <- gamplotandpred(dataset2, "pftlong", "lmlf_astbh", Hs=c(3,10))
-```
-
-#### Leaf area pipe ratio
-```{r pipeamfpftlong}
-pipelongalf <- gamplotandpred(dataset2, "pftlong", "lalf_astbh", Hs=c(3,10))
 ```
@@
 
-```{r}
-slalong <- gamplotandpred(dataset2, "pftlong", "lsla", Hs=c(3,10))
-```
+```{r eval=FALSE}
 
-### SLA and pipe model ratios across pft x biome
+dat <- subset(dataset2,!is.na(a.stbh) & !is.na(m.so))
+datsp <- split(dat, dat$Group)
+ii <- which(sapply(datsp,nrow) > 10)
+lms <- lapply(datsp[ii],
+              function(x)lm(log10(m.so) ~ log10(a.stbh), data=x))  
+getr2 <- function(x)summary(x)$adj.r.squared
 
-```{r}
-
-long <- merge(slalong, pipelongmlf)
-long$lma <- 1/long$lsla
-long <- merge(long, pipelongalf)
-
-makeci <- function(dfr, v){
-
-  y <- dfr[,v]
-  se <- dfr[,paste0(v,"_SE")]
-
-  yl <- 10^(log10(y) - 2*se)
-  yu <- 10^(log10(y) + 2*se)
-return(data.frame(uci=yu, lci=yl))
-}
-
-addse <- function(dfr, x, y, direction, transformx=NULL){
-
-  cix <- makeci(dfr,x)
-  ciy <- makeci(dfr,y)
-  if(!is.null(transformx)){
-    cix <- transformx(cix)
-    dfr[,x] <- transformx(dfr[,x])
-  }
-
-  arrows(x0=cix$lci, x1=cix$uci, y0=dfr[,y], y1=dfr[,y],
-         code=3, angle=90, length=0.025)
-  arrows(x0=dfr[,x], x1=dfr[,x], y0=ciy$lci, y1=ciy$uci,
-         code=3, angle=90, length=0.025)
-
-}
-
-with(long,  plot(lma, lmlf_astbh, pch=19, col=c("grey","black")[as.factor(H)],
-                 ylim=c(0,1000), xlim=c(0,0.4), 
-                 panel.first={
-                    addse(long, "lsla", "lmlf_astbh", transformx=function(x)1/x)
-                  }))
-
-with(long,  plot(lma, lalf_astbh, pch=19, col=c("grey","black")[as.factor(H)],
-                 ylim=c(0,4500), xlim=c(0,0.4), 
-                 panel.first={
-                    addse(long, "lsla", "lalf_astbh", transformx=function(x)1/x)
-                  }))
+r2 <- sapply(lms, getr2)
 
@@
 
-```{r}
+```{r eval=FALSE}
+lme1 <- lme(log10(m.so) ~ log10(a.stbh), random=~log10(a.stbh)|Group, 
+            data=dataset2, subset=pft=="EG", na.action=na.omit)
+lme2 <- lme(log10(m.so) ~ log10(a.stbh), random=~log10(a.stbh)|Group, 
+            data=dataset2, subset=pft=="EA", na.action=na.omit)
+lme3 <- lme(log10(m.so) ~ log10(a.stbh), random=~log10(a.stbh)|Group, 
+            data=dataset2, subset=pft=="DA", na.action=na.omit)
 
-palette(rainbow(7))
-with(dataset2, plot(log10(a.stbh), log10(a.lf), pch=19, col=pftlong))
+datsp <- split(dataset2, dataset2$Group)
+R <- sapply(datsp, function(x)range(x$a.stbh))
 
-# SLA problems!
-pinus <- dataset2[grep("Pinus", dataset2$speciesMatched),]
-pinus <- subset(pinus, !is.na(a.lf) & !is.na(m.lf))
-with(pinus, tapply(a.lf/m.lf, Group, mean))
+makedf <- function(x){
+
+  p <- coef(x)
+  g <- rownames(p)
+  r <- R[,g]
+  df <- as.data.frame(cbind(p,t(r)))
+  names(df) <- c("a","b","xmin","xmax")
+
+  return(df)
+}
+
+p <- makedf(lme1)
+pointsp <- function(x,...){
+  for(i in 1:nrow(x)){
+    ablinepiece(x$a[i], x$b[i], from=log10(x$xmin[i]), to=log10(x$xmax[i]), ...)
+  }
+}
+with(dataset2, plot(log10(a.stbh), log10(m.so), type='n'))
+pointsp(makedf(lme1), col="red")
+pointsp(makedf(lme2), col="green3")
+pointsp(makedf(lme3), col="blue")
 

### 2014-09-10 Rmd/lmfwithgam_bypftlong.Rmd
split gam analysis by pft or pftlong (Rmd)

--- 2014-09-05 Rmd/lmfwithgam.Rmd
@@
 source("R/prepareDataset.R")
-
 ```
@@
 
-### Leaf mass fraction and leaf area ratio by pft
-
-#### LMF
-```{r lmfpft}
-gamplotandpred(dataset2, "pft", "lmlf_mso")
-```
-
-#### LAR
-```{r larpft}
-gamplotandpred(dataset2, "pft", "lalf_mso")
-```
-
 ### Pipe model ratios by pft x biome
@@
 
-### Pipe model ratios by pft
-
-#### Leaf mass pipe ratio
-```{r pipemlfpft}
-pipemlf <- gamplotandpred(dataset2, "pft", "lmlf_astbh", Hs=c(3,10))
-```
-
-#### Leaf area pipe ratio
-```{r pipealfpft}
-pipealf <- gamplotandpred(dataset2, "pft", "lalf_astbh", Hs=c(3,10))
-```
-
-```{r}
-sla <- gamplotandpred(dataset2, "pft", "lsla", Hs=c(3,10))
-```
-
+#### SLA
 ```{r}
@@
 
-```{r}
+```{r eval=FALSE}
 
@@
 ```
+
+```{r eval=FALSE}
+# Make sure no artefacts : any perfect R2 of m.so = f(a.stbh) ??
+dat <- subset(dataset2,!is.na(a.stbh) & !is.na(m.so))
+datsp <- split(dat, dat$Group)
+ii <- which(sapply(datsp,nrow) > 10)
+lms <- lapply(datsp[ii],
+              function(x)lm(log10(m.so) ~ log10(a.stbh), data=x))  
+getr2 <- function(x)summary(x)$adj.r.squared
+
+r2 <- sapply(lms, getr2)
+
+```
+
+```{r}
+lme1 <- lme(log10(m.so) ~ log10(a.stbh), random=~log10(a.stbh)|Group, 
+            data=dataset2, subset=pft=="EG", na.action=na.omit)
+lme2 <- lme(log10(m.so) ~ log10(a.stbh), random=~log10(a.stbh)|Group, 
+            data=dataset2, subset=pft=="EA", na.action=na.omit)
+lme3 <- lme(log10(m.so) ~ log10(a.stbh), random=~log10(a.stbh)|Group, 
+            data=dataset2, subset=pft=="DA", na.action=na.omit)
+
+datsp <- split(dataset2, dataset2$Group)
+R <- sapply(datsp, function(x)range(x$a.stbh))
+
+makedf <- function(x){
+
+  p <- coef(x)
+  g <- rownames(p)
+  r <- R[,g]
+  df <- as.data.frame(cbind(p,t(r)))
+  names(df) <- c("a","b","xmin","xmax")
+
+return(df)
+}
+
+p <- makedf(lme1)
+pointsp <- function(x,...){
+  for(i in 1:nrow(x)){
+    ablinepiece(x$a[i], x$b[i], from=log10(x$xmin[i]), to=log10(x$xmax[i]), ...)
+  }
+}
+with(dataset2, plot(log10(a.stbh), log10(m.so), type='n'))
+pointsp(makedf(lme1), col="red")
+pointsp(makedf(lme2), col="green3")
+pointsp(makedf(lme3), col="blue")
+
+```

### 2014-09-22 Rmd/lmfwithgam_bypft.Rmd
More work on allocation vs. sla, and predicting basal diameter from dbh

@@
 ```{r functions}
-
-fitgam <- function(X,Y,dfr, k=-1, model=1){
-  dfr$Y <- dfr[,Y]
-  dfr$X <- dfr[,X]
-  dfr <- droplevels(dfr)
-  g<- gam(Y ~ s(X, k=k), data=dfr)
-
-  return(g)
-}
-addpoly <- function(x,y1,y2,col=alpha("lightgrey",0.8),...){
-  ii <- order(x)
-  y1 <- y1[ii]
-  y2 <- y2[ii]
-  x <- x[ii]
-  polygon(c(x,rev(x)), c(y1, rev(y2)), col=col, border=NA,...)
-}
-
-gamplotandpred <- function(dat, pftvar, yvar, Hs = c(1,10)){
-
-  o <- par()
-  dat$PFT <- dat[,pftvar]
-  dat$Y <- dat[,yvar]
-  dat <- droplevels(subset(dat, !is.na(lh.t) & !is.na(Y)))
-
-  d <- split(dat, dat$PFT)
-
-  hran <- lapply(d, function(x)range(x$lh.t, na.rm=TRUE))
-  fits <- lapply(d, function(x)fitgam("lh.t","Y",x, k=4))
-  pred <- lapply(fits, function(x)predict(x, newdata=data.frame(X=log10(Hs)), 
-                                          se.fit=TRUE))
-
-  palette(rainbow(nlevels(dat$PFT)))
-
-  with(dat, plot(lh.t, Y, axes=FALSE, pch=16, col=PFT, ylab=yvar))
-  magaxis(side=1:2, unlog=1:2)
-
-  for(i in 1:length(d)){
-    nd <- data.frame(X=seq(hran[[i]][1], hran[[i]][2], length=101))
-    p <- predict(fits[[i]],nd,se.fit=TRUE)
-    addpoly(nd$X, p$fit-2*p$se.fit, p$fit+2*p$se.fit, col=alpha("lightgrey",0.7))
-    lines(nd$X, p$fit, col=palette()[i], lwd=2)
-  }
-  abline(v=log10(Hs), col="dimgrey", lty=5)
-  legend("bottomleft", levels(dat$PFT), fill=palette(), cex=0.7)
-
-  logY <- sapply(pred, "[[", "fit")
-  se <- sapply(pred, "[[", "se.fit")
-  Y_cil <- 10^(logY - 2*se)
-  Y_ciu <- 10^(logY + 2*se)
-  Y <- 10^logY
-
-  par(mar=c(7,5,2,2), las=2)
-  b <- barplot2(Y[1,], beside=T, plot.ci=TRUE, ci.l=Y_cil[1,], ci.u=Y_ciu[1,],
-                main=paste0("Height = ",Hs[1],"m"), col=palette(), ylab=yvar)
-
-  b <- barplot2(Y[2,], beside=T, plot.ci=TRUE, ci.l=Y_cil[2,], ci.u=Y_ciu[2,],
-                main=paste0("Height = ",Hs[2],"m"), col=palette(), ylab=yvar)
-
-  df1 <- as.data.frame(Y)
-  df1$H <- Hs
-  df2 <- as.data.frame(se)
-  df2$H <- Hs
-  df1 <- melt(df1, id="H", value.name=yvar, variable.name=pftvar)
-  df2 <- melt(df2, id="H", value.name=paste0(yvar,"_SE"), variable.name=pftvar)  
-  df <- merge(df1, df2)
-  par(o)
-  return(invisible(df))
-}
-
+source("R/gam_functions.R")
 ```

### 2014-09-22 R/predict_dba_frombh.R

@@ -1,0 +1,27 @@
+
+
+makebasalD <- function()
+  
+  
+
+test <- subset(baad, !is.na(d.ba) & !is.na(d.bh) & !is.na(h.t) & !is.na(h.bh) & h.t > h.bh)  
+
+
+fit <- nls(d.ba ~ d.bh * h.t^(a*h.t^b) /(h.t - h.bh)^(a*h.t^b), start=list(a=0.9, b=1),
+           data=test)
+
+test$d.ba2 <- predict(fit, test)
+
+with(test, plot(log10(d.ba2), log10(d.ba), col=as.factor(studyName), pch=16))
+abline(0,1)
+
+with(test, plot(d.ba2, d.ba))
+
+with(test, plot(h.t, d.ba - d.ba2))
+abline(h=0)
+
+# look at coefs.
+hts <- seq(2, 100, length=101)
+p <- predict(fit, expand.grid(d.bh=1, h.bh=1.3, h.t=hts))
+
+plot(hts, p)

### 2014-09-22 R/sla_andpipemodel_bypftlong.R

@@ -1,28 +1,33 @@
 
 source("load.R")
 source("R/preparedataset.R")
-library(multcomp)
 
 
-meansbypftlong <- function(yvar1, yvar2, xlab=expression(Specific~leaf~area~(m^2~kg^-1)), 
-                           ylab1=NULL, ylab2=NULL, addtrend=c(TRUE,FALSE),
+meansbypftlong <- function(yvar1, yvar2, pftvar, xlab=expression(Specific~leaf~area~(m^2~kg^-1)), 
+                           xlim=c(0,25),main="",
+                           ylab1=NULL, ylab2=NULL, addtrend=c(FALSE,FALSE),
                            ylim1=NULL, ylim2=NULL, dataset){
   
-  
-  dat <- droplevels(subset(dataset, pftlong %in% 
+  if(pftvar == "pftlong"){
+    dat <- droplevels(subset(dataset, pftlong %in% 
                              c("DA-temperate","EA-temperate","EA-tropical","EG-boreal",
                                "EG-temperate")))
+  } else {
+    dat <- dataset
+  }
+  
   dat$Y1 <- dat[,yvar1]
   dat$Y2 <- dat[,yvar2]
+  dat$P <- dat[,pftvar]
   
   mixmean <- function(yvar, dataset=dat){
     
     dataset$Y <- dataset[,yvar]
-    f <- lmer(Y ~ pftlong - 1 + (1|Group), data=dataset)
+    f <- lmer(Y ~ P - 1 + (1|Group), data=dataset)
     
     ci <- suppressMessages(confint(f)[-c(1,2),])
     ci <- as.data.frame(cbind(10^fixef(f),10^ci))
-    rownames(ci) <- gsub("pftlong","", rownames(ci))
+    rownames(ci) <- gsub("P","", rownames(ci))
     names(ci) <- c("y","lci","uci")
     return(ci)
   }
@@ -31,26 +36,26 @@ meansbypftlong <- function(yvar1, yvar2, xlab=expression(Specific~leaf~area~(m^2
   mlfastbh <- mixmean(yvar1)
   alfastbh <- mixmean(yvar2)
   
-  palette(rainbow(5))
+  palette(rainbow(length(unique(dat$P))))
   
[…]

### 2014-10-01 manuscript/draftpaper.tex
More lit notes

@@
 
+Friedlingstein et al 1999 make no mention of plant functional type differences or the need to vary allocation by pft.
+
 Poorter et al.
@@
 evergreen).
+
+Enquist & Niklas (2002) report 2.6 times higher leaf mass at given stem mass for conifers compared to angiosperms, but did not separate further, nor interpreted this difference in terms of SLA.
+Their interpretation is 'that conifers typically retain three cohorts of leaves that have less well-developed [...] mesophyll'.
+Not sure how that latter bit makes sense, and leaf lifespan interpretation is no good; because it still begs the question why would conifers not simply have 1/3 less leaf mass per cohort?
+
+McCarthy et al.
+(2007) further demonstrate large difference in LMF between angiosperms and conifers, and note that a large chunk of the residuals is correlated with SLA and LL.
+They interpret this rather vaguely as 'leaf morphology may constrain leaf mass distribution'.
+Their SLA was not on the same individuals, but they could have calculated PFT average LAR anyway, but did not.
 

### 2014-10-07 manuscript/draftpaper.tex
Added bib file with many allometry studies

@@
 \geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
+\bibliographystyle{authordate1}
 
@@
 For example, vegetation with a higher leaf mass fraction will have a lower overall residence time of carbon.
+Study was \cite{friend_carbon_2014}.
 
@@
 
-Enquist & Niklas (2002) report 2.6 times higher leaf mass at given stem mass for conifers compared to angiosperms, but did not separate further, nor interpreted this difference in terms of SLA.
+Enquist and Niklas (2002) report 2.6 times higher leaf mass at given stem mass for conifers compared to angiosperms, but did not separate further, nor interpreted this difference in terms of SLA.
 Their interpretation is 'that conifers typically retain three cohorts of leaves that have less well-developed [...] mesophyll'.
@@
 
+O'Neill and DeAngelis (1981) reported much higher leaf biomass for gymnosperms vs.
+angiosperms, at a given woody biomass increment.
+This means (as Niklas reported) a much higher productivity per unit leaf biomass for angiosperms.
+
 Serious lack of biomass data to constrain models.
@@
 (2010) did not find any differences in biomass fractions (root,stem,leaf) between angiosperms and gymnosperms (but again this was based on stand biomass data, but still this is strange!).
+
+Hui et al 2014 - isometric scaling of above vs. belowground biomass in chinese forest biomass database.
+
+Sardans and Penuelas 2013 - report similiar LMF for evergreen angio. and gymnosperms, and ca. 40\% less leaf mass for decid. angio. Local study, 3000 plots in Catalonia.
+
+Reich et al 2014 : "On average, when standardized for MAT and Mstem, conifers tended to have more Mfol and less Mroot than angiosperm forests".
+Their Fig.
+2 shows trends in LMF with MAT etc., in bins of stem size.
+Also allometric analysis with MAT as covariate, showing significance of MAT.
+For Angiosperms, our explanation would be that evergreen angio is more common at high MAT.
+But what about conifers?
 
@@
 
+\section{References}
+\bibliography{allometry}
+
 \end{document}

### 2014-10-07 manuscript/draftpaper.tex
Refs properly added to manuscript

@@
 \geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
-\bibliographystyle{authordate1}
+\bibliographystyle{plain}
 
@@
 \begin{enumerate}
-  \item Importance of biomass allocation in global vegetation models, feedback on fluxes of CO2, lifetime of C in the system. 
-  \item Close relationship to allometric scaling models, which provide an important constraint on the allocation of biomass into the various plant compartments (roots, leaves and stems).
-  \item Previous theoretical work on allometric scaling has predominantly focused on the universal nature of size-biomass relationships in plants.
-This side-steps the known difference in biomass fractions at a given plant size between plant functional types (Enquist, Poorter).
-  \item There is no shortage of theoretical developments in understanding biomass allocation (Franklin).
-Other theoretical work focusing on optimal partitioning theory is promising but is yet to lead to operational models for dynamic global vegetation models.
-  \item These DGVMs typically are parameterized by plant functional type (of woody vegetation types that we here allude to include deciduous, evergreen, angiosperm, gymnosperm), current knowledge of allometry differences between pfts is very weak.
-  \item Whatever the way forward in allocation, what is lacking is a solid empirical base that can be used to constrain any model of allocation, that includes woody plants of all sizes, of all major plant functional types and biomes.
-  \item Here we use such a database (Falster et al.) to evaluate whether important differences exist in biomass partitioning between four major woody plant functional types.
-  \item We find that specific leaf area is the major determinant of the partitioning of biomass between leaves and stems.
-After correcting for important size effects, the ratio of leaf mass to total aboveground biomass is varied in such a way to keep the ratio of leaf area to aboveground biomass invariable between plant functional types.
+\item Importance of biomass allocation in global vegetation models \cite{ise_comparison_2010}, feedback on fluxes of CO2, lifetime of C in the system \cite{friend_carbon_2014}.
+
+\item There is no shortage of theoretical developments in understanding biomass allocation (WBE, \cite{franklin_modeling_2012}).
+In particular, theoretical work focusing on optimal partitioning theory is promising (e.g.
+\cite{valentine_modeling_2012}) but is yet to lead to operational models for dynamic global vegetation models.
+
+\item Previous syntheses on plant biomass relationships has predominantly focused on the universal nature of size-biomass relationships in plants.
+Little is known about differences between plant functional types (evergreen vs.
+deciduous, angiosperm vs.
+gymnosperm).
+Since most global models parameterize allocation submodels by PFT, it is imperative that we better understand these potential differences.
+
+\item Whichever is the 'right' allocation model, what is lacking is a solid empirical base that can be used to constrain any model of allocation, that includes woody plants of all sizes, of all major plant functional types and biomes.
+
+\item Here we use such a database (Falster et al.) to evaluate whether important differences exist in biomass partitioning woody plant functional types, and whether climatic differences exist.
+
 \item As a first approximation, which is useful for large scale DGVMs, models should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly.
 Specific leaf area then gives the ratio of leaf biomass to stem biomass.
-This follows because aboveground woody biomass scales closely with stem cross-sectional area and plant height, in a way that was found to not vary between pft's (*****).
 
@@
 
-Friend et al.
-(2014) showed that 30\% of the uncertainty in global vegetation carbon simulated with seven vegetation models is due to the residence time of carbon in vegetation.
+\cite{friend_carbon_2014} showed that 30\% of the uncertainty in global vegetation carbon simulated with seven vegetation models is due to the residence time of carbon in vegetation.
 This residence time depends on turnover rates of the various plant pools, and hence also on the distribution of biomass between these pools.
 For example, vegetation with a higher leaf mass fraction will have a lower overall residence time of carbon.
-Study was \cite{friend_carbon_2014}.
 
-De Kauwe et al.
-(2014) : "We have shown that allocation approaches that are constrained by biomass fractions (such as functional relationships) were more successful at capturing observed trends, and were generally more robust, than approaches based on allocation coefficients."
+\cite{de2014does} "We have shown that allocation approaches that are constrained by biomass fractions (such as functional relationships) were more successful at capturing observed trends, and were generally more robust, than approaches based on allocation coefficients."
 
-Friedlingstein et al 1999 make no mention of plant functional type differences or the need to vary allocation by pft.
+\cite{friedlingstein_toward_1999} make no mention of plant functional type differences or the need to vary allocation by pft.
 
-Poorter et al.
-(2012) found a slightly larger LMF in gymnosperms, but the difference was small and could not split between deciduous and evergreen angiosperms due to a lack of data.
+\cite{poorter_biomass_2012} found a slightly larger LMF in gymnosperms, but the difference was small and could not split between deciduous and evergreen angiosperms due to a lack of data.
 They argue that "most conifers retain leaves for two or more years.
@@
 
-Enquist and Niklas (2002) report 2.6 times higher leaf mass at given stem mass for conifers compared to angiosperms, but did not separate further, nor interpreted this difference in terms of SLA.
+\cite{enquist_global_2002} report 2.6 times higher leaf mass at given stem mass for conifers compared to angiosperms, but did not separate further, nor interpreted this difference in terms of SLA.
 Their interpretation is 'that conifers typically retain three cohorts of leaves that have less well-developed [...] mesophyll'.
@@
 
-McCarthy et al.
-(2007) further demonstrate large difference in LMF between angiosperms and conifers, and note that a large chunk of the residuals is correlated with SLA and LL.
+\cite{mccarthy_consistency_2007} further demonstrate large difference in LMF between angiosperms and conifers, and note that a large chunk of the residuals is correlated with SLA and LL.
 They interpret this rather vaguely as 'leaf morphology may constrain leaf mass distribution'.
@@
 
-O'Neill and DeAngelis (1981) reported much higher leaf biomass for gymnosperms vs.
+\cite{oneill_comparative_1981} reported much higher leaf biomass for gymnosperms vs.
 angiosperms, at a given woody biomass increment.
@@
 Serious lack of biomass data to constrain models.
-Studies like Wolf et al.
-(2011) resort to studying individual tree allometry based on dividing stand biomass by stand density.
-Wolf et al.
-(2010) did not find any differences in biomass fractions (root,stem,leaf) between angiosperms and gymnosperms (but again this was based on stand biomass data, but still this is strange!).
+Studies like \cite{wolf_forest_2011} resort to studying individual tree allometry based on dividing stand biomass by stand density.
+\cite{wolf_allometric_2010} did not find any differences in biomass fractions (root,stem,leaf) between angiosperms and gymnosperms (but again this was based on stand biomass data, but still this is strange!).
 
-Hui et al 2014 - isometric scaling of above vs. belowground biomass in chinese forest biomass database.
+\cite{hui_near_2014} Isometric scaling of above vs. belowground biomass in chinese forest biomass database.
 
-Sardans and Penuelas 2013 - report similiar LMF for evergreen angio. and gymnosperms, and ca. 40\% less leaf mass for decid. angio. Local study, 3000 plots in Catalonia.
+\cite{sardans_tree_2013} report similiar LMF for evergreen angio. and gymnosperms, and ca. 40\% less leaf mass for decid. angio. Local study, 3000 plots in Catalonia.
 
-Reich et al 2014 : "On average, when standardized for MAT and Mstem, conifers tended to have more Mfol and less Mroot than angiosperm forests".
+\cite{reich_temperature_2014} "On average, when standardized for MAT and Mstem, conifers tended to have more Mfol and less Mroot than angiosperm forests".
 Their Fig.
@@
 \section{Results and Discussion}
-(this will not be separate in the manuscript, just temporary notes here)
-
-Figure (figure\_mlfastbh\_sla\_pft.pdf) shows that across plant functional types (pfts) (but not within pfts), the ratio of leaf mass to stem cross-sectional area is negatively correlated with specific leaf area.
-This means that pfts with larger SLA display less leaf mass per unit stem area (and hence per unit aboveground biomass), in a way so that leaf area per stem cross sectional area does not differ by pft (figure\_alfastbh\_sla\_pft.pdf).
-The convergent trait, independent of plant functional type, is then the ratio of leaf area to stem area.
-
-Though leaf area per stem area did not differ between pfts (figure\_alfastbh\_sla\_pft and other figure tba), within pfts it was correlated to sla.
-In other words, for a given pft, species with higher sla tended to have larger total leaf area at a given stem cross-sectional area.
-This was in a way so that leaf mass per unit stem area did not correlate to sla within a pft, so that within pfts the more convergent trait is in fact leaf mass per unit stem area.
-
-The implication for global vegetation models that use pft as the unit of parameterization is that specific leaf area (pft-specific) should be combined with a target leaf area to stem area ratio to give a leaf mass fraction that constrains the allocation of carbon.
-This is what LPJ does, but there the pipe model ratio is actually pft-specific (and nobody knows where it comes from; see sitch and also Hickler).
-
-Fixed allocation fraction models would get the distribution of biomass right if leaf turnover rate is exactly proportional to specific leaf area (because if an extra unit of leaf area is produced when SLA is increased, this could be offset exactly by a faster turnover rate).
-However glopnet says that turnover rate scales with more than SLA (exponent 1.3, but not a great fit).
-
-\section{Analyses}
-
-A list of analyses to be included, either in the main text or as supporting information. We need to show all these bits to support the storyline.
-
-\begin{enumerate}
-  \item Leaf mass vs. aboveground biomass; line per species/studyName, colored by pft.
-  \item Leaf mass / aboveground biomass vs. plant height; line per species/studyName, colored by pft.
-  \item Barplot leaf mass / aboveground biomass at reference height with bootstrap CI, by pft (and vegetation type)
-  \item Barplot leaf area / aboveground biomass at reference height with bootstrap CI, by pft (and vegetation type)
-  \item Same ones but for leaf mass or leaf area vs. stem area (at BH or base).
-  \item Ratio mass / stem area vs. SLA, ratio area / stem area vs. SLA, as estimated from random effects of fitted lme model.
-  \item Root - shoot ratio, and barplots of pft/vegetation averages with bootstrap CI.
-  \item Histogram of estimated effects at a reference height, by plant functional type, showing the variation between species. 
-\end{enumerate}
 
@@
   \item Removed deciduous gymnosperms, we have a few of those in BAAD (6 species / studyName combinations), but only one includes estimates of leaf area and mass. 
-We did not attempt here to include SLA from another database (* although we can do this, we need to decide!).
+  \item We did not attempt here to include SLA from another database (such as TRY), instead only used SLA directly estimated for the harvested plants.
+  \item We used a generalized additive model (GAM) to visualize the relationships.
+In addition, we used linear mixed-effects models (with species within study as the random effect) to test for significance and to calculate confidence intervals for PFT-specific averages.
 \end{itemize}
 
-\section{References}
 \bibliography{allometry}

### 2014-10-07 manuscript/draftpaper.tex
More refs

@@
 
-\item There is no shortage of theoretical developments in understanding biomass allocation (WBE, \cite{franklin_modeling_2012}).
+\item There is no shortage of theoretical developments in understanding biomass allocation (\cite{cannell_carbon_1994,franklin_modeling_2012}, WBE).
 In particular, theoretical work focusing on optimal partitioning theory is promising (e.g.
@@
 
+\cite{cairns_root_1997} found no difference in root-shoot ratio between gymno and angio.
+
+\cite{delucia_climate-driven_2000} LA/SA decreases with summer VPD for Pinus but not non-Pinus gymno.
+Also Pinus has much lower LA/SA than non-Pinus; suggesting that we should separate gymnosperms into those groups as well.
+
 \section{Results and Discussion}

### 2014-10-07 manuscript/draftpaper.tex
More refs
callaway + givnish refs

--- 2014-10-07 manuscript/draftpaper.tex
@@
 \geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
-\bibliographystyle{authordate1}
 
@@
 For example, vegetation with a higher leaf mass fraction will have a lower overall residence time of carbon.
-Study was \cite{friend_carbon_2014}.
 
@@
 
-Enquist and Niklas (2002) report 2.6 times higher leaf mass at given stem mass for conifers compared to angiosperms, but did not separate further, nor interpreted this difference in terms of SLA.
+Enquist & Niklas (2002) report 2.6 times higher leaf mass at given stem mass for conifers compared to angiosperms, but did not separate further, nor interpreted this difference in terms of SLA.
 Their interpretation is 'that conifers typically retain three cohorts of leaves that have less well-developed [...] mesophyll'.
@@
 
-O'Neill and DeAngelis (1981) reported much higher leaf biomass for gymnosperms vs.
-angiosperms, at a given woody biomass increment.
-This means (as Niklas reported) a much higher productivity per unit leaf biomass for angiosperms.
-
 Serious lack of biomass data to constrain models.
@@
 
-Hui et al 2014 - isometric scaling of above vs. belowground biomass in chinese forest biomass database.
+Givnish 2002 discusses costs and benefits of evergreen vs.
+deciduous habit, in terms of constructing a growth model that takes into account photosynthesis, leaf lifespan, belowground costs.
+However he does not consider or report on differences in allocation associated with leaf lifespan.
+Instead he simply states that 'leaf allocation ratio' relates to leaf lifespan, which of course is only true if total leaf mass is constant.
 
-Sardans and Penuelas 2013 - report similiar LMF for evergreen angio. and gymnosperms, and ca. 40\% less leaf mass for decid. angio. Local study, 3000 plots in Catalonia.
-
-Reich et al 2014 : "On average, when standardized for MAT and Mstem, conifers tended to have more Mfol and less Mroot than angiosperm forests".
-Their Fig.
-2 shows trends in LMF with MAT etc., in bins of stem size.
-Also allometric analysis with MAT as covariate, showing significance of MAT.
-For Angiosperms, our explanation would be that evergreen angio is more common at high MAT.
-But what about conifers?
+Callaway et al 1994, example of lower leaf mass / sapwood area for desert population of PIPO vs.
+montane population.
+For pointing out small-scale climatic effects on allometry (that may not be obvious in global scale analyses).
 
@@
 
-\section{References}
-\bibliography{allometry}
-
 \end{document}

### 2014-10-07 manuscript/draftpaper.tex

--- 2014-10-07 manuscript/draftpaper.tex
@@
 
+Givnish 2002 discusses costs and benefits of evergreen vs.
+deciduous habit, in terms of constructing a growth model that takes into account photosynthesis, leaf lifespan, belowground costs.
+However he does not consider or report on differences in allocation associated with leaf lifespan.
+Instead he simply states that 'leaf allocation ratio' relates to leaf lifespan, which of course is only true if total leaf mass is constant.
+
+Callaway et al 1994, example of lower leaf mass / sapwood area for desert population of PIPO vs.
+montane population.
+For pointing out small-scale climatic effects on allometry (that may not be obvious in global scale analyses).
+
 \cite{poorter_biomass_2012} found a slightly larger LMF in gymnosperms, but the difference was small and could not split between deciduous and evergreen angiosperms due to a lack of data.

### 2014-10-09 manuscript/draftpaper.tex
Figures now saved in manuscript/figures and added to manuscript

@@
 \usepackage{parskip}
+\usepackage{graphicx}
+\graphicspath{ {./figures/} }
 \usepackage{geometry}
@@
 
-\section{Main text}
+\section{Introduction}
 
@@
 
+%------------------------------------------------------------------------------------%
+\section{Methods}
+
+\begin{itemize}
+  \item Removed deciduous gymnosperms, we have a few of those in BAAD (6 species / studyName combinations), but only one includes estimates of leaf area and mass. 
+  \item We did not attempt here to include SLA from another database (such as TRY), instead only used SLA directly estimated for the harvested plants.
+  \item We used a generalized additive model (GAM) to visualize the relationships.
+In addition, we used linear mixed-effects models (with species within study as the random effect) to test for significance and to calculate confidence intervals for PFT-specific averages.
+\end{itemize}
+
+%------------------------------------------------------------------------------------%
+\section{Results}
+
+%------------------------------------------------------------------------------------%
+\section{Discussion}
+
+%------------------------------------------------------------------------------------%
 \section{Lit notes}
@@
 
-\section{Results and Discussion}
+\cite{grier_notes:_1974} early study on leaf mass proportionality to sapwood area (and uses mass not area).
 
-\section{Methods}
+\section{Figures}
 
-\begin{itemize}
-  \item Removed deciduous gymnosperms, we have a few of those in BAAD (6 species / studyName combinations), but only one includes estimates of leaf area and mass. 
-  \item We did not attempt here to include SLA from another database (such as TRY), instead only used SLA directly estimated for the harvested plants.
-  \item We used a generalized additive model (GAM) to visualize the relationships.
-In addition, we used linear mixed-effects models (with species within study as the random effect) to test for significance and to calculate confidence intervals for PFT-specific averages.
-\end{itemize}
+\begin{figure}[h!]
+    \centering
+    \includegraphics{figure1_mlf_astba2_bypft.pdf}
+    \caption{Leaf mass per unit basal stem area. Each symbol is an individual plant. Lines are generalized additive model fits.}
+    \label{fig:figure1}
+\end{figure}
 
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{figure2_mlf_alf_astbaest_pftmeans.pdf}
+    \caption{Average leaf mass per unit basal stem area (panel 1), and leaf area per unit basal stem area (panel 2) for the three major plant functional types.
+Error bars are 95\% confidence interval for the mean (calculated with a mixed-effects model with species within study as the random effect).
+Letters denote significant differences.
+}
+    \label{fig:figure2}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{figure3_LMF_pft_lines.pdf}
+    \caption{Leaf mass fraction (leaf mass / aboveground biomass) by plant functional type. Each symbol is an individual plant. Lines are generalized additive model fits.}
+    \label{fig:figure3}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{figure4_LAR_pft_lines.pdf}
+    \caption{Leaf area ratio (leaf area / aboveground biomass) by plant functional type. Each symbol is an individual plant. Lines are generalized additive model fits.}
+    \label{fig:figure4}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{figure4_LAR_pft_lines.pdf}
+    \caption{Leaf area ratio (leaf area / aboveground biomass) by plant functional type. Each symbol is an individual plant. Lines are generalized additive model fits.}
+    \label{fig:figure4}
+\end{figure}
+
+\clearpage
 \bibliography{allometry}

### 2014-10-09 manuscript/draftpaper.tex
captions for all figures done

@@
 In addition, we used linear mixed-effects models (with species within study as the random effect) to test for significance and to calculate confidence intervals for PFT-specific averages.
+  \item For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception (Lang 1991, Stenberg 2006).
 \end{itemize}
@@
 
+\section{Supporting Information Figures}
+
 \begin{figure}[h!]
     \centering
-    \includegraphics[width=0.99\textwidth]{figure4_LAR_pft_lines.pdf}
-    \caption{Leaf area ratio (leaf area / aboveground biomass) by plant functional type. Each symbol is an individual plant. Lines are generalized additive model fits.}
-    \label{fig:figure4}
+    \includegraphics[width=0.99\textwidth]{figureSI-1_alf_astba2_bypft.pdf}
+    \caption{Leaf area per unit basal stem area. Each symbol is an individual plant. Lines are generalized additive model fits.}
+    \label{fig:figureSI1}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{figureSI-2_mlf_alf_astbaest_pftlongmeans.pdf}
+    \caption{Like Fig.~\ref{fig:figure2}, but for five combinations of major plant functional type and biome.
+Out of 9 possible biome / pft combinations, only five in the dataset had leaf area estimates (and two combo's are not in the dataset, e.g.
+tropical evergreen gymno).
+Note that within pft, vegetation is never significant (this result will also be in another figure!).}
+    \label{fig:figureSI2}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.5\textwidth]{figureSI-3_lmlf_astba2_hist_bypft.pdf}
+    \caption{Histograms of leaf mass per unit basal stem area grouped by the three major plant functional types.
+Vertical lines and shading are means with confidence intervals (same as in Fig.~\ref{fig:figure2})}
+    \label{fig:figureSI3}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.5\textwidth]{figureSI-4_lalf_astba2_hist_bypft.pdf}
+    \caption{Histograms of leaf area per unit basal stem area grouped by the three major plant functional types.
+Vertical lines and shading are means with confidence intervals (same as in Fig.~\ref{fig:figure2})}
+    \label{fig:figureSI4}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{figureSI-5_mlfmst_byht_pft.pdf}
+    \caption{Leaf and woody biomass scaling with plant height.}
+    \label{fig:figureSI5}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{figureSI-6_mrt_mso_bypft.pdf}
+    \caption{Near-isometric scaling of above-ground with below-ground biomass, and no effects of plant functional type. The black line is a 1-1 line.}
+    \label{fig:figureSI6}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{figureSI-7_MAT_LMFscaling.pdf}
+    \caption{For each species within study, we fit the scaling relationship of leaf mass as a function of woody biomass assuming an exponent of 3/4, using standardized major axis.
+The resulting coefficient (b0) is size-invariant (and thus superior to the simple ratio of leaf mass to woody biomass, which is strongly size-dependent, see Fig.~\ref{fig:figure3}).
+We found no effects of mean annual temperature on the coefficient.
+The triangles are from Roth et al.
+(2007), a study on Pinus eliottii with exceptionally high leaf mass per unit woody biomass.}
+    \label{fig:figureSI7}
 \end{figure}

### 2014-10-10 manuscript/draftpaper.tex
Added a simple Makefile for the manuscript
Makefile works OK now. Major update manuscript.

@@
 
-\title{Specific leaf area explains difference in biomass partitioning between woody plant functional types}
+\title{Specific leaf mass explains difference in biomass partitioning between woody plant functional types}
 
@@
 
-\item There is no shortage of theoretical developments in understanding biomass allocation (\cite{cannell_carbon_1994,franklin_modeling_2012}, WBE).
+\item There is no shortage of theoretical developments in understanding biomass allocation (\cite{cannell_carbon_1994,franklin_modeling_2012,enquist_land_2012}).
 In particular, theoretical work focusing on optimal partitioning theory is promising (e.g.
@@
 
-\item Whichever is the 'right' allocation model, what is lacking is a solid empirical base that can be used to constrain any model of allocation, that includes woody plants of all sizes, of all major plant functional types and biomes.
+\item Major plant functional types differ widely in specific leaf mass (SLM, the ratio of leaf mass to area, an index of leaf thickness) (\cite{poorter_causes_2009}), which can be expected to affect biomass partitioning in several ways.
+First, SLM correlates with leaf lifespan.
+Longer lived leaves would lead to higher plant leaf mass (MF) as long as annual production rates are similar (\cite{enquist_global_2002}).
+Another interpretation is that plants may target a certain leaf area display given a certain biomass investment in supporting tissues, because light interception and photosynthesis correlate with the exchange area of the plant.
 
-\item Here we use such a database (Falster et al.) to evaluate whether important differences exist in biomass partitioning woody plant functional types, and whether climatic differences exist.
+\item It has been reported that gymnosperms (with higher SLM) have higher leaf mass fraction than angiosperms (with lower SLM) \cite{enquist_global_2002, mccarthy_consistency_2007, poorter_biomass_2012, sardans_tree_2013}.
+But a recent study reported no difference (\cite{wolf_allometric_2010}).
+No study to date has interpreted differences in partitioning between PFTs in terms of SLM.
+This would be useful because SLM is already an important parameter in DGVMs, potentially reducing parameters required for the allocation subroutine.
 
-\item As a first approximation, which is useful for large scale DGVMs, models should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly.
-Specific leaf area then gives the ratio of leaf biomass to stem biomass.
+\item What is lacking is a solid empirical base that can be used to constrain any model of allocation, that includes woody plants of all sizes, of all major plant functional types and biomes.
+Most global efforts to date use the Cannell database of forest biomass, which is not based on individual biomass estimates.
+Here we use a new database (Falster et al.), the largest ever compiled for individual woody plants, to evaluate whether important differences exist in biomass partitioning woody plant functional types.
+We also test whether these differences are confounded (or compounded) by climate.
 
@@
 \begin{itemize}
+\item BAAD database, XXX observations etc.
+
 \item Removed deciduous gymnosperms, we have a few of those in BAAD (6 species / studyName combinations), but only one includes estimates of leaf area and mass.
+Also removed glasshouse and common garden studies, including only field grown woody plants (including plantations).
+
 \item We did not attempt here to include SLA from another database (such as TRY), instead only used SLA directly estimated for the harvested plants.
+
+\item Stem area was measured either at breast height (ca.
+1.3m) or at the base of the plant (normally for small plants).
+For the subset of the data where both were measured, we estimated basal stem area (Asba) from breast height area (Asbh) by fitting XXX equation.
+Results were similar when we used either measure of stem area (not shown).
+
 \item We used a generalized additive model (GAM) to visualize the relationships.
-In addition, we used linear mixed-effects models (with species within study as the random effect) to test for significance and to calculate confidence intervals for PFT-specific averages.
+
+\item We used linear mixed-effects models (with species within study as the random effect) to test for significance and to calculate confidence intervals for  PFT-specific averages.
+
+\item Variance partitioning with R2 for linear mixed-effects models as reported by XXX.
+
 \item For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception (Lang 1991, Stenberg 2006).
@@
 
+Note: stats done as well, but not yet incorporated into the manuscript.
+
+\begin{itemize}
+
+\item We focus here only on aboveground biomass because we found no PFT differences in root-shoot scaling (Fig.~\ref{fig:figureSI6}), and approximately isometric scaling (sma slope = xxx) (consistent with other reports, \cite{hui_near_2014, cairns_root_1997}).
+
+\item We found strong correlations between basal stem area and leaf mass across the entire dataset, with a clear separation between the three major plant functional types (Fig.~\ref{fig:figure1}).
+We analyzed the ratio MF/Asba, and found that PFT explained by far the most variation (variables tested: plant height, biome (boreal,tropical,temperate), PFT, and all interactions).
+
+\item Because of weak plant height effects, we calculated the average MF/Asba for the three pfts; there was a ca.
+3-fold difference between evergreen gymnosperms and deciduous angiosperms, with evergreen angiosperms taking an intermediate value (Fig.~\ref{fig:figure2}).
+These differences were approximately proportional to the average SLM for the PFT.
+
+\item When we calculated the ratio of leaf area to basal stem area, we found no difference between PFTs (Fig.~\ref{fig:figure2}), confirming that MF/Asba is approximately proportional to SLM across PFTs.
+(See also Fig.~\ref{fig:figureSI1}).
+
+\item We do note that there was substantial variation between individual plants (Figs.~\ref{fig:figureSI3} and~\ref{fig:figureSI3}).
+Nonetheless, PFT explained most variation in MF/Asba (to be added).
+If we also included SLM in the model, PFT was no longer significant.
+PFT explained very little variation when we partitioned the variance in AF/Asba.
+
+\item We found similar results for the whole-plant leaf mass fraction (LMF), where a clear and strong separation was found between plant functional types (Fig.~\ref{fig:figure3}).
+See also Fig.~\ref{fig:figureSI5} for the raw data.
+For the leaf area ratio, pfts were generally closer together except for small evergreen gymnosperms (< 1m height) for unknown reasons.
+Variance partitioning also confirmed that the differences between PFT could be largely explained by their differences in SLM (not shown yet!).
+
+\item We found no effect of biome on the ratio MF/Asba or AF/Asba (tropical, temperate, boreal) (Fig.~\ref{fig:figureSI2}).
+
+\item Unlike \cite{reich_temperature_2014}, we found no effect of mean annual temperature on the scaling of leaf biomass with woody biomass (Fig.~\ref{fig:figureSI7}).
+
+\end{itemize}
+
 %------------------------------------------------------------------------------------%
 \section{Discussion}
+
+\begin{itemize}
+
+\item How do current DGVMs work?
+Either constant allocation fraction, target allometric equations, or optimization \cite{de2014does}.
+Constant allocation fraction gives constant LAR by pft if leaf lifespan is exactly proportional to SLM.
+Target allometric equations can be made pft-invariant by using allometry of leaf area, not leaf mass.
+Optimization models probably already give the right answer because normally LAI is optimized (Woodward).
+
+\item As a first approximation, which is useful for large scale DGVMs, models should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly.
+Specific leaf area then gives the ratio of leaf biomass to stem biomass.
+
+\item We found no effects of climate on anything.
+This is different from \cite{reich_temperature_2014} but they studied whole-stand biomass, not individual plant biomass.
+So perhaps this means the climate effects are on stand density not individual plant effects.
+No effect of climate is good news for models that parameterize the globe by PFT, as this seems sufficient, based on our dataset.
+
+\end{itemize}
 
@@
 
+\clearpage
 \section{Supporting Information Figures}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{figureSI-6_mrt_mso_bypft.pdf}
+    \caption{Near-isometric scaling of above-ground with below-ground biomass, and no effects of plant functional type. The black line is a 1-1 line.}
+    \label{fig:figureSI6}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.5\textwidth]{figureSI-3_lmlf_astba2_hist_bypft.pdf}
+    \caption{Histograms of leaf mass per unit basal stem area grouped by the three major plant functional types.
+Vertical lines and shading are means with confidence intervals (same as in Fig.~\ref{fig:figure2})}
+    \label{fig:figureSI3}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.5\textwidth]{figureSI-4_lalf_astba2_hist_bypft.pdf}
+    \caption{Histograms of leaf area per unit basal stem area grouped by the three major plant functional types.
+Vertical lines and shading are means with confidence intervals (same as in Fig.~\ref{fig:figure2})}
+    \label{fig:figureSI4}
+\end{figure}
 
@@
     \centering
-    \includegraphics[width=0.5\textwidth]{figureSI-3_lmlf_astba2_hist_bypft.pdf}
-    \caption{Histograms of leaf mass per unit basal stem area grouped by the three major plant functional types.
-Vertical lines and shading are means with confidence intervals (same as in Fig.~\ref{fig:figure2})}
-    \label{fig:figureSI3}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.5\textwidth]{figureSI-4_lalf_astba2_hist_bypft.pdf}
-    \caption{Histograms of leaf area per unit basal stem area grouped by the three major plant functional types.
-Vertical lines and shading are means with confidence intervals (same as in Fig.~\ref{fig:figure2})}
-    \label{fig:figureSI4}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
     \includegraphics[width=0.99\textwidth]{figureSI-5_mlfmst_byht_pft.pdf}
@@
     \label{fig:figureSI5}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{figureSI-6_mrt_mso_bypft.pdf}
-    \caption{Near-isometric scaling of above-ground with below-ground biomass, and no effects of plant functional type. The black line is a 1-1 line.}
-    \label{fig:figureSI6}
 \end{figure}

### 2014-10-10 manuscript/draftpaper.tex
more refs

@@
 
-\item For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception (Lang 1991, Stenberg 2006).
+\item For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception  \cite{lang_application_1991, chen_defining_1992}.
 \end{itemize}
@@
 
-\item How do current DGVMs work?
+\item We found coordination between an important leaf trait and whole-plant biomass partitioning. This has implications for global vegetation models.
+
+\item How is allocation implemented in current DGVMs?
 Either constant allocation fraction, target allometric equations, or optimization \cite{de2014does}.
@@
 %------------------------------------------------------------------------------------%
-\section{Lit notes}
-
-\cite{friend_carbon_2014} showed that 30\% of the uncertainty in global vegetation carbon simulated with seven vegetation models is due to the residence time of carbon in vegetation.
+% \section{Lit notes}
+% 
+% \cite{friend_carbon_2014} showed that 30\% of the uncertainty in global vegetation carbon simulated with seven vegetation models is due to the residence time of carbon in vegetation.
 This residence time depends on turnover rates of the various plant pools, and hence also on the distribution of biomass between these pools.
 For example, vegetation with a higher leaf mass fraction will have a lower overall residence time of carbon.
-
-\cite{de2014does} "We have shown that allocation approaches that are constrained by biomass fractions (such as functional relationships) were more successful at capturing observed trends, and were generally more robust, than approaches based on allocation coefficients."
-
-\cite{friedlingstein_toward_1999} make no mention of plant functional type differences or the need to vary allocation by pft.
-
-Givnish 2002 discusses costs and benefits of evergreen vs.
+% 
+% \cite{de2014does} "We have shown that allocation approaches that are constrained by biomass fractions (such as functional relationships) were more successful at capturing observed trends, and were generally more robust, than approaches based on allocation coefficients."
+% 
+% \cite{friedlingstein_toward_1999} make no mention of plant functional type differences or the need to vary allocation by pft.
+% 
+% \cite{givnish_adaptive_2002} discusses costs and benefits of evergreen vs.
 deciduous habit, in terms of constructing a growth model that takes into account photosynthesis, leaf lifespan, belowground costs.
@@
 Instead he simply states that 'leaf allocation ratio' relates to leaf lifespan, which of course is only true if total leaf mass is constant.
-
-Callaway et al 1994, example of lower leaf mass / sapwood area for desert population of PIPO vs.
+% 
+% \cite{callaway_biomass_1994} example of lower leaf mass / sapwood area for desert population of PIPO vs.
 montane population.
 For pointing out small-scale climatic effects on allometry (that may not be obvious in global scale analyses).
-
-\cite{poorter_biomass_2012} found a slightly larger LMF in gymnosperms, but the difference was small and could not split between deciduous and evergreen angiosperms due to a lack of data.
+% 
+% \cite{poorter_biomass_2012} found a slightly larger LMF in gymnosperms, but the difference was small and could not split between deciduous and evergreen angiosperms due to a lack of data.
 They argue that "most conifers retain leaves for two or more years.
@@
 evergreen).
-
-\cite{enquist_global_2002} report 2.6 times higher leaf mass at given stem mass for conifers compared to angiosperms, but did not separate further, nor interpreted this difference in terms of SLA.
+% 
+% \cite{enquist_global_2002} report 2.6 times higher leaf mass at given stem mass for conifers compared to angiosperms, but did not separate further, nor interpreted this difference in terms of SLA.
 Their interpretation is 'that conifers typically retain three cohorts of leaves that have less well-developed [...] mesophyll'.
 Not sure how that latter bit makes sense, and leaf lifespan interpretation is no good; because it still begs the question why would conifers not simply have 1/3 less leaf mass per cohort?
-
-\cite{mccarthy_consistency_2007} further demonstrate large difference in LMF between angiosperms and conifers, and note that a large chunk of the residuals is correlated with SLA and LL.
+% 
+% \cite{mccarthy_consistency_2007} further demonstrate large difference in LMF between angiosperms and conifers, and note that a large chunk of the residuals is correlated with SLA and LL.
 They interpret this rather vaguely as 'leaf morphology may constrain leaf mass distribution'.
 Their SLA was not on the same individuals, but they could have calculated PFT average LAR anyway, but did not.
-
-\cite{oneill_comparative_1981} reported much higher leaf biomass for gymnosperms vs.
+% 
+% \cite{oneill_comparative_1981} reported much higher leaf biomass for gymnosperms vs.
 angiosperms, at a given woody biomass increment.
 This means (as Niklas reported) a much higher productivity per unit leaf biomass for angiosperms.
-
-Serious lack of biomass data to constrain models.
+% 
+% Serious lack of biomass data to constrain models.
 Studies like \cite{wolf_forest_2011} resort to studying individual tree allometry based on dividing stand biomass by stand density.
 \cite{wolf_allometric_2010} did not find any differences in biomass fractions (root,stem,leaf) between angiosperms and gymnosperms (but again this was based on stand biomass data, but still this is strange!).
-
-\cite{hui_near_2014} Isometric scaling of above vs. belowground biomass in chinese forest biomass database.
-
-\cite{sardans_tree_2013} report similiar LMF for evergreen angio. and gymnosperms, and ca. 40\% less leaf mass for decid. angio. Local study, 3000 plots in Catalonia.
-
-\cite{reich_temperature_2014} "On average, when standardized for MAT and Mstem, conifers tended to have more Mfol and less Mroot than angiosperm forests".
+% 
+% \cite{hui_near_2014} Isometric scaling of above vs. belowground biomass in chinese forest biomass database.
+% 
+% \cite{sardans_tree_2013} report similiar LMF for evergreen angio. and gymnosperms, and ca. 40\% less leaf mass for decid. angio. Local study, 3000 plots in Catalonia.
+% 
+% \cite{reich_temperature_2014} "On average, when standardized for MAT and Mstem, conifers tended to have more Mfol and less Mroot than angiosperm forests".
 Their Fig.
@@
 But what about conifers?
-
-\cite{cairns_root_1997} found no difference in root-shoot ratio between gymno and angio.
-
-\cite{delucia_climate-driven_2000} LA/SA decreases with summer VPD for Pinus but not non-Pinus gymno.
+% 
+% \cite{cairns_root_1997} found no difference in root-shoot ratio between gymno and angio.
+% 
+% \cite{delucia_climate-driven_2000} LA/SA decreases with summer VPD for Pinus but not non-Pinus gymno.
 Also Pinus has much lower LA/SA than non-Pinus; suggesting that we should separate gymnosperms into those groups as well.
-
-\cite{grier_notes:_1974} early study on leaf mass proportionality to sapwood area (and uses mass not area).
+% 
+% 
+% \cite{grier_notes:_1974} early study on leaf mass proportionality to sapwood area (and uses mass not area).
 

### 2014-10-13 R/baadvarpart.R
Added variance partioning tables and described in text

@@ -1,28 +1,28 @@
 
 
-
-
-
+# Variance partitioning to fixed and random effects 
 source("load.R")
 source("R/preparedataset.R")
 source("R/rsquaredglmm.R")
 
+library(lmerTest)
 
-
-r2stepwise <- function(m){
+# Sequential R2, backwards elimination of all fixed effects.
+# ((not used at the moment))
+r2stepwise <- function(m, data){
   
   tr <- attributes(terms(m))$term.labels
   ii <- lapply(1:length(tr), function(x)1:x)
   yvar <- rownames(attributes(terms(m))$factors)[1]
   
-  fitM <- function(i){
+  fitM <- function(i, dat){
     tr <- c(tr[i], "(1|Group)")
     f <- as.formula(paste(yvar, "~", paste(tr, collapse=" + ")))
     M <- lmer(formula=f, data=dat)
   return(M)
   }
   
-  models <- lapply(ii, fitM)
+  models <- lapply(ii, fitM, dat=data)
   rsq <- do.call(rbind, lapply(models, r.squared))
 rownames(rsq) <- tr
 return(rsq)
@@ -37,40 +37,90 @@ plotr2 <- function(x, which=c("Marginal","Conditional"),...){
 
 
 
-
 # MLF / AST
-dat <- droplevels(subset(dataset2, !is.na(h.t) & !is.na(pft) & !is.na(lmlf_astba2)))
-fullmodel <- lmer(lmlf_astba2 ~ log10(h.t)*pft*vegetation + (1|Group), data=dat)
+dat_mlf <- droplevels(subset(dataset2, !is.na(h.t) & !is.na(pft) & !is.na(lmlf_astba2)))
+dat_alf <- droplevels(subset(dataset2, !is.na(h.t) & !is.na(pft) & !is.na(lalf_astba2)))
+
+lmer_mlf_0 <- lmer(lmlf_astba2 ~ log10(h.t) + (1|Group), data=dat_mlf)
+lmer_mlf_1 <- lmer(lmlf_astba2 ~ log10(h.t)*bortemptrop + (1|Group), data=dat_mlf)
[…]

### 2014-10-13 manuscript/draftpaper.Rnw
Made Rnw to replace .tex, added table (for test)
Added variance partioning tables and described in text

--- 2014-10-10 manuscript/draftpaper.tex
@@
 
-\author{Remko Duursma, Daniel Falster and BAAD data contributors}
+\author{Remko Duursma, Daniel Falster and BAAD contributors}
 
 \maketitle
+
+%--------------------------------------------------------------------------------------------%
+
+<<preload, echo=FALSE>>=
+load("tables/Tables_varpart.RData")
+@
 
@@
 
-\item Variance partitioning with R2 for linear mixed-effects models as reported by XXX.
+\item Variance partitioning with marginal (fixed effects) and conditional (fixed and random effects) R2 for linear mixed-effects models as reported by \cite{nakagawa_general_2013}.
 
@@
 \section{Results}
-
-Note: stats done as well, but not yet incorporated into the manuscript.
 
@@
 \item We do note that there was substantial variation between individual plants (Figs.~\ref{fig:figureSI3} and~\ref{fig:figureSI3}).
-Nonetheless, PFT explained most variation in MF/Asba (to be added).
-If we also included SLM in the model, PFT was no longer significant.
-PFT explained very little variation when we partitioned the variance in AF/Asba.
+We estimated the variance explained by PFT, plant height, and biome in MF/Asba using a linear mixed-effects model (see Table~\ref{varpartpipe}).
+We found that plant height and biome explained only \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[1],2)}\%, but when PFT was added to the model, the variance explained increased to \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[2],2)}\%.
+If we used specific leaf mass instead of PFT, the R2 was similar (\Sexpr{100*signif(Table_pipemodel_varpart$Marginal[3],2)}\%), showing that PFT affected MF/Asba largely through differences in SLM.
+Consistent with this, PFT, plant height and biome explained very little variance in ALF/Astba (Table~\ref{varpartpipe}).
 
@@
 For the leaf area ratio, pfts were generally closer together except for small evergreen gymnosperms (< 1m height) for unknown reasons.
-Variance partitioning also confirmed that the differences between PFT could be largely explained by their differences in SLM (not shown yet!).
 
-\item We found no effect of biome on the ratio MF/Asba or AF/Asba (tropical, temperate, boreal) (Fig.~\ref{fig:figureSI2}).
+\item Results for variance partitioning of LMF and LAR gave comparable results (Table~\ref{varpartpipe}): PFT increased the variance explained in LMF from \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[1],2)} to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[2],2)}\%.
+However, SLM was not a good substitute for PFT, as the explained variance decreased to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[3],2)}\%.
+The leaf area ratio, similarly to ALF/Astba, did not depend strongly on PFT or SLA.
+
+\item Although biome was a significant predictor in all variables tested, it consistently explained very little variation (Table~\ref{varpartpipe} and~\ref{varpartlmflar}).
+Pair-wise comparisons between biomes within PFT never showed significant differences for MF/Asba or AL/Asba (for example boreal vs.
+temperate evergreen gymnosperm) (Fig.~\ref{fig:figureSI2}).
 
@@
 No effect of climate is good news for models that parameterize the globe by PFT, as this seems sufficient, based on our dataset.
+Are there experimental data that show temperature effects on allocation?
+What about water availability?
 
@@
 
+\section{Tables}
+
+<<echo=FALSE, message=FALSE, warning=FALSE>>=
+library(xtable)
+@
+
+<<results='asis', echo=FALSE>>=
+print(xtable(Table_pipemodel_varpart, label="varpartpipe", 
+             caption="Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.
+The marginal R2 is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type.
+The conditional R2 is that explained by fixed and random effects (species within study).
+All fixed effects were highly significant (P \\textless 0.001).
+In addition to the fixed effects shown, all interactions were added to each of the models."),
+      include.rownames=FALSE, caption.placement="top")
+@
+
+<<results='asis', echo=FALSE>>=
+load("tables/Tables_varpart.RData")
+print(xtable(Table_LMFLAR_varpart, label="varpartlmflar",
+      caption="Explained variance in the leaf mass fraction (leaf mass / aboveground biomass) and leaf area ratio (leaf area / aboveground biomass).
+See Table~\\ref{varpartpipe} for more explanation.
+For plant height, we also added the squared term as a fixed effect, and the interaction with PFT or SLM, because the relationship was non-linear."),
+      include.rownames=FALSE, caption.placement="top")
+@
+
+\clearpage
 \section{Figures}

### 2014-10-13 manuscript/draftpaper.tex
Added variance partioning tables and described in text

@@
-\documentclass[a4paper]{article}
+\documentclass[a4paper]{article}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 
@@
 \bibliographystyle{plain}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 
-\author{Remko Duursma, Daniel Falster and BAAD data contributors}
+\author{Remko Duursma, Daniel Falster and BAAD contributors}
 
 \maketitle
+
+%--------------------------------------------------------------------------------------------%
 
@@
 \item We do note that there was substantial variation between individual plants (Figs.~\ref{fig:figureSI3} and~\ref{fig:figureSI3}).
-Nonetheless, PFT explained most variation in MF/Asba (to be added).
-If we also included SLM in the model, PFT was no longer significant.
-PFT explained very little variation when we partitioned the variance in AF/Asba.
+We estimated the variance explained by PFT, plant height, and biome in MF/Asba using a linear mixed-effects model (see Table~\ref{varpartpipe}).
+We found that plant height and biome explained only 0.85\%, but when PFT was added to the model, the variance explained increased to 2.6\%.
+If we used specific leaf mass instead of PFT, the R2 was similar (29\%), showing that PFT affected MF/Asba largely through differences in SLM.
+Consistent with this, PFT, plant height and biome explained very little variance in ALF/Astba (Table~\ref{varpartpipe}).
 
@@
 For the leaf area ratio, pfts were generally closer together except for small evergreen gymnosperms (< 1m height) for unknown reasons.
-Variance partitioning also confirmed that the differences between PFT could be largely explained by their differences in SLM (not shown yet!).
 
-\item We found no effect of biome on the ratio MF/Asba or AF/Asba (tropical, temperate, boreal) (Fig.~\ref{fig:figureSI2}).
+\item Results for variance partitioning of LMF and LAR gave comparable results (Table~\ref{varpartpipe}): PFT increased the variance explained in LMF from 65 to 66\%.
+However, SLM was not a good substitute for PFT, as the explained variance decreased to 75\%.
+The leaf area ratio, similarly to ALF/Astba, did not depend strongly on PFT or SLA.
+
+\item Although biome was a significant predictor in all variables tested, it consistently explained very little variation (Table~\ref{varpartpipe} and~\ref{varpartlmflar}).
+Pair-wise comparisons between biomes within PFT never showed significant differences for MF/Asba or AL/Asba (for example boreal vs.
+temperate evergreen gymnosperm) (Fig.~\ref{fig:figureSI2}).
 
@@
 
+\section{Tables}
+
+% latex table generated in R 3.1.1 by xtable 1.7-4 package
+% Mon Oct 13 15:43:27 2014
+\begin{table}[ht]
+\centering
+\caption{Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.
+The marginal R2 is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type.
+The conditional R2 is that explained by fixed and random effects (species within study).
+All fixed effects were highly significant (P \textless 0.001).
+In addition to the fixed effects shown, all interactions were added to each of the models.}
+\label{varpartpipe}
+\begin{tabular}{llrr}
+  \hline
+Variable & Predictors & Marginal & Conditional \\ 
+  \hline
+LMF\_ASTBA & H & 0.01 & 0.53 \\ 
+   & H, B & 0.03 & 0.55 \\ 
+   & PFT, H, B & 0.29 & 0.64 \\ 
+   & SLM, H, B & 0.31 & 0.66 \\ 
+  LAR\_ASTBA & H & 0.01 & 0.52 \\ 
+   & H, B & 0.05 & 0.53 \\ 
+   & PFT, H, B & 0.06 & 0.54 \\ 
+   & SLM, H, B & 0.12 & 0.56 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+% latex table generated in R 3.1.1 by xtable 1.7-4 package
+% Mon Oct 13 15:43:27 2014
+\begin{table}[ht]
+\centering
+\caption{Explained variance in the leaf mass fraction (leaf mass / aboveground biomass) and leaf area ratio (leaf area / aboveground biomass).
+See Table~\ref{varpartpipe} for more explanation.
+For plant height, we also added the squared term as a fixed effect, and the interaction with PFT or SLM, because the relationship was non-linear.}
+\label{varpartlmflar}
+\begin{tabular}{llrr}
+  \hline
+Variable & Predictors & Marginal & Conditional \\ 
+  \hline
+LMF & H & 0.65 & 0.89 \\ 
+   & H, B & 0.66 & 0.89 \\ 
+   & PFT, H, B & 0.75 & 0.88 \\ 
+   & SLM, H, B & 0.68 & 0.89 \\ 
+  LAR & H & 0.69 & 0.89 \\ 
+   & H, B & 0.73 & 0.90 \\ 
+   & PFT, H, B & 0.76 & 0.90 \\ 
+   & SLM, H, B & 0.77 & 0.92 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+\clearpage
 \section{Figures}

### 2014-10-13 manuscript/draftpaper.Rnw
Added tables of counts to manuscript, showing sample sizes used by pft, biome (plants and species)

@@
 \clearpage
+\section{Supporting Tables}
+
+<<results='asis', echo=FALSE>>=
+load("tables/Table_counts.RData")
+print(xtable(tab_mlfastba, align="rlll|l",
+             caption="mlf/asba"
+             ), hline.after=c(0,3), caption.placement="top")
+@
+
+<<results='asis', echo=FALSE>>=
+print(xtable(tab_alfastba, align="rlll|l",
+             caption="alf/asba"
+             ), hline.after=c(1,3), caption.placement="top")
+@
+
+<<results='asis', echo=FALSE>>=
+print(xtable(tab_lmf, align="rlll|l",
+             caption="LMF"
+             ), hline.after=c(1,3), caption.placement="top")
+@
+
+<<results='asis', echo=FALSE>>=
+print(xtable(tab_lar, align="rlll|l",
+             caption="LAR"
+             ), hline.after=c(1,3), caption.placement="top")
+@
+
+\clearpage
 \section{Figures}

### 2014-10-13 manuscript/draftpaper.tex
Added tables of counts to manuscript, showing sample sizes used by pft, biome (plants and species)

@@
 We estimated the variance explained by PFT, plant height, and biome in MF/Asba using a linear mixed-effects model (see Table~\ref{varpartpipe}).
-We found that plant height and biome explained only 0.85\%, but when PFT was added to the model, the variance explained increased to 2.6\%.
-If we used specific leaf mass instead of PFT, the R2 was similar (29\%), showing that PFT affected MF/Asba largely through differences in SLM.
+We found that plant height and biome explained only 0.86\%, but when PFT was added to the model, the variance explained increased to 2.6\%.
+If we used specific leaf mass instead of PFT, the R2 was similar (28\%), showing that PFT affected MF/Asba largely through differences in SLM.
 Consistent with this, PFT, plant height and biome explained very little variance in ALF/Astba (Table~\ref{varpartpipe}).
@@
 
-\item Results for variance partitioning of LMF and LAR gave comparable results (Table~\ref{varpartpipe}): PFT increased the variance explained in LMF from 65 to 66\%.
-However, SLM was not a good substitute for PFT, as the explained variance decreased to 75\%.
+\item Results for variance partitioning of LMF and LAR gave comparable results (Table~\ref{varpartpipe}): PFT increased the variance explained in LMF from 60 to 64\%.
+However, SLM was not a good substitute for PFT, as the explained variance decreased to 73\%.
 The leaf area ratio, similarly to ALF/Astba, did not depend strongly on PFT or SLA.
@@
 % latex table generated in R 3.1.1 by xtable 1.7-4 package
-% Mon Oct 13 15:43:27 2014
+% Mon Oct 13 21:23:52 2014
 \begin{table}[ht]
@@
 LMF\_ASTBA & H & 0.01 & 0.53 \\ 
-   & H, B & 0.03 & 0.55 \\ 
-   & PFT, H, B & 0.29 & 0.64 \\ 
-   & SLM, H, B & 0.31 & 0.66 \\ 
+   & H, B & 0.03 & 0.54 \\ 
+   & PFT, H, B & 0.28 & 0.64 \\ 
+   & SLM, H, B & 0.31 & 0.65 \\ 
   LAR\_ASTBA & H & 0.01 & 0.52 \\ 
@@
   \hline
-LMF & H & 0.65 & 0.89 \\ 
-   & H, B & 0.66 & 0.89 \\ 
-   & PFT, H, B & 0.75 & 0.88 \\ 
-   & SLM, H, B & 0.68 & 0.89 \\ 
-  LAR & H & 0.69 & 0.89 \\ 
-   & H, B & 0.73 & 0.90 \\ 
+LMF & H & 0.60 & 0.88 \\ 
+   & H, B & 0.64 & 0.89 \\ 
+   & PFT, H, B & 0.73 & 0.88 \\ 
+   & SLM, H, B & 0.69 & 0.89 \\ 
+  LAR & H & 0.70 & 0.89 \\ 
+   & H, B & 0.74 & 0.90 \\ 
    & PFT, H, B & 0.76 & 0.90 \\ 
@@
    \hline
+\end{tabular}
+\end{table}
+
+\clearpage
+\section{Supporting Tables}
+
+% latex table generated in R 3.1.1 by xtable 1.7-4 package
+% Mon Oct 13 21:23:52 2014
+\begin{table}[ht]
+\centering
+\caption{mlf/asba} 
+\begin{tabular}{rlll|l}
+  & boreal & temperate & tropical & Sum \\ 
+  \hline
+DA & 308 (6) & 2067 (93) & 162 (14) & 2537 (113) \\ 
+  EA &  & 1064 (58) & 1696 (173) & 2760 (231) \\ 
+  EG & 1010 (10) & 1831 (31) &  & 2841 (41) \\ 
+   \hline
+Sum & 1318 (16) & 4962 (182) & 1858 (187) & 8138 (385) \\ 
+  \end{tabular}
+\end{table}
+
+% latex table generated in R 3.1.1 by xtable 1.7-4 package
+% Mon Oct 13 21:23:52 2014
+\begin{table}[ht]
+\centering
+\caption{alf/asba} 
+\begin{tabular}{rlll|l}
+  & boreal & temperate & tropical & Sum \\ 
+ DA &  & 1572 (70) & 35 (7) & 1607 (77) \\ 
+   \hline
+EA &  & 934 (53) & 1370 (118) & 2304 (171) \\ 
+  EG & 458 (8) & 1074 (21) &  & 1532 (29) \\ 
+   \hline
+Sum & 458 (8) & 3580 (144) & 1405 (125) & 5443 (277) \\ 
+  \end{tabular}
+\end{table}
+
+% latex table generated in R 3.1.1 by xtable 1.7-4 package
+% Mon Oct 13 21:23:52 2014
+\begin{table}[ht]
+\centering
+\caption{LMF} 
+\begin{tabular}{rlll|l}
+  & boreal & temperate & tropical & Sum \\ 
+ DA & 184 (4) & 1975 (94) & 162 (14) & 2321 (112) \\ 
+   \hline
+EA &  & 1271 (68) & 1674 (169) & 2945 (237) \\ 
+  EG & 669 (7) & 1157 (24) &  & 1826 (31) \\ 
+   \hline
+Sum & 853 (11) & 4403 (186) & 1836 (183) & 7092 (380) \\ 
+  \end{tabular}
+\end{table}
+
+% latex table generated in R 3.1.1 by xtable 1.7-4 package
+% Mon Oct 13 21:23:52 2014
+\begin{table}[ht]
+\centering
+\caption{LAR} 
+\begin{tabular}{rlll|l}
+  & boreal & temperate & tropical & Sum \\ 
+ DA &  & 1452 (68) & 35 (7) & 1487 (75) \\ 
+   \hline
+EA &  & 1124 (59) & 1320 (112) & 2444 (171) \\ 
+  EG & 321 (7) & 508 (15) &  & 829 (22) \\ 
+   \hline
+Sum & 321 (7) & 3084 (142) & 1355 (119) & 4760 (268) \\ 
   \end{tabular}

### 2014-10-14 manuscript/draftpaper.Rnw
Added extra MAP, MAT figure, colored by vegetation (SI). SI tables captions updated, code tidied, and moved to back of ms.

--- 2014-10-13 manuscript/draftpaper.Rnw
@@
 <<preload, echo=FALSE>>=
+library(knitr)
+opts_chunk$set(echo=FALSE)
+
 load("tables/Tables_varpart.RData")
+baad <- readRDS("../data/baad.rds")
 @
@@
 \begin{itemize}
-\item BAAD database, XXX observations etc.
+\item BAAD database (Falster et al., in review), including a total of \Sexpr{nrow(baad$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), aboveground woody biomass ($M_W$), total root biomass ($M_R$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
+1.3m) ($A_{Sbh}$).
+For each analysis, we therefore used different subsets of the data (far fewer observations are available for root biomass, in particular), sample sizes are included in the caption of each figure (\emph{to be added}).
+We further excluded deciduous gymnosperms (since very few data were available), and glasshouse and common garden studies, including only field grown woody plants (including plantations).
 
-\item Removed deciduous gymnosperms, we have a few of those in BAAD (6 species / studyName combinations), but only one includes estimates of leaf area and mass.
-Also removed glasshouse and common garden studies, including only field grown woody plants (including plantations).
+\item We considered three plant functional types : evergreen angiosperms, evergreen gymnosperms, and decidous angiosperms.
+All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes (decided by contributing author, but checked with MAT and latitude).
 
@@
 
-\item For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception  \cite{lang_application_1991, chen_defining_1992}.
+\item For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception  \cite{lang_application_1991, chen_defining_1992}, for comparison to flat leaves.
 \end{itemize}
@@
 
-<<results='asis', echo=FALSE>>=
+<<results='asis'>>=
 print(xtable(Table_pipemodel_varpart, label="varpartpipe", 
@@
 
-<<results='asis', echo=FALSE>>=
+<<results='asis'>>=
 load("tables/Tables_varpart.RData")
@@
 \section{Figures}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics{Figure1_MAPMAT_baad_vs_worldclim.pdf}
+    \caption{Mean annual precipitation and mean annual temperature for all included studies.
+Hexagons are terrestrial observations extracted from Worldclim (darker bins indicating more observations).
+Note that Worldclim observations include land areas where woody vegetation does not occur (desert, tundra).
+The deciduous gymnosperms were not included in any analysis in this paper, but are shown here.
+}
+    \label{fig:figureworldclim}
+\end{figure}
 
@@
 \section{Supporting Information Figures}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{figureSI-8_MAPMAT_vegetation.pdf}
+    \caption{Mean annual precipitation and temperature for all contributed datasets, by vegetation type. Note that in this study, we did not include glasshouse studies.}
+    \label{fig:figureSI8}
+\end{figure}
 
@@
 \clearpage
+\section{Supporting Information Tables}
+
+<<tablepreload>>=
+load("tables/Table_counts.RData")
+printit <- function(x, caption){
+  print(xtable(x, align="rlll|l",
+               caption=caption
+               ), hline.after=c(0,3), caption.placement="top")
+
+}
+@
+
+<<results='asis'>>=
+printit(tab_mlfastba, caption="Number of observations used in the analysis of Mf/As. Individual plants (number of species).")
+@
+
+<<results='asis'>>=
+printit(tab_alfastba, caption="Number of observations used in the analysis of Af/As. Individual plants (number of species).")
+@
+
+<<results='asis'>>=
+printit(tab_lmf, caption="Number of observations used in the analysis of leaf mass fraction (LMF). Individual plants (number of species).")
+@
+
+<<results='asis'>>=
+printit(tab_lar, caption="Number of observations used in the analysis of leaf area ratio (LAR). Individual plants (number of species).")
+@
+
+\clearpage
 \bibliography{allometry}

### 2014-10-14 manuscript/draftpaper.tex
Added extra MAP, MAT figure, colored by vegetation (SI). SI tables captions updated, code tidied, and moved to back of ms.

@@
 \begin{itemize}
-\item BAAD database, XXX observations etc.
+\item BAAD database (Falster et al., in review), including a total of 20848 observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), aboveground woody biomass ($M_W$), total root biomass ($M_R$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
+1.3m) ($A_{Sbh}$).
+For each analysis, we therefore used different subsets of the data (far fewer observations are available for root biomass, in particular), sample sizes are included in the caption of each figure (\emph{to be added}).
+We further excluded deciduous gymnosperms (since very few data were available), and glasshouse and common garden studies, including only field grown woody plants (including plantations).
 
-\item Removed deciduous gymnosperms, we have a few of those in BAAD (6 species / studyName combinations), but only one includes estimates of leaf area and mass.
-Also removed glasshouse and common garden studies, including only field grown woody plants (including plantations).
+\item We considered three plant functional types : evergreen angiosperms, evergreen gymnosperms, and decidous angiosperms.
+All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes (decided by contributing author, but checked with MAT and latitude).
 
@@
 % latex table generated in R 3.1.1 by xtable 1.7-4 package
-% Mon Oct 13 21:23:52 2014
+% Tue Oct 14 13:17:40 2014
 \begin{table}[ht]
@@
 % latex table generated in R 3.1.1 by xtable 1.7-4 package
-% Mon Oct 13 21:23:52 2014
+% Tue Oct 14 13:17:41 2014
 \begin{table}[ht]
@@
 \clearpage
-\section{Supporting Tables}
+\section{Figures}
 
-% latex table generated in R 3.1.1 by xtable 1.7-4 package
-% Mon Oct 13 21:23:52 2014
-\begin{table}[ht]
+\begin{figure}[h!]
     \centering
-\caption{mlf/asba} 
-\begin{tabular}{rlll|l}
-  & boreal & temperate & tropical & Sum \\ 
-  \hline
-DA & 308 (6) & 2067 (93) & 162 (14) & 2537 (113) \\ 
-  EA &  & 1064 (58) & 1696 (173) & 2760 (231) \\ 
-  EG & 1010 (10) & 1831 (31) &  & 2841 (41) \\ 
-   \hline
-Sum & 1318 (16) & 4962 (182) & 1858 (187) & 8138 (385) \\ 
-  \end{tabular}
-\end{table}
-
-% latex table generated in R 3.1.1 by xtable 1.7-4 package
-% Mon Oct 13 21:23:52 2014
-\begin{table}[ht]
-\centering
-\caption{alf/asba} 
-\begin{tabular}{rlll|l}
-  & boreal & temperate & tropical & Sum \\ 
- DA &  & 1572 (70) & 35 (7) & 1607 (77) \\ 
-   \hline
-EA &  & 934 (53) & 1370 (118) & 2304 (171) \\ 
-  EG & 458 (8) & 1074 (21) &  & 1532 (29) \\ 
-   \hline
-Sum & 458 (8) & 3580 (144) & 1405 (125) & 5443 (277) \\ 
-  \end{tabular}
-\end{table}
-
-% latex table generated in R 3.1.1 by xtable 1.7-4 package
-% Mon Oct 13 21:23:52 2014
-\begin{table}[ht]
-\centering
-\caption{LMF} 
-\begin{tabular}{rlll|l}
-  & boreal & temperate & tropical & Sum \\ 
- DA & 184 (4) & 1975 (94) & 162 (14) & 2321 (112) \\ 
-   \hline
-EA &  & 1271 (68) & 1674 (169) & 2945 (237) \\ 
-  EG & 669 (7) & 1157 (24) &  & 1826 (31) \\ 
-   \hline
-Sum & 853 (11) & 4403 (186) & 1836 (183) & 7092 (380) \\ 
-  \end{tabular}
-\end{table}
-
-% latex table generated in R 3.1.1 by xtable 1.7-4 package
-% Mon Oct 13 21:23:52 2014
-\begin{table}[ht]
-\centering
-\caption{LAR} 
-\begin{tabular}{rlll|l}
-  & boreal & temperate & tropical & Sum \\ 
- DA &  & 1452 (68) & 35 (7) & 1487 (75) \\ 
-   \hline
-EA &  & 1124 (59) & 1320 (112) & 2444 (171) \\ 
-  EG & 321 (7) & 508 (15) &  & 829 (22) \\ 
-   \hline
-Sum & 321 (7) & 3084 (142) & 1355 (119) & 4760 (268) \\ 
-  \end{tabular}
-\end{table}
-
-\clearpage
-\section{Figures}
+    \includegraphics{Figure1_MAPMAT_baad_vs_worldclim.pdf}
+    \caption{Mean annual precipitation and mean annual temperature for all included studies.
+Hexagons are terrestrial observations extracted from Worldclim (darker bins indicating more observations).
+Note that Worldclim observations include land areas where woody vegetation does not occur (desert, tundra).
+The deciduous gymnosperms were not included in any analysis in this paper, but are shown here.
+}
+    \label{fig:figureworldclim}
+\end{figure}
 
@@
 \clearpage
+\section{Supporting Information Tables}
+
+% latex table generated in R 3.1.1 by xtable 1.7-4 package
+% Tue Oct 14 13:17:41 2014
+\begin{table}[ht]
+\centering
+\caption{Number of observations used in the analysis of Mf/As. Individual plants (number of species).} 
+\begin{tabular}{rlll|l}
+  & boreal & temperate & tropical & Sum \\ 
+  \hline
+DA & 308 (6) & 2067 (93) & 162 (14) & 2537 (113) \\ 
+  EA &  & 1064 (58) & 1696 (173) & 2760 (231) \\ 
+  EG & 1010 (10) & 1831 (31) &  & 2841 (41) \\ 
+   \hline
+Sum & 1318 (16) & 4962 (182) & 1858 (187) & 8138 (385) \\ 
+  \end{tabular}
+\end{table}
+
+% latex table generated in R 3.1.1 by xtable 1.7-4 package
+% Tue Oct 14 13:17:41 2014
+\begin{table}[ht]
+\centering
+\caption{Number of observations used in the analysis of Af/As. Individual plants (number of species).} 
+\begin{tabular}{rlll|l}
+  & boreal & temperate & tropical & Sum \\ 
+  \hline
+DA &  & 1572 (70) & 35 (7) & 1607 (77) \\ 
+  EA &  & 934 (53) & 1370 (118) & 2304 (171) \\ 
+  EG & 458 (8) & 1074 (21) &  & 1532 (29) \\ 
+   \hline
+Sum & 458 (8) & 3580 (144) & 1405 (125) & 5443 (277) \\ 
+  \end{tabular}
+\end{table}
+
+% latex table generated in R 3.1.1 by xtable 1.7-4 package
+% Tue Oct 14 13:17:41 2014
+\begin{table}[ht]
+\centering
+\caption{Number of observations used in the analysis of leaf mass fraction (LMF). Individual plants (number of species).} 
+\begin{tabular}{rlll|l}
+  & boreal & temperate & tropical & Sum \\ 
+  \hline
+DA & 184 (4) & 1975 (94) & 162 (14) & 2321 (112) \\ 
+  EA &  & 1271 (68) & 1674 (169) & 2945 (237) \\ 
+  EG & 669 (7) & 1157 (24) &  & 1826 (31) \\ 
+   \hline
+Sum & 853 (11) & 4403 (186) & 1836 (183) & 7092 (380) \\ 
+  \end{tabular}
+\end{table}
+
+% latex table generated in R 3.1.1 by xtable 1.7-4 package
+% Tue Oct 14 13:17:41 2014
+\begin{table}[ht]
+\centering
+\caption{Number of observations used in the analysis of leaf area ratio (LAR). Individual plants (number of species).} 
+\begin{tabular}{rlll|l}
+  & boreal & temperate & tropical & Sum \\ 
+  \hline
+DA &  & 1452 (68) & 35 (7) & 1487 (75) \\ 
+  EA &  & 1124 (59) & 1320 (112) & 2444 (171) \\ 
+  EG & 321 (7) & 508 (15) &  & 829 (22) \\ 
+   \hline
+Sum & 321 (7) & 3084 (142) & 1355 (119) & 4760 (268) \\ 
+  \end{tabular}
+\end{table}
+
+\clearpage
 \bibliography{allometry}

### 2014-10-14 manuscript/draftpaper.Rnw
Results update

@@
 
-\item We found strong correlations between basal stem area and leaf mass across the entire dataset, with a clear separation between the three major plant functional types (Fig.~\ref{fig:figure1}).
-We analyzed the ratio MF/Asba, and found that PFT explained by far the most variation (variables tested: plant height, biome (boreal,tropical,temperate), PFT, and all interactions).
+\item We found strong correlations between basal stem area ($A_S$) and leaf mass ($M_F$) across the entire dataset, with a clear separation between the three major plant functional types (Fig.~\ref{fig:figure1}).
 
-\item Because of weak plant height effects, we calculated the average MF/Asba for the three pfts; there was a ca.
+\item Because of weak plant height effects, we calculated average $M_F/A_S$ across the entire dataset by the three main PFTs.
+We found a ca.
 3-fold difference between evergreen gymnosperms and deciduous angiosperms, with evergreen angiosperms taking an intermediate value (Fig.~\ref{fig:figure2}).
-These differences were approximately proportional to the average SLM for the PFT.
+These differences were approximately proportional to the average specific leaf mass (SLM = $M_F / A_F$)for the PFT.
 
-\item When we calculated the ratio of leaf area to basal stem area, we found no difference between PFTs (Fig.~\ref{fig:figure2}), confirming that MF/Asba is approximately proportional to SLM across PFTs.
-(See also Fig.~\ref{fig:figureSI1}).
+\item However, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure2}), confirming that $M_F/A_S$ is approximately proportional to SLM across PFTs (See also Fig.~\ref{fig:figureSI1}).
 
 \item We do note that there was substantial variation between individual plants (Figs.~\ref{fig:figureSI3} and~\ref{fig:figureSI3}).
-We estimated the variance explained by PFT, plant height, and biome in MF/Asba using a linear mixed-effects model (see Table~\ref{varpartpipe}).
-We found that plant height and biome explained only \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[1],2)}\%, but when PFT was added to the model, the variance explained increased to \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[2],2)}\%.
-If we used specific leaf mass instead of PFT, the R2 was similar (\Sexpr{100*signif(Table_pipemodel_varpart$Marginal[3],2)}\%), showing that PFT affected MF/Asba largely through differences in SLM.
-Consistent with this, PFT, plant height and biome explained very little variance in ALF/Astba (Table~\ref{varpartpipe}).
+We estimated the variance explained by PFT, plant height, and biome in $M_F/A_S$ using a linear mixed-effects model (see Table~\ref{varpartpipe}).
+We found that plant height and biome explained only \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[2],2)}\%, but when PFT was added to the model, the variance explained increased to \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[3],2)}\%.
+If we used SLM as a predictor instead of PFT, the $R^2$ was similar (\Sexpr{100*signif(Table_pipemodel_varpart$Marginal[4],2)}\%), showing that PFT affected $M_F/A_S$ largely through differences in SLM.
+Consistent with this, PFT, plant height and biome explained very little variance in $A_F/A_S$ (Table~\ref{varpartpipe}).
 
-\item We found similar results for the whole-plant leaf mass fraction (LMF), where a clear and strong separation was found between plant functional types (Fig.~\ref{fig:figure3}).
-See also Fig.~\ref{fig:figureSI5} for the raw data.
-For the leaf area ratio, pfts were generally closer together except for small evergreen gymnosperms (< 1m height) for unknown reasons.
+\item We found similar results for the above-ground leaf mass fraction (LMF), where a clear and strong separation was found between plant functional types (Fig.~\ref{fig:figure3}, see also Fig.~\ref{fig:figureSI5}).
+For the leaf area ratio (LAR), PFTs were generally closer together, with a notable exception for small evergreen gymnosperms (< 1m height) (Fig.~\ref{fig:figure4}).
 
-\item Results for variance partitioning of LMF and LAR gave comparable results (Table~\ref{varpartpipe}): PFT increased the variance explained in LMF from \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[1],2)} to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[2],2)}\%.
-However, SLM was not a good substitute for PFT, as the explained variance decreased to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[3],2)}\%.
-The leaf area ratio, similarly to ALF/Astba, did not depend strongly on PFT or SLA.
+\item Results for variance partitioning of LMF and LAR gave comparable results (Table~\ref{varpartpipe}): PFT increased the variance explained in LMF from \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[2],2)} to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[3],2)}\%.
+However, SLM was not a good substitute for PFT, as the explained variance decreased to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[4],2)}\%.
+The leaf area ratio, similarly to $A_F/A_S$, did not depend strongly on PFT or SLM.
 
 \item Although biome was a significant predictor in all variables tested, it consistently explained very little variation (Table~\ref{varpartpipe} and~\ref{varpartlmflar}).
-Pair-wise comparisons between biomes within PFT never showed significant differences for MF/Asba or AL/Asba (for example boreal vs.
-temperate evergreen gymnosperm) (Fig.~\ref{fig:figureSI2}).
+Pair-wise comparisons between biomes within PFT mostly did not show significant differences for $M_F/A_S$ or $A_F/A_S$.
+The only exception was a higher $A_F/A_S$ for tropical compared to temperate evergreen angiosperms (Fig.~\ref{fig:figureSI2}).
 

### 2014-10-14 manuscript/draftpaper.tex
Results update

@@
 
-\item We found strong correlations between basal stem area and leaf mass across the entire dataset, with a clear separation between the three major plant functional types (Fig.~\ref{fig:figure1}).
-We analyzed the ratio MF/Asba, and found that PFT explained by far the most variation (variables tested: plant height, biome (boreal,tropical,temperate), PFT, and all interactions).
+\item We found strong correlations between basal stem area ($A_S$) and leaf mass ($M_F$) across the entire dataset, with a clear separation between the three major plant functional types (Fig.~\ref{fig:figure1}).
 
-\item Because of weak plant height effects, we calculated the average MF/Asba for the three pfts; there was a ca.
+\item Because of weak plant height effects, we calculated average $M_F/A_S$ across the entire dataset by the three main PFTs.
+We found a ca.
 3-fold difference between evergreen gymnosperms and deciduous angiosperms, with evergreen angiosperms taking an intermediate value (Fig.~\ref{fig:figure2}).
-These differences were approximately proportional to the average SLM for the PFT.
+These differences were approximately proportional to the average specific leaf mass (SLM = $M_F / A_F$)for the PFT.
 
-\item When we calculated the ratio of leaf area to basal stem area, we found no difference between PFTs (Fig.~\ref{fig:figure2}), confirming that MF/Asba is approximately proportional to SLM across PFTs.
-(See also Fig.~\ref{fig:figureSI1}).
+\item However, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure2}), confirming that $M_F/A_S$ is approximately proportional to SLM across PFTs (See also Fig.~\ref{fig:figureSI1}).
 
 \item We do note that there was substantial variation between individual plants (Figs.~\ref{fig:figureSI3} and~\ref{fig:figureSI3}).
-We estimated the variance explained by PFT, plant height, and biome in MF/Asba using a linear mixed-effects model (see Table~\ref{varpartpipe}).
-We found that plant height and biome explained only 0.86\%, but when PFT was added to the model, the variance explained increased to 2.6\%.
-If we used specific leaf mass instead of PFT, the R2 was similar (28\%), showing that PFT affected MF/Asba largely through differences in SLM.
-Consistent with this, PFT, plant height and biome explained very little variance in ALF/Astba (Table~\ref{varpartpipe}).
+We estimated the variance explained by PFT, plant height, and biome in $M_F/A_S$ using a linear mixed-effects model (see Table~\ref{varpartpipe}).
+We found that plant height and biome explained only 2.6\%, but when PFT was added to the model, the variance explained increased to 28\%.
+If we used SLM as a predictor instead of PFT, the $R^2$ was similar (31\%), showing that PFT affected $M_F/A_S$ largely through differences in SLM.
+Consistent with this, PFT, plant height and biome explained very little variance in $A_F/A_S$ (Table~\ref{varpartpipe}).
 
-\item We found similar results for the whole-plant leaf mass fraction (LMF), where a clear and strong separation was found between plant functional types (Fig.~\ref{fig:figure3}).
-See also Fig.~\ref{fig:figureSI5} for the raw data.
-For the leaf area ratio, pfts were generally closer together except for small evergreen gymnosperms (< 1m height) for unknown reasons.
+\item We found similar results for the above-ground leaf mass fraction (LMF), where a clear and strong separation was found between plant functional types (Fig.~\ref{fig:figure3}, see also Fig.~\ref{fig:figureSI5}).
+For the leaf area ratio (LAR), PFTs were generally closer together, with a notable exception for small evergreen gymnosperms (< 1m height) (Fig.~\ref{fig:figure4}).
 
-\item Results for variance partitioning of LMF and LAR gave comparable results (Table~\ref{varpartpipe}): PFT increased the variance explained in LMF from 60 to 64\%.
-However, SLM was not a good substitute for PFT, as the explained variance decreased to 73\%.
-The leaf area ratio, similarly to ALF/Astba, did not depend strongly on PFT or SLA.
+\item Results for variance partitioning of LMF and LAR gave comparable results (Table~\ref{varpartpipe}): PFT increased the variance explained in LMF from 64 to 73\%.
+However, SLM was not a good substitute for PFT, as the explained variance decreased to 69\%.
+The leaf area ratio, similarly to $A_F/A_S$, did not depend strongly on PFT or SLM.
 
 \item Although biome was a significant predictor in all variables tested, it consistently explained very little variation (Table~\ref{varpartpipe} and~\ref{varpartlmflar}).
-Pair-wise comparisons between biomes within PFT never showed significant differences for MF/Asba or AL/Asba (for example boreal vs.
-temperate evergreen gymnosperm) (Fig.~\ref{fig:figureSI2}).
+Pair-wise comparisons between biomes within PFT mostly did not show significant differences for $M_F/A_S$ or $A_F/A_S$.
+The only exception was a higher $A_F/A_S$ for tropical compared to temperate evergreen angiosperms (Fig.~\ref{fig:figureSI2}).
 
@@
 % latex table generated in R 3.1.1 by xtable 1.7-4 package
-% Tue Oct 14 13:17:40 2014
+% Tue Oct 14 14:02:13 2014
 \begin{table}[ht]
@@
 % latex table generated in R 3.1.1 by xtable 1.7-4 package
-% Tue Oct 14 13:17:41 2014
+% Tue Oct 14 14:02:13 2014
 \begin{table}[ht]

### 2014-10-14 R/functions-figures.R
Added least-square means of LMF and LAR. Now all analyses use a version of the dataset where leaf mass and area are included (to avoid using different datasets).

@@ -297,7 +297,7 @@ gamplotandpred <- function(dat, pftvar, yvar,
   df2 <- melt(df2, id="H", value.name=paste0(yvar,"_SE"), variable.name=pftvar)  
   df <- merge(df1, df2)
   par(o)
-  return(invisible(df))
+  return(invisible(list(df=df, fits=fits)))
 }
 
 plotg <- function(g, which, ...){

### 2014-10-14 manuscript/draftpaper.Rnw
Added least-square means of LMF and LAR. Now all analyses use a version of the dataset where leaf mass and area are included (to avoid using different datasets).

@@
 \item We used linear mixed-effects models (with species within study as the random effect) to test for significance and to calculate confidence intervals for PFT-specific averages.
+Least-square means for LMF and LAR were calculated with the lsmeans package.
 
@@
 For the leaf area ratio (LAR), PFTs were generally closer together, with a notable exception for small evergreen gymnosperms (< 1m height) (Fig.~\ref{fig:figure4}).
+Calculated least-square means of LMF and LAR (averages accounting for variation explained by plant height) showed very similar patterns: large differences in mass-based allocation between PFTs (LMF), and small differences in area-based allocation (LAR) (Fig.~\ref{fig:figure5}).
 
-\item Results for variance partitioning of LMF and LAR gave comparable results (Table~\ref{varpartpipe}): PFT increased the variance explained in LMF from \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[2],2)} to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[3],2)}\%.
+\item We also partionined the variance of LMF and LAR (Table~\ref{varpartpipe}).
+When added to a model that included plant height and biome, PFT increased the variance explained in log-transformed LMF from \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[2],2)} to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[3],2)}\%.
 However, SLM was not a good substitute for PFT, as the explained variance decreased to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[4],2)}\%.
-The leaf area ratio, similarly to $A_F/A_S$, did not depend strongly on PFT or SLM.
+For (log-transformed) LAR, PFT and SLM explained very little additional variance.
 
@@
 
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{figure5_lsmeans_LMFLAR.pdf}
+    \caption{Average leaf mass fraction (LMF) and leaf area ratio (LAR) accounting for variation explained by plant height (least-square means) with a mixed effects model (with species within study as a random effect).
+Error bars are 95\% confidence intervals.}
+    \label{fig:figure5}
+\end{figure}
+
 \clearpage

### 2014-10-14 R/baadvarpart.R

@@ -77,6 +77,8 @@ r2_lmer_LMF_1 <- r.squared(lmer_LMF_1)
 r2_lmer_LMF_2 <- r.squared(lmer_LMF_2)
 r2_lmer_LMF_3 <- r.squared(lmer_LMF_3)
 
+
+
 # LAR
 lmer_LAR_0 <- lmer(log10(a.lf/m.so) ~ log10(h.t) + I(log10(h.t)^2) + (1|Group), data=dat_alfmso)
 lmer_LAR_1 <- lmer(log10(a.lf/m.so) ~ log10(h.t)*bortemptrop + I(log10(h.t)^2) + (1|Group), data=dat_alfmso)

### 2014-10-14 R/allocation.Rnw
Problem description for Lu

\documentclass{article}

\usepackage{parskip}

\begin{document}

Dynamics of leaf mass ($M_F$) and woody biomass ($M_S$) depend on annual allocation of NPP to these pools ($\phi_F$ and $phi_S$), and turnover rates ($yr^{-1}$) ($r_F$ and $r_S$), as in:

\begin{equation}
\frac{dM_F}{dt} = \phi_F NPP - r_F M_F
\end{equation}

\begin{equation}
\frac{dM_S}{dt} = \phi_S NPP - r_S M_S
\end{equation}
With the constraint that $\phi_F + \phi_S = 1$.

We can further assume that NPP is proportional to foliage biomass,

\begin{equation}
NPP = a_F M_F
\end{equation}
So that Eq. 1 and 2 become,

\begin{equation}
\frac{dM_F}{dt} = (\phi_F  a_F - r_F) M_F
\end{equation}

\begin{equation}
\frac{dM_S}{dt} = \phi_S  a_F  M_F - r_S M_S
\end{equation}

I don't know how to deal with this system of DE's, what I would like to have is a solution that gives me the dependence of $M_F$ on $M_S$ (and time of course),

$$M_F = f(M_S,t)$$

Or if it is more convenient,

$$M_F/M_S = f(M_S,t)$$

where either function of course also depends on the various parameters. If it helps the solution, you may also assume that $r_F >> r_S$

\end{document}

### 2014-10-14 manuscript/draftpaper.Rnw
refs, and lit notes added back

@@
 %------------------------------------------------------------------------------------%
-% \section{Lit notes}
-% 
-% \cite{friend_carbon_2014} showed that 30\% of the uncertainty in global vegetation carbon simulated with seven vegetation models is due to the residence time of carbon in vegetation.
+\section{Lit notes}
+
+Reich et al 1992, for seedlings, negative relationship leaf lifespan and LAR.
+This is different from our results, since SLM is nearly proportional to leaf lifespan, and we found no relationship between LAR and SLM (but do they have all the PFTs??).
+However, across stands, positive relationship between leaf biomass and leaf lifespan, which is consistent with our results.
+
+Cornelissen et al 1996 positive relationship SLA and LAR, but this is for tiny seedlings harvested 21 days after first leaf appearance. But check references and who cites this!
+
+\cite{friend_carbon_2014} showed that 30\% of the uncertainty in global vegetation carbon simulated with seven vegetation models is due to the residence time of carbon in vegetation.
 This residence time depends on turnover rates of the various plant pools, and hence also on the distribution of biomass between these pools.
 For example, vegetation with a higher leaf mass fraction will have a lower overall residence time of carbon.
-% 
-% \cite{de2014does} "We have shown that allocation approaches that are constrained by biomass fractions (such as functional relationships) were more successful at capturing observed trends, and were generally more robust, than approaches based on allocation coefficients."
-% 
-% \cite{friedlingstein_toward_1999} make no mention of plant functional type differences or the need to vary allocation by pft.
-% 
-% \cite{givnish_adaptive_2002} discusses costs and benefits of evergreen vs.
+
+\cite{de2014does} "We have shown that allocation approaches that are constrained by biomass fractions (such as functional relationships) were more successful at capturing observed trends, and were generally more robust, than approaches based on allocation coefficients."
+
+\cite{friedlingstein_toward_1999} make no mention of plant functional type differences or the need to vary allocation by pft.
+
+\cite{givnish_adaptive_2002} discusses costs and benefits of evergreen vs.
 deciduous habit, in terms of constructing a growth model that takes into account photosynthesis, leaf lifespan, belowground costs.
@@
 Instead he simply states that 'leaf allocation ratio' relates to leaf lifespan, which of course is only true if total leaf mass is constant.
-% 
-% \cite{callaway_biomass_1994} example of lower leaf mass / sapwood area for desert population of PIPO vs.
+
+\cite{callaway_biomass_1994} example of lower leaf mass / sapwood area for desert population of PIPO vs.
 montane population.
 For pointing out small-scale climatic effects on allometry (that may not be obvious in global scale analyses).
-% 
-% \cite{poorter_biomass_2012} found a slightly larger LMF in gymnosperms, but the difference was small and could not split between deciduous and evergreen angiosperms due to a lack of data.
+
+\cite{poorter_biomass_2012} found a slightly larger LMF in gymnosperms, but the difference was small and could not split between deciduous and evergreen angiosperms due to a lack of data.
 They argue that "most conifers retain leaves for two or more years.
@@
 evergreen).
-% 
-% \cite{enquist_global_2002} report 2.6 times higher leaf mass at given stem mass for conifers compared to angiosperms, but did not separate further, nor interpreted this difference in terms of SLA.
+
+\cite{enquist_global_2002} report 2.6 times higher leaf mass at given stem mass for conifers compared to angiosperms, but did not separate further, nor interpreted this difference in terms of SLA.
 Their interpretation is 'that conifers typically retain three cohorts of leaves that have less well-developed [...] mesophyll'.
 Not sure how that latter bit makes sense, and leaf lifespan interpretation is no good; because it still begs the question why would conifers not simply have 1/3 less leaf mass per cohort?
-% 
-% \cite{mccarthy_consistency_2007} further demonstrate large difference in LMF between angiosperms and conifers, and note that a large chunk of the residuals is correlated with SLA and LL.
+
+\cite{mccarthy_consistency_2007} further demonstrate large difference in LMF between angiosperms and conifers, and note that a large chunk of the residuals is correlated with SLA and LL.
 They interpret this rather vaguely as 'leaf morphology may constrain leaf mass distribution'.
 Their SLA was not on the same individuals, but they could have calculated PFT average LAR anyway, but did not.
-% 
-% \cite{oneill_comparative_1981} reported much higher leaf biomass for gymnosperms vs.
+
+\cite{oneill_comparative_1981} reported much higher leaf biomass for gymnosperms vs.
 angiosperms, at a given woody biomass increment.
 This means (as Niklas reported) a much higher productivity per unit leaf biomass for angiosperms.
-% 
-% Serious lack of biomass data to constrain models.
+
+Serious lack of biomass data to constrain models.
 Studies like \cite{wolf_forest_2011} resort to studying individual tree allometry based on dividing stand biomass by stand density.
 \cite{wolf_allometric_2010} did not find any differences in biomass fractions (root,stem,leaf) between angiosperms and gymnosperms (but again this was based on stand biomass data, but still this is strange!).
-% 
-% \cite{hui_near_2014} Isometric scaling of above vs. belowground biomass in chinese forest biomass database.
-% 
-% \cite{sardans_tree_2013} report similiar LMF for evergreen angio. and gymnosperms, and ca. 40\% less leaf mass for decid. angio. Local study, 3000 plots in Catalonia.
-% 
-% \cite{reich_temperature_2014} "On average, when standardized for MAT and Mstem, conifers tended to have more Mfol and less Mroot than angiosperm forests".
+
+\cite{hui_near_2014} Isometric scaling of above vs. belowground biomass in chinese forest biomass database.
+
+\cite{sardans_tree_2013} report similiar LMF for evergreen angio. and gymnosperms, and ca. 40\% less leaf mass for decid. angio. Local study, 3000 plots in Catalonia.
+
+\cite{reich_temperature_2014} "On average, when standardized for MAT and Mstem, conifers tended to have more Mfol and less Mroot than angiosperm forests".
 Their Fig.
@@
 But what about conifers?
-% 
-% \cite{cairns_root_1997} found no difference in root-shoot ratio between gymno and angio.
-% 
-% \cite{delucia_climate-driven_2000} LA/SA decreases with summer VPD for Pinus but not non-Pinus gymno.
+
+\cite{cairns_root_1997} found no difference in root-shoot ratio between gymno and angio.
+
+\cite{delucia_climate-driven_2000} LA/SA decreases with summer VPD for Pinus but not non-Pinus gymno.
 Also Pinus has much lower LA/SA than non-Pinus; suggesting that we should separate gymnosperms into those groups as well.
-% 
-% 
-% \cite{grier_notes:_1974} early study on leaf mass proportionality to sapwood area (and uses mass not area).
+
+\cite{grier_notes:_1974} early study on leaf mass proportionality to sapwood area (and uses mass not area).
 

### 2014-10-14 manuscript/draftpaper.Rnw
Makefile fixed; analysis moved to statanalysis (and added to makefile)

@@
 \begin{enumerate}
-\item Importance of biomass allocation in global vegetation models \cite{ise_comparison_2010}, feedback on fluxes of CO2, lifetime of C in the system \cite{friend_carbon_2014}.
+\item The importance of biomass allocation in global vegetation models \cite{ise_comparison_2010}, feedback on fluxes of CO2, lifetime of C in the system \cite{friend_carbon_2014}, which depends on partitioning of biomass into short-lived foliage vs.
+long-lived woody pools.
 

### 2014-10-14 dependencies.R
Makefile fixed; analysis moved to statanalysis (and added to makefile)

@@ -3,7 +3,7 @@
 pkgs <- c("lmerTest","maps","mapdata","maptools","gdata","knitr","smatr","scales",
           "xtable","magicaxis","rmarkdown","dplyr","sciplot","epade",
           "multcomp","doBy","mgcv","reshape2","lme4","Hmisc","hexbin",
-          "lattice","grid","stringr","raster","dismo","XML","rgdal","lsmeans")
+          "lattice","grid","stringr","raster","dismo","XML","rgdal","lsmeans","gplots")
 
 pkgs.missing <- setdiff(pkgs, rownames(installed.packages()))
 install.packages(pkgs.missing)

### 2014-10-14 figures.R

@@ -22,11 +22,6 @@ to.pdf({
 }, filename="manuscript/figures/figure1_mlf_astba2_bypft.pdf", width=6, height=5)
 
 
-smoothplotbypft(log10(h.t), log10(m.st/a.stba2), dataset2, xlab=expression(Basal~stem~area~~(m^2)),
-                ylab=expression(Aboveground~woody~mass~(kg)), cex=0.6,
-                pointcols=transCols,linecols=linecols)
-
-
 to.pdf({
   par(mar=c(5,5,2,2), cex.lab=1.2)
   smoothplotbypft(log10(a.stba2), log10(a.lf), dataset2, xlab=expression(Basal~stem~area~~(m^2)),
@@ -262,6 +257,74 @@ to.pdf({
   abline(0,1)
   Legend("topleft")
 }, filename="manuscript/figures/figureSI-6_mrt_mso_bypft.pdf", width=6, height=5)
+
+
+
+
+# Worldclim
+climspace <- read.csv("data/Worldclim_landcover_climspace.csv")
+map <- climspace$MAP_WC
+mat <- climspace$MAT_WC/10
+mapmat <- baad[!duplicated(baad[,c("MAP","MAT")]),]
+mapmat$vegetation <- as.factor(mapmat$vegetation)
+
+pftpoints <- function(p,i){
+  panel.points(mapmat$MAT[mapmat$pft == p],
+               mapmat$MAP[mapmat$pft == p],
+               cex=1.3, pch=19, 
+               col=Cols[i])
+}
+
+# Hexbin plot with Worldclim and BAAD (figure 1)
+to.pdf({
+  Cols <- alpha(c("blue","cyan1","red","forestgreen"),0.6)
+  h <- hexbinplot(map ~ mat, aspect = 1, bins=50, 
+                  xlab = expression("Mean annual temperature"~(degree*C)), 
+                  ylab = "Mean annual precipitation (mm)", 
+                  colorkey=FALSE,
+                  par.settings = list(par.xlab.text=list(cex=1.5),
+                                      par.ylab.text=list(cex=1.5)),
+                  panel = function(...) {
+                    panel.hexbinplot(...)
+                    pftpoints("DA",1)
+                    pftpoints("EA",3)
+                    pftpoints("EG",4)
+                    pftpoints("DG",2)
+                    panel.points(rep(-25,4),seq(6000,8000,length=4),pch=19,cex=1.3,col=alpha(Cols,0.9))
+                    panel.text(rep(-22,4), seq(6000,8000,length=4), 
[…]

### 2014-10-15 manuscript/draftpaper.Rnw
Changed font to more readable opensans
Removed two counts tables as they were identical

@@
 \usepackage{graphicx}
+\usepackage[default,osfigures,scale=0.95]{opensans}
 \graphicspath{ {./figures/} }
@@
 
-\item We used a generalized additive model (GAM) to visualize the relationships.
+\item We used a generalized additive model (GAM) to visualize the relationships, but these were not used for formal inference.
 
 \item We used linear mixed-effects models (with species within study as the random effect) to test for significance and to calculate confidence intervals for PFT-specific averages.
-Least-square means for LMF and LAR were calculated with the lsmeans package.
+Least-square means for LMF and LAR were calculated with the 'lsmeans' package.
 
-\item Variance partitioning with marginal (fixed effects) and conditional (fixed and random effects) R2 for linear mixed-effects models as reported by \cite{nakagawa_general_2013}.
+\item All variables were log-transformed prior to analysis, to stabilize variance.
+
+\item Variance partitioning with marginal (fixed effects) and conditional (fixed and random effects) $R^2$ for linear mixed-effects models as reported by \cite{nakagawa_general_2013}.
 
@@
 <<results='asis'>>=
-printit(tab_mlfastba, caption="Number of observations used in the analysis of Mf/As. Individual plants (number of species).")
+printit(tab_mlfastba, caption="Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$.
+Individual plants (number of species).
+DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
 @
 
+% <<results='asis'>>=
+% printit(tab_alfastba, caption="Number of observations used in the analysis of Af/As.
+Individual plants (number of species).
+DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
+% @
+%
+
 <<results='asis'>>=
-printit(tab_alfastba, caption="Number of observations used in the analysis of Af/As. Individual plants (number of species).")
+printit(tab_lmf, caption="Number of observations used in the analysis of leaf mass fraction (LMF) and leaf area ratio (LAR).
+Individual plants (number of species).
+DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
 @
 
-<<results='asis'>>=
-printit(tab_lmf, caption="Number of observations used in the analysis of leaf mass fraction (LMF). Individual plants (number of species).")
-@
-
-<<results='asis'>>=
-printit(tab_lar, caption="Number of observations used in the analysis of leaf area ratio (LAR). Individual plants (number of species).")
-@
+% 
+% <<results='asis'>>=
+% printit(tab_lar, caption="Number of observations used in the analysis of leaf area ratio (LAR).
+Individual plants (number of species).
+DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
+% @
 

### 2014-10-15 manuscript/draftpaper.Rnw
intro + methods

@@
 First, SLM correlates with leaf lifespan.
-Longer lived leaves would lead to higher plant leaf mass (MF) as long as annual production rates are similar (\cite{enquist_global_2002}).
+Longer lived leaves would lead to higher plant leaf mass ($M_F$) as long as annual production rates are similar (\cite{enquist_global_2002}).
 Another interpretation is that plants may target a certain leaf area display given a certain biomass investment in supporting tissues, because light interception and photosynthesis correlate with the exchange area of the plant.
 
-\item It has been reported that gymnosperms (with higher SLM) have higher leaf mass fraction than angiosperms (with lower SLM) \cite{enquist_global_2002, mccarthy_consistency_2007, poorter_biomass_2012, sardans_tree_2013}.
-But a recent study reported no difference (\cite{wolf_allometric_2010}).
-No study to date has interpreted differences in partitioning between PFTs in terms of SLM.
-This would be useful because SLM is already an important parameter in DGVMs, potentially reducing parameters required for the allocation subroutine.
+\item It has been reported numerous times that gymnosperms (with higher SLM) have higher leaf mass fraction than angiosperms (with lower SLM) \cite{enquist_global_2002, mccarthy_consistency_2007, poorter_biomass_2012, sardans_tree_2013}, although a recent study reported no difference (\cite{wolf_allometric_2010}).
+No study to date has been able to analyze differences between deciduous and evergreen angiosperms, or link differences in partitioning between PFTs to other plant traits such as SLM.
 
@@
 Most global efforts to date use the Cannell database of forest biomass, which is not based on individual biomass estimates.
-Here we use a new database (Falster et al.), the largest ever compiled for individual woody plants, to evaluate whether important differences exist in biomass partitioning woody plant functional types.
-We also test whether these differences are confounded (or compounded) by climate.
+Here we use a new database (Falster et al.), the largest ever compiled for individual woody plants, including data for major plant functional types (PFTs), and all major biomes.
+
+\item We consider here the relationship between leaf biomass and area with stem diameter (an important allometric relationship, Enquist others, and related to the pipe model if the ratio sapwood / heartwood is relatively constant, an important measure of hydraulic supply), the aboveground leaf mass fraction (partioning of biomass into leaf vs.
+woody pools), and the leaf area ratio (a measure of the photosynthetic area per unit woody biomass).
+We expect that $M_F$ and $A_F$ are approximately proportional to the square of stem diameter, which would remove size-related shifts in allometry that often confound interpretations.
+For each of these variables, we ask whether 1) there are differences between major plant functional types after removing size-related variation, 2) there are differences due to climate, compared between major biomes, or as a function of the mean annual temperature (as recently reported for whole stand biomass, \cite{reich_temperature_2014}).
 
@@
 We further excluded deciduous gymnosperms (since very few data were available), and glasshouse and common garden studies, including only field grown woody plants (including plantations).
+
+\item We focus here only on aboveground biomass because we found no PFT differences in root-shoot scaling (Fig.~\ref{fig:figureSI6}), and approximately isometric scaling (sma slope = xxx) (consistent with other reports, \cite{hui_near_2014, cairns_root_1997}).
+
+\item For the aboveground biomass pools, we calculated the leaf mass fraction ($M_F / M_W$) and leaf area ratio ($A_F / M_W$). LMF, LAR, SLM, $M_F$ and $A_F$ are related by,
+
+$$LMF = SLM * LAR = (M_F / A_F) * LAR$$
 
@@
 \begin{itemize}
-
-\item We focus here only on aboveground biomass because we found no PFT differences in root-shoot scaling (Fig.~\ref{fig:figureSI6}), and approximately isometric scaling (sma slope = xxx) (consistent with other reports, \cite{hui_near_2014, cairns_root_1997}).
 
@@
 Optimization models probably already give the right answer because normally LAI is optimized (Woodward).
+
+\item A recent study compared allocation routines in a number of leading ecosystem models \cite{de2014does}, and recommended to constrain allocation by observed biomass fractions instead of using constant allocation fractions, which is common practice.
 

### 2014-10-17 manuscript/draftpaper.Rnw
Major figures update. Also added species-level plot. Manuscript results section improved.

@@
 \maketitle
+
+% Notes from Daniel phone call
+% - make case that allocating biomass vs. targeting certain leaf area vs. stem area is very different
+% 
+% fig 1 : more grey levels. scale for grey levels
+% figure captions need to be stronger statements
+% 
+% use all data with leaf mass, explain that subset has same results
+% 
+% Fig 9 + 10 instead of Fig 2 in main text.
+% Then, Fig 3 says that SLM explains differences.
+% 
+% Fig 3 + 6 log-log axes
+% 
+% Fig 4 first, introduce PFT differences
+% 
+% No difference in m.so/a.astba2, but large difference m.lf/a.stba2
+% 
+% remove fig 5
+% 
+% combine fig 3
+% 
+% climate-related effects are captured by shift in PFT
+%
 
@@
 
-\item For the aboveground biomass pools, we calculated the leaf mass fraction ($M_F / M_W$) and leaf area ratio ($A_F / M_W$). LMF, LAR, SLM, $M_F$ and $A_F$ are related by,
+\item For the aboveground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total aboveground biomass) and leaf area ratio ($A_F / M_T$).
+LMF, LAR, SLM, $M_F$ and $A_F$ are related by,
 
-$$LMF = SLM * LAR = (M_F / A_F) * LAR$$
+$$LMF = M_F / M_T = (M_F / A_F) (A_F / M_T) = SLM * LAR$$
+
+\item We can further decompose aboveground biomass ($M_T$) into a size-related metric (basal stem area, $A_S$), and the ratio of $M_T$ to $A_S$, which is a function of wood density and plant woody volume to basal area.
+
+$$LMF = M_F / M_T = (M_F / A_S) (M_T / A_S)^{-1}$$
+
+We test for PFT differences to both these components of biomass partioning.
+We were unable to further split $A_S/M_W$ because wood density data were available only for a subset of data.
+Chave has shown that etc.
+etc.
 
@@
 
-\item We found strong correlations between basal stem area ($A_S$) and leaf mass ($M_F$) across the entire dataset, with a clear separation between the three major plant functional types (Fig.~\ref{fig:figure1}).
+\item We found large differences between PFTs in biomass partitioning between leaves and stems (Fig.~\ref{fig:figure2}, see also Fig.~\ref{fig:figureSI3}).
+Calculated least-square means of LMF and LAR (averages accounting for variation explained by plant height) showed large differences in mass-based allocation between PFTs (LMF), and small inconsistent differences in area-based allocation (LAR) (Fig.~\ref{fig:figure2}, see Fig.~\ref{fig:figureSI4} for raw data on leaf area ratio by PFT).
 
-\item Because of weak plant height effects, we calculated average $M_F/A_S$ across the entire dataset by the three main PFTs.
-We found a ca.
-3-fold difference between evergreen gymnosperms and deciduous angiosperms, with evergreen angiosperms taking an intermediate value (Fig.~\ref{fig:figure2}).
-These differences were approximately proportional to the average specific leaf mass (SLM = $M_F / A_F$)for the PFT.
+\item The ratio of aboveground biomass to basal stem area did not differ by PFT (Fig.~\ref{fig:figureSI5}).
 
-\item However, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure2}), confirming that $M_F/A_S$ is approximately proportional to SLM across PFTs (See also Fig.~\ref{fig:figureSI1}).
+\item However we found very large differences between PFTs in the ratio of leaf mass to basal stem area (Fig.~\ref{fig:figure3}), but the ratio leaf area to basal stem area converged between PFTs.
 
-\item We do note that there was substantial variation between individual plants (Figs.~\ref{fig:figureSI3} and~\ref{fig:figureSI3}).
-We estimated the variance explained by PFT, plant height, and biome in $M_F/A_S$ using a linear mixed-effects model (see Table~\ref{varpartpipe}).
+\item We then expressed the average $M_F/A_s$ as a function of the specific leaf mass (SLM = $M_F / A_F$), a measure of the leaf construction cost to the plant.
+We found that, across PFTs, $M_F/A_S$ was approximately proportional to the average specific leaf mass for the PFT.
+As a result of this approximate proportionality, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure4}) (see also Fig.~\ref{fig:figureSI6} for raw data).
+
+\item This relationship also held approximately across species (Fig.~\ref{fig:figureSI7}).
+
+\item Because there was substantial variation between individual plants (Figs.~\ref{fig:figure3}), we estimated the variance explained by PFT, plant height, and biome in $M_F/A_S$ using a linear mixed-effects model (see Table~\ref{varpartpipe}).
 We found that plant height and biome explained only \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[2],2)}\%, but when PFT was added to the model, the variance explained increased to \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[3],2)}\%.
@@
 Consistent with this, PFT, plant height and biome explained very little variance in $A_F/A_S$ (Table~\ref{varpartpipe}).
-
-\item We found similar results for the above-ground leaf mass fraction (LMF), where a clear and strong separation was found between plant functional types (Fig.~\ref{fig:figure3}, see also Fig.~\ref{fig:figureSI5}).
-For the leaf area ratio (LAR), PFTs were generally closer together, with a notable exception for small evergreen gymnosperms (< 1m height) (Fig.~\ref{fig:figure4}).
-Calculated least-square means of LMF and LAR (averages accounting for variation explained by plant height) showed very similar patterns: large differences in mass-based allocation between PFTs (LMF), and small differences in area-based allocation (LAR) (Fig.~\ref{fig:figure5}).
 
@@
 
-\item Although biome was a significant predictor in all variables tested, it consistently explained very little variation (Table~\ref{varpartpipe} and~\ref{varpartlmflar}).
+\item Although biome was a significant predictor for all variables tested, it consistently explained very little variation (Table~\ref{varpartpipe} and~\ref{varpartlmflar}).
 Pair-wise comparisons between biomes within PFT mostly did not show significant differences for $M_F/A_S$ or $A_F/A_S$.
-The only exception was a higher $A_F/A_S$ for tropical compared to temperate evergreen angiosperms (Fig.~\ref{fig:figureSI2}).
+The only exception was a higher $A_F/A_S$ for tropical compared to temperate evergreen angiosperms (Fig.~\ref{fig:figureSI8}).
 
-\item Unlike \cite{reich_temperature_2014}, we found no effect of mean annual temperature on the scaling of leaf biomass with woody biomass (Fig.~\ref{fig:figureSI7}).
+\item Unlike \cite{reich_temperature_2014}, we found no effect of mean annual temperature on the scaling of leaf biomass with woody biomass (Fig.~\ref{fig:figureSI9}).
 
@@
 }
-    \label{fig:figureworldclim}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics{figure1_mlf_astba2_bypft.pdf}
-    \caption{Leaf mass per unit basal stem area. Each symbol is an individual plant. Lines are generalized additive model fits.}
     \label{fig:figure1}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{figure2_mlf_alf_astbaest_pftmeans.pdf}
-    \caption{Average leaf mass per unit basal stem area (panel 1), and leaf area per unit basal stem area (panel 2) for the three major plant functional types.
-Error bars are 95\% confidence interval for the mean (calculated with a mixed-effects model with species within study as the random effect).
-Letters denote significant differences.
-}
+    \includegraphics[width=0.99\textwidth]{Figure2_LMF_lines_lsmeans_3panel.pdf}
+    \caption{Leaf mass fraction (leaf mass / aboveground biomass) by plant functional type.
+Each symbol is an individual plant.
+Lines are generalized additive model fits.Bottom panels show average leaf mass fraction (LMF) and leaf area ratio (LAR) accounting for variation explained by plant height (least-square means) with a mixed effects model (with species within study as a random effect).
+Error bars are 95\% confidence intervals.}
     \label{fig:figure2}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{figure3_LMF_pft_lines.pdf}
-    \caption{Leaf mass fraction (leaf mass / aboveground biomass) by plant functional type. Each symbol is an individual plant. Lines are generalized additive model fits.}
+    \includegraphics[width=0.99\textwidth]{Figure3_hist_alfast_mlfast.pdf}
+    \caption{Plant functional types diverge strongly in mass-based allocation, but converge in area-based allocation.
+Shown are histograms of leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types.
+Vertical lines and shading are means with confidence intervals (same as in Fig.~\ref{fig:figure2}).
+Histograms of $M_F/A_S$ show large differences between PFTs, while $A_F/A_S$ converges to a single average across all PFTs, though much variation exists between plants.}
     \label{fig:figure3}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{figure4_LAR_pft_lines.pdf}
-    \caption{Leaf area ratio (leaf area / aboveground biomass) by plant functional type. Each symbol is an individual plant. Lines are generalized additive model fits.}
+    \includegraphics[width=0.99\textwidth]{Figure4_mlf_alf_astbaest_pftmeans.pdf}
+    \caption{Average leaf mass per unit basal stem area (panel 1), and leaf area per unit basal stem area (panel 2) for the three major plant functional types.
+Error bars are 95\% confidence interval for the mean (calculated with a mixed-effects model with species within study as the random effect).
+Letters denote significant differences.
+}
     \label{fig:figure4}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{figure5_lsmeans_LMFLAR.pdf}
-    \caption{Average leaf mass fraction (LMF) and leaf area ratio (LAR) accounting for variation explained by plant height (least-square means) with a mixed effects model (with species within study as a random effect).
-Error bars are 95\% confidence intervals.}
-    \label{fig:figure5}
 \end{figure}
@@
 
+\renewcommand\thefigure{S\arabic{figure}}    
+\setcounter{figure}{0}
+
 \begin{figure}[h!]
     \centering
-    \includegraphics[width=0.99\textwidth]{figureSI-8_MAPMAT_vegetation.pdf}
+    \includegraphics[width=0.99\textwidth]{FigureSI-1_MAPMAT_vegetation.pdf}
     \caption{Mean annual precipitation and temperature for all contributed datasets, by vegetation type. Note that in this study, we did not include glasshouse studies.}
+    \label{fig:figureSI1}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureSI-2_mrt_mso_bypft.pdf}
+    \caption{Near-isometric scaling of above-ground with below-ground biomass, and no effects of plant functional type. The black line is a 1-1 line.}
+    \label{fig:figureSI2}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureSI-3_mlfmst_byht_pft.pdf}
+    \caption{Leaf and woody biomass scaling with plant height.}
+    \label{fig:figureSI3}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureSI-4_LAR_pft_lines.pdf}
+    \caption{Leaf area ratio (leaf area / aboveground biomass) by plant functional type. Each symbol is an individual plant. Lines are generalized additive model fits.}
+    \label{fig:figureSI4}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureSI-5_mso_astba2_twopanel.pdf}
+    \caption{Relationship between aboveground biomass and basal stem area, each point is an individual plant.
+Lines are spline fits.
+The right panel shows least-square means of the ratio of aboveground biomass to basal stem area, estimated from a linear mixed-effects model with basal stem area (linear and squared terms) and PFT.
+}
+    \label{fig:figureSI5}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureSI-6_mlf_alf_astba2_bypft.pdf}
+    \caption{Plant leaf mass as a function of basal stem area (left panel), and leaf area as a function of basal stem area (right panel).
+Each symbol is an individual plant.
+Lines are generalized additive model fits.}
+  \label{fig:figureSI6}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureSI-7_mlf_alf_ast_byspecies.pdf}
+    \caption{Plant leaf mass as a function of basal stem area (left panel), and leaf area as a function of basal stem area (right panel).
+Individual plant sata were averaged by species/study (the same species occuring in multiple datasets thus appears multiple times).
+Lines are linear regressions with 95\% confidence intervals.
+Both regressions were significant, but R2 was 0.28 in the left panel, and only 0.02 in the right panel.
+The coefficient of variation between species was 0.19 for log-transformed MF/AS, and only 0.099 for AF/AS.}
+  \label{fig:figureSI7}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureSI-8_mlf_alf_astbaest_pftlongmeans.pdf}
+    \caption{Like Fig.~\ref{fig:figure2}, but for five combinations of major plant functional type and biome.
+Out of 9 possible biome / pft combinations, only five in the dataset had leaf area estimates (and two combo's are not in the dataset, e.g.
+tropical evergreen gymno).
+Note that within pft, vegetation is never significant (this result will also be in another figure!).}
     \label{fig:figureSI8}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{figureSI-6_mrt_mso_bypft.pdf}
-    \caption{Near-isometric scaling of above-ground with below-ground biomass, and no effects of plant functional type. The black line is a 1-1 line.}
-    \label{fig:figureSI6}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.5\textwidth]{figureSI-3_lmlf_astba2_hist_bypft.pdf}
-    \caption{Histograms of leaf mass per unit basal stem area grouped by the three major plant functional types.
-Vertical lines and shading are means with confidence intervals (same as in Fig.~\ref{fig:figure2})}
-    \label{fig:figureSI3}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.5\textwidth]{figureSI-4_lalf_astba2_hist_bypft.pdf}
-    \caption{Histograms of leaf area per unit basal stem area grouped by the three major plant functional types.
-Vertical lines and shading are means with confidence intervals (same as in Fig.~\ref{fig:figure2})}
-    \label{fig:figureSI4}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{figureSI-1_alf_astba2_bypft.pdf}
-    \caption{Leaf area per unit basal stem area. Each symbol is an individual plant. Lines are generalized additive model fits.}
-    \label{fig:figureSI1}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{figureSI-2_mlf_alf_astbaest_pftlongmeans.pdf}
-    \caption{Like Fig.~\ref{fig:figure2}, but for five combinations of major plant functional type and biome.
-Out of 9 possible biome / pft combinations, only five in the dataset had leaf area estimates (and two combo's are not in the dataset, e.g.
-tropical evergreen gymno).
-Note that within pft, vegetation is never significant (this result will also be in another figure!).}
-    \label{fig:figureSI2}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{figureSI-5_mlfmst_byht_pft.pdf}
-    \caption{Leaf and woody biomass scaling with plant height.}
-    \label{fig:figureSI5}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{figureSI-7_MAT_LMFscaling.pdf}
+    \includegraphics[width=0.99\textwidth]{FigureSI-9_MAT_LMFscaling.pdf}
     \caption{For each species within study, we fit the scaling relationship of leaf mass as a function of woody biomass assuming an exponent of 3/4, using standardized major axis.
@@
 (2007), a study on Pinus eliottii with exceptionally high leaf mass per unit woody biomass.}
-    \label{fig:figureSI7}
+  \label{fig:figureSI9}
 \end{figure}

### 2014-10-17 R/functions-figures.R

@@ -15,12 +15,13 @@ mixmean <- function(yvar, g, dataset=dat){
 
 
 
-meansbypft <- function(yvar1, yvar2, pftvar, 
+meansbypft <- function(yvar1, yvar2=NULL, pftvar, 
                            xvar="lsla",
                            panel1.expr=NULL,
                            panel2.expr=NULL,
                            setpar=TRUE,
                            Cols=NULL,
+                           panel1only=FALSE,
                            legend.text=NULL,
                            legend.where="topright",
                            legend.cex=0.7,
@@ -42,30 +43,32 @@ meansbypft <- function(yvar1, yvar2, pftvar,
   if(siglets == "symbol")library(maptools)
   
   dat$Y1 <- dat[,yvar1]
-  dat$Y2 <- dat[,yvar2]
+  if(!panel1only)dat$Y2 <- dat[,yvar2]
   dat$P <- dat[,pftvar]  
   
   sla <- mixmean(xvar,pftvar,dat)
   mlfastbh <- mixmean(yvar1,pftvar,dat)
-  alfastbh <- mixmean(yvar2,pftvar,dat)
+  if(!panel1only)alfastbh <- mixmean(yvar2,pftvar,dat)
   
   if(is.null(Cols))
     Cols <- rainbow(length(unique(dat$P)))
   
   # Simple trend lines; not used at the moment though.
   lmfit1 <- lm(mlfastbh$y ~ sla$y)
-  lmfit2 <- lm(alfastbh$y ~ sla$y)
+  if(!panel1only)lmfit2 <- lm(alfastbh$y ~ sla$y)
   
   # model fit for multiple comparisons
   f1 <- lmer(Y1 ~ P - 1 + (1|Group), data=dat)
-  f2 <- lmer(Y2 ~ P - 1 + (1|Group), data=dat)
+  if(!panel1only)f2 <- lmer(Y2 ~ P - 1 + (1|Group), data=dat)
   
   # Multiple comparison letters.
   lets1 <- cld(glht(f1, linfct=mcp(P="Tukey")))$mcletters$Letters
-  lets2 <- cld(glht(f2, linfct=mcp(P="Tukey")))$mcletters$Letters
[…]

### 2014-10-17 R/ownhexbinplot.R
working on own hexbinplot. manuscript discussion update

@@ -1,0 +1,8 @@
+
+
+
+h <- hexbin(map ~ mat)
+cells <- hcell2xy(h, check.erosion = T)
+
+library(plotrix)
+hexagon(cells$x, cells$y)

### 2014-10-17 manuscript/draftpaper.Rnw
working on own hexbinplot. manuscript discussion update

@@
 \maketitle
-
-% Notes from Daniel phone call
-% - make case that allocating biomass vs. targeting certain leaf area vs. stem area is very different
-% 
-% fig 1 : more grey levels. scale for grey levels
-% figure captions need to be stronger statements
-% 
-% use all data with leaf mass, explain that subset has same results
-% 
-% Fig 9 + 10 instead of Fig 2 in main text.
-% Then, Fig 3 says that SLM explains differences.
-% 
-% Fig 3 + 6 log-log axes
-% 
-% Fig 4 first, introduce PFT differences
-% 
-% No difference in m.so/a.astba2, but large difference m.lf/a.stba2
-% 
-% remove fig 5
-% 
-% combine fig 3
-% 
-% climate-related effects are captured by shift in PFT
-%
 
@@
 
-\item We found no effects of climate on anything.
-This is different from \cite{reich_temperature_2014} but they studied whole-stand biomass, not individual plant biomass.
+\item We found no effects of climate on anything, when analyzed in addition to PFT.
+This means that climate-related differences are captured by a shift in PFT (see also Fig.~\ref{figure1}, showing PFTs occupy different parts of the climate space).
+Our results contrast with those of \cite{reich_temperature_2014} but they studied whole-stand biomass, not individual plant biomass.
 So perhaps this means the climate effects are on stand density not individual plant effects.

### 2014-10-20 manuscript/draftpaper.Rnw
Prep manuscript for NCC, wrote first para, moved litnotes to manuscript/litnotes.txt

@@
 \usepackage{geometry}
+\usepackage{authblk}
+\usepackage{gensymb}
 \geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
@@
 
-\author{Remko Duursma, Daniel Falster and BAAD contributors}
+\author[1]{Remko Duursma}
+\author[2]{Daniel Falster}
+\author[3]{BAAD contributors}
 
+\affil[1]{Hawkesbury Institute for the Environment}
+\affil[2]{Macquarie University}
+\affil[3]{See Supporting Information}
+
+\renewcommand\Authands{ and }
 \maketitle
@@
 %--------------------------------------------------------------------------------------------%
+%% Main text. Max words = 2000, first para = 200.
 
-\section{Introduction}
+\textbf{The partitioning of biomass in terrestrial vegetation between leaves, stems and roots is a critical component of the global carbon cycle.
+Global vegetation models consistently point to biomass partioning as one of the most uncertain and important drivers for the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, and global predictions of carbon storage \cite{ise_comparison_2010}.
+Little is known about direct and indirect effects of climate on biomass partitioning, largely due to the lack of a global individual-based biomass dataset.
+Here, we analyze the largest database on plant biomass to date.
+We show that partitioning of biomass to leaves vs.
+woody stems is strongly dependent on the plant functional type.
+Evergreen gymnosperms maintain three times more leaf biomass at a given total aboveground biomass than deciduous angiosperms, with evergreen angiosperms taking an intermediate value.
+We found that specific leaf mass, an indicator of leaf construction cost, explains these differences so that the PFTs converged in the ratio  of leaf area to aboveground biomass.
+We found no effects of mean annual temperature or precipitation on biomass partitioning, suggesting that climate-related differences are captured by a shift in PFT.
+These results point to a simple algorithm for biomass partitioning in the world’s forests.}
 
-\begin{enumerate}
-\item The importance of biomass allocation in global vegetation models \cite{ise_comparison_2010}, feedback on fluxes of CO2, lifetime of C in the system \cite{friend_carbon_2014}, which depends on partitioning of biomass into short-lived foliage vs.
-long-lived woody pools.
+\section{flow}
 
-\item There is no shortage of theoretical developments in understanding biomass allocation (\cite{cannell_carbon_1994,franklin_modeling_2012,enquist_land_2012}).
+\begin{itemize}
+
+\item First paragraph
+\item Allocation in models, main uncertainties
+\item Climate effects on partioning : direct (MAT, MAP), indirect (PFT).
+\item Known: gymnosperm vs. angiosperm. Unknown: leaf habit (deciduous vs. evergreen), direct climate, patterns for individual plants
+\item BAAD : Largest database, other large datasets are stands or only small plants
+\item Root scaling no PFT differences, focus on aboveground
+\item LMF, LAR
+\item Found very large LMF differences, but convergent LAR
+\item decompose into MF/AS, MT/AS, find similar differences in MF/AS but none in MT/AS, suggesting convergent AF/AS between PFTs.
+\item No direct climate effects, so shift in PFT captures global patterns in biomass partioning
+\item This has implications for allocation in DGVMs, they can be divided into roughly three categories in terms of biomass allocation (allocation fractions, allometric constraints, LAI optimization).
+Allocation vs.
+partitioning: the role of leaf and stem turnover.
+
+\end{itemize}
+
+\section{textbits}
+
+There is no shortage of theoretical developments in understanding biomass allocation (\cite{cannell_carbon_1994,franklin_modeling_2012,enquist_land_2012}).
 In particular, theoretical work focusing on optimal partitioning theory is promising (e.g.
@@
 
-\item Previous syntheses on plant biomass relationships has predominantly focused on the universal nature of size-biomass relationships in plants.
+Previous syntheses on plant biomass relationships has predominantly focused on the universal nature of size-biomass relationships in plants.
 Little is known about differences between plant functional types (evergreen vs.
@@
 
-\item Major plant functional types differ widely in specific leaf mass (SLM, the ratio of leaf mass to area, an index of leaf thickness) (\cite{poorter_causes_2009}), which can be expected to affect biomass partitioning in several ways.
+Major plant functional types differ widely in specific leaf mass (SLM, the ratio of leaf mass to area, an index of leaf thickness) (\cite{poorter_causes_2009}), which can be expected to affect biomass partitioning in several ways.
 First, SLM correlates with leaf lifespan.
@@
 
-\item It has been reported numerous times that gymnosperms (with higher SLM) have higher leaf mass fraction than angiosperms (with lower SLM) \cite{enquist_global_2002, mccarthy_consistency_2007, poorter_biomass_2012, sardans_tree_2013}, although a recent study reported no difference (\cite{wolf_allometric_2010}).
+It has been reported numerous times that gymnosperms (with higher SLM) have higher leaf mass fraction than angiosperms (with lower SLM) \cite{enquist_global_2002, cheng_relationships_2007, mccarthy_consistency_2007, poorter_biomass_2012, sardans_tree_2013}, although a recent study reported no difference (\cite{wolf_allometric_2010}).
 No study to date has been able to analyze differences between deciduous and evergreen angiosperms, or link differences in partitioning between PFTs to other plant traits such as SLM.
 
-\item What is lacking is a solid empirical base that can be used to constrain any model of allocation, that includes woody plants of all sizes, of all major plant functional types and biomes.
+What is lacking is a solid empirical base that can be used to constrain any model of allocation, that includes woody plants of all sizes, of all major plant functional types and biomes.
 Most global efforts to date use the Cannell database of forest biomass, which is not based on individual biomass estimates.
@@
 
-\item We consider here the relationship between leaf biomass and area with stem diameter (an important allometric relationship, Enquist others, and related to the pipe model if the ratio sapwood / heartwood is relatively constant, an important measure of hydraulic supply), the aboveground leaf mass fraction (partioning of biomass into leaf vs.
+We consider here the relationship between leaf biomass and area with stem diameter (an important allometric relationship, Enquist others, and related to the pipe model if the ratio sapwood / heartwood is relatively constant, an important measure of hydraulic supply), the aboveground leaf mass fraction (partioning of biomass into leaf vs.
 woody pools), and the leaf area ratio (a measure of the photosynthetic area per unit woody biomass).
@@
 
-\end{enumerate}
+ We found large differences between PFTs in biomass partitioning between leaves and stems (Fig.~\ref{fig:figure2}, see also Fig.~\ref{fig:figureSI3}).
+Calculated least-square means of LMF and LAR (averages accounting for variation explained by plant height) showed large differences in mass-based allocation between PFTs (LMF), and small inconsistent differences in area-based allocation (LAR) (Fig.~\ref{fig:figure2}, see Fig.~\ref{fig:figureSI4} for raw data on leaf area ratio by PFT).
+
+ The ratio of aboveground biomass to basal stem area did not differ by PFT (Fig.~\ref{fig:figureSI5}).
+
+ However we found very large differences between PFTs in the ratio of leaf mass to basal stem area (Fig.~\ref{fig:figure3}), but the ratio leaf area to basal stem area converged between PFTs.
+
+ We then expressed the average $M_F/A_s$ as a function of the specific leaf mass (SLM = $M_F / A_F$), a measure of the leaf construction cost to the plant.
+We found that, across PFTs, $M_F/A_S$ was approximately proportional to the average specific leaf mass for the PFT.
+As a result of this approximate proportionality, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure4}) (see also Fig.~\ref{fig:figureSI6} for raw data).
+
+ This relationship also held approximately across species (Fig.~\ref{fig:figureSI7}).
+
+ Because there was substantial variation between individual plants (Figs.~\ref{fig:figure3}), we estimated the variance explained by PFT, plant height, and biome in $M_F/A_S$ using a linear mixed-effects model (see Table~\ref{varpartpipe}).
+We found that plant height and biome explained only \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[2],2)}\%, but when PFT was added to the model, the variance explained increased to \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[3],2)}\%.
+If we used SLM as a predictor instead of PFT, the $R^2$ was similar (\Sexpr{100*signif(Table_pipemodel_varpart$Marginal[4],2)}\%), showing that PFT affected $M_F/A_S$ largely through differences in SLM.
+Consistent with this, PFT, plant height and biome explained very little variance in $A_F/A_S$ (Table~\ref{varpartpipe}).
+
+ We also partionined the variance of LMF and LAR (Table~\ref{varpartpipe}).
+When added to a model that included plant height and biome, PFT increased the variance explained in log-transformed LMF from \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[2],2)} to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[3],2)}\%.
+However, SLM was not a good substitute for PFT, as the explained variance decreased to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[4],2)}\%.
+For (log-transformed) LAR, PFT and SLM explained very little additional variance.
+
+ Although biome was a significant predictor for all variables tested, it consistently explained very little variation (Table~\ref{varpartpipe} and~\ref{varpartlmflar}).
+Pair-wise comparisons between biomes within PFT mostly did not show significant differences for $M_F/A_S$ or $A_F/A_S$.
+The only exception was a higher $A_F/A_S$ for tropical compared to temperate evergreen angiosperms (Fig.~\ref{fig:figureSI8}).
+
+ Unlike \cite{reich_temperature_2014}, we found no effect of mean annual temperature on the scaling of leaf biomass with woody biomass (Fig.~\ref{fig:figureSI9}).
+
+ We found coordination between an important leaf trait and whole-plant biomass partitioning. This has implications for global vegetation models.
+
+ How is allocation implemented in current DGVMs?
+Either constant allocation fraction, target allometric equations, or optimization \cite{de2014does}.
+Constant allocation fraction gives constant LAR by pft if leaf lifespan is exactly proportional to SLM.
+Target allometric equations can be made pft-invariant by using allometry of leaf area, not leaf mass.
+Optimization models probably already give the right answer because normally LAI is optimized (Woodward).
+
+ A recent study compared allocation routines in a number of leading ecosystem models \cite{de2014does}, and recommended to constrain allocation by observed biomass fractions instead of using constant allocation fractions, which is common practice.
+
+ As a first approximation, which is useful for large scale DGVMs, models should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly.
+Specific leaf area then gives the ratio of leaf biomass to stem biomass.
+
+ We found no effects of climate on anything, when analyzed in addition to PFT.
+This means that climate-related differences are captured by a shift in PFT (see also Fig.~\ref{figure1}, showing PFTs occupy different parts of the climate space).
+Our results contrast with those of \cite{reich_temperature_2014} but they studied whole-stand biomass, not individual plant biomass.
+So perhaps this means the climate effects are on stand density not individual plant effects.
+No effect of climate is good news for models that parameterize the globe by PFT, as this seems sufficient, based on our dataset.
+Are there experimental data that show temperature effects on allocation?
+What about water availability?
 
 %------------------------------------------------------------------------------------%
-\section{Methods}
+\section*{Methods}
 
-\begin{itemize}
-\item BAAD database (Falster et al., in review), including a total of \Sexpr{nrow(baad$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), aboveground woody biomass ($M_W$), total root biomass ($M_R$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
+ BAAD database (Falster et al., in review), including a total of \Sexpr{nrow(baad$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), aboveground woody biomass ($M_W$), total root biomass ($M_R$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
 1.3m) ($A_{Sbh}$).
@@
 
-\item We focus here only on aboveground biomass because we found no PFT differences in root-shoot scaling (Fig.~\ref{fig:figureSI6}), and approximately isometric scaling (sma slope = xxx) (consistent with other reports, \cite{hui_near_2014, cairns_root_1997}).
+ We focus here only on aboveground biomass because we found no PFT differences in root-shoot scaling (Fig.~\ref{fig:figureSI6}), and approximately isometric scaling (sma slope = xxx) (consistent with other reports, \cite{hui_near_2014, cairns_root_1997}).
 
-\item For the aboveground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total aboveground biomass) and leaf area ratio ($A_F / M_T$).
+ For the aboveground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total aboveground biomass) and leaf area ratio ($A_F / M_T$).
 LMF, LAR, SLM, $M_F$ and $A_F$ are related by,
@@
 
-\item We can further decompose aboveground biomass ($M_T$) into a size-related metric (basal stem area, $A_S$), and the ratio of $M_T$ to $A_S$, which is a function of wood density and plant woody volume to basal area.
+ We can further decompose aboveground biomass ($M_T$) into a size-related metric (basal stem area, $A_S$), and the ratio of $M_T$ to $A_S$, which is a function of wood density and plant woody volume to basal area.
 
@@
 
-\item We considered three plant functional types : evergreen angiosperms, evergreen gymnosperms, and decidous angiosperms.
+ We considered three plant functional types : evergreen angiosperms, evergreen gymnosperms, and decidous angiosperms.
 All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes (decided by contributing author, but checked with MAT and latitude).
 
-\item We did not attempt here to include SLA from another database (such as TRY), instead only used SLA directly estimated for the harvested plants.
+ We did not attempt here to include SLA from another database (such as TRY), instead only used SLA directly estimated for the harvested plants.
 
-\item Stem area was measured either at breast height (ca.
+ Stem area was measured either at breast height (ca.
 1.3m) or at the base of the plant (normally for small plants).
@@
 
-\item We used a generalized additive model (GAM) to visualize the relationships, but these were not used for formal inference.
+ We used a generalized additive model (GAM) to visualize the relationships, but these were not used for formal inference.
 
-\item We used linear mixed-effects models (with species within study as the random effect) to test for significance and to calculate confidence intervals for PFT-specific averages.
+ We used linear mixed-effects models (with species within study as the random effect) to test for significance and to calculate confidence intervals for PFT-specific averages.
 Least-square means for LMF and LAR were calculated with the 'lsmeans' package.
 
-\item All variables were log-transformed prior to analysis, to stabilize variance.
+ All variables were log-transformed prior to analysis, to stabilize variance.
 
-\item Variance partitioning with marginal (fixed effects) and conditional (fixed and random effects) $R^2$ for linear mixed-effects models as reported by \cite{nakagawa_general_2013}.
+ Variance partitioning with marginal (fixed effects) and conditional (fixed and random effects) $R^2$ for linear mixed-effects models as reported by \cite{nakagawa_general_2013}.
 
-\item For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception  \cite{lang_application_1991, chen_defining_1992}, for comparison to flat leaves.
-\end{itemize}
-
-%------------------------------------------------------------------------------------%
-\section{Results}
-
-\begin{itemize}
-
-\item We found large differences between PFTs in biomass partitioning between leaves and stems (Fig.~\ref{fig:figure2}, see also Fig.~\ref{fig:figureSI3}).
-Calculated least-square means of LMF and LAR (averages accounting for variation explained by plant height) showed large differences in mass-based allocation between PFTs (LMF), and small inconsistent differences in area-based allocation (LAR) (Fig.~\ref{fig:figure2}, see Fig.~\ref{fig:figureSI4} for raw data on leaf area ratio by PFT).
-
-\item The ratio of aboveground biomass to basal stem area did not differ by PFT (Fig.~\ref{fig:figureSI5}).
-
-\item However we found very large differences between PFTs in the ratio of leaf mass to basal stem area (Fig.~\ref{fig:figure3}), but the ratio leaf area to basal stem area converged between PFTs.
-
-\item We then expressed the average $M_F/A_s$ as a function of the specific leaf mass (SLM = $M_F / A_F$), a measure of the leaf construction cost to the plant.
-We found that, across PFTs, $M_F/A_S$ was approximately proportional to the average specific leaf mass for the PFT.
-As a result of this approximate proportionality, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure4}) (see also Fig.~\ref{fig:figureSI6} for raw data).
-
-\item This relationship also held approximately across species (Fig.~\ref{fig:figureSI7}).
-
-\item Because there was substantial variation between individual plants (Figs.~\ref{fig:figure3}), we estimated the variance explained by PFT, plant height, and biome in $M_F/A_S$ using a linear mixed-effects model (see Table~\ref{varpartpipe}).
-We found that plant height and biome explained only \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[2],2)}\%, but when PFT was added to the model, the variance explained increased to \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[3],2)}\%.
-If we used SLM as a predictor instead of PFT, the $R^2$ was similar (\Sexpr{100*signif(Table_pipemodel_varpart$Marginal[4],2)}\%), showing that PFT affected $M_F/A_S$ largely through differences in SLM.
-Consistent with this, PFT, plant height and biome explained very little variance in $A_F/A_S$ (Table~\ref{varpartpipe}).
-
-\item We also partionined the variance of LMF and LAR (Table~\ref{varpartpipe}).
-When added to a model that included plant height and biome, PFT increased the variance explained in log-transformed LMF from \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[2],2)} to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[3],2)}\%.
-However, SLM was not a good substitute for PFT, as the explained variance decreased to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[4],2)}\%.
-For (log-transformed) LAR, PFT and SLM explained very little additional variance.
-
-\item Although biome was a significant predictor for all variables tested, it consistently explained very little variation (Table~\ref{varpartpipe} and~\ref{varpartlmflar}).
-Pair-wise comparisons between biomes within PFT mostly did not show significant differences for $M_F/A_S$ or $A_F/A_S$.
-The only exception was a higher $A_F/A_S$ for tropical compared to temperate evergreen angiosperms (Fig.~\ref{fig:figureSI8}).
-
-\item Unlike \cite{reich_temperature_2014}, we found no effect of mean annual temperature on the scaling of leaf biomass with woody biomass (Fig.~\ref{fig:figureSI9}).
-
-\end{itemize}
-
-%------------------------------------------------------------------------------------%
-\section{Discussion}
-
-\begin{itemize}
-
-\item We found coordination between an important leaf trait and whole-plant biomass partitioning. This has implications for global vegetation models.
-
-\item How is allocation implemented in current DGVMs?
-Either constant allocation fraction, target allometric equations, or optimization \cite{de2014does}.
-Constant allocation fraction gives constant LAR by pft if leaf lifespan is exactly proportional to SLM.
-Target allometric equations can be made pft-invariant by using allometry of leaf area, not leaf mass.
-Optimization models probably already give the right answer because normally LAI is optimized (Woodward).
-
-\item A recent study compared allocation routines in a number of leading ecosystem models \cite{de2014does}, and recommended to constrain allocation by observed biomass fractions instead of using constant allocation fractions, which is common practice.
-
-\item As a first approximation, which is useful for large scale DGVMs, models should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly.
-Specific leaf area then gives the ratio of leaf biomass to stem biomass.
-
-\item We found no effects of climate on anything, when analyzed in addition to PFT.
-This means that climate-related differences are captured by a shift in PFT (see also Fig.~\ref{figure1}, showing PFTs occupy different parts of the climate space).
-Our results contrast with those of \cite{reich_temperature_2014} but they studied whole-stand biomass, not individual plant biomass.
-So perhaps this means the climate effects are on stand density not individual plant effects.
-No effect of climate is good news for models that parameterize the globe by PFT, as this seems sufficient, based on our dataset.
-Are there experimental data that show temperature effects on allocation?
-What about water availability?
-
-\end{itemize}
-
-%------------------------------------------------------------------------------------%
-\section{Lit notes}
-
-Reich et al 1992, for seedlings, negative relationship leaf lifespan and LAR.
-This is different from our results, since SLM is nearly proportional to leaf lifespan, and we found no relationship between LAR and SLM (but do they have all the PFTs??).
-However, across stands, positive relationship between leaf biomass and leaf lifespan, which is consistent with our results.
-
-Cornelissen et al 1996 positive relationship SLA and LAR, but this is for tiny seedlings harvested 21 days after first leaf appearance. But check references and who cites this!
-
-\cite{friend_carbon_2014} showed that 30\% of the uncertainty in global vegetation carbon simulated with seven vegetation models is due to the residence time of carbon in vegetation.
-This residence time depends on turnover rates of the various plant pools, and hence also on the distribution of biomass between these pools.
-For example, vegetation with a higher leaf mass fraction will have a lower overall residence time of carbon.
-
-\cite{de2014does} "We have shown that allocation approaches that are constrained by biomass fractions (such as functional relationships) were more successful at capturing observed trends, and were generally more robust, than approaches based on allocation coefficients."
-
-\cite{friedlingstein_toward_1999} make no mention of plant functional type differences or the need to vary allocation by pft.
-
-\cite{givnish_adaptive_2002} discusses costs and benefits of evergreen vs.
-deciduous habit, in terms of constructing a growth model that takes into account photosynthesis, leaf lifespan, belowground costs.
-However he does not consider or report on differences in allocation associated with leaf lifespan.
-Instead he simply states that 'leaf allocation ratio' relates to leaf lifespan, which of course is only true if total leaf mass is constant.
-
-\cite{callaway_biomass_1994} example of lower leaf mass / sapwood area for desert population of PIPO vs.
-montane population.
-For pointing out small-scale climatic effects on allometry (that may not be obvious in global scale analyses).
-
-\cite{poorter_biomass_2012} found a slightly larger LMF in gymnosperms, but the difference was small and could not split between deciduous and evergreen angiosperms due to a lack of data.
-They argue that "most conifers retain leaves for two or more years.
-Thus, the higher LMF may simply be the consequence of leaf longevity rather than higher assimilate partitioning per se.
-In this way, conifers are able to compensate for their lower physiological activity per gram of leaf." This raises the question whether LMF differs by angiosperms and gymnosperms, or as a consequence of leaf longevity (especially deciduous vs.
-evergreen).
-
-\cite{enquist_global_2002} report 2.6 times higher leaf mass at given stem mass for conifers compared to angiosperms, but did not separate further, nor interpreted this difference in terms of SLA.
-Their interpretation is 'that conifers typically retain three cohorts of leaves that have less well-developed [...] mesophyll'.
-Not sure how that latter bit makes sense, and leaf lifespan interpretation is no good; because it still begs the question why would conifers not simply have 1/3 less leaf mass per cohort?
-
-\cite{mccarthy_consistency_2007} further demonstrate large difference in LMF between angiosperms and conifers, and note that a large chunk of the residuals is correlated with SLA and LL.
-They interpret this rather vaguely as 'leaf morphology may constrain leaf mass distribution'.
-Their SLA was not on the same individuals, but they could have calculated PFT average LAR anyway, but did not.
-
-\cite{oneill_comparative_1981} reported much higher leaf biomass for gymnosperms vs.
-angiosperms, at a given woody biomass increment.
-This means (as Niklas reported) a much higher productivity per unit leaf biomass for angiosperms.
-
-Serious lack of biomass data to constrain models.
-Studies like \cite{wolf_forest_2011} resort to studying individual tree allometry based on dividing stand biomass by stand density.
-\cite{wolf_allometric_2010} did not find any differences in biomass fractions (root,stem,leaf) between angiosperms and gymnosperms (but again this was based on stand biomass data, but still this is strange!).
-
-\cite{hui_near_2014} Isometric scaling of above vs. belowground biomass in chinese forest biomass database.
-
-\cite{sardans_tree_2013} report similiar LMF for evergreen angio. and gymnosperms, and ca. 40\% less leaf mass for decid. angio. Local study, 3000 plots in Catalonia.
-
-\cite{reich_temperature_2014} "On average, when standardized for MAT and Mstem, conifers tended to have more Mfol and less Mroot than angiosperm forests".
-Their Fig.
-2 shows trends in LMF with MAT etc., in bins of stem size.
-Also allometric analysis with MAT as covariate, showing significance of MAT.
-For Angiosperms, our explanation would be that evergreen angio is more common at high MAT.
-But what about conifers?
-
-\cite{cairns_root_1997} found no difference in root-shoot ratio between gymno and angio.
-
-\cite{delucia_climate-driven_2000} LA/SA decreases with summer VPD for Pinus but not non-Pinus gymno.
-Also Pinus has much lower LA/SA than non-Pinus; suggesting that we should separate gymnosperms into those groups as well.
-
-\cite{grier_notes:_1974} early study on leaf mass proportionality to sapwood area (and uses mass not area).
+ For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception  \cite{lang_application_1991, chen_defining_1992}, for comparison to flat leaves.
 
@@
     \includegraphics{Figure1_MAPMAT_baad_vs_worldclim.pdf}
-    \caption{Mean annual precipitation and mean annual temperature for all included studies.
-Hexagons are terrestrial observations extracted from Worldclim (darker bins indicating more observations).
-Note that Worldclim observations include land areas where woody vegetation does not occur (desert, tundra).
-The deciduous gymnosperms were not included in any analysis in this paper, but are shown here.
+    \caption{Global coverage of the climate space by the dataset.
+Hexagons are terrestrial observations extracted from Worldclim (colored by number of 0.5\degree~cells).
+Note that Worldclim observations include land areas where woody vegetation does not occur (e.g., desert, tundra).
+The deciduous gymnosperms were not included in any analysis in this paper, but are shown in the figure.
 }

### 2014-10-21 manuscript/draftpaper.Rnw
Found very minor effects of MAT on LMF, added to Fig 9, and MAP panel. smoothplotbypft can do lm instead of gam.

@@
 
- Unlike \cite{reich_temperature_2014}, we found no effect of mean annual temperature on the scaling of leaf biomass with woody biomass (Fig.~\ref{fig:figureSI9}).
+ Unlike \cite{reich_temperature_2014}, we found only very weak effects of mean annual temperature the scaling of leaf biomass with woody biomass (Fig.~\ref{fig:figureSI9}).
+The size-invariant relationship was only significant for deciduous angiosperms, but it was a very weak correlation that explained very little additional variation.
+These results were confirmed with a mixed-effects model of MF vs.
+MS, with MAT and PFT as covariates (see Table XXX).
+We found no significant effects of mean annual precipitation of leaf-woody biomass relationships (Table XXX).
 
@@
 
- We found no effects of climate on anything, when analyzed in addition to PFT.
+ We found no appreciable effects of climate on anything, when analyzed in addition to PFT.
 This means that climate-related differences are captured by a shift in PFT (see also Fig.~\ref{figure1}, showing PFTs occupy different parts of the climate space).

### 2014-10-21 R/functions-figures.R

@@ -142,10 +142,12 @@ lsmeansPlot <- function(x,lma,lets,ylim=NULL,...){
 
 
 smoothplotbypft <- function(x,y,data,pointcols=alpha(c("blue","red","forestgreen"),0.3),
+                            fittype=c("gam","lm"),
                             linecols=c("deepskyblue3","red","chartreuse3"), 
                             xlab=NULL, ylab=NULL,logaxes=TRUE,
                             ...){
   
+  fittype <- match.arg(fittype)
   data$pft <- as.factor(data$pft)
   data <- droplevels(data)
   data$X <- eval(substitute(x),data)
@@ -157,18 +159,30 @@ smoothplotbypft <- function(x,y,data,pointcols=alpha(c("blue","red","forestgreen
   d <- split(data, data$pft)
   
   hran <- lapply(d, function(x)range(x$X, na.rm=TRUE))
-  fits <- lapply(d, function(x)try(fitgam("X","Y",x, k=4)))
+  if(fittype == "gam"){
+    fits <- lapply(d, function(x)try(fitgam("X","Y",x, k=4)))
+  } else {
+    fits <- lapply(d, function(x)lm(Y ~ X, data=x))
+  }
   
   with(data, plot(X, Y, axes=FALSE, pch=16, col=pointcols[pft],
                   xlab=xlab, ylab=ylab, ...)) 
   if(logaxes)magaxis(side=1:2, unlog=1:2)
   
   for(i in 1:length(d)){
-    nd <- data.frame(X=seq(hran[[i]][1], hran[[i]][2], length=101))
-    if(!inherits(fits[[i]], "try-error")){
-      p <- predict(fits[[i]],nd,se.fit=TRUE)
-      addpoly(nd$X, p$fit-2*p$se.fit, p$fit+2*p$se.fit, col=alpha("lightgrey",0.7))
-      lines(nd$X, p$fit, col=linecols[i], lwd=2)
+    
+    if(fittype == "gam"){
+      nd <- data.frame(X=seq(hran[[i]][1], hran[[i]][2], length=101))
+      if(!inherits(fits[[i]], "try-error")){
+        p <- predict(fits[[i]],nd,se.fit=TRUE)
+        addpoly(nd$X, p$fit-2*p$se.fit, p$fit+2*p$se.fit, col=alpha("lightgrey",0.7))
+        lines(nd$X, p$fit, col=linecols[i], lwd=2)
+      }
+    }
[…]

### 2014-10-21 statanalysis.R

@@ -1,10 +1,11 @@
 
 
+
+#---------------------------------------------------------------------------------------#
 # Variance partitioning to fixed and random effects 
 source("load.R")
 source("R/preparedataset.R")
 source("R/rsquaredglmm.R")
-options(warn=-1)
 
 # Sequential R2, backwards elimination of all fixed effects.
 # ((not used at the moment))
@@ -151,8 +152,29 @@ tab_lar <- tabFun(dat_alfmso, c("lalf_mso","h.t"))
 save(tab_mlfastba,tab_alfastba,tab_lmf,tab_lar,
      file="manuscript/tables/Table_counts.RData")
 
-options(warn=0)
+
+#-----------------------------------------------------------------------------------------#
+  
 
 
-  
-  
+# Test of effect of MAT and MAP on leaf - stem scaling
+
+# Data subset
+d <- droplevels(subset(dataset2, !is.na(m.st) & !is.na(m.lf) & !is.na(MAT)))
+
+lme0 <- lme(log10(m.lf) ~ log10(m.st)*pft, random=~log10(m.st)|Group, data=d,method="ML",
+            na.action=na.omit)
+lme1 <- lme(log10(m.lf) ~ log10(m.st)*pft*MAT, random=~log10(m.st)|Group, data=d,method="ML",
+            na.action=na.omit)
+lme2 <- lme(log10(m.lf) ~ log10(m.st)*pft*MAP, random=~log10(m.st)|Group, data=d,method="ML",
+            na.action=na.omit)
+
+
+# MAT significant
+anova(lme0, lme1)
+
+# MAP is not
+anova(lme0, lme2)
+
+
+

### 2014-10-21 manuscript/draftpaper.Rnw
Text improvements, new appendix on predicting basal D from DBH, code reorganized a bit. Figure 1 has PFT averages for MAP and MAT.

@@
 
-There is no shortage of theoretical developments in understanding biomass allocation (\cite{cannell_carbon_1994,franklin_modeling_2012,enquist_land_2012}).
-In particular, theoretical work focusing on optimal partitioning theory is promising (e.g.
-\cite{valentine_modeling_2012}) but is yet to lead to operational models for dynamic global vegetation models.
-
 Previous syntheses on plant biomass relationships has predominantly focused on the universal nature of size-biomass relationships in plants.
@@
 It has been reported numerous times that gymnosperms (with higher SLM) have higher leaf mass fraction than angiosperms (with lower SLM) \cite{enquist_global_2002, cheng_relationships_2007, mccarthy_consistency_2007, poorter_biomass_2012, sardans_tree_2013}, although a recent study reported no difference (\cite{wolf_allometric_2010}).
-No study to date has been able to analyze differences between deciduous and evergreen angiosperms, or link differences in partitioning between PFTs to other plant traits such as SLM.
+No study to date has been able to further tease apart PFT differences, or test the potentially confounding effects of climate.
 
@@
 
-We consider here the relationship between leaf biomass and area with stem diameter (an important allometric relationship, Enquist others, and related to the pipe model if the ratio sapwood / heartwood is relatively constant, an important measure of hydraulic supply), the aboveground leaf mass fraction (partioning of biomass into leaf vs.
-woody pools), and the leaf area ratio (a measure of the photosynthetic area per unit woody biomass).
-We expect that $M_F$ and $A_F$ are approximately proportional to the square of stem diameter, which would remove size-related shifts in allometry that often confound interpretations.
-For each of these variables, we ask whether 1) there are differences between major plant functional types after removing size-related variation, 2) there are differences due to climate, compared between major biomes, or as a function of the mean annual temperature (as recently reported for whole stand biomass, \cite{reich_temperature_2014}).
+We found no PFT differences in root-shoot scaling (Fig.~\ref{fig:figureSI6}), and approximately isometric scaling (sma slope = xxx), which is consistent with other reports, \cite{hui_near_2014, cairns_root_1997}).
+Because of this simple scaling between above- and below-ground biomass pools, we further focus on biomass partioning between leaves and aboveground woody pools.
+
+The main objective was to test whether PFT-specific differences exist in the ratio of leaf mass to total above-ground biomass (the 'leaf mass fraction', LMF), and the leaf area to aboveground biomass (the 'leaf area ratio', LAR).
+Because PFT and climate are confounded to some extent (Fig.~\ref{figure1}), we also test whether climate has additional effects on biomass partitioning, or whether these effects are captured by PFT alone.
 
@@
 
- The ratio of aboveground biomass to basal stem area did not differ by PFT (Fig.~\ref{fig:figureSI5}).
+We can further decompose LMF into two components representing different aspects of plant function and structure.
+The ratio of leaf biomass ($M_F$) to stem basal area ($A_S$) relates to hydraulic supply, and the ratio of total aboveground biomass ($M_T$) and $A_S$ is a function of plant shape (stem taper, plant height at given stem diameter), and wood density \cite{chave_tree_2005}.
 
- However we found very large differences between PFTs in the ratio of leaf mass to basal stem area (Fig.~\ref{fig:figure3}), but the ratio leaf area to basal stem area converged between PFTs.
+$$LMF = M_F / M_T = (M_F / A_S) (M_T / A_S)^{-1}$$
 
- We then expressed the average $M_F/A_s$ as a function of the specific leaf mass (SLM = $M_F / A_F$), a measure of the leaf construction cost to the plant.
+We found that ratio of aboveground biomass to basal stem area ($M_T/A_S$) did not differ by PFT (Fig.~\ref{fig:figureSI5}).
+However we found very large differences between PFTs in the ratio $M_F/A_S$(Fig.~\ref{fig:figure3}).
+
+ We then expressed the average $M_F/A_S$ as a function of the specific leaf mass (SLM = $M_F / A_F$), a measure of the leaf construction cost to the plant.
 We found that, across PFTs, $M_F/A_S$ was approximately proportional to the average specific leaf mass for the PFT.
@@
  Unlike \cite{reich_temperature_2014}, we found only very weak effects of mean annual temperature the scaling of leaf biomass with woody biomass (Fig.~\ref{fig:figureSI9}).
-The size-invariant relationship was only significant for deciduous angiosperms, but it was a very weak correlation that explained very little additional variation.
+However, the variance explained by MAT was very small, compared to the large difference between PFTs.
 These results were confirmed with a mixed-effects model of MF vs.
-MS, with MAT and PFT as covariates (see Table XXX).
-We found no significant effects of mean annual precipitation of leaf-woody biomass relationships (Table XXX).
+MS, with MAT and PFT as covariates (see Table~\ref{table:matlmeanova}).
+We found no significant effects of mean annual precipitation of leaf-woody biomass relationships (Table~\ref{table:maplmeanova}).
 
- We found coordination between an important leaf trait and whole-plant biomass partitioning. This has implications for global vegetation models.
+In summary, we found coordination between an important leaf trait and whole-plant biomass partitioning. This has implications for global vegetation models.
 
@@
 A recent study compared allocation routines in a number of leading ecosystem models \cite{de2014does}, and recommended to constrain allocation by observed biomass fractions instead of using constant allocation fractions, which is common practice.
-
 As a first approximation, which is useful for large scale DGVMs, models should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly.
@@
 
- We found no appreciable effects of climate on anything, when analyzed in addition to PFT.
+There is no shortage of theoretical developments in understanding biomass allocation (\cite{cannell_carbon_1994,franklin_modeling_2012,enquist_land_2012}).
+In particular, theoretical work focusing on optimal partitioning theory is promising (e.g.
+\cite{valentine_modeling_2012}) but is yet to lead to operational models for dynamic global vegetation models.
+
+We found no effects of MAP on biomass partitioning, and only very weak effects of MAT on LMF, when analyzed in addition to PFT.
 This means that climate-related differences are captured by a shift in PFT (see also Fig.~\ref{figure1}, showing PFTs occupy different parts of the climate space).
@@
 No effect of climate is good news for models that parameterize the globe by PFT, as this seems sufficient, based on our dataset.
-Are there experimental data that show temperature effects on allocation?
-What about water availability?
 
@@
 
- BAAD database (Falster et al., in review), including a total of \Sexpr{nrow(baad$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), aboveground woody biomass ($M_W$), total root biomass ($M_R$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
+We used the recently compiled Biomass And Allometry Database (BAAD) (Falster et al., in review), including a total of \Sexpr{nrow(baad$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), aboveground woody biomass ($M_W$), total root biomass ($M_R$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
 1.3m) ($A_{Sbh}$).
-For each analysis, we therefore used different subsets of the data (far fewer observations are available for root biomass, in particular), sample sizes are included in the caption of each figure (\emph{to be added}).
-We further excluded deciduous gymnosperms (since very few data were available), and glasshouse and common garden studies, including only field grown woody plants (including plantations).
-
- We focus here only on aboveground biomass because we found no PFT differences in root-shoot scaling (Fig.~\ref{fig:figureSI6}), and approximately isometric scaling (sma slope = xxx) (consistent with other reports, \cite{hui_near_2014, cairns_root_1997}).
+For each analysis, we therefore used different subsets of the data (far fewer observations are available for root biomass, in particular).
+Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4} We exluded glasshouse and common garden studies, including only field grown woody plants (including plantations).
+We considered three plant functional types : evergreen angiosperms, evergreen gymnosperms, and decidous angiosperms.
+We excluded  deciduous gymnosperms because few data were available.
+All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes (decided by contributing author, but checked with MAT and latitude).
 
@@
 
- We can further decompose aboveground biomass ($M_T$) into a size-related metric (basal stem area, $A_S$), and the ratio of $M_T$ to $A_S$, which is a function of wood density and plant woody volume to basal area.
+We were unable to further decompose $A_S/M_W$ into wood volume and wood density components (cf. \cite{chave_tree_2005}) because wood density data were available only for a small subset of data.
+We did not include SLM estimates from another database (such as TRY), instead only used SLA directly estimated for the harvested plants (typically for a subsample of leaves).
+For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception  \cite{lang_application_1991, chen_defining_1992}, for comparison to flat leaves.
 
-$$LMF = M_F / M_T = (M_F / A_S) (M_T / A_S)^{-1}$$
-
-We test for PFT differences to both these components of biomass partioning.
-We were unable to further split $A_S/M_W$ because wood density data were available only for a subset of data.
-Chave has shown that etc.
-etc.
-
- We considered three plant functional types : evergreen angiosperms, evergreen gymnosperms, and decidous angiosperms.
-All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes (decided by contributing author, but checked with MAT and latitude).
-
- We did not attempt here to include SLA from another database (such as TRY), instead only used SLA directly estimated for the harvested plants.
-
- Stem area was measured either at breast height (ca.
-1.3m) or at the base of the plant (normally for small plants).
-For the subset of the data where both were measured, we estimated basal stem area (Asba) from breast height area (Asbh) by fitting XXX equation.
-Results were similar when we used either measure of stem area (not shown).
+Stem cross-sectional area was measured either at breast height (ca.
+1.3m) and/or at the base of the plant.
+For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using non-linear regression (see~SIXXX).
+The main results were similar when we repeated it for data where either $A_S$ or $A_{Sbh}$ were directly measured (not shown).
 
 We used a generalized additive model (GAM) to visualize the relationships, but these were not used for formal inference.
-
- We used linear mixed-effects models (with species within study as the random effect) to test for significance and to calculate confidence intervals for PFT-specific averages.
-Least-square means for LMF and LAR were calculated with the 'lsmeans' package.
-
- All variables were log-transformed prior to analysis, to stabilize variance.
-
+Instead we used linear mixed-effects models (with species within study as the random effect) to test for significance and to calculate confidence intervals for PFT-specific averages.
+Least-square means for LMF and LAR were calculated with the 'lsmeans' package in R.
+All variables were log-transformed prior to analysis.
 Variance partitioning with marginal (fixed effects) and conditional (fixed and random effects) $R^2$ for linear mixed-effects models as reported by \cite{nakagawa_general_2013}.
-
- For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception  \cite{lang_application_1991, chen_defining_1992}, for comparison to flat leaves.
 
@@
     \caption{Global coverage of the climate space by the dataset.
-Hexagons are terrestrial observations extracted from Worldclim (colored by number of 0.5\degree~cells).
+Hexagons are terrestrial observations extracted from Worldclim (colored by number of 0.5\degree~cells, cut off at 6000mm MAP).
 Note that Worldclim observations include land areas where woody vegetation does not occur (e.g., desert, tundra).
-The deciduous gymnosperms were not included in any analysis in this paper, but are shown in the figure.
+Triangles are averages for the three PFTs across all contributed studies.
 }
@@
 
-<<tablepreload>>=
+<<tablepreload1>>=
 load("tables/Table_counts.RData")
-printit <- function(x, caption){
+printit <- function(x, caption, label=NULL){
   print(xtable(x, align="rlll|l",
-               caption=caption
+               caption=caption, label=label
                ), hline.after=c(0,3), caption.placement="top")
@@
 <<results='asis'>>=
-printit(tab_mlfastba, caption="Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$.
+printit(tab_mlfastba, 
+        label="table:count1",
+        caption="Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$.
 Individual plants (number of species).
@@
 
-% <<results='asis'>>=
-% printit(tab_alfastba, caption="Number of observations used in the analysis of Af/As.
-Individual plants (number of species).
-DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
-% @
-%
+<<results='asis'>>=
+printit(tab_alfastba, 
+        label="table:count2",
+        caption="Number of observations used in the analysis of Af/As. Individual plants (number of species). DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
+@
 
 <<results='asis'>>=
-printit(tab_lmf, caption="Number of observations used in the analysis of leaf mass fraction (LMF) and leaf area ratio (LAR).
+printit(tab_lmf, 
+        label="table:count3",
+        caption="Number of observations used in the analysis of leaf mass fraction (LMF) and leaf area ratio (LAR).
 Individual plants (number of species).
@@
 
-% 
-% <<results='asis'>>=
-% printit(tab_lar, caption="Number of observations used in the analysis of leaf area ratio (LAR).
+<<results='asis'>>=
+printit(tab_lar, 
+        label="table:count4",
+        caption="Number of observations used in the analysis of leaf area ratio (LAR).
 Individual plants (number of species).
 DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
-% @
+@
+
+<<tablepreload2>>=
+library(car)
+load("tables/Fits_lme_mlfmst_MAPMAT.RData")
+printanova <- function(x, caption, label=NULL){
+  rownames(x) <- gsub("log10(m.st)", "MS", rownames(x), fixed=TRUE) 
+  rownames(x) <- gsub("pft", "PFT", rownames(x)) 
+  print(xtable(x, caption=caption, label=label),
+                                        caption.placement="top")
+}
+@
+
+<<results='asis'>>=
+printanova(Anova(mlfmst_lme1), 
+           label="table:matlmeanova",
+           caption="Type-II Anova table for a linear mixed-effects model of $M_F$ as a function of $M_S$ (both log-transformed), PFT and mean annual temperature (MAT).
+Species by dataset was used as the random effect.
+Further analysis of the interactions revealed that MAT only significantly affected the relationship between $M_F$ and $M_S$ for the deciduous angiosperms (not shown).")
+@
+
+<<results='asis'>>=
+printanova(Anova(mlfmst_lme2), 
+           label="table:maplmeanova",
+           caption="Type-II Anova table for a linear mixed-effects model of $M_F$ as a function of $M_S$ (both log-transformed), PFT and mean annual precipitation (MAT).
+Species by dataset was used as the random effect.
+We did not analyze MAT and MAP together in one model, as they were strongly correlated, making interpretation of the fitted model difficult.")
+@
+
+\clearpage
+\section{Supporting Information Methods}
+
+\subsection*{Predicting basal diameter}
+
+<<>>=
+load("tables/basalafit.RData")
+@
+
+Using the subset of data where basal stem diameter ($A_S$) and diameter at breast height ($A_{Sbh}$) were measured (a total of \Sexpr{length(residuals(basalafit))} observations covering the three major PFTs), we developed a non-linear model to predict $A_S$ when only $A_{Sbh}$ and plant height ($H$) were measured.
+We fit the following equation, using non-linear regression.
+
+$$
+D_S = D_{Sbh} H^a / (H - H_{bh})^a
+$$
+
+where $D_S$ is the basal stem diameter (m), $D_{Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{Sbh}$ was measured (typically 1.3 or 1.34m), and $a$ was further expressed as a function of plant height:
+
+$$
+a = c_0H^c_1
+$$
+
+The estimated coefficients were $c_0$ = \Sexpr{coef(basalafit)[1]}, $c_1$ = \Sexpr{coef(basalafit)[2]}, root-mean square error = \Sexpr{summary(basalafit)$sigma}.
 

### 2014-10-21 figures.R

@@ -31,6 +31,8 @@ map <- climspace$MAP_WC
 mat <- climspace$MAT_WC/10
 mapmat <- baad[!duplicated(baad[,c("MAP","MAT")]),]
 mapmat$vegetation <- as.factor(mapmat$vegetation)
+mapmat <- droplevels(subset(mapmat, pft != "DG"))
+mmpft <- summaryBy(MAP + MAT ~ pft, data=mapmat, FUN=mean, na.rm=TRUE)
 
 h <- hexbin(map ~ mat)
 cells <- hcell2xy(h)
@@ -47,7 +49,7 @@ greyCols <- grey(seq(0.85,0.2,length=nlevels(hcut)))
 
 to.pdf({
   par(pty='s', cex.lab=1.2)
-  plot(cells, type='n',
+  plot(cells, type='n', ylim=c(0,6000),
        xlab = expression("Mean annual temperature"~(degree*C)), 
        ylab = "Mean annual precipitation (mm)")
   for(i in 1:nhex){
@@ -56,15 +58,16 @@ to.pdf({
             col=greyCols[hcut[i]])
   }
   l <- legend("topleft", levels(hcut), fill=greyCols, cex=0.7, title="Nr cells", bty='n')
-  
+  box()
   
   mapmat$pft <- as.factor(mapmat$pft)
-  hCols <- alpha(c("blue","cyan1","red","forestgreen"),0.6)
-  with(mapmat, points(MAT, MAP, pch=19, col=hCols[pft], cex=1.2))
+  hCols <- alpha(c("blue","red","forestgreen"),0.6)  #"cyan1",
+  with(mapmat, points(MAT, MAP, pch=19, col=hCols[pft], cex=1.1))
+  with(mmpft, points(MAT.mean, MAP.mean, col=hCols[pft], pch=24, cex=1.5, bg="white", lwd=2))
   with(subset(mapmat, pft=="DG"), points(MAT, MAP, pch=19, col=hCols[pft], cex=1.2))
   legend(l$rect$left + l$rect$w, 
          l$rect$top, title="Plant functional type",
-         c("Deciduous Angiosperm", "Deciduous Evergreen", 
+         c("Deciduous Angiosperm", #"Deciduous Evergreen", 
            "Evergreen Angiosperm", "Evergreen Gymnosperm"),
          pch=19, col=hCols, pt.cex=1.2, cex=0.7, bty='n')
 }, filename="manuscript/figures/Figure1_MAPMAT_baad_vs_worldclim.pdf", width=6, height=6)

### 2014-10-21 R/predictBasalA.R
Text improvements, new appendix on predicting basal D from DBH, code reorganized a bit. Figure 1 has PFT averages for MAP and MAT.

@@ -1,0 +1,19 @@
+predictBasalA <- function(dat=NULL, alldat=baad, returnwhat=c("prediction","fit")){
+  
+  returnwhat <- match.arg(returnwhat)
+  
+  # Predicted basal diameter. See R/predict_dba...R
+  test <- subset(alldat, !is.na(d.ba) & !is.na(d.bh) & !is.na(h.t) & !is.na(h.bh) & h.t > h.bh)
+  fit <- nls(d.ba ~ d.bh * h.t^(c0*h.t^c1) /(h.t - h.bh)^(c0*h.t^c1), start=list(c0=0.9, c1=0.7),
+             data=test)
+  
+  if(returnwhat == "fit")return(fit)
+  
+  d.ba2 <- predict(fit, dat)
+  d.ba2[dat$h.bh >= dat$h.t] <- NA
+  d.ba2[!is.na(dat$d.ba)] <- dat$d.ba[!is.na(dat$d.ba)]
+  a.stba2 <- (pi/4)*d.ba2^2
+  
+  return(a.stba2)
+  
+}

### 2014-10-21 R/prepareDataset.R

@@ -31,20 +31,6 @@ dataset <- within(dataset, {
 })
 
 # Predict basal diameter from breast-height
-predictBasalA <- function(dat, alldat=baad){
-  # Predicted basal diameter. See R/predict_dba...R
-  test <- subset(alldat, !is.na(d.ba) & !is.na(d.bh) & !is.na(h.t) & !is.na(h.bh) & h.t > h.bh)
-  fit <- nls(d.ba ~ d.bh * h.t^(a*h.t^b) /(h.t - h.bh)^(a*h.t^b), start=list(a=0.9, b=1),
-             data=test)
-  
-  d.ba2 <- predict(fit, dat)
-  d.ba2[dat$h.bh >= dat$h.t] <- NA
-  d.ba2[!is.na(dat$d.ba)] <- dat$d.ba[!is.na(dat$d.ba)]
-  a.stba2 <- (pi/4)*d.ba2^2
-  
-return(a.stba2)
-}
-
 dataset$a.stba2 <- predictBasalA(dataset)
 
 

### 2014-10-21 dependencies.R

@@ -1,9 +1,10 @@
 #!/usr/bin/env Rscript
 
 pkgs <- c("lmerTest","maps","mapdata","maptools","gdata","knitr","smatr","scales",
-          "xtable","magicaxis","rmarkdown","dplyr","sciplot","epade",
+          "xtable","magicaxis","rmarkdown","dplyr",
           "multcomp","doBy","mgcv","reshape2","lme4","Hmisc","hexbin",
-          "lattice","grid","stringr","raster","dismo","XML","rgdal","lsmeans","gplots")
+          "lattice","grid","stringr","raster","dismo","XML","rgdal","lsmeans",
+          "gplots","car")
 
 pkgs.missing <- setdiff(pkgs, rownames(installed.packages()))
 install.packages(pkgs.missing)

### 2014-10-22 R/functions-figures.R
Root-shoot ratio analysis update, Figure SI2, lsmeansPlot calculates significance letters

@@ -119,26 +119,48 @@ meansbypft <- function(yvar1, yvar2=NULL, pftvar,
 }
 
 
-lsmeansPlot <- function(x,lma,lets,ylim=NULL,...){
+lsmeansPlot <- function(y,x,lets,ylim=NULL,...){
   
-  uci <- 10^x$lsmeans.table[["Upper CI"]]
-  lci <- 10^x$lsmeans.table[["Lower CI"]]
-  Y <- 10^x$lsmeans.table[["Estimate"]]
+  
+  if(class(rootlsmeans)[1] != "lsmeans")
+    stop("Need object returned by lmerTest::lsmeans")
+  
+  uci <- 10^y$lsmeans.table[["Upper CI"]]
+  lci <- 10^y$lsmeans.table[["Lower CI"]]
+  Y <- 10^y$lsmeans.table[["Estimate"]]
   
   if(is.null(ylim))ylim <- c(0, max(uci))
   
-  plot(lma$y, Y, 
-       ylim=ylim, 
-       panel.first={
-         arrows(x0=lma$lci, x1=lma$uci, y0=Y, y1=Y,
-                code=3,angle=90,length=0.025,col=Cols)
-         arrows(x0=lma$y, x1=lma$y, y0=lci, y1=uci, angle=90, code=3, 
-                length=0.025, col=Cols)
-       }, pch=19, col=Cols,...)
+  numx <- is.list(x) & "lci" %in% names(x)
+  
+  lets <- cld.lsmeans(y)$Letters
+  
+  if(numx){
+    plot(x$y, Y, 
+         ylim=ylim, 
+         panel.first={
+           arrows(x0=x$lci, x1=x$uci, y0=Y, y1=Y,
+                  code=3,angle=90,length=0.025,col=Cols)
+           arrows(x0=x$y, x1=x$y, y0=lci, y1=uci, angle=90, code=3, 
+                  length=0.025, col=Cols)
+         }, pch=19, col=Cols,...)
+    u <- par()$usr
+    text(x$y, u[3] + 0.0*(u[4]-u[3]), lets, pos=3, cex=0.9)
+    
+  } else {
+    
+    plot(x, Y, ylim=ylim, col=Cols, pch=19, 
+         panel.first= arrows(x0=x, x1=x, y0=lci, y1=uci, angle=90, code=3, 
+                                       length=0.025, col=Cols),
[…]

### 2014-10-22 manuscript/draftpaper.Rnw
Root-shoot ratio analysis update, Figure SI2, lsmeansPlot calculates significance letters

@@
 
-\textbf{The partitioning of biomass in terrestrial vegetation between leaves, stems and roots is a critical component of the global carbon cycle.
-Global vegetation models consistently point to biomass partioning as one of the most uncertain and important drivers for the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, and global predictions of carbon storage \cite{ise_comparison_2010}.
+\textbf{The partitioning of biomass in terrestrial vegetation between leaves, stems and roots is a critical component of dynamic global vegetation models (DGVMs).
+Biomass partioning is consistently shown to be one of the most uncertain and important drivers for the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, and global predictions of carbon storage \cite{ise_comparison_2010}.
 Little is known about direct and indirect effects of climate on biomass partitioning, largely due to the lack of a global individual-based biomass dataset.
@@
 We found that specific leaf mass, an indicator of leaf construction cost, explains these differences so that the PFTs converged in the ratio  of leaf area to aboveground biomass.
-We found no effects of mean annual temperature or precipitation on biomass partitioning, suggesting that climate-related differences are captured by a shift in PFT.
+We found weak effects of mean annual temperature and no effects of precipitation on biomass partitioning, suggesting that climate-related differences are captured largely by a shift in PFT.
 These results point to a simple algorithm for biomass partitioning in the world’s forests.}
@@
 
-We found no PFT differences in root-shoot scaling (Fig.~\ref{fig:figureSI6}), and approximately isometric scaling (sma slope = xxx), which is consistent with other reports, \cite{hui_near_2014, cairns_root_1997}).
-Because of this simple scaling between above- and below-ground biomass pools, we further focus on biomass partioning between leaves and aboveground woody pools.
+We found some evidence for a higher root-shoot ratio ($M_R/M_T$) in deciduous angiosperms (Fig.~\ref{fig:figureSI2}), however differences were small and depended on plant height.
+We found near-isometric scaling ($M_R \propto H^0.93) (cf.
+\cite{hui_near_2014, cairns_root_1997}), leading to only small shifts in the root-shoot ratio with increasing height.
+We did not test for direct climate effects on $M_R/M_T$ due to limited data availability for root biomass.
 
@@
 We found large differences between PFTs in biomass partitioning between leaves and stems (Fig.~\ref{fig:figure2}, see also Fig.~\ref{fig:figureSI3}).
-Calculated least-square means of LMF and LAR (averages accounting for variation explained by plant height) showed large differences in mass-based allocation between PFTs (LMF), and small inconsistent differences in area-based allocation (LAR) (Fig.~\ref{fig:figure2}, see Fig.~\ref{fig:figureSI4} for raw data on leaf area ratio by PFT).
+Least-square means of LMF and LAR (averages accounting for variation explained by plant height) showed large differences in mass-based allocation between PFTs (LMF), and small inconsistent differences in area-based allocation (LAR) (Fig.~\ref{fig:figure2}, see Fig.~\ref{fig:figureSI4} for raw data on leaf area ratio by PFT).
 
@@
 We found that, across PFTs, $M_F/A_S$ was approximately proportional to the average specific leaf mass for the PFT.
-As a result of this approximate proportionality, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure4}) (see also Fig.~\ref{fig:figureSI6} for raw data).
-
-This relationship also held approximately across species (Fig.~\ref{fig:figureSI7}).
+As a result of this approximate proportionality, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure4}) (see also Fig.~\ref{fig:figureSI6}).
+This relationship also held approximately when data were first averaged by species (Fig.~\ref{fig:figureSI7}).
 
@@
 Unlike \cite{reich_temperature_2014}, we found only very weak effects of mean annual temperature the scaling of leaf biomass with woody biomass (Fig.~\ref{fig:figureSI9}).
-However, the variance explained by MAT was very small, compared to the large difference between PFTs.
+The variance explained by MAT was very small, especially compared to the large difference between PFTs.
 These results were confirmed with a mixed-effects model of MF vs.

### 2014-10-22 R/prepareDataset.R

@@ -65,6 +65,10 @@ dataset2$pftlong <- as.factor(with(dataset2, paste(pft, bortemptrop, sep='-')))
 
 
 
+# Root dataset, excluding three studies with very poor root estimates
+datroot <- subset(dataset2, !is.na(m.rt) & !is.na(m.so) &
+                    !studyName %in% c("Gargaglione2010","Rodriguez2003","Albrektson1984"))
+
 
 
 

### 2014-10-22 manuscript/draftpaper.Rnw
Used more data throughout (previously used dataset2, now dataset which also includes woodland, savanna). Bug fixes in manuscript, glitch fixes figures, SI5 better right panel X axis).

@@
 \usepackage{gensymb}
+%\usepackage{lineno}
 \geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
@@
 \begin{document}
+
+%\linenumbers
 
@@
 
+\section*{First paragraph}
+
 \textbf{The partitioning of biomass in terrestrial vegetation between leaves, stems and roots is a critical component of dynamic global vegetation models (DGVMs).
@@
 
-\section{flow}
-
-\begin{itemize}
-
-\item First paragraph
-\item Allocation in models, main uncertainties
-\item Climate effects on partioning : direct (MAT, MAP), indirect (PFT).
-\item Known: gymnosperm vs. angiosperm. Unknown: leaf habit (deciduous vs. evergreen), direct climate, patterns for individual plants
-\item BAAD : Largest database, other large datasets are stands or only small plants
-\item Root scaling no PFT differences, focus on aboveground
-\item LMF, LAR
-\item Found very large LMF differences, but convergent LAR
-\item decompose into MF/AS, MT/AS, find similar differences in MF/AS but none in MT/AS, suggesting convergent AF/AS between PFTs.
-\item No direct climate effects, so shift in PFT captures global patterns in biomass partioning
-\item This has implications for allocation in DGVMs, they can be divided into roughly three categories in terms of biomass allocation (allocation fractions, allometric constraints, LAI optimization).
+% \section{flow}
+% 
+% \begin{itemize}
+% 
+% \item First paragraph
+% \item Allocation in models, main uncertainties
+% \item Climate effects on partioning : direct (MAT, MAP), indirect (PFT).
+% \item Known: gymnosperm vs. angiosperm. Unknown: leaf habit (deciduous vs. evergreen), direct climate, patterns for individual plants
+% \item BAAD : Largest database, other large datasets are stands or only small plants
+% \item Root scaling no PFT differences, focus on aboveground
+% \item LMF, LAR
+% \item Found very large LMF differences, but convergent LAR
+% \item decompose into MF/AS, MT/AS, find similar differences in MF/AS but none in MT/AS, suggesting convergent AF/AS between PFTs.
+% \item No direct climate effects, so shift in PFT captures global patterns in biomass partioning
+% \item This has implications for allocation in DGVMs, they can be divided into roughly three categories in terms of biomass allocation (allocation fractions, allometric constraints, LAI optimization).
 Allocation vs.
 partitioning: the role of leaf and stem turnover.
+% 
+% \end{itemize}
 
-\end{itemize}
-
-\section{textbits}
+\section*{Main text}
 
@@
 We found some evidence for a higher root-shoot ratio ($M_R/M_T$) in deciduous angiosperms (Fig.~\ref{fig:figureSI2}), however differences were small and depended on plant height.
-We found near-isometric scaling ($M_R \propto H^0.93) (cf.
+We found near-isometric scaling ($M_R \propto H^{0.93}$) (cf.
 \cite{hui_near_2014, cairns_root_1997}), leading to only small shifts in the root-shoot ratio with increasing height.

### 2014-10-23 manuscript/draftpaper.Rnw

@@
 \usepackage{gensymb}
-%\usepackage{lineno}
+\usepackage{lineno}
 \geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
@@
 
-%\linenumbers
+\linenumbers
 
-\title{Specific leaf mass explains difference in biomass partitioning between woody plant functional types}
+\title{A global compilation reveals divergent biomass partitioning between plant functional types }
 
@@
 \textbf{The partitioning of biomass in terrestrial vegetation between leaves, stems and roots is a critical component of dynamic global vegetation models (DGVMs).
-Biomass partioning is consistently shown to be one of the most uncertain and important drivers for the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, and global predictions of carbon storage \cite{ise_comparison_2010}.
+Biomass partioning is consistently shown to be one of the most uncertain and important drivers for global predictions of carbon storage \cite{ise_comparison_2010}, the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, and the CO$_2$ response of net primary production \cite{dekauwe2014does} .
 Little is known about direct and indirect effects of climate on biomass partitioning, largely due to the lack of a global individual-based biomass dataset.
-Here, we analyze the largest database on plant biomass to date.
+Here, we analyze the largest database on plant biomass to date [cite].
 We show that partitioning of biomass to leaves vs.

### 2014-10-24 manuscript/draftpaper.Rnw
fig SI-8 updated

@@
 We found some evidence for a higher root-shoot ratio ($M_R/M_T$) in deciduous angiosperms (Fig.~\ref{fig:figureSI2}), however differences were small and depended on plant height.
-We found near-isometric scaling ($M_R \propto H^{0.93}$) (cf.
-\cite{hui_near_2014, cairns_root_1997}), leading to only small shifts in the root-shoot ratio with increasing height.
-We did not test for direct climate effects on $M_R/M_T$ due to limited data availability for root biomass.
+We found near-isometric scaling ($M_R \propto H^{0.93}$) \cite{hui_near_2014, cairns_root_1997, mokany_critical_2006}, leading to only small shifts in the root-shoot ratio with increasing height.
+We found no effects of climate on root-shoot ratio, consistent with a recent global comparison \cite{mokany_critical_2006}.
+Relation precip to root biomass is probably more complex \cite{schenk_rooting_2002}.
 

### 2014-10-24 R/functions-figures.R

@@ -18,11 +18,13 @@ mixmean <- function(yvar, g, dataset=dat){
 meansbypft <- function(yvar1, yvar2=NULL, pftvar, 
                            xvar="llma",
                            dataset,
+                           axis1=TRUE,
                            panel1.expr=NULL,
                            panel2.expr=NULL,
                            setpar=TRUE,
                            Cols=NULL,
                            panel1only=FALSE,
+                          addlegend=TRUE,
                            legend.text=NULL,
                            legend.where="topright",
                            legend.cex=0.7,
@@ -51,7 +53,7 @@ meansbypft <- function(yvar1, yvar2=NULL, pftvar,
   if(is.character(xvar)){
     X <- mixmean(xvar,pftvar,dat)
   } else {
-    X <- list(y = X, lci=rep(NA,3), uci=rep(NA,3))
+    X <- list(y = xvar, lci=rep(NA,length(xvar)), uci=rep(NA,length(xvar)))
   }
   
   # mix means of Y variables.
@@ -61,7 +63,6 @@ meansbypft <- function(yvar1, yvar2=NULL, pftvar,
   if(is.null(Cols))
     Cols <- rainbow(length(unique(dat$P)))
   
-
   # model fit for multiple comparisons
   f1 <- lmer(Y1 ~ P - 1 + (1|Group), data=dat)
   if(!panel1only)f2 <- lmer(Y2 ~ P - 1 + (1|Group), data=dat)
@@ -83,6 +84,7 @@ meansbypft <- function(yvar1, yvar2=NULL, pftvar,
                 y1=y1$uci,code=3,angle=90,length=0.025,col=Cols)
        })
   axis(2)
+  if(axis1)axis(1,labels=FALSE)
   box()
   
   u <- par()$usr
@@ -93,22 +95,26 @@ meansbypft <- function(yvar1, yvar2=NULL, pftvar,
     pointLabel(X$y, y1$y, lets1, cex=0.9)
   }
   
-  axis(1,labels=FALSE)
   if(!is.null(panel1.expr))eval(panel1.expr)
   
   if(is.null(legend.text))legend.text <- rownames(X)
-  legend(legend.where, legend.text,pch=19,col=Cols, 
+  if(addlegend){
+    legend(legend.where, legend.text,pch=19,col=Cols, 
          cex=legend.cex,pt.cex=1.2, bty='n')
+  }
   
[…]

### 2014-10-24 figures.R
fig SI-8 updated

@@ -137,7 +137,9 @@ to.pdf({
 to.pdf({
   par(cex.axis=0.85, mfrow=c(1,2), mar=c(5,5,1,1), cex=1.1)
   meansbypft("lmlf_astba2","lalf_astba2", "pft", 
-             xvar="llma",setpar=FALSE,
+             xvar="llma",
+             dataset=dataset, 
+             setpar=FALSE,
              legend.where="topleft",
              legend.cex=0.6,
              legend.text=c("Decid. Angio.","Evergr. Angio.","Evergr. Gymno."),
@@ -147,7 +149,6 @@ to.pdf({
              xlab=expression("Specific leaf mass"~~(kg~m^-2)),
              ylab2=expression(A[F]/A[S]~~(m^2~m^-2)),
              ylab1=expression(M[F]/A[S]~~(kg~m^-2)), 
-             dataset=dataset, 
              xlim=c(0,0.2),
              ylim1=c(0,250),ylim2=c(0,2000))
 }, width=8, height=4, filename="manuscript/figures/Figure4_mlf_alf_astbaest_pftmeans.pdf")
@@ -334,23 +335,23 @@ to.pdf({
 # SI 8
 # Means of leaf mass, area per stem area by PFT - biome combination.
 to.pdf({
-  par(cex.axis=0.85, mfrow=c(1,2), mar=c(5,5,1,1), cex=1.1)
+  Labs <- c("Temp. Decid. Angio.","Temp. Evergr. Angio.",
+    "Trop. Evergr. Angio.","Boreal Evergr. Gymno.",
+    "Temp. Evergr. Gymno.")
+  par(cex.axis=0.85, mfrow=c(1,2), mar=c(8,5,1,1), cex=1.1, las=2, cex.axis=0.9)
   meansbypft("lmlf_astba2","lalf_astba2", "pftlong", 
-             xvar="llma",setpar=FALSE,
-             legend.where="bottomright",
-             legend.cex=0.6,
-             legend.text=c("Temp. Decid. Angio.","Temp. Evergr. Angio.",
-                           "Trop. Evergr. Angio.","Boreal Evergr. Gymno.",
-                           "Temp. Evergr. Gymno."),
-             panel1.expr={axis(1);axis(2)},
-             panel2.expr={axis(1);axis(2)},
-             Cols=rainbow(5),
-             siglets="symbol",
-             xlab=expression("Specific leaf mass"~~(kg~m^-2)),
[…]

### 2014-10-26 manuscript/draftpaper.Rnw
Specific leaf mass now called leaf mass per area; tidied up various bits; analysis of MAP MAT R2 saved to manuscript/tables; removed cfg CSV file

@@
 Evergreen gymnosperms maintain three times more leaf biomass at a given total aboveground biomass than deciduous angiosperms, with evergreen angiosperms taking an intermediate value.
-We found that specific leaf mass, an indicator of leaf construction cost, explains these differences so that the PFTs converged in the ratio  of leaf area to aboveground biomass.
+We found that leaf mass per area, an indicator of leaf construction cost, explains these differences so that the PFTs converged in the ratio  of leaf area to aboveground biomass.
 We found weak effects of mean annual temperature and no effects of precipitation on biomass partitioning, suggesting that climate-related differences are captured largely by a shift in PFT.
@@
 
-Major plant functional types differ widely in specific leaf mass (SLM, the ratio of leaf mass to area, an index of leaf thickness) (\cite{poorter_causes_2009}), which can be expected to affect biomass partitioning in several ways.
+Major plant functional types differ widely in leaf mass per area (SLM, the ratio of leaf mass to area, an index of leaf thickness) (\cite{poorter_causes_2009}), which can be expected to affect biomass partitioning in several ways.
 First, SLM correlates with leaf lifespan.
@@
 
-We then expressed the average $M_F/A_S$ as a function of the specific leaf mass (SLM = $M_F / A_F$), a measure of the leaf construction cost to the plant.
-We found that, across PFTs, $M_F/A_S$ was approximately proportional to the average specific leaf mass for the PFT.
+We then expressed the average $M_F/A_S$ as a function of the leaf mass per area (SLM = $M_F / A_F$), a measure of the leaf construction cost to the plant.
+We found that, across PFTs, $M_F/A_S$ was approximately proportional to the average leaf mass per area for the PFT.
 As a result of this approximate proportionality, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure4}) (see also Fig.~\ref{fig:figureSI6}).

### 2014-10-26 R/obs_barplot.R
Specific leaf mass now called leaf mass per area; tidied up various bits; analysis of MAP MAT R2 saved to manuscript/tables; removed cfg CSV file

@@ -1,17 +1,19 @@
 
+# 1. Barplot with number of observations by variable.
+# 2. Barplot with number of observations by PFT, vegetation and growingCondition.
+# Nice for presentations etc.
 source("load.R")
 library(Hmisc)
 fn <- function(x)sum(!is.na(baad[,x]))
 
-cfg <- read.csv("data/variableDefinitions.csv", stringsAsFactors=FALSE)
-
-
-Vars <- cfg$Variable[13:47]
-Desc <- capitalize(cfg$label[13:47])
+# Choose variables, get names and labels.
+thesevars <- 15:50
+Vars <- cfg$variable[thesevars]
+Desc <- capitalize(cfg$label[thesevars])
 names(Vars) <- Desc
 Vars <- Vars[!Vars %in% c("d.bh","h.c","d.ba")]
 vn <- sort(sapply(Vars,fn),F)
-vn <- vn[vn >0]
+vn <- vn[vn >0] # exclude vars with 0 observations
 
 to.pdf({
   par(mar=c(1,13,3,1), las=1, mgp=c(2,0.5,0), cex=0.8, tcl=0.4)

### 2014-10-29 R/functions-figures.R
Major update mixmean, incorporates significance letters, and meansbypft (slightly more generic and clean). Major update FigS8 (all pft-biome combos included when data present; caption update).

@@ -1,19 +1,31 @@
 
 
 #' Means by some grouping variable g, accounting for random effect R.
-mixmean <- function(yvar, g, R="Group", dataset){
+mixmean <- function(yvar, g, data, R="Group"){
   
-  dataset$P <- dataset[,g]
-  dataset$Y <- dataset[,yvar]
-  dataset$R <- dataset[,R]
+  data$P <- data[,g]
+  data$Y <- data[,yvar]
+  data$R <- data[,R]
   
-  f <- lmer(Y ~ P - 1 + (1|R), data=dataset)
+  f <- lmer(Y ~ P - 1 + (1|R), data=data)
   
   ci <- suppressMessages(confint(f)[-c(1,2),])
   ci <- as.data.frame(cbind(10^fixef(f),10^ci))
-  rownames(ci) <- gsub("P","", rownames(ci))
   names(ci) <- c("y","lci","uci")
-  return(ci)
+  ci[,g] <- gsub("P","", rownames(ci))
+  rownames(ci) <- NULL
+  
+  df <- data.frame(unique(data$P))
+  names(df) <- g
+  df <- merge(df, ci, all=TRUE)
+  
+  lets <- as.data.frame(cld(glht(f, linfct=mcp(P="Tukey")))$mcletters$Letters)
+  names(lets) <- "signifletters"
+  lets[,g] <- rownames(lets)
+  rownames(lets) <- NULL
+  df <- merge(df, lets, all=TRUE)
+
+  return(df)
 }
 
 
@@ -38,13 +50,7 @@ meansbypft <- function(yvar1, yvar2=NULL, pftvar,
                            ylab1=NULL, ylab2=NULL, 
                            ylim1=NULL, ylim2=NULL){
     
-  if(pftvar == "pftlong"){
-    dat <- droplevels(subset(dataset, pftlong %in% 
-                               c("DA-temperate","EA-temperate","EA-tropical","EG-boreal",
-                                 "EG-temperate")))
-  } else {
-    dat <- dataset
-  }
+  dat <- dataset
   
   siglets <- match.arg(siglets)
   if(siglets == "symbol")library(maptools)
@@ -67,13 +73,9 @@ meansbypft <- function(yvar1, yvar2=NULL, pftvar,
   if(is.null(Cols))
     Cols <- rainbow(length(unique(dat$P)))
   
[…]

### 2014-10-29 manuscript/draftpaper.Rnw
one ref
Major update mixmean, incorporates significance letters, and meansbypft (slightly more generic and clean). Major update FigS8 (all pft-biome combos included when data present; caption update).

@@
 
-Major plant functional types differ widely in leaf mass per area (SLM, the ratio of leaf mass to area, an index of leaf thickness) (\cite{poorter_causes_2009}), which can be expected to affect biomass partitioning in several ways.
-First, SLM correlates with leaf lifespan.
+Major plant functional types differ widely in leaf mass per area (LMA, the ratio of leaf mass to area, an index of leaf thickness) (\cite{poorter_causes_2009}), which can be expected to affect biomass partitioning in several ways.
+First, LMA correlates with leaf lifespan.
 Longer lived leaves would lead to higher plant leaf mass ($M_F$) as long as annual production rates are similar (\cite{enquist_global_2002}).
@@
 
-It has been reported numerous times that gymnosperms (with higher SLM) have higher leaf mass fraction than angiosperms (with lower SLM) \cite{enquist_global_2002, cheng_relationships_2007, mccarthy_consistency_2007, poorter_biomass_2012, sardans_tree_2013}, although a recent study reported no difference (\cite{wolf_allometric_2010}).
+It has been reported numerous times that gymnosperms (with higher LMA) have higher leaf mass fraction than angiosperms (with lower LMA) \cite{enquist_global_2002, cheng_relationships_2007, mccarthy_consistency_2007, poorter_biomass_2012, sardans_tree_2013}, although a recent study reported no difference (\cite{wolf_allometric_2010}).
 No study to date has been able to further tease apart PFT differences, or test the potentially confounding effects of climate.
@@
 
-We then expressed the average $M_F/A_S$ as a function of the leaf mass per area (SLM = $M_F / A_F$), a measure of the leaf construction cost to the plant.
+We then expressed the average $M_F/A_S$ as a function of the leaf mass per area (LMA = $M_F / A_F$), a measure of the leaf construction cost to the plant.
 We found that, across PFTs, $M_F/A_S$ was approximately proportional to the average leaf mass per area for the PFT.
@@
 We found that plant height and biome explained only \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[2],2)}\%, but when PFT was added to the model, the variance explained increased to \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[3],2)}\%.
-If we used SLM as a predictor instead of PFT, the $R^2$ was similar (\Sexpr{100*signif(Table_pipemodel_varpart$Marginal[4],2)}\%), showing that PFT affected $M_F/A_S$ largely through differences in SLM.
+If we used LMA as a predictor instead of PFT, the $R^2$ was similar (\Sexpr{100*signif(Table_pipemodel_varpart$Marginal[4],2)}\%), showing that PFT affected $M_F/A_S$ largely through differences in LMA.
 Consistent with this, PFT, plant height and biome explained very little variance in $A_F/A_S$ (Table~\ref{varpartpipe}).
@@
 When added to a model that included plant height and biome, PFT increased the variance explained in log-transformed LMF from \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[2],2)} to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[3],2)}\%.
-However, SLM was not a good substitute for PFT, as the explained variance decreased to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[4],2)}\%.
-For (log-transformed) LAR, PFT and SLM explained very little additional variance.
+However, LMA was not a good substitute for PFT, as the explained variance decreased to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[4],2)}\%.
+For (log-transformed) LAR, PFT and LMA explained very little additional variance.
 
@@
 
+More plasticity in LMA than biomass partitioning has been demonstrated \cite{poorter_biomass_2012}.
+
 In summary, we found coordination between an important leaf trait and whole-plant biomass partitioning. This has implications for global vegetation models.
@@
 Either constant allocation fraction, target allometric equations, or optimization \cite{de2014does}.
-Constant allocation fraction gives constant LAR by pft if leaf lifespan is exactly proportional to SLM.
+Constant allocation fraction gives constant LAR by pft if leaf lifespan is exactly proportional to LMA.
 Target allometric equations can be made pft-invariant by using allometry of leaf area, not leaf mass.
@@
 For the aboveground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total aboveground biomass) and leaf area ratio ($A_F / M_T$).
-LMF, LAR, SLM, $M_F$ and $A_F$ are related by,
+LMF, LAR, LMA, $M_F$ and $A_F$ are related by,
 
-$$LMF = M_F / M_T = (M_F / A_F) (A_F / M_T) = SLM * LAR$$
+$$LMF = M_F / M_T = (M_F / A_F) (A_F / M_T) = LMA * LAR$$
 
 We were unable to further decompose $A_S/M_W$ into wood volume and wood density components (cf. \cite{chave_tree_2005}) because wood density data were available only for a small subset of data.
-We did not include SLM estimates from another database (such as TRY), instead only used SLA directly estimated for the harvested plants (typically for a subsample of leaves).
+We did not include LMA estimates from another database (such as TRY), instead only used SLA directly estimated for the harvested plants (typically for a subsample of leaves).
 For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception  \cite{lang_application_1991, chen_defining_1992}, for comparison to flat leaves.
@@
 See Table~\\ref{varpartpipe} for more explanation.
-For plant height, we also added the squared term as a fixed effect, and the interaction with PFT or SLM, because the relationship was non-linear."),
+For plant height, we also added the squared term as a fixed effect, and the interaction with PFT or LMA, because the relationship was non-linear."),
       include.rownames=FALSE, caption.placement="top")
@@
     \includegraphics[width=0.99\textwidth]{FigureSI-8_mlf_alf_astbaest_pftlongmeans.pdf}
-    \caption{Like Fig.~\ref{fig:figure2}, but for five combinations of major plant functional type and biome.
-Out of 9 possible biome / pft combinations, only five in the dataset had leaf area estimates (and two combo's are not in the dataset, e.g.
-tropical evergreen gymno).
-Note that within pft, vegetation is never significant (this result will also be in another figure!).}
+    \caption{Mean leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/A_S$) by PFT and major biome.
+No leaf area estimates were available for boreal deciduous angiosperms.
+DA = deciduous angiosperm, EA = evergreen angiosperm, EG = evergreen gymnosperm, Trop.
+= tropical, Temp.
+= temperate.
+Error bars are 95\% confidence intervals for the mean, estimated with a linear mixed-effects model using species within study as the random effect.
+Letters denote significant differences (at $\alpha = 0.05$)}
     \label{fig:figureSI8}

### 2014-10-31 manuscript/draftpaper.Rnw
Added supporting information on behaviour of simple two-pool allocation model.

@@
 \usepackage{graphicx}
+\usepackage[T1]{fontenc}
 \usepackage[default,osfigures,scale=0.95]{opensans}
@@
 
+\subsection*{A simple biomass allocation model}
+
+We have shown that biomass partitioning to leaf biomass decreases quickly with increasing plant size.
+However, this does not yield a model for the allocation of net primary production (NPP) to leaf and stem components, since partitioning depends on allocation as well as turnover.
+Here we construct a minimal model of biomass allocation to evaluate the fixed allocation model (a model that allocates a fixed proportion of NPP to leaves and stems, not accounting for possible shifts with plant size).
+We consider two pools: a leaf pool with high turnover, and an aboveground woody pool with slow turnover.
+Belowground allocation is not considered here, it is implicitly assumed that the allocation to belowground is a fixed proportion of NPP.
+
+Dynamics of leaf mass ($M_F$) and aboveground woody biomass ($M_S$) depend on annual allocation of NPP to these pools ($\phi_F$ and $phi_S$), and turnover rates ($yr^{-1}$) ($r_F$ and $r_S$), as in:
+
+\begin{equation}\label{eq:dmfdt}
+\frac{dM_F}{dt} = \phi_F NPP - r_F M_F
+\end{equation}
+
+\begin{equation}\label{eq:dmsdt}
+\frac{dM_S}{dt} = \phi_S NPP - r_S M_S
+\end{equation}
+With the constraint that $\phi_F + \phi_S = 1$. We further assume that NPP is proportional to foliage biomass,
+
+\begin{equation}
+NPP = a_F M_F
+\end{equation}
+
+So that Eqs.~\ref{eq:dmfdt} and~\ref{eq:dmsdt} become,
+\begin{equation}
+\frac{dM_F}{dt} = (\phi_F  a_F - r_F) M_F
+\end{equation}
+
+\begin{equation}
+\frac{dM_S}{dt} = (1 - \phi_F)  a_F  M_F - r_S M_S
+\end{equation}
+
+These two differential equations may be integrated (using the method of variation of parameters) to yield the solution,
+
+\begin{equation}\label{eq:mfsolution}
+M_F(t) = c_2 e^{t(\phi_F a_F - r_F)}
+\end{equation}
+
+and
+
+\begin{equation}\label{eq:mssolution}
+M_S(t) = \frac{\phi_S a_F c_2}{\phi_F a_F - r_F} \frac{1}{\frac{r_S}{\phi_F a_F - r_F} + 1} e^{t(\phi_F a_F - r_F)}
++ c_4 e^{-r_S t}
+\end{equation}
+
+where $c_2 = M_f(0)$ and
+
+$$
+c_4 = M_S(0) - \frac{\phi_S a_F c_2}{\phi_F a_F - r_F} \frac{1}{\frac{r_S}{\phi_F a_F - r_F} + 1}
+$$
+
+We are here interested in the ratio $M_F/M_S$ and its dependence on plant size.
+We compared simulations using Eqs.~\ref{eq:mfsolution} and~\ref{eq:mssolution} to the average relationship estimated from our database (see Fig.~\ref{fig:figureSI10}).
+It is clear from these simulations that $M_F/M_S$ quickly reaches a steady state value, and shows very little decline with increasing plant size.
+In contrast, the data show a very steep (note the log-log axes) and persistent decline in $M_F/M_S$.
+These results indicate that fixed allocation model can not adequately be used to model $M_F/M_S$, some of the parameters must be a function of plant size (most likely, the allocation to foliage mass should decline with size).
+
+Finally, it can easily be shown that $M_F/M_S$ converges to the following equation (found by setting $t = \infty$ in Eqs.~\ref{eq:mfsolution} and~\ref{eq:mssolution}, and taking the ratio $M_F/M_S$).
+To simplify the result, we have also assumed that $r_S = 0$ (because we can assume that $r_S << r_F$).
+
+\begin{equation}
+\lim_{t\to \infty} \frac{M_F}{M_S} = \frac{\phi_F - r_F/a_F}{1 - \phi_F|}
+\end{equation}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureSI-10_allocationmodel.pdf}
+    \caption{caption text}
+  \label{fig:figureSI10}
+\end{figure}
+
 \clearpage

### 2014-10-31 manuscript/draftpaper.Rnw
minor glitch fix

@@
 \usepackage{graphicx}
-\usepackage[T1]{fontenc}
+% \usepackage[T1]{fontenc}
 \usepackage[default,osfigures,scale=0.95]{opensans}
@@
 We found weak effects of mean annual temperature and no effects of precipitation on biomass partitioning, suggesting that climate-related differences are captured largely by a shift in PFT.
-These results point to a simple algorithm for biomass partitioning in the world’s forests.}
+These results point to a simple algorithm for biomass partitioning in the world's forests.}
 
@@
 
-Dynamics of leaf mass ($M_F$) and aboveground woody biomass ($M_S$) depend on annual allocation of NPP to these pools ($\phi_F$ and $phi_S$), and turnover rates ($yr^{-1}$) ($r_F$ and $r_S$), as in:
+Dynamics of leaf mass ($M_F$) and aboveground woody biomass ($M_S$) depend on annual allocation of NPP to these pools ($\phi_F$ and $\phi_S$), and turnover rates ($yr^{-1}$) ($r_F$ and $r_S$), as in:
 
@@
 \begin{equation}
-\lim_{t\to \infty} \frac{M_F}{M_S} = \frac{\phi_F - r_F/a_F}{1 - \phi_F|}
+\lim_{t\to \infty} \frac{M_F}{M_S} = \frac{\phi_F - r_F/a_F}{1 - \phi_F}
 \end{equation}

### 2014-11-16 manuscript/draftpaper.Rnw
Removed old climate analysis, added new GAM-based general one. Added as new Table 1.

@@
 load("tables/Tables_varpart.RData")
+load("tables/gamr2MIgdd0.RData")
 baad <- readRDS("../data/baad.rds")
@@
 We found that leaf mass per area, an indicator of leaf construction cost, explains these differences so that the PFTs converged in the ratio  of leaf area to aboveground biomass.
-We found weak effects of mean annual temperature and no effects of precipitation on biomass partitioning, suggesting that climate-related differences are captured largely by a shift in PFT.
-These results point to a simple algorithm for biomass partitioning in the world's forests.}
+We found weak effects of climate while accounting for PFT effects, suggesting that climate-related differences are captured largely by a shift in PFT.
+These results can be used to constrain biomass partitioning routines in global dynamic vegetation models.}
 
@@
 <<results='asis'>>=
-print(xtable(Table_pipemodel_varpart, label="varpartpipe", 
-             caption="Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.
-The marginal R2 is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type.
-The conditional R2 is that explained by fixed and random effects (species within study).
-All fixed effects were highly significant (P \\textless 0.001).
-In addition to the fixed effects shown, all interactions were added to each of the models."),
-      include.rownames=FALSE, caption.placement="top")
-@
-
-<<results='asis'>>=
-load("tables/Tables_varpart.RData")
-print(xtable(Table_LMFLAR_varpart, label="varpartlmflar",
-      caption="Explained variance in the leaf mass fraction (leaf mass / aboveground biomass) and leaf area ratio (leaf area / aboveground biomass).
-See Table~\\ref{varpartpipe} for more explanation.
-For plant height, we also added the squared term as a fixed effect, and the interaction with PFT or LMA, because the relationship was non-linear."),
+print(xtable(g0, label="table/gamr2table",
+      caption="Explained variance by plant functional type (PFT), plant height (H), moisture index (MI) and growing season average temperature (mgdd0), for six whole-plant variables."),
       include.rownames=FALSE, caption.placement="top")
@@
 
-<<tablepreload2>>=
-library(car)
-load("tables/Fits_lme_mlfmst_MAPMAT.RData")
-printanova <- function(x, caption, label=NULL){
-  rownames(x) <- gsub("log10(m.st)", "MS", rownames(x), fixed=TRUE) 
-  rownames(x) <- gsub("pft", "PFT", rownames(x)) 
-  print(xtable(x, caption=caption, label=label),
-                                        caption.placement="top")
-}
+<<results='asis'>>=
+print(xtable(Table_pipemodel_varpart, label="varpartpipe", 
+             caption="Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.
+The marginal R2 is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type.
+The conditional R2 is that explained by fixed and random effects (species within study).
+All fixed effects were highly significant (P \\textless 0.001).
+In addition to the fixed effects shown, all interactions were added to each of the models."),
+      include.rownames=FALSE, caption.placement="top")
 @
@@
 <<results='asis'>>=
-printanova(Anova(mlfmst_lme1), 
-           label="table:matlmeanova",
-           caption="Type-II Anova table for a linear mixed-effects model of $M_F$ as a function of $M_S$ (both log-transformed), PFT and mean annual temperature (MAT).
-Species by dataset was used as the random effect.
-Further analysis of the interactions revealed that MAT only significantly affected the relationship between $M_F$ and $M_S$ for the deciduous angiosperms (not shown).")
-@
-
-<<results='asis'>>=
-printanova(Anova(mlfmst_lme2), 
-           label="table:maplmeanova",
-           caption="Type-II Anova table for a linear mixed-effects model of $M_F$ as a function of $M_S$ (both log-transformed), PFT and mean annual precipitation (MAT).
-Species by dataset was used as the random effect.
-We did not analyze MAT and MAP together in one model, as they were strongly correlated, making interpretation of the fitted model difficult.")
+load("tables/Tables_varpart.RData")
+print(xtable(Table_LMFLAR_varpart, label="varpartlmflar",
+      caption="Explained variance in the leaf mass fraction (leaf mass / aboveground biomass) and leaf area ratio (leaf area / aboveground biomass).
+See Table~\\ref{varpartpipe} for more explanation.
+For plant height, we also added the squared term as a fixed effect, and the interaction with PFT or LMA, because the relationship was non-linear."),
+      include.rownames=FALSE, caption.placement="top")
 @

### 2014-11-16 R/climateeffects.R

@@ -1,7 +1,13 @@
 
 
+source("load.R")
+source("R/preparedataset.R")
+source("R/rsquaredglmm.R")
+library(visreg)
 
+#----------------------------------------------------------------------------------#
 
+# lmer with H, H^2, PFT, MAP, MAT
 
 testmapmat <- function(yvar){
   
@@ -31,6 +37,8 @@ l <- lapply(vars,fx)
 tab <- cbind(as.data.frame(vars), as.data.frame(do.call(rbind,l)))
 names(tab) <- c("Variable","H,PFT,MAT,MAP","H,PFT","H,MAT,MAP","H")
 
+#-----------------------------------------------------------------------------------#
+# Apparent effect for gymnosperms, but this is likely spurious.
 
 lme1 <- testmapmat("lalf_astba2")
 visreg(lme1[[1]], "MAT", by="pft")
@@ -49,32 +57,17 @@ r.squared(lme_eg2)
 r.squared(lme_eg3)
 r.squared(lme_eg4)
 
+# Partial regression plot, actually very weak effect of MAP,
+# overpredicted by lme above.
 dateg$lalf_astba2_p <- predict(lme_eg, dateg, re.form=NA)
 dateg$lalf_astba2_p2 <- predict(lme_eg2, dateg, re.form=NA)
 
 with(dateg, plot(MAP, lalf_astba2 - lalf_astba2_p, pch=19, col=matbin))
 abline(h=0)
 
-palette(rainbow(11))
-
-
-
-with(dateg, plot(MAT, lalf_astba2 - lalf_astba2_p, pch=19, col=matbin))
-with(dateg, plot(MAT, lalf_astba2 - lalf_astba2_p2, pch=19, col=matbin))
-
-ii <- with(dateg, identify(MAT, lalf_astba2 - lalf_astba2_p))
-
-
-dateg2 <- droplevels(subset(dataset, pft =="EG" & studyName != "Reid2004"))
-lme_egb <- lmer(lalf_astba2 ~ h.t*I(h.t^2) + (1|Group), data=dateg2)
-lme_egb2 <- lmer(lalf_astba2 ~ h.t*I(h.t^2)*MAT*MAP + (1|Group), data=dateg2)
-r.squared(lme_egb)
-r.squared(lme_egb2)
-
 
 #------------------------------------------------------------------------------------#
-# explained variance with gam
-
+# Explained variance with gam (H, PFT, MAT, MAP)
 
[…]

### 2014-11-16 statanalysis.R

@@ -20,36 +20,6 @@ smroot <- sma(m.rt ~ m.so*pft, data=datroot, log="xy")
 
 
 #---------------------------------------------------------------------------------------#
-# Variance partitioning to fixed and random effects 
-
-
-# Sequential R2, backwards elimination of all fixed effects.
-# ((not used at the moment))
-r2stepwise <- function(m, data){
-  
-  tr <- attributes(terms(m))$term.labels
-  ii <- lapply(1:length(tr), function(x)1:x)
-  yvar <- rownames(attributes(terms(m))$factors)[1]
-  
-  fitM <- function(i, dat){
-    tr <- c(tr[i], "(1|Group)")
-    f <- as.formula(paste(yvar, "~", paste(tr, collapse=" + ")))
-    M <- lmer(formula=f, data=dat)
-  return(M)
-  }
-  
-  models <- lapply(ii, fitM, dat=data)
-  rsq <- do.call(rbind, lapply(models, r.squared))
-rownames(rsq) <- tr
-return(rsq)
-}
-
-plotr2 <- function(x, which=c("Marginal","Conditional"),...){
-
-  which <- match.arg(which)
-  par(mar=c(4,11,2,2))
-  barplot(rev(x[[which]]), names.arg=rev(rownames(x)), horiz=T, las=2,...)
-}
 
 # Datasets with NAs removed, of key variables.
 dat_mlf <- droplevels(subset(dataset2, !is.na(h.t) & !is.na(pft) & !is.na(lmlf_astba2)))
@@ -119,7 +89,7 @@ makem <- function(..., variable=NULL){
   m <- do.call(rbind, lapply(l,s))  
   m <- as.data.frame(m)
   if(!is.null(variable))m <- cbind(data.frame(Variable=c(variable,NA,NA,NA), 
-                                              Predictors=c("H", "H, B", "PFT, H, B", "SLM, H, B")),
+                                              Predictors=c("H", "H, B", "PFT, H, B", "LMA, H, B")),
                                    m)
                                               
   
@@ -192,38 +162,56 @@ save(d, mlfmst_lme0,mlfmst_lme1,mlfmst_lme2,
      file="manuscript/tables/Fits_lme_mlfmst_MAPMAT.RData")
 
 
[…]

### 2014-11-18 manuscript/draftpaper.Rnw
added some author instructions, changed title
New figure SI-9, using MI and mgdd0.

@@
 
-\title{A global compilation reveals divergent biomass partitioning between plant functional types }
+\title{Indirect climate effects on biomass partitioning}
 
@@
 
+% First paragraph The first paragraph should be 180 to 200 words and should contain references.
+The paragraph should serve both as a general introduction to the topic and as a brief, non-technical summary of your main results and their implications.
+Ideally, the paragraph starts with two to three sentences of background explaining the topic and current understanding.
+This is followed by a phrase like “Here we present” to introduce the new work.
+In the second part, the main methods, findings, conclusions and implications should be described clearly and accessibly.
+
 \section*{First paragraph}
@@
 % \item Climate effects on partioning : direct (MAT, MAP), indirect (PFT).
-% \item Known: gymnosperm vs. angiosperm. Unknown: leaf habit (deciduous vs. evergreen), direct climate, patterns for individual plants
+% \item Known: gymnosperm vs. angiosperm. Unknown: leaf habit (deciduous vs. evergreen), direct climate, patterns for individual plants.
 % \item BAAD : Largest database, other large datasets are stands or only small plants
-% \item Root scaling no PFT differences, focus on aboveground
-% \item LMF, LAR
+% \item Root scaling no PFT and very weak climate differences, focus on aboveground
 % \item Found very large LMF differences, but convergent LAR
 % \item decompose into MF/AS, MT/AS, find similar differences in MF/AS but none in MT/AS, suggesting convergent AF/AS between PFTs.
-% \item No direct climate effects, so shift in PFT captures global patterns in biomass partioning
+% \item Only small direct climate effects, so shift in PFT captures global patterns in biomass partioning.
 % \item This has implications for allocation in DGVMs, they can be divided into roughly three categories in terms of biomass allocation (allocation fractions, allometric constraints, LAI optimization).
@@
 
-Unlike \cite{reich_temperature_2014}, we found only very weak effects of mean annual temperature the scaling of leaf biomass with woody biomass (Fig.~\ref{fig:figureSI9}).
+Unlike \cite{reich_temperature_2014}, we found only very weak effects of climate on biomass relationships.
 The variance explained by MAT was very small, especially compared to the large difference between PFTs.
@@
 \section{Figures}
+
+% Figures A Nature Climate Change Letter can contain up to four display items (any combination of figures and/or tables).
+Figures should be sized in single or double column format.
+Multi-panel figures should be arranged so that the figure forms a neat rectangle.
 
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureSI-9_MAT_LMFscaling.pdf}
-    \caption{For each species within study, we fit the scaling relationship of leaf mass as a function of woody biomass assuming an exponent of 3/4, using standardized major axis.
-The resulting coefficient (b0) is size-invariant (and thus superior to the simple ratio of leaf mass to woody biomass, which is strongly size-dependent, see Fig.~\ref{fig:figure3}).
-We found no effects of mean annual temperature on the coefficient.
-The triangles are from Roth et al.
-(2007), a study on Pinus eliottii with exceptionally high leaf mass per unit woody biomass.}
+    \includegraphics[width=0.99\textwidth]{FigureSI-9_climateeffects.pdf}
+    \caption{Modest effects of climate on three size-invariant variables, compared to the very large effect of plant functional type on mass-based variables (leaf mass per area, LMA, and $M_F/A_S$), and convergence in area-based variables ($A_F/A_S$).
+See SI Methods for calculation of MI and growing season temperature.
+Solid lines are generalized additive model fits (with a base dimension of 3, larger values resulted in biologically implausible relationships).
+Grey areas are approximate 95\% confidence intervals, calculated by the gam after averaging each variable by species/dataset combination (but all raw individual plant data are shown in the plots).}
   \label{fig:figureSI9}

### 2014-11-18 figures.R

@@ -17,6 +17,7 @@ palette(Cols)
 # Plot settings.
 lmaLabel <- expression("Leaf mass per area"~~(kg~m^-2))
 
+
 Legend <- function(where, labels=c("short","long"), bty='n', ...){
   lab <- if(match.arg(labels) == "short")
             c("Decid. Angio.", "Evergr. Angio.", "Evergr. Gymno.")
@@ -409,48 +410,36 @@ to.pdf({
 
 
 # SI 9
-# cf. Reich et al 2014. Does MAT influence LMF?
-# Here, I 'correct' for size by estimating b0 in MF = b0*MS^(3/4).
-# This is estimated with sma for each Group. MAT is averaged within Group,
-# this takes several locations per Group (not that many though!)
-# Roth2007 is highlighted.
-dat <- studyWithVars(dataset, c("m.lf","m.st","MAT", "MAP"))
-sm1 <- sma(m.lf ~ m.st*Group, data=dat, log="xy", slope.test=3/4, quiet=TRUE)
+# Three size-invariant variables as a function of MI and mgdd0
+figure_SI9 <- function(K=3){
+  milab <- "Moisture Index (-)"
+  mgdlab <- expression(Growing~season~T~(degree*C))
+  par(mfrow=c(3,2), cex.lab=1.2, mar=c(5,5,1,1))
+  smoothplot(MI, llma, pft, dataset, log="y", kgam=K, R="Group", randommethod = "agg",
+             xlab=milab,
+             ylim=c(-2,0), ylab=expression(LMA~~(kg~m^-2)))
+  Legend("topright")
+  smoothplot(mgdd0, llma, pft, dataset, log="y", kgam=K, R="Group", randommethod = "agg",
+             xlab=mgdlab,
+             ylim=c(-2,0), ylab=expression(LMA~~(kg~m^-2)))
+  smoothplot(MI, lmlf_astba2, pft, dataset, log="y", kgam=K, R="Group", randommethod = "agg",
+             xlab=milab,
+             ylab=expression(M[F]/A[S]~~(kg~m^-2)))
+  smoothplot(mgdd0, lmlf_astba2, pft, dataset, log="y", kgam=K, R="Group", randommethod = "agg",
+             xlab=mgdlab,
+             ylab=expression(M[F]/A[S]~~(kg~m^-2)))
[…]

### 2014-11-18 manuscript/draftpaper.Rnw
Removed simple allocation model; should be part of follow-up paper

@@
 
-\subsection*{A simple biomass allocation model}
-
-We have shown that biomass partitioning to leaf biomass decreases quickly with increasing plant size.
-However, this does not yield a model for the allocation of net primary production (NPP) to leaf and stem components, since partitioning depends on allocation as well as turnover.
-Here we construct a minimal model of biomass allocation to evaluate the fixed allocation model (a model that allocates a fixed proportion of NPP to leaves and stems, not accounting for possible shifts with plant size).
-We consider two pools: a leaf pool with high turnover, and an aboveground woody pool with slow turnover.
-Belowground allocation is not considered here, it is implicitly assumed that the allocation to belowground is a fixed proportion of NPP.
-
-Dynamics of leaf mass ($M_F$) and aboveground woody biomass ($M_S$) depend on annual allocation of NPP to these pools ($\phi_F$ and $\phi_S$), and turnover rates ($yr^{-1}$) ($r_F$ and $r_S$), as in:
-
-\begin{equation}\label{eq:dmfdt}
-\frac{dM_F}{dt} = \phi_F NPP - r_F M_F
-\end{equation}
-
-\begin{equation}\label{eq:dmsdt}
-\frac{dM_S}{dt} = \phi_S NPP - r_S M_S
-\end{equation}
-With the constraint that $\phi_F + \phi_S = 1$. We further assume that NPP is proportional to foliage biomass,
-
-\begin{equation}
-NPP = a_F M_F
-\end{equation}
-
-So that Eqs.~\ref{eq:dmfdt} and~\ref{eq:dmsdt} become,
-\begin{equation}
-\frac{dM_F}{dt} = (\phi_F  a_F - r_F) M_F
-\end{equation}
-
-\begin{equation}
-\frac{dM_S}{dt} = (1 - \phi_F)  a_F  M_F - r_S M_S
-\end{equation}
-
-These two differential equations may be integrated (using the method of variation of parameters) to yield the solution,
-
-\begin{equation}\label{eq:mfsolution}
-M_F(t) = c_2 e^{t(\phi_F a_F - r_F)}
-\end{equation}
-
-and
-
-\begin{equation}\label{eq:mssolution}
-M_S(t) = \frac{\phi_S a_F c_2}{\phi_F a_F - r_F} \frac{1}{\frac{r_S}{\phi_F a_F - r_F} + 1} e^{t(\phi_F a_F - r_F)}
-+ c_4 e^{-r_S t}
-\end{equation}
-
-where $c_2 = M_f(0)$ and
-
-$$
-c_4 = M_S(0) - \frac{\phi_S a_F c_2}{\phi_F a_F - r_F} \frac{1}{\frac{r_S}{\phi_F a_F - r_F} + 1}
-$$
-
-We are here interested in the ratio $M_F/M_S$ and its dependence on plant size.
-We compared simulations using Eqs.~\ref{eq:mfsolution} and~\ref{eq:mssolution} to the average relationship estimated from our database (see Fig.~\ref{fig:figureSI10}).
-It is clear from these simulations that $M_F/M_S$ quickly reaches a steady state value, and shows very little decline with increasing plant size.
-In contrast, the data show a very steep (note the log-log axes) and persistent decline in $M_F/M_S$.
-These results indicate that fixed allocation model can not adequately be used to model $M_F/M_S$, some of the parameters must be a function of plant size (most likely, the allocation to foliage mass should decline with size).
-
-Finally, it can easily be shown that $M_F/M_S$ converges to the following equation (found by setting $t = \infty$ in Eqs.~\ref{eq:mfsolution} and~\ref{eq:mssolution}, and taking the ratio $M_F/M_S$).
-To simplify the result, we have also assumed that $r_S = 0$ (because we can assume that $r_S << r_F$).
-
-\begin{equation}
-\lim_{t\to \infty} \frac{M_F}{M_S} = \frac{\phi_F - r_F/a_F}{1 - \phi_F}
-\end{equation}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureSI-10_allocationmodel.pdf}
-    \caption{caption text}
-  \label{fig:figureSI10}
-\end{figure}
-
 \clearpage

### 2014-11-19 manuscript/draftpaper.Rnw
Climate space figure now removes areas without woody vegetation

@@
 
+Moisture index and growing season temperature were taken from Lin et al. (in review).
+
 \section{Tables}
@@
     \caption{Global coverage of the climate space by the dataset.
-Hexagons are terrestrial observations extracted from Worldclim (colored by number of 0.5\degree~cells, cut off at 6000mm MAP).
-Note that Worldclim observations include land areas where woody vegetation does not occur (e.g., desert, tundra).
-Triangles are averages for the three PFTs across all contributed studies.
+Hexagons are terrestrial observations extracted from Worldclim \cite{worldclim} (colored by number of 0.5\degree~cells, cut off at 6000mm MAP), excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \cite{lathamglobal}) Triangles are averages for the three PFTs across all contributed studies.
 }

### 2014-11-20 manuscript/draftpaper.Rnw
Changed order of figures, minor changes to (new) Fig 4 and (new) Fig 1
Fixed Figure SI2 (colors, removed glasshouse studies)

@@
     \centering
-    \includegraphics{Figure1_MAPMAT_baad_vs_worldclim.pdf}
-    \caption{Global coverage of the climate space by the dataset.
-Hexagons are terrestrial observations extracted from Worldclim \cite{worldclim} (colored by number of 0.5\degree~cells, cut off at 6000mm MAP), excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \cite{lathamglobal}) Triangles are averages for the three PFTs across all contributed studies.
-}
-    \label{fig:figure1}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
     \includegraphics[width=0.99\textwidth]{Figure2_LMF_lines_lsmeans_3panel.pdf}
@@
 
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureSI-9_climateeffects.pdf}
+    \caption{Modest effects of climate on three size-invariant variables, compared to the very large effect of plant functional type on mass-based variables (leaf mass per area, LMA, and $M_F/A_S$), and convergence in area-based variables ($A_F/A_S$).
+See SI Methods for calculation of MI and growing season temperature.
+Solid lines are generalized additive model fits (with a base dimension of 3, larger values resulted in biologically implausible relationships).
+Grey areas are approximate 95\% confidence intervals, calculated by the gam after averaging each variable by species/dataset combination (but all raw individual plant data are shown in the plots).}
+  \label{fig:figureSI9}
+\end{figure}
+
 \clearpage
@@
     \centering
+    \includegraphics{Figure1_MAPMAT_baad_vs_worldclim.pdf}
+    \caption{Global coverage of the climate space by the dataset.
+Hexagons are terrestrial observations extracted from Worldclim \cite{worldclim} (colored by number of 0.5\degree~cells, cut off at 6000mm MAP), excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \cite{lathamglobal}) Triangles are averages for the three PFTs across all contributed studies.
+}
+    \label{fig:figure1}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
     \includegraphics[width=0.99\textwidth]{FigureSI-1_MAPMAT_vegetation.pdf}
-    \caption{Mean annual precipitation and temperature for all contributed datasets, by vegetation type. Note that in this study, we did not include glasshouse studies.}
+    \caption{Mean annual precipitation (MAP) and temperature (MAT) for all contributed datasets, by vegetation type.
+MAP and MAT were taken from Worldclim \cite{worldclim}, using latitude and longitude reported for each contributed dataset.}
     \label{fig:figureSI1}
@@
     \label{fig:figureSI8}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureSI-9_climateeffects.pdf}
-    \caption{Modest effects of climate on three size-invariant variables, compared to the very large effect of plant functional type on mass-based variables (leaf mass per area, LMA, and $M_F/A_S$), and convergence in area-based variables ($A_F/A_S$).
-See SI Methods for calculation of MI and growing season temperature.
-Solid lines are generalized additive model fits (with a base dimension of 3, larger values resulted in biologically implausible relationships).
-Grey areas are approximate 95\% confidence intervals, calculated by the gam after averaging each variable by species/dataset combination (but all raw individual plant data are shown in the plots).}
-  \label{fig:figureSI9}
 \end{figure}

### 2014-11-20 R/prepareDataset.R

@@ -79,7 +79,7 @@ make_baadmapmat <- function(){
   mapmat <- baad[!duplicated(baad[,c("MAP","MAT")]),]
   mapmat$vegetation <- as.factor(mapmat$vegetation)
   mapmat$pft <- as.factor(mapmat$pft)
-  mapmat <- droplevels(subset(mapmat, pft != "DG"))
+  mapmat <- droplevels(subset(mapmat, pft != "DG" & growingCondition != "GH"))
   
   return(mapmat)
 }

### 2014-11-20 manuscript/draftpaper.Rnw
Methods update

@@
 The main objective was to test whether PFT-specific differences exist in the ratio of leaf mass to total above-ground biomass (the 'leaf mass fraction', LMF), and the leaf area to aboveground biomass (the 'leaf area ratio', LAR).
-Because PFT and climate are confounded to some extent (Fig.~\ref{figure1}), we also test whether climate has additional effects on biomass partitioning, or whether these effects are captured by PFT alone.
+Because climate and PFT are correlated, (Fig.~\ref{fig:figure1}), we also test whether climate has additional effects on biomass partitioning, or whether these effects are captured by PFT alone.
 
@@
 
-We can further decompose LMF into two components representing different aspects of plant function and structure.
+We can further decompose LMF into two components representing different aspects of plant function and structure (see Methods).
 The ratio of leaf biomass ($M_F$) to stem basal area ($A_S$) relates to hydraulic supply, and the ratio of total aboveground biomass ($M_T$) and $A_S$ is a function of plant shape (stem taper, plant height at given stem diameter), and wood density \cite{chave_tree_2005}.
-
-$$LMF = M_F / M_T = (M_F / A_S) (M_T / A_S)^{-1}$$
-
 We found that ratio of aboveground biomass to basal stem area ($M_T/A_S$) did not differ by PFT (Fig.~\ref{fig:figureSI5}).
@@
 The variance explained by MAT was very small, especially compared to the large difference between PFTs.
-These results were confirmed with a mixed-effects model of MF vs.
-MS, with MAT and PFT as covariates (see Table~\ref{table:matlmeanova}).
-We found no significant effects of mean annual precipitation of leaf-woody biomass relationships (Table~\ref{table:maplmeanova}).
 
-More plasticity in LMA than biomass partitioning has been demonstrated \cite{poorter_biomass_2012}.
+Leaf mass per area generally shows a more plastic response to climate than biomass partitioning  \cite{poorter_biomass_2012}.
 
@@
 
-A recent study compared allocation routines in a number of leading ecosystem models \cite{de2014does}, and recommended to constrain allocation by observed biomass fractions instead of using constant allocation fractions, which is common practice.
+A recent study compared allocation routines in a number of leading ecosystem models \cite{dekauwe2014}, and recommended to constrain allocation by observed biomass fractions instead of using constant allocation fractions, which is common practice.
 As a first approximation, which is useful for large scale DGVMs, models should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly.
@@
 We found no effects of MAP on biomass partitioning, and only very weak effects of MAT on LMF, when analyzed in addition to PFT.
-This means that climate-related differences are captured by a shift in PFT (see also Fig.~\ref{figure1}, showing PFTs occupy different parts of the climate space).
+This means that climate-related differences are captured by a shift in PFT (see also Fig.~\ref{fig:figure1}, showing PFTs occupy different parts of the climate space).
 Our results contrast with those of \cite{reich_temperature_2014} but they studied whole-stand biomass, not individual plant biomass.
@@
 1.3m) ($A_{Sbh}$).
-For each analysis, we therefore used different subsets of the data (far fewer observations are available for root biomass, in particular).
-Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4} We exluded glasshouse and common garden studies, including only field grown woody plants (including plantations).
-We considered three plant functional types : evergreen angiosperms, evergreen gymnosperms, and decidous angiosperms.
+For each analysis, we used different subsets of the data.
+Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4} We exluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
+We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and decidous angiosperms.
 We excluded  deciduous gymnosperms because few data were available.
-All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes (decided by contributing author, but checked with MAT and latitude).
+All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes.
 
-For the aboveground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total aboveground biomass) and leaf area ratio ($A_F / M_T$).
-LMF, LAR, LMA, $M_F$ and $A_F$ are related by,
+For the aboveground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total aboveground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,
 
@@
 
-We were unable to further decompose $A_S/M_W$ into wood volume and wood density components (cf. \cite{chave_tree_2005}) because wood density data were available only for a small subset of data.
-We did not include LMA estimates from another database (such as TRY), instead only used SLA directly estimated for the harvested plants (typically for a subsample of leaves).
-For conifers, leaf area was converted to half-total surface area, which is most appropriate in terms of light interception  \cite{lang_application_1991, chen_defining_1992}, for comparison to flat leaves.
+We further decomposed LMF and LAR by considering the ratio of leaf mass (or area) to plant basal area, and the ratio of aboveground biomass to plant basal area. For LMF,
+
+$$LMF = M_F / M_T = (M_F / A_S) (M_T / A_S)^{-1}$$
+
+And similarly for LAR, by replacing $M_F$ with $A_F$.
+We were unable to further decompose $A_S/M_T$ into wood volume and wood density components (cf.
+\cite{chave_tree_2005}) because wood density data were available only for a small subset of data.
+
+We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves).
+For conifers, leaf area was converted to half-total surface area, which is most appropriate for comparison to flat leaves in terms of light interception  \cite{lang_application_1991, chen_defining_1992}.
 
@@
 1.3m) and/or at the base of the plant.
-For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using non-linear regression (see~SIXXX).
+For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using non-linear regression (see~\ref{sec:predas}).
 The main results were similar when we repeated it for data where either $A_S$ or $A_{Sbh}$ were directly measured (not shown).
 
-We used a generalized additive model (GAM) to visualize the relationships, but these were not used for formal inference.
-Instead we used linear mixed-effects models (with species within study as the random effect) to test for significance and to calculate confidence intervals for PFT-specific averages.
-Least-square means for LMF and LAR were calculated with the 'lsmeans' package in R.
+We used a generalized additive model (GAM) to visualize the relationships between biomass and plant size variables.
+In addition, we corrected for plant size by fitting linear mixed-effects models (with species within study as the random effect) using plant height (linear and squared terms), and calculating least-square means with the 'lsmeans' package in R.
 All variables were log-transformed prior to analysis.
@@
 
-Moisture index and growing season temperature were taken from Lin et al. (in review).
+To test for direct climate effects in addition to the PFT effect, we used moisture index and growing season temperature as estimated by Lin et al. (in review).
 
@@
 Error bars are 95\% confidence interval for the mean (calculated with a mixed-effects model with species within study as the random effect).
-Letters denote significant differences.
+Letters denote significant differences (at $\alpha = 0.05$).
 }
@@
     \includegraphics[width=0.99\textwidth]{FigureSI-2_mrt_mso_bypft.pdf}
-    \caption{Near-isometric scaling of above-ground with below-ground biomass, and no effects of plant functional type. The black line is a 1-1 line.}
+    \caption{Relationship between total root biomass ($M_R$) and aboveground biomass ($M_T$) separated by PFT.
+The right panel shows least-square means of the root shoot ratio ($M_R/M_T$).
+Letters denote significant differences (at $\alpha = 0.05$).}
     \label{fig:figureSI2}
@@
     \includegraphics[width=0.99\textwidth]{FigureSI-3_mlfmst_byht_pft.pdf}
-    \caption{Leaf and woody biomass scaling with plant height.}
+    \caption{Raw data for leaf biomass and total above-ground biomass for each of the PFTs. Each point is an individual plant.}
     \label{fig:figureSI3}
@@
     \includegraphics[width=0.99\textwidth]{FigureSI-4_LAR_pft_lines.pdf}
-    \caption{Leaf area ratio (leaf area / aboveground biomass) by plant functional type. Each symbol is an individual plant. Lines are generalized additive model fits.}
+    \caption{Leaf area ratio ($A_F/M_T$) by PFT.
+Each symbol is an individual plant.
+Lines are generalized additive model fits using species/dataset as a random effect.
+Shaded areas are approximate 95\% confidence intervals.}
     \label{fig:figureSI4}
@@
 = temperate.
-Error bars are 95\% confidence intervals for the mean, estimated with a linear mixed-effects model using species within study as the random effect.
+Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect.
 Letters denote significant differences (at $\alpha = 0.05$)}
@@
         caption="Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$.
-Individual plants (number of species).
+Numbers denote individual plants, with number of species in parentheses.
 DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@@
         label="table:count2",
-        caption="Number of observations used in the analysis of Af/As. Individual plants (number of species). DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
+        caption="Number of observations used in the analysis of $A_F/A_S$.
+Numbers denote individual plants, with number of species in parentheses.
+DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
 @
@@
         label="table:count3",
-        caption="Number of observations used in the analysis of leaf mass fraction (LMF) and leaf area ratio (LAR).
-Individual plants (number of species).
+        caption="Number of observations used in the analysis of leaf mass fraction ($M_F/M_T$) and leaf area ratio ($A_F/M_T$).
+Numbers denote individual plants, with number of species in parentheses.
 DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@@
         caption="Number of observations used in the analysis of leaf area ratio (LAR).
-Individual plants (number of species).
+Numbers denote individual plants, with number of species in parentheses.
 DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@@
 \subsection*{Predicting basal diameter}
+\label{sec:predas}
 

### 2014-12-01 manuscript/draftpaper.Rnw
added sections
proper textsubscript

@@
 \usepackage{graphicx}
+\usepackage{fixltx2e}
 % \usepackage[T1]{fontenc}
@@
 \textbf{The partitioning of biomass in terrestrial vegetation between leaves, stems and roots is a critical component of dynamic global vegetation models (DGVMs).
-Biomass partioning is consistently shown to be one of the most uncertain and important drivers for global predictions of carbon storage \cite{ise_comparison_2010}, the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, and the CO$_2$ response of net primary production \cite{dekauwe2014does} .
+Biomass partioning is consistently shown to be one of the most uncertain and important drivers for global predictions of carbon storage \cite{ise_comparison_2010}, the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014does} .
 Little is known about direct and indirect effects of climate on biomass partitioning, largely due to the lack of a global individual-based biomass dataset.
@@
 To test for direct climate effects in addition to the PFT effect, we used moisture index and growing season temperature as estimated by Lin et al. (in review).
+
+\section{Acknowledgments}
+
+\section{Author contributions}
 

### 2014-12-01 manuscript/draftpaper.Rnw
Added some code to replace PDFs with jpegs, for (much) quicker printing

@@
 \textbf{The partitioning of biomass in terrestrial vegetation between leaves, stems and roots is a critical component of dynamic global vegetation models (DGVMs).
-Biomass partioning is consistently shown to be one of the most uncertain and important drivers for global predictions of carbon storage \cite{ise_comparison_2010}, the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014does} .
+Biomass partioning is consistently shown to be one of the most uncertain and important drivers for global predictions of carbon storage \cite{ise_comparison_2010}, the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
 Little is known about direct and indirect effects of climate on biomass partitioning, largely due to the lack of a global individual-based biomass dataset.
@@
 How is allocation implemented in current DGVMs?
-Either constant allocation fraction, target allometric equations, or optimization \cite{de2014does}.
+Either constant allocation fraction, target allometric equations, or optimization \cite{dekauwe2014}.
 Constant allocation fraction gives constant LAR by pft if leaf lifespan is exactly proportional to LMA.

### 2014-12-01 R/prepare_for_print.R

@@ -1,0 +1,26 @@
+
+
+# Make all figures into JPG
+o <- getwd()
+setwd("manuscript/figures")
+pdfs <- dir(pattern="[.]pdf")
+
+for(i in seq_along(pdfs)){
+  
+  fn <- gsub("[.]pdf",".jpg",pdfs[i])
+  shell(paste("convert -density 600",pdfs[i],fn))
+  message(i)
+}
+
+
+setwd(o)
+
+# use .jpg in manuscript
+setwd("manuscript")
+shell("sed -i 's/.pdf/.jpg/g' draftpaper.Rnw")
+setwd(o)
+
+# Now compile the Rnw and print!
+
+# and go back,
+shell("sed -i 's/.jpg/.pdf/g' manuscript/draftpaper.Rnw")

### 2014-12-08 manuscript/draftpaper.Rnw
Reshuffling and renaming of figures (again), removed roots

--- 2014-11-20 manuscript/draftpaper.Rnw
@@
 
-\title{Indirect climate effects on biomass partitioning}
+\title{Plant functional type effects on biomass partitioning}
 
@@
 
-\textbf{The partitioning of biomass in terrestrial vegetation between leaves, stems and roots is a critical component of dynamic global vegetation models (DGVMs).
+\textbf{The partitioning of biomass in terrestrial vegetation between short-lived leaves and long-lived stems is a critical component of dynamic global vegetation models (DGVMs).
 Biomass partioning is consistently shown to be one of the most uncertain and important drivers for global predictions of carbon storage \cite{ise_comparison_2010}, the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, and the CO$_2$ response of net primary production \cite{dekauwe2014does} .
@@
 % \item BAAD : Largest database, other large datasets are stands or only small plants
-% \item Root scaling no PFT and very weak climate differences, focus on aboveground
 % \item Found very large LMF differences, but convergent LAR
@@
 
-We found some evidence for a higher root-shoot ratio ($M_R/M_T$) in deciduous angiosperms (Fig.~\ref{fig:figureSI2}), however differences were small and depended on plant height.
-We found near-isometric scaling ($M_R \propto H^{0.93}$) \cite{hui_near_2014, cairns_root_1997, mokany_critical_2006}, leading to only small shifts in the root-shoot ratio with increasing height.
-We found no effects of climate on root-shoot ratio, consistent with a recent global comparison \cite{mokany_critical_2006}.
-Relation precip to root biomass is probably more complex \cite{schenk_rooting_2002}.
-
 The main objective was to test whether PFT-specific differences exist in the ratio of leaf mass to total above-ground biomass (the 'leaf mass fraction', LMF), and the leaf area to aboveground biomass (the 'leaf area ratio', LAR).
@@
 
-We found large differences between PFTs in biomass partitioning between leaves and stems (Fig.~\ref{fig:figure2}, see also Fig.~\ref{fig:figureSI3}).
-Least-square means of LMF and LAR (averages accounting for variation explained by plant height) showed large differences in mass-based allocation between PFTs (LMF), and small inconsistent differences in area-based allocation (LAR) (Fig.~\ref{fig:figure2}, see Fig.~\ref{fig:figureSI4} for raw data on leaf area ratio by PFT).
+We found large differences between PFTs in biomass partitioning between leaves and stems (Fig.~\ref{fig:figure1}b, see also Fig.~\ref{fig:figureS2}).
+After accounting for the strong effects of plant height, we found large differences in LMF between PFTs (Fig.~\ref{fig:figure1}c), and small inconsistent differences in LAR (Fig.~\ref{fig:figure1}d, see also Fig.~\ref{fig:figureS3}).
+This suggests that PFTs do not differ in the amount of leaf area displayed per unit aboveground biomass.
+Since the LMA differs strongly, this implies a difference in the biomass partitioning between leaves and stems.
 
@@
 The ratio of leaf biomass ($M_F$) to stem basal area ($A_S$) relates to hydraulic supply, and the ratio of total aboveground biomass ($M_T$) and $A_S$ is a function of plant shape (stem taper, plant height at given stem diameter), and wood density \cite{chave_tree_2005}.
-We found that ratio of aboveground biomass to basal stem area ($M_T/A_S$) did not differ by PFT (Fig.~\ref{fig:figureSI5}).
-However we found very large differences between PFTs in the ratio $M_F/A_S$(Fig.~\ref{fig:figure3}).
+We found that ratio of aboveground biomass to basal stem area ($M_T/A_S$) was nearly constant between PFTs (Fig.~\ref{fig:figureS4}), with deciduous angiosperms having slightly higher $M_T$ at a given $A_S$.
+However we found very large differences between PFTs in the ratio $M_F/A_S$(Fig.~\ref{fig:figure2}).
 
@@
 We found that, across PFTs, $M_F/A_S$ was approximately proportional to the average leaf mass per area for the PFT.
-As a result of this approximate proportionality, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure4}) (see also Fig.~\ref{fig:figureSI6}).
-This relationship also held approximately when data were first averaged by species (Fig.~\ref{fig:figureSI7}).
+As a result of this approximate proportionality, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure2}) (see also Fig.~\ref{fig:figureS5}).
+This relationship also held approximately when data were first averaged by species (Fig.~\ref{fig:figureS6}).
 
-Because there was substantial variation between individual plants (Figs.~\ref{fig:figure3}), we estimated the variance explained by PFT, plant height, and biome in $M_F/A_S$ using a linear mixed-effects model (see Table~\ref{varpartpipe}).
+Although PFTs converged in the amount of leaf area displayed per unit aboveground biomass and stem basal area, there was a substantial amount of variation between plants (Fig.~\ref{fig:figure3}).
+We partitioned the variance into PFT, plant height, and biome effects in $M_F/A_S$ using a linear mixed-effects model (see Table~\ref{varpartpipe}).
 We found that plant height and biome explained only \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[2],2)}\%, but when PFT was added to the model, the variance explained increased to \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[3],2)}\%.
@@
 Pair-wise comparisons between biomes within PFT mostly did not show significant differences for $M_F/A_S$ or $A_F/A_S$.
-The only exception was a higher $A_F/A_S$ for tropical compared to temperate evergreen angiosperms (Fig.~\ref{fig:figureSI8}).
+The only exception was a higher $A_F/A_S$ for tropical compared to temperate evergreen angiosperms (Fig.~\ref{fig:figureS7}).
 
@@
 
-We used the recently compiled Biomass And Allometry Database (BAAD) (Falster et al., in review), including a total of \Sexpr{nrow(baad$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), aboveground woody biomass ($M_W$), total root biomass ($M_R$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
+We used the recently compiled Biomass And Allometry Database (BAAD) (Falster et al., in review), including a total of \Sexpr{nrow(baad$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), aboveground woody biomass ($M_W$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
 1.3m) ($A_{Sbh}$).
@@
 
+\section{Acknowledgments}
+
+\section{Author contributions}
+
 \section{Tables}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{Figure2_LMF_lines_lsmeans_3panel.pdf}
-    \caption{Leaf mass fraction (leaf mass / aboveground biomass) by plant functional type.
+    \includegraphics[width=0.99\textwidth]{Figure1.pdf}
+    \caption{(a) Global coverage of the climate space by the dataset.
+Hexagons are terrestrial observations extracted from Worldclim \cite{worldclim} (colored by number of 0.5\degree~cells, cut off at 6000mm MAP), excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \cite{lathamglobal}) Triangles are averages for the three PFTs across all contributed studies.
+(b) Leaf mass fraction (leaf mass / aboveground biomass) by plant functional type.
 Each symbol is an individual plant.
-Lines are generalized additive model fits.Bottom panels show average leaf mass fraction (LMF) and leaf area ratio (LAR) accounting for variation explained by plant height (least-square means) with a mixed effects model (with species within study as a random effect).
-Error bars are 95\% confidence intervals.}
+Lines are generalized additive model fits.
+(c) and (d) Leaf mass fraction (LMF) and leaf area ratio (LAR) at the average plant height in the dataset, estimated from panel (b).
+Error bars are 95\% confidence intervals.
+Letters denote significant differences.}
+    \label{fig:figure1}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
+    \caption{Average leaf mass per unit basal stem area (panel 1), and leaf area per unit basal stem area (panel 2) for the three major plant functional types.
+Error bars are 95\% confidence interval for the mean (calculated with a mixed-effects model with species within study as the random effect).
+Letters denote significant differences (at $\alpha = 0.05$).
+}
     \label{fig:figure2}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{Figure3_hist_alfast_mlfast.pdf}
+    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
     \caption{Plant functional types diverge strongly in mass-based allocation, but converge in area-based allocation.
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{Figure4_mlf_alf_astbaest_pftmeans.pdf}
-    \caption{Average leaf mass per unit basal stem area (panel 1), and leaf area per unit basal stem area (panel 2) for the three major plant functional types.
-Error bars are 95\% confidence interval for the mean (calculated with a mixed-effects model with species within study as the random effect).
-Letters denote significant differences (at $\alpha = 0.05$).
-}
-    \label{fig:figure4}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureSI-9_climateeffects.pdf}
+    \includegraphics[width=0.99\textwidth]{Figure4.pdf}
     \caption{Modest effects of climate on three size-invariant variables, compared to the very large effect of plant functional type on mass-based variables (leaf mass per area, LMA, and $M_F/A_S$), and convergence in area-based variables ($A_F/A_S$).
@@
 Grey areas are approximate 95\% confidence intervals, calculated by the gam after averaging each variable by species/dataset combination (but all raw individual plant data are shown in the plots).}
-  \label{fig:figureSI9}
+  \label{fig:figure4}
 \end{figure}
@@
     \centering
-    \includegraphics{Figure1_MAPMAT_baad_vs_worldclim.pdf}
-    \caption{Global coverage of the climate space by the dataset.
-Hexagons are terrestrial observations extracted from Worldclim \cite{worldclim} (colored by number of 0.5\degree~cells, cut off at 6000mm MAP), excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \cite{lathamglobal}) Triangles are averages for the three PFTs across all contributed studies.
-}
-    \label{fig:figure1}
+    \includegraphics[width=0.99\textwidth]{FigureS1.pdf}
+    \caption{Mean annual precipitation (MAP) and temperature (MAT) for all contributed datasets, by vegetation type.
+MAP and MAT were taken from Worldclim \cite{worldclim}, using latitude and longitude reported for each contributed dataset.}
+    \label{fig:figureS1}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureSI-1_MAPMAT_vegetation.pdf}
-    \caption{Mean annual precipitation (MAP) and temperature (MAT) for all contributed datasets, by vegetation type.
-MAP and MAT were taken from Worldclim \cite{worldclim}, using latitude and longitude reported for each contributed dataset.}
-    \label{fig:figureSI1}
+    \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
+    \caption{Raw data for leaf biomass and total above-ground biomass for each of the PFTs. Each point is an individual plant.}
+    \label{fig:figureS2}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureSI-2_mrt_mso_bypft.pdf}
-    \caption{Relationship between total root biomass ($M_R$) and aboveground biomass ($M_T$) separated by PFT.
-The right panel shows least-square means of the root shoot ratio ($M_R/M_T$).
-Letters denote significant differences (at $\alpha = 0.05$).}
-    \label{fig:figureSI2}
+    \includegraphics[width=0.99\textwidth]{FigureS3.pdf}
+    \caption{Leaf area ratio ($A_F/M_T$) by PFT.
+Each symbol is an individual plant.
+Lines are generalized additive model fits using species/dataset as a random effect.
+Shaded areas are approximate 95\% confidence intervals.}
+    \label{fig:figureS3}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureSI-3_mlfmst_byht_pft.pdf}
-    \caption{Raw data for leaf biomass and total above-ground biomass for each of the PFTs. Each point is an individual plant.}
-    \label{fig:figureSI3}
+    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
+    \caption{Relationship between aboveground biomass and basal stem area, each point is an individual plant.
+Lines are spline fits.
+The right panel shows least-square means of the ratio of aboveground biomass to basal stem area, estimated from a linear mixed-effects model with basal stem area (linear and squared terms) and PFT.
+}
+    \label{fig:figureS4}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureSI-4_LAR_pft_lines.pdf}
-    \caption{Leaf area ratio ($A_F/M_T$) by PFT.
+    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
+    \caption{Plant leaf mass as a function of basal stem area (left panel), and leaf area as a function of basal stem area (right panel).
 Each symbol is an individual plant.
-Lines are generalized additive model fits using species/dataset as a random effect.
-Shaded areas are approximate 95\% confidence intervals.}
-    \label{fig:figureSI4}
+Lines are generalized additive model fits.}
+  \label{fig:figureS5}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureSI-5_mso_astba2_twopanel.pdf}
-    \caption{Relationship between aboveground biomass and basal stem area, each point is an individual plant.
-Lines are spline fits.
-The right panel shows least-square means of the ratio of aboveground biomass to basal stem area, estimated from a linear mixed-effects model with basal stem area (linear and squared terms) and PFT.
-}
-    \label{fig:figureSI5}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureSI-6_mlf_alf_astba2_bypft.pdf}
-    \caption{Plant leaf mass as a function of basal stem area (left panel), and leaf area as a function of basal stem area (right panel).
-Each symbol is an individual plant.
-Lines are generalized additive model fits.}
-  \label{fig:figureSI6}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureSI-7_mlf_alf_ast_byspecies.pdf}
+    \includegraphics[width=0.99\textwidth]{FigureS6.pdf}
     \caption{Plant leaf mass as a function of basal stem area (left panel), and leaf area as a function of basal stem area (right panel).
@@
 The coefficient of variation between species was 0.19 for log-transformed MF/AS, and only 0.099 for AF/AS.}
-  \label{fig:figureSI7}
+  \label{fig:figureS6}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureSI-8_mlf_alf_astbaest_pftlongmeans.pdf}
+    \includegraphics[width=0.99\textwidth]{FigureS7.pdf}
     \caption{Mean leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/A_S$) by PFT and major biome.
@@
 Letters denote significant differences (at $\alpha = 0.05$)}
-    \label{fig:figureSI8}
+    \label{fig:figureS7}
 \end{figure}
@@
 
-The estimated coefficients were $c_0$ = \Sexpr{coef(basalafit)[1]}, $c_1$ = \Sexpr{coef(basalafit)[2]}, root-mean square error = \Sexpr{summary(basalafit)$sigma}.
+The estimated coefficients were $c_0$ = \Sexpr{signif(coef(basalafit)[1],3)}, $c_1$ = \Sexpr{signif(coef(basalafit)[2],3)}, root-mean square error = \Sexpr{signif(summary(basalafit)$sigma,3)}.
 

### 2014-12-08 manuscript.Rnw
Move to maker, ruthless cull & restructure
Tried to reduce to only materials currently being used
Use maker workflow instead of Makefile

--- 2014-12-08 manuscript/draftpaper.Rnw
@@
 \usepackage{lineno}
-\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
+\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
 \bibliographystyle{plain}
@@
 
-\author[1]{Remko Duursma}
-\author[2]{Daniel Falster}
+\author[1]{Remko A. Duursma}
+\author[2]{Daniel S. Falster}
 \author[3]{BAAD contributors}
@@
 
-load("tables/Tables_varpart.RData")
-load("tables/gamr2MIgdd0.RData")
-baad <- readRDS("../data/baad.rds")
 @
@@
 
-\textbf{The partitioning of biomass in terrestrial vegetation between short-lived leaves and long-lived stems is a critical component of dynamic global vegetation models (DGVMs).
-Biomass partioning is consistently shown to be one of the most uncertain and important drivers for global predictions of carbon storage \cite{ise_comparison_2010}, the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, and the CO$_2$ response of net primary production \cite{dekauwe2014does} .
+\textbf{The partitioning of biomass in terrestrial vegetation between leaves, stems and roots is a critical component of dynamic global vegetation models (DGVMs).
+Biomass partioning is consistently shown to be one of the most uncertain and important drivers for global predictions of carbon storage \cite{ise_comparison_2010}, the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
 Little is known about direct and indirect effects of climate on biomass partitioning, largely due to the lack of a global individual-based biomass dataset.
@@
 We partitioned the variance into PFT, plant height, and biome effects in $M_F/A_S$ using a linear mixed-effects model (see Table~\ref{varpartpipe}).
-We found that plant height and biome explained only \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[2],2)}\%, but when PFT was added to the model, the variance explained increased to \Sexpr{100*signif(Table_pipemodel_varpart$Marginal[3],2)}\%.
-If we used LMA as a predictor instead of PFT, the $R^2$ was similar (\Sexpr{100*signif(Table_pipemodel_varpart$Marginal[4],2)}\%), showing that PFT affected $M_F/A_S$ largely through differences in LMA.
+We found that plant height and biome explained only \Sexpr{100*signif(table_pipemodel_varpart$r2[2],2)}\%, but when PFT was added to the model, the variance explained increased to \Sexpr{100*signif(table_pipemodel_varpart$r2[3],2)}\%.
+If we used LMA as a predictor instead of PFT, the $R^2$ was similar (\Sexpr{100*signif(table_pipemodel_varpart$r2[4],2)}\%), showing that PFT affected $M_F/A_S$ largely through differences in LMA.
 Consistent with this, PFT, plant height and biome explained very little variance in $A_F/A_S$ (Table~\ref{varpartpipe}).
@@
 We also partionined the variance of LMF and LAR (Table~\ref{varpartpipe}).
-When added to a model that included plant height and biome, PFT increased the variance explained in log-transformed LMF from \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[2],2)} to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[3],2)}\%.
-However, LMA was not a good substitute for PFT, as the explained variance decreased to \Sexpr{100*signif(Table_LMFLAR_varpart$Marginal[4],2)}\%.
+When added to a model that included plant height and biome, PFT increased the variance explained in log-transformed LMF from \Sexpr{100*signif(table_LMFLAR_varpart$r2[2],2)} to \Sexpr{100*signif(table_LMFLAR_varpart$r2[3],2)}\%.
+However, LMA was not a good substitute for PFT, as the explained variance decreased to \Sexpr{100*signif(table_LMFLAR_varpart$r2[4],2)}\%.
 For (log-transformed) LAR, PFT and LMA explained very little additional variance.
@@
 
-We used the recently compiled Biomass And Allometry Database (BAAD) (Falster et al., in review), including a total of \Sexpr{nrow(baad$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), aboveground woody biomass ($M_W$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
+We used the recently compiled Biomass And Allometry Database (BAAD) (Falster et al., in review), including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), aboveground woody biomass ($M_W$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
 1.3m) ($A_{Sbh}$).
@@
 
-To test for direct climate effects in addition to the PFT effect, we used moisture index and growing season temperature as estimated by Lin et al. (in review).
+To test for direct climate effects in addition to the PFT effect, we used moisture index (rainfall / potential evapotranspiration) and growing season temperature (average air temperature during the period of the year when air temperature is larger than 0) as estimated by Lin et al.
+(in review).
+To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \cite{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \cite{lathamglobal}).
 
@@
 \section{Author contributions}
+
+R.A.D and D.S.F. conceived the ideas and methods, and co-wrote the paper. R.A.D. analysed data. BAAD contributors shared data.
 
@@
 <<results='asis'>>=
-print(xtable(g0, label="table/gamr2table",
-      caption="Explained variance by plant functional type (PFT), plant height (H), moisture index (MI) and growing season average temperature (mgdd0), for six whole-plant variables."),
-      include.rownames=FALSE, caption.placement="top")
+print(xtable(table_gamr2MIgdd0, label="table/gamr2table",
+      caption="Explained variance by plant functional type (PFT), plant height (H), moisture index (MI) and growing season average temperature (mgdd0), for six whole-plant variables.
+Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."), caption.placement="top")
 @
@@
     \caption{(a) Global coverage of the climate space by the dataset.
-Hexagons are terrestrial observations extracted from Worldclim \cite{worldclim} (colored by number of 0.5\degree~cells, cut off at 6000mm MAP), excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \cite{lathamglobal}) Triangles are averages for the three PFTs across all contributed studies.
-(b) Leaf mass fraction (leaf mass / aboveground biomass) by plant functional type.
+Hexagons are terrestrial observations, excluding areas without woody vegetation, colored by number of 0.5\degree~cells, cut off at 6000mm MAP.
+Triangles are averages for the three PFTs across all contributed studies.
+(b) Leaf mass fraction (LMF = leaf mass / aboveground biomass) by PFT.
 Each symbol is an individual plant.
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
-    \caption{Average leaf mass per unit basal stem area (panel 1), and leaf area per unit basal stem area (panel 2) for the three major plant functional types.
+    \includegraphics[width=0.5\textwidth]{Figure2.pdf}
+    \caption{Average leaf mass per unit basal stem area (panel 1), and leaf area per unit basal stem area (panel 2) for the three major PFTs.
 Error bars are 95\% confidence interval for the mean (calculated with a mixed-effects model with species within study as the random effect).
@@
     \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
-    \caption{Raw data for leaf biomass and total above-ground biomass for each of the PFTs. Each point is an individual plant.}
+    \caption{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_W$) for each of the PFTs, as a function of total plant height ($H$). Each point is an individual plant.}
     \label{fig:figureS2}
@@
 
+\renewcommand\thetable{S\arabic{table}}
+\setcounter{table}{0}
+
 <<tablepreload1>>=
-load("tables/Table_counts.RData")
 printit <- function(x, caption, label=NULL){
@@
 <<results='asis'>>=
-printit(tab_mlfastba, 
+printit(table_mlfastba,
         label="table:count1",
@@
 <<results='asis'>>=
-printit(tab_alfastba, 
+printit(table_alfastba,
         label="table:count2",
@@
 <<results='asis'>>=
-printit(tab_lmf, 
+printit(table_lmf,
         label="table:count3",
@@
 <<results='asis'>>=
-printit(tab_lar, 
+printit(table_lar,
         label="table:count4",
@@
 <<results='asis'>>=
-print(xtable(Table_pipemodel_varpart, label="varpartpipe", 
+print(xtable(table_pipemodel_varpart, label="varpartpipe",
              caption="Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.
@@
 <<results='asis'>>=
-load("tables/Tables_varpart.RData")
-print(xtable(Table_LMFLAR_varpart, label="varpartlmflar",
+print(xtable(table_LMFLAR_varpart, label="varpartlmflar",
       caption="Explained variance in the leaf mass fraction (leaf mass / aboveground biomass) and leaf area ratio (leaf area / aboveground biomass).
@@
 <<>>=
-load("tables/basalafit.RData")
 @
@@
 \clearpage
-\bibliography{allometry}
+\bibliography{references}
 

### 2014-12-08 R/figures-extra.R

@@ -1,51 +1,0 @@
-
-
-source("load.R")
-dataset <- droplevels(subset(baad, pft != "DG"))
-dataset$pft <- as.factor(dataset$pft)
-source("R/functions-figures.R")
-
-# Cairns et al 1997 reported lower fine root / total root mass ratio for
-# gymnosperms
-dat <- droplevels(studyWithVars(dataset, c("m.rt","m.rf")))
-smoothplotbypft(log10(m.rt), log10(m.rf),dat)
-
-
-# Root-shoot ratio
-# See Cheng and Niklas 2007. They have near isometry,
-# relationships are ca. MA = 0.42*MR^1.05
-# Also no sign. effects of pft.
-dat <- droplevels(studyWithVars(dataset, c("m.rt","m.so")))
-smoothplotbypft(log10(m.rt), log10(m.so),dat)
-abline(0.4, 1, lwd=2)  # ca. Cheng & Niklas.
-
-
-# McCarthy et al 2007.
-# Residuals of MF vs. MS correlated with SLA?
-# Not the way to go, since the relationship is not allometric to begin with...
-dat <- droplevels(studyWithVars(dataset, c("m.st","m.lf")))
-dat <- subset(dat, !is.na(m.st) & !is.na(m.lf))
-smoothplotbypft(log10(m.st), log10(m.lf),dat)
-
-dat$mlffit <- fitted(sma(m.lf ~ m.st, data=dat, log="xy"))
-with(dat, plot(log10(a.lf/m.lf), log10(m.lf-mlffit)))
-
-
-# McMahon 1973
-# Niklas and Spatz 2004
-smoothplotbypft(log10(d.bh), log10(h.t),dataset)
-smoothplotbypft(log10(d.ba), log10(h.t),dataset)
-
-# Enquist Niklas 2002
-# Chave?
-smoothplotbypft(log10(d.bh^2*h.t), log10(m.st), dataset)
-smoothplotbypft(log10(d.ba^2*h.t), log10(m.st), dataset)
-
-
-
-# Poorter1999
-p <- droplevels(subset(baad, studyName == "Poorter1999"))
-smoothplotbypft(log10(m.st), log10(m.lf),p)
-smoothplotbypft(log10(m.st), log10(a.lf),p)
-
-

### 2014-12-08 R/plot-utils.R

@@ -1,130 +1,0 @@
-
-blank_plot <- function(xlim, ylim,
-  xlab=NULL, ylab=NULL,line=4,
-  xtick=NULL,ytick=NULL,
-  xtick.lab=xtick,ytick.lab=tick,...){
-
-  plot(NA, xlim = xlim, ylim = ylim, xaxt='n', yaxt='n', ann=FALSE,...)
-
-  if(!is.null(xtick))
-   axis(1, at=xtick, labels=xtick.lab)
-
- if(!is.null(ytick))
-   axis(2, at=ytick, labels=ytick.lab, las=1)
-
- if(!is.null(xlab))
-  mtext(xlab, 1, line=line)
-
- if(!is.null(ylab))
-  mtext(ylab, 2, line=line)
-
-}
-
-
-to.dev <- function(expr, dev, filename, ..., verbose=TRUE){
-  if(!file.exists(dirname(filename)))
-    dir.create(dirname(filename), recursive=TRUE)
-  if ( verbose )
-    cat(sprintf("Creating %s\n", filename))
-  dev(filename, ...)
-  on.exit(dev.off())
-  eval.parent(substitute(expr))
-}
-
-to.pdf <- function(expr, filename, ..., verbose=TRUE) {
-  if(!file.exists(dirname(filename)))
-    dir.create(dirname(filename), recursive=TRUE)
-  if ( verbose )
-    cat(sprintf("Creating %s\n", filename))
-  pdf(filename, ...)
-  on.exit(dev.off())
-  eval.parent(substitute(expr))
-}
-
-na.clean <-function(x){x[!is.na(x)]}
-
-#returns vector from lo to hi with multiplication steps of incr. Used for making ticks to a log-scaled axis
-seq.log <- function(from, to, base)
-{base^(log(from,base):log(to,base))}
-
-#returns vector of minor tick spacings approrpaite for log 10 scaled axis with major ticks given by 'major'.
-seq.log.minor<-function(major)
-{
-  temp<-NULL;
-  if(length(major) > 1)
-    for (i in 1:(length(major)-1))
-      temp<-c(temp, seq(major[i], major[i+1]-major[i], major[i]));
-  temp<-c(temp, major[length(major)]);
-  return (temp);
-}
-
-
-
-#returns up to 80 unique, nice colors, generated using http://tools.medialab.sciences-po.fr/iwanthue/
[…]

### 2014-12-08 R/predict_dba_frombh.R

@@ -1,27 +1,0 @@
-
-
-makebasalD <- function()
-  
-  
-
-test <- subset(baad, !is.na(d.ba) & !is.na(d.bh) & !is.na(h.t) & !is.na(h.bh) & h.t > h.bh)  
-
-
-fit <- nls(d.ba ~ d.bh * h.t^(a*h.t^b) /(h.t - h.bh)^(a*h.t^b), start=list(a=0.9, b=1),
-           data=test)
-
-test$d.ba2 <- predict(fit, test)
-
-with(test, plot(log10(d.ba2), log10(d.ba), col=as.factor(studyName), pch=16))
-abline(0,1)
-
-with(test, plot(d.ba2, d.ba))
-
-with(test, plot(h.t, d.ba - d.ba2))
-abline(h=0)
-
-# look at coefs.
-hts <- seq(2, 100, length=101)
-p <- predict(fit, expand.grid(d.bh=1, h.bh=1.3, h.t=hts))
-
-plot(hts, p)

### 2014-12-08 dependencies.R

@@ -1,10 +1,0 @@
-#!/usr/bin/env Rscript
-
-pkgs <- c("lmerTest","maps","mapdata","maptools","gdata","knitr","smatr","scales",
-          "xtable","magicaxis","rmarkdown","dplyr",
-          "multcomp","doBy","mgcv","reshape2","lme4","Hmisc","hexbin",
-          "lattice","grid","stringr","raster","dismo","XML","rgdal","lsmeans",
-          "gplots","car","RColorBrewer")
-
-pkgs.missing <- setdiff(pkgs, rownames(installed.packages()))
-install.packages(pkgs.missing)

### 2014-12-08 R/LMF_SLA_bystudy.R
Move to maker, ruthless cull & restructure
Tried to reduce to only materials currently being used
Use maker workflow instead of Makefile

@@ -1,78 +1,0 @@
-
-
-source("load.R")
-
-b <- studyWithVars(baad, c("m.so","m.lf","a.lf"))
-b$lma <- with(b, m.lf/a.lf)
-b <- subset(b, !is.na(m.lf) & !is.na(a.lf) & !is.na(m.so))
-tab <- with(b, table(pft, studyName))
-
-st <- colnames(tab)[apply(tab, 2, function(x)length(x[x>0]) > 1)]
-
-dat <- subset(b, studyName %in% st)
-dats <- split(dat, dat$studyName)
-
-pdf("LMF_LAR_bystudy.pdf", width=8, height=4)
-for(i in 1:length(dats)){
-  message(i)
-  par(mfrow=c(1,2))
-  smoothplotbypft(log10(m.so), log10(m.lf), dats[[i]])
-  smoothplotbypft(log10(m.so), log10(a.lf), dats[[i]])
-  title(names(dats)[i], outer=T, line=-2, cex.main=0.9)
-}
-dev.off()
-
-
-
-#-------------------------------------------------------------------------------------------#
-# Averages. Not successful because of size difference (within and between datasets)
-fx <- function(x){
-  x <- x[complete.cases(x[,c("m.lf","m.so","lma","a.lf")]),]
-  x$pft <- as.factor(x$pft)
-  lmf <- function(p)with(p, 10^mean(log10(m.lf / m.so)))
-  lar <- function(p)with(p, 10^mean(log10(a.lf / m.so)))
-  lma <- function(p)with(p, 10^mean(log10(m.lf / a.lf)))
-  mso <- function(p)with(p, 10^mean(log10(m.so)))
-  X <- split(x,x$pft)
-  df <- data.frame(pft=levels(x$pft),
-                   LMF=sapply(X,lmf),
-                   LAR=sapply(X,lar),
-                   LMA=sapply(X,lma),
-                   MSO=sapply(X,mso)
-                   )
-return(df)
-}
-
-pftm <- lapply(dats,fx)
-
-
-palette(rainbow(13))
-
-plot(1, type='n', xlim=c(0,0.3), ylim=c(-3,0))
-for(i in 1:13){
-  with(pftm[[i]], points(LMA,log10(LMF), type='o',
-                         pch=15, col=i))
-  
-}
-
-
-plot(1, type='n', xlim=c(0,0.3), ylim=c(-2,1.5))
-for(i in 1:13){
[…]

### 2014-12-08 R/WorldMap.R

@@ -1,27 +1,0 @@
-
-to.pdf({
-  
-  # Colored by vegetation type
-  cols <- alpha(c("blue","brown","forestgreen","darkorange","dodgerblue2",
-                    "navajowhite4","lightgoldenrod3","lavender"),0.67)
-  vegs <- c("BorF","Sav","TempF","TempRF","TropRF","TropSF","Wo")
-  vegs_labels <- c("Boreal forest","Savannah","Temperate forest","Temperate rainforest","Tropical rainforest",
-                   "Tropical seasonal forest","Woodland","Glasshouse / Common garden")
-  
-  drawWorldPlot(baad[baad$vegetation == vegs[1],], horlines=FALSE, sizebyn=TRUE, pchcol=cols[1])
-  for(i in 2:length(vegs)){
-    drawWorldPlot(baad[baad$vegetation == vegs[i],], horlines=FALSE, sizebyn=TRUE, pchcol=cols[i], add=TRUE)
-  }
-  drawWorldPlot(baad[baad$growingCondition == "GH",], horlines=FALSE, sizebyn=TRUE, 
-                pchcol=cols[8], add=TRUE)
-  par(xpd=NA)
-  legend(-160, -100, vegs_labels[1:4], fill=cols[1:4], bty='n')
-  legend(-40, -100, vegs_labels[5:8], fill=cols[5:8], bty='n')
-}, "output/pdf/WorldMap.pdf", width=10, height=8)
-
-
-
-
-
-
-

### 2014-12-08 R/build.R

@@ -1,0 +1,11 @@
+
+tex_2_pdf <- function(texfile){
+
+  filename <- tools::file_path_sans_ext(texfile)
+  system(sprintf("pdflatex %s", texfile))
+  system(sprintf("bibtex %s",   filename))
+  system(sprintf("pdflatex %s", texfile))
+  system(sprintf("pdflatex %s", texfile))
+  aux.files <- paste0(filename, c(".log", ".aux", ".bbl", ".blg"))
+  file.remove(aux.files[file.exists(aux.files)])
+}

### 2014-12-08 R/data_processing.R

@@ -1,0 +1,298 @@
+
+extract_baad_data <- function(baad) {
+  baad$data
+}
+
+extract_baad_dictionary <- function(baad) {
+  baad$dictionary
+}
+
+
+# Convert conifer leaf area to projected leaf area
+convertConiferLA <- function(baad) {
+
+  # One sided total leaf area (half total surface area)
+  # to projjected area. Average of species in Barclay & Goodman 2000
+  # For Pinus we use the value for lodgepole pine (see below)
+  lambda1 <- c(0.873, 0.92, 0.879, 0.864, 0.839)
+  ola_pla <- mean(1/lambda1)
+
+  conv_pine <- function(x, method){
+    if(method %in% c("","?","ax"))
+      method <- "unknown"
+
+    cv <- 1/0.778
+    switch(method,
+           a4 = x * cv,
+           a5 = x/2,
+           a6 = x,
+           a7 = x/2,
+           unknown = x
+           )
+  }
+  conv_nonpine <- function(x, method){
+    if(method %in% c("","?","ax"))
+      method <- "unknown"
+
+    switch(method,
+           a4 = x * ola_pla,
+           a5 = x/2,
+           a6 = x,
+           a7 = x/2,
+           unknown = x
+    )
+  }
+
+  convf <- function(x, method, species, pft){
+
+    if(pft %in% c("DA","EA"))return(x)
+
+    if(grepl("Pinus", species,ignore.case=TRUE))
+      conv_pine(x, method)
+    else
+      conv_nonpine(x, method)
+  }
+  conv <- Vectorize(convf)
+
+  alfmeth <- baad$methods[,c("studyName","a.lf")]
+  alfmeth$method_alf <- str_extract(alfmeth$a.lf,"a[4-7]{1}")
+  alfmeth$method_alf[is.na(alfmeth$method_alf)] <- ""
+  baad$data <- merge(baad$data, alfmeth[,c("studyName","method_alf")],all=T)
+
+  with(baad$data, conv(a.lf, method_alf, speciesMatched, pft))
+
+  baad
+}
+
+
+getWorldClim <- function(longitude, latitude, varname, worldclim_dataloc = "c:/data/worldclim"){
+
+  here <- data.frame(lon=longitude,lat=latitude)
+  coordinates(here) <- c("lon", "lat")
[…]

### 2014-12-08 R/figures.R

@@ -1,28 +1,37 @@
 
-# Load BAAD and add climate data etc.
-source("load.R")
 
-# Prepare 'dataset' and 'dataset2'.
-source("R/preparedataset.R")
-source("R/functions-figures.R")
-source("R/signifletters.R")  # cld.lsmeans, based on cld.glht
 
 # Color settings
-Cols <- c("blue","red","forestgreen")
-transCols <- alpha(Cols,0.4)
-linecols <- c("deepskyblue3","firebrick2","chartreuse3")
-palette(Cols)
+my_cols <- function() {
+  c("blue","red","forestgreen")
+}
+
+my_cols_transparent <- function(a=0.4) {
+  alpha(my_cols(),a)
+}
+
+my_linecols <- function() {
+  c("deepskyblue3","firebrick2","chartreuse3")
+}
+
+#palette(Cols)
 
 
 # Plot settings.
-lmaLabel <- expression("Leaf mass per area"~~(kg~m^-2))
-lmaLabelshort <- expression(M[F]/A[F]~~(kg~m^-2))
+lmaLabel <- function() {
+  expression("Leaf mass per area"~~(kg~m^-2))
+}
 
-Legend <- function(where, labels=c("short","long"), bty='n', rev=TRUE, ...){
+lmaLabel_short <- function() {
+  expression(M[F]/A[F]~~(kg~m^-2))
+}
+
+my_legend <- function(where, labels=c("short","long"), bty='n', rev=TRUE, ...){
   lab <- if(match.arg(labels) == "short")
             c("Decid. Angio.", "Evergr. Angio.", "Evergr. Gymno.")
          else
            c("Deciduous Angiosperm", "Evergreen Angiosperm", "Evergreen Gymnosperm")
+  Cols <- my_cols()
   if(rev){
     lab <- rev(lab)
     Cols <- rev(Cols)
@@ -31,11 +40,8 @@ Legend <- function(where, labels=c("short","long"), bty='n', rev=TRUE, ...){
          pch=19, col=Cols, bty=bty, ...)
 }
 
-# Average SLM by PFT, used in several plots.
-lma <- mixmean("llma", "pft", dataset)
-
-# Base parameter for all gam fits.
-KGAM <- 4
+# # Average SLM by PFT, used in several plots.
+# lma <- mixmean("llma", "pft", dataset)
 
 
 
[…]

### 2014-12-08 R/glasshouse_vs_field.R

@@ -1,47 +1,0 @@
-
-
-dat <- studyWithVars(baad, c("m.lf", "m.so", "h.t"))
-dat$Group <- as.factor(with(dat, paste(studyName, speciesMatched)))
-tab <- table(dat$Group)
-
-dat <- subset(dat, h.t < 3)
-
-N <- 15
-dat <- droplevels(subset(dat, Group %in% names(tab[tab >= N]) & growingCondition != ""))
-dat$mlfmso <- with(dat, m.lf / m.so)
-sm1 <- sma(mlfmso ~ h.t*Group, data=dat, log="xy")
-
-dat$newgrow <- ifelse(dat$growingCondition %in% c("GH","CG"), "Pot", "Field")
-
-
-p <- coef(sm1)
-p$Group <- rownames(p)
-
-cis <- do.call(rbind, lapply(sm1$slopetest, "[[", "ci"))
-cis <- as.data.frame(cis)
-names(cis) <- c("slope_lci","slope_uci")
-p <- cbind(p,cis)
-m <- dat[,c("Group","pft","newgrow","vegetation")]
-m <- m[!duplicated(m),]
-p <- merge(p,m, all=FALSE)
-
-
-p$slopesignif <- "smaller"
-p$slopesignif[p$slope_lci < 0 & p$slope_uci > 0] <- "equal"
-p$slopesignif[p$slope_lci > 0 & p$slope_uci > 0] <- "larger"
-
-tab <- with(p, table(slopesignif, newgrow))
-tabprop <- sweep(tab, 2, colSums(tab), FUN="/")
-
-
-ran <- function(x,...)max(x, na.rm=TRUE) - min(x, na.rm=TRUE)
-R <- summaryBy(h.t ~ Group, FUN=c(ran,min), data=dat, na.rm=TRUE)
-names(R)[2] <- "ht_range"
-p <- merge(p, R)
-
-p <- subset(p, slope1 > 0 & ht_range > 0.1)
-
-
-morethan1 <- function(x)length(x[x>1])/length(x)
-barplot(with(p, tapply(slope1/slope2,growingCondition,FUN=morethan1)))
-

### 2014-12-08 R/publications_byyear.R

@@ -1,19 +1,0 @@
-
-
-library(stringr)
-ref <- baad$references
-
-getY <- function(x)str_extract(x, "([0-9]{4})")
-
-y <- as.numeric(unname(sapply(ref$citation, getY)))
-tab <- table(y)
-tabdf <- data.frame(year=names(tab), npub=as.vector(tab))
-dfr <- data.frame(year=min(y,na.rm=T):max(y,na.rm=T))
-dfr <- merge(dfr, tabdf,all=TRUE)
-dfr$npub[is.na(dfr$npub)] <- 0
-plot(dfr, type='h', xlim=c(1950, 2015),
-     ylab="Number of publications",
-     xlab="Year published")
-
-
-

### 2014-12-08 R/tables_stats.R

@@ -1,0 +1,159 @@
+
+make_r2_table <- function(data,  variable_name ){
+  data.frame(Variable = c(variable_name, rep(NA, length(data)-1)),
+    Predictors=names(data),
+    r2=sapply(data,function(x) x[,"Conditional"]))
+}
+
+
+make_table_pipemodel_varpart <- function(dataset2) {
+
+  # Datasets with NAs removed, of key variables.
+  assign("dat_mlf", droplevels(subset(dataset2, !is.na(h.t) & !is.na(pft) & !is.na(lmlf_astba2))), envir = .GlobalEnv)
+  assign("dat_alf", droplevels(subset(dataset2, !is.na(h.t) & !is.na(pft) & !is.na(lalf_astba2))), envir = .GlobalEnv)
+
+  mlf_formulas <- list()
+  mlf_formulas[["H"]]         <- lmlf_astba2 ~ log10(h.t) + (1|Group)
+  mlf_formulas[["H, B"]]      <- lmlf_astba2 ~ log10(h.t)*bortemptrop + (1|Group)
+  mlf_formulas[["PFT, H, B"]] <- lmlf_astba2 ~ log10(h.t)*pft*bortemptrop + (1|Group)
+  mlf_formulas[["LMA, H, B"]] <- lmlf_astba2 ~ log10(h.t)*lsla*bortemptrop + (1|Group)
+
+  mlf_lmer_fits <- lapply(mlf_formulas, function(x) lmer(x, data=dat_mlf))
+  mlf_lmer_r2 <- lapply( mlf_lmer_fits, r.squared.merMod)
+
+  alf_formulas <- gsub("lmlf_astba2", "lalf_astba2", mlf_formulas)
+  names(alf_formulas) <-   names(mlf_formulas)
+  alf_lmer_fits <- lapply(alf_formulas, function(x) lmer(x, data=dat_alf))
+  alf_lmer_r2 <- lapply( alf_lmer_fits, r.squared.merMod)
+
+  rm(dat_mlf, dat_alf, envir = .GlobalEnv)
+  rbind(make_r2_table(mlf_lmer_r2, "LMF_ASTBA"),
+        make_r2_table(alf_lmer_r2, "LAR_ASTBA"))
+}
+
+
+make_table_LMFLAR_varpart <- function(dataset2) {
+
+assign("dat_mlf", droplevels(subset(dataset2, !is.na(h.t) & !is.na(pft) & !is.na(lmlf_astba2))), envir = .GlobalEnv)
[…]

### 2014-12-08 R/functions-figures.R
Fixed figure2 formatting

@@ -164,9 +164,9 @@ meansbypft <- function(yvar1, yvar2=NULL, pftvar,
     if(siglets == "symbol"){
       pointLabel(X$y, y2$y, lets2, cex=0.9)
     }
-    mtext(side=1, text=xlab, line=3, outer=T)
-    mtext(side=2, at = 0.25, text=ylab2, line=3, outer=T)
-    mtext(side=2, at = 0.75, text=ylab1, line=3, outer=T)
+    mtext(side=1, text=xlab, line=3, outer=T, las=0)
+    mtext(side=2, at = 0.25, text=ylab2, line=3, outer=T, las=0)
+    mtext(side=2, at = 0.75, text=ylab1, line=3, outer=T, las=0)
   }
   
   mtext(side=4, text=main, line=2, outer=TRUE)

### 2014-12-08 manuscript/draftpaper.Rnw
Fixed figure2 formatting
SI tables now properly numbered

@@
     \centering
-    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
+    \includegraphics[width=0.5\textwidth]{Figure2.pdf}
     \caption{Average leaf mass per unit basal stem area (panel 1), and leaf area per unit basal stem area (panel 2) for the three major plant functional types.
@@
 
+\renewcommand\thetable{S\arabic{table}}    
+\setcounter{table}{0}
+
 <<tablepreload1>>=

### 2014-12-08 manuscript/draftpaper.Rnw
Minor figure formatting and captions

@@
 print(xtable(g0, label="table/gamr2table",
-      caption="Explained variance by plant functional type (PFT), plant height (H), moisture index (MI) and growing season average temperature (mgdd0), for six whole-plant variables."),
+      caption="Explained variance by plant functional type (PFT), plant height (H), moisture index (MI) and growing season average temperature (mgdd0), for six whole-plant variables.
+Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),
       include.rownames=FALSE, caption.placement="top")
@@
     \includegraphics[width=0.5\textwidth]{Figure2.pdf}
-    \caption{Average leaf mass per unit basal stem area (panel 1), and leaf area per unit basal stem area (panel 2) for the three major plant functional types.
+    \caption{Average leaf mass per unit basal stem area (panel 1), and leaf area per unit basal stem area (panel 2) for the three major PFTs.
 Error bars are 95\% confidence interval for the mean (calculated with a mixed-effects model with species within study as the random effect).

### 2014-12-09 manuscript.Rnw
Some rewriting of opening para

@@
 
-\section*{First paragraph}
-
-\textbf{The partitioning of biomass in terrestrial vegetation between leaves, stems and roots is a critical component of dynamic global vegetation models (DGVMs).
-Biomass partioning is consistently shown to be one of the most uncertain and important drivers for global predictions of carbon storage \cite{ise_comparison_2010}, the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
-Little is known about direct and indirect effects of climate on biomass partitioning, largely due to the lack of a global individual-based biomass dataset.
-Here, we analyze the largest database on plant biomass to date [cite].
-We show that partitioning of biomass to leaves vs.
-woody stems is strongly dependent on the plant functional type.
-Evergreen gymnosperms maintain three times more leaf biomass at a given total aboveground biomass than deciduous angiosperms, with evergreen angiosperms taking an intermediate value.
-We found that leaf mass per area, an indicator of leaf construction cost, explains these differences so that the PFTs converged in the ratio  of leaf area to aboveground biomass.
-We found weak effects of climate while accounting for PFT effects, suggesting that climate-related differences are captured largely by a shift in PFT.
-These results can be used to constrain biomass partitioning routines in global dynamic vegetation models.}
-
 % \section{flow}
@@
 partitioning: the role of leaf and stem turnover.
+% \item Another implication is that, as a first approximation, using LMA is more appropriate than splitting by PFT.
+Using LMA, together with the assumption of invariant leaf area display, gives biomass partitioning between leaves and stems.
+This even worked, to some extent, across species.
 %
@@
 
+\section*{First paragraph}
+
+\textbf{The partitioning of biomass in terrestrial vegetation between long-lived stems and short-lived leaves affects the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, model predictions of global carbon storage \cite{ise_comparison_2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
+Little is known about direct and indirect effects of climate on biomass partitioning, largely due to the lack of a global individual-based biomass dataset.
+Here, we analyze a new database on plant biomass fractions, the largest to date [cite].
+We show that partitioning of biomass to leaves vs.
+woody stems is strongly dependent on the plant functional type, and proportional to the leaf mass per area, an important leaf functional trait.
+As a result, plant functional types converged in the ratio of leaf area to aboveground biomass.
+We found only modest effects of climate after accounting for PFT, suggesting that climate-related differences in biomass partitioning are captured largely by a shift in PFT.
+These results form an important constraint on biomass partitioning routines in global dynamic vegetation models.}
+
 \section*{Main text}
 
-Previous syntheses on plant biomass relationships has predominantly focused on the universal nature of size-biomass relationships in plants.
-Little is known about differences between plant functional types (evergreen vs.
+% 1.
+Importance of allocation in global vegetation models, short-lived leaves vs.
+long-lived stems affect ecosystem-scale turnover of C, as well as productivity (as partitioning to leaves affects LAI).
+DGVMs split plant diversity into a few plant functional types (PFTs).
+This begs two questions : how does biomass partitioning dffer between PFTs, and can we relate it to commonly measured leaf traits, to simplify application of DGVMs?
+
+Previous syntheses on plant biomass relationships have predominantly focused on the universal nature of size-biomass relationships in plants.
+Relatively little is known about differences in biomass partitioning between PFTs(evergreen vs.
 deciduous, angiosperm vs.
 gymnosperm).
-Since most global models parameterize allocation submodels by PFT, it is imperative that we better understand these potential differences.
 
@@
 
-%------------------------------------------------------------------------------------%
 \section*{Methods}

### 2014-12-09 R/figures.R
Fig 4b with lm instead (less confusing), ms uses MAP, MAT

@@ -225,7 +225,7 @@ figure4 <- function(dataset, KGAM=3, climvar1="MI", climvar2="mgdd0", fitoneline
 }
 
 
-figure4b <- function(dataset, climvar1="MAP", climvar2="MAT", fitoneline=FALSE, fittype="gam"){
+figure4b <- function(dataset, climvar1="MAP", climvar2="MAT", fitoneline=FALSE, fittype="lm"){
   figure4(dataset, climvar1=climvar1, climvar2=climvar2, fitoneline=fitoneline,fittype=fittype)  
 }
 

### 2014-12-09 manuscript.Rnw
Fig 4b with lm instead (less confusing), ms uses MAP, MAT
Figure3b has histograms on top of each other; now in ms as well. Closes #7

@@
     \centering
-    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
+    \includegraphics[width=0.99\textwidth]{Figure3b.pdf}
     \caption{Plant functional types diverge strongly in mass-based allocation, but converge in area-based allocation.
 Shown are histograms of leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types.
-Vertical lines and shading are means with confidence intervals (same as in Fig.~\ref{fig:figure2}).
-Histograms of $M_F/A_S$ show large differences between PFTs, while $A_F/A_S$ converges to a single average across all PFTs, though much variation exists between plants.}
+Vertical lines and shading are means with confidence intervals (same as in Fig.~\ref{fig:figure2}).}
     \label{fig:figure3}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{Figure4.pdf}
+    \includegraphics[width=0.99\textwidth]{Figure4b.pdf}
     \caption{Modest effects of climate on three size-invariant variables, compared to the very large effect of plant functional type on mass-based variables (leaf mass per area, LMA, and $M_F/A_S$), and convergence in area-based variables ($A_F/A_S$).

### 2014-12-10 manuscript.Rnw
first para, second version

@@
 opts_chunk$set(echo=FALSE)
-
+source("R/functions-ms.R")
 @
@@
 
+% Remko version new
+\textbf{
+- The partitioning of biomass into short-lived leaves and long-lived stems is one of the most uncertain  and influential components of dynamic global vegetation models (DGVMs), affecting the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, predictions of global carbon storage \cite{ise_comparison_2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
+Despite its importance, little is known about the variation in biomass partitioning due to plant functional type (PFT), biome, or climate.
+Here, we use a new large compilation of global woody plant biomass data to quantify this variation.
+We show that across three major PFTs, the partitioning to total plant leaf mass is proportional to leaf mass per area, an important leaf functional trait.
+As a result, plant functional types converged in the ratio of leaf area to aboveground biomass.
+We found only modest effects of biome or climate after accounting for PFT.
+These results suggest that leaf mass per area can be substituted for PFT to predict biomass partitioning to leaves and stems in DGVMs.
+We recommend that these results are used as a benchmark to test the many existing biomass allocation routines \cite{dekauwe2014}.}
+
+% Remko version old
 \textbf{The partitioning of biomass in terrestrial vegetation between long-lived stems and short-lived leaves affects the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, model predictions of global carbon storage \cite{ise_comparison_2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.

### 2014-12-10 manuscript.Rnw

@@
 These results form an important constraint on biomass partitioning routines in global dynamic vegetation models.}
+
+\section*{First paragraph - Daniel}
+
+\textbf{Leaf traits drive differences in biomass partitioning among major plant functional types}
+
+The partitioning of plant primary productivity among fleshy leaves and woody stems is a key factor driving carbon cycles and the flow of energy within terrestrial ecosystems.
+Although researchers have long assumed key plant functional types (PFTs) to differ in biomass partitioning, empirical data have been lacking.
+Divergent approaches to modelling allocation is thus a major source of uncertainty in terrestrial ecosystem models (TEMs) and global carbon cycles.
+Using a new database on Biomass partitioning combining data from over 175 studies, we here show a strong effect of PFT on biomass partitioning.
+Contrary to prior expectation, we show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass.
+Moreover, we find that such differences relate to systematic shifts in leaf-mass per area, a leading trait within the leaf economics spectrum.
+Another core trait, leaf area per sapwood area, explained further variation among species in biomass partitioning, but did not differ systematically among PFTs.
+Combined, these results uncover fundamental differences in how plants are constructed and allow for systematic testing of allocation routines with current TEMs.
 

### 2014-12-10 manuscript.Rnw
Title change ms; caption fig 1

@@
 
-\title{Plant functional type effects on biomass partitioning}
+\title{Leaf traits drive differences in biomass partitioning among major plant functional types}
 
@@
 
-\textbf{Leaf traits drive differences in biomass partitioning among major plant functional types}
-
+\textbf{
 The partitioning of plant primary productivity among fleshy leaves and woody stems is a key factor driving carbon cycles and the flow of energy within terrestrial ecosystems.
@@
 Another core trait, leaf area per sapwood area, explained further variation among species in biomass partitioning, but did not differ systematically among PFTs.
-Combined, these results uncover fundamental differences in how plants are constructed and allow for systematic testing of allocation routines with current TEMs.
+Combined, these results uncover fundamental differences in how plants are constructed and allow for systematic testing of allocation routines with current TEMs.}
 
@@
 
+%(a) Global coverage of the climate space by the dataset.
+Hexagons are terrestrial observations, excluding areas without woody vegetation, colored by number of 0.5\degree~cells, cut off at 6000mm MAP.
+Triangles are averages for the three PFTs across all contributed studies.
+
 \begin{figure}[h!]
@@
     \includegraphics[width=0.99\textwidth]{Figure1.pdf}
-    \caption{(a) Global coverage of the climate space by the dataset.
-Hexagons are terrestrial observations, excluding areas without woody vegetation, colored by number of 0.5\degree~cells, cut off at 6000mm MAP.
-Triangles are averages for the three PFTs across all contributed studies.
+    \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf construction cost.
 (b) Leaf mass fraction (LMF = leaf mass / aboveground biomass) by PFT.
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureS1.pdf}
+    \includegraphics[width=0.99\textwidth]{FigureS1b.pdf}
     \caption{Mean annual precipitation (MAP) and temperature (MAT) for all contributed datasets, by vegetation type.

### 2014-12-10 manuscript.Rnw
Figure 3 updated (old fig 3 removed)

@@
     \centering
-    \includegraphics[width=0.99\textwidth]{Figure3b.pdf}
-    \caption{Plant functional types diverge strongly in mass-based allocation, but converge in area-based allocation.
-Shown are histograms of leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types.
-Vertical lines and shading are means with confidence intervals (same as in Fig.~\ref{fig:figure2}).}
+    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
+    \caption{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.
+Shown are histograms of leaf mass per area (an index of leaf construction cost) ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types.
+Arrows indicate means (same as in Fig.~\ref{fig:figure2}).}
     \label{fig:figure3}

### 2014-12-10 R/figures.R

@@ -208,45 +208,24 @@ figure2 <- function(dataset){
 
 # Figure 3. Histograms of MF/AS and AF/AS.
 figure3 <- function(dataset){
-
-  par(mfcol=c(3,2), mar=c(0,0,0,0), oma=c(5,5,2,2), las=1)
-  histbypft(lmlf_astba2, pft, dataset, xaxis=3,legend.cex=1,col=my_cols(),
-            xlab="",
-            Means=mixmean("lmlf_astba2","pft",dataset),
-            legend.text=c("Deciduous Angiosperm",
-                          "Evergreen Angiosperm",
-                          "Evergreen Gymnosperm"))
-  histbypft(lalf_astba2, pft, dataset, xaxis=3,legend.cex=1,col=my_cols(),
-            xlab="",
-            Means=mixmean("lalf_astba2","pft",dataset),
-            legend.text=rep("",3))
-  mtext(side=1, line=3, text=expression(M[F]/A[S]~~(kg~m^-2)),
-        outer=TRUE, at=0.25, cex=0.9)
-  mtext(side=1, line=3, text=expression(A[F]/A[S]~~(m^2~m^-2)),
-        outer=TRUE, at=0.75, cex=0.9)
-}
-
-
-# Figure 3. Histograms of MF/AS and AF/AS.
-figure3b <- function(dataset){
   
-  par(mar=c(0,0,0,0), oma=c(5,5,1,1), las=1, mfrow=c(1,3), cex.axis=0.85)
+  par(mar=c(0,0,0,2), oma=c(5,5,1,1), las=1, cex.axis=0.85, mfrow=c(1,3), mgp=c(3,1.5,0))
   
   histbypft(llma, pft, dataset, xaxis=3,legend.cex=1,col=my_cols_transparent(), 
             xlab="", overlay=TRUE,plotwhat="density",ylab="Density",
-            Means=mixmean("llma","pft",dataset),cicol=alpha("white",0.5),
+            Means=mixmean("llma","pft",dataset),cicol=alpha("grey",0.5),
             legend.text="", meanlinecol=my_linecols())
   plotlabel("(a)","topleft")
   
   histbypft(lmlf_astba2, pft, dataset, xaxis=3,legend.cex=1,col=my_cols_transparent(), 
             xlab="", overlay=TRUE,plotwhat="density",ylab="Density",
-            Means=mixmean("lmlf_astba2","pft",dataset),cicol=alpha("white",0.5),
[…]

### 2014-12-10 manuscript.Rnw
Updated Figures 1 and 2. Fig 1 has crappy sketch of tree as placeholder.

@@
 
-%(a) Global coverage of the climate space by the dataset.
-Hexagons are terrestrial observations, excluding areas without woody vegetation, colored by number of 0.5\degree~cells, cut off at 6000mm MAP.
-Triangles are averages for the three PFTs across all contributed studies.
-
 \begin{figure}[h!]
@@
     \includegraphics[width=0.99\textwidth]{Figure1.pdf}
-    \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf construction cost.
-(b) Leaf mass fraction (LMF = leaf mass / aboveground biomass) by PFT.
-Each symbol is an individual plant.
-Lines are generalized additive model fits.
-(c) and (d) Leaf mass fraction (LMF) and leaf area ratio (LAR) at the average plant height in the dataset, estimated from panel (b).
-Error bars are 95\% confidence intervals.
-Letters denote significant differences.}
+    \caption{(a) Global coverage of the climate space by the dataset.
+Hexagons are terrestrial observations, excluding areas without woody vegetation, colored by number of 0.5\degree~cells, cut off at 6000mm MAP.
+Triangles are averages for the three PFTs across all contributed studies.
+(b) Schematic of a woody plant, and the variables considered in this study.
+H = plant height, $A_S$ basal stem area, $A_sbh$ stem area at 'breast height', $M_W$ woody biomass, $M_F$ leaf biomass (total aboveground biomass $M_T = M_F + M_W$), $A_F$ total leaf area.}
     \label{fig:figure1}
@@
     \centering
-    \includegraphics[width=0.5\textwidth]{Figure2.pdf}
-    \caption{Average leaf mass per unit basal stem area (panel 1), and leaf area per unit basal stem area (panel 2) for the three major PFTs.
-Error bars are 95\% confidence interval for the mean (calculated with a mixed-effects model with species within study as the random effect).
-Letters denote significant differences (at $\alpha = 0.05$).
-}
+    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
+    \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf construction cost.
+(a) Leaf mass fraction (LMF = leaf mass / aboveground biomass) by PFT.
+Each symbol is an individual plant.
+Lines are generalized additive model fits.
+(b) and (c) Leaf mass fraction (LMF) and leaf area ratio (LAR) at the average plant height in the dataset, estimated from panel (a).
+(d) Average leaf mass per unit basal stem area, and (e) leaf area per unit basal stem area for the three major PFTs confirm that the between-PFT variation in leaf mass fraction is due to leaf mass per unit basal area.
+Error bars are 95\% confidence intervals.
+Letters denote significant differences (at $\alpha = 0.05$).}
     \label{fig:figure2}

### 2014-12-11 manuscript.Rnw
New panel in SI4, some caption updates

@@
     \caption{(a) Global coverage of the climate space by the dataset.
-Hexagons are terrestrial observations, excluding areas without woody vegetation, colored by number of 0.5\degree~cells, cut off at 6000mm MAP.
-Triangles are averages for the three PFTs across all contributed studies.
+Hexagons are terrestrial observations, excluding areas without woody vegetation, colored by number of 0.5\degree~cells.
 (b) Schematic of a woody plant, and the variables considered in this study.
-H = plant height, $A_S$ basal stem area, $A_sbh$ stem area at 'breast height', $M_W$ woody biomass, $M_F$ leaf biomass (total aboveground biomass $M_T = M_F + M_W$), $A_F$ total leaf area.}
+H = plant height, $A_S$ basal stem area, $A_{sbh}$ stem area at 'breast height', $M_W$ woody biomass, $M_F$ leaf biomass (total aboveground biomass $M_T = M_F + M_W$), $A_F$ total leaf area.}
     \label{fig:figure1}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureS1b.pdf}
-    \caption{Mean annual precipitation (MAP) and temperature (MAT) for all contributed datasets, by vegetation type.
-MAP and MAT were taken from Worldclim \cite{worldclim}, using latitude and longitude reported for each contributed dataset.}
+    \includegraphics[width=0.99\textwidth]{FigureS1.pdf}
+    \caption{Global coverage of the climate space by the dataset. Methods are as in Fig 1 in main text, symbols are colored by vegetation type as reported by the data contributors.}
     \label{fig:figureS1}

### 2014-12-12 manuscript.Rnw
Table with collaborators. Couple of small glitches but not bad. Closes #1

--- 2014-12-10 manuscript.Rnw
@@
 \usepackage{gensymb}
+\usepackage{longtable}
 \usepackage{lineno}
@@
 
-\affil[1]{Hawkesbury Institute for the Environment}
-\affil[2]{Macquarie University}
+\affil[1]{1 Hawkesbury Insitute for the Environment, University of Western Sydney, Hawkesbury Campus, First Floor, Building L9, Locked Bag 1797, Penrith NSW 2751, Australia}
+\affil[2]{Biological Sciences, Macquarie University NSW 2109, Australia}
 \affil[3]{See Supporting Information}
@@
 source("R/functions-ms.R")
+baad <- readRDS("data/baad.rds")
+dataauthors <- author_details(baad)
 @
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureS1b.pdf}
+    \includegraphics[width=0.99\textwidth]{FigureS1.pdf}
     \caption{Mean annual precipitation (MAP) and temperature (MAT) for all contributed datasets, by vegetation type.
@@
 
+<<results='asis'>>=
+print(xtable(dataauthors,
+             caption="Contributors to the BAAD"
+             ), include.rownames=FALSE, tabular.environment='longtable', floating=FALSE, size="small",
+      caption.placement="top")
+@
+
 <<tablepreload1>>=
@@
 
-<<>>=
-@
-
 Using the subset of data where basal stem diameter ($A_S$) and diameter at breast height ($A_{Sbh}$) were measured (a total of \Sexpr{length(residuals(basalafit))} observations covering the three major PFTs), we developed a non-linear model to predict $A_S$ when only $A_{Sbh}$ and plant height ($H$) were measured.

### 2014-12-12 R/functions-figures.R
Table with collaborators. Couple of small glitches but not bad. Closes #1

@@ -5,7 +5,7 @@ plotlabel <- function(txt, where, inset=0.08, ...){
   if(grepl("right",where))x <- u[2] - inset*(u[2]-u[1])
   if(grepl("bottom",where))y <- u[3] + inset*(u[4]-u[3])
   if(grepl("top",where))y <- u[4] - inset*(u[4]-u[3])
-
+  
   text(x,y,txt,...)
 }
 

### 2014-12-14 manuscript.Rnw
New text in manuscript

@@
 
-% \section{flow}
-%
-% \begin{itemize}
-%
-% \item First paragraph
-% \item Allocation in models, main uncertainties
-% \item Climate effects on partioning : direct (MAT, MAP), indirect (PFT).
-% \item Known: gymnosperm vs. angiosperm. Unknown: leaf habit (deciduous vs. evergreen), direct climate, patterns for individual plants.
-% \item BAAD : Largest database, other large datasets are stands or only small plants
-% \item Found very large LMF differences, but convergent LAR
-% \item decompose into MF/AS, MT/AS, find similar differences in MF/AS but none in MT/AS, suggesting convergent AF/AS between PFTs.
-% \item Only small direct climate effects, so shift in PFT captures global patterns in biomass partioning.
-% \item This has implications for allocation in DGVMs, they can be divided into roughly three categories in terms of biomass allocation (allocation fractions, allometric constraints, LAI optimization).
-Allocation vs.
-partitioning: the role of leaf and stem turnover.
-% \item Another implication is that, as a first approximation, using LMA is more appropriate than splitting by PFT.
-Using LMA, together with the assumption of invariant leaf area display, gives biomass partitioning between leaves and stems.
-This even worked, to some extent, across species.
-%
-% \end{itemize}
-
 \section*{First paragraph}
 
-% Remko version new
 \textbf{
-- The partitioning of biomass into short-lived leaves and long-lived stems is one of the most uncertain  and influential components of dynamic global vegetation models (DGVMs), affecting the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, predictions of global carbon storage \cite{ise_comparison_2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
-Despite its importance, little is known about the variation in biomass partitioning due to plant functional type (PFT), biome, or climate.
-Here, we use a new large compilation of global woody plant biomass data to quantify this variation.
-We show that across three major PFTs, the partitioning to total plant leaf mass is proportional to leaf mass per area, an important leaf functional trait.
-As a result, plant functional types converged in the ratio of leaf area to aboveground biomass.
-We found only modest effects of biome or climate after accounting for PFT.
-These results suggest that leaf mass per area can be substituted for PFT to predict biomass partitioning to leaves and stems in DGVMs.
-We recommend that these results are used as a benchmark to test the many existing biomass allocation routines \cite{dekauwe2014}.}
-
-% Remko version old
-\textbf{The partitioning of biomass in terrestrial vegetation between long-lived stems and short-lived leaves affects the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, model predictions of global carbon storage \cite{ise_comparison_2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
-Little is known about direct and indirect effects of climate on biomass partitioning, largely due to the lack of a global individual-based biomass dataset.
-Here, we analyze a new database on plant biomass fractions, the largest to date [cite].
-We show that partitioning of biomass to leaves vs.
-woody stems is strongly dependent on the plant functional type, and proportional to the leaf mass per area, an important leaf functional trait.
-As a result, plant functional types converged in the ratio of leaf area to aboveground biomass.
-We found only modest effects of climate after accounting for PFT, suggesting that climate-related differences in biomass partitioning are captured largely by a shift in PFT.
-These results form an important constraint on biomass partitioning routines in global dynamic vegetation models.}
-
-\section*{First paragraph - Daniel}
-
-\textbf{
-The partitioning of plant primary productivity among fleshy leaves and woody stems is a key factor driving carbon cycles and the flow of energy within terrestrial ecosystems.
-Although researchers have long assumed key plant functional types (PFTs) to differ in biomass partitioning, empirical data have been lacking.
-Divergent approaches to modelling allocation is thus a major source of uncertainty in terrestrial ecosystem models (TEMs) and global carbon cycles.
-Using a new database on Biomass partitioning combining data from over 175 studies, we here show a strong effect of PFT on biomass partitioning.
+The partitioning of biomass into short-lived leaves and long-lived stems is one of the most uncertain  and influential components of dynamic global vegetation models (DGVMs), affecting the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, predictions of global carbon storage \cite{ise_comparison_2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
+Although DGVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
+Using a new database on biomass partitioning from over 175 studies, we here show a strong effect of PFT on biomass partitioning.
 Contrary to prior expectation, we show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass.
-Moreover, we find that such differences relate to systematic shifts in leaf-mass per area, a leading trait within the leaf economics spectrum.
-Another core trait, leaf area per sapwood area, explained further variation among species in biomass partitioning, but did not differ systematically among PFTs.
-Combined, these results uncover fundamental differences in how plants are constructed and allow for systematic testing of allocation routines with current TEMs.}
+Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum.
+Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current DGVMs.
 
@@
 
-% 1.
-Importance of allocation in global vegetation models, short-lived leaves vs.
-long-lived stems affect ecosystem-scale turnover of C, as well as productivity (as partitioning to leaves affects LAI).
-DGVMs split plant diversity into a few plant functional types (PFTs).
-This begs two questions : how does biomass partitioning dffer between PFTs, and can we relate it to commonly measured leaf traits, to simplify application of DGVMs?
+\begin{itemize}
 
-Previous syntheses on plant biomass relationships have predominantly focused on the universal nature of size-biomass relationships in plants.
-Relatively little is known about differences in biomass partitioning between PFTs(evergreen vs.
-deciduous, angiosperm vs.
-gymnosperm).
+\item DGVMs are used to predict the trajectory of the global vegetation carbon sink in a changing climate.
+Biomass partitioning (related to, but not to be confused with allocation of net primary production) is important because short-lived leaves re-enter the decomposition cycle quickly, releasing much of its carbon back to the atmosphere.
+Long-lived stems act as a very long carbon storage term (Luyssaert).
+As a result, partitioning affects the overall lifetime of carbon in the ecosystem (Friend).
+\item DGVMs differ widely in how partitioning is calculated (de Kauwe) (though it should be mentioned that partitioning and allocation are not the same!
+- see Methods S2).
+As the global diversity in plants is typically simplified in a handful of plant functional types, it is common to assume that biomass partitioning indeed varies predictably between PFTs, as a basis for model parameterization.
+However, it is largely unknown whether PFTs differ systematically in BP.
+This is largely because an appropriate database has been lacking.
+\item What is known is that gymnosperms have more leaf biomass per unit aboveground biomass (the 'leaf mass fraction', LMF) than angiosperms (Enquist etc.).
+But it is not even clear whether this is due to the angiosperm/evergreen divergence, or related to inherent differences in leaf life span.
+If the latter, we would expect evergreen angiosperms to have more leaf biomass than deciduous angiosperms.
+\item It is also unclear whether PFTs with higher leaf biomass will have higher total leaf area, because PFTs differ in the ratio of leaf mass to area (LMA), a measure of the leaf construction cost.
+Leaf area itself is relevant because light interception acts on leaf area, which drives photosynthesis.
+A few classic studies have assumed that gymnosperms indeed carry much more leaf area than angiosperms (Chabot Hicks, plus another).
+\item Finally it is also unknown whether potential differences between PFTs in BP are a result of inherent differences in morphology, or whether climate has direct effects on BP, and the PFTs simply occupy different climate spaces.
+\item Here we use BAAD, large database blah blah, three main PFTs, many biomes, wide range in plant size, good coverage of climate space where woody plants occur (Fig 1a and S1).
+We included several key variables from BAAD (Fig 1b).
+\item We hypothesize that gymnosperms have higher LMF than angiosperms, and that evergreen angiosperms have higher LMF than deciduous angiosperms (we did not have sufficient data to test deciduous evergreens).
+\item We found that LMF differed strongly between the three PFTs tested, with evergreen gymnosperms > evergreen angiosperms > deciduous angiosperms (Fig 2a).
+Because of allometric relationships between the studied variables (niklas), there was a very strong effect of plant height on LMF (Fig 2a and S2).
+We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships.
+\item We then found that LMF was proportional to the leaf construction cost, as measured by leaf mass per area (Fig 2b).
+\item The consequence of this is that the leaf area ratio (the amount of leaf area supported per unit aboveground biomass), was invariant between PFTs, because $LAR = LMF / LMA$ (Fig 2c and S3).
+We thus supported the view that PFTs differ in LMF, but for the first time demonstrated that they converge in the amount of leaf area supported.
+\item Next, we split LMF into two components, $LMF = M_F/M_T = M_F/A_S * A_S/MT$, where $M_F/A_S$ is the ratio of leaf biomass to stem basal area (see Methods S1), and $A_S/M_T$ the ratio of stem basal area to total aboveground biomass.
+We found that the first term showed identical PFT patterns as LMF (Fig 2d and e), $M_F/A_S$ was highly variable between PFT and proportional to LMA, so that $A_F/A_S$ is invariant between PFTs.
+\item This split is also convenient because the first term is nearly independent of plant height, allowing a more straightforward analysis of climate effects (not having to correct for height every time).
+The second term was found to be nearly unaffected by PFT (S4), but strongly height-dependent.
+It thus absorbs variation related to height and aboveground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species (Chaves).
+\item We analyzed climate effects in two ways, biome, quantitative variables (MAP,MAT,MI,mgdd0)
+\item We did not find differences between biomes in $M_F/A_S$ (Fig S5 - now S7!), but some small inconsistent differences in $A_F/A_S$ (corresponding to slightly lower LMA in tropical vs.
+temperate evergreen angiosperms).
+SHOULD THIS BE FIGURE 4?
+Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height (Table S7 SHOULD BE IN MAIN TEXT).
+\item We found statistically significant but weak and inconsistent effects of climate variables (MAP and MAT) on $M_F/A_S$ and $A_F/A_S$ (Fig 4).
+These variables explained very little variation when added to a statistical model in addition to PFT and plant height (Table 1).
+
+\item We found that PFTs differed in LMF, and this was proportional to leaf mass per area, a leading trait within the leaf economics spectrum.
+Indeed, if LMA was used instead of PFT, we were able to explain more variation in $M_F/A_S$ and $A_F/A_S$ (Table S7).
+It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.
+\item We found very modest effects of climate on biomass partitioning, somewhat inconsistent with Reich et al 2014.
+It is consistent with the idea that LMA is more flexible than biomass partitioning (Poorter).
+We argue that climate directly or indirectly affects LMA, which then leads to differences in LMF, because LAR is less variable.
+Plants seem to target a certain total leaf area display per unit aboveground biomass investment, but we do stress that there is still much variation unaccounted for (Figure S histograms), and that other studies certainly have uncovered small-scale effects of climate on leaf area to sapwood area (pine studies etc), an important component of $A_F/A_S$.
+Our results hold at the PFT level, which is relevant for DGVMs.
+\item What consequences for DGVMs?
+There is a multitude of allocation models (de Kauwe).
+Together with assumptions on leaf and stem turnover rates, this gives estimates of biomass partitioning (S2 methods).
+So far it is common to choose between different models largely based on philosophical arguments (Franklin 2012).
+We argue that relationships such as in Fig 2a should be used to benchmark and constrain the various allocation models.
+The data we use here is freely available (Falster), which should aid in the models being tested, with the hope that they perhaps converge.
+
+\end{itemize}
+
+% Fig 3 to SI.
+% Fig S7 to main text
+
+\subsubsection*{Old text bit still useful}
 
@@
 No study to date has been able to further tease apart PFT differences, or test the potentially confounding effects of climate.
-
-What is lacking is a solid empirical base that can be used to constrain any model of allocation, that includes woody plants of all sizes, of all major plant functional types and biomes.
-Most global efforts to date use the Cannell database of forest biomass, which is not based on individual biomass estimates.
-Here we use a new database (Falster et al.), the largest ever compiled for individual woody plants, including data for major plant functional types (PFTs), and all major biomes.
-
-The main objective was to test whether PFT-specific differences exist in the ratio of leaf mass to total above-ground biomass (the 'leaf mass fraction', LMF), and the leaf area to aboveground biomass (the 'leaf area ratio', LAR).
-Because climate and PFT are correlated, (Fig.~\ref{fig:figure1}), we also test whether climate has additional effects on biomass partitioning, or whether these effects are captured by PFT alone.
-
-We found large differences between PFTs in biomass partitioning between leaves and stems (Fig.~\ref{fig:figure1}b, see also Fig.~\ref{fig:figureS2}).
-After accounting for the strong effects of plant height, we found large differences in LMF between PFTs (Fig.~\ref{fig:figure1}c), and small inconsistent differences in LAR (Fig.~\ref{fig:figure1}d, see also Fig.~\ref{fig:figureS3}).
-This suggests that PFTs do not differ in the amount of leaf area displayed per unit aboveground biomass.
-Since the LMA differs strongly, this implies a difference in the biomass partitioning between leaves and stems.
 
@@
 This relationship also held approximately when data were first averaged by species (Fig.~\ref{fig:figureS6}).
-
-Although PFTs converged in the amount of leaf area displayed per unit aboveground biomass and stem basal area, there was a substantial amount of variation between plants (Fig.~\ref{fig:figure3}).
-We partitioned the variance into PFT, plant height, and biome effects in $M_F/A_S$ using a linear mixed-effects model (see Table~\ref{varpartpipe}).
-We found that plant height and biome explained only \Sexpr{100*signif(table_pipemodel_varpart$r2[2],2)}\%, but when PFT was added to the model, the variance explained increased to \Sexpr{100*signif(table_pipemodel_varpart$r2[3],2)}\%.
-If we used LMA as a predictor instead of PFT, the $R^2$ was similar (\Sexpr{100*signif(table_pipemodel_varpart$r2[4],2)}\%), showing that PFT affected $M_F/A_S$ largely through differences in LMA.
-Consistent with this, PFT, plant height and biome explained very little variance in $A_F/A_S$ (Table~\ref{varpartpipe}).
-
-We also partionined the variance of LMF and LAR (Table~\ref{varpartpipe}).
-When added to a model that included plant height and biome, PFT increased the variance explained in log-transformed LMF from \Sexpr{100*signif(table_LMFLAR_varpart$r2[2],2)} to \Sexpr{100*signif(table_LMFLAR_varpart$r2[3],2)}\%.
-However, LMA was not a good substitute for PFT, as the explained variance decreased to \Sexpr{100*signif(table_LMFLAR_varpart$r2[4],2)}\%.
-For (log-transformed) LAR, PFT and LMA explained very little additional variance.
-
-Although biome was a significant predictor for all variables tested, it consistently explained very little variation (Table~\ref{varpartpipe} and~\ref{varpartlmflar}).
-Pair-wise comparisons between biomes within PFT mostly did not show significant differences for $M_F/A_S$ or $A_F/A_S$.
-The only exception was a higher $A_F/A_S$ for tropical compared to temperate evergreen angiosperms (Fig.~\ref{fig:figureS7}).
 

### 2014-12-14 manuscript.Rnw
Table with co-authors compiles OK again

@@
 baad <- readRDS("data/baad.rds")
-dataauthors <- author_details(baad)
+dataauthors <- author_details(baad, "data/contact.csv")
 @
@@
 Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current DGVMs.
+}
 

### 2014-12-15 manuscript.Rnw
Removed figure1b, figure1c, figureS5
Reshuffle of figures (removed Fig 3, replaced with S7); updated var part tables.

@@
 \usepackage{fixltx2e}
-% \usepackage[T1]{fontenc}
 \usepackage[default,osfigures,scale=0.95]{opensans}
@@
 We included several key variables from BAAD (Fig 1b).
+Crucially, the data are for individuals, which is more relevant to DGVMs than plot-level estimates, because ...
 \item We hypothesize that gymnosperms have higher LMF than angiosperms, and that evergreen angiosperms have higher LMF than deciduous angiosperms (we did not have sufficient data to test deciduous evergreens).
@@
 It thus absorbs variation related to height and aboveground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species (Chaves).
-\item We analyzed climate effects in two ways, biome, quantitative variables (MAP,MAT,MI,mgdd0)
-\item We did not find differences between biomes in $M_F/A_S$ (Fig S5 - now S7!), but some small inconsistent differences in $A_F/A_S$ (corresponding to slightly lower LMA in tropical vs.
+\item We analyzed climate effects in two ways, biome, quantitative variables (MAP,MAT), and by biome (tropical / temperate/ boreal), which is a simple classification taking into account both MAT and MAP.
+\item Regardless of how we analyzed it, we found only weak climate effects.
+We did not find differences between biomes in $M_F/A_S$ (Fig 3), but some small inconsistent differences in $A_F/A_S$ (corresponding to slightly lower LMA in tropical vs.
 temperate evergreen angiosperms).
-SHOULD THIS BE FIGURE 4?
 Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height (Table S7 SHOULD BE IN MAIN TEXT).
@@
 These variables explained very little variation when added to a statistical model in addition to PFT and plant height (Table 1).
-
 \item We found that PFTs differed in LMF, and this was proportional to leaf mass per area, a leading trait within the leaf economics spectrum.
@@
 \end{itemize}
-
-% Fig 3 to SI.
-% Fig S7 to main text
 
@@
 
+\clearpage
 \section{Tables}
@@
 Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."), caption.placement="top")
+@
+
+<<results='asis'>>=
+print(xtable(table_varpart2, label="varpartpipe",
+             caption="Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.
+The R2 shown is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type.
+All fixed effects were highly significant (P \\textless 0.001).
+In addition to the fixed effects shown, all interactions were added to each of the models."),
+      include.rownames=FALSE, caption.placement="top")
 @
@@
     \includegraphics[width=0.99\textwidth]{Figure3.pdf}
-    \caption{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.
-Shown are histograms of leaf mass per area (an index of leaf construction cost) ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types.
-Arrows indicate means (same as in Fig.~\ref{fig:figure2}).}
+    \caption{Mean leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/A_S$) by PFT and major biome.
+No leaf area estimates were available for boreal deciduous angiosperms.
+DA = deciduous angiosperm, EA = evergreen angiosperm, EG = evergreen gymnosperm, Trop.
+= tropical, Temp.
+= temperate.
+Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect.
+Letters denote significant differences (at $\alpha = 0.05$)}
     \label{fig:figure3}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{Figure4b.pdf}
+    \includegraphics[width=0.99\textwidth]{Figure4.pdf}
     \caption{Modest effects of climate on three size-invariant variables, compared to the very large effect of plant functional type on mass-based variables (leaf mass per area, LMA, and $M_F/A_S$), and convergence in area-based variables ($A_F/A_S$).
@@
     \includegraphics[width=0.99\textwidth]{FigureS1.pdf}
-    \caption{Global coverage of the climate space by the dataset. Methods are as in Fig 1 in main text, symbols are colored by vegetation type as reported by the data contributors.}
+    \caption{Global coverage of the climate space by the dataset. Methods are as in Fig. 1 in main text, symbols are colored by vegetation type as reported by the data contributors.}
 \label{fig:figureS1}
@@
     \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
-    \caption{Plant leaf mass as a function of basal stem area (left panel), and leaf area as a function of basal stem area (right panel).
-Each symbol is an individual plant.
-Lines are generalized additive model fits.}
-  \label{fig:figureS5}
+    \caption{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.
+Shown are histograms of leaf mass per area (an index of leaf construction cost) ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types.
+Arrows indicate means (same as in Fig.~\ref{fig:figure2}).}
+    \label{fig:figures5}
 \end{figure}
@@
     \caption{Plant leaf mass as a function of basal stem area (left panel), and leaf area as a function of basal stem area (right panel).
-Individual plant sata were averaged by species/study (the same species occuring in multiple datasets thus appears multiple times).
+Individual plant data were averaged by species/study (the same species occuring in multiple datasets thus appears multiple times).
 Lines are linear regressions with 95\% confidence intervals.
-Both regressions were significant, but R2 was 0.28 in the left panel, and only 0.02 in the right panel.
-The coefficient of variation between species was 0.19 for log-transformed MF/AS, and only 0.099 for AF/AS.}
+Both regressions were significant, but $R^2$ was 0.28 in the left panel, and only 0.02 in the right panel.}
   \label{fig:figureS6}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureS7.pdf}
-    \caption{Mean leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/A_S$) by PFT and major biome.
-No leaf area estimates were available for boreal deciduous angiosperms.
-DA = deciduous angiosperm, EA = evergreen angiosperm, EG = evergreen gymnosperm, Trop.
-= tropical, Temp.
-= temperate.
-Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect.
-Letters denote significant differences (at $\alpha = 0.05$)}
-    \label{fig:figureS7}
 \end{figure}
@@
 @
+
+\clearpage
 
@@
 
-<<results='asis'>>=
-print(xtable(table_pipemodel_varpart, label="varpartpipe",
-             caption="Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.
-The marginal R2 is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type.
-The conditional R2 is that explained by fixed and random effects (species within study).
-All fixed effects were highly significant (P \\textless 0.001).
-In addition to the fixed effects shown, all interactions were added to each of the models."),
-      include.rownames=FALSE, caption.placement="top")
-@
-
-<<results='asis'>>=
-print(xtable(table_LMFLAR_varpart, label="varpartlmflar",
-      caption="Explained variance in the leaf mass fraction (leaf mass / aboveground biomass) and leaf area ratio (leaf area / aboveground biomass).
-See Table~\\ref{varpartpipe} for more explanation.
-For plant height, we also added the squared term as a fixed effect, and the interaction with PFT or LMA, because the relationship was non-linear."),
-      include.rownames=FALSE, caption.placement="top")
-@
-
 \clearpage

### 2014-12-15 R/figures.R
Reshuffle of figures (removed Fig 3, replaced with S7); updated var part tables.

@@ -170,42 +170,40 @@ figure2 <- function(dataset, KGAM=4){
 }
 
 
-# Figure 3. Histograms of MF/AS and AF/AS.
-figure3 <- function(dataset){
-  
-  par(mar=c(0,0,0,2), oma=c(5,5,1,1), las=1, cex.axis=0.85, mfrow=c(1,3), mgp=c(3,1.5,0))
-  
-  histbypft(llma, pft, dataset, xaxis=3,legend.cex=1,col=my_cols_transparent(), 
-            xlab="", overlay=TRUE,plotwhat="density",ylab="Density",
-            Means=mixmean("llma","pft",dataset),cicol=alpha("grey",0.5),
-            legend.text="", meanlinecol=my_linecols())
-  plotlabel("(a)","topleft")
-  
-  histbypft(lmlf_astba2, pft, dataset, xaxis=3,legend.cex=1,col=my_cols_transparent(), 
-            xlab="", overlay=TRUE,plotwhat="density",ylab="Density",
-            Means=mixmean("lmlf_astba2","pft",dataset),cicol=alpha("grey",0.5),
-            legend.text="", meanlinecol=my_linecols())
-  plotlabel("(b)","topleft")
-  
-  histbypft(lalf_astba2, pft, dataset, xaxis=3,legend.cex=1,col=my_cols_transparent(), 
-            xlab="",overlay=TRUE,plotwhat="density",ylab="Density",
-            Means=mixmean("lalf_astba2","pft",dataset),cicol=alpha("grey",0.65),
-            legend.text=rep("",3), meanlinecol=my_linecols())
-  plotlabel("(c)","topleft")
-  
-  mtext(side=1, line=3, text=expression(M[F]/A[F]~~(kg~m^-2)),
-        outer=TRUE, at=1/6, cex=0.9)
-  mtext(side=1, line=3, text=expression(M[F]/A[S]~~(kg~m^-2)),
-        outer=TRUE, at=3/6, cex=0.9)
-  mtext(side=1, line=3, text=expression(A[F]/A[S]~~(m^2~m^-2)),
-        outer=TRUE, at=5/6, cex=0.9)
-}
 
+# Means of leaf mass, area per stem area by PFT - biome combination.
+figure3 <- function(dataset2){
+  
+  Labs <- c(
+    "Boreal DA",
+    "Temp. DA",
+    "Trop. DA",
+    "Temp. EA",
+    "Trop. EA",
+    "Boreal EG",
[…]

### 2014-12-15 R/tables_stats.R
Updated Table S1, replaces S1 and S2 - much more consistent with Table 1 (which has also been updated a bit)

@@ -13,10 +13,11 @@ make_table_pipemodel_varpart <- function(dataset2) {
   assign("dat_alf", droplevels(subset(dataset2, !is.na(h.t) & !is.na(pft) & !is.na(lalf_astba2))), envir = .GlobalEnv)
 
   mlf_formulas <- list()
+  mlf_formulas[["PFT"]]       <- lmlf_astba2 ~ pft + (1|Group)
   mlf_formulas[["H"]]         <- lmlf_astba2 ~ log10(h.t) + (1|Group)
-  mlf_formulas[["H, B"]]      <- lmlf_astba2 ~ log10(h.t)*bortemptrop + (1|Group)
-  mlf_formulas[["PFT, H, B"]] <- lmlf_astba2 ~ log10(h.t)*pft*bortemptrop + (1|Group)
-  mlf_formulas[["LMA, H, B"]] <- lmlf_astba2 ~ log10(h.t)*lsla*bortemptrop + (1|Group)
+  mlf_formulas[["H, PFT"]]    <- lmlf_astba2 ~ log10(h.t)*pft + (1|Group)
+  mlf_formulas[["H, PFT, B"]] <- lmlf_astba2 ~ log10(h.t)*pft*bortemptrop + (1|Group)
+  mlf_formulas[["H, LMA, B"]] <- lmlf_astba2 ~ log10(h.t)*llma*bortemptrop + (1|Group)
 
   mlf_lmer_fits <- lapply(mlf_formulas, function(x) lmer(x, data=dat_mlf))
   mlf_lmer_r2 <- lapply( mlf_lmer_fits, r.squared.merMod)
@@ -34,7 +35,7 @@ make_table_pipemodel_varpart <- function(dataset2) {
 
 make_table_LMFLAR_varpart <- function(dataset2) {
 
-assign("dat_mlf", droplevels(subset(dataset2, !is.na(h.t) & !is.na(pft) & !is.na(lmlf_astba2))), envir = .GlobalEnv)
+  assign("dat_mlf", droplevels(subset(dataset2, !is.na(h.t) & !is.na(pft) & !is.na(lmlf_astba2))), envir = .GlobalEnv)
   assign("dat_alf", droplevels(subset(dataset2, !is.na(h.t) & !is.na(pft) & !is.na(lalf_astba2))), envir = .GlobalEnv)
 
[…]

### 2014-12-15 manuscript.Rnw
Updated Table S1, replaces S1 and S2 - much more consistent with Table 1 (which has also been updated a bit)

@@
 \author[2]{Daniel S. Falster}
-\author[3]{BAAD contributors}
+\author[3]{the BAAD consortium}
 
@@
 <<results='asis'>>=
-print(xtable(table_gamr2MIgdd0, label="table/gamr2table",
-      caption="Explained variance by plant functional type (PFT), plant height (H), moisture index (MI) and growing season average temperature (mgdd0), for six whole-plant variables.
-Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."), caption.placement="top")
-@
-
-<<results='asis'>>=
-print(xtable(table_varpart2, label="varpartpipe",
-             caption="Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.
-The R2 shown is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type.
-All fixed effects were highly significant (P \\textless 0.001).
-In addition to the fixed effects shown, all interactions were added to each of the models."),
+print(xtable(table_gamr2MATMAP, label="table/gamr2table",
+      caption="Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.
+Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),
       include.rownames=FALSE, caption.placement="top")
@@
 print(xtable(dataauthors,
-             caption="Contributors to the BAAD"
+             caption="Co-authors in the BAAD consortium."
              ), include.rownames=FALSE, tabular.environment='longtable', floating=FALSE, size="small",
@@
 
+<<results='asis'>>=
+print(xtable(table_varpart2, label="varpart2",
+             caption="Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.
+The R2 shown is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type.
+All fixed effects were highly significant (P \\textless 0.001).
+In addition to the fixed effects shown, all interactions were added to each of the models."),
+      include.rownames=FALSE, caption.placement="top")
+@
+
 \clearpage

### 2014-12-15 manuscript.Rnw
Added Methods S1, explaining allocation vs. partitioning

@@
 
-\affil[1]{1 Hawkesbury Insitute for the Environment, University of Western Sydney, Hawkesbury Campus, First Floor, Building L9, Locked Bag 1797, Penrith NSW 2751, Australia}
-\affil[2]{Biological Sciences, Macquarie University NSW 2109, Australia}
+\affil[1]{1 Hawkesbury Insitute for the Environment, University of Western Sydney, Hawkesbury Campus, First Floor, Building L9, Locked Bag 1797, Penrith, NSW 2751, Australia}
+\affil[2]{Biological Sciences, Macquarie University, North Ryde, NSW 2109, Australia}
 \affil[3]{See Supporting Information}
@@
 \textbf{
-The partitioning of biomass into short-lived leaves and long-lived stems is one of the most uncertain  and influential components of dynamic global vegetation models (DGVMs), affecting the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, predictions of global carbon storage \cite{ise_comparison_2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
-Although DGVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
-Using a new database on biomass partitioning from over 175 studies, we here show a strong effect of PFT on biomass partitioning.
+The partitioning of biomass into short-lived leaves and long-lived stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, predictions of global carbon storage \cite{ise_comparison_2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
+Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
+Using a new global database on biomass partitioning from over 175 studies, we here show a strong effect of PFT on partitioning between stems and leaves.
 Contrary to prior expectation, we show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass.
 Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum.
-Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current DGVMs.
+Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
 }
@@
 
-\item DGVMs are used to predict the trajectory of the global vegetation carbon sink in a changing climate.
+\item GVMs (including dynamic GVMs, which model the distribution of vegetation types, and those who use a prescribed vegetation distributions) are used to predict the trajectory of the global vegetation carbon sink in a changing climate.
 Biomass partitioning (related to, but not to be confused with allocation of net primary production) is important because short-lived leaves re-enter the decomposition cycle quickly, releasing much of its carbon back to the atmosphere.
 Long-lived stems act as a very long carbon storage term (Luyssaert).
-As a result, partitioning affects the overall lifetime of carbon in the ecosystem (Friend).
+As a result, partitioning affects the overall lifetime of carbon in the ecosystem, which has been shown to be an important determinant of overall carbon storage projections(Friend).
 \item DGVMs differ widely in how partitioning is calculated (de Kauwe) (though it should be mentioned that partitioning and allocation are not the same!
@@
 
-\subsection*{Predicting basal diameter}
+\subsection*{Methods S1 - Biomass partitioning terminology}
+\label{sec:partterm}
+
+The terms 'partitioning' and 'allocation' have been used in various ways, confusing comparisons between studies \cite{litton_carbon_2007}.
+Here, we define biomass partitioning as the actual distribution of biomass between compartments (e.g., leaves vs.
+stems), and biomass allocation as the proportion of net primary production (NPP) that is allocated to some compartment.
+The two processes are different because of continuous turnover of biomass, which differs strongly between compartments.
+We may write (cf.
+McMurtrie and Wolf 1983),
+
+$$
+\frac{dM_i}{dt} = a_i NPP - \tau_i M_i
+$$
+
+where $M_i$ is the biomass in some compartment (leaves, stems or roots), $a_i$ the annual allocation of NPP to that compartment, and $\tau_i$ the annual turnover of compartment $i$.
+It is thus easy to see that partitioning ($M_i/M_T$, where $M_T$ is total biomass) can be different from allocation ($a_i$) because turnover ($\tau_i$) differs strongly between compartments.
+
+\subsection*{Methods S2 - Predicting basal diameter}
 \label{sec:predas}

### 2014-12-15 R/figures.R
Added Methods S1, explaining allocation vs. partitioning

@@ -13,9 +13,6 @@ my_linecols <- function() {
   c("deepskyblue3","firebrick2","chartreuse3")
 }
 
-#palette(Cols)
-
-
 # Plot settings.
 lmaLabel <- function() {
   expression("Leaf mass per area"~~(kg~m^-2))

### 2014-12-15 manuscript.Rnw
Proper formatting of variables in tables

@@
 Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),
-      include.rownames=FALSE, caption.placement="top")
+      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
 @
@@
 In addition to the fixed effects shown, all interactions were added to each of the models."),
-      include.rownames=FALSE, caption.placement="top")
+      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
 @

### 2014-12-16 manuscript.Rnw
Added a couple of notes, two tables next to each other for comparison

@@
 \geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
+
 \bibliographystyle{plain}
@@
 This is largely because an appropriate database has been lacking.
+\item It is our intent to evaluate and critique alternative models of biomass allocation (wbe etc), but rather to establish, empirically, patterns in biomass partitioning across plant functional types and plant size.
 \item What is known is that gymnosperms have more leaf biomass per unit aboveground biomass (the 'leaf mass fraction', LMF) than angiosperms (Enquist etc.).
@@
 If the latter, we would expect evergreen angiosperms to have more leaf biomass than deciduous angiosperms.
-\item It is also unclear whether PFTs with higher leaf biomass will have higher total leaf area, because PFTs differ in the ratio of leaf mass to area (LMA), a measure of the leaf construction cost.
+\item Novelty in this study, a) individual plants, b) entire size range including saplings and the largest trees in the world, c) estimates of leaf area as well as mass d) further division into three plant functional types, and three biomes
+\item It is unclear whether PFTs with higher leaf biomass will have higher total leaf area, because PFTs differ in the ratio of leaf mass to area (LMA), a measure of the leaf construction cost.
 Leaf area itself is relevant because light interception acts on leaf area, which drives photosynthesis.
-A few classic studies have assumed that gymnosperms indeed carry much more leaf area than angiosperms (Chabot Hicks, plus another).
+A few classic studies have assumed that gymnosperms indeed carry much more leaf area than angiosperms (Bond; Chabot and Hicks).
 \item Finally it is also unknown whether potential differences between PFTs in BP are a result of inherent differences in morphology, or whether climate has direct effects on BP, and the PFTs simply occupy different climate spaces.
@@
 temperate evergreen angiosperms).
-Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height (Table S7 SHOULD BE IN MAIN TEXT).
+Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height (Table S6).
 \item We found statistically significant but weak and inconsistent effects of climate variables (MAP and MAT) on $M_F/A_S$ and $A_F/A_S$ (Fig 4).
@@
 Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),
+      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
+@
+
+<<results='asis'>>=
+print(xtable(table_varpart2, label="varpart2",
+             caption="Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.
+The R2 shown is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type.
+All fixed effects were highly significant (P \\textless 0.001).
+In addition to the fixed effects shown, all interactions were added to each of the models."),
       include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
@@
 
-<<results='asis'>>=
-print(xtable(table_varpart2, label="varpart2",
-             caption="Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.
-The R2 shown is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type.
-All fixed effects were highly significant (P \\textless 0.001).
-In addition to the fixed effects shown, all interactions were added to each of the models."),
-      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
-@
-
 \clearpage

### 2014-12-16 manuscript.Rnw
Fairly major update Figure 3

@@
     \centering
-    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
-    \caption{Mean leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/A_S$) by PFT and major biome.
+    \includegraphics[width=0.6\textwidth]{Figure3.pdf}
+    \caption{No biome differences in leaf mass per unit stem basal stem area for $M_F/A_S$ and weak, inconsistent effects of biome on leaf area per unit basal stem area ($A_F/A_S$).
 No leaf area estimates were available for boreal deciduous angiosperms.
-DA = deciduous angiosperm, EA = evergreen angiosperm, EG = evergreen gymnosperm, Trop.
-= tropical, Temp.
-= temperate.
+DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.
 Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect.
@@
     \includegraphics[width=0.99\textwidth]{Figure4.pdf}
-    \caption{Modest effects of climate on three size-invariant variables, compared to the very large effect of plant functional type on mass-based variables (leaf mass per area, LMA, and $M_F/A_S$), and convergence in area-based variables ($A_F/A_S$).
-See SI Methods for calculation of MI and growing season temperature.
-Solid lines are generalized additive model fits (with a base dimension of 3, larger values resulted in biologically implausible relationships).
-Grey areas are approximate 95\% confidence intervals, calculated by the gam after averaging each variable by species/dataset combination (but all raw individual plant data are shown in the plots).}
+    \caption{Modest effects of climate on three size-invariant variables, compared to the very large effect of plant functional type on mass-based variables ($M_F/A_S$), and convergence in area-based variables ($A_F/A_S$).
+Solid lines are generalized additive model (GAM) fits (with a base dimension of 3, larger values resulted in biologically implausible relationships).
+Grey areas are approximate 95\% confidence intervals, calculated by the GAM after averaging each variable by species/dataset combination.}
   \label{fig:figure4}
@@
 We may write (cf.
-McMurtrie and Wolf 1983),
+\cite{McMurtrieWolf1983}),
 

### 2014-12-16 R/figures.R

@@ -172,22 +172,26 @@ figure2 <- function(dataset, KGAM=4){
 figure3 <- function(dataset2){
   
   Labs <- c(
-    "Boreal DA",
-    "Temp. DA",
-    "Trop. DA",
-    "Temp. EA",
-    "Trop. EA",
-    "Boreal EG",
-    "Temp. EG")
+    "Boreal",
+    "Temperate",
+    "Tropical",
+    "Temperate",
+    "Tropical",
+    "Boreal",
+    "Temperate")
   
-  par(cex.axis=0.85, mfrow=c(1,2), mar=c(8,5,1,1), cex=1.1, las=1, cex.axis=0.9)
+  par(cex.axis=0.85, mfrow=c(2,1), mar=c(0.35,5,7,1), pty="m", cex=1.1, las=1)
+  
   meansbypft("lmlf_astba2","lalf_astba2", "pftlong", panel1only=F,
              xvar=1:7,setpar=FALSE,
              addlegend=FALSE,
              Cols=c("dodgerblue2","blue", "lightskyblue", "red","hotpink2",
                     "forestgreen","chartreuse3"),
-             panel1.exp={axis(1, at=1:7, labels=Labs, las=2, cex.axis=0.8)},
-             panel2.exp={axis(1, at=1:7, labels=Labs, las=2, cex.axis=0.8)},
+             panel1.exp={axis(1, at=1:7, labels=FALSE)
+                         plotlabel("(a)","topleft")
+                         par(mar=c(7,5,0.35,1))},
+             panel2.exp={axis(1, at=1:7, labels=Labs, las=2, cex.axis=0.8)
+                         plotlabel("(b)","topleft")},
              siglets="bottom",
              xlab="",
              axis1=FALSE,
@@ -196,6 +200,26 @@ figure3 <- function(dataset2){
              dataset=dataset2,
              xlim=c(0,8),
              ylim1=c(0,250),ylim2=c(0,2000))
+  mtext(side=1, at=c(2,4.5,6.5), text=c("DA","EA","EG"), line=5)
+  
+  plotBracket <- function(x0, x1, rely=0.3, uprely=0.05){
+      
+    h0 <- grconvertY(0, "ndc","user")
+    usr <- par("usr")
+    marh <- usr[3] - h0
+    y <- h0 + rely * marh
+    upy <- y + uprely * marh
[…]

### 2014-12-16 manuscript.Rnw
caption and labels update figure 4

@@
     \includegraphics[width=0.99\textwidth]{Figure4.pdf}
-    \caption{Modest effects of climate on three size-invariant variables, compared to the very large effect of plant functional type on mass-based variables ($M_F/A_S$), and convergence in area-based variables ($A_F/A_S$).
+    \caption{Modest effects of climate on leaf mass and area per unit stem basal area, compared to the very large effect of plant functional type on mass-based variables ($M_F/A_S$).
 Solid lines are generalized additive model (GAM) fits (with a base dimension of 3, larger values resulted in biologically implausible relationships).
-Grey areas are approximate 95\% confidence intervals, calculated by the GAM after averaging each variable by species/dataset combination.}
+Grey areas are approximate 95\% confidence intervals, calculated by the GAM.
+To account for within-species and contributed dataset correlations, data were first averaged by species/dataset combination.}
   \label{fig:figure4}

### 2014-12-16 R/figures.R

@@ -229,10 +229,10 @@ clim_figure <- function(dataset, KGAM=3, climvar1="MI", climvar2="mgdd0", fitone
   gcol <- alpha("lightgrey",0.5)
   
   if(climvar1 == "MI")Xlab1 <- "Moisture Index (-)"
-  if(climvar1 == "MAP")Xlab1 <- "Mean annual precipitation (mm)"
+  if(climvar1 == "MAP")Xlab1 <- "MAP (mm)"
   
   if(climvar2 == "mgdd0")Xlab2 <- expression(Growing~season~T~(degree*C))
-  if(climvar2 == "MAT")Xlab2 <- expression(Mean~annual~temperature~~(degree*C))
+  if(climvar2 == "MAT")Xlab2 <- expression(MAT~(degree*C))
   
   dataset$XVAR1 <- dataset[,climvar1]
   dataset$XVAR2 <- dataset[,climvar2]
@@ -251,11 +251,11 @@ clim_figure <- function(dataset, KGAM=3, climvar1="MI", climvar2="mgdd0", fitone
              ylab=expression(M[F]/A[S]~~(kg~m^-2)))
   smoothplot(XVAR1, lalf_astba2, pft, dataset, log="y", kgam=KGAM, R=R, randommethod = "agg", fittype=fittype,
              xlab=Xlab1,polycolor=gcol,pointcols=my_cols_transparent(),linecols=my_linecols(),
-             fitoneline=fitoneline,
+             fitoneline=fitoneline,ylim=c(2,4),
              ylab=expression(A[F]/A[S]~~(m^2~m^-2)))
   smoothplot(XVAR2, lalf_astba2, pft, dataset, log="y", kgam=KGAM, R=R, randommethod = "agg", fittype=fittype,
              xlab=Xlab2,polycolor=gcol,pointcols=my_cols_transparent(),linecols=my_linecols(),
-             fitoneline=fitoneline,
+             fitoneline=fitoneline,ylim=c(2,4),
              ylab=expression(A[F]/A[S]~~(m^2~m^-2)))
 }
 

### 2014-12-16 manuscript.Rnw
Updated Figure 4

@@
     \includegraphics[width=0.99\textwidth]{Figure4.pdf}
-    \caption{Modest effects of climate on leaf mass and area per unit stem basal area, compared to the very large effect of plant functional type on mass-based variables ($M_F/A_S$).
+    \caption{Modest effects of climate (mean annual temperature, MAT, and mean annual precipitation, MAP) on leaf mass and area per unit stem basal area, compared to the very large effect of plant functional type on mass-based variables ($M_F/A_S$).
 Solid lines are generalized additive model (GAM) fits (with a base dimension of 3, larger values resulted in biologically implausible relationships).

### 2014-12-16 R/functions-figures.R

@@ -319,6 +319,7 @@ smoothplot <- function(x,y,g=NULL,data,
                             kgam=4,
                             R=NULL,
                             randommethod=c("lmer","aggregate"),
+                            axes=TRUE,
                             log="xy",
                             fitoneline=FALSE,
                             pointcols=NULL,
@@ -384,24 +385,26 @@ smoothplot <- function(x,y,g=NULL,data,
   if(plotit){
     with(data, plot(X, Y, axes=FALSE, pch=16, col=pointcols[G],
                     xlab=xlab, ylab=ylab, ...))
-
-    if(log=="xy")magaxis(side=1:2, unlog=1:2)
-    if(log=="x"){
-      magaxis(side=1, unlog=1)
-      axis(2)
-      box()
+  
+    if(axes){
+      if(log=="xy")magaxis(side=1:2, unlog=1:2)
+      if(log=="x"){
+        magaxis(side=1, unlog=1)
+        axis(2)
+        box()
+      }
+      if(log=="y"){
+        magaxis(side=2, unlog=2)
+        axis(1)
+        box()
+      }
+      if(log==""){
+        axis(1)
+        axis(2)
+        box()
+      }
     }
-    if(log=="y"){
-      magaxis(side=2, unlog=2)
-      axis(1)
-      box()
-    }
-    if(log==""){
-      axis(1)
-      axis(2)
-      box()
-    }
-
+    
     for(i in 1:length(fits)){
 
       if(fittype == "gam"){

### 2014-12-16 manuscript.Rnw
Add tree to Figure 1. Closes #8

@@
 
+Figure 1 coutresy of Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary.
+
 \section{Author contributions}
@@
     \includegraphics[width=0.99\textwidth]{Figure1.pdf}
-    \caption{(a) Global coverage of the climate space by the dataset.
-Hexagons are terrestrial observations, excluding areas without woody vegetation, colored by number of 0.5\degree~cells.
-(b) Schematic of a woody plant, and the variables considered in this study.
-H = plant height, $A_S$ basal stem area, $A_{sbh}$ stem area at 'breast height', $M_W$ woody biomass, $M_F$ leaf biomass (total aboveground biomass $M_T = M_F + M_W$), $A_F$ total leaf area.}
+    \caption{\textbf{Overview of the global biomass and allometry database.} (a) Variables were measured on \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species.
+(b) Coverage of the dataset across global climate space.
+Grey hexagons indicate the number of 0.5\degree~cells with woody vegetation across the space.
+Colour symbols show the locations of sampled individuals for three dominat woody functional types.}
     \label{fig:figure1}

### 2014-12-16 R/figures.R

@@ -45,10 +45,65 @@ my_legend <- function(where, labels=c("short","long"), bty='n', rev=TRUE, ...){
 #:: Main figures.
 
 
+figure1 <- function(baad_mapmat, world_mapmat, tree_image){
+
+  par(oma=c(3,1,1,1), mfrow=c(1,2), mar=c(1,1,2,1))
+  plot(1:2, type='n', ann=FALSE, axes=FALSE, xlim=c(-1,1), ylim=c(-1,1))
+  plotlabel("(a)", "topleft", inset.y= -0.08, inset.x = -0.04, xpd=NA)
+  # note we are only plotting to last viewport, but calls to other two needed to make figure work.
+  vps <- baseViewports()
+  pushViewport(vps$inner)
+  pushViewport(vps$figure)
+  pushViewport(vps$plot)
+ # grid.rect(gp=gpar(lwd=3, col="blue"))  # show grid
+  fig.tree(tree_image)
+
+  par(mar=c(1,4,2,1))
+  figureMAPMATworldclim(baad_mapmat, world_mapmat, setpar=FALSE, legend2=TRUE,
+                        meanpoints=FALSE)
+  plotlabel("(b)", "topleft", inset.y= -0.08, inset.x = -0.3, xpd=NA)
+}
+
+fig.tree <- function(filename) {
+
+  x0 <- 0.35
+  gp0=gpar(cex=0.8)
+  img <- readPNG(filename)
+  grid.raster(img, unit(x0, "npc"),  y = unit(0.4, "npc"), just=c("centre"), height=unit(0.80, "npc"))
+
+  # height
+  grid.lines(x = c(0.02, 0.02), y = c(0.0, 0.8),arrow = arrow(ends = "last"), gp=gpar(lwd=2))
+  grid.text("H, height of plant", x = 0.05 , y = 0.85, just="left", gp=gp0)
+
+  # stems areas
+  grid.draw(ellipseGrob(x0 - 0.025, 0.03, size=6,ar=4,angle=0, def="npc"))
+  grid.text(expression(paste(A[S],", stem area at base")),
+    x = x0 + 0.03 , y = 0.03, just="left", gp=gp0)
+
+  grid.draw(ellipseGrob(x0 - 0.03, 0.08, size=4,ar=4,angle=0, def="npc"))
+  grid.text(expression(paste(A[Sbh],", stem area at breast height")),
+    x = x0 +0.023 , y = 0.08, just="left", gp=gp0)
[…]

### 2014-12-16 R/data_processing.R
Add tree to Figure 1. Closes #8

@@ -1,4 +1,10 @@
 
+download_tree_png <- function(destination_filename) {
+  url <-
+    "http://ian.umces.edu/imagelibrary/albums/userpics/12789/normal_ian-symbol-eucalyptus-spp-1.png"
+  download.file(url, destination_filename)
+}
+
 extract_baad_data <- function(baad) {
   baad$data
 }

### 2014-12-16 R/functions-figures.R

@@ -1,11 +1,11 @@
 # Simple function for placing labels on a figure.
-plotlabel <- function(txt, where, inset=0.08, ...){
+plotlabel <- function(txt, where, inset=0.08, inset.x=inset, inset.y=inset,...){
   u <- par()$usr
-  if(grepl("left",where))x <- u[1] + inset*(u[2]-u[1])
-  if(grepl("right",where))x <- u[2] - inset*(u[2]-u[1])
-  if(grepl("bottom",where))y <- u[3] + inset*(u[4]-u[3])
-  if(grepl("top",where))y <- u[4] - inset*(u[4]-u[3])
-  
+  if(grepl("left",where))x <- u[1] + inset.x*(u[2]-u[1])
+  if(grepl("right",where))x <- u[2] - inset.x*(u[2]-u[1])
+  if(grepl("bottom",where))y <- u[3] + inset.y*(u[4]-u[3])
+  if(grepl("top",where))y <- u[4] - inset.y*(u[4]-u[3])
+
   text(x,y,txt,...)
 }
 

### 2014-12-16 R/figures.R
Consistent xlabs, bold first sentence in figures and tables

@@ -269,11 +269,11 @@ clim_figure <- function(dataset, KGAM=3, climvar1="MI", climvar2="mgdd0", fitone
   gcol <- alpha("lightgrey",0.5)
   
   if(climvar1 == "MI")Xlab1 <- "Moisture Index (-)"
-  if(climvar1 == "MAP")Xlab1 <- "MAP (mm)"
+  if(climvar1 == "MAP")Xlab1 <- expression("Mean annual precipitation"~(mm))
   
   if(climvar2 == "mgdd0")Xlab2 <- expression(Growing~season~T~(degree*C))
-  if(climvar2 == "MAT")Xlab2 <- expression(MAT~(degree*C))
-  
+  if(climvar2 == "MAT")Xlab2 <- expression("Mean annual temperature"~(degree*C))
+
   Ylab2 <- expression(M[F]/A[S]~~(kg~m^-2))
   Ylab1 <- expression(A[F]/A[S]~~(m^2~m^-2))
   

### 2014-12-16 manuscript.Rnw
Consistent xlabs, bold first sentence in figures and tables

@@
 print(xtable(table_gamr2MATMAP, label="table/gamr2table",
-      caption="Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.
-Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),
+      caption="\\textbf{Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.} Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),
       include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
@@
 print(xtable(table_varpart2, label="varpart2",
-             caption="Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.
-The R2 shown is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type.
+             caption="\\textbf{Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.} The $r^2$ shown is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type.
 All fixed effects were highly significant (P \\textless 0.001).
@@
     \includegraphics[width=0.99\textwidth]{Figure2.pdf}
-    \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf construction cost.
-(a) Leaf mass fraction (LMF = leaf mass / aboveground biomass) by PFT.
+    \caption{\textbf{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf construction cost.
+} (a) Leaf mass fraction (LMF = leaf mass / aboveground biomass) by PFT.
 Each symbol is an individual plant.
@@
     \includegraphics[width=0.6\textwidth]{Figure3.pdf}
-    \caption{No biome differences in leaf mass per unit stem basal stem area for $M_F/A_S$ and weak, inconsistent effects of biome on leaf area per unit basal stem area ($A_F/A_S$).
+    \caption{\textbf{No biome differences in leaf mass per unit stem basal stem area for $M_F/A_S$ and weak, inconsistent effects of biome on leaf area per unit basal stem area ($A_F/A_S$)}.
 No leaf area estimates were available for boreal deciduous angiosperms.
@@
     \includegraphics[width=0.99\textwidth]{Figure4.pdf}
-    \caption{Modest effects of climate (mean annual temperature, MAT, and mean annual precipitation, MAP) on leaf mass and area per unit stem basal area, compared to the very large effect of plant functional type on mass-based variables ($M_F/A_S$).
+    \caption{\textbf{Modest effects of climate on leaf mass and area per unit stem basal area, compared to the very large effect of plant functional type on mass-based variables ($M_F/A_S$)}.
 Solid lines are generalized additive model (GAM) fits (with a base dimension of 3, larger values resulted in biologically implausible relationships).
@@
     \includegraphics[width=0.99\textwidth]{FigureS1.pdf}
-    \caption{Global coverage of the climate space by the dataset. Methods are as in Fig. 1 in main text, symbols are colored by vegetation type as reported by the data contributors.}
+    \caption{\textbf{Global coverage of the climate space by the dataset.} Methods are as in Fig. 1 in main text, symbols are colored by vegetation type as reported by the data contributors.}
 \label{fig:figureS1}
@@
     \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
-    \caption{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_W$) for each of the PFTs, as a function of total plant height ($H$). Each point is an individual plant.}
+    \caption{\textbf{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_W$) for each of the PFTs, as a function of total plant height ($H$).} Each point is an individual plant.}
     \label{fig:figureS2}
@@
     \includegraphics[width=0.99\textwidth]{FigureS3.pdf}
-    \caption{Leaf area ratio ($A_F/M_T$) by PFT.
-Each symbol is an individual plant.
+    \caption{\textbf{Leaf area ratio ($A_F/M_T$) by PFT.} Each symbol is an individual plant.
 Lines are generalized additive model fits using species/dataset as a random effect.
@@
     \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
-    \caption{Relationship between aboveground biomass and basal stem area, each point is an individual plant.
-Lines are spline fits.
+    \caption{\textbf{Relationship between aboveground biomass and basal stem area, each point is an individual plant.} Lines are spline fits.
 The right panel shows least-square means of the ratio of aboveground biomass to basal stem area, estimated from a linear mixed-effects model with basal stem area (linear and squared terms) and PFT.
@@
     \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
-    \caption{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.
-Shown are histograms of leaf mass per area (an index of leaf construction cost) ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types.
+    \caption{\textbf{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.} Shown are histograms of leaf mass per area (an index of leaf construction cost) ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types.
 Arrows indicate means (same as in Fig.~\ref{fig:figure2}).}
@@
     \includegraphics[width=0.99\textwidth]{FigureS6.pdf}
-    \caption{Plant leaf mass as a function of basal stem area (left panel), and leaf area as a function of basal stem area (right panel).
-Individual plant data were averaged by species/study (the same species occuring in multiple datasets thus appears multiple times).
+    \caption{\textbf{Plant leaf mass as a function of basal stem area (left panel), and leaf area as a function of basal stem area (right panel).} Individual plant data were averaged by species/study (the same species occuring in multiple datasets thus appears multiple times).
 Lines are linear regressions with 95\% confidence intervals.
@@
 print(xtable(dataauthors,
-             caption="Co-authors in the BAAD consortium."
+             caption="\\textbf{Co-authors in the BAAD consortium.}"
              ), include.rownames=FALSE, tabular.environment='longtable', floating=FALSE, size="small",
@@
         label="table:count1",
-        caption="Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$.
-Numbers denote individual plants, with number of species in parentheses.
+        caption="\\textbf{Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$.} Numbers denote individual plants, with number of species in parentheses.
 DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@@
         label="table:count2",
-        caption="Number of observations used in the analysis of $A_F/A_S$.
-Numbers denote individual plants, with number of species in parentheses.
+        caption="\\textbf{Number of observations used in the analysis of $A_F/A_S$.} Numbers denote individual plants, with number of species in parentheses.
 DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@@
         label="table:count3",
-        caption="Number of observations used in the analysis of leaf mass fraction ($M_F/M_T$) and leaf area ratio ($A_F/M_T$).
-Numbers denote individual plants, with number of species in parentheses.
+        caption="\\textbf{Number of observations used in the analysis of leaf mass fraction ($M_F/M_T$) and leaf area ratio ($A_F/M_T$).} Numbers denote individual plants, with number of species in parentheses.
 DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@@
         label="table:count4",
-        caption="Number of observations used in the analysis of leaf area ratio (LAR).
-Numbers denote individual plants, with number of species in parentheses.
+        caption="\\textbf{Number of observations used in the analysis of leaf area ratio (LAR).} Numbers denote individual plants, with number of species in parentheses.
 DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")

### 2014-12-16 manuscript.Rnw
Download BAAD database

@@
 source("R/functions-ms.R")
-baad <- readRDS("data/baad.rds")
+baad <- readRDS("downloads/baad.rds")
 dataauthors <- author_details(baad, "data/contact.csv")

### 2014-12-16 R/data_processing.R

@@ -1,4 +1,10 @@
 
+download_baad <- function(destination_filename) {
+  url <-
+    "https://github.com/dfalster/baad/releases/download/v2-Ecology-check/baad.rds"
+  download.file(url, destination_filename, method="wget")
+}
+
 download_tree_png <- function(destination_filename) {
   url <-
     "http://ian.umces.edu/imagelibrary/albums/userpics/12789/normal_ian-symbol-eucalyptus-spp-1.png"

### 2014-12-16 manuscript.Rnw
Minor ms edits; small glitch in equation in Methods S1

@@
 
-\affil[1]{1 Hawkesbury Insitute for the Environment, University of Western Sydney, Hawkesbury Campus, First Floor, Building L9, Locked Bag 1797, Penrith, NSW 2751, Australia}
+\affil[1]{1 Hawkesbury Insitute for the Environment, University of Western Sydney, Locked Bag 1797, Penrith, NSW 2751, Australia}
 \affil[2]{Biological Sciences, Macquarie University, North Ryde, NSW 2109, Australia}
@@
 Grey hexagons indicate the number of 0.5\degree~cells with woody vegetation across the space.
-Colour symbols show the locations of sampled individuals for three dominat woody functional types.}
+Colour symbols show the locations of sampled individuals for three dominant woody functional types.}
     \label{fig:figure1}
@@
 where $M_i$ is the biomass in some compartment (leaves, stems or roots), $a_i$ the annual allocation of NPP to that compartment, and $\tau_i$ the annual turnover of compartment $i$.
-It is thus easy to see that partitioning ($M_i/M_T$, where $M_T$ is total biomass) can be different from allocation ($a_i$) because turnover ($\tau_i$) differs strongly between compartments.
+It is thus easy to see that partitioning ($M_i/M_T$, where $M_T$ is total biomass) can be different from allocation ($a_i$) because turnover ($\tau_i$) differs between leaf and wood biomass.
 
@@
 $$
-a = c_0H^c_1
+a = c_0H^{c_1}
 $$

### 2014-12-17 manuscript.Rnw
Tidied references.bib

@@
 
-\affil[1]{1 Hawkesbury Insitute for the Environment, University of Western Sydney, Locked Bag 1797, Penrith, NSW 2751, Australia}
+\affil[1]{Hawkesbury Institute for the Environment, University of Western Sydney, Locked Bag 1797, Penrith, NSW 2751, Australia}
 \affil[2]{Biological Sciences, Macquarie University, North Ryde, NSW 2109, Australia}
@@
 \textbf{
-The partitioning of biomass into short-lived leaves and long-lived stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, predictions of global carbon storage \cite{ise_comparison_2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
+The partitioning of biomass into short-lived leaves and long-lived stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the lifetime of carbon in ecosystems \cite{friend2014}, predictions of global carbon storage \cite{ise2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
 Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
@@
 Biomass partitioning (related to, but not to be confused with allocation of net primary production) is important because short-lived leaves re-enter the decomposition cycle quickly, releasing much of its carbon back to the atmosphere.
-Long-lived stems act as a very long carbon storage term (Luyssaert).
-As a result, partitioning affects the overall lifetime of carbon in the ecosystem, which has been shown to be an important determinant of overall carbon storage projections(Friend).
+Long-lived stems act as a very long carbon storage term \cite{luyssaert2008}.
+As a result, partitioning affects the overall lifetime of carbon in the ecosystem, which has been shown to be an important determinant of overall carbon storage projections \cite{friend2014}.
 \item DGVMs differ widely in how partitioning is calculated (de Kauwe) (though it should be mentioned that partitioning and allocation are not the same!
@@
 \item It is our intent to evaluate and critique alternative models of biomass allocation (wbe etc), but rather to establish, empirically, patterns in biomass partitioning across plant functional types and plant size.
-\item What is known is that gymnosperms have more leaf biomass per unit aboveground biomass (the 'leaf mass fraction', LMF) than angiosperms (Enquist etc.).
+\item What is known is that gymnosperms have more leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF) than angiosperms \cite{oneilldeangelis1981,enquistniklas2002}.
 But it is not even clear whether this is due to the angiosperm/evergreen divergence, or related to inherent differences in leaf life span.
@@
 \item We then found that LMF was proportional to the leaf construction cost, as measured by leaf mass per area (Fig 2b).
-\item The consequence of this is that the leaf area ratio (the amount of leaf area supported per unit aboveground biomass), was invariant between PFTs, because $LAR = LMF / LMA$ (Fig 2c and S3).
+\item The consequence of this is that the leaf area ratio (the amount of leaf area supported per unit above-ground biomass), was invariant between PFTs, because $LAR = LMF / LMA$ (Fig 2c and S3).
 We thus supported the view that PFTs differ in LMF, but for the first time demonstrated that they converge in the amount of leaf area supported.
-\item Next, we split LMF into two components, $LMF = M_F/M_T = M_F/A_S * A_S/MT$, where $M_F/A_S$ is the ratio of leaf biomass to stem basal area (see Methods S1), and $A_S/M_T$ the ratio of stem basal area to total aboveground biomass.
+\item Next, we split LMF into two components, $LMF = M_F/M_T = M_F/A_S * A_S/MT$, where $M_F/A_S$ is the ratio of leaf biomass to stem basal area (see Methods S1), and $A_S/M_T$ the ratio of stem basal area to total above-ground biomass.
 We found that the first term showed identical PFT patterns as LMF (Fig 2d and e), $M_F/A_S$ was highly variable between PFT and proportional to LMA, so that $A_F/A_S$ is invariant between PFTs.
@@
 The second term was found to be nearly unaffected by PFT (S4), but strongly height-dependent.
-It thus absorbs variation related to height and aboveground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species (Chaves).
+It thus absorbs variation related to height and above-ground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species (Chaves).
 \item We analyzed climate effects in two ways, biome, quantitative variables (MAP,MAT), and by biome (tropical / temperate/ boreal), which is a simple classification taking into account both MAT and MAP.
@@
 We argue that climate directly or indirectly affects LMA, which then leads to differences in LMF, because LAR is less variable.
-Plants seem to target a certain total leaf area display per unit aboveground biomass investment, but we do stress that there is still much variation unaccounted for (Figure S histograms), and that other studies certainly have uncovered small-scale effects of climate on leaf area to sapwood area (pine studies etc), an important component of $A_F/A_S$.
+Plants seem to target a certain total leaf area display per unit above-ground biomass investment, but we do stress that there is still much variation unaccounted for (Figure S histograms), and that other studies certainly have uncovered small-scale effects of climate on leaf area to sapwood area (pine studies etc), an important component of $A_F/A_S$.
 Our results hold at the PFT level, which is relevant for DGVMs.
@@
 Together with assumptions on leaf and stem turnover rates, this gives estimates of biomass partitioning (S2 methods).
-So far it is common to choose between different models largely based on philosophical arguments (Franklin 2012).
+So far it is common to choose between different models largely based on philosophical arguments \cite{franklin2012}.
 We argue that relationships such as in Fig 2a should be used to benchmark and constrain the various allocation models.
@@
 
-Major plant functional types differ widely in leaf mass per area (LMA, the ratio of leaf mass to area, an index of leaf thickness) (\cite{poorter_causes_2009}), which can be expected to affect biomass partitioning in several ways.
+Major plant functional types differ widely in leaf mass per area (LMA, the ratio of leaf mass to area, an index of leaf thickness) (\cite{poorter2009}), which can be expected to affect biomass partitioning in several ways.
 First, LMA correlates with leaf lifespan.
-Longer lived leaves would lead to higher plant leaf mass ($M_F$) as long as annual production rates are similar (\cite{enquist_global_2002}).
+Longer lived leaves would lead to higher plant leaf mass ($M_F$) as long as annual production rates are similar (\cite{enquistniklas2002}).
 Another interpretation is that plants may target a certain leaf area display given a certain biomass investment in supporting tissues, because light interception and photosynthesis correlate with the exchange area of the plant.
 
-It has been reported numerous times that gymnosperms (with higher LMA) have higher leaf mass fraction than angiosperms (with lower LMA) \cite{enquist_global_2002, cheng_relationships_2007, mccarthy_consistency_2007, poorter_biomass_2012, sardans_tree_2013}, although a recent study reported no difference (\cite{wolf_allometric_2010}).
+It has been reported numerous times that gymnosperms (with higher LMA) have higher leaf mass fraction than angiosperms (with lower LMA) \cite{enquistniklas2002, cheng2007a, mccarthyetal2007, poorter2012, sardans2013}, although a recent study reported no difference (\cite{wolf2010}).
 No study to date has been able to further tease apart PFT differences, or test the potentially confounding effects of climate.
 
+It is customary to estimate total above-ground biomass from stem diameter \cite{jenkins2003,chave2005}.
+
 We can further decompose LMF into two components representing different aspects of plant function and structure (see Methods).
-The ratio of leaf biomass ($M_F$) to stem basal area ($A_S$) relates to hydraulic supply, and the ratio of total aboveground biomass ($M_T$) and $A_S$ is a function of plant shape (stem taper, plant height at given stem diameter), and wood density \cite{chave_tree_2005}.
-We found that ratio of aboveground biomass to basal stem area ($M_T/A_S$) was nearly constant between PFTs (Fig.~\ref{fig:figureS4}), with deciduous angiosperms having slightly higher $M_T$ at a given $A_S$.
+The ratio of leaf biomass ($M_F$) to stem basal area ($A_S$) relates to hydraulic supply, and the ratio of total above-ground biomass ($M_T$) and $A_S$ is a function of plant shape (stem taper, plant height at given stem diameter), and wood density \cite{chave2005}.
+We found that ratio of above-ground biomass to basal stem area ($M_T/A_S$) was nearly constant between PFTs (Fig.~\ref{fig:figureS4}), with deciduous angiosperms having slightly higher $M_T$ at a given $A_S$.
 However we found very large differences between PFTs in the ratio $M_F/A_S$(Fig.~\ref{fig:figure2}).
@@
 
-Unlike \cite{reich_temperature_2014}, we found only very weak effects of climate on biomass relationships.
-The variance explained by MAT was very small, especially compared to the large difference between PFTs.
+Unlike \cite{reich2014}, we found only very weak effects of climate on biomass relationships. The variance explained by MAT was very small, especially compared to the large difference between PFTs.
 
-Leaf mass per area generally shows a more plastic response to climate than biomass partitioning  \cite{poorter_biomass_2012}.
+Leaf mass per area generally shows a more plastic response to climate than biomass partitioning  \cite{poorter2012}.
 
@@
 
-There is no shortage of theoretical developments in understanding biomass allocation (\cite{cannell_carbon_1994,franklin_modeling_2012,enquist_land_2012}).
+There is no shortage of theoretical developments in understanding biomass allocation (\cite{cannelldewar1994,franklin2012,enquist2012}).
 In particular, theoretical work focusing on optimal partitioning theory is promising (e.g.
-\cite{valentine_modeling_2012}) but is yet to lead to operational models for dynamic global vegetation models.
+\cite{valentinemakela2012}) but is yet to lead to operational models for dynamic global vegetation models.
 
@@
 This means that climate-related differences are captured by a shift in PFT (see also Fig.~\ref{fig:figure1}, showing PFTs occupy different parts of the climate space).
-Our results contrast with those of \cite{reich_temperature_2014} but they studied whole-stand biomass, not individual plant biomass.
+Our results contrast with those of \cite{reich2014} but they studied whole-stand biomass, not individual plant biomass.
 So perhaps this means the climate effects are on stand density not individual plant effects.
@@
 
-We used the recently compiled Biomass And Allometry Database (BAAD) (Falster et al., in review), including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), aboveground woody biomass ($M_W$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
+We used the recently compiled Biomass And Allometry Database (BAAD) (Falster et al., in review), including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_W$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
 1.3m) ($A_{Sbh}$).
 For each analysis, we used different subsets of the data.
-Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4} We exluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
-We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and decidous angiosperms.
+Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4} We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
+We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms.
 We excluded  deciduous gymnosperms because few data were available.
@@
 
-For the aboveground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total aboveground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,
+For the above-ground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total above-ground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,
 
@@
 
-We further decomposed LMF and LAR by considering the ratio of leaf mass (or area) to plant basal area, and the ratio of aboveground biomass to plant basal area. For LMF,
+We further decomposed LMF and LAR by considering the ratio of leaf mass (or area) to plant basal area, and the ratio of above-ground biomass to plant basal area. For LMF,
 
@@
 We were unable to further decompose $A_S/M_T$ into wood volume and wood density components (cf.
-\cite{chave_tree_2005}) because wood density data were available only for a small subset of data.
+\cite{chave2005}) because wood density data were available only for a small subset of data.
 
 We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves).
-For conifers, leaf area was converted to half-total surface area, which is most appropriate for comparison to flat leaves in terms of light interception  \cite{lang_application_1991, chen_defining_1992}.
+For conifers, leaf area was converted to half-total surface area, which is most appropriate for comparison to flat leaves in terms of light interception  \cite{lang1991, chen1992}.
 
@@
 All variables were log-transformed prior to analysis.
-Variance partitioning with marginal (fixed effects) and conditional (fixed and random effects) $R^2$ for linear mixed-effects models as reported by \cite{nakagawa_general_2013}.
+Variance partitioning with marginal (fixed effects) and conditional (fixed and random effects) $R^2$ for linear mixed-effects models as reported by \cite{nakagawa2013}.
 
@@
 
-Figure 1 coutresy of Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary.
+Figure 1a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary.
 
@@
 
-R.A.D and D.S.F. conceived the ideas and methods, and co-wrote the paper. R.A.D. analysed data. BAAD contributors shared data.
+R.A.D and D.S.F. conceived the ideas and methods, and co-wrote the paper. R.A.D. analysed data. Co-authors in the BAAD consortium shared data.
 
@@
     \caption{\textbf{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf construction cost.
-} (a) Leaf mass fraction (LMF = leaf mass / aboveground biomass) by PFT.
+} (a) Leaf mass fraction (LMF = leaf mass / above-ground biomass) by PFT.
 Each symbol is an individual plant.
@@
     \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
-    \caption{\textbf{Relationship between aboveground biomass and basal stem area, each point is an individual plant.} Lines are spline fits.
-The right panel shows least-square means of the ratio of aboveground biomass to basal stem area, estimated from a linear mixed-effects model with basal stem area (linear and squared terms) and PFT.
+    \caption{\textbf{Relationship between above-ground biomass and basal stem area, each point is an individual plant.} Lines are spline fits.
+The right panel shows least-square means of the ratio of above-ground biomass to basal stem area, estimated from a linear mixed-effects model with basal stem area (linear and squared terms) and PFT.
 }
@@
     \includegraphics[width=0.99\textwidth]{FigureS6.pdf}
-    \caption{\textbf{Plant leaf mass as a function of basal stem area (left panel), and leaf area as a function of basal stem area (right panel).} Individual plant data were averaged by species/study (the same species occuring in multiple datasets thus appears multiple times).
+    \caption{\textbf{Plant leaf mass as a function of basal stem area (left panel), and leaf area as a function of basal stem area (right panel).} Individual plant data were averaged by species/study (the same species occurring in multiple datasets thus appears multiple times).
 Lines are linear regressions with 95\% confidence intervals.

### 2015-01-06 manuscript.Rnw
Much writing. This is a full draft!

@@
 \documentclass[a4paper]{article}
+
+%texcount manuscript.tex > wordcount.txt
 
@@
 
-\textbf{
-The partitioning of biomass into short-lived leaves and long-lived stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the lifetime of carbon in ecosystems \cite{friend2014}, predictions of global carbon storage \cite{ise2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
+\textbf{The partitioning of biomass into short-lived leaves and long-lived stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the lifetime of carbon in ecosystems \cite{friend2014}, predictions of global carbon storage \cite{ise2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
 Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
@@
 
-\begin{itemize}
+Globally, forests store ca.
+360 Pg C in live biomass \cite{panetal2011}, or about 40 years of current antropogenic carbon emissions (Quere).
+The partioning of this biomass among plant components is a major determinant of vegetation productivity and the rate of carbon cycling.
+Biomass stored in stems has a long residence time \cite{luyssaert2008},  thus representing a significant long-term carbon storage term.
+Leaf biomass turns over quickly, and is a major input to the soil carbon cycle where the majority of carbon is released back to the atmosphere.
+Recently, the residence time of carbon in vegetation has been identified as a major source of uncertainty in model projections of effects of climate change on the global carbon sink \cite{friend2014}.
+There is little consensus in these global vegetation models (GVMs) on how biomass partitioning should be modelled \cite{dekauwe2014}, but there is no shortage of theory underling the complex process of biomass allocation \cite{franklin2012} (we use allocation to mean the flux of carbon into plant components, and partitioning the actual distribution of biomass among components, see Methods S1).
 
-\item GVMs (including dynamic GVMs, which model the distribution of vegetation types, and those who use a prescribed vegetation distributions) are used to predict the trajectory of the global vegetation carbon sink in a changing climate.
-Biomass partitioning (related to, but not to be confused with allocation of net primary production) is important because short-lived leaves re-enter the decomposition cycle quickly, releasing much of its carbon back to the atmosphere.
-Long-lived stems act as a very long carbon storage term \cite{luyssaert2008}.
-As a result, partitioning affects the overall lifetime of carbon in the ecosystem, which has been shown to be an important determinant of overall carbon storage projections \cite{friend2014}.
-\item DGVMs differ widely in how partitioning is calculated (de Kauwe) (though it should be mentioned that partitioning and allocation are not the same!
-- see Methods S2).
-As the global diversity in plants is typically simplified in a handful of plant functional types, it is common to assume that biomass partitioning indeed varies predictably between PFTs, as a basis for model parameterization.
-However, it is largely unknown whether PFTs differ systematically in BP.
-This is largely because an appropriate database has been lacking.
-\item It is our intent to evaluate and critique alternative models of biomass allocation (wbe etc), but rather to establish, empirically, patterns in biomass partitioning across plant functional types and plant size.
-\item What is known is that gymnosperms have more leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF) than angiosperms \cite{oneilldeangelis1981,enquistniklas2002}.
-But it is not even clear whether this is due to the angiosperm/evergreen divergence, or related to inherent differences in leaf life span.
-If the latter, we would expect evergreen angiosperms to have more leaf biomass than deciduous angiosperms.
-\item Novelty in this study, a) individual plants, b) entire size range including saplings and the largest trees in the world, c) estimates of leaf area as well as mass d) further division into three plant functional types, and three biomes
-\item It is unclear whether PFTs with higher leaf biomass will have higher total leaf area, because PFTs differ in the ratio of leaf mass to area (LMA), a measure of the leaf construction cost.
+It is difficult to decide which partioning models and parameterisations are superior, since GVMs are normally tested against productivity or carbon storage data, and both are the outcome of multiple processes.
+The vast majority of GVMs simplify parameterization by considering a few plant functional types (PFTs).
+It is also common to assume that biomass partitioning varies between PFTs, however, parameterization of these differences have been based on little empirical evidence.
+Based on previous compilations of forest biomass estimates, it is known that gymnosperms have more leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF) than angiosperms \cite{oneilldeangelis1981,enquistniklas2002,reich2014}.
+However, it is not known whether this is due to the angiosperm/evergreen divergence, or related to inherent differences in leaf life span.
+If the latter is true, we would expect evergreen angiosperms to have more leaf biomass than deciduous angiosperms.
+
+We argue that the lack of a large global database on biomass partitioning at the individual plant level has hampered progress in this field.
+Here, we use such a database to lay the empirical base against which models can be benchmarked.
+Our new database (BAAD, Falster) includes over 20,000 observations of biomass and size metrics for individual plants, covering the entire size range including saplings and the largest trees in the world, three major woody PFTs, and three biomes (tropical, temperate, and boreal forests).
+Another unique aspect is that the database contains leaf mass as well as area estimates, previous compilations only considered mass.
+It is unclear whether PFTs with higher leaf biomass will have higher total leaf area, because PFTs differ in the ratio of leaf mass to area (LMA), a measure of the leaf construction cost \cite{poorter2009}.
 Leaf area itself is relevant because light interception acts on leaf area, which drives photosynthesis.
-A few classic studies have assumed that gymnosperms indeed carry much more leaf area than angiosperms (Bond; Chabot and Hicks).
-\item Finally it is also unknown whether potential differences between PFTs in BP are a result of inherent differences in morphology, or whether climate has direct effects on BP, and the PFTs simply occupy different climate spaces.
-\item Here we use BAAD, large database blah blah, three main PFTs, many biomes, wide range in plant size, good coverage of climate space where woody plants occur (Fig 1a and S1).
-We included several key variables from BAAD (Fig 1b).
-Crucially, the data are for individuals, which is more relevant to DGVMs than plot-level estimates, because ...
-\item We hypothesize that gymnosperms have higher LMF than angiosperms, and that evergreen angiosperms have higher LMF than deciduous angiosperms (we did not have sufficient data to test deciduous evergreens).
-\item We found that LMF differed strongly between the three PFTs tested, with evergreen gymnosperms > evergreen angiosperms > deciduous angiosperms (Fig 2a).
-Because of allometric relationships between the studied variables (niklas), there was a very strong effect of plant height on LMF (Fig 2a and S2).
+A few classic studies have assumed that gymnosperms indeed carry much more leaf area than angiosperms \cite{chabothicks1982, bond1989}.
+Finally it is also unknown whether potential differences between PFTs in biomass partitioning are a result of inherent differences in morphology, or whether climate directly affects paritioning, and the PFTs simply occupy different climate spaces.
+
+We hypothesize that gymnosperms have higher LMF than angiosperms, and that evergreen angiosperms have higher LMF than deciduous angiosperms (we did not have sufficient data to test deciduous evergreens).
+We further test whether differences between PFTs are related to LMA.
+We use several variables from BAAD (Fig.
+1a) to test this main hypothesis.
+We also test whether climate has direct influence on biomass partitioning.
+This is possible because the database has very good global coverage of the climate space (Fig.
+1b).
+
+We found that LMF differed strongly between the three PFTs tested, with evergreen gymnosperms > evergreen angiosperms > deciduous angiosperms (Fig.
+2a).
+Because of allometric relationships between the studied variables \cite{niklas1994}, there was a very strong effect of plant height on LMF (Fig 2a and S2).
 We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships.
-\item We then found that LMF was proportional to the leaf construction cost, as measured by leaf mass per area (Fig 2b).
-\item The consequence of this is that the leaf area ratio (the amount of leaf area supported per unit above-ground biomass), was invariant between PFTs, because $LAR = LMF / LMA$ (Fig 2c and S3).
+We then found that LMF at a common plant height was proportional to the leaf construction cost, as measured by leaf mass per area (Fig 2b).
+The consequence of this is that the leaf area ratio (the amount of leaf area supported per unit above-ground biomass), was invariant between PFTs, because $LAR = LMF / LMA$ (Fig 2c and S3).
 We thus supported the view that PFTs differ in LMF, but for the first time demonstrated that they converge in the amount of leaf area supported.
-\item Next, we split LMF into two components, $LMF = M_F/M_T = M_F/A_S * A_S/MT$, where $M_F/A_S$ is the ratio of leaf biomass to stem basal area (see Methods S1), and $A_S/M_T$ the ratio of stem basal area to total above-ground biomass.
+
+Next, we decomposed LMF into two components, $LMF = M_F/M_T = M_F/A_S * A_S/MT$, where $M_F/A_S$ is the ratio of leaf biomass to stem basal area (see Methods S2), and $A_S/M_T$ the ratio of stem basal area to total above-ground biomass.
 We found that the first term showed identical PFT patterns as LMF (Fig 2d and e), $M_F/A_S$ was highly variable between PFT and proportional to LMA, so that $A_F/A_S$ is invariant between PFTs.
-\item This split is also convenient because the first term is nearly independent of plant height, allowing a more straightforward analysis of climate effects (not having to correct for height every time).
-The second term was found to be nearly unaffected by PFT (S4), but strongly height-dependent.
-It thus absorbs variation related to height and above-ground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species (Chaves).
-\item We analyzed climate effects in two ways, biome, quantitative variables (MAP,MAT), and by biome (tropical / temperate/ boreal), which is a simple classification taking into account both MAT and MAP.
-\item Regardless of how we analyzed it, we found only weak climate effects.
-We did not find differences between biomes in $M_F/A_S$ (Fig 3), but some small inconsistent differences in $A_F/A_S$ (corresponding to slightly lower LMA in tropical vs.
+This decomposition is also convenient because the first term is nearly independent of plant height, allowing a more straightforward analysis of climate effects (not having to correct for height every time).
+The second term was found to be nearly unaffected by PFT (Fig.
+S4), but strongly height-dependent.
+It thus absorbs variation related to height and above-ground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species \cite{chave2005}.
+
+We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation (MAP), mean annual temperature (MAT)), and by biome (tropical / temperate/ boreal), which is a simple classification taking into account both MAT and MAP.
+Regardless of how we analyzed it, we found only weak climate effects.
+We did not find differences between biomes in $M_F/A_S$ (Fig.
+3), but some small inconsistent differences in $A_F/A_S$ (corresponding to slightly lower LMA in tropical vs.
 temperate evergreen angiosperms).
 Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height (Table S6).
-\item We found statistically significant but weak and inconsistent effects of climate variables (MAP and MAT) on $M_F/A_S$ and $A_F/A_S$ (Fig 4).
+Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) on $M_F/A_S$ and $A_F/A_S$ (Fig 4).
 These variables explained very little variation when added to a statistical model in addition to PFT and plant height (Table 1).
-\item We found that PFTs differed in LMF, and this was proportional to leaf mass per area, a leading trait within the leaf economics spectrum.
+
+We found that PFTs differed in LMF, and this was proportional to leaf mass per area (LMA), a leading trait within the leaf economics spectrum.
 Indeed, if LMA was used instead of PFT, we were able to explain more variation in $M_F/A_S$ and $A_F/A_S$ (Table S7).
 It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.
-\item We found very modest effects of climate on biomass partitioning, somewhat inconsistent with Reich et al 2014.
-It is consistent with the idea that LMA is more flexible than biomass partitioning (Poorter).
-We argue that climate directly or indirectly affects LMA, which then leads to differences in LMF, because LAR is less variable.
-Plants seem to target a certain total leaf area display per unit above-ground biomass investment, but we do stress that there is still much variation unaccounted for (Figure S histograms), and that other studies certainly have uncovered small-scale effects of climate on leaf area to sapwood area (pine studies etc), an important component of $A_F/A_S$.
-Our results hold at the PFT level, which is relevant for DGVMs.
-\item What consequences for DGVMs?
-There is a multitude of allocation models (de Kauwe).
-Together with assumptions on leaf and stem turnover rates, this gives estimates of biomass partitioning (S2 methods).
-So far it is common to choose between different models largely based on philosophical arguments \cite{franklin2012}.
-We argue that relationships such as in Fig 2a should be used to benchmark and constrain the various allocation models.
-The data we use here is freely available (Falster), which should aid in the models being tested, with the hope that they perhaps converge.
-
-\end{itemize}
-
-\subsubsection*{Old text bit still useful}
-
-Major plant functional types differ widely in leaf mass per area (LMA, the ratio of leaf mass to area, an index of leaf thickness) (\cite{poorter2009}), which can be expected to affect biomass partitioning in several ways.
-First, LMA correlates with leaf lifespan.
-Longer lived leaves would lead to higher plant leaf mass ($M_F$) as long as annual production rates are similar (\cite{enquistniklas2002}).
+One explanation for this proportionality may be the correlation between LMA and leaf lifespan.
+If annual leaf production is constant, longer leaf lifespan will simply result in the accumulation of more leaf cohorts, and thus biomass \cite{enquistniklas2002}.
+However, leaf lifespan is correlated to but not proportional to LMA \cite{wrightetal2004}.
 Another interpretation is that plants may target a certain leaf area display given a certain biomass investment in supporting tissues, because light interception and photosynthesis correlate with the exchange area of the plant.
 
-It has been reported numerous times that gymnosperms (with higher LMA) have higher leaf mass fraction than angiosperms (with lower LMA) \cite{enquistniklas2002, cheng2007a, mccarthyetal2007, poorter2012, sardans2013}, although a recent study reported no difference (\cite{wolf2010}).
-No study to date has been able to further tease apart PFT differences, or test the potentially confounding effects of climate.
+We found very only weak and inconsistent effects of climate on biomass partitioning, in contrast to a recent study where LMF was found to be correlated with MAT \cite{reich2014}.
+However, they studied whole-stand biomass, not individual plant biomass.
+This begs the question whether climate effects on stand-level biomass partitioning are on stand density or individual plants.
+Our finding that climate does not appreciably affect biomass partitioning is good news for models that parameterize the globe by PFT alone.
+We do stress that there is still much variation between species and studies unaccounted for (Fig.
+S5), and that other studies certainly have uncovered small-scale effects of climate on leaf area to sapwood area \cite{callaway1994}, an important component of $A_F/A_S$.
+Coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning.
 
-It is customary to estimate total above-ground biomass from stem diameter \cite{jenkins2003,chave2005}.
-
-We can further decompose LMF into two components representing different aspects of plant function and structure (see Methods).
-The ratio of leaf biomass ($M_F$) to stem basal area ($A_S$) relates to hydraulic supply, and the ratio of total above-ground biomass ($M_T$) and $A_S$ is a function of plant shape (stem taper, plant height at given stem diameter), and wood density \cite{chave2005}.
-We found that ratio of above-ground biomass to basal stem area ($M_T/A_S$) was nearly constant between PFTs (Fig.~\ref{fig:figureS4}), with deciduous angiosperms having slightly higher $M_T$ at a given $A_S$.
-However we found very large differences between PFTs in the ratio $M_F/A_S$(Fig.~\ref{fig:figure2}).
-
-We then expressed the average $M_F/A_S$ as a function of the leaf mass per area (LMA = $M_F / A_F$), a measure of the leaf construction cost to the plant.
-We found that, across PFTs, $M_F/A_S$ was approximately proportional to the average leaf mass per area for the PFT.
-As a result of this approximate proportionality, the ratio of leaf area to basal stem area ($A_F/A_S$) was not different between PFTs (Fig.~\ref{fig:figure2}) (see also Fig.~\ref{fig:figureS5}).
-This relationship also held approximately when data were first averaged by species (Fig.~\ref{fig:figureS6}).
-
-Unlike \cite{reich2014}, we found only very weak effects of climate on biomass relationships. The variance explained by MAT was very small, especially compared to the large difference between PFTs.
-
-Leaf mass per area generally shows a more plastic response to climate than biomass partitioning  \cite{poorter2012}.
-
-In summary, we found coordination between an important leaf trait and whole-plant biomass partitioning. This has implications for global vegetation models.
-
-How is allocation implemented in current DGVMs?
-Either constant allocation fraction, target allometric equations, or optimization \cite{dekauwe2014}.
-Constant allocation fraction gives constant LAR by pft if leaf lifespan is exactly proportional to LMA.
-Target allometric equations can be made pft-invariant by using allometry of leaf area, not leaf mass.
-Optimization models probably already give the right answer because normally LAI is optimized (Woodward).
-
-A recent study compared allocation routines in a number of leading ecosystem models \cite{dekauwe2014}, and recommended to constrain allocation by observed biomass fractions instead of using constant allocation fractions, which is common practice.
-As a first approximation, which is useful for large scale DGVMs, models should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly.
-Specific leaf area then gives the ratio of leaf biomass to stem biomass.
-
-There is no shortage of theoretical developments in understanding biomass allocation (\cite{cannelldewar1994,franklin2012,enquist2012}).
-In particular, theoretical work focusing on optimal partitioning theory is promising (e.g.
-\cite{valentinemakela2012}) but is yet to lead to operational models for dynamic global vegetation models.
-
-We found no effects of MAP on biomass partitioning, and only very weak effects of MAT on LMF, when analyzed in addition to PFT.
-This means that climate-related differences are captured by a shift in PFT (see also Fig.~\ref{fig:figure1}, showing PFTs occupy different parts of the climate space).
-Our results contrast with those of \cite{reich2014} but they studied whole-stand biomass, not individual plant biomass.
-So perhaps this means the climate effects are on stand density not individual plant effects.
-No effect of climate is good news for models that parameterize the globe by PFT, as this seems sufficient, based on our dataset.
+A recent study compared allocation routines in a number of leading ecosystem models \cite{dekauwe2014}, and recommended to constrain allocation by observed biomass fractions instead of using constant allocation fractions.
+We argue that these models should be constrained by patterns with plant size and PFT presented in our study.
+As a first approximation, which is useful for GVMs, models should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly.
+The leaf mass per area, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem biomass.
+The data we use here is freely available (Falster), which should aid in the models being tested, with the hope that they perhaps converge.
 
@@
 
-We used the recently compiled Biomass And Allometry Database (BAAD) (Falster et al., in review), including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_W$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
+We used the recently compiled Biomass And Allometry Database (BAAD) (Falster), including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_W$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
 1.3m) ($A_{Sbh}$).
@@
 All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes.
+To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \cite{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \cite{lathamglobal}).
 
@@
 We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves).
-For conifers, leaf area was converted to half-total surface area, which is most appropriate for comparison to flat leaves in terms of light interception  \cite{lang1991, chen1992}.
-
+For conifers, leaf area was converted to half-total surface area using published conversion factors \cite{barclay2000}, which is most appropriate for comparison to flat leaves in terms of light interception  \cite{lang1991, chen1992}.
 Stem cross-sectional area was measured either at breast height (ca.
@@
 For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using non-linear regression (see~\ref{sec:predas}).
-The main results were similar when we repeated it for data where either $A_S$ or $A_{Sbh}$ were directly measured (not shown).
 
-We used a generalized additive model (GAM) to visualize the relationships between biomass and plant size variables.
-In addition, we corrected for plant size by fitting linear mixed-effects models (with species within study as the random effect) using plant height (linear and squared terms), and calculating least-square means with the 'lsmeans' package in R.
+We used generalized additive models (GAM) to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height.
+In all fitted GAMs, we used species-dataset combination as a random effect.
+We used the \emph{mgcv} package in R \cite{wood2006}.
 All variables were log-transformed prior to analysis.
-Variance partitioning with marginal (fixed effects) and conditional (fixed and random effects) $R^2$ for linear mixed-effects models as reported by \cite{nakagawa2013}.
 
-To test for direct climate effects in addition to the PFT effect, we used moisture index (rainfall / potential evapotranspiration) and growing season temperature (average air temperature during the period of the year when air temperature is larger than 0) as estimated by Lin et al.
-(in review).
-To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \cite{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \cite{lathamglobal}).
+Variance explained by quantitative climate variables (MAP and MAT) were tested with GAMs where variables were sequentially added to the model, and the explained variance ($R^2$) calculated.
+For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect.
+We calculated the $R^2$ for linear mixed-effects models as reported by \cite{nakagawa2013} for the fixed effects only.
 

### 2015-01-06 manuscript.Rnw
Text polishing / more refs / minor glitch figure S6.

@@
 Globally, forests store ca.
-360 Pg C in live biomass \cite{panetal2011}, or about 40 years of current antropogenic carbon emissions (Quere).
-The partioning of this biomass among plant components is a major determinant of vegetation productivity and the rate of carbon cycling.
-Biomass stored in stems has a long residence time \cite{luyssaert2008},  thus representing a significant long-term carbon storage term.
+360 Pg C in live biomass \cite{panetal2011}, or about 40 years of current athropogenic emissions \cite{friedlingsteinetal2014}.
+The partitioning of this biomass among plant components is a major determinant of vegetation productivity and the rate of carbon cycling.
+Biomass stored in woody stems has a long residence time \cite{luyssaert2008},  thus representing a significant long-term carbon storage term.
 Leaf biomass turns over quickly, and is a major input to the soil carbon cycle where the majority of carbon is released back to the atmosphere.
 Recently, the residence time of carbon in vegetation has been identified as a major source of uncertainty in model projections of effects of climate change on the global carbon sink \cite{friend2014}.
-There is little consensus in these global vegetation models (GVMs) on how biomass partitioning should be modelled \cite{dekauwe2014}, but there is no shortage of theory underling the complex process of biomass allocation \cite{franklin2012} (we use allocation to mean the flux of carbon into plant components, and partitioning the actual distribution of biomass among components, see Methods S1).
+There is little consensus in these global vegetation models (GVMs) on how biomass partitioning should be modelled \cite{franklin2012,dekauwe2014}, but there is no shortage of underlying theory \cite{cannelldewar1994,franklin2012}.
+We stress that biomass partitioning, the actual distribution of biomass among components, is distinct from biomass allocation, the flux of carbon into plant components.
+Biomass partitioning is estimated from allocation by including the rates of turnover of the various plant components (see Methods S1).
 
-It is difficult to decide which partioning models and parameterisations are superior, since GVMs are normally tested against productivity or carbon storage data, and both are the outcome of multiple processes.
-The vast majority of GVMs simplify parameterization by considering a few plant functional types (PFTs).
-It is also common to assume that biomass partitioning varies between PFTs, however, parameterization of these differences have been based on little empirical evidence.
+It is difficult to decide which partitioning models are superior, since GVMs are usually only tested against productivity or carbon storage data, and both are the outcome of multiple processes.
+The vast majority of GVMs simplify parameterization by considering a few plant functional types (PFTs) \cite{harrisonetal2010}.
+It is also common to assume that parameters affection biomass partitioning varies between PFTs, however, parameterization of these differences have been based on little empirical evidence.
 Based on previous compilations of forest biomass estimates, it is known that gymnosperms have more leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF) than angiosperms \cite{oneilldeangelis1981,enquistniklas2002,reich2014}.
-However, it is not known whether this is due to the angiosperm/evergreen divergence, or related to inherent differences in leaf life span.
+However, it is not known whether this is due to the angiosperm/evergreen divergence, or to inherent differences in leaf life span.
 If the latter is true, we would expect evergreen angiosperms to have more leaf biomass than deciduous angiosperms.
@@
 We argue that the lack of a large global database on biomass partitioning at the individual plant level has hampered progress in this field.
-Here, we use such a database to lay the empirical base against which models can be benchmarked.
-Our new database (BAAD, Falster) includes over 20,000 observations of biomass and size metrics for individual plants, covering the entire size range including saplings and the largest trees in the world, three major woody PFTs, and three biomes (tropical, temperate, and boreal forests).
-Another unique aspect is that the database contains leaf mass as well as area estimates, previous compilations only considered mass.
-It is unclear whether PFTs with higher leaf biomass will have higher total leaf area, because PFTs differ in the ratio of leaf mass to area (LMA), a measure of the leaf construction cost \cite{poorter2009}.
+Here, we use a unique database to lay the empirical base against which models can be benchmarked.
+Our new database (BAAD, Falster) includes over 20,000 observations contributed by 175 studies of biomass and size metrics for individual plants, covering the entire size range including saplings and the largest trees in the world, three major woody PFTs, and three biomes (tropical, temperate, and boreal forests).
+Another unique aspect is that the database contains leaf mass as well as leaf area estimates, whereas previous large compilations only considered leaf mass.
+It is unclear whether PFTs with higher leaf biomass will also have higher total leaf area, because PFTs differ in the ratio of leaf mass to area (LMA), a measure of the leaf construction cost \cite{poorter2009}.
 Leaf area itself is relevant because light interception acts on leaf area, which drives photosynthesis.
 A few classic studies have assumed that gymnosperms indeed carry much more leaf area than angiosperms \cite{chabothicks1982, bond1989}.
-Finally it is also unknown whether potential differences between PFTs in biomass partitioning are a result of inherent differences in morphology, or whether climate directly affects paritioning, and the PFTs simply occupy different climate spaces.
+Finally it is also unknown whether potential differences between PFTs in biomass partitioning are a result of inherent differences in morphology, or whether climate directly affects partitioning, and the PFTs simply occupy different climate spaces.
 
-We hypothesize that gymnosperms have higher LMF than angiosperms, and that evergreen angiosperms have higher LMF than deciduous angiosperms (we did not have sufficient data to test deciduous evergreens).
+We hypothesize that gymnosperms have higher LMF than angiosperms, and that evergreen angiosperms have higher LMF than deciduous angiosperms.
 We further test whether differences between PFTs are related to LMA.
@@
 This is possible because the database has very good global coverage of the climate space (Fig.
-1b).
+1b), and includes nine major vegetation types (Fig.
+S1).
 
@@
 Because of allometric relationships between the studied variables \cite{niklas1994}, there was a very strong effect of plant height on LMF (Fig 2a and S2).
-We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships.
-We then found that LMF at a common plant height was proportional to the leaf construction cost, as measured by leaf mass per area (Fig 2b).
+We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \cite{wood2006}.
+We then found that LMF at a common plant height was proportional to the leaf construction cost, as measured by leaf mass per area (Fig.
+2b).
 The consequence of this is that the leaf area ratio (the amount of leaf area supported per unit above-ground biomass), was invariant between PFTs, because $LAR = LMF / LMA$ (Fig 2c and S3).
@@
 We found that the first term showed identical PFT patterns as LMF (Fig 2d and e), $M_F/A_S$ was highly variable between PFT and proportional to LMA, so that $A_F/A_S$ is invariant between PFTs.
-This decomposition is also convenient because the first term is nearly independent of plant height, allowing a more straightforward analysis of climate effects (not having to correct for height every time).
-The second term was found to be nearly unaffected by PFT (Fig.
-S4), but strongly height-dependent.
+This decomposition is also convenient because the first term is nearly independent of plant height (Tables 1 and 2), allowing a more straightforward analysis of climate effects (not having to correct for height every time).
+The second term ($A_S/MT$) was found to be nearly unaffected by PFT, but strongly height-dependent (Fig.
+S4).
 It thus absorbs variation related to height and above-ground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species \cite{chave2005}.
 
-We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation (MAP), mean annual temperature (MAT)), and by biome (tropical / temperate/ boreal), which is a simple classification taking into account both MAT and MAP.
+We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation; MAP, mean annual temperature; MAT), and by biome (tropical / temperate/ boreal), which is a simple classification taking into account both MAT and MAP (Fig.
+S1).
 Regardless of how we analyzed it, we found only weak climate effects.
@@
 temperate evergreen angiosperms).
-Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height (Table S6).
-Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) on $M_F/A_S$ and $A_F/A_S$ (Fig 4).
+Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height (Table 2).
+Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) on $M_F/A_S$ and $A_F/A_S$ (Fig.
+4).
 These variables explained very little variation when added to a statistical model in addition to PFT and plant height (Table 1).
@@
 It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.
-One explanation for this proportionality may be the correlation between LMA and leaf lifespan.
+One explanation for this proportionality may be the correlation between LMA and leaf lifespan \cite{wrightetal2004}.
 If annual leaf production is constant, longer leaf lifespan will simply result in the accumulation of more leaf cohorts, and thus biomass \cite{enquistniklas2002}.
-However, leaf lifespan is correlated to but not proportional to LMA \cite{wrightetal2004}.
-Another interpretation is that plants may target a certain leaf area display given a certain biomass investment in supporting tissues, because light interception and photosynthesis correlate with the exchange area of the plant.
+However, leaf lifespan (LL) is correlated to but not proportional to LMA ($LL \propto LMA^{0.58}$, from \cite{wrightetal2004}.
+Another interpretation is that plants may target a certain leaf area given a certain biomass investment in supporting tissues, because light interception and photosynthesis correlate with the exchange area of the plant.
+We also found that this pattern broadly held across species (rather than across PFTs), the height-invariant variable $M_F/A_S$ was strongly correlated to LMA ($R^2 = 0.28$), whereas $A_F/A_S$ was only weakly ($R^2 = 0.02$) correlated with LMA (Fig.
+S6), demonstrating that leaf area is more conserved between species than leaf mass.
 
@@
 However, they studied whole-stand biomass, not individual plant biomass.
-This begs the question whether climate effects on stand-level biomass partitioning are on stand density or individual plants.
+This begs the question whether climate effects on stand-level biomass partitioning are on stand density (number of individuals per unit area) or individual plants.
 Our finding that climate does not appreciably affect biomass partitioning is good news for models that parameterize the globe by PFT alone.
+However, it is inconsistent with studies demonstrating small-scale effects of climate on leaf area to sapwood area \cite{callaway1994,poyatosetal2007}, an important component of $A_F/A_S$.
 We do stress that there is still much variation between species and studies unaccounted for (Fig.
-S5), and that other studies certainly have uncovered small-scale effects of climate on leaf area to sapwood area \cite{callaway1994}, an important component of $A_F/A_S$.
-Coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning.
+S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning.
+Nonetheless, we can firmly conclude that effects of plant functional type were much stronger than direct climate effects (Tables 1 and 2).
 
@@
 The leaf mass per area, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem biomass.
+An example of such a model is LPJ \cite{sitchetal2003}, where the ratio of leaf area to sapwood area is a constant parameter, but LMA differs between PFTs.
+Models that use constant allocation fractions may also be able to mimic the patterns between PFTs observed here, but only if leaf life span differs in such a way to maintain a proportionality between LMF and LMA.
+In any case, it is relatively straightforward to compare model simulations to broad patterns in LMF with PFT and plant height as presented in this paper.
 The data we use here is freely available (Falster), which should aid in the models being tested, with the hope that they perhaps converge.
@@
 print(xtable(table_varpart2, label="varpart2",
-             caption="\\textbf{Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.} The $r^2$ shown is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type.
+             caption="\\textbf{Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.} Each variable was added to a linear mixed-effects model, using species within dataset as a random effect.
+The $r^2$ shown is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type).
 All fixed effects were highly significant (P \\textless 0.001).
@@
 Lines are linear regressions with 95\% confidence intervals.
-Both regressions were significant, but $R^2$ was 0.28 in the left panel, and only 0.02 in the right panel.}
+Both regressions were significant ($P \textless 0.01$), but $R^2$ was 0.28 in the left panel, and only 0.02 in the right panel.}
   \label{fig:figureS6}

### 2015-01-06 R/figures.R

@@ -497,7 +497,7 @@ figureS6 <- function(dataset){
 
   with(agg, plot(llma, lalf_astba2, pch=19, col=my_cols_transparent()[pft], axes=FALSE, ylim=log10(c(90,10000)),
                  xlim=log10(c(0.01,1)),
-                 xlab=lmaLabel(), ylab=expression(A[F]/A[S]~~(kg~m^-2)),))
+                 xlab=lmaLabel_short(), ylab=expression(A[F]/A[S]~~(kg~m^-2)),))
   magaxis(1:2, unlog=1:2)
   predline(lm2)
   box()

### 2015-01-07 manuscript.Rnw
Edits based on MDK comments

@@
 
-\textbf{The partitioning of biomass into short-lived leaves and long-lived stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the lifetime of carbon in ecosystems \cite{friend2014}, predictions of global carbon storage \cite{ise2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
+\textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time of carbon in ecosystems \cite{friend2014}, predictions of global carbon storage \cite{ise2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
 Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
 Using a new global database on biomass partitioning from over 175 studies, we here show a strong effect of PFT on partitioning between stems and leaves.
-Contrary to prior expectation, we show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass.
-Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum.
+We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass.
+Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum \cite{wrightetal2004}.
 Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
@@
 The partitioning of this biomass among plant components is a major determinant of vegetation productivity and the rate of carbon cycling.
-Biomass stored in woody stems has a long residence time \cite{luyssaert2008},  thus representing a significant long-term carbon storage term.
-Leaf biomass turns over quickly, and is a major input to the soil carbon cycle where the majority of carbon is released back to the atmosphere.
+Biomass stored in woody stems has a long residence time \cite{luyssaert2008},  thus representing a significant long-term carbon storage component.
+By contrast, leaf biomass turns over quickly, and is a major input to the soil carbon cycle where the majority of carbon is released back to the atmosphere.
 Recently, the residence time of carbon in vegetation has been identified as a major source of uncertainty in model projections of effects of climate change on the global carbon sink \cite{friend2014}.
@@
 The vast majority of GVMs simplify parameterization by considering a few plant functional types (PFTs) \cite{harrisonetal2010}.
-It is also common to assume that parameters affection biomass partitioning varies between PFTs, however, parameterization of these differences have been based on little empirical evidence.
+It is also common to assume that parameters related to biomass partitioning varies between PFTs, however, parameterization of these differences have been based on little empirical evidence.
 Based on previous compilations of forest biomass estimates, it is known that gymnosperms have more leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF) than angiosperms \cite{oneilldeangelis1981,enquistniklas2002,reich2014}.
@@
 Here, we use a unique database to lay the empirical base against which models can be benchmarked.
-Our new database (BAAD, Falster) includes over 20,000 observations contributed by 175 studies of biomass and size metrics for individual plants, covering the entire size range including saplings and the largest trees in the world, three major woody PFTs, and three biomes (tropical, temperate, and boreal forests).
-Another unique aspect is that the database contains leaf mass as well as leaf area estimates, whereas previous large compilations only considered leaf mass.
+Our new database (BAAD, Falster) includes over 20,000 observations contributed by 175 studies of biomass and size metrics for individual plants, covering the entire size range including saplings and the largest trees in the world, three major woody PFTs, three biomes (tropical, temperate, and boreal forests), and nine vegetation types (Fig.
+S1).
+Another unique aspect is that the database contains leaf mass as well as leaf area estimates, whereas previous large compilations only considered leaf mass \cite{oneilldeangelis1981,enquistniklas2002,reich2014}.
 It is unclear whether PFTs with higher leaf biomass will also have higher total leaf area, because PFTs differ in the ratio of leaf mass to area (LMA), a measure of the leaf construction cost \cite{poorter2009}.
@@
 1a) to test this main hypothesis.
-We also test whether climate has direct influence on biomass partitioning.
+We also test whether climate has a direct influence on biomass partitioning.
 This is possible because the database has very good global coverage of the climate space (Fig.
-1b), and includes nine major vegetation types (Fig.
-S1).
+1b).
 
@@
 The consequence of this is that the leaf area ratio (the amount of leaf area supported per unit above-ground biomass), was invariant between PFTs, because $LAR = LMF / LMA$ (Fig 2c and S3).
-We thus supported the view that PFTs differ in LMF, but for the first time demonstrated that they converge in the amount of leaf area supported.
+We thus supported the hypothesis that PFTs differ in LMF, and for the first time demonstrated that they converge in the amount of leaf area supported.
 
@@
 S1).
-Regardless of how we analyzed it, we found only weak climate effects.
+Regardless of how we analyzed the data, we found only weak climate effects.
 We did not find differences between biomes in $M_F/A_S$ (Fig.
@@
 temperate evergreen angiosperms).
-Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height (Table 2).
+Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table 2).
 Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) on $M_F/A_S$ and $A_F/A_S$ (Fig.
 4).
-These variables explained very little variation when added to a statistical model in addition to PFT and plant height (Table 1).
+These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table 1).
+The only exception was LMA, where the $R^2$ increased from 0.47 to 0.61 when climate variables were added, consistent with previous reports that LMA shows higher plasticity than biomass partitioning \cite{poorter2012}.
 
 We found that PFTs differed in LMF, and this was proportional to leaf mass per area (LMA), a leading trait within the leaf economics spectrum.
-Indeed, if LMA was used instead of PFT, we were able to explain more variation in $M_F/A_S$ and $A_F/A_S$ (Table S7).
+In addition, LMA alone explained 28\% of the variation in $M_F/A_S$ across species, without accounting for PFT (Fig.
+S6) (and consistent with averages by PFT, $A_F/A_S$ was very weakly related to LMA, with an $R^2$ of 0.02).
 It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.
@@
 Another interpretation is that plants may target a certain leaf area given a certain biomass investment in supporting tissues, because light interception and photosynthesis correlate with the exchange area of the plant.
-We also found that this pattern broadly held across species (rather than across PFTs), the height-invariant variable $M_F/A_S$ was strongly correlated to LMA ($R^2 = 0.28$), whereas $A_F/A_S$ was only weakly ($R^2 = 0.02$) correlated with LMA (Fig.
-S6), demonstrating that leaf area is more conserved between species than leaf mass.
 
 We found very only weak and inconsistent effects of climate on biomass partitioning, in contrast to a recent study where LMF was found to be correlated with MAT \cite{reich2014}.
-However, they studied whole-stand biomass, not individual plant biomass.
+One explanation for these differences is that they studied whole-stand biomass, not individual plant biomass.
 This begs the question whether climate effects on stand-level biomass partitioning are on stand density (number of individuals per unit area) or individual plants.
@@
 
-A recent study compared allocation routines in a number of leading ecosystem models \cite{dekauwe2014}, and recommended to constrain allocation by observed biomass fractions instead of using constant allocation fractions.
+A recent study compared allocation routines in a number of leading ecosystem models \cite{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions.
 We argue that these models should be constrained by patterns with plant size and PFT presented in our study.
-As a first approximation, which is useful for GVMs, models should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly.
+As a first approximation, GVMS should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly.
 The leaf mass per area, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem biomass.
@@
 In any case, it is relatively straightforward to compare model simulations to broad patterns in LMF with PFT and plant height as presented in this paper.
-The data we use here is freely available (Falster), which should aid in the models being tested, with the hope that they perhaps converge.
+The data we use here is freely available (Falster), which should aid model testing.
 
@@
 
-The terms 'partitioning' and 'allocation' have been used in various ways, confusing comparisons between studies \cite{litton_carbon_2007}.
+The terms 'partitioning' and 'allocation' have been used in various ways, confusing comparisons between studies \cite{litton2007}.
 Here, we define biomass partitioning as the actual distribution of biomass between compartments (e.g., leaves vs.

### 2015-01-07 manuscript.Rnw
citation style like nature

@@
 
-\bibliographystyle{plain}
+\usepackage[sort&compress]{natbib}
+\citestyle{nature}
 
@@
 
+\bibliographystyle{unsrtnat}
 \linenumbers
@@
 
-\textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time of carbon in ecosystems \cite{friend2014}, predictions of global carbon storage \cite{ise2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}.
+\textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time of carbon in ecosystems \citep{friend2014}, predictions of global carbon storage \citep{ise2010}, and the CO\textsubscript{2} response of net primary production \citep{dekauwe2014}.
 Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
@@
 We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass.
-Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum \cite{wrightetal2004}.
+Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum \citep{wrightetal2004}.
 Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
@@
 Globally, forests store ca.
-360 Pg C in live biomass \cite{panetal2011}, or about 40 years of current athropogenic emissions \cite{friedlingsteinetal2014}.
+360 Pg C in live biomass \citep{panetal2011}, or about 40 years of current athropogenic emissions \citep{friedlingsteinetal2014}.
 The partitioning of this biomass among plant components is a major determinant of vegetation productivity and the rate of carbon cycling.
-Biomass stored in woody stems has a long residence time \cite{luyssaert2008},  thus representing a significant long-term carbon storage component.
+Biomass stored in woody stems has a long residence time \citep{luyssaert2008},  thus representing a significant long-term carbon storage component.
 By contrast, leaf biomass turns over quickly, and is a major input to the soil carbon cycle where the majority of carbon is released back to the atmosphere.
-Recently, the residence time of carbon in vegetation has been identified as a major source of uncertainty in model projections of effects of climate change on the global carbon sink \cite{friend2014}.
-There is little consensus in these global vegetation models (GVMs) on how biomass partitioning should be modelled \cite{franklin2012,dekauwe2014}, but there is no shortage of underlying theory \cite{cannelldewar1994,franklin2012}.
+Recently, the residence time of carbon in vegetation has been identified as a major source of uncertainty in model projections of effects of climate change on the global carbon sink \citep{friend2014}.
+There is little consensus in these global vegetation models (GVMs) on how biomass partitioning should be modelled \citep{franklin2012,dekauwe2014}, but there is no shortage of underlying theory \citep{cannelldewar1994,franklin2012}.
 We stress that biomass partitioning, the actual distribution of biomass among components, is distinct from biomass allocation, the flux of carbon into plant components.
@@
 It is difficult to decide which partitioning models are superior, since GVMs are usually only tested against productivity or carbon storage data, and both are the outcome of multiple processes.
-The vast majority of GVMs simplify parameterization by considering a few plant functional types (PFTs) \cite{harrisonetal2010}.
+The vast majority of GVMs simplify parameterization by considering a few plant functional types (PFTs) \citep{harrisonetal2010}.
 It is also common to assume that parameters related to biomass partitioning varies between PFTs, however, parameterization of these differences have been based on little empirical evidence.
-Based on previous compilations of forest biomass estimates, it is known that gymnosperms have more leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF) than angiosperms \cite{oneilldeangelis1981,enquistniklas2002,reich2014}.
+Based on previous compilations of forest biomass estimates, it is known that gymnosperms have more leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF) than angiosperms \citep{oneilldeangelis1981,enquistniklas2002,reich2014}.
 However, it is not known whether this is due to the angiosperm/evergreen divergence, or to inherent differences in leaf life span.
@@
 S1).
-Another unique aspect is that the database contains leaf mass as well as leaf area estimates, whereas previous large compilations only considered leaf mass \cite{oneilldeangelis1981,enquistniklas2002,reich2014}.
-It is unclear whether PFTs with higher leaf biomass will also have higher total leaf area, because PFTs differ in the ratio of leaf mass to area (LMA), a measure of the leaf construction cost \cite{poorter2009}.
+Another unique aspect is that the database contains leaf mass as well as leaf area estimates, whereas previous large compilations only considered leaf mass \citep{oneilldeangelis1981,enquistniklas2002,reich2014}.
+It is unclear whether PFTs with higher leaf biomass will also have higher total leaf area, because PFTs differ in the ratio of leaf mass to area (LMA), a measure of the leaf construction cost \citep{poorter2009}.
 Leaf area itself is relevant because light interception acts on leaf area, which drives photosynthesis.
-A few classic studies have assumed that gymnosperms indeed carry much more leaf area than angiosperms \cite{chabothicks1982, bond1989}.
+A few classic studies have assumed that gymnosperms indeed carry much more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
 Finally it is also unknown whether potential differences between PFTs in biomass partitioning are a result of inherent differences in morphology, or whether climate directly affects partitioning, and the PFTs simply occupy different climate spaces.
@@
 2a).
-Because of allometric relationships between the studied variables \cite{niklas1994}, there was a very strong effect of plant height on LMF (Fig 2a and S2).
-We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \cite{wood2006}.
+Because of allometric relationships between the studied variables \citep{niklas1994}, there was a very strong effect of plant height on LMF (Fig 2a and S2).
+We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}.
 We then found that LMF at a common plant height was proportional to the leaf construction cost, as measured by leaf mass per area (Fig.
@@
 S4).
-It thus absorbs variation related to height and above-ground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species \cite{chave2005}.
+It thus absorbs variation related to height and above-ground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species \citep{chave2005}.
 
@@
 These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table 1).
-The only exception was LMA, where the $R^2$ increased from 0.47 to 0.61 when climate variables were added, consistent with previous reports that LMA shows higher plasticity than biomass partitioning \cite{poorter2012}.
+The only exception was LMA, where the $R^2$ increased from 0.47 to 0.61 when climate variables were added, consistent with previous reports that LMA shows higher plasticity than biomass partitioning \citep{poorter2012}.
 
@@
 It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.
-One explanation for this proportionality may be the correlation between LMA and leaf lifespan \cite{wrightetal2004}.
-If annual leaf production is constant, longer leaf lifespan will simply result in the accumulation of more leaf cohorts, and thus biomass \cite{enquistniklas2002}.
-However, leaf lifespan (LL) is correlated to but not proportional to LMA ($LL \propto LMA^{0.58}$, from \cite{wrightetal2004}.
+One explanation for this proportionality may be the correlation between LMA and leaf lifespan \citep{wrightetal2004}.
+If annual leaf production is constant, longer leaf lifespan will simply result in the accumulation of more leaf cohorts, and thus biomass \citep{enquistniklas2002}.
+However, leaf lifespan (LL) is correlated to but not proportional to LMA ($LL \propto LMA^{0.58}$, from Ref.
+\citep{wrightetal2004}).
 Another interpretation is that plants may target a certain leaf area given a certain biomass investment in supporting tissues, because light interception and photosynthesis correlate with the exchange area of the plant.
 
-We found very only weak and inconsistent effects of climate on biomass partitioning, in contrast to a recent study where LMF was found to be correlated with MAT \cite{reich2014}.
+We found very only weak and inconsistent effects of climate on biomass partitioning, in contrast to a recent study where LMF was found to be correlated with MAT \citep{reich2014}.
 One explanation for these differences is that they studied whole-stand biomass, not individual plant biomass.
@@
 Our finding that climate does not appreciably affect biomass partitioning is good news for models that parameterize the globe by PFT alone.
-However, it is inconsistent with studies demonstrating small-scale effects of climate on leaf area to sapwood area \cite{callaway1994,poyatosetal2007}, an important component of $A_F/A_S$.
+However, it is inconsistent with studies demonstrating small-scale effects of climate on leaf area to sapwood area \citep{callaway1994,poyatosetal2007}, an important component of $A_F/A_S$.
 We do stress that there is still much variation between species and studies unaccounted for (Fig.
@@
 
-A recent study compared allocation routines in a number of leading ecosystem models \cite{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions.
+A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions.
 We argue that these models should be constrained by patterns with plant size and PFT presented in our study.
@@
 The leaf mass per area, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem biomass.
-An example of such a model is LPJ \cite{sitchetal2003}, where the ratio of leaf area to sapwood area is a constant parameter, but LMA differs between PFTs.
+An example of such a model is LPJ \citep{sitchetal2003}, where the ratio of leaf area to sapwood area is a constant parameter, but LMA differs between PFTs.
 Models that use constant allocation fractions may also be able to mimic the patterns between PFTs observed here, but only if leaf life span differs in such a way to maintain a proportionality between LMF and LMA.
@@
 All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes.
-To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \cite{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \cite{lathamglobal}).
+To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).
 
@@
 We were unable to further decompose $A_S/M_T$ into wood volume and wood density components (cf.
-\cite{chave2005}) because wood density data were available only for a small subset of data.
+\citep{chave2005}) because wood density data were available only for a small subset of data.
 
 We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves).
-For conifers, leaf area was converted to half-total surface area using published conversion factors \cite{barclay2000}, which is most appropriate for comparison to flat leaves in terms of light interception  \cite{lang1991, chen1992}.
+For conifers, leaf area was converted to half-total surface area using published conversion factors \citep{barclay2000}, which is most appropriate for comparison to flat leaves in terms of light interception  \citep{lang1991, chen1992}.
 Stem cross-sectional area was measured either at breast height (ca.
@@
 In all fitted GAMs, we used species-dataset combination as a random effect.
-We used the \emph{mgcv} package in R \cite{wood2006}.
+We used the \emph{mgcv} package in R \citep{wood2006}.
 All variables were log-transformed prior to analysis.
@@
 For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect.
-We calculated the $R^2$ for linear mixed-effects models as reported by \cite{nakagawa2013} for the fixed effects only.
+We calculated the $R^2$ for linear mixed-effects models as reported by \citep{nakagawa2013} for the fixed effects only.
 
@@
 
-The terms 'partitioning' and 'allocation' have been used in various ways, confusing comparisons between studies \cite{litton2007}.
+The terms 'partitioning' and 'allocation' have been used in various ways, confusing comparisons between studies \citep{litton2007}.
 Here, we define biomass partitioning as the actual distribution of biomass between compartments (e.g., leaves vs.
@@
 We may write (cf.
-\cite{McMurtrieWolf1983}),
+\citep{McMurtrieWolf1983}),
 

### 2015-01-07 manuscript.Rnw
minor fix

@@
 
-Next, we decomposed LMF into two components, $LMF = M_F/M_T = M_F/A_S * A_S/MT$, where $M_F/A_S$ is the ratio of leaf biomass to stem basal area (see Methods S2), and $A_S/M_T$ the ratio of stem basal area to total above-ground biomass.
+Next, we decomposed LMF into two components, $LMF = M_F/M_T = M_F/A_S * A_S/M_T$, where $M_F/A_S$ is the ratio of leaf biomass to stem basal area (see Methods S2), and $A_S/M_T$ the ratio of stem basal area to total above-ground biomass.
 We found that the first term showed identical PFT patterns as LMF (Fig 2d and e), $M_F/A_S$ was highly variable between PFT and proportional to LMA, so that $A_F/A_S$ is invariant between PFTs.

### 2015-01-07 manuscript.Rnw
More tweaking to citations and bibliography.
more citations

@@
 
-\bibliographystyle{unsrtnat}
+\bibliographystyle{ieeetr}
 \linenumbers
@@
 360 Pg C in live biomass \citep{panetal2011}, or about 40 years of current athropogenic emissions \citep{friedlingsteinetal2014}.
-The partitioning of this biomass among plant components is a major determinant of vegetation productivity and the rate of carbon cycling.
+The partitioning of this biomass among plant components is a major determinant of vegetation productivity and the rate of carbon cycling \cite{ise2010}.
 Biomass stored in woody stems has a long residence time \citep{luyssaert2008},  thus representing a significant long-term carbon storage component.
@@
 One explanation for this proportionality may be the correlation between LMA and leaf lifespan \citep{wrightetal2004}.
-If annual leaf production is constant, longer leaf lifespan will simply result in the accumulation of more leaf cohorts, and thus biomass \citep{enquistniklas2002}.
+If annual leaf production is constant, longer leaf lifespan will simply result in the accumulation of more leaf cohorts, and thus biomass \citep{chabothicks1982,enquistniklas2002,poorter2012}.
 However, leaf lifespan (LL) is correlated to but not proportional to LMA ($LL \propto LMA^{0.58}$, from Ref.

### 2015-01-07 manuscript.Rnw
More edits and refs.

@@
 Globally, forests store ca.
-360 Pg C in live biomass \citep{panetal2011}, or about 40 years of current athropogenic emissions \citep{friedlingsteinetal2014}.
+360 Pg C in live biomass \citep{panetal2011}, or about 40 years of current athropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}.
 The partitioning of this biomass among plant components is a major determinant of vegetation productivity and the rate of carbon cycling \cite{ise2010}.
 Biomass stored in woody stems has a long residence time \citep{luyssaert2008},  thus representing a significant long-term carbon storage component.
-By contrast, leaf biomass turns over quickly, and is a major input to the soil carbon cycle where the majority of carbon is released back to the atmosphere.
+By contrast, leaf biomass turns over quickly, and is a major input to the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}.
 Recently, the residence time of carbon in vegetation has been identified as a major source of uncertainty in model projections of effects of climate change on the global carbon sink \citep{friend2014}.
@@
 The vast majority of GVMs simplify parameterization by considering a few plant functional types (PFTs) \citep{harrisonetal2010}.
-It is also common to assume that parameters related to biomass partitioning varies between PFTs, however, parameterization of these differences have been based on little empirical evidence.
+It is also common to assume that parameters related to biomass partitioning vary between PFTs, however, estimatesof these parameters have been based on little empirical evidence.
 Based on previous compilations of forest biomass estimates, it is known that gymnosperms have more leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF) than angiosperms \citep{oneilldeangelis1981,enquistniklas2002,reich2014}.
@@
 Here, we use a unique database to lay the empirical base against which models can be benchmarked.
-Our new database (BAAD, Falster) includes over 20,000 observations contributed by 175 studies of biomass and size metrics for individual plants, covering the entire size range including saplings and the largest trees in the world, three major woody PFTs, three biomes (tropical, temperate, and boreal forests), and nine vegetation types (Fig.
+Our new database (BAAD, Falster) includes over 21,000 observations contributed by 175 studies of biomass and size metrics for individual plants, covering the entire size range including saplings and the largest trees in the world, three major woody PFTs, three biomes (tropical, temperate, and boreal forests), and nine vegetation types (Fig.
 S1).
@@
 We found that the first term showed identical PFT patterns as LMF (Fig 2d and e), $M_F/A_S$ was highly variable between PFT and proportional to LMA, so that $A_F/A_S$ is invariant between PFTs.
-This decomposition is also convenient because the first term is nearly independent of plant height (Tables 1 and 2), allowing a more straightforward analysis of climate effects (not having to correct for height every time).
+This decomposition is also convenient because the first term is nearly independent of plant height (Tables 1 and 2), allowing a more straightforward analysis of climate effects (avoiding the need to correct for height every time).
 The second term ($A_S/MT$) was found to be nearly unaffected by PFT, but strongly height-dependent (Fig.
@@
 
-We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation; MAP, mean annual temperature; MAT), and by biome (tropical / temperate/ boreal), which is a simple classification taking into account both MAT and MAP (Fig.
+We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation; MAP, mean annual temperature; MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig.
 S1).
@@
 A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions.
-We argue that these models should be constrained by patterns with plant size and PFT presented in our study.
+We propose that these models are constrained by patterns with plant size and PFT presented in this study.
 As a first approximation, GVMS should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly.
@@
 Models that use constant allocation fractions may also be able to mimic the patterns between PFTs observed here, but only if leaf life span differs in such a way to maintain a proportionality between LMF and LMA.
-In any case, it is relatively straightforward to compare model simulations to broad patterns in LMF with PFT and plant height as presented in this paper.
+In any case, it is relatively straightforward to compare model simulations to broad patterns in LMF with PFT and plant height as presented here.
 The data we use here is freely available (Falster), which should aid model testing.

### 2015-01-09 manuscript.Rnw
Refine author list
Reworking of main text

--- 2015-01-07 manuscript.Rnw
@@
 \author[2]{Daniel S. Falster}
-\author[3]{the BAAD consortium}
 
@@
 \affil[2]{Biological Sciences, Macquarie University, North Ryde, NSW 2109, Australia}
-\affil[3]{See Supporting Information}
 
@@
 Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum \citep{wrightetal2004}.
+In addition, we show how a second trait, leaf area per stem basal area, contributes additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs.
 Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
@@
 
-Globally, forests store ca.
-360 Pg C in live biomass \citep{panetal2011}, or about 40 years of current athropogenic emissions \citep{friedlingsteinetal2014}.
-The partitioning of this biomass among plant components is a major determinant of vegetation productivity and the rate of carbon cycling \cite{ise2010}.
-Biomass stored in woody stems has a long residence time \citep{luyssaert2008},  thus representing a significant long-term carbon storage component.
-By contrast, leaf biomass turns over quickly, and is a major input to the soil carbon cycle where the majority of carbon is released back to the atmosphere.
-Recently, the residence time of carbon in vegetation has been identified as a major source of uncertainty in model projections of effects of climate change on the global carbon sink \citep{friend2014}.
-There is little consensus in these global vegetation models (GVMs) on how biomass partitioning should be modelled \citep{franklin2012,dekauwe2014}, but there is no shortage of underlying theory \citep{cannelldewar1994,franklin2012}.
-We stress that biomass partitioning, the actual distribution of biomass among components, is distinct from biomass allocation, the flux of carbon into plant components.
-Biomass partitioning is estimated from allocation by including the rates of turnover of the various plant components (see Methods S1).
+The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\cite{ise2010}.
+Globally, forests store approximately 360Pg of carbon\citep{panetal2011}, equivalent to about 40years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}.
+Biomass stored in woody stems is unproductive but has a long residence time \citep{luyssaert2008}, whereas leaf biomass is enhances photosynthesis but turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}.
+Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014}.
 
-It is difficult to decide which partitioning models are superior, since GVMs are usually only tested against productivity or carbon storage data, and both are the outcome of multiple processes.
-The vast majority of GVMs simplify parameterization by considering a few plant functional types (PFTs) \citep{harrisonetal2010}.
-It is also common to assume that parameters related to biomass partitioning varies between PFTs, however, parameterization of these differences have been based on little empirical evidence.
-Based on previous compilations of forest biomass estimates, it is known that gymnosperms have more leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF) than angiosperms \citep{oneilldeangelis1981,enquistniklas2002,reich2014}.
-However, it is not known whether this is due to the angiosperm/evergreen divergence, or to inherent differences in leaf life span.
-If the latter is true, we would expect evergreen angiosperms to have more leaf biomass than deciduous angiosperms.
+Perhaps and the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising world vegetation.
+While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots) and the quality with which these are constructed.
+The approach taken by most GVMs for handling this functional diversity, is to consider only a few archetypal PFTs \citep{harrisonetal2010}, assumed to differ in key physiological attributes.
+While it is common to assume differences in biomass partitioning between PFTs, these estimates are based on limited empirical data.
+Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm}) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
+These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared.
 
-We argue that the lack of a large global database on biomass partitioning at the individual plant level has hampered progress in this field.
-Here, we use a unique database to lay the empirical base against which models can be benchmarked.
-Our new database (BAAD, Falster) includes over 20,000 observations contributed by 175 studies of biomass and size metrics for individual plants, covering the entire size range including saplings and the largest trees in the world, three major woody PFTs, three biomes (tropical, temperate, and boreal forests), and nine vegetation types (Fig.
+Here, we use a unique, new database\citep{Falster2015} (Fig.
+1) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms.
+The dataset includes over 21,000 observations of biomass and size metrics for individual plants, contributed by 175 studies across nine vegetation types (Fig.
 S1).
-Another unique aspect is that the database contains leaf mass as well as leaf area estimates, whereas previous large compilations only considered leaf mass \citep{oneilldeangelis1981,enquistniklas2002,reich2014}.
-It is unclear whether PFTs with higher leaf biomass will also have higher total leaf area, because PFTs differ in the ratio of leaf mass to area (LMA), a measure of the leaf construction cost \citep{poorter2009}.
-Leaf area itself is relevant because light interception acts on leaf area, which drives photosynthesis.
-A few classic studies have assumed that gymnosperms indeed carry much more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
-Finally it is also unknown whether potential differences between PFTs in biomass partitioning are a result of inherent differences in morphology, or whether climate directly affects partitioning, and the PFTs simply occupy different climate spaces.
+Previous work based on whole stands\citep{oneilldeangelis1981,enquistniklas2002,reich2014} and very small plants\citep{poorter2012}, reported differences among angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
+Here we extend this finding to show that LMF differs not only across the angiosperm-gymnosperm divide, but also within angiosperms, at the level of individual plants, and across the entire size range of woody plants in natural forests (0.1-100m) (Fig.
+2a).
 
-We hypothesize that gymnosperms have higher LMF than angiosperms, and that evergreen angiosperms have higher LMF than deciduous angiosperms.
-We further test whether differences between PFTs are related to LMA.
-We use several variables from BAAD (Fig.
-1a) to test this main hypothesis.
-We also test whether climate has a direct influence on biomass partitioning.
-This is possible because the database has very good global coverage of the climate space (Fig.
-1b).
+Moreover, we show that differences in LMF among woody PFTs arise due to differences in leaf morphology, not in the amount of leaf area held on the plant (Fig.
+2b-c).
+Previously, it was unclear whether PFTs with higher leaf biomass also had higher total leaf area.
+Leaf area is relevant because it drives total light interception and thus photosynthesis.
+Several classic studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}, suggesting a growth advantage.
+We found no evidence for this, instead we found that the three PFTS maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR).
+Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig.
+2a and S2).
+We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}.
+We then found that LMF at a common height was proportional to the leaf construction cost, as measured by leaf mass per area (LMA) (Fig.
+2b).
+The consequence of this is that the leaf area ratio (the amount of leaf area supported per unit above-ground biomass), was invariant between PFTs, because $\textrm{LAR} = LMF / \textrm{LMA}$ (Fig.
+2c and S3).
 
-We found that LMF differed strongly between the three PFTs tested, with evergreen gymnosperms > evergreen angiosperms > deciduous angiosperms (Fig.
-2a).
-Because of allometric relationships between the studied variables \citep{niklas1994}, there was a very strong effect of plant height on LMF (Fig 2a and S2).
-We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}.
-We then found that LMF at a common plant height was proportional to the leaf construction cost, as measured by leaf mass per area (Fig.
-2b).
-The consequence of this is that the leaf area ratio (the amount of leaf area supported per unit above-ground biomass), was invariant between PFTs, because $LAR = LMF / LMA$ (Fig 2c and S3).
-We thus supported the hypothesis that PFTs differ in LMF, and for the first time demonstrated that they converge in the amount of leaf area supported.
+Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem; quantifying the amount of light intercepting tissue relative to the amount of tissue available to transport water through the stem.
+This measure has long been assumed to ..(reason why it's important)...
+Mirroring the results for LMF, we find that the amount of leaf mass per unit stem area differs among the three PFTs, that these differences also relate to LMA (Fig.
+2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area (Fig.
+2e).
 
-Next, we decomposed LMF into two components, $LMF = M_F/M_T = M_F/A_S * A_S/M_T$, where $M_F/A_S$ is the ratio of leaf biomass to stem basal area (see Methods S2), and $A_S/M_T$ the ratio of stem basal area to total above-ground biomass.
-We found that the first term showed identical PFT patterns as LMF (Fig 2d and e), $M_F/A_S$ was highly variable between PFT and proportional to LMA, so that $A_F/A_S$ is invariant between PFTs.
-This decomposition is also convenient because the first term is nearly independent of plant height (Tables 1 and 2), allowing a more straightforward analysis of climate effects (not having to correct for height every time).
-The second term ($A_S/MT$) was found to be nearly unaffected by PFT, but strongly height-dependent (Fig.
+The above results highlight how differences in LMA -- a central trait of the leaf economics spectrum\citep{wrightetal2004} -- drives differences in biomass partitioning among dominant woody PFTs.
+Based on a XXX statistical model, we found that LMA alone could explain almost as much  (?more?) variation as PFT, even though LMA only captures aspects of leaf physiology.
+These results suggest promising avenues for parameterizing and simplifying GVMs.
+
+Given these results, we were also interested to know whether additional variation in biomass partitioning could be explained by either climate, within PFT variation in LMA or by other traits.
+Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons among species and across climatic gradients, due to the small sample size within each species or site.
+Thus, to answer these additional questions we found it helpful to decompose LMF as the product,
+\begin{equation}\label{eq:M1}\textrm{LMF} = \left(\frac{M_F}{A_S}\right) \left(\frac{A_S}{M_T}\right).\end{equation}
+An advantage of eq.
+\ref{eq:M1} is that the second term absorbs nearly all of the size related variation in LMF (Tables 1-2) and is otherwise fairly conserved between species\citep{chave2005} and across PFTs (Fig.
 S4).
-It thus absorbs variation related to height and above-ground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species \citep{chave2005}.
+As shown in Fig.
+2b and Fig.
+2d, the remaining term $M_F/A_S$ exhibits comparable differences to LMF, but unlike $M_F/A_S$ is independent of plant height (Tables 1-2).
+$M_F/A_S$ is thus a useful of LMF that can be compared across species and climates.
 
-We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation; MAP, mean annual temperature; MAT), and by biome (tropical / temperate/ boreal), which is a simple classification taking into account both MAT and MAP (Fig.
+We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation, MAP; mean annual temperature, MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig.
 S1).
-Regardless of how we analyzed the data, we found only weak climate effects.
-We did not find differences between biomes in $M_F/A_S$ (Fig.
-3), but some small inconsistent differences in $A_F/A_S$ (corresponding to slightly lower LMA in tropical vs.
-temperate evergreen angiosperms).
+Regardless of how we analyzed the data, we found only weak climatic effects on biomass partitioning within PFTs (Fig.
+3).
 Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table 2).
-Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) on $M_F/A_S$ and $A_F/A_S$ (Fig.
-4).
+Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) (Fig.
+3b-c).
 These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table 1).
-The only exception was LMA, where the $R^2$ increased from 0.47 to 0.61 when climate variables were added, consistent with previous reports that LMA shows higher plasticity than biomass partitioning \citep{poorter2012}.
 
-We found that PFTs differed in LMF, and this was proportional to leaf mass per area (LMA), a leading trait within the leaf economics spectrum.
-In addition, LMA alone explained 28\% of the variation in $M_F/A_S$ across species, without accounting for PFT (Fig.
-S6) (and consistent with averages by PFT, $A_F/A_S$ was very weakly related to LMA, with an $R^2$ of 0.02).
+The finding that climate does not appreciably affect biomass partitioning is good news for models that parameterize the globe by PFT.
+Yet, these results contrast those of a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}.
+One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands.
+This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants.
+With the caveats that there is still variation between species and studies unaccounted for (Fig.
+S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT seem stronger than any direct climate effect on biomass partitioning within individual plants (Tables 1 and 2).
+
+In contrast, we found that variation among species in leaf and stem traits could explain additional variation in biomass partitioning.
+The effect of LMA was found to hold also within PFTs (Fig.
+4a), explaining 28\% of the variation in $M_F/A_S$ across species.
+The fact that PFTs differ in LMA is evident in the separation of the three groups along the x-axis of Fig.
+4a.
+Yet these clouds were broad and overlapping, with a more general relationship extending across the entire LMA axis.
+The remaining (??
+what is number??) variation in $M_F/A_S$  was explained by species-level differences in $A_F/A_S$ (Fig.
+4b).
+As previously noted, $A_F/A_S$ does not differ systematically among PFTs (Fig.
+2e), but does vary close to two orders of magnitude across species.
+The separation among PFTs along the Y-axis arises due to differences in LMA.
+
+Overall, our results establish general patterns about plant construction and thus lay an empirical base against which models can be benchmarked.
+We found that PFTs differed in LMF, and this was proportional to LMA.
 It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.
-One explanation for this proportionality may be the correlation between LMA and leaf lifespan \citep{wrightetal2004}.
-If annual leaf production is constant, longer leaf lifespan will simply result in the accumulation of more leaf cohorts, and thus biomass \citep{chabothicks1982,enquistniklas2002,poorter2012}.
-However, leaf lifespan (LL) is correlated to but not proportional to LMA ($LL \propto LMA^{0.58}$, from Ref.
-\citep{wrightetal2004}).
-Another interpretation is that plants may target a certain leaf area given a certain biomass investment in supporting tissues, because light interception and photosynthesis correlate with the exchange area of the plant.
-
-We found very only weak and inconsistent effects of climate on biomass partitioning, in contrast to a recent study where LMF was found to be correlated with MAT \citep{reich2014}.
-One explanation for these differences is that they studied whole-stand biomass, not individual plant biomass.
-This begs the question whether climate effects on stand-level biomass partitioning are on stand density (number of individuals per unit area) or individual plants.
-Our finding that climate does not appreciably affect biomass partitioning is good news for models that parameterize the globe by PFT alone.
-However, it is inconsistent with studies demonstrating small-scale effects of climate on leaf area to sapwood area \citep{callaway1994,poyatosetal2007}, an important component of $A_F/A_S$.
-We do stress that there is still much variation between species and studies unaccounted for (Fig.
-S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning.
-Nonetheless, we can firmly conclude that effects of plant functional type were much stronger than direct climate effects (Tables 1 and 2).
-
 A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions.
-We argue that these models should be constrained by patterns with plant size and PFT presented in our study.
-As a first approximation, GVMS should aim to keep leaf area to stem cross-sectional area constant, and vary biomass allocation accordingly.
-The leaf mass per area, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem biomass.
+A first approximation within GVMs would be to keep leaf area to stem cross-sectional area constant for a given PFT, and vary biomass allocation accordingly.
+LMA, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem biomass.
 An example of such a model is LPJ \citep{sitchetal2003}, where the ratio of leaf area to sapwood area is a constant parameter, but LMA differs between PFTs.
-Models that use constant allocation fractions may also be able to mimic the patterns between PFTs observed here, but only if leaf life span differs in such a way to maintain a proportionality between LMF and LMA.
-In any case, it is relatively straightforward to compare model simulations to broad patterns in LMF with PFT and plant height as presented in this paper.
-The data we use here is freely available (Falster), which should aid model testing.
+More fundamentally, our results suggest PFTs defined on the basis of two key traits, LMA and $A_F/A_S$ would better capture patterns of biomass partitioning than existing groups, which only account for differences in one of the two traits identified here.
+In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{Falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz_towards_2012, dekauwe2014} of GVMs against empirical data.
 
@@
 
-We used the recently compiled Biomass And Allometry Database (BAAD) (Falster), including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_W$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
+We used the recently compiled Biomass And Allometry Database (BAAD) \citep{Falster2015}, including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_W$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
 1.3m) ($A_{Sbh}$).
@@
 
-$$LMF = M_F / M_T = (M_F / A_F) (A_F / M_T) = LMA * LAR$$
+$$\textrm{LMF} = \frac{M_F}{M_T} = \left(\frac{M_F}{A_F}\right) \left(\frac{A_F}{M_T}\right) = \textrm{LMA} \times \textrm{LAR}.$$
 
@@
 
-$$LMF = M_F / M_T = (M_F / A_S) (M_T / A_S)^{-1}$$
+$$\textrm{LMF} = \frac{M_F}{M_T} = \left(\frac{M_F}{A_S}\right) \left(\frac{M_T}{A_S}\right)^{-1}.$$
 
@@
 
-R.A.D and D.S.F. conceived the ideas and methods, and co-wrote the paper. R.A.D. analysed data. Co-authors in the BAAD consortium shared data.
+R.A.D and D.S.F. conceived the ideas and methods, and co-wrote the paper. R.A.D. analysed data.
 
@@
     \centering
-    \includegraphics[width=0.6\textwidth]{Figure3.pdf}
-    \caption{\textbf{No biome differences in leaf mass per unit stem basal stem area for $M_F/A_S$ and weak, inconsistent effects of biome on leaf area per unit basal stem area ($A_F/A_S$)}.
+    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
+    \caption{\textbf{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning}.
+a) LMF for each PFT within three different biomes.
 No leaf area estimates were available for boreal deciduous angiosperms.
-DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.
 Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect.
-Letters denote significant differences (at $\alpha = 0.05$)}
+Letters denote significant differences (at $\alpha = 0.05$).
+b-c) Relationships between climate and LMF within each PFT.
+Solid lines are generalized additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
     \label{fig:figure3}
@@
     \includegraphics[width=0.99\textwidth]{Figure4.pdf}
-    \caption{\textbf{Modest effects of climate on leaf mass and area per unit stem basal area, compared to the very large effect of plant functional type on mass-based variables ($M_F/A_S$)}.
-Solid lines are generalized additive model (GAM) fits (with a base dimension of 3, larger values resulted in biologically implausible relationships).
-Grey areas are approximate 95\% confidence intervals, calculated by the GAM.
-To account for within-species and contributed dataset correlations, data were first averaged by species/dataset combination.}
+     \caption{\textbf{Leaf and stem traits capture variation in biomass partitioning across species.} Individual plant data were averaged by species and study combinations.
+Lines are linear regressions with 95\% confidence intervals.
+Both regressions were significant ($P \textless 0.01$).
+$R^2$ for fitted relationships are 0.28 and XXX in the right panel.}
   \label{fig:figure4}
@@
 
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureS6.pdf}
-    \caption{\textbf{Plant leaf mass as a function of basal stem area (left panel), and leaf area as a function of basal stem area (right panel).} Individual plant data were averaged by species/study (the same species occurring in multiple datasets thus appears multiple times).
-Lines are linear regressions with 95\% confidence intervals.
-Both regressions were significant ($P \textless 0.01$), but $R^2$ was 0.28 in the left panel, and only 0.02 in the right panel.}
-  \label{fig:figureS6}
-\end{figure}
-
 \clearpage
@@
 \setcounter{table}{0}
-
-<<results='asis'>>=
-print(xtable(dataauthors,
-             caption="\\textbf{Co-authors in the BAAD consortium.}"
-             ), include.rownames=FALSE, tabular.environment='longtable', floating=FALSE, size="small",
-      caption.placement="top")
-@
 
@@
 $$
-\frac{dM_i}{dt} = a_i NPP - \tau_i M_i
+\frac{dM_i}{dt} = a_i \textrm{NPP} - \tau_i M_i
 $$

### 2015-01-12 manuscript.Rnw
Text editing; more refs.

@@
 baad <- readRDS("downloads/baad.rds")
-dataauthors <- author_details(baad, "data/contact.csv")
 @
@@
 \textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time of carbon in ecosystems \citep{friend2014}, predictions of global carbon storage \citep{ise2010}, and the CO\textsubscript{2} response of net primary production \citep{dekauwe2014}.
-Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
+Although GVMs typically predict that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
 Using a new global database on biomass partitioning from over 175 studies, we here show a strong effect of PFT on partitioning between stems and leaves.
@@
 Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum \citep{wrightetal2004}.
-In addition, we show how a second trait, leaf area per stem basal area, contributes additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs.
+In addition, we show how a second trait, leaf area per stem basal area, explains additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs.
 Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
@@
 The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\cite{ise2010}.
-Globally, forests store approximately 360Pg of carbon\citep{panetal2011}, equivalent to about 40years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}.
-Biomass stored in woody stems is unproductive but has a long residence time \citep{luyssaert2008}, whereas leaf biomass is enhances photosynthesis but turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}.
+Globally, forests store approximately 360Pg of carbon\citep{panetal2011} in living biomass, equivalent to about 40years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}.
+Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass determines turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}.
 Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014}.
 
-Perhaps and the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising world vegetation.
-While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots) and the quality with which these are constructed.
-The approach taken by most GVMs for handling this functional diversity, is to consider only a few archetypal PFTs \citep{harrisonetal2010}, assumed to differ in key physiological attributes.
-While it is common to assume differences in biomass partitioning between PFTs, these estimates are based on limited empirical data.
-Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm}) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
+Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation.
+While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots).
+The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010}, assumed to differ in key physiological attributes.
+While GMVs predict differences in biomass partitioning between PFTs (Methods \ref{sec:gvmsummary}), these predictions are constrained by limited empirical data.
+Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
 These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared.
@@
 S1).
-Previous work based on whole stands\citep{oneilldeangelis1981,enquistniklas2002,reich2014} and very small plants\citep{poorter2012}, reported differences among angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
+Previous work based on whole stands\citep{oneilldeangelis1981,enquistniklas2002,reich2014} and very small plants\citep{poorter2012}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
 Here we extend this finding to show that LMF differs not only across the angiosperm-gymnosperm divide, but also within angiosperms, at the level of individual plants, and across the entire size range of woody plants in natural forests (0.1-100m) (Fig.
@@
 
-Moreover, we show that differences in LMF among woody PFTs arise due to differences in leaf morphology, not in the amount of leaf area held on the plant (Fig.
+Moreover, we show that differences in LMF among woody PFTs arise due to differences in leaf morphology, but they converge in the amount of leaf area held on the plant (Fig.
 2b-c).
@@
 Leaf area is relevant because it drives total light interception and thus photosynthesis.
-Several classic studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}, suggesting a growth advantage.
+Several classic studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
 We found no evidence for this, instead we found that the three PFTS maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR).
@@
 2b).
-The consequence of this is that the leaf area ratio (the amount of leaf area supported per unit above-ground biomass), was invariant between PFTs, because $\textrm{LAR} = LMF / \textrm{LMA}$ (Fig.
+As a consequence, LAR was invariant between PFTs, because $\textrm{LAR} = LMF / \textrm{LMA}$ (Fig.
 2c and S3).
 
-Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem; quantifying the amount of light intercepting tissue relative to the amount of tissue available to transport water through the stem.
-This measure has long been assumed to ..(reason why it's important)...
-Mirroring the results for LMF, we find that the amount of leaf mass per unit stem area differs among the three PFTs, that these differences also relate to LMA (Fig.
-2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area (Fig.
+Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$), a quantity closely related to hydraulic supply capacity \cite{enquist2003}.
+This ratio is also important because plant leaf mass is frequently estimated from $A_S$ with allometric equations \cite{shinozaki1964b, chave2005}.
+Mirroring the results for LMF, we find that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig.
+2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig.
 2e).
@@
 The above results highlight how differences in LMA -- a central trait of the leaf economics spectrum\citep{wrightetal2004} -- drives differences in biomass partitioning among dominant woody PFTs.
-Based on a XXX statistical model, we found that LMA alone could explain almost as much  (?more?) variation as PFT, even though LMA only captures aspects of leaf physiology.
+Based on a XXX statistical model, we found that LMA alone could explain almost as much  (?more?) variation as PFT, even though LMA only captures aspects of leaf morphology.
 These results suggest promising avenues for parameterizing and simplifying GVMs.
@@
 Thus, to answer these additional questions we found it helpful to decompose LMF as the product,
-\begin{equation}\label{eq:M1}\textrm{LMF} = \left(\frac{M_F}{A_S}\right) \left(\frac{A_S}{M_T}\right).\end{equation}
+
+\begin{equation}\label{eq:M1}
+\textrm{LMF} = \left(\frac{M_F}{A_S}\right) \left(\frac{A_S}{M_T}\right).
+\end{equation}
+
 An advantage of eq.
@@
 2d, the remaining term $M_F/A_S$ exhibits comparable differences to LMF, but unlike $M_F/A_S$ is independent of plant height (Tables 1-2).
-$M_F/A_S$ is thus a useful of LMF that can be compared across species and climates.
+$M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and climates.
 
@@
 With the caveats that there is still variation between species and studies unaccounted for (Fig.
-S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT seem stronger than any direct climate effect on biomass partitioning within individual plants (Tables 1 and 2).
+S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables 1 and 2).
 
@@
 An example of such a model is LPJ \citep{sitchetal2003}, where the ratio of leaf area to sapwood area is a constant parameter, but LMA differs between PFTs.
+
 More fundamentally, our results suggest PFTs defined on the basis of two key traits, LMA and $A_F/A_S$ would better capture patterns of biomass partitioning than existing groups, which only account for differences in one of the two traits identified here.
@@
 
-We further decomposed LMF and LAR by considering the ratio of leaf mass (or area) to plant basal area, and the ratio of above-ground biomass to plant basal area. For LMF,
-
-$$\textrm{LMF} = \frac{M_F}{M_T} = \left(\frac{M_F}{A_S}\right) \left(\frac{M_T}{A_S}\right)^{-1}.$$
-
-And similarly for LAR, by replacing $M_F$ with $A_F$.
-We were unable to further decompose $A_S/M_T$ into wood volume and wood density components (cf.
-\citep{chave2005}) because wood density data were available only for a small subset of data.
-
 We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves).
@@
 For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect.
-We calculated the $R^2$ for linear mixed-effects models as reported by \citep{nakagawa2013} for the fixed effects only.
+We calculated the $R^2$ for linear mixed-effects models as reported by for the fixed effects only \citep{nakagawa2013}.
 
@@
 
-Figure 1a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary.
+Figure 1a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary. We thank Martin de Kauwe for discussions and commenting on an earlier version of this paper.
 
@@
 a) LMF for each PFT within three different biomes.
+DA: deciduous angiosperms, EA: evergreen angiosperms, EG: evergreen gymnosperms.
 No leaf area estimates were available for boreal deciduous angiosperms.
@@
 
+<<echo=FALSE>>=
+agg <- summaryBy(lalf_astba2 + lmlf_astba2 + llma ~ Group,
+                   data=dataset, FUN=mean, na.rm=TRUE, keep.names=TRUE,
+                   id=~pft)
+agg <- subset(agg, !is.na(llma))
+lm1 <- lm(lmlf_astba2 ~ llma, data=agg)
+lm2 <- lm(lmlf_astba2 ~ lalf_astba2 * pft, data=agg)
+r2panela <- summary(lm1)$adj.r.squared
+r2panelb <- summary(lm2)$adj.r.squared
+@
+
 \begin{figure}[h!]
@@
 Both regressions were significant ($P \textless 0.01$).
-$R^2$ for fitted relationships are 0.28 and XXX in the right panel.}
+$R^2$ for fitted relationships are \Sexpr{round(100*r2panela,1)}\% in panel (a) and \Sexpr{round(100*r2panelb,1)}\% in panel (b) (for a linear regression model including PFT as a factor variable).}
   \label{fig:figure4}
@@
 
-\clearpage
-
 <<tablepreload1>>=
@@
 
-\subsection*{Methods S1 - Biomass partitioning terminology}
+\renewcommand\thesubsection{S\arabic{subsection}}
+\setcounter{subsection}{0}
+
+\subsection{Modelling of biomass partitioning in global vegetation models (GVMs)}
+\label{sec:gvmsummary}
+
+Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types.
+However, there is no consensus on how these differences are parameterized, or whether they are linked to inherent differences in leaf mass per area (LMA) or leaf lifespan (LL).
+We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them.
+We base this classification on a recent model intercomparison study \cite{dekauwe2014}.
+See Methods \ref{sec:partterm} for use of the terms 'partitioning' and 'allocation'.
+
+\subsubsection*{Constant allocation fractions}
+
+Some models assume that a constant fraction of net primary production (NPP) is allocated to each component (leaves, stems, roots).
+CABLE and EALCO are examples of models that allocate a PFT-specific fraction of NPP to each plant component each timestep \cite{dekauwe2014} (though, for both models, alternative model formulations have been used as well).
+Constant allocation fractions still give rise to differences in biomass partitioning between PFTs, since turnover of leaves (determined by LL) will always differ between at least deciduous and evergreen PFTs.
+
+\subsubsection*{Allometric constraints}
+
+In this class of models, some allometric (or similar) relationship between biomass components is used to calculate biomass allocation in each timestep.
+For example, in ED2 \cite{medvigy2009}, biomass partitioning to leaves, stem sapwood and roots is constrained by a set of equations with PFT-specific parameters.
+Biomass allocation is then calculated at each timestep to satisfy these equations.
+Other models use leaf area as a constraint.
+For example, LPJ \cite{sitchetal2003} uses PFT-specific values of leaf area to sapwood area ratio to constrain estimates of biomass allocation.
+Further differences in biomass partitioning between PFTs are then the result of differences in LMA (leaf mass at a given leaf area), as well as LL (shorter lived leaves require higher annual allocation of biomass for upkeep).
+
+\subsubsection*{Optimality}
+
+Few GVMs use a scheme in which carbon allocation is determined by assuming biomass fractions are the result of optimization, however, this scheme is widely used in smaller-scale vegetation models \cite{valentinemakela2012,franklin2012}.
+The SDGVM model \cite{woodward1995} optimizes leaf area index by considering the carbon balance of the lowest leaf layer in the canopy.
+As with LPJ above, differences in biomass partitioning between PFTs then arise as a result of differences in LMA as well as LL.
+
+\subsection{Biomass partitioning terminology}
 \label{sec:partterm}
@@
 We may write (cf.
+Ref.
 \citep{McMurtrieWolf1983}),
@@
 
-\subsection*{Methods S2 - Predicting basal diameter}
+\subsection{Predicting basal diameter}
 \label{sec:predas}

### 2015-01-12 R/figures.R

@@ -294,15 +294,17 @@ figure3 <- function(dataset, dataset2){
 
 # SI6
 # MF/AS and AF/AS at species level by PFT.
-figure4<- function(dataset){
+figure4 <- function(dataset){
   agg <- summaryBy(lalf_astba2 + lmlf_astba2 + llma ~ Group,
                    data=dataset, FUN=mean, na.rm=TRUE, keep.names=TRUE,
                    id=~pft)
   agg <- subset(agg, !is.na(llma))
 
   lm1 <- lm(lmlf_astba2 ~ llma, data=agg)
-  lm2 <- lm(lmlf_astba2 ~ lalf_astba2, data=agg)
-
+  
+  # For R2.
+  lm2s <- lapply(split(agg, agg$pft), function(x)lm(lmlf_astba2 ~ lalf_astba2, data=x))
+  
   par(mfrow=c(1,2), mar=c(4,0.2,0.2,0.2),oma=c(1,4,1,1), cex.lab=1.1, las=1)
 
   with(agg, plot(llma, lmlf_astba2, pch=19, col=my_cols_transparent()[pft],
@@ -321,7 +323,7 @@ figure4<- function(dataset){
   magaxis(1, unlog=1)
   magaxis(2, unlog=2, labels=FALSE)
 
-  predline(lm2)
+  for(i in 1:3)predline(lm2s[[i]], col=my_linecols()[i])
   box()
   plotlabel("(b)","topleft", log.y=FALSE, log.x=FALSE)
   my_legend("bottomright", cex=0.8, pt.cex=1)

### 2015-01-13 manuscript.Rnw
Added R2 for LMA instead of PFT in models of LMF and MF/AS. Editing to text; added comment to paragraph about AF/AS.

@@
 The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\cite{ise2010}.
-Globally, forests store approximately 360Pg of carbon\citep{panetal2011} in living biomass, equivalent to about 40years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}.
+Globally, forests store approximately 360Pg of carbon in living biomass\citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}.
 Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass determines turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}.
@@
 We found no evidence for this, instead we found that the three PFTS maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR).
+
 Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig.
@@
 This ratio is also important because plant leaf mass is frequently estimated from $A_S$ with allometric equations \cite{shinozaki1964b, chave2005}.
-Mirroring the results for LMF, we find that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig.
+Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig.
 2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig.
@@
 The above results highlight how differences in LMA -- a central trait of the leaf economics spectrum\citep{wrightetal2004} -- drives differences in biomass partitioning among dominant woody PFTs.
-Based on a XXX statistical model, we found that LMA alone could explain almost as much  (?more?) variation as PFT, even though LMA only captures aspects of leaf morphology.
-These results suggest promising avenues for parameterizing and simplifying GVMs.
+To assess whether LMA could replace PFT when predicting biomass partitioning, we fit a linear mixed-effects model to LMF and $M_F/A_S$ using PFTa dn plant height (and the quadratic term) as predictors (and all interactions).
+We then replaced PFT with LMA, and found that LMA alone could explain almost as much variation in LMF as PFT ($R^2$ = 0.74 with PFT vs.
+0.62 with LMA) and $M_F/A_S$ ($R^2$ = 0.29 vs.
+0.28).
+This is surprising because LMA only captures aspects of leaf morphology.
+These results suggest promising avenues for parameterizing and simplifying biomass partitioning routines in GVMs.
 
@@
 
+\emph{I think the point is that we found MF/AS to be proportional to LMA across PFTs, however, the other component of MF/AS is AF/AS and that in fact, across species, explained more variation in MF/AS.
+So, though LMA is a good substitute for PFT, lots of other variation still exists, which should be taken into account.
+We just have shown that AF/AS does not depend on PFT, and almost not on climate - so it has to be some unidentified factor instead.
+??}.
 In contrast, we found that variation among species in leaf and stem traits could explain additional variation in biomass partitioning.
@@
 The remaining (??
-what is number??) variation in $M_F/A_S$  was explained by species-level differences in $A_F/A_S$ (Fig.
+what is number??
+\emph{the remainder should be 100 - 28}) variation in $M_F/A_S$  was explained by species-level differences in $A_F/A_S$ (Fig.
 4b).
@@
 
-We used the recently compiled Biomass And Allometry Database (BAAD) \citep{Falster2015}, including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_W$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
+We used the recently compiled Biomass And Allometry Database (BAAD) \citep{Falster2015}, including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
 1.3m) ($A_{Sbh}$).
@@
 We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves).
-For conifers, leaf area was converted to half-total surface area using published conversion factors \citep{barclay2000}, which is most appropriate for comparison to flat leaves in terms of light interception  \citep{lang1991, chen1992}.
+For conifers, leaf area was converted to half-total surface area using published conversion factors \citep{barclay2000}, which is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}.
 Stem cross-sectional area was measured either at breast height (ca.
@@
 
-Figure 1a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary. We thank Martin de Kauwe for discussions and commenting on an earlier version of this paper.
+Figure 1a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary. We thank Martin de Kauwe for discussions and comments on an earlier version of this paper.
 
@@
     \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
-    \caption{\textbf{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_W$) for each of the PFTs, as a function of total plant height ($H$).} Each point is an individual plant.}
+    \caption{\textbf{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$).} Each point is an individual plant.}
     \label{fig:figureS2}

### 2015-01-13 manuscript.Rnw
Figure 4 has 3 regression lines in panel a

@@
 agg <- subset(agg, !is.na(llma))
-lm1 <- lm(lmlf_astba2 ~ llma, data=agg)
+lm1 <- lm(lmlf_astba2 ~ llma * pft, data=agg)
 lm2 <- lm(lmlf_astba2 ~ lalf_astba2 * pft, data=agg)

### 2015-01-14 manuscript.Rnw
Deleted functions-ms.R; no longer needed
All log axes are formatted consistently. Closes #3

@@
 opts_chunk$set(echo=FALSE)
-source("R/functions-ms.R")
 baad <- readRDS("downloads/baad.rds")
@@
     \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
-    \caption{\textbf{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.} Shown are histograms of leaf mass per area (an index of leaf construction cost) ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types.
+    \caption{\textbf{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.} Shown are histograms (as probability density functions) of leaf mass per area (an index of leaf construction cost) ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types.
 Arrows indicate means (same as in Fig.~\ref{fig:figure2}).}

### 2015-01-14 R/functions-figures.R
All log axes are formatted consistently. Closes #3

@@ -631,10 +631,36 @@ ablinepiece <- function(a=NULL,b=NULL,reg=NULL,from=NULL,to=NULL,...){
 }
 
 
-## Position label at a fractional x/y position on a plot
-label <- function(px, py, lab, ..., adj=c(0, 1)) {
-  usr <- par("usr")
-  text(usr[1] + px*(usr[2] - usr[1]),
-       usr[3] + py*(usr[4] - usr[3]),
-       lab, adj=adj, ...)
+
+# Simple axes function, bases on magaxis and code therein. 
+log10axes <- function(side=1:2, logged=NULL, labels=TRUE){
+  
+  magaxis(side=side, unlog=side, labels=FALSE)
+  
+  loggedaxes <- c(FALSE,FALSE)
+  if(!is.null(logged)){
+    loggedaxes[logged] <- TRUE  
+  }
+  
+  for(ii in side){
+    lims <- sort(par("usr")[(1 + (ii-1)*2):(2 + (ii-1)*2)])
+    sci.tick <- maglab(10^lims, n = 5, log = TRUE, exptext = TRUE, 
+                      crunch = TRUE, logpretty = TRUE, usemultloc = FALSE, 
+                      prettybase = 10, hersh = FALSE)
+    
+    if(labels){
+      lab <- do.call(expression, lapply(log10(sci.tick$labat), function(i) bquote(10^.(i))))
+      
+      if(loggedaxes[ii]){
+        axis(side = ii, at = sci.tick$labat, tick = FALSE, 
+                labels = lab, mgp = par("mgp"))
+      } else {
+        axis(side = ii, at = log10(sci.tick$labat), tick = FALSE, 
+             labels = lab, mgp = par("mgp"))
+      } 
+    }
+  }
 }
+
+# with(dataset, plot(llma, lmlf_astba2, pch=16, axes=FALSE, xlim=c(-2,0)))
+# log10axes(2)

### 2015-01-14 manuscript.Rnw
edits to text, attempt to better integrate climate effects and AF/AS

@@
 The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010}, assumed to differ in key physiological attributes.
-While GMVs predict differences in biomass partitioning between PFTs (Methods \ref{sec:gvmsummary}), these predictions are constrained by limited empirical data.
+While GVMs predict differences in biomass partitioning between PFTs (Methods \ref{sec:gvmsummary}), these predictions are constrained by limited empirical data.
 Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
@@
 2b).
-As a consequence, LAR was invariant between PFTs, because $\textrm{LAR} = LMF / \textrm{LMA}$ (Fig.
+As a consequence, LAR was invariant between PFTs, because $\textrm{LAR} = \textrm{LMF} / \textrm{LMA}$ (Fig.
 2c and S3).
@@
 The above results highlight how differences in LMA -- a central trait of the leaf economics spectrum\citep{wrightetal2004} -- drives differences in biomass partitioning among dominant woody PFTs.
-To assess whether LMA could replace PFT when predicting biomass partitioning, we fit a linear mixed-effects model to LMF and $M_F/A_S$ using PFTa dn plant height (and the quadratic term) as predictors (and all interactions).
+To assess whether LMA could replace PFT when predicting biomass partitioning, we fit a linear mixed-effects model to LMF and $M_F/A_S$ using PFT and plant height (and the quadratic term) as predictors (and all interactions).
 We then replaced PFT with LMA, and found that LMA alone could explain almost as much variation in LMF as PFT ($R^2$ = 0.74 with PFT vs.
@@
 
-Given these results, we were also interested to know whether additional variation in biomass partitioning could be explained by either climate, within PFT variation in LMA or by other traits.
-Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons among species and across climatic gradients, due to the small sample size within each species or site.
-Thus, to answer these additional questions we found it helpful to decompose LMF as the product,
+Many GVMs assume that climate influences biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Methods \ref{sec:gvmsummary}).
+We were thus interested to know whether additional variation in biomass partitioning could be explained by climate, as a basis for model formulation.
+Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons across climatic gradients, due to the small sample size within each species or site.
+Thus, we decomposed LMF as the product,
 
@@
 2b and Fig.
-2d, the remaining term $M_F/A_S$ exhibits comparable differences to LMF, but unlike $M_F/A_S$ is independent of plant height (Tables 1-2).
-$M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and climates.
+2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables 1-2).
+$M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.
 
@@
 
-\emph{I think the point is that we found MF/AS to be proportional to LMA across PFTs, however, the other component of MF/AS is AF/AS and that in fact, across species, explained more variation in MF/AS.
-So, though LMA is a good substitute for PFT, lots of other variation still exists, which should be taken into account.
-We just have shown that AF/AS does not depend on PFT, and almost not on climate - so it has to be some unidentified factor instead.
-??}.
-In contrast, we found that variation among species in leaf and stem traits could explain additional variation in biomass partitioning.
-The effect of LMA was found to hold also within PFTs (Fig.
-4a), explaining 28\% of the variation in $M_F/A_S$ across species.
-The fact that PFTs differ in LMA is evident in the separation of the three groups along the x-axis of Fig.
-4a.
-Yet these clouds were broad and overlapping, with a more general relationship extending across the entire LMA axis.
-The remaining (??
-what is number??
-\emph{the remainder should be 100 - 28}) variation in $M_F/A_S$  was explained by species-level differences in $A_F/A_S$ (Fig.
+We found that LMF differs predictably among PFTs, however there was considerable variation in all studied variables between species within PFTs (Fig.
+S5).
+It is thus pertinent to understand whether we can move beyond a PFT-based classification by considering which traits affect biomass partitioning.
+To do this, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq.
+\ref{eq:M2}).
+
+\begin{equation}\label{eq:M2}
+\frac{M_F}{A_S} = \left(\frac{M_F}{A_F}\right) \left(\frac{A_F}{A_S}\right).
+\end{equation}
+
+The correlation between $M_F/A_S$ and LMA was found to hold also within PFTs (Fig.
+4a), explaining 30\% of the variation in $M_F/A_S$ across species.
+The regressions across species by PFT were broad and overlapping, with a more general relationship extending across the entire LMA axis.
+A larger fraction of the variation in $M_F/A_S$ was explained by species-level differences in $A_F/A_S$ (Fig.
 4b).
@@
 2e), but does vary close to two orders of magnitude across species.
-The separation among PFTs along the Y-axis arises due to differences in LMA.
+In Fig.
+4b, The separation among PFTs in $M_F/A_S$ arises due to differences in LMA.
+We thus show that in a PFT-based classification, LMA is a good first estimate of biomass partitioning, however, more variation can be explained by another trait ($A_F/A_S$), which varies appreciably between species.
 

### 2015-01-14 manuscript.Rnw
Edits to Methods S1, summary of climate effects in GVMs on biomass partitioning

@@
 
-Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types.
+Most leading GVMs predict that biomass partitioning between leaves and woody stems differs between major plant functional types.
 However, there is no consensus on how these differences are parameterized, or whether they are linked to inherent differences in leaf mass per area (LMA) or leaf lifespan (LL).
-We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them.
+We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them (See Methods \ref{sec:partterm} for use of the terms 'partitioning' and 'allocation').
 We base this classification on a recent model intercomparison study \cite{dekauwe2014}.
-See Methods \ref{sec:partterm} for use of the terms 'partitioning' and 'allocation'.
+We then briefly summarize how current GVMs include effects of climate on biomass partitioning.
 
@@
 Some models assume that a constant fraction of net primary production (NPP) is allocated to each component (leaves, stems, roots).
-CABLE and EALCO are examples of models that allocate a PFT-specific fraction of NPP to each plant component each timestep \cite{dekauwe2014} (though, for both models, alternative model formulations have been used as well).
-Constant allocation fractions still give rise to differences in biomass partitioning between PFTs, since turnover of leaves (determined by LL) will always differ between at least deciduous and evergreen PFTs.
+CABLE, EALCO \cite{dekauwe2014} and IBIS \cite{foley1996} are examples of models that allocate a PFT-specific fraction of NPP to each plant component each timestep  (though, for both models, alternative model formulations have been used as well).
+Constant allocation fractions still give rise to differences in biomass partitioning between PFTs (see Methods \ref{sec:partterm}), since turnover of leaves (determined by LL) will always differ between at least deciduous and evergreen PFTs.
 
@@
 
-In this class of models, some allometric (or similar) relationship between biomass components is used to calculate biomass allocation in each timestep.
-For example, in ED2 \cite{medvigy2009}, biomass partitioning to leaves, stem sapwood and roots is constrained by a set of equations with PFT-specific parameters.
+In this class of models, some allometric (or other size-dependent) relationship between biomass components is used to calculate biomass allocation in each timestep.
+For example, in ED2 \cite{medvigy2009}, biomass partitioning to leaves, stem sapwood and roots is constrained by a set of equations with PFT-specific parameters, estimated through model calibration.
 Biomass allocation is then calculated at each timestep to satisfy these equations.
@@
 The SDGVM model \cite{woodward1995} optimizes leaf area index by considering the carbon balance of the lowest leaf layer in the canopy.
-As with LPJ above, differences in biomass partitioning between PFTs then arise as a result of differences in LMA as well as LL.
+As with LPJ described above, differences in biomass partitioning between PFTs then arise as a result of differences in LMA as well as LL.
+
+\subsubsection*{Climate effects}
+
+Models differ widely in how climate directly (or indirectly) affects biomass partitioning beyond differences between PFTs.
+In one class of models, biomass partitioning is modelled based on relative resource limitations, consistent with the idea that biomass allocation proceeds to the plant component most limited by the resource it acquires \cite{friedlingstein1999}.
+For example, in the allocation model of Arora and Boer \cite{aroraboer2005}, which is used in the ISAM model \cite{dekauwe2014}, allocation to leaves, stems and roots depends on relative limitations of light and water.
+Likewise, Friedlingstein et al.
+\cite{friedlingstein1999} developed a resource-limitation framework for the CASA model, where allocation depends on limitations of light, water and nutrients.
+It is clear that these types of models predict apparent climate effects on biomass partitioning along temperature and moisture gradients.
+
+Some models do not assume effects of stress on biomass allocation directly, but differences in allocation may result from effects of stress on biomass production, and subsequent priority rules for the allocation of biomass.
+For example, in LPJ \cite{sitchetal2003}, biomass production may be low enough in very dry years so that intended allocation to leaves and stem sapwood cannot be supported.
+In this case, allocation proceeds to leaves only.
+In dry climates, where water stress will be more common, this would gradually lead to different partitioning of biomass to leaves and stems.
 

### 2015-01-22 manuscript.Rnw
Edits to manuscript, captions. Moved Table 1 to SI1

@@
 \textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time of carbon in ecosystems \citep{friend2014}, predictions of global carbon storage \citep{ise2010}, and the CO\textsubscript{2} response of net primary production \citep{dekauwe2014}.
-Although GVMs typically predict that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
+Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
 Using a new global database on biomass partitioning from over 175 studies, we here show a strong effect of PFT on partitioning between stems and leaves.
@@
 The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\cite{ise2010}.
+Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass determines turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}.
 Globally, forests store approximately 360Pg of carbon in living biomass\citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}.
-Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass determines turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}.
 Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014}.
@@
 The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010}, assumed to differ in key physiological attributes.
-While GVMs predict differences in biomass partitioning between PFTs (Methods \ref{sec:gvmsummary}), these predictions are constrained by limited empirical data.
+While GVMs assume differences in biomass partitioning between PFTs (Methods \ref{sec:gvmsummary}), these predictions are constrained by limited empirical data.
 Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
@@
 Several classic studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
-We found no evidence for this, instead we found that the three PFTS maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR).
+We found no evidence for this, instead we found that the three PFTs maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR).
 
@@
 2b).
-As a consequence, LAR was invariant between PFTs, because $\textrm{LAR} = \textrm{LMF} / \textrm{LMA}$ (Fig.
+As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig.
 2c and S3).
@@
 
-The above results highlight how differences in LMA -- a central trait of the leaf economics spectrum\citep{wrightetal2004} -- drives differences in biomass partitioning among dominant woody PFTs.
+The above results highlight how LMA -- a central trait of the leaf economics spectrum (LES)\citep{wrightetal2004} -- drives differences in biomass partitioning among dominant woody PFTs.
 To assess whether LMA could replace PFT when predicting biomass partitioning, we fit a linear mixed-effects model to LMF and $M_F/A_S$ using PFT and plant height (and the quadratic term) as predictors (and all interactions).
-We then replaced PFT with LMA, and found that LMA alone could explain almost as much variation in LMF as PFT ($R^2$ = 0.74 with PFT vs.
+We then replaced PFT with LMA, and found that the model with LMA could explain almost as much variation in LMF as the model with PFT ($R^2$ = 0.74 with PFT vs.
 0.62 with LMA) and $M_F/A_S$ ($R^2$ = 0.29 vs.
 0.28).
-This is surprising because LMA only captures aspects of leaf morphology.
+This is surprising because LMA only captures aspects of leaf morphology, in indirectly leaf lifespan and photosynthetic capacity via the LES\citep{wrightetal2004}.
 These results suggest promising avenues for parameterizing and simplifying biomass partitioning routines in GVMs.
@@
 Many GVMs assume that climate influences biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Methods \ref{sec:gvmsummary}).
-We were thus interested to know whether additional variation in biomass partitioning could be explained by climate, as a basis for model formulation.
+We were thus interested to know whether additional variation in biomass partitioning could be explained by climate.
 Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons across climatic gradients, due to the small sample size within each species or site.
@@
 An advantage of eq.
-\ref{eq:M1} is that the second term absorbs nearly all of the size related variation in LMF (Tables 1-2) and is otherwise fairly conserved between species\citep{chave2005} and across PFTs (Fig.
+\ref{eq:M1} is that the second term absorbs nearly all of the size related variation in LMF (Tables 1 and S1) and is otherwise fairly conserved between species\citep{chave2005} and across PFTs (Fig.
 S4).
@@
 2b and Fig.
-2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables 1-2).
+2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Table 1 and S1).
 $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.
@@
 S5).
-It is thus pertinent to understand whether we can move beyond a PFT-based classification by considering which traits affect biomass partitioning.
+It is thus pertinent to understand whether we can move beyond a PFT-based classification by considering other traits that affect biomass partitioning.
 To do this, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq.
@@
 A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions.
-A first approximation within GVMs would be to keep leaf area to stem cross-sectional area constant for a given PFT, and vary biomass allocation accordingly.
-LMA, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem biomass.
-An example of such a model is LPJ \citep{sitchetal2003}, where the ratio of leaf area to sapwood area is a constant parameter, but LMA differs between PFTs.
-
-More fundamentally, our results suggest PFTs defined on the basis of two key traits, LMA and $A_F/A_S$ would better capture patterns of biomass partitioning than existing groups, which only account for differences in one of the two traits identified here.
+Based on our findings, a first approximation within GVMs would be to assume leaf area to stem cross-sectional area a parameter that does not vary between PFTs, and vary biomass allocation accordingly.
+LMA, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem area.
+Some models already incorporate a similar algorithm (see Methods~\ref{sec:gvmsummary}), while other algorithms may be tuned to yield similar patterns between PFTs as we have presented.
 In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{Falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz_towards_2012, dekauwe2014} of GVMs against empirical data.
@@
 For each analysis, we used different subsets of the data.
-Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4} We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
+Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4}.
+We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
 We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms.
@@
 1.3m) and/or at the base of the plant.
-For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using non-linear regression (see~\ref{sec:predas}).
+For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using a non-linear regression model (see~\ref{sec:predas}).
 
@@
 We used the \emph{mgcv} package in R \citep{wood2006}.
-All variables were log-transformed prior to analysis.
+All variables (except MAP and MAT) were log-transformed prior to analysis.
 
@@
 For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect.
-We calculated the $R^2$ for linear mixed-effects models as reported by for the fixed effects only \citep{nakagawa2013}.
+We calculated the $R^2$ for linear mixed-effects models for the fixed effects only \citep{nakagawa2013}.
 
@@
 <<results='asis'>>=
-print(xtable(table_gamr2MATMAP, label="table/gamr2table",
-      caption="\\textbf{Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.} Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),
-      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
-@
-
-<<results='asis'>>=
 print(xtable(table_varpart2, label="varpart2",
-             caption="\\textbf{Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.} Each variable was added to a linear mixed-effects model, using species within dataset as a random effect.
-The $r^2$ shown is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type).
+             caption="\\textbf{Explained variance by plant functional type (PFT), plant height (H) and biome (B), for five whole-plant variables.} Each variable was added to a linear mixed-effects model, using species within dataset as a random effect.
+The $R^2$ shown is that explained by the fixed effects only.
 All fixed effects were highly significant (P \\textless 0.001).
@@
     \caption{\textbf{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning}.
-a) LMF for each PFT within three different biomes.
+a) Leaf mass per unit stem cross-sectional area ($M_F/A_S$) for each PFT within three different biomes.
 DA: deciduous angiosperms, EA: evergreen angiosperms, EG: evergreen gymnosperms.
-No leaf area estimates were available for boreal deciduous angiosperms.
 Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect.
 Letters denote significant differences (at $\alpha = 0.05$).
-b-c) Relationships between climate and LMF within each PFT.
+b-c) Relationships between climate and $M_F/A_S$ within each PFT.
 Solid lines are generalized additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
@@
     \caption{\textbf{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.} Shown are histograms (as probability density functions) of leaf mass per area (an index of leaf construction cost) ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types.
-Arrows indicate means (same as in Fig.~\ref{fig:figure2}).}
+Arrows indicate means by PFT.}
     \label{fig:figures5}
@@
 \setcounter{table}{0}
+
+<<results='asis'>>=
+print(xtable(table_gamr2MATMAP, label="table/gamr2table",
+      caption="\\textbf{Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.} Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),
+      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
+@
 
@@
 
-Most leading GVMs predict that biomass partitioning between leaves and woody stems differs between major plant functional types.
+Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types.
 However, there is no consensus on how these differences are parameterized, or whether they are linked to inherent differences in leaf mass per area (LMA) or leaf lifespan (LL).
-We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them (See Methods \ref{sec:partterm} for use of the terms 'partitioning' and 'allocation').
+Typically, parameters related to biomass allocation differ between PFTs (often based on little empirical evidence), which then lead to differences in biomass partitioning.
+We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them.
 We base this classification on a recent model intercomparison study \cite{dekauwe2014}.
-We then briefly summarize how current GVMs include effects of climate on biomass partitioning.
+Note that we use 'partioning' to indicate the actual distribution of biomass among plant components, and 'allocation' the fraction of net primary production to each component (see Methods \ref{sec:partterm}).
+Finally, we briefly summarize how current GVMs include effects of climate on biomass partitioning.
 
@@
 
-In this class of models, some allometric (or other size-dependent) relationship between biomass components is used to calculate biomass allocation in each timestep.
+In this class of models,  allometric (or other size-dependent) relationships between biomass components are used to calculate biomass allocation in each timestep.
 For example, in ED2 \cite{medvigy2009}, biomass partitioning to leaves, stem sapwood and roots is constrained by a set of equations with PFT-specific parameters, estimated through model calibration.
@@
 Few GVMs use a scheme in which carbon allocation is determined by assuming biomass fractions are the result of optimization, however, this scheme is widely used in smaller-scale vegetation models \cite{valentinemakela2012,franklin2012}.
-The SDGVM model \cite{woodward1995} optimizes leaf area index by considering the carbon balance of the lowest leaf layer in the canopy.
+One example of a global model is SDGVM \cite{woodward1995}, which optimizes leaf area index by considering the carbon balance of the lowest leaf layer in the canopy.
 As with LPJ described above, differences in biomass partitioning between PFTs then arise as a result of differences in LMA as well as LL.

### 2015-01-22 manuscript.Rnw
table refs fixed

@@
 An advantage of eq.
-\ref{eq:M1} is that the second term absorbs nearly all of the size related variation in LMF (Tables 1 and S1) and is otherwise fairly conserved between species\citep{chave2005} and across PFTs (Fig.
+\ref{eq:M1} is that the second term absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}) and is otherwise fairly conserved between species\citep{chave2005} and across PFTs (Fig.
 S4).
@@
 2b and Fig.
-2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Table 1 and S1).
+2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}).
 $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.
@@
 3).
-Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table 2).
+Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table \ref{table/varpartB}).
 Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) (Fig.
 3b-c).
-These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table 1).
+These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table \ref{table/varpartMAPMAT}).
 
@@
 With the caveats that there is still variation between species and studies unaccounted for (Fig.
-S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables 1 and 2).
+S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}).
 
@@
 <<results='asis'>>=
-print(xtable(table_varpart2, label="varpart2",
+print(xtable(table_varpart2, label="table/varpartB",
              caption="\\textbf{Explained variance by plant functional type (PFT), plant height (H) and biome (B), for five whole-plant variables.} Each variable was added to a linear mixed-effects model, using species within dataset as a random effect.
@@
 <<results='asis'>>=
-print(xtable(table_gamr2MATMAP, label="table/gamr2table",
+print(xtable(table_gamr2MATMAP, label="table/varpartMAPMAT",
       caption="\\textbf{Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.} Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),

### 2015-02-09 manuscript.Rnw
added Wolf2011, and minor tidy refs
small tweaks to abstract, add ref to BAAD

@@
 
-\textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time of carbon in ecosystems \citep{friend2014}, predictions of global carbon storage \citep{ise2010}, and the CO\textsubscript{2} response of net primary production \citep{dekauwe2014}.
+\textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time and storage of carbon in ecosystems, and the response of net primary production to rising CO\textsubscript{2}\citep{dekauwe2014,friend2014,ise2010}.
 Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
-Using a new global database on biomass partitioning from over 175 studies, we here show a strong effect of PFT on partitioning between stems and leaves.
+Using a new global database on biomass partitioning from over 175 studies\citep{Falster2015}, we here show a strong effect of PFT on partitioning between stems and leaves.
 We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass.
@@
 Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
-These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared.
+These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared \citep{wolf2011}.
 

### 2015-05-27 manuscript.Rnw
Moved to Ecology Letters format; fixed a reference

@@
 
-\usepackage[sort&compress]{natbib}
-\citestyle{nature}
+\usepackage{natbib}
+\bibliographystyle{ecol_let}
 
@@
 
-\bibliographystyle{ieeetr}
 \linenumbers
@@
 
-\section*{First paragraph}
+\section*{Abstract}
 
-\textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time and storage of carbon in ecosystems, and the response of net primary production to rising CO\textsubscript{2}\citep{dekauwe2014,friend2014,ise2010}.
+% Should be 150 words - can move some to introduction
+The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time and storage of carbon in ecosystems, and the response of net primary production to rising CO\textsubscript{2}\citep{dekauwe2014,friend2014,ise2010}.
 Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
@@
 Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
-}
 
-\section*{Main text}
+\section*{Introduction}
 
@@
 
+\section*{Methods}
+
+We used the recently compiled Biomass And Allometry Database (BAAD) \citep{Falster2015}, including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
+1.3m) ($A_{Sbh}$).
+For each analysis, we used different subsets of the data.
+Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4}.
+We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
+We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms.
+We excluded  deciduous gymnosperms because few data were available.
+All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes.
+To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).
+
+For the above-ground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total above-ground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,
+
+$$\textrm{LMF} = \frac{M_F}{M_T} = \left(\frac{M_F}{A_F}\right) \left(\frac{A_F}{M_T}\right) = \textrm{LMA} \times \textrm{LAR}.$$
+
+We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves).
+For conifers, leaf area was converted to half-total surface area using published conversion factors \citep{barclay2000}, which is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}.
+Stem cross-sectional area was measured either at breast height (ca.
+1.3m) and/or at the base of the plant.
+For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using a non-linear regression model (see~\ref{sec:predas}).
+
+We used generalized additive models (GAM) to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height.
+In all fitted GAMs, we used species-dataset combination as a random effect.
+We used the \emph{mgcv} package in R \citep{wood2006}.
+All variables (except MAP and MAT) were log-transformed prior to analysis.
+
+Variance explained by quantitative climate variables (MAP and MAT) were tested with GAMs where variables were sequentially added to the model, and the explained variance ($R^2$) calculated.
+For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect.
+We calculated the $R^2$ for linear mixed-effects models for the fixed effects only \citep{nakagawa2013}.
+
+\section*{Results}
+
+% Placed section cut somewhat arbitrarily!
 Moreover, we show that differences in LMF among woody PFTs arise due to differences in leaf morphology, but they converge in the amount of leaf area held on the plant (Fig.
@@
 
+\section*{Discussion}
+
+% Placed section cut somewhat arbitrarily!
 The finding that climate does not appreciably affect biomass partitioning is good news for models that parameterize the globe by PFT.
@@
 
-\section*{Methods}
-
-We used the recently compiled Biomass And Allometry Database (BAAD) \citep{Falster2015}, including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
-1.3m) ($A_{Sbh}$).
-For each analysis, we used different subsets of the data.
-Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4}.
-We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
-We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms.
-We excluded  deciduous gymnosperms because few data were available.
-All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes.
-To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).
-
-For the above-ground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total above-ground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,
-
-$$\textrm{LMF} = \frac{M_F}{M_T} = \left(\frac{M_F}{A_F}\right) \left(\frac{A_F}{M_T}\right) = \textrm{LMA} \times \textrm{LAR}.$$
-
-We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves).
-For conifers, leaf area was converted to half-total surface area using published conversion factors \citep{barclay2000}, which is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}.
-Stem cross-sectional area was measured either at breast height (ca.
-1.3m) and/or at the base of the plant.
-For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using a non-linear regression model (see~\ref{sec:predas}).
-
-We used generalized additive models (GAM) to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height.
-In all fitted GAMs, we used species-dataset combination as a random effect.
-We used the \emph{mgcv} package in R \citep{wood2006}.
-All variables (except MAP and MAT) were log-transformed prior to analysis.
-
-Variance explained by quantitative climate variables (MAP and MAT) were tested with GAMs where variables were sequentially added to the model, and the explained variance ($R^2$) calculated.
-For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect.
-We calculated the $R^2$ for linear mixed-effects models for the fixed effects only \citep{nakagawa2013}.
-
 \section{Acknowledgments}
@@
 Figure 1a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary. We thank Martin de Kauwe for discussions and comments on an earlier version of this paper.
-
-\section{Author contributions}
-
-R.A.D and D.S.F. conceived the ideas and methods, and co-wrote the paper. R.A.D. analysed data.
 

### 2015-07-01 manuscript.Rnw
Minor edits based on NCC reviews.

--- 2015-02-09 manuscript.Rnw
@@
 The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\cite{ise2010}.
-Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass determines turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}.
+Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}.
 Globally, forests store approximately 360Pg of carbon in living biomass\citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}.
@@
 S1).
-Previous work based on whole stands\citep{oneilldeangelis1981,enquistniklas2002,reich2014} and very small plants\citep{poorter2012}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
+Previous work based on whole stands\citep{oneilldeangelis1981,enquistniklas2002,reich2014} and individual plants\citep{poorter2012}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
 Here we extend this finding to show that LMF differs not only across the angiosperm-gymnosperm divide, but also within angiosperms, at the level of individual plants, and across the entire size range of woody plants in natural forests (0.1-100m) (Fig.
@@
 Leaf area is relevant because it drives total light interception and thus photosynthesis.
-Several classic studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
+Several classic studies have assumed, based only on differences in measurements of leaf mass, that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
 We found no evidence for this, instead we found that the three PFTs maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR).
@@
 We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves).
-For conifers, leaf area was converted to half-total surface area using published conversion factors \citep{barclay2000}, which is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}.
+For conifers, leaf area was converted to half-total surface area using the average of a set of published conversion factors \citep{barclay2000}.
+We used a different conversion for pines (\emph{Pinus} spp.) than non-pines, because the conversion factors were very different for these two groups \citep{barclay2000}.
+This conversion was necessary because half-total surface area is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}.
 Stem cross-sectional area was measured either at breast height (ca.
@@
 
-Figure 1a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary. We thank Martin de Kauwe for discussions and comments on an earlier version of this paper.
+Figure 1a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary.
+Our sincere thanks to everyone who contributed data to the Biomass and Allometry Database.
+We thank Martin de Kauwe for discussions and comments on an earlier version of this paper.
 
@@
     \caption{\textbf{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf construction cost.
-} (a) Leaf mass fraction (LMF = leaf mass / above-ground biomass) by PFT.
+} (a) Leaf mass fraction ($M_F/M_T$ = leaf mass / above-ground biomass) by PFT.
 Each symbol is an individual plant.
 Lines are generalized additive model fits.
-(b) and (c) Leaf mass fraction (LMF) and leaf area ratio (LAR) at the average plant height in the dataset, estimated from panel (a).
+(b) and (c) Leaf mass fraction and leaf area ratio ($A_F/M_T$) at the average plant height in the dataset, estimated from panel (a).
 (d) Average leaf mass per unit basal stem area, and (e) leaf area per unit basal stem area for the three major PFTs confirm that the between-PFT variation in leaf mass fraction is due to leaf mass per unit basal area.

### 2015-07-01 manuscript.Rnw
minor edits to text

@@
 Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
-Using a new global database on biomass partitioning from over 175 studies\citep{Falster2015}, we here show a strong effect of PFT on partitioning between stems and leaves.
+Using a new global database on biomass partitioning from over 175 studies\citep{Falster2015}, we here show that partitioning between stems and leaves varies strongly between PFTs (as defined by deciduous vs.
+evergreen and gymnosperm vs.
+angiosperm).
 We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass.
@@
 The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010}, assumed to differ in key physiological attributes.
-While GVMs assume differences in biomass partitioning between PFTs (Methods \ref{sec:gvmsummary}), these predictions are constrained by limited empirical data.
+While GVMs assume or predict differences in biomass partitioning between PFTs (Methods \ref{sec:gvmsummary}), these differences are constrained by limited empirical data.
 Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
@@
 Several classic studies have assumed, based only on differences in measurements of leaf mass, that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
-We found no evidence for this, instead we found that the three PFTs maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR).
+We found no evidence for this, instead we found that the three PFTs maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR), despite large differences in the leaf mass fraction.
 
@@
 
-Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$), a quantity closely related to hydraulic supply capacity \cite{enquist2003}.
+Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$), a quantity  related to hydraulic supply capacity \cite{enquist2003}.
 This ratio is also important because plant leaf mass is frequently estimated from $A_S$ with allometric equations \cite{shinozaki1964b, chave2005}.
@@
 
-The finding that climate does not appreciably affect biomass partitioning is good news for models that parameterize the globe by PFT.
-Yet, these results contrast those of a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}.
+The finding that climate does not appreciably affect biomass partitioning contrast those of a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}.
 One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands.
@@
 S5).
-It is thus pertinent to understand whether we can move beyond a PFT-based classification by considering other traits that affect biomass partitioning.
+It is thus pertinent to understand whether we can improve on a PFT-based classification by including other traits that affect biomass partitioning.
 To do this, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq.
@@
 Overall, our results establish general patterns about plant construction and thus lay an empirical base against which models can be benchmarked.
-We found that PFTs differed in LMF, and this was proportional to LMA.
+We found that PFTs differed in LMF, and that LMF was proportional to LMA.
 It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.

### 2015-07-01 manuscript.Rnw
minor edits to text

@@
 A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions.
-Based on our findings, a first approximation within GVMs would be to assume leaf area to stem cross-sectional area a parameter that does not vary between PFTs, and vary biomass allocation accordingly.
+Based on our findings, a first approximation within GVMs would be to assume leaf area to stem cross-sectional area a parameter that does not vary between PFTs (notwithstanding effects of plant height on this variable), and vary biomass allocation accordingly.
 LMA, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem area.

### 2015-07-16 manuscript.Rnw
reformatted to 'normal' ms for NewPhyt or whatever
Newphyt bibtex style complete
not bold first para

@@
 
-\usepackage[sort&compress]{natbib}
-\citestyle{nature}
+\usepackage{natbib}
+\bibliographystyle{newphyt}
 
@@
 
-%--------------------------------------------------------------------------------------------%
-%% Main text. Max words = 2000, first para = 200.
+\section*{Abstract}
 
-% First paragraph The first paragraph should be 180 to 200 words and should contain references.
-The paragraph should serve both as a general introduction to the topic and as a brief, non-technical summary of your main results and their implications.
-Ideally, the paragraph starts with two to three sentences of background explaining the topic and current understanding.
-This is followed by a phrase like “Here we present” to introduce the new work.
-In the second part, the main methods, findings, conclusions and implications should be described clearly and accessibly.
-
-\section*{First paragraph}
-
-\textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time and storage of carbon in ecosystems, and the response of net primary production to rising CO\textsubscript{2}\citep{dekauwe2014,friend2014,ise2010}.
+The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time and storage of carbon in ecosystems, and the response of net primary production to rising CO\textsubscript{2}\citep{dekauwe2014,friend2014,ise2010}.
 Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
@@
 2a).
+
+\section*{Materials and Methods}
+
+We used the recently compiled Biomass And Allometry Database (BAAD) \citep{Falster2015}, including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
+1.3m) ($A_{Sbh}$).
+For each analysis, we used different subsets of the data.
+Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4}.
+We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
+We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms.
+We excluded  deciduous gymnosperms because few data were available.
+All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes.
+To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).
+
+For the above-ground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total above-ground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,
+
+$$\textrm{LMF} = \frac{M_F}{M_T} = \left(\frac{M_F}{A_F}\right) \left(\frac{A_F}{M_T}\right) = \textrm{LMA} \times \textrm{LAR}.$$
+
+We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves).
+For conifers, leaf area was converted to half-total surface area using the average of a set of published conversion factors \citep{barclay2000}.
+We used a different conversion for pines (\emph{Pinus} spp.) than non-pines, because the conversion factors were very different for these two groups \citep{barclay2000}.
+This conversion was necessary because half-total surface area is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}.
+Stem cross-sectional area was measured either at breast height (ca.
+1.3m) and/or at the base of the plant.
+For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using a non-linear regression model (see~\ref{sec:predas}).
+
+We used generalized additive models (GAM) to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height.
+In all fitted GAMs, we used species-dataset combination as a random effect.
+We used the \emph{mgcv} package in R \citep{wood2006}.
+All variables (except MAP and MAT) were log-transformed prior to analysis.
+
+Variance explained by quantitative climate variables (MAP and MAT) were tested with GAMs where variables were sequentially added to the model, and the explained variance ($R^2$) calculated.
+For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect.
+We calculated the $R^2$ for linear mixed-effects models for the fixed effects only \citep{nakagawa2013}.
+
+\section*{Results}
 
@@
 
+\section*{Discussion}
+
+The finding that climate does not appreciably affect biomass partitioning contrast those of a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}.
+One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands.
+This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants.
+With the caveats that there is still variation between species and studies unaccounted for (Fig.
+S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}).
+
 The finding that climate does not appreciably affect biomass partitioning contrast those of a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}.
@@
 
-\section*{Methods}
-
-We used the recently compiled Biomass And Allometry Database (BAAD) \citep{Falster2015}, including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
-1.3m) ($A_{Sbh}$).
-For each analysis, we used different subsets of the data.
-Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4}.
-We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
-We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms.
-We excluded  deciduous gymnosperms because few data were available.
-All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes.
-To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).
-
-For the above-ground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total above-ground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,
-
-$$\textrm{LMF} = \frac{M_F}{M_T} = \left(\frac{M_F}{A_F}\right) \left(\frac{A_F}{M_T}\right) = \textrm{LMA} \times \textrm{LAR}.$$
-
-We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves).
-For conifers, leaf area was converted to half-total surface area using the average of a set of published conversion factors \citep{barclay2000}.
-We used a different conversion for pines (\emph{Pinus} spp.) than non-pines, because the conversion factors were very different for these two groups \citep{barclay2000}.
-This conversion was necessary because half-total surface area is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}.
-Stem cross-sectional area was measured either at breast height (ca.
-1.3m) and/or at the base of the plant.
-For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using a non-linear regression model (see~\ref{sec:predas}).
-
-We used generalized additive models (GAM) to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height.
-In all fitted GAMs, we used species-dataset combination as a random effect.
-We used the \emph{mgcv} package in R \citep{wood2006}.
-All variables (except MAP and MAT) were log-transformed prior to analysis.
-
-Variance explained by quantitative climate variables (MAP and MAT) were tested with GAMs where variables were sequentially added to the model, and the explained variance ($R^2$) calculated.
-For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect.
-We calculated the $R^2$ for linear mixed-effects models for the fixed effects only \citep{nakagawa2013}.
-
 \section{Acknowledgments}
@@
 We thank Martin de Kauwe for discussions and comments on an earlier version of this paper.
-
-\section{Author contributions}
-
-R.A.D and D.S.F. conceived the ideas and methods, and co-wrote the paper. R.A.D. analysed data.
 
@@
 \section{Figures}
-
-% Figures A Nature Climate Change Letter can contain up to four display items (any combination of figures and/or tables).
-Figures should be sized in single or double column format.
-Multi-panel figures should be arranged so that the figure forms a neat rectangle.
 

### 2015-07-20 manuscript.Rnw
More conversion to newphyt; dummy new paragraphs added

@@
 
-The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\cite{ise2010}.
-Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}.
+The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\citep{ise2010}.
+Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \citep{ryanlaw2005}.
 Globally, forests store approximately 360Pg of carbon in living biomass\citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}.
@@
 
+% paragraph on what we do know about differences between PFTs; what we might expect (LMF vs. LAR; LMA and leaf lifespan differences)
+Previous work based on whole stands\citep{oneilldeangelis1981,enquistniklas2002,reich2014} and individual plants\citep{poorter2012}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
+However, it is unclear whether PFTs with higher leaf biomass also have higher total leaf area.
+Leaf area is relevant because it drives total light interception and thus photosynthesis.
+Several classic studies have assumed, based only on differences in measurements of leaf mass, that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
+
+%paragraph on LMF, it is an interesting variable because it is an important component of turnover of biomass. Link to leaf lifespan across PFTs as well. 
+For the above-ground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total above-ground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,
+
+$$\textrm{LMF} = \frac{M_F}{M_T} = \left(\frac{M_F}{A_F}\right) \left(\frac{A_F}{M_T}\right) = \textrm{LMA} \times \textrm{LAR}.$$
+
+For LMF we hypothesize that it is higher for EG than DA, with EA taking an intermediate value as determined by its mean LMA.
+This would be consistent both with leaf lifespan correlate as well as minimizing variation in leaf area per unit biomass (which determines light interception).
+
+We also wonder whether differences in LMF and LAR (and its component variables) are sensitive to climate.
+One may expect this to be the case because short-term experiments have shown responsiveness of biomass partitioning to soil water etc.
+etc.; or this variation may wash out at a large scale.
+It is thus the question whether variation within PFT can be explained by climate or not.
+Indeed, many GVMs assume that climate affects biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Methods \ref{sec:gvmsummary}).
+
 Here, we use a unique, new database\citep{Falster2015} (Fig.
@@
 S1).
-Previous work based on whole stands\citep{oneilldeangelis1981,enquistniklas2002,reich2014} and individual plants\citep{poorter2012}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
-Here we extend this finding to show that LMF differs not only across the angiosperm-gymnosperm divide, but also within angiosperms, at the level of individual plants, and across the entire size range of woody plants in natural forests (0.1-100m) (Fig.
-2a).
+
+One key challenge is to account for the very large size variation in any allometric variable, also LMF.
+We do this by fitting an assumption-free model (GAM) and studying PFT patterns at a common height.
 
 \section*{Materials and Methods}
+
+\subsection*{Data}
 
@@
 
-For the above-ground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total above-ground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,
-
-$$\textrm{LMF} = \frac{M_F}{M_T} = \left(\frac{M_F}{A_F}\right) \left(\frac{A_F}{M_T}\right) = \textrm{LMA} \times \textrm{LAR}.$$
-
 We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves).
@@
 For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using a non-linear regression model (see~\ref{sec:predas}).
+
+\subsection*{Data analysis}
 
@@
 
+Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons across climatic gradients, due to the small sample size within each species or site.
+To study climate effects on biomass partitioning, we further decomposed LMF as the product,
+
+\begin{equation}\label{eq:M1}
+\textrm{LMF} = \left(\frac{M_F}{A_S}\right) \left(\frac{A_S}{M_T}\right).
+\end{equation}
+
+We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation, MAP; mean annual temperature, MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig.
+S1).
+
+For Eq.\ref{eq:M1} we need the basal diameter, but this is not normally measured for large plants.
+Using the subset of data where basal stem diameter ($A_S$) and diameter at breast height ($A_{Sbh}$) were measured (a total of \Sexpr{length(residuals(basalafit))} observations covering the three major PFTs), we developed a non-linear model to predict $A_S$ when only $A_{Sbh}$ and plant height ($H$) were measured.
+We fit the following equation, using non-linear regression.
+
+$$
+D_S = D_{Sbh} H^a / (H - H_{bh})^a
+$$
+
+where $D_S$ is the basal stem diameter (m), $D_{Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{Sbh}$ was measured (typically 1.3 or 1.34m), and $a$ was further expressed as a function of plant height:
+
+$$
+a = c_0H^{c_1}
+$$
+
+The estimated coefficients were $c_0$ = \Sexpr{signif(coef(basalafit)[1],3)}, $c_1$ = \Sexpr{signif(coef(basalafit)[2],3)}, root-mean square error = \Sexpr{signif(summary(basalafit)$sigma,3)}.
+
+\subsection*{Terminology}
+\label{sec:partterm}
+
+The terms 'partitioning' and 'allocation' have been used in various ways, confusing comparisons between studies \citep{litton2007}.
+Here, we define biomass partitioning as the actual distribution of biomass between compartments (e.g., leaves vs.
+stems), and biomass allocation as the proportion of net primary production (NPP) that is allocated to some compartment.
+The two processes are different because of continuous turnover of biomass, which differs strongly between compartments.
+We may write \citep{McMurtrieWolf1983},
+
+$$
+\frac{dM_i}{dt} = a_i \textrm{NPP} - \tau_i M_i
+$$
+
+where $M_i$ is the biomass in some compartment (leaves, stems or roots), $a_i$ the annual allocation of NPP to that compartment, and $\tau_i$ the annual turnover of compartment $i$.
+It is thus easy to see that partitioning ($M_i/M_T$, where $M_T$ is total biomass) can be different from allocation ($a_i$) because turnover ($\tau_i$) differs between leaf and wood biomass.
+
 \section*{Results}
-
-Moreover, we show that differences in LMF among woody PFTs arise due to differences in leaf morphology, but they converge in the amount of leaf area held on the plant (Fig.
-2b-c).
-Previously, it was unclear whether PFTs with higher leaf biomass also had higher total leaf area.
-Leaf area is relevant because it drives total light interception and thus photosynthesis.
-Several classic studies have assumed, based only on differences in measurements of leaf mass, that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
-We found no evidence for this, instead we found that the three PFTs maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR), despite large differences in the leaf mass fraction.
 
@@
 
-Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$), a quantity  related to hydraulic supply capacity \cite{enquist2003}.
-This ratio is also important because plant leaf mass is frequently estimated from $A_S$ with allometric equations \cite{shinozaki1964b, chave2005}.
+Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$).
+This ratio is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}.
 Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig.
@@
 
-The above results highlight how LMA -- a central trait of the leaf economics spectrum (LES)\citep{wrightetal2004} -- drives differences in biomass partitioning among dominant woody PFTs.
+The above results highlight how LMA drives differences in biomass partitioning among dominant woody PFTs.
 To assess whether LMA could replace PFT when predicting biomass partitioning, we fit a linear mixed-effects model to LMF and $M_F/A_S$ using PFT and plant height (and the quadratic term) as predictors (and all interactions).
@@
 These results suggest promising avenues for parameterizing and simplifying biomass partitioning routines in GVMs.
-
-Many GVMs assume that climate influences biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Methods \ref{sec:gvmsummary}).
-We were thus interested to know whether additional variation in biomass partitioning could be explained by climate.
-Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons across climatic gradients, due to the small sample size within each species or site.
-Thus, we decomposed LMF as the product,
-
-\begin{equation}\label{eq:M1}
-\textrm{LMF} = \left(\frac{M_F}{A_S}\right) \left(\frac{A_S}{M_T}\right).
-\end{equation}
-
-An advantage of eq.
-\ref{eq:M1} is that the second term absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}) and is otherwise fairly conserved between species\citep{chave2005} and across PFTs (Fig.
-S4).
-As shown in Fig.
-2b and Fig.
-2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}).
-$M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.
-
-We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation, MAP; mean annual temperature, MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig.
-S1).
-Regardless of how we analyzed the data, we found only weak climatic effects on biomass partitioning within PFTs (Fig.
-3).
-Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table \ref{table/varpartB}).
-Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) (Fig.
-3b-c).
-These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table \ref{table/varpartMAPMAT}).
-
-\section*{Discussion}
-
-The finding that climate does not appreciably affect biomass partitioning contrast those of a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}.
-One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands.
-This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants.
-With the caveats that there is still variation between species and studies unaccounted for (Fig.
-S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}).
-
-The finding that climate does not appreciably affect biomass partitioning contrast those of a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}.
-One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands.
-This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants.
-With the caveats that there is still variation between species and studies unaccounted for (Fig.
-S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}).
 
@@
 We thus show that in a PFT-based classification, LMA is a good first estimate of biomass partitioning, however, more variation can be explained by another trait ($A_F/A_S$), which varies appreciably between species.
+
+We found that decomposing as in Eq.
+\ref{eq:M1} was very useful to study climate effects because the second term absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}) and is otherwise fairly conserved across PFTs (Fig.
+S4).
+As shown in Fig.
+2b and Fig.
+2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}).
+$M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.
+
+Regardless of how we analyzed the data, we found only weak climatic effects on biomass partitioning within PFTs (Fig.
+3).
+Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table \ref{table/varpartB}).
+Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) (Fig.
+3b-c).
+These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table \ref{table/varpartMAPMAT}).
+
+\section*{Discussion}
+
+We found that PFTs differed in LMF, and that LMF was proportional to LMA.
+This was true especially across the three PFTs studied, but also to a lesser extent across species within PFT.
+The consequence is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) did not differ between PFTs, and was much less variable within PFTs than LMF.
+This result is particularly convincing because we studied the entire size range of woody plants in natural forests (0.1-100m) (Fig.
+2a).
+Other work has already demonstrated a large difference in LMF between angiosperms and gymnosperms, but have not linked this directly to differences in LMA, nor was able to tell if it is because of angiosperm-gymnosperm divide or because of deciduous vs.
+evergreen habit.
+We show that LMA -- a central trait of the leaf economics spectrum (LES)\citep{wrightetal2004} -- explains these differences in a consistent manner.
+
+paragraph - Why would LAR be less variable than LMF?
+Because light interception.
+Lulu effect paper.
+Need to discuss linkage between LMA, leaf lifespan, leaf area vs.
+leaf mass at given MT (and given height!).
+
+paragraph - variation in LMF was the same as MF/AS, suggesting that MT/AS is a fairly conserved quantity \ref{eq:M2}.
+Consistent with \citep{chave2005} who showed that aboveground biomass can be well estimated with Ht*AS*wood density.
+This works because, for large plants, most of aboveground biomass is wood.
+
+Climate (temperature, rainfall) did not appreciably affect biomass partitioning between leaves and stems.
+This finding is in contrast with a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}.
+One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands.
+This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants.
+With the caveats that there is still variation between species and studies unaccounted for (Fig.
+S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}).
 
@@
 \section{Figures}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureS1.pdf}
+    \caption{\textbf{Global coverage of the climate space by the dataset.} Methods are as in Fig. 1 in main text, symbols are colored by vegetation type as reported by the data contributors.}
+\label{fig:figureS1}
+\end{figure}
 
@@
 \setcounter{figure}{0}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureS1.pdf}
-    \caption{\textbf{Global coverage of the climate space by the dataset.} Methods are as in Fig. 1 in main text, symbols are colored by vegetation type as reported by the data contributors.}
-\label{fig:figureS1}
-\end{figure}
 
@@
 We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them.
-We base this classification on a recent model intercomparison study \cite{dekauwe2014}.
+We base this classification on a recent model intercomparison study \citep{dekauwe2014}.
 Note that we use 'partioning' to indicate the actual distribution of biomass among plant components, and 'allocation' the fraction of net primary production to each component (see Methods \ref{sec:partterm}).
@@
 Some models assume that a constant fraction of net primary production (NPP) is allocated to each component (leaves, stems, roots).
-CABLE, EALCO \cite{dekauwe2014} and IBIS \cite{foley1996} are examples of models that allocate a PFT-specific fraction of NPP to each plant component each timestep  (though, for both models, alternative model formulations have been used as well).
+CABLE, EALCO \citep{dekauwe2014} and IBIS \citep{foley1996} are examples of models that allocate a PFT-specific fraction of NPP to each plant component each timestep  (though, for both models, alternative model formulations have been used as well).
 Constant allocation fractions still give rise to differences in biomass partitioning between PFTs (see Methods \ref{sec:partterm}), since turnover of leaves (determined by LL) will always differ between at least deciduous and evergreen PFTs.
@@
 In this class of models,  allometric (or other size-dependent) relationships between biomass components are used to calculate biomass allocation in each timestep.
-For example, in ED2 \cite{medvigy2009}, biomass partitioning to leaves, stem sapwood and roots is constrained by a set of equations with PFT-specific parameters, estimated through model calibration.
+For example, in ED2 \citep{medvigy2009}, biomass partitioning to leaves, stem sapwood and roots is constrained by a set of equations with PFT-specific parameters, estimated through model calibration.
 Biomass allocation is then calculated at each timestep to satisfy these equations.
 Other models use leaf area as a constraint.
-For example, LPJ \cite{sitchetal2003} uses PFT-specific values of leaf area to sapwood area ratio to constrain estimates of biomass allocation.
+For example, LPJ \citep{sitchetal2003} uses PFT-specific values of leaf area to sapwood area ratio to constrain estimates of biomass allocation.
 Further differences in biomass partitioning between PFTs are then the result of differences in LMA (leaf mass at a given leaf area), as well as LL (shorter lived leaves require higher annual allocation of biomass for upkeep).
@@
 
-Few GVMs use a scheme in which carbon allocation is determined by assuming biomass fractions are the result of optimization, however, this scheme is widely used in smaller-scale vegetation models \cite{valentinemakela2012,franklin2012}.
-One example of a global model is SDGVM \cite{woodward1995}, which optimizes leaf area index by considering the carbon balance of the lowest leaf layer in the canopy.
+Few GVMs use a scheme in which carbon allocation is determined by assuming biomass fractions are the result of optimization, however, this scheme is widely used in smaller-scale vegetation models \citep{valentinemakela2012,franklin2012}.
+One example of a global model is SDGVM \citep{woodward1995}, which optimizes leaf area index by considering the carbon balance of the lowest leaf layer in the canopy.
 As with LPJ described above, differences in biomass partitioning between PFTs then arise as a result of differences in LMA as well as LL.
@@
 Models differ widely in how climate directly (or indirectly) affects biomass partitioning beyond differences between PFTs.
-In one class of models, biomass partitioning is modelled based on relative resource limitations, consistent with the idea that biomass allocation proceeds to the plant component most limited by the resource it acquires \cite{friedlingstein1999}.
-For example, in the allocation model of Arora and Boer \cite{aroraboer2005}, which is used in the ISAM model \cite{dekauwe2014}, allocation to leaves, stems and roots depends on relative limitations of light and water.
+In one class of models, biomass partitioning is modelled based on relative resource limitations, consistent with the idea that biomass allocation proceeds to the plant component most limited by the resource it acquires \citep{friedlingstein1999}.
+For example, in the allocation model of Arora and Boer \citep{aroraboer2005}, which is used in the ISAM model \citep{dekauwe2014}, allocation to leaves, stems and roots depends on relative limitations of light and water.
 Likewise, Friedlingstein et al.
-\cite{friedlingstein1999} developed a resource-limitation framework for the CASA model, where allocation depends on limitations of light, water and nutrients.
+\citep{friedlingstein1999} developed a resource-limitation framework for the CASA model, where allocation depends on limitations of light, water and nutrients.
 It is clear that these types of models predict apparent climate effects on biomass partitioning along temperature and moisture gradients.
@@
 Some models do not assume effects of stress on biomass allocation directly, but differences in allocation may result from effects of stress on biomass production, and subsequent priority rules for the allocation of biomass.
-For example, in LPJ \cite{sitchetal2003}, biomass production may be low enough in very dry years so that intended allocation to leaves and stem sapwood cannot be supported.
+For example, in LPJ \citep{sitchetal2003}, biomass production may be low enough in very dry years so that intended allocation to leaves and stem sapwood cannot be supported.
 In this case, allocation proceeds to leaves only.
 In dry climates, where water stress will be more common, this would gradually lead to different partitioning of biomass to leaves and stems.
-
-\subsection{Biomass partitioning terminology}
-\label{sec:partterm}
-
-The terms 'partitioning' and 'allocation' have been used in various ways, confusing comparisons between studies \citep{litton2007}.
-Here, we define biomass partitioning as the actual distribution of biomass between compartments (e.g., leaves vs.
-stems), and biomass allocation as the proportion of net primary production (NPP) that is allocated to some compartment.
-The two processes are different because of continuous turnover of biomass, which differs strongly between compartments.
-We may write (cf.
-Ref.
-\citep{McMurtrieWolf1983}),
-
-$$
-\frac{dM_i}{dt} = a_i \textrm{NPP} - \tau_i M_i
-$$
-
-where $M_i$ is the biomass in some compartment (leaves, stems or roots), $a_i$ the annual allocation of NPP to that compartment, and $\tau_i$ the annual turnover of compartment $i$.
-It is thus easy to see that partitioning ($M_i/M_T$, where $M_T$ is total biomass) can be different from allocation ($a_i$) because turnover ($\tau_i$) differs between leaf and wood biomass.
-
-\subsection{Predicting basal diameter}
-\label{sec:predas}
-
-Using the subset of data where basal stem diameter ($A_S$) and diameter at breast height ($A_{Sbh}$) were measured (a total of \Sexpr{length(residuals(basalafit))} observations covering the three major PFTs), we developed a non-linear model to predict $A_S$ when only $A_{Sbh}$ and plant height ($H$) were measured.
-We fit the following equation, using non-linear regression.
-
-$$
-D_S = D_{Sbh} H^a / (H - H_{bh})^a
-$$
-
-where $D_S$ is the basal stem diameter (m), $D_{Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{Sbh}$ was measured (typically 1.3 or 1.34m), and $a$ was further expressed as a function of plant height:
-
-$$
-a = c_0H^{c_1}
-$$
-
-The estimated coefficients were $c_0$ = \Sexpr{signif(coef(basalafit)[1],3)}, $c_1$ = \Sexpr{signif(coef(basalafit)[2],3)}, root-mean square error = \Sexpr{signif(summary(basalafit)$sigma,3)}.
 

### 2015-07-20 manuscript.Rnw
more writing

@@
 
-The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time and storage of carbon in ecosystems, and the response of net primary production to rising CO\textsubscript{2}\citep{dekauwe2014,friend2014,ise2010}.
+The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs).
 Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
-Using a new global database on biomass partitioning from over 175 studies\citep{Falster2015}, we here show that partitioning between stems and leaves varies strongly between PFTs (as defined by deciduous vs.
+Using a new global database on biomass partitioning from over 175 studies, we here show that partitioning between stems and leaves varies strongly between PFTs (as defined by deciduous vs.
 evergreen and gymnosperm vs.
@@
 We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass.
-Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum \citep{wrightetal2004}.
+Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum.
 In addition, we show how a second trait, leaf area per stem basal area, explains additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs.
@@
 The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\citep{ise2010}.
+It affects the residence time and storage of carbon in ecosystems, and the response of net primary production to rising CO\textsubscript{2}\citep{dekauwe2014,friend2014,ise2010}.
 Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \citep{ryanlaw2005}.
@@
 
-%paragraph on LMF, it is an interesting variable because it is an important component of turnover of biomass. Link to leaf lifespan across PFTs as well. 
-For the above-ground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total above-ground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,
+Despite considerable advances in allometric theory (enquist etc.
+etc.) we don't have a clear picture of differences between PFTs.
+Instead, most of the work has focussed on understanding size-related shifts in biomass partitioning.
+Longer leaf lifespan would logically lead to higher LMF (as argued by enquist, niklas) only if the annual allocation does not vary with leaf lifespan, but there is no a priori reason why this should be so.
+We take the view that a broader evidence base is needed to elucidate patterns which can subsequently be used to refine and improve existing models of allometry.
+
+In this study we focus on the partitioning of above-ground biomass between short-lived leaves and long-lives woody pools.
+We are also interested in a related variable describing the amount of leaf area per unit aboveground biomass, which is more closely related to light interception and hence productivity.
+These variables are related as follows,
 
@@
 
+Where $M_T$ is total above-ground biomass, $M_F$ total leaf biomass, $A_F$ total leaf area, the leaf mass fraction is $M_F / M_T$, and leaf area ratio is $A_F / M_T$.
 For LMF we hypothesize that it is higher for EG than DA, with EA taking an intermediate value as determined by its mean LMA.
@@
 One may expect this to be the case because short-term experiments have shown responsiveness of biomass partitioning to soil water etc.
-etc.; or this variation may wash out at a large scale.
-It is thus the question whether variation within PFT can be explained by climate or not.
+etc.
+Alternatively observed variation in LMF across the globe may be more closely related to PFT.
 Indeed, many GVMs assume that climate affects biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Methods \ref{sec:gvmsummary}).
+Approaches taken by these models differ widely, thus it is pertinent that global-scale patterns are described to decide between the different assumptions.
 
@@
 The dataset includes over 21,000 observations of biomass and size metrics for individual plants, contributed by 175 studies across nine vegetation types (Fig.
-S1).
-
-One key challenge is to account for the very large size variation in any allometric variable, also LMF.
+1).
+Since we are here interested in the differences between PFTs and relation to climate, we do not study the dependence on size per se.
 We do this by fitting an assumption-free model (GAM) and studying PFT patterns at a common height.
@@
 \subsection*{Terminology}
-\label{sec:partterm}
 
 The terms 'partitioning' and 'allocation' have been used in various ways, confusing comparisons between studies \citep{litton2007}.
-Here, we define biomass partitioning as the actual distribution of biomass between compartments (e.g., leaves vs.
-stems), and biomass allocation as the proportion of net primary production (NPP) that is allocated to some compartment.
+Here, we define biomass \emph{partitioning} as the actual distribution of biomass between compartments (e.g., leaves vs.
+stems), and biomass \emph{allocation} as the proportion of net primary production (NPP) that is allocated to some compartment.
 The two processes are different because of continuous turnover of biomass, which differs strongly between compartments.
@@
 We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}.
-We then found that LMF at a common height was proportional to the leaf construction cost, as measured by leaf mass per area (LMA) (Fig.
+We then found that LMF at a common height was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig.
 2b).
@@
 2c and S3).
-
-Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$).
-This ratio is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}.
 Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig.
@@
 This works because, for large plants, most of aboveground biomass is wood.
+Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$).
+This ratio is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}.
 
@@
     \includegraphics[width=0.99\textwidth]{Figure1.pdf}
-    \caption{\textbf{Overview of the global biomass and allometry database.} (a) Variables were measured on \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species.
+    \caption{Overview of the global biomass and allometry database.
+(a) Variables were measured on \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species.
 (b) Coverage of the dataset across global climate space.
@@
     \includegraphics[width=0.99\textwidth]{Figure2.pdf}
-    \caption{\textbf{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf construction cost.
-} (a) Leaf mass fraction ($M_F/M_T$ = leaf mass / above-ground biomass) by PFT.
+    \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf construction cost.
+(a) Leaf mass fraction ($M_F/M_T$ = leaf mass / above-ground biomass) by PFT.
 Each symbol is an individual plant.
@@
     \includegraphics[width=0.99\textwidth]{Figure3.pdf}
-    \caption{\textbf{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning}.
+    \caption{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning.
 a) Leaf mass per unit stem cross-sectional area ($M_F/A_S$) for each PFT within three different biomes.
@@
 We base this classification on a recent model intercomparison study \citep{dekauwe2014}.
-Note that we use 'partioning' to indicate the actual distribution of biomass among plant components, and 'allocation' the fraction of net primary production to each component (see Methods \ref{sec:partterm}).
+Note that we use 'partioning' to indicate the actual distribution of biomass among plant components, and 'allocation' the fraction of net primary production to each component (see Methods).
 Finally, we briefly summarize how current GVMs include effects of climate on biomass partitioning.
@@
 CABLE, EALCO \citep{dekauwe2014} and IBIS \citep{foley1996} are examples of models that allocate a PFT-specific fraction of NPP to each plant component each timestep  (though, for both models, alternative model formulations have been used as well).
-Constant allocation fractions still give rise to differences in biomass partitioning between PFTs (see Methods \ref{sec:partterm}), since turnover of leaves (determined by LL) will always differ between at least deciduous and evergreen PFTs.
+Constant allocation fractions still give rise to differences in biomass partitioning between PFTs (see Methods in main text), since turnover of leaves (determined by LL) will always differ between at least deciduous and evergreen PFTs.
 

### 2015-07-21 manuscript.Rnw
References fixed; more added and text fixed in intro

--- 2015-07-20 manuscript.Rnw
@@
 \usepackage{fixltx2e}
-\usepackage[default,osfigures,scale=0.95]{opensans}
+% \usepackage[default,osfigures,scale=0.95]{opensans}
 \graphicspath{ {./figures/} }
@@
 
-% paragraph on what we do know about differences between PFTs; what we might expect (LMF vs. LAR; LMA and leaf lifespan differences)
-Previous work based on whole stands\citep{oneilldeangelis1981,enquistniklas2002,reich2014} and individual plants\citep{poorter2012}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
-However, it is unclear whether PFTs with higher leaf biomass also have higher total leaf area.
-Leaf area is relevant because it drives total light interception and thus photosynthesis.
+In this paper we are primarily interested in the distribution of biomass ('partitioning') between leaves and stems, an important component of the residence time of carbon in ecosystems \citep{friend2014}.
+Previous work based on whole stands \citep{oneilldeangelis1981,enquistniklas2002,reich2014} and individual plants \citep{poorter2012}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
+\citet{poorter2012} and \citet{enquist2002} attributed higher LMF in gymnosperms to longer leaf lifespan (LL) compared to typical angiosperms, but this begs the question whether differences in LMF are also apparent between deciduous and evergreen functional types within angiosperms.
+It is also unknown whether PFTs with higher leaf biomass also have higher total leaf area, which is relevant because it drives total light interception and thus productivity.
 Several classic studies have assumed, based only on differences in measurements of leaf mass, that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
 
-%paragraph on LMF, it is an interesting variable because it is an important component of turnover of biomass. Link to leaf lifespan across PFTs as well. 
-For the above-ground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total above-ground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,
-
-$$\textrm{LMF} = \frac{M_F}{M_T} = \left(\frac{M_F}{A_F}\right) \left(\frac{A_F}{M_T}\right) = \textrm{LMA} \times \textrm{LAR}.$$
-
-For LMF we hypothesize that it is higher for EG than DA, with EA taking an intermediate value as determined by its mean LMA.
-This would be consistent both with leaf lifespan correlate as well as minimizing variation in leaf area per unit biomass (which determines light interception).
-
-We also wonder whether differences in LMF and LAR (and its component variables) are sensitive to climate.
-One may expect this to be the case because short-term experiments have shown responsiveness of biomass partitioning to soil water etc.
-etc.; or this variation may wash out at a large scale.
+Little is known about global-scale patterns in LMF and LAR in relation to climate.
+One may expect that biomass partitioning is correlated with precipitation or mean annual temperature \citep{reich2014} because short-term experiments (cite) and climate gradients \citep{callaway1994,poyatos2007} have shown responsiveness of biomass partitioning and related traits.
 It is thus the question whether variation within PFT can be explained by climate or not.
@@
 
-Here, we use a unique, new database\citep{Falster2015} (Fig.
+Here, we use a unique, new database \citep{Falster2015} (Fig.
 1) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms.
-The dataset includes over 21,000 observations of biomass and size metrics for individual plants, contributed by 175 studies across nine vegetation types (Fig.
-S1).
-
-One key challenge is to account for the very large size variation in any allometric variable, also LMF.
-We do this by fitting an assumption-free model (GAM) and studying PFT patterns at a common height.
+The dataset includes observations of biomass and size metrics for individual plants, contributed by 175 studies across nine vegetation types (Fig.
+1).
+In this paper we include field-grown plants only, including the entire size range of woody plants (0.01 - \textgreater 100m height).
+One key challenge is thus to account for the very large size variation commonly found in any allometric variable \citep{niklas1994}.
+We do this by fitting a statistical model that does no assume a functional dependence on plant size, allowing us to study PFT patterns at a common plant height.
 
@@
 1.3m) and/or at the base of the plant.
-For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using a non-linear regression model (see~\ref{sec:predas}).
+For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using a non-linear regression model, as follows.
+
+Using the subset of data where basal stem diameter ($A_S$) and diameter at breast height ($A_{Sbh}$) were measured (a total of \Sexpr{length(residuals(basalafit))} observations covering the three major PFTs), we developed a non-linear model to predict $A_S$ when only $A_{Sbh}$ and plant height ($H$) were measured.
+We fit the following equation, using non-linear regression.
+
+$$
+D_S = D_{Sbh} H^a / (H - H_{bh})^a
+$$
+
+where $D_S$ is the basal stem diameter (m), $D_{Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{Sbh}$ was measured (typically 1.3 or 1.34m), and $a$ was further expressed as a function of plant height:
+
+$$
+a = c_0H^{c_1}
+$$
+
+The estimated coefficients were $c_0$ = \Sexpr{signif(coef(basalafit)[1],3)}, $c_1$ = \Sexpr{signif(coef(basalafit)[2],3)}, root-mean square error = \Sexpr{signif(summary(basalafit)$sigma,3)}.
 
@@
 S1).
-
-For Eq.\ref{eq:M1} we need the basal diameter, but this is not normally measured for large plants.
-Using the subset of data where basal stem diameter ($A_S$) and diameter at breast height ($A_{Sbh}$) were measured (a total of \Sexpr{length(residuals(basalafit))} observations covering the three major PFTs), we developed a non-linear model to predict $A_S$ when only $A_{Sbh}$ and plant height ($H$) were measured.
-We fit the following equation, using non-linear regression.
-
-$$
-D_S = D_{Sbh} H^a / (H - H_{bh})^a
-$$
-
-where $D_S$ is the basal stem diameter (m), $D_{Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{Sbh}$ was measured (typically 1.3 or 1.34m), and $a$ was further expressed as a function of plant height:
-
-$$
-a = c_0H^{c_1}
-$$
-
-The estimated coefficients were $c_0$ = \Sexpr{signif(coef(basalafit)[1],3)}, $c_1$ = \Sexpr{signif(coef(basalafit)[2],3)}, root-mean square error = \Sexpr{signif(summary(basalafit)$sigma,3)}.
 
@@
 
+Poorter et al 2012 did not find a clear relationship between LMF and LMA across 8 PFTs, perhaps not enough data?
+Also, we only include field-grown plants, they have everything.
+It also seems they did not account for plant size in that analysis.
+
 paragraph - Why would LAR be less variable than LMF?
@@
 leaf mass at given MT (and given height!).
+One option is that LAR is targeted to be more constant, because it drives light interception.
+Very high leaf areas lead to very shaded foliage (etc.).
+Other option is that this arises through the correlation with leaf lifepan.
+This is how Enquist \& Niklas explained it, but it is only true when all else is equal.
+In support, \citet{reich1992} showed positive correlation between leaf life span and total stand leaf biomass across diverse forest stands.
+Our results are certainly consistent with longer leaf lifespan leading to higher leaf biomass, because of the strong nearly 1/x correlation between LL and LMA ()
 
 paragraph - variation in LMF was the same as MF/AS, suggesting that MT/AS is a fairly conserved quantity \ref{eq:M2}.
-Consistent with \citep{chave2005} who showed that aboveground biomass can be well estimated with Ht*AS*wood density.
+Consistent with \citep{chave2005} who showed that aboveground biomass can be well estimated with c*Ht*AS*wood density (works because geometry).
 This works because, for large plants, most of aboveground biomass is wood.

### 2015-07-21 manuscript.Rnw
Split supporting information into separate doc (also in remake.yml)

@@
 
+\renewcommand{\baselinestretch}{1.2}   % line spacing
 \usepackage{amssymb,amsmath}
@@
 
-We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves).
+For the above-ground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total above-ground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,
+
+$$\textrm{LMF} = \frac{M_F}{M_T} = \left(\frac{M_F}{A_F}\right) \left(\frac{A_F}{M_T}\right) = \textrm{LMA} \times \textrm{LAR}.$$
+
+We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves, see \citet{Falster2015} for methods details for each contributed study).
 For conifers, leaf area was converted to half-total surface area using the average of a set of published conversion factors \citep{barclay2000}.
@@
 
-Figure 1a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary.
+Figure 2a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary.
 Our sincere thanks to everyone who contributed data to the Biomass and Allometry Database.
@@
 \clearpage
-\section{Supporting Information Tables}
-
-\renewcommand\thetable{S\arabic{table}}
-\setcounter{table}{0}
-
-<<results='asis'>>=
-print(xtable(table_gamr2MATMAP, label="table/varpartMAPMAT",
-      caption="\\textbf{Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.} Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),
-      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
-@
-
-<<tablepreload1>>=
-printit <- function(x, caption, label=NULL){
-  print(xtable(x, align="rlll|l",
-               caption=caption, label=label
-               ), hline.after=c(0,3), caption.placement="top")
-
-}
-@
-
-<<results='asis'>>=
-printit(table_mlfastba,
-        label="table:count1",
-        caption="\\textbf{Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$.} Numbers denote individual plants, with number of species in parentheses.
-DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
-@
-
-<<results='asis'>>=
-printit(table_alfastba,
-        label="table:count2",
-        caption="\\textbf{Number of observations used in the analysis of $A_F/A_S$.} Numbers denote individual plants, with number of species in parentheses.
-DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
-@
-
-<<results='asis'>>=
-printit(table_lmf,
-        label="table:count3",
-        caption="\\textbf{Number of observations used in the analysis of leaf mass fraction ($M_F/M_T$) and leaf area ratio ($A_F/M_T$).} Numbers denote individual plants, with number of species in parentheses.
-DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
-@
-
-<<results='asis'>>=
-printit(table_lar,
-        label="table:count4",
-        caption="\\textbf{Number of observations used in the analysis of leaf area ratio (LAR).} Numbers denote individual plants, with number of species in parentheses.
-DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
-@
-
-\clearpage
-\section{Supporting Information Methods}
-
-\renewcommand\thesubsection{S\arabic{subsection}}
-\setcounter{subsection}{0}
-
-\subsection{Modelling of biomass partitioning in global vegetation models (GVMs)}
-\label{sec:gvmsummary}
-
-Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types.
-However, there is no consensus on how these differences are parameterized, or whether they are linked to inherent differences in leaf mass per area (LMA) or leaf lifespan (LL).
-Typically, parameters related to biomass allocation differ between PFTs (often based on little empirical evidence), which then lead to differences in biomass partitioning.
-We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them.
-We base this classification on a recent model intercomparison study \citep{dekauwe2014}.
-Note that we use 'partioning' to indicate the actual distribution of biomass among plant components, and 'allocation' the fraction of net primary production to each component (see Methods \ref{sec:partterm}).
-Finally, we briefly summarize how current GVMs include effects of climate on biomass partitioning.
-
-\subsubsection*{Constant allocation fractions}
-
-Some models assume that a constant fraction of net primary production (NPP) is allocated to each component (leaves, stems, roots).
-CABLE, EALCO \citep{dekauwe2014} and IBIS \citep{foley1996} are examples of models that allocate a PFT-specific fraction of NPP to each plant component each timestep  (though, for both models, alternative model formulations have been used as well).
-Constant allocation fractions still give rise to differences in biomass partitioning between PFTs (see Methods \ref{sec:partterm}), since turnover of leaves (determined by LL) will always differ between at least deciduous and evergreen PFTs.
-
-\subsubsection*{Allometric constraints}
-
-In this class of models,  allometric (or other size-dependent) relationships between biomass components are used to calculate biomass allocation in each timestep.
-For example, in ED2 \citep{medvigy2009}, biomass partitioning to leaves, stem sapwood and roots is constrained by a set of equations with PFT-specific parameters, estimated through model calibration.
-Biomass allocation is then calculated at each timestep to satisfy these equations.
-Other models use leaf area as a constraint.
-For example, LPJ \citep{sitchetal2003} uses PFT-specific values of leaf area to sapwood area ratio to constrain estimates of biomass allocation.
-Further differences in biomass partitioning between PFTs are then the result of differences in LMA (leaf mass at a given leaf area), as well as LL (shorter lived leaves require higher annual allocation of biomass for upkeep).
-
-\subsubsection*{Optimality}
-
-Few GVMs use a scheme in which carbon allocation is determined by assuming biomass fractions are the result of optimization, however, this scheme is widely used in smaller-scale vegetation models \citep{valentinemakela2012,franklin2012}.
-One example of a global model is SDGVM \citep{woodward1995}, which optimizes leaf area index by considering the carbon balance of the lowest leaf layer in the canopy.
-As with LPJ described above, differences in biomass partitioning between PFTs then arise as a result of differences in LMA as well as LL.
-
-\subsubsection*{Climate effects}
-
-Models differ widely in how climate directly (or indirectly) affects biomass partitioning beyond differences between PFTs.
-In one class of models, biomass partitioning is modelled based on relative resource limitations, consistent with the idea that biomass allocation proceeds to the plant component most limited by the resource it acquires \citep{friedlingstein1999}.
-For example, in the allocation model of Arora and Boer \citep{aroraboer2005}, which is used in the ISAM model \citep{dekauwe2014}, allocation to leaves, stems and roots depends on relative limitations of light and water.
-Likewise, Friedlingstein et al.
-\citep{friedlingstein1999} developed a resource-limitation framework for the CASA model, where allocation depends on limitations of light, water and nutrients.
-It is clear that these types of models predict apparent climate effects on biomass partitioning along temperature and moisture gradients.
-
-Some models do not assume effects of stress on biomass allocation directly, but differences in allocation may result from effects of stress on biomass production, and subsequent priority rules for the allocation of biomass.
-For example, in LPJ \citep{sitchetal2003}, biomass production may be low enough in very dry years so that intended allocation to leaves and stem sapwood cannot be supported.
-In this case, allocation proceeds to leaves only.
-In dry climates, where water stress will be more common, this would gradually lead to different partitioning of biomass to leaves and stems.
-
-\clearpage
 \bibliography{references}

### 2015-07-21 manuscript_suppinfo.Rnw
Split supporting information into separate doc (also in remake.yml)

\documentclass{article}

\usepackage{amssymb,amsmath}
\usepackage{palatino}
\usepackage{geometry}
\usepackage{authblk}
\usepackage{gensymb}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}

\usepackage{parskip}
\usepackage{natbib}
\bibliographystyle{newphyt}

\begin{document}

\clearpage
\section*{Supporting Information}

\renewcommand\thesubsection{S\arabic{subsection}}
\setcounter{subsection}{0}

\subsection{Sample sizes by plant functional type}

\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}

<<preload, echo=FALSE>>=
library(knitr)
library(xtable)
opts_chunk$set(echo=FALSE)
@

<<results='asis'>>=
print(xtable(table_gamr2MATMAP, label="table/varpartMAPMAT",
      caption="\\textbf{Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.} Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),
      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
@

<<tablepreload1>>=
printit <- function(x, caption, label=NULL){
  print(xtable(x, align="rlll|l",
               caption=caption, label=label
               ), hline.after=c(0,3), caption.placement="top")

}
@

<<results='asis'>>=
printit(table_mlfastba,
        label="table:count1",
        caption="\\textbf{Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$.} Numbers denote individual plants, with number of species in parentheses. DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@

<<results='asis'>>=
printit(table_alfastba,
        label="table:count2",
        caption="\\textbf{Number of observations used in the analysis of $A_F/A_S$.} Numbers denote individual plants, with number of species in parentheses. DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@

<<results='asis'>>=
printit(table_lmf,
        label="table:count3",
        caption="\\textbf{Number of observations used in the analysis of leaf mass fraction ($M_F/M_T$) and leaf area ratio ($A_F/M_T$).} Numbers denote individual plants, with number of species in parentheses. DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@

<<results='asis'>>=
printit(table_lar,
        label="table:count4",
        caption="\\textbf{Number of observations used in the analysis of leaf area ratio (LAR).} Numbers denote individual plants, with number of species in parentheses. DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@

\clearpage
\subsection{Modelling of biomass partitioning in global vegetation models (GVMs)}
\label{sec:gvmsummary}

Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types. However, there is no consensus on how these differences are parameterized, or whether they are linked to inherent differences in leaf mass per area (LMA) or leaf lifespan (LL). Typically, parameters related to biomass allocation differ between PFTs (often based on little empirical evidence), which then lead to differences in biomass partitioning. We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them. We base this classification on a recent model intercomparison study \citep{dekauwe2014}. Note that we use 'partitioning' to indicate the actual distribution of biomass among plant components, and 'allocation' the fraction of net primary production to each component (see Methods in main text). Finally, we briefly summarize how current GVMs include effects of climate on biomass partitioning.

\subsubsection*{Constant allocation fractions}

Some models assume that a constant fraction of net primary production (NPP) is allocated to each component (leaves, stems, roots). CABLE, EALCO \citep{dekauwe2014} and IBIS \citep{foley1996} are examples of models that allocate a PFT-specific fraction of NPP to each plant component each timestep  (though, for both models, alternative model formulations have been used as well). Constant allocation fractions still give rise to differences in biomass partitioning between PFTs (see Methods in main text), since turnover of leaves (determined by LL) will always differ between at least deciduous and evergreen PFTs.

\subsubsection*{Allometric constraints}

In this class of models,  allometric (or other size-dependent) relationships between biomass components are used to calculate biomass allocation in each timestep. For example, in ED2 \citep{medvigy2009}, biomass partitioning to leaves, stem sapwood and roots is constrained by a set of equations with PFT-specific parameters, estimated through model calibration. Biomass allocation is then calculated at each timestep to satisfy these equations. Other models use leaf area as a constraint. For example, LPJ \citep{sitchetal2003} uses PFT-specific values of leaf area to sapwood area ratio to constrain estimates of biomass allocation. Further differences in biomass partitioning between PFTs are then the result of differences in LMA (leaf mass at a given leaf area), as well as LL (shorter lived leaves require higher annual allocation of biomass for upkeep).

\subsubsection*{Optimality}

Few GVMs use a scheme in which carbon allocation is determined by assuming biomass fractions are the result of optimization, however, this scheme is widely used in smaller-scale vegetation models \citep{valentinemakela2012,franklin2012}. One example of a global model is SDGVM \citep{woodward1995}, which optimizes leaf area index by considering the carbon balance of the lowest leaf layer in the canopy. As with LPJ described above, differences in biomass partitioning between PFTs then arise as a result of differences in LMA as well as LL.

\subsubsection*{Climate effects}

Models differ widely in how climate directly (or indirectly) affects biomass partitioning beyond differences between PFTs. In one class of models, biomass partitioning is modelled based on relative resource limitations, consistent with the idea that biomass allocation proceeds to the plant component most limited by the resource it acquires \citep{friedlingstein1999}. For example, in the allocation model of Arora and Boer \citep{aroraboer2005}, which is used in the ISAM model \citep{dekauwe2014}, allocation to leaves, stems and roots depends on relative limitations of light and water. Likewise, Friedlingstein et al. \citep{friedlingstein1999} developed a resource-limitation framework for the CASA model, where allocation depends on limitations of light, water and nutrients. It is clear that these types of models predict apparent climate effects on biomass partitioning along temperature and moisture gradients.

Some models do not assume effects of stress on biomass allocation directly, but differences in allocation may result from effects of stress on biomass production, and subsequent priority rules for the allocation of biomass. For example, in LPJ \citep{sitchetal2003}, biomass production may be low enough in very dry years so that intended allocation to leaves and stem sapwood cannot be supported. In this case, allocation proceeds to leaves only. In dry climates, where water stress will be more common, this would gradually lead to different partitioning of biomass to leaves and stems.

\bibliography{references}

\end{document}

### 2015-07-21 manuscript.Rnw
Figures moved to SuppInfo; references special characters fixed; figure caption correct formatting

@@
 
-\renewcommand{\baselinestretch}{1.2}   % line spacing
+\renewcommand{\baselinestretch}{1.6}   % line spacing
 \usepackage{amssymb,amsmath}
@@
 \usepackage{lineno}
-\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
+\geometry{verbose,tmargin=2.4cm,bmargin=2.4cm,lmargin=2.4cm,rmargin=2.4cm}
+\usepackage{natbib}
 
-\usepackage{natbib}
 \bibliographystyle{newphyt}
+\renewcommand{\figurename}{Fig.}
+\usepackage[labelfont=bf, labelsep=quad]{caption}
 
@@
 
-The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\citep{ise2010}.
+The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation \citep{ise2010}.
 Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \citep{ryanlaw2005}.
-Globally, forests store approximately 360Pg of carbon in living biomass\citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}.
-Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014}.
+Globally, forests store approximately 360Pg of carbon in living biomass \citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}.
+Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014,negron2015}.
 
@@
 Previous work based on whole stands \citep{oneilldeangelis1981,enquistniklas2002,reich2014} and individual plants \citep{poorter2012}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
-\citet{poorter2012} and \citet{enquist2002} attributed higher LMF in gymnosperms to longer leaf lifespan (LL) compared to typical angiosperms, but this begs the question whether differences in LMF are also apparent between deciduous and evergreen functional types within angiosperms.
+\citet{poorter2012} and \citet{enquistniklas2002} attributed higher LMF in gymnosperms to longer leaf lifespan (LL) compared to typical angiosperms, but this begs the question whether differences in LMF are also apparent between deciduous and evergreen functional types within angiosperms.
 It is also unknown whether PFTs with higher leaf biomass also have higher total leaf area, which is relevant because it drives total light interception and thus productivity.
@@
 The dataset includes observations of biomass and size metrics for individual plants, contributed by 175 studies across nine vegetation types (Fig.
-1).
+S1).
 In this paper we include field-grown plants only, including the entire size range of woody plants (0.01 - \textgreater 100m height).
@@
 
-\section{Acknowledgments}
+\section*{Acknowledgments}
 
@@
 \clearpage
-\section{Tables}
+\section*{Tables}
 
@@
 print(xtable(table_varpart2, label="table/varpartB",
-             caption="\\textbf{Explained variance by plant functional type (PFT), plant height (H) and biome (B), for five whole-plant variables.} Each variable was added to a linear mixed-effects model, using species within dataset as a random effect.
+             caption="Explained variance by plant functional type (PFT), plant height (H) and biome (B), for five whole-plant variables.
+Each variable was added to a linear mixed-effects model, using species within dataset as a random effect.
 The $R^2$ shown is that explained by the fixed effects only.
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureS1.pdf}
-    \caption{\textbf{Global coverage of the climate space by the dataset.} Methods are as in Fig. 1 in main text, symbols are colored by vegetation type as reported by the data contributors.}
-\label{fig:figureS1}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
     \includegraphics[width=0.99\textwidth]{Figure1.pdf}
-    \caption{\textbf{Overview of the global biomass and allometry database.} (a) Variables were measured on \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species.
+    \caption{Overview of the global biomass and allometry database.
+(a) Variables were measured on \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species.
 (b) Coverage of the dataset across global climate space.
@@
     \includegraphics[width=0.99\textwidth]{Figure4.pdf}
-     \caption{\textbf{Leaf and stem traits capture variation in biomass partitioning across species.} Individual plant data were averaged by species and study combinations.
+     \caption{Leaf and stem traits capture variation in biomass partitioning across species.
+Individual plant data were averaged by species and study combinations.
 Lines are linear regressions with 95\% confidence intervals.
@@
 \clearpage
-\section{Supporting Information Figures}
-
-\renewcommand\thefigure{S\arabic{figure}}
-\setcounter{figure}{0}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
-    \caption{\textbf{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$).} Each point is an individual plant.}
-    \label{fig:figureS2}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureS3.pdf}
-    \caption{\textbf{Leaf area ratio ($A_F/M_T$) by PFT.} Each symbol is an individual plant.
-Lines are generalized additive model fits using species/dataset as a random effect.
-Shaded areas are approximate 95\% confidence intervals.}
-    \label{fig:figureS3}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
-    \caption{\textbf{Relationship between above-ground biomass and basal stem area, each point is an individual plant.} Lines are spline fits.
-The right panel shows least-square means of the ratio of above-ground biomass to basal stem area, estimated from a linear mixed-effects model with basal stem area (linear and squared terms) and PFT.
-}
-    \label{fig:figureS4}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
-    \caption{\textbf{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.} Shown are histograms (as probability density functions) of leaf mass per area (an index of leaf construction cost) ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types.
-Arrows indicate means by PFT.}
-    \label{fig:figures5}
-\end{figure}
-
-\clearpage
 \bibliography{references}

### 2015-07-21 manuscript_suppinfo.Rnw
Figures moved to SuppInfo; references special characters fixed; figure caption correct formatting

@@
 \geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
-
+\graphicspath{ {./figures/} }
 \usepackage{parskip}
 \usepackage{natbib}
+
 \bibliographystyle{newphyt}
+\renewcommand{\figurename}{Fig.}
+\usepackage[labelfont=bf, labelsep=quad]{caption}
 
@@
 print(xtable(table_gamr2MATMAP, label="table/varpartMAPMAT",
-      caption="\\textbf{Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.} Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),
+      caption="Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.
+Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),
       include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
@@
         label="table:count1",
-        caption="\\textbf{Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$.} Numbers denote individual plants, with number of species in parentheses.
+        caption="Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$.
+Numbers denote individual plants, with number of species in parentheses.
 DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@@
         label="table:count2",
-        caption="\\textbf{Number of observations used in the analysis of $A_F/A_S$.} Numbers denote individual plants, with number of species in parentheses.
+        caption="Number of observations used in the analysis of $A_F/A_S$.
+Numbers denote individual plants, with number of species in parentheses.
 DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@@
         label="table:count3",
-        caption="\\textbf{Number of observations used in the analysis of leaf mass fraction ($M_F/M_T$) and leaf area ratio ($A_F/M_T$).} Numbers denote individual plants, with number of species in parentheses.
+        caption="Number of observations used in the analysis of leaf mass fraction ($M_F/M_T$) and leaf area ratio ($A_F/M_T$).
+Numbers denote individual plants, with number of species in parentheses.
 DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
@@
         label="table:count4",
-        caption="\\textbf{Number of observations used in the analysis of leaf area ratio (LAR).} Numbers denote individual plants, with number of species in parentheses.
+        caption="Number of observations used in the analysis of leaf area ratio (LAR).
+Numbers denote individual plants, with number of species in parentheses.
 DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
 @
+
+\clearpage
+\subsection{Supporting figures}
+
+\renewcommand\thefigure{S\arabic{figure}}
+\setcounter{figure}{0}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureS1.pdf}
+    \caption{Global coverage of the climate space by the dataset. Methods are as in Fig. 1 in main text, symbols are colored by vegetation type as reported by the data contributors.}
+\label{fig:figureS1}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
+    \caption{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$). Each point is an individual plant.}
+    \label{fig:figureS2}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureS3.pdf}
+    \caption{Leaf area ratio ($A_F/M_T$) by PFT.
+Each symbol is an individual plant.
+Lines are generalized additive model fits using species/dataset as a random effect.
+Shaded areas are approximate 95\% confidence intervals.}
+    \label{fig:figureS3}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
+    \caption{Relationship between above-ground biomass and basal stem area, each point is an individual plant.
+Lines are spline fits.
+The right panel shows least-square means of the ratio of above-ground biomass to basal stem area, estimated from a linear mixed-effects model with basal stem area (linear and squared terms) and PFT.
+}
+    \label{fig:figureS4}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
+    \caption{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.
+Shown are histograms (as probability density functions) of leaf mass per area (an index of leaf construction cost) ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types.
+Arrows indicate means by PFT.}
+    \label{fig:figures5}
+\end{figure}
 

### 2015-07-21 manuscript.Rnw
More writing!; Formatting is now complete, I think.

@@
 The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010}, assumed to differ in key physiological attributes.
-While GVMs assume or predict differences in biomass partitioning between PFTs (Methods \ref{sec:gvmsummary}), these differences are constrained by limited empirical data.
+While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are constrained by limited empirical data.
 Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
@@
 Little is known about global-scale patterns in LMF and LAR in relation to climate.
-One may expect that biomass partitioning is correlated with precipitation or mean annual temperature \citep{reich2014} because short-term experiments (cite) and climate gradients \citep{callaway1994,poyatos2007} have shown responsiveness of biomass partitioning and related traits.
-It is thus the question whether variation within PFT can be explained by climate or not.
-Indeed, many GVMs assume that climate affects biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Methods \ref{sec:gvmsummary}).
+It can be expected that biomass partitioning is correlated with precipitation or mean annual temperature \citep{reich2014} because short-term experiments (cite) and climate gradients \citep{callaway1994,poyatos2007} have shown responsiveness of biomass partitioning to environmental drivers.
+A recent study \citep{reich2014} also demonstrated that stand-scale biomass partitioning was related to mean annual temperature across diverse forest stands.
+Indeed, many GVMs assume that climate affects biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Notes S1).
+Again, however, these models do not agree on the degree of plasticity in biomass partitioning, or which variables should it respond to.
 
@@
 \section*{Materials and Methods}
-
-\subsection*{Data}
-
-We used the recently compiled Biomass And Allometry Database (BAAD) \citep{Falster2015}, including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca.
-1.3m) ($A_{Sbh}$).
-For each analysis, we used different subsets of the data.
-Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4}.
-We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
-We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms.
-We excluded  deciduous gymnosperms because few data were available.
-All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes.
-To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).
-
-For the above-ground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total above-ground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,
-
-$$\textrm{LMF} = \frac{M_F}{M_T} = \left(\frac{M_F}{A_F}\right) \left(\frac{A_F}{M_T}\right) = \textrm{LMA} \times \textrm{LAR}.$$
-
-We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves, see \citet{Falster2015} for methods details for each contributed study).
-For conifers, leaf area was converted to half-total surface area using the average of a set of published conversion factors \citep{barclay2000}.
-We used a different conversion for pines (\emph{Pinus} spp.) than non-pines, because the conversion factors were very different for these two groups \citep{barclay2000}.
-This conversion was necessary because half-total surface area is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}.
-Stem cross-sectional area was measured either at breast height (ca.
-1.3m) and/or at the base of the plant.
-For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using a non-linear regression model, as follows.
-
-Using the subset of data where basal stem diameter ($A_S$) and diameter at breast height ($A_{Sbh}$) were measured (a total of \Sexpr{length(residuals(basalafit))} observations covering the three major PFTs), we developed a non-linear model to predict $A_S$ when only $A_{Sbh}$ and plant height ($H$) were measured.
-We fit the following equation, using non-linear regression.
-
-$$
-D_S = D_{Sbh} H^a / (H - H_{bh})^a
-$$
-
-where $D_S$ is the basal stem diameter (m), $D_{Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{Sbh}$ was measured (typically 1.3 or 1.34m), and $a$ was further expressed as a function of plant height:
-
-$$
-a = c_0H^{c_1}
-$$
-
-The estimated coefficients were $c_0$ = \Sexpr{signif(coef(basalafit)[1],3)}, $c_1$ = \Sexpr{signif(coef(basalafit)[2],3)}, root-mean square error = \Sexpr{signif(summary(basalafit)$sigma,3)}.
-
-\subsection*{Data analysis}
-
-We used generalized additive models (GAM) to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height.
-In all fitted GAMs, we used species-dataset combination as a random effect.
-We used the \emph{mgcv} package in R \citep{wood2006}.
-All variables (except MAP and MAT) were log-transformed prior to analysis.
-
-Variance explained by quantitative climate variables (MAP and MAT) were tested with GAMs where variables were sequentially added to the model, and the explained variance ($R^2$) calculated.
-For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect.
-We calculated the $R^2$ for linear mixed-effects models for the fixed effects only \citep{nakagawa2013}.
-
-Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons across climatic gradients, due to the small sample size within each species or site.
-To study climate effects on biomass partitioning, we further decomposed LMF as the product,
-
-\begin{equation}\label{eq:M1}
-\textrm{LMF} = \left(\frac{M_F}{A_S}\right) \left(\frac{A_S}{M_T}\right).
-\end{equation}
-
-We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation, MAP; mean annual temperature, MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig.
-S1).
 
@@
 
+\subsection*{Data}
+
+We used the recently compiled Biomass And Allometry Database (BAAD) \citep{Falster2015}, including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$).
+For each analysis, we used different subsets of the data.
+Sample sizes by PFT are summarized in Tables S2-S5.
+We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
+We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms.
+We excluded  deciduous gymnosperms because few data were available.
+All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes.
+To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).
+
+For the above-ground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total above-ground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,
+
+$$\textrm{LMF} = \frac{M_F}{M_T} = \left(\frac{M_F}{A_F}\right) \left(\frac{A_F}{M_T}\right) = \textrm{LMA} \times \textrm{LAR}.$$
+
+We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves, see \citet{Falster2015} for methods details for each contributed study).
+For conifers, leaf area was converted to half-total surface area using the average of a set of published conversion factors \citep{barclay2000}.
+We used a different conversion for pines (\emph{Pinus} spp.) than non-pines, because the conversion factors were very different for these two groups \citep{barclay2000}.
+This conversion was necessary because half-total surface area is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}.
+Stem cross-sectional area was measured either at breast height and/or at the base of the plant.
+For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using a non-linear regression model, as follows.
+
+Using the subset of data where basal stem diameter ($A_S$) and diameter at breast height ($A_{Sbh}$) were measured (a total of \Sexpr{length(residuals(basalafit))} observations covering the three major PFTs), we developed a non-linear model to predict $A_S$ when only $A_{Sbh}$ and plant height ($H$) were measured.
+We fit the following equation, using non-linear regression.
+
+$$D_S = D_{Sbh} H^a / (H - H_{bh})^a$$
+
+where $D_S$ is the basal stem diameter (m), $D_{Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{Sbh}$ was measured (typically 1.3 or 1.34m), and $a$ was further expressed as a function of plant height:
+
+$$a = c_0H^{c_1}$$
+
+The estimated coefficients were $c_0$ = \Sexpr{signif(coef(basalafit)[1],3)}, $c_1$ = \Sexpr{signif(coef(basalafit)[2],3)}, root-mean square error = \Sexpr{signif(summary(basalafit)$sigma,3)}.
+
+\subsection*{Data analysis}
+
+We used generalized additive models (GAM) to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height.
+In all fitted GAMs, we used a cubic regression spline.
+For the smoothed term in the model (plant height), we used up to 3 or 4 degrees of freedom, which resulted in biologically realistic smoothed relationships.
+Within the GAM, we used a penalized regression smoother [@Wood2006] to allow the final degree of smoothness to be estimated from the data.
+In all fitted GAMs, we used species-dataset combination as a random effect.We used the \emph{mgcv} package in R \citep{wood2006}.
+All variables (except MAP and MAT) were log-transformed prior to analysis.
+
+Variance explained by quantitative climate variables (MAP and MAT) were tested with GAMs where variables were sequentially added to the model, and the explained variance ($R^2$) calculated.
+For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect.
+We calculated the $R^2$ for linear mixed-effects models for the fixed effects only \citep{nakagawa2013}.
+
+Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons across climatic gradients, due to the small sample size within each species or site.
+To study climate effects on biomass partitioning, we further decomposed LMF as the product,
+
+\begin{equation}\label{eq:M1}
+\textrm{LMF} = \left(\frac{M_F}{A_S}\right) \left(\frac{A_S}{M_T}\right).
+\end{equation}
+
+We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation, MAP; mean annual temperature, MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig.
+S1).
+
 \section*{Results}
@@
 2a and S2).
-We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}.
+We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships.
 We then found that LMF at a common height was proportional to the leaf construction cost, as measured by leaf mass per area (LMA) (Fig.
 2b).
-As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig.
+As a consequence, LAR was invariant between PFTs (Eq.
+1) (Fig.
 2c and S3).
@@
 Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$).
-This ratio is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}.
 Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig.
@@
 0.28).
-This is surprising because LMA only captures aspects of leaf morphology, in indirectly leaf lifespan and photosynthetic capacity via the LES\citep{wrightetal2004}.
-These results suggest promising avenues for parameterizing and simplifying biomass partitioning routines in GVMs.
 
@@
 
+The ratio $M_F/A_S$ is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}.
 The correlation between $M_F/A_S$ and LMA was found to hold also within PFTs (Fig.
-4a), explaining 30\% of the variation in $M_F/A_S$ across species.
+3a), explaining 30\% of the variation in $M_F/A_S$ across species.
 The regressions across species by PFT were broad and overlapping, with a more general relationship extending across the entire LMA axis.
 A larger fraction of the variation in $M_F/A_S$ was explained by species-level differences in $A_F/A_S$ (Fig.
-4b).
+3b).
 As previously noted, $A_F/A_S$ does not differ systematically among PFTs (Fig.
@@
 In Fig.
-4b, The separation among PFTs in $M_F/A_S$ arises due to differences in LMA.
-We thus show that in a PFT-based classification, LMA is a good first estimate of biomass partitioning, however, more variation can be explained by another trait ($A_F/A_S$), which varies appreciably between species.
+3b, The separation among PFTs in $M_F/A_S$ arises due to differences in LMA.
 
 We found that decomposing as in Eq.
-\ref{eq:M1} was very useful to study climate effects because the second term absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}) and is otherwise fairly conserved across PFTs (Fig.
+\ref{eq:M1} was very useful to study climate effects because the second term absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB} and S1) and is otherwise fairly conserved across PFTs (Fig.
 S4).
@@
 2b and Fig.
-2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}).
+2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and S1.
 $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.
@@
 Regardless of how we analyzed the data, we found only weak climatic effects on biomass partitioning within PFTs (Fig.
-3).
+4).
 Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table \ref{table/varpartB}).
 Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) (Fig.
-3b-c).
-These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table \ref{table/varpartMAPMAT}).
+4b-c).
+These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table S1).
 
@@
 2a).
-Other work has already demonstrated a large difference in LMF between angiosperms and gymnosperms, but have not linked this directly to differences in LMA, nor was able to tell if it is because of angiosperm-gymnosperm divide or because of deciduous vs.
-evergreen habit.
+Previous studies have demonstrated a large difference in LMF between angiosperms and gymnosperms \citep{poorter2012}, but have not further separated functional types into deciduous vs.
+evergreen angiosperms, or provided an explanation in terms of leaf traits.
 We show that LMA -- a central trait of the leaf economics spectrum (LES)\citep{wrightetal2004} -- explains these differences in a consistent manner.
 
-Poorter et al 2012 did not find a clear relationship between LMF and LMA across 8 PFTs, perhaps not enough data?
-Also, we only include field-grown plants, they have everything.
-It also seems they did not account for plant size in that analysis.
+A previous large compilation of plant biomass partitioning \citep{poorter2012} (not restricted to field-grown plants) did not find a correlation between LMF and LMA across 8 PFTs (including non-woody PFTs).
+In their comparison, they did not correct for plant size, which has a very large effect on LMF (Fig.
+2).
+It is certainly necessary to correct for plant size when comparing biomass partitioning parameters across PFTs.
 
-paragraph - Why would LAR be less variable than LMF?
-Because light interception.
-Lulu effect paper.
-Need to discuss linkage between LMA, leaf lifespan, leaf area vs.
-leaf mass at given MT (and given height!).
-One option is that LAR is targeted to be more constant, because it drives light interception.
-Very high leaf areas lead to very shaded foliage (etc.).
-Other option is that this arises through the correlation with leaf lifepan.
-This is how Enquist \& Niklas explained it, but it is only true when all else is equal.
-In support, \citet{reich1992} showed positive correlation between leaf life span and total stand leaf biomass across diverse forest stands.
-Our results are certainly consistent with longer leaf lifespan leading to higher leaf biomass, because of the strong nearly 1/x correlation between LL and LMA ()
+The finding that LMF is proportional to LMA across PFTs implies that LAR does not differ.
+Why would this be the case?
+One explanation is that it is beneficial for plants to target a more constant LAR rather than LMF, because leaf area drives light interception.
+Because canopy size closely correlates with aboveground biomass \citep{duursma2010}, and the ratio of leaf area to canopy size determines self-shading and thus production efficiency of the foliage \citep{duursma20}, a low LAR is wasteful of woody biomass, and a high LAR wasteful of the foliage because self-shading becomes too high.
+Another, not mutually exclusive, explanation lies in the strong positive correlation between LMA and leaf lifespan (LL) \citep{wright2004}.
+Both \cite{poorter2012} and \citet{enquistniklas2002} explained higher LMF in gymnosperms by higher LL; plants simply maintain more foliage.
+However it is not not self-evident that else else is indeed equal, that is, annual production of foliage could differ between PFTs.
+Nonetheless, further supporting this explanation, \citet{reich1992} showed a positive correlation between LL and total stand leaf biomass across diverse forest stands.
 
-paragraph - variation in LMF was the same as MF/AS, suggesting that MT/AS is a fairly conserved quantity \ref{eq:M2}.
-Consistent with \citep{chave2005} who showed that aboveground biomass can be well estimated with c*Ht*AS*wood density (works because geometry).
-This works because, for large plants, most of aboveground biomass is wood.
+We found that, as for LMF, $M_F/A_S$ was also proportional to LMA across PFTs (Fig.
+2).
+Because LMF may be decomposed into the terms $M_F/A_S$ and $M_T/A_S$ (\ref{eq:M2}), this result suggests that $M_T/A_S$ does not vary between PFTs (see also Fig.
+S4).
+If this result holds, it means that aboveground biomass may be estimated across PFTs with a single equation.
+For tropical forests, \citet{chave2005} showed that aboveground biomass can be well estimated as,
+
+$$M_T = F \rho A_S H_T$$
+
+where F is a form factor, and $\rho$ wood density (kg m$^3$).
+In this equation leaf mass is subsumed in F, or assumed to be negligible (which hold for somewhat tall trees, see Fig.
+2).
+Our result seems to indicate something like this works for all PFTs (see Fig.
+S4).
+However, there are differences in wood density between angiosperms and gymnosperms (Zanne + TRY categories), does this mean F is different between PFTs?
+It is timely to datamine BAAD to develop better general equations to estimate aboveground biomass.
 
@@
 With the caveats that there is still variation between species and studies unaccounted for (Fig.
-S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}).
+S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and S1).
+
+We found that PFTs differed in LMF, and that LMF was proportional to LMA.
+It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.
+Indeed, nearly as much variation was explained across species when using LMA instead of PFT in a statistical model.
+This is surprising because LMA only captures aspects of leaf morphology, in indirectly leaf lifespan and photosynthetic capacity via the leaf economics spectrum \citep{wrightetal2004}.
+We thus show that in a PFT-based classification, LMA is a good first estimate of biomass partitioning, however, more variation can be explained by another trait ($A_F/A_S$), which varies appreciably between species.
+These results suggest promising avenues for parameterizing and simplifying biomass partitioning routines in GVMs.
 
 Overall, our results establish general patterns about plant construction and thus lay an empirical base against which models can be benchmarked.
-We found that PFTs differed in LMF, and that LMF was proportional to LMA.
-It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.
 A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions.
@@
 LMA, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem area.
-Some models already incorporate a similar algorithm (see Methods~\ref{sec:gvmsummary}), while other algorithms may be tuned to yield similar patterns between PFTs as we have presented.
+Some models already incorporate a similar algorithm (see Notes S1), while other algorithms may be tuned to yield similar patterns between PFTs as we have presented.
 In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{Falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz_towards_2012, dekauwe2014} of GVMs against empirical data.
@@
 
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
-    \caption{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning.
-a) Leaf mass per unit stem cross-sectional area ($M_F/A_S$) for each PFT within three different biomes.
-DA: deciduous angiosperms, EA: evergreen angiosperms, EG: evergreen gymnosperms.
-Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect.
-Letters denote significant differences (at $\alpha = 0.05$).
-b-c) Relationships between climate and $M_F/A_S$ within each PFT.
-Solid lines are generalized additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
-    \label{fig:figure3}
-\end{figure}
-
 <<echo=FALSE>>=
@@
 
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
+    \caption{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning.
+a) Leaf mass per unit stem cross-sectional area ($M_F/A_S$) for each PFT within three different biomes.
+DA: deciduous angiosperms, EA: evergreen angiosperms, EG: evergreen gymnosperms.
+Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect.
+Letters denote significant differences (at $\alpha = 0.05$).
+b-c) Relationships between climate and $M_F/A_S$ within each PFT.
+Solid lines are generalized additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
+    \label{fig:figure3}
+\end{figure}
+
 \clearpage
@@
 
+\section*{Supporting Information}
+Additional supporting information may be found in the online version of this article.\\
+
+\textbf{Table S1} Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.\\
+\textbf{Table S2} Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$.\\
+\textbf{Table S3} Number of observations used in the analysis of $A_F/A_S$.\\
+\textbf{Table S4} Number of observations used in the analysis of leaf mass fraction ($M_F/M_T$) and leaf area ratio ($A_F/M_T$).\\
+\textbf{Table S5} Number of observations used in the analysis of leaf area ratio (LAR).\\
+\textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\
+\textbf{Fig. S2} Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$).\\
+\textbf{Fig. S3} Leaf area ratio ($A_F/M_T$) by PFT.\\
+\textbf{Fig. S4} Relationship between above-ground biomass and basal stem area.\\
+\textbf{Fig. S5} Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.\\
+\textbf{Notes S1} Modelling of biomass partitioning in global vegetation models (GVMs)
+
 \end{document}

### 2015-07-21 manuscript_suppinfo.Rnw
More writing!; Formatting is now complete, I think.

@@
 \setcounter{subsection}{0}
-
-\subsection{Sample sizes by plant functional type}
 
@@
 \clearpage
-\subsection{Supporting figures}
 
@@
     \includegraphics[width=0.99\textwidth]{FigureS1.pdf}
-    \caption{Global coverage of the climate space by the dataset. Methods are as in Fig. 1 in main text, symbols are colored by vegetation type as reported by the data contributors.}
+    \caption{Global coverage of the climate space by the dataset, labelled by vegetation type.
+Methods for climate variables are as in Fig.
+1 in main text.
+Symbols are colored by vegetation type as reported by the data contributors.}
 \label{fig:figureS1}
@@
     \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
-    \caption{Relationship between above-ground biomass and basal stem area, each point is an individual plant.
+    \caption{Relationship between above-ground biomass and basal stem area.
+Each point is an individual plant.
 Lines are spline fits.
@@
 \clearpage
-\subsection{Modelling of biomass partitioning in global vegetation models (GVMs)}
-\label{sec:gvmsummary}
+\subsection*{Notes S1. Modelling of biomass partitioning in global vegetation models (GVMs)}
 

### 2015-07-21 manuscript.Rnw

@@
 The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation \citep{ise2010}.
+It affects the residence time and storage of carbon in ecosystems, and the response of net primary production to rising CO\textsubscript{2}\citep{dekauwe2014,friend2014,ise2010}.
 Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \citep{ryanlaw2005}.
@@
 These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared \citep{wolf2011}.
+
+Despite considerable advances in allometric theory (enquist etc.
+etc.) we don't have a clear picture of differences between PFTs.
+Instead, most of the work has focussed on understanding size-related shifts in biomass partitioning.
+Longer leaf lifespan would logically lead to higher LMF (as argued by enquist, niklas) only if the annual allocation does not vary with leaf lifespan, but there is no a priori reason why this should be so.
+We take the view that a broader evidence base is needed to elucidate patterns which can subsequently be used to refine and improve existing models of allometry.
 
@@
 2a and S2).
-We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships.
-We then found that LMF at a common height was proportional to the leaf construction cost, as measured by leaf mass per area (LMA) (Fig.
+We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}.
+We then found that LMF at a common height was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig.
 2b).
-As a consequence, LAR was invariant between PFTs (Eq.
-1) (Fig.
+As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig.
 2c and S3).
-
-Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$).
 Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig.

### 2015-07-23 manuscript.Rnw
Another revision. This one is very close, all paragraphs cleaned.

@@
 \usepackage[labelfont=bf, labelsep=quad]{caption}
+
+\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
 
@@
 The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs).
-Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking.
+Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical studies have not been able to justify these differences.
 Using a new global database on biomass partitioning from over 175 studies, we here show that partitioning between stems and leaves varies strongly between PFTs (as defined by deciduous vs.
@@
 Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
+
+\emph{Keywords}: allocation, plant allometry, biomass estimation, specific leaf area, dynamic global vegetation model
 
@@
 
-Despite considerable advances in allometric theory (enquist etc.
-etc.) we don't have a clear picture of differences between PFTs.
-Instead, most of the work has focussed on understanding size-related shifts in biomass partitioning.
-Longer leaf lifespan would logically lead to higher LMF (as argued by enquist, niklas) only if the annual allocation does not vary with leaf lifespan, but there is no a priori reason why this should be so.
-We take the view that a broader evidence base is needed to elucidate patterns which can subsequently be used to refine and improve existing models of allometry.
+Despite considerable advances in theory underlying allometric scaling in plants \citep{enquist2003, west1999, savage2010} we do not yet have a clear understanding of potential differences between PFTs in terms of biomass partitioning.
+Instead, most previous work has focussed on understanding size-related shifts in biomass partitioning, as governed by constraints including hydraulic supply and mechanical stability.
+Rather than to advance a new model of biomass partitioning or allocation, we take the view that a broader evidence base is needed to elucidate patterns between PFTs.
+It is then up to the individual GVMs to refine the algorithms and parameter values to more closely match the observations.
 
@@
 \citet{poorter2012} and \citet{enquistniklas2002} attributed higher LMF in gymnosperms to longer leaf lifespan (LL) compared to typical angiosperms, but this begs the question whether differences in LMF are also apparent between deciduous and evergreen functional types within angiosperms.
-It is also unknown whether PFTs with higher leaf biomass also have higher total leaf area, which is relevant because it drives total light interception and thus productivity.
-Several classic studies have assumed, based only on differences in measurements of leaf mass, that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
-
+It is also unknown whether PFTs with higher leaf biomass also have higher total leaf area, which is relevant because leaf area drives total light interception and thus productivity.
+Several classic, oft-cited, studies have assumed, based only on differences in measurements of leaf mass, that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
 Little is known about global-scale patterns in LMF and LAR in relation to climate.
-It can be expected that biomass partitioning is correlated with precipitation or mean annual temperature \citep{reich2014} because short-term experiments (cite) and climate gradients \citep{callaway1994,poyatos2007} have shown responsiveness of biomass partitioning to environmental drivers.
-A recent study \citep{reich2014} also demonstrated that stand-scale biomass partitioning was related to mean annual temperature across diverse forest stands.
+It can be expected that biomass partitioning is correlated with precipitation or mean annual temperature because experiments (cite) and comparisons across sites with contrasting climate \citep{callaway1994,poyatos2007} have shown responsiveness of biomass partitioning to environmental drivers.
+A recent study \citep{reich2014} demonstrated that stand-scale biomass partitioning was related to mean annual temperature across diverse forest stands.
 Indeed, many GVMs assume that climate affects biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Notes S1).
-Again, however, these models do not agree on the degree of plasticity in biomass partitioning, or which variables should it respond to.
+Again, however, these models do not agree on the degree of plasticity in biomass partitioning, or which climate variables it should respond to.
 
@@
 1) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms.
+A second objective was to study relationships between biomass partitioning and climate variables or biome at a global scale.
 The dataset includes observations of biomass and size metrics for individual plants, contributed by 175 studies across nine vegetation types (Fig.
-S1).
-In this paper we include field-grown plants only, including the entire size range of woody plants (0.01 - \textgreater 100m height).
+S1), across the three major biomes (boreal, temperate and tropical).
+In this paper we include field-grown woody plants only, including the entire size range of woody plants (0.01 - \textgreater 100m height).
 One key challenge is thus to account for the very large size variation commonly found in any allometric variable \citep{niklas1994}.
@@
 For the smoothed term in the model (plant height), we used up to 3 or 4 degrees of freedom, which resulted in biologically realistic smoothed relationships.
-Within the GAM, we used a penalized regression smoother [@Wood2006] to allow the final degree of smoothness to be estimated from the data.
-In all fitted GAMs, we used species-dataset combination as a random effect.We used the \emph{mgcv} package in R \citep{wood2006}.
+Within the GAM, we used a penalized regression smoother \citep{wood2006} to allow the final degree of smoothness to be estimated from the data.
+In all fitted GAMs, we used species-dataset combination as a random effect.
+We used the \emph{mgcv} package in R \citep{wood2006}.
 All variables (except MAP and MAT) were log-transformed prior to analysis.
@@
 
-We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation, MAP; mean annual temperature, MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig.
+because we found that the vast majority of size-related variation was captured by $A_S/M_T$ alone.
+Similarly, we decomposed LAR as the product of $A_F/A_S$ and $A_S/M_T$.
+We analyzed climate effects on $M_F/A_S$ and $A_F/A_S$ in two ways: with quantitative variables (mean annual precipitation, MAP; mean annual temperature, MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig.
 S1).
+In both cases we used PFT and plant height as covariates.
 
@@
 2a and S2).
-We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}.
-We then found that LMF at a common height was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig.
+When corrected for plant height by estimating LMF at a common plant height, we found that LMF was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig.
 2b).
@@
 
-We found that decomposing as in Eq.
-\ref{eq:M1} was very useful to study climate effects because the second term absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB} and S1) and is otherwise fairly conserved across PFTs (Fig.
+We found that decomposing LMF and LAR as shown in Eq.
+\ref{eq:M1} was very useful to study climate effects because the second term ($A_S/M_T$)absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB}, Table S1 and Fig.
+S4) and is otherwise fairly conserved across PFTs (Fig.
 S4).
@@
 2b and Fig.
-2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and S1.
+2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and Table S1).
 $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.
@@
 We found that PFTs differed in LMF, and that LMF was proportional to LMA.
-This was true especially across the three PFTs studied, but also to a lesser extent across species within PFT.
-The consequence is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) did not differ between PFTs, and was much less variable within PFTs than LMF.
-This result is particularly convincing because we studied the entire size range of woody plants in natural forests (0.1-100m) (Fig.
+This conclusion was particularly strong across the three PFTs studied, but it also held across species within PFT.
+The consequence is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) did not differ between PFTs.
+This result is particularly convincing because the data included individual plants from the entire size range of woody plants in natural forests (0.1-100m) (Fig.
 2a).
@@
 evergreen angiosperms, or provided an explanation in terms of leaf traits.
-We show that LMA -- a central trait of the leaf economics spectrum (LES)\citep{wrightetal2004} -- explains these differences in a consistent manner.
+We show that LMA -- a central trait of the leaf economics spectrum (LES) \citep{wrightetal2004} -- explains these differences in a consistent manner.
 
-A previous large compilation of plant biomass partitioning \citep{poorter2012} (not restricted to field-grown plants) did not find a correlation between LMF and LMA across 8 PFTs (including non-woody PFTs).
-In their comparison, they did not correct for plant size, which has a very large effect on LMF (Fig.
-2).
-It is certainly necessary to correct for plant size when comparing biomass partitioning parameters across PFTs.
+A previous large compilation of plant biomass partitioning \citep{poorter2012} (which included many non field-grown plants) did not find a correlation between LMF and LMA across 8 PFTs (including non-woody PFTs).
+However, they did not correct for plant size in their comparison.
+We found that, as expected, plant size strongly influences LMF (Fig.
+2) and LAR (Fig.
+S3).
+It is certainly necessary to correct for plant size when comparing biomass partitioning parameters across PFTs\footnote{some better ending to this para is needed - not just want to bash poorter's crappy analysis, is there a positive note we can end on?
+we suggest a method to do so?}.
 
-The finding that LMF is proportional to LMA across PFTs implies that LAR does not differ.
-Why would this be the case?
-One explanation is that it is beneficial for plants to target a more constant LAR rather than LMF, because leaf area drives light interception.
-Because canopy size closely correlates with aboveground biomass \citep{duursma2010}, and the ratio of leaf area to canopy size determines self-shading and thus production efficiency of the foliage \citep{duursma20}, a low LAR is wasteful of woody biomass, and a high LAR wasteful of the foliage because self-shading becomes too high.
-Another, not mutually exclusive, explanation lies in the strong positive correlation between LMA and leaf lifespan (LL) \citep{wright2004}.
-Both \cite{poorter2012} and \citet{enquistniklas2002} explained higher LMF in gymnosperms by higher LL; plants simply maintain more foliage.
-However it is not not self-evident that else else is indeed equal, that is, annual production of foliage could differ between PFTs.
+The finding that LMF is proportional to LMA across PFTs implies that LAR does not differ between PFTs.
+One explanation for this pattern lies in the strong positive correlation between LMA and leaf lifespan (LL) \citep{wright2004}.
+Both \cite{poorter2012} and \citet{enquistniklas2002} explained higher LMF in gymnosperms by higher LL; plants simply maintain more cohorts of foliage, but this explanation demands that all else is equal.
+It is not not self-evident that else else is indeed equal, that is, annual production of foliage could differ between PFTs.
 Nonetheless, further supporting this explanation, \citet{reich1992} showed a positive correlation between LL and total stand leaf biomass across diverse forest stands.
+Another, not mutually exclusive, possible explanation is that it may be beneficial for plants to target a more constant LAR rather than LMF, when considering the trade-off between leaf area and $M_T$.
+Canopy size scales with $M_T$ \citep{duursma2010}, and a larger canopy means higher light interception per unit leaf area \citep{duursma2007}.
+It can thus be argued that an optimal LAR exists that balances investment in supporting woody biomass (reflected by $M_T$) and the efficiency of foliage in terms of light interception.
 
@@
 S4).
-If this result holds, it means that aboveground biomass may be estimated across PFTs with a single equation.
+If this result holds, it means that aboveground biomass may be estimated across PFTs with an equation that does not differ between PFTs.
 For tropical forests, \citet{chave2005} showed that aboveground biomass can be well estimated as,
@@
 
-where F is a form factor, and $\rho$ wood density (kg m$^3$).
-In this equation leaf mass is subsumed in F, or assumed to be negligible (which hold for somewhat tall trees, see Fig.
-2).
-Our result seems to indicate something like this works for all PFTs (see Fig.
+where $F$ is a form factor, and $\rho$ wood density (kg m$^3$).
+Rearranging, this equation predicts that $M_T/A_S$ is proportional to plant height.
+We found some support for this prediction (Fig.
 S4).
-However, there are differences in wood density between angiosperms and gymnosperms (Zanne + TRY categories), does this mean F is different between PFTs?
-It is timely to datamine BAAD to develop better general equations to estimate aboveground biomass.
+It is timely to use the BAAD to develop better, more general, equations to estimate individual plant aboveground biomass across the globe.
 
-Climate (temperature, rainfall) did not appreciably affect biomass partitioning between leaves and stems.
+We found that climate did not appreciably affect biomass partitioning between leaves and stems.
 This finding is in contrast with a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}.
-One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands.
+One explanation for this difference is that our measurements were taken on individual plants, whereas in  \citet{reich2014} they were on whole stands.
 This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants.
 With the caveats that there is still variation between species and studies unaccounted for (Fig.
-S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and S1).
+S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and Table S1).
 
-We found that PFTs differed in LMF, and that LMF was proportional to LMA.
-It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.
+Overall, our results indicate that it is possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.
 Indeed, nearly as much variation was explained across species when using LMA instead of PFT in a statistical model.

### 2015-07-23 manuscript_suppinfo.Rnw
Another revision. This one is very close, all paragraphs cleaned.

@@
 \usepackage{gensymb}
-\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
+\geometry{verbose,tmargin=2.4cm,bmargin=2.4cm,lmargin=2.4cm,rmargin=2.4cm}
 \graphicspath{ {./figures/} }
@@
 Each point is an individual plant.
-Lines are spline fits.
-The right panel shows least-square means of the ratio of above-ground biomass to basal stem area, estimated from a linear mixed-effects model with basal stem area (linear and squared terms) and PFT.
+Lines are GAM fits by PFT.
+(a) Scaling of aboveground biomass ($M_T$) with basal stem area ($A_S$).
+(b) The ratio of $M_T/A_S$ as a function of plant height shows no clear differences between PFTs.
+The solid black line is a line with a slope of 1, which is expected if the aboveground biomass equation from \citet{chave2005} hold across the dataset.
+(c) Least-square means of the ratio of above-ground biomass to basal stem area, estimated from a linear mixed-effects model with basal stem area (linear and squared terms) and PFT.
 }

### 2015-07-23 R/figures.R

@@ -442,6 +442,8 @@ figureS4 <- function(dataset, KGAM=4){
              ylab=expression(M[T]/A[S]~~(kg~m^-2)), axes=FALSE,
              linecols=my_linecols(), pointcols=my_cols_transparent(), R="Group",kgam=KGAM,
              cex=0.6, xlim=c(-2,2.5), ylim=c(0,4.5))
+  abline(2.5,1)
+
   log10axes()
   box()
   plotlabel("(b)","topright")

### 2015-07-23 manuscript.Rnw
Automatic word count on first page

@@
 
+Total number of words: \Sexpr{nrw$Total} \\
+Abstract: \Sexpr{nrw$Abstract} \\
+Introduction: \Sexpr{nrw$Introduction} \\
+Methods and Materials: \Sexpr{nrw$Methods} \\
+Results: \Sexpr{nrw$Results} \\
+Discussion: \Sexpr{nrw$Discussion} \\
+
+Number of figures: 4 (all in colour) \\
+Number of tables: 1 \\
+Supporting information : 5 Tables, 5 Figures and 1 Notes.
+
 %--------------------------------------------------------------------------------------------%
@@
 
+\clearpage
 \section*{Abstract}
@@
 Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig.
-2a and S2).
+2a and Fig.
+S2).
 When corrected for plant height by estimating LMF at a common plant height, we found that LMF was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig.
@@
 As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig.
-2c and S3).
+2c and Fig.
+S3).
 Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig.

### 2015-07-23 R/build.R

@@ -7,5 +7,7 @@ tex_2_pdf <- function(texfile){
   system(sprintf("pdflatex %s", texfile))
   system(sprintf("pdflatex %s", texfile))
   aux.files <- paste0(filename, c(".log", ".aux", ".bbl", ".blg"))
+  shell("texcount manuscript.tex > wordcount.txt")
   file.remove(aux.files[file.exists(aux.files)])
 }
+

### 2015-07-23 R/manuscript_functions.R

@@ -1,0 +1,32 @@
+
+get_wordcount <- function(fn){
+  
+  # gets first numeric string
+  getcount <- function(x)as.numeric(str_extract(x, "[0-9]{1,4}"))
+  
+  # not robust; will find first matched argument
+  count <- function(txt)getcount(r[grep(txt,r)[1]])
+  
+  if(file.exists(fn)){
+    
+    r <- readLines(fn)
+    
+    # should do this more cleverly, but annoyingly the subsections are not added to a total
+    # for the section. So more work to make this general
+    
+    l <- list()
+    l$Abstract <- count("Abstract")
+    l$Introduction <- count("Introduction")
+    l$Methods <- count("Terminology") + count("Data")  + 
+                 count("Data analysis")
+    l$Results <- count("Results")
+    l$Discussion <- count("Discussion")
+    
+    l$Total <- count("Words in text")
+    
+    return(l)
+    
+  } else {
+    return(list(NA))
+  }
+}

### 2015-07-23 R/build.R
Removed wordcount from remake, does not work since it is a recursive dependency.

@@ -7,7 +7,6 @@ tex_2_pdf <- function(texfile){
   system(sprintf("pdflatex %s", texfile))
   system(sprintf("pdflatex %s", texfile))
   aux.files <- paste0(filename, c(".log", ".aux", ".bbl", ".blg"))
-  shell("texcount manuscript.tex > wordcount.txt")
   file.remove(aux.files[file.exists(aux.files)])
 }
 

### 2015-07-24 manuscript.Rnw
Removed wordcount from remake, does not work since it is a recursive dependency.
Minor formatting; intro on new page; bit of info in terminology
added a reference

--- 2015-07-23 manuscript.Rnw
@@
   %texcount manuscript.tex > wordcount.txt
+  %shell("texcount manuscript.tex > wordcount.txt")
 
@@
 
-%--------------------------------------------------------------------------------------------%
-
 <<preload, echo=FALSE>>=
@@
 
+%% uncomment this after compiling the ms, then compile again. This is a recursive dependency :(
+% Also, requires Perl.
+% #' <<>>=
+% #' shell("texcount manuscript.tex > wordcount.txt")
+% #' nrw <- get_wordcount("wordcount.txt")
+% #' @
+% #' 
+% #' Total number of words: \Sexpr{nrw$Total} \\
+% #' Abstract: \Sexpr{nrw$Abstract} \\
+% #' Introduction: \Sexpr{nrw$Introduction} \\
+% #' Methods and Materials: \Sexpr{nrw$Methods} \\
+% #' Results: \Sexpr{nrw$Results} \\
+% #' Discussion: \Sexpr{nrw$Discussion} \\
+% #' 
+% #' Number of figures: 4 (all in colour) \\
+% #' Number of tables: 1 \\
+% #' Supporting information : 5 Tables, 5 Figures and 1 Notes.
+
+%--------------------------------------------------------------------------------------------%
+
+\clearpage
 \section*{Abstract}
@@
 
+\clearpage
 \section*{Introduction}
@@
 Several classic, oft-cited, studies have assumed, based only on differences in measurements of leaf mass, that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
+
 Little is known about global-scale patterns in LMF and LAR in relation to climate.
-It can be expected that biomass partitioning is correlated with precipitation or mean annual temperature because experiments (cite) and comparisons across sites with contrasting climate \citep{callaway1994,poyatos2007} have shown responsiveness of biomass partitioning to environmental drivers.
+It can be expected that biomass partitioning is correlated with precipitation or mean annual temperature because small-scale comparisons across sites with contrasting climate have shown responsiveness of biomass partitioning to environmental drivers \citep{callaway1994,delucia2000,poyatos2007}.
 A recent study \citep{reich2014} demonstrated that stand-scale biomass partitioning was related to mean annual temperature across diverse forest stands.
@@
 It is thus easy to see that partitioning ($M_i/M_T$, where $M_T$ is total biomass) can be different from allocation ($a_i$) because turnover ($\tau_i$) differs between leaf and wood biomass.
+Here, we present data on biomass partitioning, which can inform models of allocation only when estimates of turnover are available.
 
@@
 
+\clearpage
 \section*{Supporting Information}

### 2015-07-24 manuscript.Rnw
More references; minor editing

@@
 
-In this paper we are primarily interested in the distribution of biomass ('partitioning') between leaves and stems, an important component of the residence time of carbon in ecosystems \citep{friend2014}.
+In this paper we are primarily interested in the distribution of biomass ('partitioning') between leaves and woody stems, an important component of the residence time of carbon in ecosystems \citep{friend2014}.
 Previous work based on whole stands \citep{oneilldeangelis1981,enquistniklas2002,reich2014} and individual plants \citep{poorter2012}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
@@
 It is also unknown whether PFTs with higher leaf biomass also have higher total leaf area, which is relevant because leaf area drives total light interception and thus productivity.
-Several classic, oft-cited, studies have assumed, based only on differences in measurements of leaf mass, that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
+Several classic, oft-cited, studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}, but this was based only on differences in measurements of leaf mass.
 
 Little is known about global-scale patterns in LMF and LAR in relation to climate.
-It can be expected that biomass partitioning is correlated with precipitation or mean annual temperature because small-scale comparisons across sites with contrasting climate have shown responsiveness of biomass partitioning to environmental drivers \citep{callaway1994,delucia2000,poyatos2007}.
+It can be expected that biomass partitioning is correlated with precipitation or mean annual temperature because smaller-scale comparisons have shown responsiveness of biomass partitioning to environmental drivers \citep{berninger1994,callaway1994,delucia2000,poyatos2007,}.
 A recent study \citep{reich2014} demonstrated that stand-scale biomass partitioning was related to mean annual temperature across diverse forest stands.
@@
 We used the recently compiled Biomass And Allometry Database (BAAD) \citep{Falster2015}, including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$).
-For each analysis, we used different subsets of the data.
+The database contains many more variables, for example root biomass for a much smaller subset of studies.
+Here we limit the analysis to patterns in aboveground biomass distribution.
+For each analysis, we used different subsets of the data because not all variables were measured in each study.
 Sample sizes by PFT are summarized in Tables S2-S5.
@@
 
-A previous large compilation of plant biomass partitioning \citep{poorter2012} (which included many non field-grown plants) did not find a correlation between LMF and LMA across 8 PFTs (including non-woody PFTs).
-However, they did not correct for plant size in their comparison.
+A previous compilation of plant biomass partitioning \citep{poorter2012} (which included many non field-grown plants) did not find a correlation between LMF and LMA across 8 PFTs (including non-woody PFTs).
+One possible reason for this discrepancy, apart from differences in sample size and growing conditions, is that they did not correct for plant size in their comparison between LMF and LMA.
 We found that, as expected, plant size strongly influences LMF (Fig.
 2) and LAR (Fig.
-S3).
-It is certainly necessary to correct for plant size when comparing biomass partitioning parameters across PFTs\footnote{some better ending to this para is needed - not just want to bash poorter's crappy analysis, is there a positive note we can end on?
-we suggest a method to do so?}.
+S3) (see also \citet{poorter2012}).
+It is thus necessary to correct for plant size when comparing biomass partitioning parameters, as has been noted many times \citep{mcconnaughay1999}.
+We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship.
+This was useful because both LMF and LAR showed very non-linear patterns with plant height, even on a logarithmic scale.
 
@@
 where $F$ is a form factor, and $\rho$ wood density (kg m$^3$).
-Rearranging, this equation predicts that $M_T/A_S$ is proportional to plant height.
+After rearranging, we find that this equation predicts that $M_T/A_S$ is proportional to plant height.
 We found some support for this prediction (Fig.
 S4).
-It is timely to use the BAAD to develop better, more general, equations to estimate individual plant aboveground biomass across the globe.
+It is timely to use the database made public by \citet{falster2015} (BAAD) to develop more general equations to estimate individual plant aboveground biomass across the globe.
 
@@
 Indeed, nearly as much variation was explained across species when using LMA instead of PFT in a statistical model.
-This is surprising because LMA only captures aspects of leaf morphology, in indirectly leaf lifespan and photosynthetic capacity via the leaf economics spectrum \citep{wrightetal2004}.
+This is surprising because LMA only captures aspects of leaf morphology, and indirectly leaf lifespan and photosynthetic capacity via the leaf economics spectrum \citep{wrightetal2004}.
 We thus show that in a PFT-based classification, LMA is a good first estimate of biomass partitioning, however, more variation can be explained by another trait ($A_F/A_S$), which varies appreciably between species.
@@
 
-Figure 2a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary.
 Our sincere thanks to everyone who contributed data to the Biomass and Allometry Database.
 We thank Martin de Kauwe for discussions and comments on an earlier version of this paper.
+Figure 2a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary.
 

### 2015-07-24 manuscript_suppinfo.Rnw
More references; minor editing

@@
 
-Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types.
+Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types (PFTs), although some algorithms make no mention of these potential differences \citep{friedlingstein1999, chen2013}.
 However, there is no consensus on how these differences are parameterized, or whether they are linked to inherent differences in leaf mass per area (LMA) or leaf lifespan (LL).

### 2015-08-06 manuscript.Rnw
Added notes on Poorter et al 2015 - we will have to integrate their results
Minor editts

@@
+% Notes on Poorter et al. 2015.
+% - we have to demonstrate that our work is still novel
+% - their focus is firstly on the size-related shifts in biomass partitioning, we don't study that in detail, and can now justify that by referring back to Poorter.
+% - they don't have SLA data for the same plants, and in fact predict what we found: 
+% "It is also known that, on average, the specific
+% leaf area (SLA; leaf area per unit leaf mass) of evergreen woody
+% species is 2–3 times higher than that of deciduous species
+% (Poorter et al., 2009). Hence, all else being equal, the 2–3 times
+% lower SLA in evergreens would have to be compensated by a 2–3
+% time higher LMF to arrive at the more or less similar LAI."
+% - so basically we can say in the intro that we wish to test this hypothesis (as we already say), and also that Poorter speculated that this may be so (although here they contrast deciduous vs.
+evergreen, and not angiosperm vs.
+gymnosperm, so our results are still novel in that case as well)
+% - we have to clearly state that our data has *no overlap* whatsoever with Poorter, this is a good thing.
+% - their data includes potted plants in glass houses and field-grown, as well as stand biomass data divided by stocking.
+Our data are more homogenous, with only field-grown plants, individual estimates.
+% - we have larger trees included; in fact the largest in the world, we need to state that in the intro to again show the novelty (our max mass is ca. 300t, theirs is 1t)
+% - they don't actually show LMF vs. plant size split by PFT (their Fig 2a), so our result is also not scooped there.
+% - they found no difference in the LMF residual between evergreen angiosperm and gymnosperm (Fig. 4), or at least I don't think so.
+
 \documentclass[a4paper]{article}
@@
 We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass.
+Similarly we find that PFTs differ in the amount of leaf mass per unit stem cross section, but not in the amount of leaf area per unit stem cross section.
 Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum.
@@
 
-The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation \citep{ise2010}.
-It affects the residence time and storage of carbon in ecosystems, and the response of net primary production to rising CO\textsubscript{2}\citep{dekauwe2014,friend2014,ise2010}.
+The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation \citep{ise2010, dekauwe2014,friend2014,ise2010}.
 Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \citep{ryanlaw2005}.
@@
 The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010}, assumed to differ in key physiological attributes.
-While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are constrained by limited empirical data.
+While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are only constrained, due to limited available data.
 Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
@@
 
+% This para feels out of place. Perhaps better incorporated as part of material for last para in intro? Or towards the end in the first paragraph of the discussion?
 Despite considerable advances in theory underlying allometric scaling in plants \citep{enquist2003, west1999, savage2010} we do not yet have a clear understanding of potential differences between PFTs in terms of biomass partitioning.
@@
 Rather than to advance a new model of biomass partitioning or allocation, we take the view that a broader evidence base is needed to elucidate patterns between PFTs.
-It is then up to the individual GVMs to refine the algorithms and parameter values to more closely match the observations.
+This should enable those building GVMs to refine algorithms and parameter values to more closely match the observations.
 
@@
 It is also unknown whether PFTs with higher leaf biomass also have higher total leaf area, which is relevant because leaf area drives total light interception and thus productivity.
-Several classic, oft-cited, studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}, but this was based only on differences in measurements of leaf mass.
+Several classic, oft-cited, studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
 
 Little is known about global-scale patterns in LMF and LAR in relation to climate.
-It can be expected that biomass partitioning is correlated with precipitation or mean annual temperature because smaller-scale comparisons have shown responsiveness of biomass partitioning to environmental drivers \citep{berninger1994,callaway1994,delucia2000,poyatos2007,}.
-A recent study \citep{reich2014} demonstrated that stand-scale biomass partitioning was related to mean annual temperature across diverse forest stands.
+It can be expected that biomass partitioning is correlated with precipitation or mean annual temperature because smaller-scale comparisons have shown responsiveness of biomass partitioning to environmental drivers \citep{berninger1994,callaway1994,delucia2000,poyatos2007}.
+A recent study by \citet{reich2014} demonstrated that stand-scale biomass partitioning was related to mean annual temperature across diverse forest stands.
 Indeed, many GVMs assume that climate affects biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Notes S1).
@@
 
-Here, we use a unique, new database \citep{Falster2015} (Fig.
+Here, we use a unique, new database \citep{falster2015} (Fig.
 1) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms.
 A second objective was to study relationships between biomass partitioning and climate variables or biome at a global scale.
-The dataset includes observations of biomass and size metrics for individual plants, contributed by 175 studies across nine vegetation types (Fig.
+The dataset includes observations of biomass and size metrics for individual plants, compiled from 175 studies across nine vegetation types (Fig.
 S1), across the three major biomes (boreal, temperate and tropical).
-In this paper we include field-grown woody plants only, including the entire size range of woody plants (0.01 - \textgreater 100m height).
+In this paper we focus on field-grown woody plants, spanning the entire size range of woody plants (0.01 - \textgreater 100 m height).
 One key challenge is thus to account for the very large size variation commonly found in any allometric variable \citep{niklas1994}.
@@
 
-We used the recently compiled Biomass And Allometry Database (BAAD) \citep{Falster2015}, including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$).
+We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total include records for \Sexpr{nrow(baad_all$data)} individuals.
+In this paper we restrict our analysis to records that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$) (n=\Sexpr{nrow(dataset)}).
 The database contains many more variables, for example root biomass for a much smaller subset of studies.
@@
 
-We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves, see \citet{Falster2015} for methods details for each contributed study).
-For conifers, leaf area was converted to half-total surface area using the average of a set of published conversion factors \citep{barclay2000}.
-We used a different conversion for pines (\emph{Pinus} spp.) than non-pines, because the conversion factors were very different for these two groups \citep{barclay2000}.
+We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves, see \citet{falster2015} for methods details for each contributed study).
+For conifers, leaf area was converted to half-total surface area using the average of a set of published conversion factors \citep{barclay2000}, with  different conversion factors applied to pines (\emph{Pinus} spp.) and non-pines.
 This conversion was necessary because half-total surface area is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}.
@@
 
-$$a = c_0H^{c_1}$$
+$$a = c_0H^{c_1.}$$
 
@@
 
-We used generalized additive models (GAM) to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height.
+We used generalised additive models (GAM) to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height.
 In all fitted GAMs, we used a cubic regression spline.
@@
 Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons across climatic gradients, due to the small sample size within each species or site.
-To study climate effects on biomass partitioning, we further decomposed LMF as the product,
+To study climate effects on biomass partitioning, we therefore further decomposed LMF as the product,
 
@@
 
-because we found that the vast majority of size-related variation was captured by $A_S/M_T$ alone.
+This decomposition showed that the vast majority of size-related variation was captured by $A_S/M_T$ alone.
 Similarly, we decomposed LAR as the product of $A_F/A_S$ and $A_S/M_T$.
@@
 
-The ratio $M_F/A_S$ is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}.
+The ratio $M_F/A_S$ is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}, using records of stem diameter commonly recorded on long-term monitoring plots.
 The correlation between $M_F/A_S$ and LMA was found to hold also within PFTs (Fig.
@@
 In Fig.
-3b, The separation among PFTs in $M_F/A_S$ arises due to differences in LMA.
+3b, the separation among PFTs in $M_F/A_S$ arises due to differences in LMA.
 
 We found that decomposing LMF and LAR as shown in Eq.
-\ref{eq:M1} was very useful to study climate effects because the second term ($A_S/M_T$)absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB}, Table S1 and Fig.
+\ref{eq:M1} was very useful to study climate effects because the second term ($A_S/M_T$) absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB}, Table S1 and Fig.
 S4) and is otherwise fairly conserved across PFTs (Fig.
@@
 This conclusion was particularly strong across the three PFTs studied, but it also held across species within PFT.
-The consequence is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) did not differ between PFTs.
-This result is particularly convincing because the data included individual plants from the entire size range of woody plants in natural forests (0.1-100m) (Fig.
+The implication is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) do not differ between PFTs.
+This result seems robust, as our data include individual plants spanning the entire size range of woody plants in natural forests (0.1-100m) (Fig.
 2a).
@@
 
-The finding that LMF is proportional to LMA across PFTs implies that LAR does not differ between PFTs.
-One explanation for this pattern lies in the strong positive correlation between LMA and leaf lifespan (LL) \citep{wright2004}.
+% Is it really necessary to invoke differences in LL?
+
+The finding that LMF is proportional to LMA across PFTs implies that LAR does not differ systematically between PFTs.
+One explanation for this pattern lies in the strong positive correlation between LMA and leaf lifespan (LL) \citep{wrightetal2004}.
 Both \cite{poorter2012} and \citet{enquistniklas2002} explained higher LMF in gymnosperms by higher LL; plants simply maintain more cohorts of foliage, but this explanation demands that all else is equal.
-It is not not self-evident that else else is indeed equal, that is, annual production of foliage could differ between PFTs.
+Yet it is not self-evident that everything else is indeed equal, that is, annual production of foliage could differ between PFTs.
 Nonetheless, further supporting this explanation, \citet{reich1992} showed a positive correlation between LL and total stand leaf biomass across diverse forest stands.
-Another, not mutually exclusive, possible explanation is that it may be beneficial for plants to target a more constant LAR rather than LMF, when considering the trade-off between leaf area and $M_T$.
-Canopy size scales with $M_T$ \citep{duursma2010}, and a larger canopy means higher light interception per unit leaf area \citep{duursma2007}.
+Another, not mutually exclusive, possible explanation is that plants may allocate biomass in a manner that targets a more constant LAR rather than LMF.
+Canopy size scales with $M_T$ \citep{duursma2010}, and a larger canopy means higher light interception per unit total biomass \citep{duursma2007}.
 It can thus be argued that an optimal LAR exists that balances investment in supporting woody biomass (reflected by $M_T$) and the efficiency of foliage in terms of light interception.
 
+% The first sentence here feels a bit repetitive 
 We found that, as for LMF, $M_F/A_S$ was also proportional to LMA across PFTs (Fig.
@@
 Indeed, nearly as much variation was explained across species when using LMA instead of PFT in a statistical model.
-This is surprising because LMA only captures aspects of leaf morphology, and indirectly leaf lifespan and photosynthetic capacity via the leaf economics spectrum \citep{wrightetal2004}.
+This is surprising because LMA only captures one aspect of functional differentiation among PFT -- leaf morphology (but including correlated effects on leaf lifespan and photosynthetic capacity via the leaf economics spectrum \citep{wrightetal2004}).
 We thus show that in a PFT-based classification, LMA is a good first estimate of biomass partitioning, however, more variation can be explained by another trait ($A_F/A_S$), which varies appreciably between species.
-These results suggest promising avenues for parameterizing and simplifying biomass partitioning routines in GVMs.
+These results suggest promising avenues for parameterising and simplifying biomass partitioning routines in GVMs.
 
@@
 Some models already incorporate a similar algorithm (see Notes S1), while other algorithms may be tuned to yield similar patterns between PFTs as we have presented.
-In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{Falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz_towards_2012, dekauwe2014} of GVMs against empirical data.
+In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz_towards_2012, dekauwe2014} of GVMs against empirical data.
 
@@
     \includegraphics[width=0.99\textwidth]{Figure1.pdf}
-    \caption{Overview of the global biomass and allometry database.
+    \caption{Overview of the data.
 (a) Variables were measured on \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species.
@@
 Each symbol is an individual plant.
-Lines are generalized additive model fits.
+Lines are generalised additive model fits.
 (b) and (c) Leaf mass fraction and leaf area ratio ($A_F/M_T$) at the average plant height in the dataset, estimated from panel (a).
@@
 b-c) Relationships between climate and $M_F/A_S$ within each PFT.
-Solid lines are generalized additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
+Solid lines are generalised additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
     \label{fig:figure3}

### 2015-08-11 manuscript.Rnw
Moved sample size tables to one combined table in main manuscript.

@@
 <<results='asis'>>=
+bigtable <- rbind(cbind(c("LMF","","",""),c("DA","EA","EG","Sum"), table_lmf),
+                  cbind(c("LAR","","",""),c("DA","EA","EG","Sum"), table_lar), 
+                  cbind(c("MF/AS","","",""),c("DA","EA","EG","Sum"), table_mlfastba),
+                  cbind(c("AF/AS","","",""),c("DA","EA","EG","Sum"), table_alfastba))
+rownames(bigtable) <- NULL
+colnames(bigtable)[1:2] <- c("Variable","PFT")
+
+print(xtable(bigtable, align="llrlll|l",
+               caption="Sample sizes used in the analyses of the four variables considered in the study (LMF, LAR, MF/AS and AF/AS), by plant functional type (PFT) and biome.
+Sample sizes denote number of individuals, with number of unique species in parentheses.
+",
+             label="table:counttable"
+               ), hline.after=c(0,4,8,12,16), caption.placement="top",
+      include.rownames=FALSE)
+@
+
+<<results='asis'>>=
 print(xtable(table_varpart2, label="table/varpartB",

### 2015-08-11 manuscript_suppinfo.Rnw
Moved sample size tables to one combined table in main manuscript.

@@
       include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
-@
-
-<<tablepreload1>>=
-printit <- function(x, caption, label=NULL){
-  print(xtable(x, align="rlll|l",
-               caption=caption, label=label
-               ), hline.after=c(0,3), caption.placement="top")
-
-}
-@
-
-<<results='asis'>>=
-printit(table_mlfastba,
-        label="table:count1",
-        caption="Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$.
-Numbers denote individual plants, with number of species in parentheses.
-DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
-@
-
-<<results='asis'>>=
-printit(table_alfastba,
-        label="table:count2",
-        caption="Number of observations used in the analysis of $A_F/A_S$.
-Numbers denote individual plants, with number of species in parentheses.
-DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
-@
-
-<<results='asis'>>=
-printit(table_lmf,
-        label="table:count3",
-        caption="Number of observations used in the analysis of leaf mass fraction ($M_F/M_T$) and leaf area ratio ($A_F/M_T$).
-Numbers denote individual plants, with number of species in parentheses.
-DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
-@
-
-<<results='asis'>>=
-printit(table_lar,
-        label="table:count4",
-        caption="Number of observations used in the analysis of leaf area ratio (LAR).
-Numbers denote individual plants, with number of species in parentheses.
-DA = Deciduous Angiosperm, EA = Evergreen Angiosperm, EG = Evergreen Gymnosperm.")
 @

### 2015-08-11 manuscript.Rnw
Moved FigS2 to main manuscript; fixed caption table1

--- 2015-08-06 manuscript.Rnw
@@
 <<results='asis'>>=
+bigtable <- rbind(cbind(c("$M_F/M_T$","","",""),c("DA","EA","EG","Sum"), table_lmf),
+                  cbind(c("$A_F/M_T$","","",""),c("DA","EA","EG","Sum"), table_lar), 
+                  cbind(c("$M_F/A_S$","","",""),c("DA","EA","EG","Sum"), table_mlfastba),
+                  cbind(c("$A_F/A_S$","","",""),c("DA","EA","EG","Sum"), table_alfastba))
+rownames(bigtable) <- NULL
+colnames(bigtable)[1:2] <- c("Variable","PFT")
+
+print(xtable(bigtable, align="llrlll|l",
+               caption="Sample sizes used in the analyses of the four variables considered in the study by plant functional type (PFT) and biome.
+Sample sizes denote number of individuals, with number of unique species in parentheses.
+",
+             label="table:counttable"
+               ), hline.after=c(0,4,8,12,16), caption.placement="top",
+      include.rownames=FALSE)
+@
+
+<<results='asis'>>=
 print(xtable(table_varpart2, label="table/varpartB",
@@
     \centering
+    \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
+    \caption{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$).
+Each point is an individual plant.
+Sample sizes are listed in Table~\ref{table:counttable}}
+    \label{fig:figureS2}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
     \includegraphics[width=0.99\textwidth]{Figure2.pdf}
-    \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf construction cost.
+    \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf mass per area.
 (a) Leaf mass fraction ($M_F/M_T$ = leaf mass / above-ground biomass) by PFT.

### 2015-08-13 manuscript.Rnw
Figure refs fixed

@@
 
-Here, we use a unique, new database \citep{falster2015} (Fig.
-1) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms.
+Here, we use a unique, new database \citep{falster2015} (Fig.~\ref{fig:figoverview}) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms.
 A second objective was to study relationships between biomass partitioning and climate variables or biome at a global scale.
@@
 
-Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig.
-2a and Fig.
+The raw data in Fig.~\ref{fig:figrawdata} show a steeper increase of woody aboveground biomass ($M_S$) with plant height than for foliage biomass ($M_F$).
+As a result, there was a strong effect of plant height on LMF , with the three plant functional types clearly differing in LMF across nearly the entire size range (Fig.~\ref{fig:figlmf}a).
+When corrected for plant height by estimating LMF at a common plant height, we found that LMF was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig.~\ref{fig:figlmf}b).
+As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig.
+~\ref{fig:figlmf}c and Fig.
 S2).
-When corrected for plant height by estimating LMF at a common plant height, we found that LMF was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig.
-2b).
-As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig.
-2c and Fig.
-S3).
+
 Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig.
-2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig.
-2e).
+~\ref{fig:figlmf}d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig.~\ref{fig:figlmf}e).
 
@@
 We found that LMF differs predictably among PFTs, however there was considerable variation in all studied variables between species within PFTs (Fig.
-S5).
+S4).
 It is thus pertinent to understand whether we can improve on a PFT-based classification by including other traits that affect biomass partitioning.
@@
 The ratio $M_F/A_S$ is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}, using records of stem diameter commonly recorded on long-term monitoring plots.
-The correlation between $M_F/A_S$ and LMA was found to hold also within PFTs (Fig.
-3a), explaining 30\% of the variation in $M_F/A_S$ across species.
+The correlation between $M_F/A_S$ and LMA was found to hold also within PFTs (Fig.~\ref{fig:figlmfspecies}a), explaining 30\% of the variation in $M_F/A_S$ across species.
 The regressions across species by PFT were broad and overlapping, with a more general relationship extending across the entire LMA axis.
-A larger fraction of the variation in $M_F/A_S$ was explained by species-level differences in $A_F/A_S$ (Fig.
-3b).
-As previously noted, $A_F/A_S$ does not differ systematically among PFTs (Fig.
-2e), but does vary close to two orders of magnitude across species.
-In Fig.
-3b, the separation among PFTs in $M_F/A_S$ arises due to differences in LMA.
+A larger fraction of the variation in $M_F/A_S$ was explained by species-level differences in $A_F/A_S$ (Fig.~\ref{fig:figlmfspecies}b).
+As previously noted, $A_F/A_S$ does not differ systematically among PFTs (Fig.~\ref{fig:figlmf}e), but does vary close to two orders of magnitude across species.
+In Fig.~\ref{fig:figlmfspecies}b, the separation among PFTs in $M_F/A_S$ arises due to differences in LMA.
 
@@
 \ref{eq:M1} was very useful to study climate effects because the second term ($A_S/M_T$) absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB}, Table S1 and Fig.
-S4) and is otherwise fairly conserved across PFTs (Fig.
-S4).
-As shown in Fig.
-2b and Fig.
-2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and Table S1).
+S3) and is otherwise fairly conserved across PFTs (Fig.
+S3).
+As shown in Fig.~\ref{fig:figlmfspecies}b and Fig.~\ref{fig:figlmf}d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and Table S1).
 $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.
 
-Regardless of how we analyzed the data, we found only weak climatic effects on biomass partitioning within PFTs (Fig.
-4).
+Regardless of how we analyzed the data, we found only weak climatic effects on biomass partitioning within PFTs (Fig.~\ref{fig:figclimate}).
 Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table \ref{table/varpartB}).
-Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) (Fig.
-4b-c).
+Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) (Fig.~\ref{fig:figclimate}b-c).
 These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table S1).
@@
 The implication is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) do not differ between PFTs.
-This result seems robust, as our data include individual plants spanning the entire size range of woody plants in natural forests (0.1-100m) (Fig.
-2a).
+This result seems robust, as our data include individual plants spanning the entire size range of woody plants in natural forests (0.1-100m) (Fig.~\ref{fig:figlmf}a).
 Previous studies have demonstrated a large difference in LMF between angiosperms and gymnosperms \citep{poorter2012}, but have not further separated functional types into deciduous vs.
@@
 One possible reason for this discrepancy, apart from differences in sample size and growing conditions, is that they did not correct for plant size in their comparison between LMF and LMA.
-We found that, as expected, plant size strongly influences LMF (Fig.
-2) and LAR (Fig.
+We found that, as expected, plant size strongly influences LMF (Fig.~\ref{fig:figlmf}) and LAR (Fig.
 S3) (see also \citet{poorter2012}).
@@
 % The first sentence here feels a bit repetitive 
-We found that, as for LMF, $M_F/A_S$ was also proportional to LMA across PFTs (Fig.
-2).
+We found that, as for LMF, $M_F/A_S$ was also proportional to LMA across PFTs (Fig.~\ref{fig:figlmf}).
 Because LMF may be decomposed into the terms $M_F/A_S$ and $M_T/A_S$ (\ref{eq:M2}), this result suggests that $M_T/A_S$ does not vary between PFTs (see also Fig.
@@
 Colour symbols show the locations of sampled individuals for three dominant woody functional types.}
-    \label{fig:figure1}
+    \label{fig:figoverview}
 \end{figure}
@@
 Sample sizes are listed in Table~\ref{table:counttable}}
-    \label{fig:figureS2}
+    \label{fig:figrawdata}
 \end{figure}
@@
 Letters denote significant differences (at $\alpha = 0.05$).}
-    \label{fig:figure2}
+    \label{fig:figlmf}
 \end{figure}
@@
 $R^2$ for fitted relationships are \Sexpr{round(100*r2panela,1)}\% in panel (a) and \Sexpr{round(100*r2panelb,1)}\% in panel (b) (for a linear regression model including PFT as a factor variable).}
-  \label{fig:figure4}
+  \label{fig:figlmfspecies}
 \end{figure}
@@
 Solid lines are generalised additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
-    \label{fig:figure3}
+    \label{fig:figclimate}
 \end{figure}
@@
 \textbf{Table S1} Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.\\
-\textbf{Table S2} Number of observations used in the analysis of $M_F/A_S$ and $A_F/A_S$.\\
-\textbf{Table S3} Number of observations used in the analysis of $A_F/A_S$.\\
-\textbf{Table S4} Number of observations used in the analysis of leaf mass fraction ($M_F/M_T$) and leaf area ratio ($A_F/M_T$).\\
 \textbf{Table S5} Number of observations used in the analysis of leaf area ratio (LAR).\\
 \textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\
-\textbf{Fig. S2} Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$).\\
-\textbf{Fig. S3} Leaf area ratio ($A_F/M_T$) by PFT.\\
-\textbf{Fig. S4} Relationship between above-ground biomass and basal stem area.\\
-\textbf{Fig. S5} Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.\\
+\textbf{Fig. S2} Leaf area ratio ($A_F/M_T$) by PFT.\\
+\textbf{Fig. S3} Relationship between above-ground biomass and basal stem area.\\
+\textbf{Fig. S4} Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.\\
 \textbf{Notes S1} Modelling of biomass partitioning in global vegetation models (GVMs)

### 2015-08-14 manuscript.Rnw
Moved histogram figure to main text

@@
 % #' 
-% #' Number of figures: 4 (all in colour) \\
-% #' Number of tables: 1 \\
-% #' Supporting information : 5 Tables, 5 Figures and 1 Notes.
+% #' Number of figures: 6 (all in colour) \\
+% #' Number of tables: 2 \\
+% #' Supporting information : 1 Tables, 3 Figures and 1 Notes.
 
@@
 
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
+    \caption{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.
+Shown are histograms (as probability density functions) of leaf mass per area ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three PFTs.
+Arrows indicate means by PFT.}
+    \label{fig:fighist}
+\end{figure}
+
 <<echo=FALSE>>=

### 2015-08-14 manuscript_suppinfo.Rnw
Moved FigS2 to main manuscript; fixed caption table1
Moved histogram figure to main text

@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
-    \caption{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$). Each point is an individual plant.}
-    \label{fig:figureS2}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
     \includegraphics[width=0.99\textwidth]{FigureS3.pdf}
-    \caption{Leaf area ratio ($A_F/M_T$) by PFT.
+    \caption{Leaf area ratio ($A_F/M_T$) by plant height (H) and PFT.
 Each symbol is an individual plant.
@@
     \label{fig:figureS4}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
-    \caption{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.
-Shown are histograms (as probability density functions) of leaf mass per area (an index of leaf construction cost) ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types.
-Arrows indicate means by PFT.}
-    \label{fig:figures5}
 \end{figure}

### 2015-08-14 manuscript.Rnw
fixed table1

@@
 print(xtable(bigtable, align="llrlll|l",
-               caption="Sample sizes used in the analyses of the four variables considered in the study by plant functional type (PFT) and biome.
+               caption="Sample sizes used in the analyses of the four variables considered in the study by plant functional type (PFT, DA = deciduous angiosperm, EA = evergreen angiosperm, EG = evergreen gymnosperm) and biome.
 Sample sizes denote number of individuals, with number of unique species in parentheses.
 ",
-             label="table:counttable"
-               ), hline.after=c(0,4,8,12,16), caption.placement="top",
+             label="table:counttable"), 
+      hline.after=c(0,4,8,12,16), caption.placement="top",
+      sanitize.text.function=function(x){x},
       include.rownames=FALSE)

### 2015-08-14 manuscript.Rnw
Fixed figure references; some minor text edits

@@
 
-We found that LMF differs predictably among PFTs, however there was considerable variation in all studied variables between species within PFTs (Fig.
-S4).
+We found that LMF differs predictably among PFTs, however there was considerable variation in all studied variables between species within PFTs (Fig.\ref{fig:fighist}).
 It is thus pertinent to understand whether we can improve on a PFT-based classification by including other traits that affect biomass partitioning.
@@
 \ref{eq:M1} was very useful to study climate effects because the second term ($A_S/M_T$) absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB}, Table S1 and Fig.
-S3) and is otherwise fairly conserved across PFTs (Fig.
-S3).
+S3) and is otherwise fairly conserved across PFTs (Fig.~\ref{fig:fighist}).
 As shown in Fig.~\ref{fig:figlmfspecies}b and Fig.~\ref{fig:figlmf}d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and Table S1).
@@
 This conclusion was particularly strong across the three PFTs studied, but it also held across species within PFT.
-The implication is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) do not differ between PFTs.
+The implication is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) does not differ between PFTs.
 This result seems robust, as our data include individual plants spanning the entire size range of woody plants in natural forests (0.1-100m) (Fig.~\ref{fig:figlmf}a).
-Previous studies have demonstrated a large difference in LMF between angiosperms and gymnosperms \citep{poorter2012}, but have not further separated functional types into deciduous vs.
+Previous studies have demonstrated a large difference in LMF between angiosperms and gymnosperms \citep{poorter2012,poorter2015}, but have not further separated functional types into deciduous vs.
 evergreen angiosperms, or provided an explanation in terms of leaf traits.
@@
 We found that, as expected, plant size strongly influences LMF (Fig.~\ref{fig:figlmf}) and LAR (Fig.
-S3) (see also \citet{poorter2012}).
+S2) (see also \citet{poorter2012}).
 It is thus necessary to correct for plant size when comparing biomass partitioning parameters, as has been noted many times \citep{mcconnaughay1999}.
@@
 
-% The first sentence here feels a bit repetitive 
-We found that, as for LMF, $M_F/A_S$ was also proportional to LMA across PFTs (Fig.~\ref{fig:figlmf}).
+We found that patterns in $M_F/A_S$ across PFTs mirrored those of LMF (Fig.~\ref{fig:figlmf}).
 Because LMF may be decomposed into the terms $M_F/A_S$ and $M_T/A_S$ (\ref{eq:M2}), this result suggests that $M_T/A_S$ does not vary between PFTs (see also Fig.
-S4).
-If this result holds, it means that aboveground biomass may be estimated across PFTs with an equation that does not differ between PFTs.
+S3).
+If this result holds, it means that aboveground biomass may be estimated across PFTs with a single equation that does not differ between PFTs.
 For tropical forests, \citet{chave2005} showed that aboveground biomass can be well estimated as,
@@
 We found some support for this prediction (Fig.
-S4).
-It is timely to use the database made public by \citet{falster2015} (BAAD) to develop more general equations to estimate individual plant aboveground biomass across the globe.
+S3).
+It is timely to use database made public by \citet{falster2015} (BAAD) and \citet{poorter2015} to develop more general equations to estimate individual plant aboveground biomass across the globe.
 
@@
 This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants.
-With the caveats that there is still variation between species and studies unaccounted for (Fig.
-S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and Table S1).
+With the caveats that there is still variation between species and studies unaccounted for (Fig.~\ref{fig:fighist}), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and Table S1).
 
-Overall, our results indicate that it is possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.
+Our results indicate that it is possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.
 Indeed, nearly as much variation was explained across species when using LMA instead of PFT in a statistical model.
@@
 \textbf{Fig. S3} Relationship between above-ground biomass and basal stem area.\\
-\textbf{Fig. S4} Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.\\
 \textbf{Notes S1} Modelling of biomass partitioning in global vegetation models (GVMs)

### 2015-08-19 manuscript.Rnw
added a reference

@@
 While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots).
-The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010}, assumed to differ in key physiological attributes.
+The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010,wullschleger2014}, assumed to differ in key physiological attributes.
 While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are only constrained, due to limited available data.
@@
 Some models already incorporate a similar algorithm (see Notes S1), while other algorithms may be tuned to yield similar patterns between PFTs as we have presented.
-In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz_towards_2012, dekauwe2014} of GVMs against empirical data.
+In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data.
 

### 2015-08-20 manuscript.Rnw
Combined both explained variance tables into main text

@@
 gymnosperm, so our results are still novel in that case as well)
+
 % - we have to clearly state that our data has *no overlap* whatsoever with Poorter, this is a good thing.
@@
 The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010,wullschleger2014}, assumed to differ in key physiological attributes.
-While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are only constrained, due to limited available data.
+While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are poorly constrained, due to limited available data.
 Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
 These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared \citep{wolf2011}.
-
-% This para feels out of place. Perhaps better incorporated as part of material for last para in intro? Or towards the end in the first paragraph of the discussion?
-Despite considerable advances in theory underlying allometric scaling in plants \citep{enquist2003, west1999, savage2010} we do not yet have a clear understanding of potential differences between PFTs in terms of biomass partitioning.
-Instead, most previous work has focussed on understanding size-related shifts in biomass partitioning, as governed by constraints including hydraulic supply and mechanical stability.
-Rather than to advance a new model of biomass partitioning or allocation, we take the view that a broader evidence base is needed to elucidate patterns between PFTs.
-This should enable those building GVMs to refine algorithms and parameter values to more closely match the observations.
 
@@
 
+Despite considerable advances in theory underlying allometric scaling in plants \citep{enquist2003, west1999, savage2010} we do not yet have a clear understanding of potential differences between PFTs in terms of biomass partitioning.
+Instead, most previous work has focussed on understanding size-related shifts in biomass partitioning, as governed by constraints including hydraulic supply and mechanical stability \citep{savage2010}.
+Rather than to advance a new model of biomass partitioning or allocation, we take the view that a broader evidence base is needed to elucidate patterns between PFTs.
+This should enable those building GVMs to refine algorithms and parameter values to more closely match the observations.
 Here, we use a unique, new database \citep{falster2015} (Fig.~\ref{fig:figoverview}) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms.
+A recent compilation of plant biomass data \citep{poorter2015} speculated that differences between gymnosperms and angiosperms in distribution of biomass between leaves and stems could be related to differences in leaf mass per area.
+Here we can directly test this hypothesis as our database includes many observations of leaf area as well as leaf mass measured on the same plants.
+
 A second objective was to study relationships between biomass partitioning and climate variables or biome at a global scale.
@@
 We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total include records for \Sexpr{nrow(baad_all$data)} individuals.
+The database does not overlap with the recent compilation of \citet{poorter2015}.
 In this paper we restrict our analysis to records that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$) (n=\Sexpr{nrow(dataset)}).
@@
 For each analysis, we used different subsets of the data because not all variables were measured in each study.
-Sample sizes by PFT are summarized in Tables S2-S5.
+Sample sizes by PFT are summarized in Table 1.
 We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
@@
 
-We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves, see \citet{falster2015} for methods details for each contributed study).
+We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves, see \citet{falster2015} for details on the methods for each contributed study).
 For conifers, leaf area was converted to half-total surface area using the average of a set of published conversion factors \citep{barclay2000}, with  different conversion factors applied to pines (\emph{Pinus} spp.) and non-pines.
@@
 
+<<results='asis'>>=
+tab1 <- table_varpart2
+tab2 <- table_gamr2MATMAP
+
+tidytab <- function(x, labeltopleft=""){
+  names(x)[1] <- "Variable"
+  x[,1] <- paste0(x[,1],"(",x[,2],")")
+  x[,2] <- NULL
+  x[,2:ncol(x)] <- round(x[,2:ncol(x)], 2)  
+  x <- unname(rbind(names(x),as.matrix(x)))  
+  x <- rbind(c(labeltopleft,"","Predictors",""),x)
+return(x)
+}
+
+tab1 <- tidytab(tab1, "Mixed-effects model")
+tab2 <- tidytab(tab2, "GAM")
+
+tab <- rbind(tab1, tab2)
+
+print(xtable(tab, align="lrrrr",
+             caption="Explained variance by plant functional type (PFT), plant height (H) and biome (B), for five whole-plant variables.
+Each variable was added to a linear mixed-effects model, using species within dataset as a random effect.
+The $R^2$ shown is that explained by the fixed effects only.
+All fixed effects were highly significant (P \\textless 0.001).
+In addition to the fixed effects shown, all interactions were added to each of the models."),
+             hline.after=c(0,1,7,8,14),
+             include.rownames=FALSE, 
+             include.colnames=FALSE, 
+             caption.placement="top", sanitize.text.function=function(x){x})
+@
+
 \clearpage

### 2015-08-20 manuscript_suppinfo.Rnw
Combined both explained variance tables into main text

@@
 \setcounter{subsection}{0}
-
-\renewcommand\thetable{S\arabic{table}}
-\setcounter{table}{0}
-
-<<preload, echo=FALSE>>=
-library(knitr)
-library(xtable)
-opts_chunk$set(echo=FALSE)
-@
-
-<<results='asis'>>=
-print(xtable(table_gamr2MATMAP, label="table/varpartMAPMAT",
-      caption="Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.
-Each variable was added as a smooth term in a generalized additive model, with the exception of PFT (a categorical variable)."),
-      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
-@
-
-\clearpage
 

### 2015-08-20 manuscript.Rnw
removed notes on poorter, I think I have accounted for it now

@@
-% Notes on Poorter et al. 2015.
-% - we have to demonstrate that our work is still novel
-% - their focus is firstly on the size-related shifts in biomass partitioning, we don't study that in detail, and can now justify that by referring back to Poorter.
-% - they don't have SLA data for the same plants, and in fact predict what we found: 
-% "It is also known that, on average, the specific
-% leaf area (SLA; leaf area per unit leaf mass) of evergreen woody
-% species is 2–3 times higher than that of deciduous species
-% (Poorter et al., 2009). Hence, all else being equal, the 2–3 times
-% lower SLA in evergreens would have to be compensated by a 2–3
-% time higher LMF to arrive at the more or less similar LAI."
-% - so basically we can say in the intro that we wish to test this hypothesis (as we already say), and also that Poorter speculated that this may be so (although here they contrast deciduous vs.
-evergreen, and not angiosperm vs.
-gymnosperm, so our results are still novel in that case as well)
-
-% - we have to clearly state that our data has *no overlap* whatsoever with Poorter, this is a good thing.
-% - their data includes potted plants in glass houses and field-grown, as well as stand biomass data divided by stocking.
-Our data are more homogenous, with only field-grown plants, individual estimates.
-% - we have larger trees included; in fact the largest in the world, we need to state that in the intro to again show the novelty (our max mass is ca. 300t, theirs is 1t)
-% - they don't actually show LMF vs. plant size split by PFT (their Fig 2a), so our result is also not scooped there.
-% - they found no difference in the LMF residual between evergreen angiosperm and gymnosperm (Fig. 4), or at least I don't think so.
-
 \documentclass[a4paper]{article}
@@
 
-A previous compilation of plant biomass partitioning \citep{poorter2012} (which included many non field-grown plants) did not find a correlation between LMF and LMA across 8 PFTs (including non-woody PFTs).
-One possible reason for this discrepancy, apart from differences in sample size and growing conditions, is that they did not correct for plant size in their comparison between LMF and LMA.
 We found that, as expected, plant size strongly influences LMF (Fig.~\ref{fig:figlmf}) and LAR (Fig.
-S2) (see also \citet{poorter2012}).
+S2) (see also \citet{poorter2012,poorter2015}).
 It is thus necessary to correct for plant size when comparing biomass partitioning parameters, as has been noted many times \citep{mcconnaughay1999}.
 We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship.
-This was useful because both LMF and LAR showed very non-linear patterns with plant height, even on a logarithmic scale.
-
-% Is it really necessary to invoke differences in LL?
+This was useful because both LMF and LAR showed very non-linear patterns with plant height, even on a logarithmic scale, consistent with \citet{poorter2015}.
 
@@
 <<results='asis'>>=
-print(xtable(table_varpart2, label="table/varpartB",
-             caption="Explained variance by plant functional type (PFT), plant height (H) and biome (B), for five whole-plant variables.
-Each variable was added to a linear mixed-effects model, using species within dataset as a random effect.
-The $R^2$ shown is that explained by the fixed effects only.
-All fixed effects were highly significant (P \\textless 0.001).
-In addition to the fixed effects shown, all interactions were added to each of the models."),
-      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
-@
-
-<<results='asis'>>=
 tab1 <- table_varpart2
@@
 print(xtable(tab, align="lrrrr",
-             caption="Explained variance by plant functional type (PFT), plant height (H) and biome (B), for five whole-plant variables.
-Each variable was added to a linear mixed-effects model, using species within dataset as a random effect.
+             caption="Explained variance by plant functional type (PFT), plant height (H) and climate variables (either biome (B), or MAP and MAT) for five whole-plant variables.
+For the analysis with biome, each variable was added to a linear mixed-effects model, using species within dataset as a random effect.
 The $R^2$ shown is that explained by the fixed effects only.
 All fixed effects were highly significant (P \\textless 0.001).
-In addition to the fixed effects shown, all interactions were added to each of the models."),
+In addition to the fixed effects shown, all interactions were added to each of the models.
+For the analysis with continuous climate variables MAP and MAT, each variable was added as a smooth term in a generalized additive model (GAM), with the exception of PFT (a categorical variable).
+Note that explained variance for the linear mixed-effects model and GAM is not necessarily the same even with the same predictors, due to different methods for fitting and estimating the explained variance."),
              hline.after=c(0,1,7,8,14),
@@
     \caption{Overview of the data.
-(a) Variables were measured on \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species.
+(a) Variables were measured on up to \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species.
 (b) Coverage of the dataset across global climate space.

### 2015-08-20 manuscript.Rnw
Fixed table references

@@
 \usepackage{gensymb}
-\usepackage{longtable}
 \usepackage{lineno}
@@
 We found that decomposing LMF and LAR as shown in Eq.
-\ref{eq:M1} was very useful to study climate effects because the second term ($A_S/M_T$) absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB}, Table S1 and Fig.
+\ref{eq:M1} was very useful to study climate effects because the second term ($A_S/M_T$) absorbs nearly all of the size related variation in LMF (Table 2 and Fig.
 S3) and is otherwise fairly conserved across PFTs (Fig.~\ref{fig:fighist}).
-As shown in Fig.~\ref{fig:figlmfspecies}b and Fig.~\ref{fig:figlmf}d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and Table S1).
+As shown in Fig.~\ref{fig:figlmfspecies}b and Fig.~\ref{fig:figlmf}d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Table 2).
 $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.
@@
 Regardless of how we analyzed the data, we found only weak climatic effects on biomass partitioning within PFTs (Fig.~\ref{fig:figclimate}).
-Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table \ref{table/varpartB}).
+Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table 2).
 Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) (Fig.~\ref{fig:figclimate}b-c).
-These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table S1).
+These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table 2).
 
@@
 This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants.
-With the caveats that there is still variation between species and studies unaccounted for (Fig.~\ref{fig:fighist}), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and Table S1).
+With the caveats that there is still variation between species and studies unaccounted for (Fig.~\ref{fig:fighist}), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Table 2).
 
@@
 Each point is an individual plant.
-Sample sizes are listed in Table~\ref{table:counttable}}
+Sample sizes are listed in Table 1}
     \label{fig:figrawdata}
@@
 
-\textbf{Table S1} Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.\\
-\textbf{Table S5} Number of observations used in the analysis of leaf area ratio (LAR).\\
 \textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\

### 2015-08-24 manuscript.Rnw
Rewrote abstract cf. NewPhyt requirements; more editing for clarity

@@
 
-The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs).
+% 200 word max.
+(1) The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs).
 Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical studies have not been able to justify these differences.
-Using a new global database on biomass partitioning from over 175 studies, we here show that partitioning between stems and leaves varies strongly between PFTs (as defined by deciduous vs.
+Here we test for differences between PFTs in partitioning of biomass between leaves and stems.
+(2) We use the recently published Biomass And Allometry Database (BAAD), a large database including observations for individual plants.
+The database covers the global climate space, allowing us to test for direct climate effects in addition to PFT.
+(3) The leaf mass fraction (LMF, leaf / total aboveground biomass) varied strongly between PFTs (as defined by deciduous vs.
 evergreen and gymnosperm vs.
 angiosperm).
-We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass.
-Similarly we find that PFTs differ in the amount of leaf mass per unit stem cross section, but not in the amount of leaf area per unit stem cross section.
-Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum.
-In addition, we show how a second trait, leaf area per stem basal area, explains additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs.
-Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
+We found that LMF, once corrected for plant height, was proportional to leaf mass per area across PFTs.
+As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass.
+We found only weak and inconsistent effects of climate on biomass partitioning.
+(4) Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in GVMs.
 
@@
 While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are poorly constrained, due to limited available data.
-Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
+Moreover, there is little consensus on how biomass partitioning and allocation (see Methods for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
 These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared \citep{wolf2011}.
@@
 In this paper we are primarily interested in the distribution of biomass ('partitioning') between leaves and woody stems, an important component of the residence time of carbon in ecosystems \citep{friend2014}.
-Previous work based on whole stands \citep{oneilldeangelis1981,enquistniklas2002,reich2014} and individual plants \citep{poorter2012}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
+Previous work based on whole stands \citep{oneilldeangelis1981,enquistniklas2002,reich2014} and individual plants \citep{poorter2012,poorter2015}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
 \citet{poorter2012} and \citet{enquistniklas2002} attributed higher LMF in gymnosperms to longer leaf lifespan (LL) compared to typical angiosperms, but this begs the question whether differences in LMF are also apparent between deciduous and evergreen functional types within angiosperms.
@@
 Here, we use a unique, new database \citep{falster2015} (Fig.~\ref{fig:figoverview}) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms.
-A recent compilation of plant biomass data \citep{poorter2015} speculated that differences between gymnosperms and angiosperms in distribution of biomass between leaves and stems could be related to differences in leaf mass per area.
-Here we can directly test this hypothesis as our database includes many observations of leaf area as well as leaf mass measured on the same plants.
+A recent compilation of plant biomass data \citep{poorter2015} speculated that differences between gymnosperms and angiosperms in distribution of biomass between leaves and stems is related to differences in leaf mass per area.
+Here we can directly test this hypothesis as our database, unlike that of \citet{poorter2015}, includes many observations of leaf area as well as leaf mass measured on the same plants.
 
@@
 The two processes are different because of continuous turnover of biomass, which differs strongly between compartments.
-We may write \citep{McMurtrieWolf1983},
+We may write \citep{mcmurtriewolf1983},
 
@@
 We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total include records for \Sexpr{nrow(baad_all$data)} individuals.
-The database does not overlap with the recent compilation of \citet{poorter2015}.
+The database does not overlap with the recent large compilation of \citet{poorter2015}.
 In this paper we restrict our analysis to records that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$) (n=\Sexpr{nrow(dataset)}).
@@
 
-The raw data in Fig.~\ref{fig:figrawdata} show a steeper increase of woody aboveground biomass ($M_S$) with plant height than for foliage biomass ($M_F$).
-As a result, there was a strong effect of plant height on LMF , with the three plant functional types clearly differing in LMF across nearly the entire size range (Fig.~\ref{fig:figlmf}a).
+The raw data in Fig.~\ref{fig:figrawdata} show a steeper increase of woody aboveground biomass ($M_S$) with plant height compared to foliage biomass ($M_F$).
+As a result, LMF decreased with plant height, with the three plant functional types clearly differing in LMF across nearly the entire size range (Fig.~\ref{fig:figlmf}a).
 When corrected for plant height by estimating LMF at a common plant height, we found that LMF was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig.~\ref{fig:figlmf}b).
@@
 
-We found that LMF differs predictably among PFTs, however there was considerable variation in all studied variables between species within PFTs (Fig.\ref{fig:fighist}).
-It is thus pertinent to understand whether we can improve on a PFT-based classification by including other traits that affect biomass partitioning.
-To do this, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq.
+There was considerable variation in all studied variables between species within PFTs (Fig.\ref{fig:fighist}).
+To understand whether we can improve on a PFT-based classification by including other traits that affect biomass partitioning, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq.
 \ref{eq:M2}).
+The ratio $M_F/A_S$ is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}, using records of stem diameter commonly recorded on long-term monitoring plots.
 
@@
 
-The ratio $M_F/A_S$ is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}, using records of stem diameter commonly recorded on long-term monitoring plots.
 The correlation between $M_F/A_S$ and LMA was found to hold also within PFTs (Fig.~\ref{fig:figlmfspecies}a), explaining 30\% of the variation in $M_F/A_S$ across species.
@@
 This result seems robust, as our data include individual plants spanning the entire size range of woody plants in natural forests (0.1-100m) (Fig.~\ref{fig:figlmf}a).
-Previous studies have demonstrated a large difference in LMF between angiosperms and gymnosperms \citep{poorter2012,poorter2015}, but have not further separated functional types into deciduous vs.
-evergreen angiosperms, or provided an explanation in terms of leaf traits.
-We show that LMA -- a central trait of the leaf economics spectrum (LES) \citep{wrightetal2004} -- explains these differences in a consistent manner.
+Previous studies have demonstrated a large difference in LMF between angiosperms and gymnosperms \citep{poorter2012,poorter2015}, but have not been able to explain these differences in terms of leaf traits.
+We show that LMA -- a central trait of the leaf economics spectrum (LES) \citep{wrightetal2004} -- explains differences in LMF in a consistent manner.
 

### 2015-08-24 manuscript.Rnw
Abstract numbering; remake fix

@@
 \clearpage
-\section*{Abstract}
+\section*{Summary}
 
 % 200 word max.
-(1) The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs).
+\begin{enumerate}
+\item The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs).
 Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical studies have not been able to justify these differences.
 Here we test for differences between PFTs in partitioning of biomass between leaves and stems.
-(2) We use the recently published Biomass And Allometry Database (BAAD), a large database including observations for individual plants.
+\item We use the recently published Biomass And Allometry Database (BAAD), a large database including observations for individual plants.
 The database covers the global climate space, allowing us to test for direct climate effects in addition to PFT.
-(3) The leaf mass fraction (LMF, leaf / total aboveground biomass) varied strongly between PFTs (as defined by deciduous vs.
+\item The leaf mass fraction (LMF, leaf / total aboveground biomass) varied strongly between PFTs (as defined by deciduous vs.
 evergreen and gymnosperm vs.
@@
 We found only weak and inconsistent effects of climate on biomass partitioning.
-(4) Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in GVMs.
+\item Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in GVMs.
+\end{enumerate}
 

### 2015-08-24 manuscript.Rnw
Minor edits throughout

@@
 
-The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation \citep{ise2010, dekauwe2014,friend2014,ise2010}.
+The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation \citep{ise2010, dekauwe2014,friend2014}.
 Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \citep{ryanlaw2005}.
@@
 In this paper we are primarily interested in the distribution of biomass ('partitioning') between leaves and woody stems, an important component of the residence time of carbon in ecosystems \citep{friend2014}.
-Previous work based on whole stands \citep{oneilldeangelis1981,enquistniklas2002,reich2014} and individual plants \citep{poorter2012,poorter2015}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
+Previous work based on either whole stands \citep{oneilldeangelis1981,enquistniklas2002,reich2014} and or a mix on stand-- and individual-based measurements \citep{poorter2012, poorter2015}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
 \citet{poorter2012} and \citet{enquistniklas2002} attributed higher LMF in gymnosperms to longer leaf lifespan (LL) compared to typical angiosperms, but this begs the question whether differences in LMF are also apparent between deciduous and evergreen functional types within angiosperms.
 It is also unknown whether PFTs with higher leaf biomass also have higher total leaf area, which is relevant because leaf area drives total light interception and thus productivity.
-Several classic, oft-cited, studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
+Some oft-cited studies have also assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
 
@@
 It can be expected that biomass partitioning is correlated with precipitation or mean annual temperature because smaller-scale comparisons have shown responsiveness of biomass partitioning to environmental drivers \citep{berninger1994,callaway1994,delucia2000,poyatos2007}.
-A recent study by \citet{reich2014} demonstrated that stand-scale biomass partitioning was related to mean annual temperature across diverse forest stands.
+Moreover, a recent study by \citet{reich2014} demonstrated that stand-scale biomass partitioning was related to mean annual temperature across diverse forest stands.
 Indeed, many GVMs assume that climate affects biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Notes S1).
@@
 Instead, most previous work has focussed on understanding size-related shifts in biomass partitioning, as governed by constraints including hydraulic supply and mechanical stability \citep{savage2010}.
-Rather than to advance a new model of biomass partitioning or allocation, we take the view that a broader evidence base is needed to elucidate patterns between PFTs.
+Rather than to advance a specific model of biomass partitioning or allocation, our view is that a broader evidence base is needed first to elucidate patterns between PFTs.
 This should enable those building GVMs to refine algorithms and parameter values to more closely match the observations.
@@
 A recent compilation of plant biomass data \citep{poorter2015} speculated that differences between gymnosperms and angiosperms in distribution of biomass between leaves and stems is related to differences in leaf mass per area.
-Here we can directly test this hypothesis as our database, unlike that of \citet{poorter2015}, includes many observations of leaf area as well as leaf mass measured on the same plants.
+Here we can directly test this hypothesis as our database, unlike those of \citet{enquistniklas2002}, \citet{reich2014}, and \cite{poorter2015}, includes many observations of leaf area as well as leaf mass measured on the same plants.
 
@@
 One key challenge is thus to account for the very large size variation commonly found in any allometric variable \citep{niklas1994}.
-We do this by fitting a statistical model that does no assume a functional dependence on plant size, allowing us to study PFT patterns at a common plant height.
+We do this by fitting a semi-parametric data-driven statistical model, i.e.
+one that does not assume a particular functional form, thereby allowing us to study PFT patterns at a common plant height.
 
@@
 
-where $M_i$ is the biomass in some compartment (leaves, stems or roots), $a_i$ the annual allocation of NPP to that compartment, and $\tau_i$ the annual turnover of compartment $i$.
+where $M_i$ is the biomass in some compartment (leaves, stems or roots) remaining on the plant, $a_i$ the annual allocation of NPP to that compartment, and $\tau_i$ the annual turnover (or loss) of compartment $i$ from the plant.
 It is thus easy to see that partitioning ($M_i/M_T$, where $M_T$ is total biomass) can be different from allocation ($a_i$) because turnover ($\tau_i$) differs between leaf and wood biomass.
@@
 
-We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total include records for \Sexpr{nrow(baad_all$data)} individuals.
-The database does not overlap with the recent large compilation of \citet{poorter2015}.
+We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total includes records for \Sexpr{nrow(baad_all$data)} individuals.
+The database has limited overlap with the recent large compilation of \citet{poorter2015} and differs in that a majority of measurements for large plants are individual rather than stand-based.
 In this paper we restrict our analysis to records that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$) (n=\Sexpr{nrow(dataset)}).
@@
 For each analysis, we used different subsets of the data because not all variables were measured in each study.
-Sample sizes by PFT are summarized in Table 1.
+Sample sizes by PFT are summarised in Table \ref{table:counttable}.
 We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
@@
 
-$$a = c_0H^{c_1.}$$
+$$a = c_0H^{c_1}.$$
 
@@
 
-We used generalised additive models (GAM) to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height.
+We used generalised additive models (GAM) to capture  the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height.
 In all fitted GAMs, we used a cubic regression spline.
 For the smoothed term in the model (plant height), we used up to 3 or 4 degrees of freedom, which resulted in biologically realistic smoothed relationships.
-Within the GAM, we used a penalized regression smoother \citep{wood2006} to allow the final degree of smoothness to be estimated from the data.
+Within the GAM, we used a penalised regression smoother \citep{wood2006} to allow the final degree of smoothness to be estimated from the data.
 In all fitted GAMs, we used species-dataset combination as a random effect.
-We used the \emph{mgcv} package in R \citep{wood2006}.
 All variables (except MAP and MAT) were log-transformed prior to analysis.
@@
 
-Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons across climatic gradients, due to the small sample size within each species or site.
+Despite the exceptional size of our dataset, the strong size-dependence of LMF still hinders comparisons across climatic gradients, due to the small sample size available within each species or site when sampling at a common height.
 To study climate effects on biomass partitioning, we therefore further decomposed LMF as the product,
@@
 
-This decomposition showed that the vast majority of size-related variation was captured by $A_S/M_T$ alone.
-Similarly, we decomposed LAR as the product of $A_F/A_S$ and $A_S/M_T$.
-We analyzed climate effects on $M_F/A_S$ and $A_F/A_S$ in two ways: with quantitative variables (mean annual precipitation, MAP; mean annual temperature, MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig.
+This decomposition showed that the vast majority of size-related variation was captured by $A_S/M_T$ alone, indicating that comparisons among $M_F/A_S$ could be made without needing to compare at a common height.
+A similar decomposition  of LAR as the product of $A_F/A_S$ and $A_S/M_T$ produced the same outcome.
+We therefore analysed for climatic effects on $M_F/A_S$ and $A_F/A_S$ in two ways: with quantitative variables (mean annual precipitation, MAP; mean annual temperature, MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig.
 S1).
-In both cases we used PFT and plant height as covariates.
+In both cases we used PFT and plant height as covariates varying across climate space.
+
+All analyses were conducted in R v3.2.0 \citep{R-2015}.
+GAMs were fitted using \emph{mgcv} package \citep{wood2006}.
+The code to replicate this analysis are available on GitHub at http://github.com/RemkoDuursma/baadanalysis.
 
@@
 As a result, LMF decreased with plant height, with the three plant functional types clearly differing in LMF across nearly the entire size range (Fig.~\ref{fig:figlmf}a).
-When corrected for plant height by estimating LMF at a common plant height, we found that LMF was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig.~\ref{fig:figlmf}b).
+When we corrected for plant height by estimating LMF at a common plant height, we found that LMF was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig.~\ref{fig:figlmf}b).
 As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig.
@@
 
-Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig.
-~\ref{fig:figlmf}d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig.~\ref{fig:figlmf}e).
+Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differed among the three PFTs, that these differences were also proportional to LMA (Fig.
+~\ref{fig:figlmf}d), and that PFTs were similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig.~\ref{fig:figlmf}e).
 
@@
 We then replaced PFT with LMA, and found that the model with LMA could explain almost as much variation in LMF as the model with PFT ($R^2$ = 0.74 with PFT vs.
-0.62 with LMA) and $M_F/A_S$ ($R^2$ = 0.29 vs.
+0.62 with LMA), likewise for $M_F/A_S$ ($R^2$ = 0.29 vs.
 0.28).
 
-There was considerable variation in all studied variables between species within PFTs (Fig.\ref{fig:fighist}).
-To understand whether we can improve on a PFT-based classification by including other traits that affect biomass partitioning, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq.
-\ref{eq:M2}).
-The ratio $M_F/A_S$ is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}, using records of stem diameter commonly recorded on long-term monitoring plots.
+There was considerable variation in all studied variables between species within PFTs (Fig.
+~\ref{fig:fighist}).
+To understand whether we can improve on a PFT-based classification by including other traits that affect biomass partitioning, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$):
 
@@
 
+The ratio $M_F/A_S$ is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}, using records of stem diameter commonly recorded on long-term monitoring plots.
 The correlation between $M_F/A_S$ and LMA was found to hold also within PFTs (Fig.~\ref{fig:figlmfspecies}a), explaining 30\% of the variation in $M_F/A_S$ across species.
@@
 We found that decomposing LMF and LAR as shown in Eq.
-\ref{eq:M1} was very useful to study climate effects because the second term ($A_S/M_T$) absorbs nearly all of the size related variation in LMF (Table 2 and Fig.
-S3) and is otherwise fairly conserved across PFTs (Fig.~\ref{fig:fighist}).
-As shown in Fig.~\ref{fig:figlmfspecies}b and Fig.~\ref{fig:figlmf}d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Table 2).
+\ref{eq:M1} was very useful to study climate effects because the second term ($A_S/M_T$) absorbed nearly all of the size-related variation in LMF (Table \ref{table:statstable} and Fig.
+S3) and is otherwise fairly conserved across PFTs (Fig.
+S3).
+As shown in Fig.~\ref{fig:figlmfspecies}b and Fig.~\ref{fig:figlmf}d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Table \ref{table:statstable}).
 $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.
 
-Regardless of how we analyzed the data, we found only weak climatic effects on biomass partitioning within PFTs (Fig.~\ref{fig:figclimate}).
-Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table 2).
+Regardless of how we analysed the data, we found only weak climatic effects on biomass partitioning within PFTs (Fig.~\ref{fig:figclimate}).
+Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table \ref{table:statstable}).
 Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) (Fig.~\ref{fig:figclimate}b-c).
-These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table 2).
+These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table \ref{table:statstable}).
 
@@
 We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship.
-This was useful because both LMF and LAR showed very non-linear patterns with plant height, even on a logarithmic scale, consistent with \citet{poorter2015}.
+This was useful because both LMF and LAR showed very non-linear patterns with plant height, even on a logarithmic scale, consistent with recent other results \citep{poorter2015}.
 
@@
 We found that patterns in $M_F/A_S$ across PFTs mirrored those of LMF (Fig.~\ref{fig:figlmf}).
-Because LMF may be decomposed into the terms $M_F/A_S$ and $M_T/A_S$ (\ref{eq:M2}), this result suggests that $M_T/A_S$ does not vary between PFTs (see also Fig.
+Because LMF may be decomposed into the terms $M_F/A_S$ and $M_T/A_S$ (eq.
+\ref{eq:M2}), this result suggests that $M_T/A_S$ does not vary between PFTs (see also Fig.
 S3).
 If this result holds, it means that aboveground biomass may be estimated across PFTs with a single equation that does not differ between PFTs.
-For tropical forests, \citet{chave2005} showed that aboveground biomass can be well estimated as,
+For tropical forests, \citet{chave2005} showed that aboveground biomass can be estimated as,
 
@@
 where $F$ is a form factor, and $\rho$ wood density (kg m$^3$).
-After rearranging, we find that this equation predicts that $M_T/A_S$ is proportional to plant height.
+After rearranging, this equation predicts that $M_T/A_S$ is proportional to plant height.
 We found some support for this prediction (Fig.
-S3).
+S3), suggesting further improvements could be made by incorporating this dependence.
 It is timely to use database made public by \citet{falster2015} (BAAD) and \citet{poorter2015} to develop more general equations to estimate individual plant aboveground biomass across the globe.
@@
 This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants.
-With the caveats that there is still variation between species and studies unaccounted for (Fig.~\ref{fig:fighist}), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Table 2).
+With the caveats that there is still variation between species and studies unaccounted for (Fig.~\ref{fig:fighist}), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Table \ref{table:statstable}).
 
@@
 Our sincere thanks to everyone who contributed data to the Biomass and Allometry Database.
-We thank Martin de Kauwe for discussions and comments on an earlier version of this paper.
+We thank Martin De Kauwe and anonymous referees for comments on an earlier version of this paper.
+Thanks also to Rich FitzJohn for advice about using the remake package.
 Figure 2a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary.
@@
 Each point is an individual plant.
-Sample sizes are listed in Table 1}
+Sample sizes are listed in Table \ref{table:counttable}}
     \label{fig:figrawdata}
@@
 Shown are histograms (as probability density functions) of leaf mass per area ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three PFTs.
-Arrows indicate means by PFT.}
+Arrows indicate means by PFT.
+Colours as in Fig.
+\ref{fig:figlmf}.}
     \label{fig:fighist}

### 2015-08-25 R/figures.R
Renumbered all figures in sequence; also thanks grid for breaking backwards compatibility in ellipseGrob; grid added to required packages

@@ -77,11 +77,11 @@ fig.tree <- function(filename) {
   grid.text("H, height of plant", x = 0.05 , y = 0.85, just="left", gp=gp0)
 
   # stems areas
-  grid.draw(ellipseGrob(x0 - 0.025, 0.03, size=6,ar=4,angle=0, def="npc"))
+  grid.draw(ellipseGrob(x0 - 0.023, 0.024, size=1.8,rho=1/3,angle=0, def="npc"))
   grid.text(expression(paste(A[S],", stem area at base")),
     x = x0 + 0.03 , y = 0.03, just="left", gp=gp0)
 
-  grid.draw(ellipseGrob(x0 - 0.03, 0.08, size=4,ar=4,angle=0, def="npc"))
+  grid.draw(ellipseGrob(x0 - 0.03, 0.08, size=1.1,rho=1/3,angle=0, def="npc"))
   grid.text(expression(paste(A[Sbh],", stem area at breast height")),
     x = x0 +0.023 , y = 0.08, just="left", gp=gp0)
 
@@ -101,7 +101,7 @@ fig.tree <- function(filename) {
 }
 
 
-# Figure 1. panel a.
+# Figure 1. panel b.
 # MAP MAT vs. Worldclim
 figureMAPMATworldclim <- function(baad_mapmat, world_mapmat, groupvar="pft",
                                   col=my_cols_transparent(0.7), setpar=TRUE,
@@ -157,9 +157,40 @@ figureMAPMATworldclim <- function(baad_mapmat, world_mapmat, groupvar="pft",
 }
 
 
+# Leaf mass, woody mass raw data
+figure2 <- function(dataset){
+  
+  par(mfrow=c(1,3), mar=c(0,0,0,0), oma=c(5,5,2,2), las=1)
+  labels <- c("Deciduous Angiosperm", "Evergreen Angiosperm", "Evergreen Gymnosperm")
+  
+  for(i in 1:3){
+    p <- levels(dataset$pft)[i]
+    dat <- dataset[dataset$pft == p,]
+    
+    with(dat, plot(log10(h.t), log10(m.st), pch=16,cex=0.5,
+                   xlim=log10(c(0.01,105)), ylim=c(-6,6),
+                   col=alpha("brown",0.5),
+                   axes=FALSE))
+    with(dat, points(log10(h.t), log10(m.lf), pch=16,cex=0.5,
[…]

### 2015-08-25 manuscript.Rnw
Renumbered all figures in sequence; also thanks grid for breaking backwards compatibility in ellipseGrob; grid added to required packages
Table label was missing

@@
 For the analysis with continuous climate variables MAP and MAT, each variable was added as a smooth term in a generalized additive model (GAM), with the exception of PFT (a categorical variable).
-Note that explained variance for the linear mixed-effects model and GAM is not necessarily the same even with the same predictors, due to different methods for fitting and estimating the explained variance."),
+Note that explained variance for the linear mixed-effects model and GAM is not necessarily the same even with the same predictors, due to different methods for fitting and estimating the explained variance.",
+             label="table:statstable"),
              hline.after=c(0,1,7,8,14),
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
+    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
     \caption{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$).
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
+    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
     \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf mass per area.
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
+    \includegraphics[width=0.99\textwidth]{Figure4.pdf}
     \caption{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{Figure4.pdf}
+    \includegraphics[width=0.99\textwidth]{Figure5.pdf}
      \caption{Leaf and stem traits capture variation in biomass partitioning across species.
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
+    \includegraphics[width=0.99\textwidth]{Figure6.pdf}
     \caption{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning.

### 2015-08-25 manuscript.Rnw
Flag to include figures into main document
Quantified overlap with Poorter2015

@@
 \renewcommand*{\thefootnote}{\fnsymbol{footnote}}
+
+% Flag to include figures in the main text
+\newif\ifincludefigs
+\includefigsfalse  % or \includefigstrue
 
@@
 We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total includes records for \Sexpr{nrow(baad_all$data)} individuals.
-The database has limited overlap with the recent large compilation of \citet{poorter2015} and differs in that a majority of measurements for large plants are individual rather than stand-based.
+The database has very limited overlap (n = 261, 1.7 \%) with the recent large compilation of \citet{poorter2015} and differs in that measurements are all for individual plants (where \citet{poorter2015} included many stand-based estimates converted back to invididuals).
 In this paper we restrict our analysis to records that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$) (n=\Sexpr{nrow(dataset)}).
@@
     \centering
+    \ifincludefigs 
     \includegraphics[width=0.99\textwidth]{Figure1.pdf}
+    \fi
     \caption{Overview of the data.
@@
     \centering
+    \ifincludefigs 
     \includegraphics[width=0.99\textwidth]{Figure2.pdf}
+    \fi
     \caption{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$).
@@
     \centering
+    \ifincludefigs 
     \includegraphics[width=0.99\textwidth]{Figure3.pdf}
+    \fi
     \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf mass per area.
@@
     \centering
+    \ifincludefigs 
     \includegraphics[width=0.99\textwidth]{Figure4.pdf}
+    \fi
     \caption{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.
@@
     \centering
+    \ifincludefigs 
     \includegraphics[width=0.99\textwidth]{Figure5.pdf}
+    \fi
      \caption{Leaf and stem traits capture variation in biomass partitioning across species.
@@
     \centering
+    \ifincludefigs 
     \includegraphics[width=0.99\textwidth]{Figure6.pdf}
+    \fi
     \caption{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning.

### 2015-08-25 manuscript.Rnw
more horizontal lines in table (ugly but converts better)

@@
 
-\linenumbers
+% \linenumbers
 
@@
              label="table:counttable"), 
-      hline.after=c(0,4,8,12,16), caption.placement="top",
+      hline.after=0:16, caption.placement="top",  #c(0,4,8,12,16)
       sanitize.text.function=function(x){x},
@@
              label="table:statstable"),
-             hline.after=c(0,1,7,8,14),
+             hline.after=0:14, #c(0,1,7,8,14),
              include.rownames=FALSE, 

### 2015-11-04 manuscript.Rnw
removed unnecessary dependency on framed latex package
New fig 3 Mf/Ms; hierarchical partitioning notes in run.R

@@
 \usepackage{palatino}
-\usepackage{framed}
 \usepackage{parskip}
@@
 
-\title{Leaf traits drive differences in biomass partitioning among major plant functional types}
+\title{Leaf traits drive differences in above-ground biomass distribution among major plant functional types}
 
@@
 \begin{enumerate}
-\item The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs).
+\item The partitioning of biomass into leaves and stems is an important component of  global vegetation models (GVMs).
 Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical studies have not been able to justify these differences.
-Here we test for differences between PFTs in partitioning of biomass between leaves and stems.
+Here we test for differences between PFTs in distribution of biomass between leaves and stems.
 \item We use the recently published Biomass And Allometry Database (BAAD), a large database including observations for individual plants.
@@
 We found only weak and inconsistent effects of climate on biomass partitioning.
-\item Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in GVMs.
+\item Combined, these results will allow for systematic benchmarking of biomass partitioning routines in GVMs.
 \end{enumerate}
@@
 
-In this paper we are primarily interested in the distribution of biomass ('partitioning') between leaves and woody stems, an important component of the residence time of carbon in ecosystems \citep{friend2014}.
+In this paper we are primarily interested in the distribution of biomass between leaves and woody stems, an important component of the residence time of carbon in ecosystems \citep{friend2014}.
 Previous work based on either whole stands \citep{oneilldeangelis1981,enquistniklas2002,reich2014} and or a mix on stand-- and individual-based measurements \citep{poorter2012, poorter2015}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).

### 2015-11-04 R/data_processing.R

@@ -195,8 +195,10 @@ prepare_dataset_1 <- function(baad){
     lmlf_astba <- log10(m.lf/a.stba)
     lalf_astba <- log10(a.lf/a.stba)
     lh.t <- log10(h.t)
+    lmlf_mst <- log10(m.lf / m.st)
     lmlf_mso <- log10(m.lf / m.so)
     lalf_mso <- log10(a.lf / m.so)
+    lalf_mst <- log10(a.lf / m.st)
     lmrt_mso <- log10(m.rt / m.so)
     lmso <- log10(m.so)
     lmrt <- log10(m.rt)

### 2015-11-04 run.R

@@ -6,8 +6,11 @@ source("R/functions-figures.R")
 source("R/signifletters.R")
 source("R/build.R")
 source("R/manuscript_functions.R")
+
 dir.create("downloads", FALSE, TRUE)
 dir.create("figures", FALSE, TRUE)
+
+# data
 download_baad("downloads/baad.rds")
 download_tree_png("downloads/ian-symbol-eucalyptus-spp-1.png")
 world_mapmat <- prepare_worldmapmat("data/Worldclim_landcover_climspace_withcover.rds")
@@ -22,15 +25,62 @@ dat_mlfmso <- prepare_dat_mlfmso(dataset2)
 dat_mlf <- prepare_dat_mlf(dataset2)
 dat_alf <- prepare_dat_alf(dataset2)
 basalafit <- BasalA_fit(baad_all)
+
+# tables
 table_gamr2MATMAP <- make_table_gamr2MATMAP(dataset)
 table_mlfastba <- make_table_mlfastba(dat_mlf)
 table_alfastba <- make_table_alfastba(dat_alf)
 table_lmf <- make_table_lmf(dat_mlfmso)
 table_lar <- make_table_lar(dat_alfmso)
-knitr::knit("manuscript_suppinfo.Rnw", "manuscript_suppinfo.tex")
 table_varpart2 <- make_table_varpart2(dataset2)
 table_r2lmaforpft <- make_table_lmaforpft(dataset2)
+
+# ms.
+knitr::knit("manuscript_suppinfo.Rnw", "manuscript_suppinfo.tex")
 knitr::knit("manuscript.Rnw", "manuscript.tex")
+
+
+# new figures
+pdf("figures/Figure3new.pdf", width = 8L, height = 4L)
+figure3new(dataset)
+dev.off()
+
+
+
+df <- dataset2[,c("lmlf_mst", "lh.t","MAP","MAT","pft")]
+df <- df[complete.cases(df),]
+
+y <- df[,"lmlf_mst"]
+m <- df[,c("lh.t","MAP","MAT","pft")]
+
+library(hier.part)
+hier.part(y, m)
+
+
+do_hierpart <- function(dep, indep, data, ...){
+  
+  df <- data[,c(dep,indep)]
+  df <- df[complete.cases(df),]
+  
+  y <- df[,dep]
+  m <- df[,indep]
+  
+  h <- hier.part(y, m, ...) 
+  
[…]

### 2015-11-05 manuscript.Rnw
notes for revision

@@
 \clearpage
+
+\section*{Revision notes}
+\begin{enumerate}
+\item Reduce focus on GVMs in introduction (Reviewer 2).
+\item Better linking to Poorter 2015 and Reich 2014 (editor, reviewer 2, reviewer 3, poorter pers comm).
+\item Remove LMF and LAR, and only analyze MF/MS and AF/MS.
+This avoids confusion of the terms LMF and LAR which normally include root biomass.
+We should more clearly focus on above-ground biomass distribution patterns (editor + reviewer 2).
+\item Partition variance in MF/MS directly, not the proxy variable MF/AS.
+We did that for statistical convenience, as reviewer 1 is correct in pointing out.
+But we can look at partitioning variance in MF/MS directly, to plant height, PFT and climate.
+Reviewer 3 (though incorrectly) says 'also have the same variable in the numerator and denominator, which causes statistical issues' - but then we can say we avoid this apparent problem.
+\item We claim that climate does not affect biomass distribution, but reviewer 1 correctly points out that climate affects plant height, so it may have an indirect effect.
+Our dataset is not appropriate for testing effects of climate on typical (or maximum) plant height, because individuals of all kinds of sizes were sampled; there will be sampling bias.
+Moles et al.
+2009 reported that maximum plant height shows a strong latitudinal pattern (taller plants at low latitude), and is correlated with climate variables.
+We just have to describe their results.
+\item Can follow reviewer 1's suggestion to partition MF/MS like this: $\frac{M_F}{M_S} = \frac{A_F}{A_S}\frac{M_F}{A_F}\frac{A_S}{M_S}$
+\item Reviewer 2 has some strange comments about MS/AS, in that we expect them to be correlated.
+Yes, but we need to make the point more clearly that there could be PFT effects in this correlation, largely through wood density (and tree shape).
+In fact we bring this up, but partitioning the variance as above may make this stand out more.
+\item "The methods do not describe if the individual data points came from trees in closed-canopy or open stands.  That would also make a huge difference." (Reviewer 3)
+\end{enumerate}
+
+\section{New list of figures}
+\begin{enumerate}
+\item Tree diagram plus climate space of dataset. Unchanged from before, except remove MT (?).
+\item MF and MS as function of tree height.
+Reviewer 2 asks "But are differences in Figure 2 statistically signficant?", weird question and subsequent comments.
+Figure could add trend lines (as smoothers), but refer to later figures for significant differences.
+The comment that sampling may have been uneven is correct, which is why we correct for plant height in the analyses.
+\item (a) MF/MS vs.
+plant height by PFT (was LMF) (b) and (c) as before, except for MF/MS.
+Reviewer 2 wants us to remove panels d and e, which we can do because we won't need MF/AS as a proxy anymore (should be put in SuppInfo instead).
+\item Histograms by PFT. This was nice when we used size-invariant MF/AS, but we can still show LMA and AF/AS? The first has large PFT differences, the second not any.
+\item Not sure how current Fig 5 will fit in the new version
+\item Not sure how current Fig 6 will fit in the new version, since we won't be able to use MF/AS.
+Can make this plot for AF/AS though, since reviewer 1 thinks that has biological relevance (and MF/AS doesn't, for some reason).
+\end{enumerate}
+
 \section*{Summary}
@@
 
+% less focus on GVMs; focus on ecological/biological reasons to test these things
+% ... but make sense to look at PFTs though, and this is the actual reason for doing that?
 Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation.
@@
 
+%explain why stem/leaf biomass distribution is important here.
 In this paper we are primarily interested in the distribution of biomass between leaves and woody stems, an important component of the residence time of carbon in ecosystems \citep{friend2014}.
@@
 
+% integrate poorter2015 and reich2014 better
 Despite considerable advances in theory underlying allometric scaling in plants \citep{enquist2003, west1999, savage2010} we do not yet have a clear understanding of potential differences between PFTs in terms of biomass partitioning.
@@
 
+% no longer needed? we will use biomass distribution instead of partitioning to be consistent with Poorter 2015.
 \subsection*{Terminology}
@@
 
+% we used basal area because many plants were <1.3m tall
 Using the subset of data where basal stem diameter ($A_S$) and diameter at breast height ($A_{Sbh}$) were measured (a total of \Sexpr{length(residuals(basalafit))} observations covering the three major PFTs), we developed a non-linear model to predict $A_S$ when only $A_{Sbh}$ and plant height ($H$) were measured.
@@
 <<results='asis'>>=
-bigtable <- rbind(cbind(c("$M_F/M_T$","","",""),c("DA","EA","EG","Sum"), table_lmf),
-                  cbind(c("$A_F/M_T$","","",""),c("DA","EA","EG","Sum"), table_lar), 
-                  cbind(c("$M_F/A_S$","","",""),c("DA","EA","EG","Sum"), table_mlfastba),
-                  cbind(c("$A_F/A_S$","","",""),c("DA","EA","EG","Sum"), table_alfastba))
-rownames(bigtable) <- NULL
-colnames(bigtable)[1:2] <- c("Variable","PFT")
 
-print(xtable(bigtable, align="llrlll|l",
-               caption="Sample sizes used in the analyses of the four variables considered in the study by plant functional type (PFT, DA = deciduous angiosperm, EA = evergreen angiosperm, EG = evergreen gymnosperm) and biome.
+print(xtable(table_hierpart),sanitize.text.function=function(x){x},
+      include.rownames=FALSE)
+
+@
+
+%' 
+%' <<results='asis'>>=
+%' bigtable <- rbind(cbind(c("$M_F/M_T$","","",""),c("DA","EA","EG","Sum"), table_lmf),
+%'                   cbind(c("$A_F/M_T$","","",""),c("DA","EA","EG","Sum"), table_lar), 
+%'                   cbind(c("$M_F/A_S$","","",""),c("DA","EA","EG","Sum"), table_mlfastba),
+%'                   cbind(c("$A_F/A_S$","","",""),c("DA","EA","EG","Sum"), table_alfastba))
+%' rownames(bigtable) <- NULL
+%' colnames(bigtable)[1:2] <- c("Variable","PFT")
+%' 
+%' print(xtable(bigtable, align="llrlll|l",
+%'                caption="Sample sizes used in the analyses of the four variables considered in the study by plant functional type (PFT, DA = deciduous angiosperm, EA = evergreen angiosperm, EG = evergreen gymnosperm) and biome.
 Sample sizes denote number of individuals, with number of unique species in parentheses.
 ",
-             label="table:counttable"), 
-      hline.after=0:16, caption.placement="top",  #c(0,4,8,12,16)
-      sanitize.text.function=function(x){x},
-      include.rownames=FALSE)
-@
-
-<<results='asis'>>=
-tab1 <- table_varpart2
-tab2 <- table_gamr2MATMAP
-
-tidytab <- function(x, labeltopleft=""){
-  names(x)[1] <- "Variable"
-  x[,1] <- paste0(x[,1],"(",x[,2],")")
-  x[,2] <- NULL
-  x[,2:ncol(x)] <- round(x[,2:ncol(x)], 2)  
-  x <- unname(rbind(names(x),as.matrix(x)))  
-  x <- rbind(c(labeltopleft,"","Predictors",""),x)
-return(x)
-}
-
-tab1 <- tidytab(tab1, "Mixed-effects model")
-tab2 <- tidytab(tab2, "GAM")
-
-tab <- rbind(tab1, tab2)
-
-print(xtable(tab, align="lrrrr",
-             caption="Explained variance by plant functional type (PFT), plant height (H) and climate variables (either biome (B), or MAP and MAT) for five whole-plant variables.
+%'              label="table:counttable"), 
+%'       hline.after=0:16, caption.placement="top",  #c(0,4,8,12,16)
+%'       sanitize.text.function=function(x){x},
+%'       include.rownames=FALSE)
+%' @
+%' 
+%' 
+%' 
+%' <<results='asis'>>=
+%' tab1 <- table_varpart2
+%' tab2 <- table_gamr2MATMAP
+%' 
+%' tidytab <- function(x, labeltopleft=""){
+%'   names(x)[1] <- "Variable"
+%'   x[,1] <- paste0(x[,1],"(",x[,2],")")
+%'   x[,2] <- NULL
+%'   x[,2:ncol(x)] <- round(x[,2:ncol(x)], 2)  
+%'   x <- unname(rbind(names(x),as.matrix(x)))  
+%'   x <- rbind(c(labeltopleft,"","Predictors",""),x)
+%' return(x)
+%' }
+%' 
+%' tab1 <- tidytab(tab1, "Mixed-effects model")
+%' tab2 <- tidytab(tab2, "GAM")
+%' 
+%' tab <- rbind(tab1, tab2)
+%' 
+%' print(xtable(tab, align="lrrrr",
+%'              caption="Explained variance by plant functional type (PFT), plant height (H) and climate variables (either biome (B), or MAP and MAT) for five whole-plant variables.
 For the analysis with biome, each variable was added to a linear mixed-effects model, using species within dataset as a random effect.
@@
 Note that explained variance for the linear mixed-effects model and GAM is not necessarily the same even with the same predictors, due to different methods for fitting and estimating the explained variance.",
-             label="table:statstable"),
-             hline.after=0:14, #c(0,1,7,8,14),
-             include.rownames=FALSE, 
-             include.colnames=FALSE, 
-             caption.placement="top", sanitize.text.function=function(x){x})
-@
+%'              label="table:statstable"),
+%'              hline.after=0:14, #c(0,1,7,8,14),
+%'              include.rownames=FALSE, 
+%'              include.colnames=FALSE, 
+%'              caption.placement="top", sanitize.text.function=function(x){x})
+%' @
+%'
 
-\clearpage
-\section{Figures}
-
-\begin{figure}[h!]
-    \centering
-    \ifincludefigs 
-    \includegraphics[width=0.99\textwidth]{Figure1.pdf}
-    \fi
-    \caption{Overview of the data.
+%' 
+%' \clearpage
+%' \section{Figures}
+%' 
+%' 
+%' \begin{figure}[h!]
+%'     \centering
+%'     \ifincludefigs 
+%'     \includegraphics[width=0.99\textwidth]{Figure1.pdf}
+%'     \fi
+%'     \caption{Overview of the data.
 (a) Variables were measured on up to \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species.
@@
 Colour symbols show the locations of sampled individuals for three dominant woody functional types.}
-    \label{fig:figoverview}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \ifincludefigs 
-    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
-    \fi
-    \caption{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$).
+%'     \label{fig:figoverview}
+%' \end{figure}
+%' 
+%' 
+%' \begin{figure}[h!]
+%'     \centering
+%'     \ifincludefigs 
+%'     \includegraphics[width=0.99\textwidth]{Figure2.pdf}
+%'     \fi
+%'     \caption{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$).
 Each point is an individual plant.
 Sample sizes are listed in Table \ref{table:counttable}}
-    \label{fig:figrawdata}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \ifincludefigs 
-    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
-    \fi
-    \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf mass per area.
+%'     \label{fig:figrawdata}
+%' \end{figure}
+%' 
+%' \begin{figure}[h!]
+%'     \centering
+%'     \ifincludefigs 
+%'     \includegraphics[width=0.99\textwidth]{Figure3.pdf}
+%'     \fi
+%'     \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf mass per area.
 (a) Leaf mass fraction ($M_F/M_T$ = leaf mass / above-ground biomass) by PFT.
@@
 Letters denote significant differences (at $\alpha = 0.05$).}
-    \label{fig:figlmf}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \ifincludefigs 
-    \includegraphics[width=0.99\textwidth]{Figure4.pdf}
-    \fi
-    \caption{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.
+%'     \label{fig:figlmf}
+%' \end{figure}
+%' 
+%' \begin{figure}[h!]
+%'     \centering
+%'     \ifincludefigs 
+%'     \includegraphics[width=0.99\textwidth]{Figure4.pdf}
+%'     \fi
+%'     \caption{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.
 Shown are histograms (as probability density functions) of leaf mass per area ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three PFTs.
@@
 \ref{fig:figlmf}.}
-    \label{fig:fighist}
-\end{figure}
-
-<<echo=FALSE>>=
-agg <- summaryBy(lalf_astba2 + lmlf_astba2 + llma ~ Group,
-                   data=dataset, FUN=mean, na.rm=TRUE, keep.names=TRUE,
-                   id=~pft)
-agg <- subset(agg, !is.na(llma))
-lm1 <- lm(lmlf_astba2 ~ llma * pft, data=agg)
-lm2 <- lm(lmlf_astba2 ~ lalf_astba2 * pft, data=agg)
-r2panela <- summary(lm1)$adj.r.squared
-r2panelb <- summary(lm2)$adj.r.squared
-@
-
-\begin{figure}[h!]
-    \centering
-    \ifincludefigs 
-    \includegraphics[width=0.99\textwidth]{Figure5.pdf}
-    \fi
-     \caption{Leaf and stem traits capture variation in biomass partitioning across species.
+%'     \label{fig:fighist}
+%' \end{figure}
+%' 
+%' 
+%' <<echo=FALSE>>=
+%' agg <- summaryBy(lalf_astba2 + lmlf_astba2 + llma ~ Group,
+%'                    data=dataset, FUN=mean, na.rm=TRUE, keep.names=TRUE,
+%'                    id=~pft)
+%' agg <- subset(agg, !is.na(llma))
+%' lm1 <- lm(lmlf_astba2 ~ llma * pft, data=agg)
+%' lm2 <- lm(lmlf_astba2 ~ lalf_astba2 * pft, data=agg)
+%' r2panela <- summary(lm1)$adj.r.squared
+%' r2panelb <- summary(lm2)$adj.r.squared
+%' @
+%' 
+%' \begin{figure}[h!]
+%'     \centering
+%'     \ifincludefigs 
+%'     \includegraphics[width=0.99\textwidth]{Figure5.pdf}
+%'     \fi
+%'      \caption{Leaf and stem traits capture variation in biomass partitioning across species.
 Individual plant data were averaged by species and study combinations.
@@
 $R^2$ for fitted relationships are \Sexpr{round(100*r2panela,1)}\% in panel (a) and \Sexpr{round(100*r2panelb,1)}\% in panel (b) (for a linear regression model including PFT as a factor variable).}
-  \label{fig:figlmfspecies}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \ifincludefigs 
-    \includegraphics[width=0.99\textwidth]{Figure6.pdf}
-    \fi
-    \caption{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning.
+%'   \label{fig:figlmfspecies}
+%' \end{figure}
+%' 
+%' 
+%' \begin{figure}[h!]
+%'     \centering
+%'     \ifincludefigs 
+%'     \includegraphics[width=0.99\textwidth]{Figure6.pdf}
+%'     \fi
+%'     \caption{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning.
 a) Leaf mass per unit stem cross-sectional area ($M_F/A_S$) for each PFT within three different biomes.
@@
 Solid lines are generalised additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
-    \label{fig:figclimate}
-\end{figure}
+%'     \label{fig:figclimate}
+%' \end{figure}
+%' 
+%'
 

### 2015-11-05 R/tables_stats.R

@@ -1,4 +1,43 @@
 
+do_hierpart <- function(dep, indep, data, ...){
+  
+  df <- data[,c(dep,indep)]
+  df <- df[complete.cases(df),]
+  
+  y <- df[,dep]
+  m <- df[,indep]
+  
+  h <- hier.part(y, m, ...) 
+  
+  h$N <- nrow(df)
+  
+  return(h)
+}
+
+make_table_hierpart <- function(dataset){
+  vars <- c("lmlf_mst","lalf_astba2","llma","lastba2_mst")
+  indep_vars <- c("lh.t","pft","MAT","MAP","MATMAP")
+  hp <- lapply(vars, function(x){
+    do_hierpart(x, indep_vars, dataset, gof="Rsqu", barplot=FALSE)
+  })
+  
+  hp_maketablerow <- function(x){
+    iperc <- c(x$I.perc[c("lh.t","pft"),], sum(x$I.perc[c("MAP","MAT","MATMAP"),]))
+    r2 <- max(x$gfs)
+    
+    ipercrel <- r2* iperc / 100
+    
+    c(ipercrel,r2)
+  }
+  
+  tab <- as.data.frame(t(sapply(hp, hp_maketablerow)))
+  tab <- cbind(c("$M_F/M_S$","$A_F/A_S$","$M_F/A_F$","$M_S/A_S$"), tab)
+  colnames(tab) <- c("Variable","$H_T$","PFT","Climate","$R^2$ total")
+  tab <- as.data.frame(tab)
+  return(tab)
+}
+
+
 make_r2_table <- function(data,  variable_name ){
   data.frame(Variable = c(variable_name, rep(NA, length(data)-1)),
     Predictors=names(data),

### 2015-11-06 manuscript.Rnw
Figures are updated.

@@
 \item "The methods do not describe if the individual data points came from trees in closed-canopy or open stands.  That would also make a huge difference." (Reviewer 3)
+\item We used MAT, MAP as well as the interaction.
+Probably better climate variables out there, but having the interaction in a model does distinguish between warm-dry and warm-wet climates (just like aridity indices).
+Because we found no appreciable effects of climate, we did not think it was worthwhile testing all possible climate variables.
 \end{enumerate}

### 2015-11-06 R/figures.R

@@ -191,59 +191,8 @@ figure2 <- function(dataset){
 
 
 
-# Figure 3 - a) leaf mass fraction by PFT, b) average LMF and LAR at mean H by PFT, c) average MF/AS and AF/AS
+
 figure3 <- function(dataset, KGAM=4){
-
-  l <- layout(matrix(c(1,2,4,1,3,5), byrow=T, ncol=3),
-              widths=c(1,0.67,0.67), heights=c(1,1))
-
-  par(mar=c(4,4,4,1), cex.axis=0.9, cex.lab=1.3, mgp=c(2.3,0.5,0), tcl=-0.35, las=1)
-  obj1 <- smoothplot(lh.t, lmlf_mso, pft, dataset, R="Group",linecols=my_linecols(),
-                     pointcols=my_cols_transparent(),axes=FALSE,
-             xlab="Plant height (m)",kgam=KGAM,
-             ylab=expression(M[F]/M[T]~~(kg~kg^-1))
-  )
-  log10axes()
-  my_legend("bottomleft", "long")
-  box()
-  plotlabel("(a)","topright")
-
-  obj2 <- smoothplot(lh.t, lalf_mso, pft, dataset, R="Group",plotit=FALSE)
-
-  par(mar=c(0.35,4,4,1), pty="m") #, oma=c(0,0,0,0))
-
-  xpred <- mean(dataset$lh.t, na.rm=TRUE)
-  plotGamPred(obj1, dataset, xpred=xpred,
-              ylab=expression(M[F]/M[T]~~(kg~kg^-1)),ylim=c(0,0.35),
-              xlim=c(0,0.2), xlab="", xaxislabels=FALSE)
-  plotlabel("(b)","topright")
-
-  par(mar=c(4,4,0.35,1)) #, oma=c(0,0,0,0))
-  plotGamPred(obj2, dataset, xpred=xpred,
-              ylab=expression(A[F]/M[T]~~(kg~kg^-1)),ylim=c(0,2.5),
-              xlim=c(0,0.2), xlab=lmaLabel_short())
-  plotlabel("(c)","topright")
-
-  par(mar=c(0.35,4,4,1), pty="m")
-  meansbypft("lmlf_astba2","lalf_astba2", "pft",
-             xvar="llma",
-             dataset=dataset,
-             setpar=FALSE,
-             addlegend=FALSE,
-             panel1.expr={axis(2);plotlabel("(d)","topleft");par(mar=c(4,4,0.35,1))},
[…]

### 2015-11-06 run.R

@@ -58,9 +58,18 @@ table_hierpart <- make_table_hierpart(dataset)
 
 
 
-# new figures
-pdf("figures/Figure3new.pdf", width = 8L, height = 4L)
-figure3new(dataset)
+# Figures
+pdf("figures/Figure1.pdf", width = 8L, height = 4L)
+figure1(baad_mapmat, world_mapmat, "downloads/ian-symbol-eucalyptus-spp-1.png")
+dev.off()
+pdf("figures/Figure2.pdf", width = 8L, height = 4L)
+figure2(dataset)
+dev.off()
+pdf("figures/Figure3.pdf", width = 8L, height = 4L)
+figure3(dataset)
+dev.off()
+pdf("figures/Figure4.pdf", width = 8L, height = 4L)
+figure4(dataset)
 dev.off()
 
 
@@ -80,18 +89,8 @@ tex_2_pdf("manuscript.tex")
 
 
 # old figures
-pdf("figures/Figure1.pdf", width = 8L, height = 4L)
-figure1(baad_mapmat, world_mapmat, "downloads/ian-symbol-eucalyptus-spp-1.png")
-dev.off()
-pdf("figures/Figure2.pdf", width = 8L, height = 4L)
-figure2(dataset)
-dev.off()
-pdf("figures/Figure3.pdf", width = 8L, height = 4L)
-figure3(dataset)
-dev.off()
-pdf("figures/Figure4.pdf", width = 8L, height = 4L)
-figure4(dataset)
-dev.off()
+
+
 pdf("figures/Figure5.pdf", width = 8L, height = 4L)
 figure5(dataset)
 dev.off()

### 2015-11-11 manuscript.Rnw
almost there

@@
 Reviewer 2 wants us to remove panels d and e, which we can do because we won't need MF/AS as a proxy anymore (should be put in SuppInfo instead).
-\item Histograms by PFT. This was nice when we used size-invariant MF/AS, but we can still show LMA and AF/AS? The first has large PFT differences, the second not any.
-\item Not sure how current Fig 5 will fit in the new version
-\item Not sure how current Fig 6 will fit in the new version, since we won't be able to use MF/AS.
-Can make this plot for AF/AS though, since reviewer 1 thinks that has biological relevance (and MF/AS doesn't, for some reason).
+\item AF vs. AS by pft, for all data, with fitted SMA. This shows AF is proportional to AS (approx, except larger plants), and no/very very small effect of PFT.
+\item Stem mass as function of AS, PFT and height.
+a) Raw data (to link to main equation) show no clear PFT effect.
+b) MS / AS is nearly proportional to H, as expected from simple stem mass equation.
+c) for larger plants (H*AS > 0.01m3), there is a significant PFT effect, with EG having lower MS / (AS*H) (see d) for averages across entire subset shown in c).
+This can be explained from differences in wood density; gymnosperms have lower wood density on average.
+\item Histograms of the three main variables: AF/AS, MF/AF, and height-corrected stem mass index, MS/(AS*H).
 \end{enumerate}
@@
 
+Much too long, of course!
 % 200 word max.
@@
 \item The partitioning of biomass into leaves and stems is an important component of  global vegetation models (GVMs).
-Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical studies have not been able to justify these differences.
-Here we test for differences between PFTs in distribution of biomass between leaves and stems.
 \item We use the recently published Biomass And Allometry Database (BAAD), a large database including observations for individual plants.
 The database covers the global climate space, allowing us to test for direct climate effects in addition to PFT.
-\item The leaf mass fraction (LMF, leaf / total aboveground biomass) varied strongly between PFTs (as defined by deciduous vs.
+\item The ratio of leaf mass to woody above-ground mass (MF/AS) varied strongly between PFTs (as defined by deciduous vs.
 evergreen and gymnosperm vs.
 angiosperm).
-We found that LMF, once corrected for plant height, was proportional to leaf mass per area across PFTs.
+We found that MF/MS, once corrected for plant height, was proportional to leaf mass per area across PFTs.
 As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass.
-We found only weak and inconsistent effects of climate on biomass partitioning.
-\item Combined, these results will allow for systematic benchmarking of biomass partitioning routines in GVMs.
+\item We partitioned the variance of MF/MS into three component variables (AF/AS, MF/AF and MS/AS), and each of these four variables into PFT, climate and plant height.
+Plant height explained most variation in MF/MS and MS/AS (but not H or MF/AF).
+PFT explained much variation in MF/MS and MF/AF (but not AF/AS and only for larger plants in MS/AS).
+\item Climate variables (MAT, MAP and its interaction) consistently explained very little additional variation, demonstrating very weak control of climate on biomass distribution and leaf area.
+We argue that climate effects will largely be indirectly through plant height.
+\item Combined, these results show consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain GVMs.
 \end{enumerate}
 
-\emph{Keywords}: allocation, plant allometry, biomass estimation, specific leaf area, dynamic global vegetation model
+\emph{Keywords}: allocation, plant allometry, biomass estimation, specific leaf area
 
@@
 
-The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation \citep{ise2010, dekauwe2014,friend2014}.
+\begin{itemize}
+\item Biomass allocation is important; need to improve GVMs and make them more consistently benchmarked against global observations. 
+\item Plant functional types; can expect differences in biomass partitioning and leaf area relationships.
+Useful from viewpoint of ecology (functional consequences of angiosperm vs.
+gymnosperm divide; evergreen vs.
+deciduous?).
+Also useful for GVMs which use PFT for parameterization.
+\item Enquist&Niklas showed differences in leaf mass at given stem mass between gymnosperms and angiosperms.Poorter2015 showed X.
+Reich2014 showed Y.
+Open questions are 1,2,3.
+One major gap is that these previous studies did not include estimates of whole-plant leaf area.
+Leaf area is relevant because bla, but may follow very different patterns from leaf mass.
+Poorter2015 speculated that X.
+\item What about climate?
+We know from small-scale experiments and observational studies that biomass distribution varies with climate drivers.
+But unclear if global patterns persist.
+Reich2014 showed X, consistent with glasshouse stuff from Poorter2015.
+\item We have the following objectives:
+\begin{enumerate}
+\item Test whether MF/MS is proportional to leaf mass per area (as suggested by Poorter)
+\item Partition the variance of leaf/stem biomass distribution into component variables, and into plant height (which is important, obviously!), plant functional type and climate. 
+\end{enumerate}
+
+\end{itemize}
+
+% The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation \citep{ise2010, dekauwe2014,friend2014}.
 Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \citep{ryanlaw2005}.
@@
 Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014,negron2015}.
-
-% less focus on GVMs; focus on ecological/biological reasons to test these things
-% ... but make sense to look at PFTs though, and this is the actual reason for doing that?
-Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation.
+% 
+% % less focus on GVMs; focus on ecological/biological reasons to test these things
+% % ... but make sense to look at PFTs though, and this is the actual reason for doing that?
+% Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation.
 While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots).
@@
 These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared \citep{wolf2011}.
-
-%explain why stem/leaf biomass distribution is important here.
-In this paper we are primarily interested in the distribution of biomass between leaves and woody stems, an important component of the residence time of carbon in ecosystems \citep{friend2014}.
+% 
+% %explain why stem/leaf biomass distribution is important here.
+% In this paper we are primarily interested in the distribution of biomass between leaves and woody stems, an important component of the residence time of carbon in ecosystems \citep{friend2014}.
 Previous work based on either whole stands \citep{oneilldeangelis1981,enquistniklas2002,reich2014} and or a mix on stand-- and individual-based measurements \citep{poorter2012, poorter2015}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
@@
 Some oft-cited studies have also assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
-
-Little is known about global-scale patterns in LMF and LAR in relation to climate.
+% 
+% Little is known about global-scale patterns in LMF and LAR in relation to climate.
 It can be expected that biomass partitioning is correlated with precipitation or mean annual temperature because smaller-scale comparisons have shown responsiveness of biomass partitioning to environmental drivers \citep{berninger1994,callaway1994,delucia2000,poyatos2007}.
@@
 Again, however, these models do not agree on the degree of plasticity in biomass partitioning, or which climate variables it should respond to.
-
-% integrate poorter2015 and reich2014 better
-Despite considerable advances in theory underlying allometric scaling in plants \citep{enquist2003, west1999, savage2010} we do not yet have a clear understanding of potential differences between PFTs in terms of biomass partitioning.
+% 
+% % integrate poorter2015 and reich2014 better
+% Despite considerable advances in theory underlying allometric scaling in plants \citep{enquist2003, west1999, savage2010} we do not yet have a clear understanding of potential differences between PFTs in terms of biomass partitioning.
 Instead, most previous work has focussed on understanding size-related shifts in biomass partitioning, as governed by constraints including hydraulic supply and mechanical stability \citep{savage2010}.
@@
 Here we can directly test this hypothesis as our database, unlike those of \citet{enquistniklas2002}, \citet{reich2014}, and \cite{poorter2015}, includes many observations of leaf area as well as leaf mass measured on the same plants.
-
-A second objective was to study relationships between biomass partitioning and climate variables or biome at a global scale.
+% 
+% A second objective was to study relationships between biomass partitioning and climate variables or biome at a global scale.
 The dataset includes observations of biomass and size metrics for individual plants, compiled from 175 studies across nine vegetation types (Fig.
@@
 one that does not assume a particular functional form, thereby allowing us to study PFT patterns at a common plant height.
+%
 
 \section*{Materials and Methods}
-
-% no longer needed? we will use biomass distribution instead of partitioning to be consistent with Poorter 2015.
-\subsection*{Terminology}
-\label{sec:partterm}
-
-The terms 'partitioning' and 'allocation' have been used in various ways, confusing comparisons between studies \citep{litton2007}.
-Here, we define biomass partitioning as the actual distribution of biomass between compartments (e.g., leaves vs.
-stems), and biomass allocation as the proportion of net primary production (NPP) that is allocated to some compartment.
-The two processes are different because of continuous turnover of biomass, which differs strongly between compartments.
-We may write \citep{mcmurtriewolf1983},
-
-$$
-\frac{dM_i}{dt} = a_i \textrm{NPP} - \tau_i M_i
-$$
-
-where $M_i$ is the biomass in some compartment (leaves, stems or roots) remaining on the plant, $a_i$ the annual allocation of NPP to that compartment, and $\tau_i$ the annual turnover (or loss) of compartment $i$ from the plant.
-It is thus easy to see that partitioning ($M_i/M_T$, where $M_T$ is total biomass) can be different from allocation ($a_i$) because turnover ($\tau_i$) differs between leaf and wood biomass.
-Here, we present data on biomass partitioning, which can inform models of allocation only when estimates of turnover are available.
 
@@
 To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).
-
-For the above-ground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total above-ground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,
-
-$$\textrm{LMF} = \frac{M_F}{M_T} = \left(\frac{M_F}{A_F}\right) \left(\frac{A_F}{M_T}\right) = \textrm{LMA} \times \textrm{LAR}.$$
 
@@
 
-Despite the exceptional size of our dataset, the strong size-dependence of LMF still hinders comparisons across climatic gradients, due to the small sample size available within each species or site when sampling at a common height.
-To study climate effects on biomass partitioning, we therefore further decomposed LMF as the product,
-
-\begin{equation}\label{eq:M1}
-\textrm{LMF} = \left(\frac{M_F}{A_S}\right) \left(\frac{A_S}{M_T}\right).
-\end{equation}
-
-This decomposition showed that the vast majority of size-related variation was captured by $A_S/M_T$ alone, indicating that comparisons among $M_F/A_S$ could be made without needing to compare at a common height.
-A similar decomposition  of LAR as the product of $A_F/A_S$ and $A_S/M_T$ produced the same outcome.
-We therefore analysed for climatic effects on $M_F/A_S$ and $A_F/A_S$ in two ways: with quantitative variables (mean annual precipitation, MAP; mean annual temperature, MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig.
-S1).
-In both cases we used PFT and plant height as covariates varying across climate space.
-
 All analyses were conducted in R v3.2.0 \citep{R-2015}.
@@
 
-The raw data in Fig.~\ref{fig:figrawdata} show a steeper increase of woody aboveground biomass ($M_S$) with plant height compared to foliage biomass ($M_F$).
-As a result, LMF decreased with plant height, with the three plant functional types clearly differing in LMF across nearly the entire size range (Fig.~\ref{fig:figlmf}a).
-When we corrected for plant height by estimating LMF at a common plant height, we found that LMF was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig.~\ref{fig:figlmf}b).
-As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig.
-~\ref{fig:figlmf}c and Fig.
-S2).
-
-Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differed among the three PFTs, that these differences were also proportional to LMA (Fig.
-~\ref{fig:figlmf}d), and that PFTs were similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig.~\ref{fig:figlmf}e).
-
-The above results highlight how LMA drives differences in biomass partitioning among dominant woody PFTs.
-To assess whether LMA could replace PFT when predicting biomass partitioning, we fit a linear mixed-effects model to LMF and $M_F/A_S$ using PFT and plant height (and the quadratic term) as predictors (and all interactions).
-We then replaced PFT with LMA, and found that the model with LMA could explain almost as much variation in LMF as the model with PFT ($R^2$ = 0.74 with PFT vs.
-0.62 with LMA), likewise for $M_F/A_S$ ($R^2$ = 0.29 vs.
-0.28).
-
-There was considerable variation in all studied variables between species within PFTs (Fig.
-~\ref{fig:fighist}).
-To understand whether we can improve on a PFT-based classification by including other traits that affect biomass partitioning, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$):
-
-\begin{equation}\label{eq:M2}
-\frac{M_F}{A_S} = \left(\frac{M_F}{A_F}\right) \left(\frac{A_F}{A_S}\right).
-\end{equation}
-
-The ratio $M_F/A_S$ is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}, using records of stem diameter commonly recorded on long-term monitoring plots.
-The correlation between $M_F/A_S$ and LMA was found to hold also within PFTs (Fig.~\ref{fig:figlmfspecies}a), explaining 30\% of the variation in $M_F/A_S$ across species.
-The regressions across species by PFT were broad and overlapping, with a more general relationship extending across the entire LMA axis.
-A larger fraction of the variation in $M_F/A_S$ was explained by species-level differences in $A_F/A_S$ (Fig.~\ref{fig:figlmfspecies}b).
-As previously noted, $A_F/A_S$ does not differ systematically among PFTs (Fig.~\ref{fig:figlmf}e), but does vary close to two orders of magnitude across species.
-In Fig.~\ref{fig:figlmfspecies}b, the separation among PFTs in $M_F/A_S$ arises due to differences in LMA.
-
-We found that decomposing LMF and LAR as shown in Eq.
-\ref{eq:M1} was very useful to study climate effects because the second term ($A_S/M_T$) absorbed nearly all of the size-related variation in LMF (Table \ref{table:statstable} and Fig.
-S3) and is otherwise fairly conserved across PFTs (Fig.
-S3).
-As shown in Fig.~\ref{fig:figlmfspecies}b and Fig.~\ref{fig:figlmf}d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Table \ref{table:statstable}).
-$M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.
-
-Regardless of how we analysed the data, we found only weak climatic effects on biomass partitioning within PFTs (Fig.~\ref{fig:figclimate}).
-Biome consistently explained very little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by only 0.01 - 0.06, see Table \ref{table:statstable}).
-Likewise, we found weak and inconsistent effects of climate variables (MAP and MAT) (Fig.~\ref{fig:figclimate}b-c).
-These variables explained little variation when added to a statistical model in addition to PFT and plant height ($R^2$ increased by 0.01 and 0.11, respectively, see Table \ref{table:statstable}).
+\begin{itemize}
+\item Dataset covers climate space pretty well. 
+\item MF and MS are non-linearly related to plant height (Fig 2), as expected. Also note stem mass increases with height more quickly than leaf mass.
+\item Leaf/stem biomass distribution was very different between PFTs (Fig 3a), and proportional to LMA when estimated at a common height, so that leaf area ratio did not differ between PFTs.
+Consistent with variance partitioning: MF/MS mostly explained by H and PFT, but LAR not much explained by PFT.
+\item 
+\end{itemize}
 
@@
 
-\textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\
-\textbf{Fig. S2} Leaf area ratio ($A_F/M_T$) by PFT.\\
-\textbf{Fig. S3} Relationship between above-ground biomass and basal stem area.\\
-\textbf{Notes S1} Modelling of biomass partitioning in global vegetation models (GVMs)
+% \textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\
+% \textbf{Fig. S2} Leaf area ratio ($A_F/M_T$) by PFT.\\
+% \textbf{Fig. S3} Relationship between above-ground biomass and basal stem area.\\
+% \textbf{Notes S1} Modelling of biomass partitioning in global vegetation models (GVMs)
 

### 2015-11-11 R/figures.R

@@ -191,7 +191,7 @@ figure2 <- function(dataset){
 
 
 
-
+# Main partitioning figure (MF/MS)
 figure3 <- function(dataset, KGAM=4){
   
   l <- layout(matrix(c(1,2,1,3), byrow=T, ncol=2),
@@ -201,7 +201,7 @@ figure3 <- function(dataset, KGAM=4){
   obj1 <- smoothplot(lh.t, lmlf_mst, pft, dataset, R="Group",linecols=my_linecols(),
                      pointcols=my_cols_transparent(),axes=FALSE, ylim=c(-3,1),
                      xlab="Plant height (m)",kgam=KGAM,
-                     ylab=expression(M[F]/M[W]~(kg~kg^-1))
+                     ylab=expression(M[F]/M[S]~(kg~kg^-1))
   )
   log10axes()
   my_legend("bottomleft", "long")
@@ -213,7 +213,7 @@ figure3 <- function(dataset, KGAM=4){
   
   xpred <- mean(dataset$lh.t, na.rm=TRUE)
   plotGamPred(obj1, dataset, xpred=xpred,
-              ylab=expression(M[F]/M[W]~(kg~kg^-1)),ylim=c(0,0.5),
+              ylab=expression(M[F]/M[S]~(kg~kg^-1)),ylim=c(0,0.5),
               xlim=c(0,0.2), xlab="", xaxislabels=FALSE)
   plotlabel("(b)","topright")
   
@@ -221,111 +221,122 @@ figure3 <- function(dataset, KGAM=4){
   
   obj2 <- smoothplot(lh.t, lalf_mst, pft, dataset, R="Group",plotit=FALSE)
   plotGamPred(obj2, dataset, xpred=xpred,
-              ylab=expression(A[F]/M[W]~(kg~kg^-1)),ylim=c(0,4),
+              ylab=expression(A[F]/M[S]~(m^2~kg^-1)),ylim=c(0,4),
               xlim=c(0,0.2), xlab=lmaLabel_short())
   plotlabel("(c)","topright")
+
+}
+
+
+# new figure 4
+figure4 <- function(dataset){
   
+  # random reorder dataset to avoid plotting artefact
+  dataset <- dataset[sample(nrow(dataset)),]
   
+  f <- sma(lalf ~ lastba2*pft, data=dataset)
+  par(mar=c(5,5,1,1), cex.lab=1.2)
[…]

### 2015-11-11 run.R

@@ -44,6 +44,17 @@ dat_mlf <- prepare_dat_mlf(dataset2)
 dat_alf <- prepare_dat_alf(dataset2)
 basalafit <- BasalA_fit(baad_all)
 
+# see new small plant test figure
+delsmallbh <- function(d){
+  
+  # plants close to breast-height height but no basal diameter; delete
+  d$a.stba2[d$h.t < 1.8 & is.na(d$d.ba) & !is.na(d$d.bh)] <- NA
+  
+return(d)
+}
+dataset <- delsmallbh(dataset)
+dataset2 <- delsmallbh(dataset2)
+
 # # tables
 # table_gamr2MATMAP <- make_table_gamr2MATMAP(dataset)
 # table_mlfastba <- make_table_mlfastba(dat_mlf)
@@ -56,6 +67,7 @@ basalafit <- BasalA_fit(baad_all)
 # new
 table_hierpart <- make_table_hierpart(dataset)
 
+table_varpart_gam <- make_table_gamr2MATMAP(dataset)
 
 
 # Figures
@@ -68,20 +80,32 @@ dev.off()
 pdf("figures/Figure3.pdf", width = 8L, height = 4L)
 figure3(dataset)
 dev.off()
-pdf("figures/Figure4.pdf", width = 8L, height = 4L)
+
+pdf("figures/Figure4.pdf", width = 6L, height = 5L)
 figure4(dataset)
 dev.off()
 
+pdf("figures/Figure5.pdf", width = 8L, height = 8L)
+figure5(dataset)
+dev.off()
 
-
-# SuppInfo???
-pdf("figures/Figure5.pdf", width = 8L, height = 4L)
-figure5(dataset, dataset2)
+pdf("figures/Figure6.pdf", width = 8L, height = 4L)
+figure6(dataset, nbin=75)
 dev.off()
 
 
+# SI
+pdf("figures/FigureS1.pdf", width = 6L, height = 6L)
+figureS1(baad_mapmat, world_mapmat)
+dev.off()
 
+pdf("figures/FigureS2.pdf", width = 6L, height = 5L)
+figureS2(dataset)
+dev.off()
 
+pdf("figures/FigureS3.pdf", width = 5L, height = 4L)
+figureS3(dataset)
+dev.off()
 
 # ms.
 # knitr::knit("manuscript_suppinfo.Rnw", "manuscript_suppinfo.tex")
[…]

### 2015-11-11 manuscript.Rnw
More notes in manuscript

@@
 
-\section*{Revision notes}
-\begin{enumerate}
-\item Reduce focus on GVMs in introduction (Reviewer 2).
-\item Better linking to Poorter 2015 and Reich 2014 (editor, reviewer 2, reviewer 3, poorter pers comm).
-\item Remove LMF and LAR, and only analyze MF/MS and AF/MS.
-This avoids confusion of the terms LMF and LAR which normally include root biomass.
-We should more clearly focus on above-ground biomass distribution patterns (editor + reviewer 2).
-\item Partition variance in MF/MS directly, not the proxy variable MF/AS.
-We did that for statistical convenience, as reviewer 1 is correct in pointing out.
-But we can look at partitioning variance in MF/MS directly, to plant height, PFT and climate.
-Reviewer 3 (though incorrectly) says 'also have the same variable in the numerator and denominator, which causes statistical issues' - but then we can say we avoid this apparent problem.
-\item We claim that climate does not affect biomass distribution, but reviewer 1 correctly points out that climate affects plant height, so it may have an indirect effect.
-Our dataset is not appropriate for testing effects of climate on typical (or maximum) plant height, because individuals of all kinds of sizes were sampled; there will be sampling bias.
-Moles et al.
-2009 reported that maximum plant height shows a strong latitudinal pattern (taller plants at low latitude), and is correlated with climate variables.
-We just have to describe their results.
-\item Can follow reviewer 1's suggestion to partition MF/MS like this: $\frac{M_F}{M_S} = \frac{A_F}{A_S}\frac{M_F}{A_F}\frac{A_S}{M_S}$
-\item Reviewer 2 has some strange comments about MS/AS, in that we expect them to be correlated.
-Yes, but we need to make the point more clearly that there could be PFT effects in this correlation, largely through wood density (and tree shape).
-In fact we bring this up, but partitioning the variance as above may make this stand out more.
-\item "The methods do not describe if the individual data points came from trees in closed-canopy or open stands.  That would also make a huge difference." (Reviewer 3)
-\item We used MAT, MAP as well as the interaction.
-Probably better climate variables out there, but having the interaction in a model does distinguish between warm-dry and warm-wet climates (just like aridity indices).
-Because we found no appreciable effects of climate, we did not think it was worthwhile testing all possible climate variables.
-\end{enumerate}
-
 \section{New list of figures}
@@
 This can be explained from differences in wood density; gymnosperms have lower wood density on average.
-\item Histograms of the three main variables: AF/AS, MF/AF, and height-corrected stem mass index, MS/(AS*H).
+\item Histograms of the three main variables: AF/AS, MF/AF, and height-corrected stem mass index, MS/(AS*H). Recaps PFT differences and highlights large amount of variation yet to explain.
 \end{enumerate}
@@
 
-Much too long, of course!
 % 200 word max.
@@
 gymnosperm divide; evergreen vs.
-deciduous?).
+deciduous).
 Also useful for GVMs which use PFT for parameterization.
@@
 One major gap is that these previous studies did not include estimates of whole-plant leaf area.
-Leaf area is relevant because bla, but may follow very different patterns from leaf mass.
-Poorter2015 speculated that X.
+Leaf area is relevant because of light interception.
+It may follow very different patterns from leaf mass.
+Poorter2015 speculated that differences in leaf mass fraction are related to PFT-differences in specific leaf area, but did not have the data to directly test this prediction.
 \item What about climate?
@@
 But unclear if global patterns persist.
-Reich2014 showed X, consistent with glasshouse stuff from Poorter2015.
+Reich2014 showed X, consistent with glasshouse stuff from Poorter2015/2012.
 \item We have the following objectives:
 \begin{enumerate}
-\item Test whether MF/MS is proportional to leaf mass per area (as suggested by Poorter)
-\item Partition the variance of leaf/stem biomass distribution into component variables, and into plant height (which is important, obviously!), plant functional type and climate. 
+\item Test whether MF/MS is proportional to leaf mass per area across plant functional types (as suggested by Poorter)
+\item Partition the variance of leaf/stem biomass distribution into component variables with the objective to determine which aspects of plant allometry drive differences in biomass distributions between PFTs and with plant size.
+\item Partition variance of component variables into plant height, plant functional type and climate. 
 \end{enumerate}
-
 \end{itemize}
@@
 Consistent with variance partitioning: MF/MS mostly explained by H and PFT, but LAR not much explained by PFT.
-\item 
+\item Partitioning MF/MS into three component variables, and partition each of the four into H, PFT and climate.
+\item First one is AF/AS, which is neither related to height (and, apart from large trees, AF is proportional to AS) or PFT, and weakly to climate. 
+\item Stem mass should be proportional to basal area, height, and the product of wood density and stem shape.
+Across all data, the relationship between MS and AS was not related to PFT, but nearly proportional to height (as expected).
+For larger trees, MS at a given AS*H was clearly lower for evergreen gymnosperms, reflecting differences in wood density.
+Again no climate effects
+\item Histograms summarize the findings, for three variables that do not, or weakly, depend on height. Large PFT effects on LMA, small ones on stem mass index, none on AF/AS. 
 \end{itemize}
@@
 
-We found that PFTs differed in LMF, and that LMF was proportional to LMA.
+\begin{itemize}
+\item MF/MS differed strongly between PFTs and was proportional to leaf mass per area, and steeply declined with plant height as expected.
+\item This was driven by
+
+\end{itemize}
+
+% We found that PFTs differed in LMF, and that LMF was proportional to LMA.
 This conclusion was particularly strong across the three PFTs studied, but it also held across species within PFT.
@@
 We show that LMA -- a central trait of the leaf economics spectrum (LES) \citep{wrightetal2004} -- explains differences in LMF in a consistent manner.
-
-We found that, as expected, plant size strongly influences LMF (Fig.~\ref{fig:figlmf}) and LAR (Fig.
+% 
+% We found that, as expected, plant size strongly influences LMF (Fig.~\ref{fig:figlmf}) and LAR (Fig.
 S2) (see also \citet{poorter2012,poorter2015}).
@@
 This was useful because both LMF and LAR showed very non-linear patterns with plant height, even on a logarithmic scale, consistent with recent other results \citep{poorter2015}.
-
-The finding that LMF is proportional to LMA across PFTs implies that LAR does not differ systematically between PFTs.
+% 
+% The finding that LMF is proportional to LMA across PFTs implies that LAR does not differ systematically between PFTs.
 One explanation for this pattern lies in the strong positive correlation between LMA and leaf lifespan (LL) \citep{wrightetal2004}.
@@
 It can thus be argued that an optimal LAR exists that balances investment in supporting woody biomass (reflected by $M_T$) and the efficiency of foliage in terms of light interception.
-
-We found that patterns in $M_F/A_S$ across PFTs mirrored those of LMF (Fig.~\ref{fig:figlmf}).
+% 
+% We found that patterns in $M_F/A_S$ across PFTs mirrored those of LMF (Fig.~\ref{fig:figlmf}).
 Because LMF may be decomposed into the terms $M_F/A_S$ and $M_T/A_S$ (eq.
@@
 For tropical forests, \citet{chave2005} showed that aboveground biomass can be estimated as,
-
-$$M_T = F \rho A_S H_T$$
-
-where $F$ is a form factor, and $\rho$ wood density (kg m$^3$).
+% 
+% $$M_T = F \rho A_S H_T$$
+% 
+% where $F$ is a form factor, and $\rho$ wood density (kg m$^3$).
 After rearranging, this equation predicts that $M_T/A_S$ is proportional to plant height.
@@
 It is timely to use database made public by \citet{falster2015} (BAAD) and \citet{poorter2015} to develop more general equations to estimate individual plant aboveground biomass across the globe.
-
-We found that climate did not appreciably affect biomass partitioning between leaves and stems.
+% 
+% We found that climate did not appreciably affect biomass partitioning between leaves and stems.
 This finding is in contrast with a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}.
@@
 With the caveats that there is still variation between species and studies unaccounted for (Fig.~\ref{fig:fighist}), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Table \ref{table:statstable}).
-
-Our results indicate that it is possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.
+% 
+% Our results indicate that it is possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.
 Indeed, nearly as much variation was explained across species when using LMA instead of PFT in a statistical model.
@@
 These results suggest promising avenues for parameterising and simplifying biomass partitioning routines in GVMs.
-
-Overall, our results establish general patterns about plant construction and thus lay an empirical base against which models can be benchmarked.
+% 
+% Overall, our results establish general patterns about plant construction and thus lay an empirical base against which models can be benchmarked.
 A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions.

### 2015-11-12 manuscript.Rnw
manuscript compiles
more bullet points intro and discussion

@@
 \usepackage{palatino}
+% \usepackage[default,osfigures,scale=0.95]{opensans}
+
 \usepackage{parskip}
@@
 \usepackage{fixltx2e}
-% \usepackage[default,osfigures,scale=0.95]{opensans}
 \graphicspath{ {./figures/} }
@@
 
-\title{Leaf traits drive differences in above-ground biomass distribution among major plant functional types}
+\title{Leaf traits drive differences in above-ground biomass distribution among woody plant functional types}
 
@@
 
-\section{New list of figures}
-\begin{enumerate}
-\item Tree diagram plus climate space of dataset. Unchanged from before, except remove MT (?).
-\item MF and MS as function of tree height.
-Reviewer 2 asks "But are differences in Figure 2 statistically signficant?", weird question and subsequent comments.
-Figure could add trend lines (as smoothers), but refer to later figures for significant differences.
-The comment that sampling may have been uneven is correct, which is why we correct for plant height in the analyses.
-\item (a) MF/MS vs.
-plant height by PFT (was LMF) (b) and (c) as before, except for MF/MS.
-Reviewer 2 wants us to remove panels d and e, which we can do because we won't need MF/AS as a proxy anymore (should be put in SuppInfo instead).
-\item AF vs. AS by pft, for all data, with fitted SMA. This shows AF is proportional to AS (approx, except larger plants), and no/very very small effect of PFT.
-\item Stem mass as function of AS, PFT and height.
-a) Raw data (to link to main equation) show no clear PFT effect.
-b) MS / AS is nearly proportional to H, as expected from simple stem mass equation.
-c) for larger plants (H*AS > 0.01m3), there is a significant PFT effect, with EG having lower MS / (AS*H) (see d) for averages across entire subset shown in c).
-This can be explained from differences in wood density; gymnosperms have lower wood density on average.
-\item Histograms of the three main variables: AF/AS, MF/AF, and height-corrected stem mass index, MS/(AS*H). Recaps PFT differences and highlights large amount of variation yet to explain.
-\end{enumerate}
-
 \section*{Summary}
@@
 Also useful for GVMs which use PFT for parameterization.
-\item Enquist&Niklas showed differences in leaf mass at given stem mass between gymnosperms and angiosperms.Poorter2015 showed X.
-Reich2014 showed Y.
-Open questions are 1,2,3.
-One major gap is that these previous studies did not include estimates of whole-plant leaf area.
+\item Enquist and Niklas showed differences in leaf mass at given stem mass between gymnosperms and angiosperms.Poorter2015 showed X. Reich2014 showed Y. Open questions are 1,2,3. 
+\item One major gap is that these previous studies did not include estimates of whole-plant leaf area.
 Leaf area is relevant because of light interception.
@@
 We know from small-scale experiments and observational studies that biomass distribution varies with climate drivers.
-But unclear if global patterns persist.
+But unclear if global patterns exist.
 Reich2014 showed X, consistent with glasshouse stuff from Poorter2015/2012.
@@
 \begin{enumerate}
-\item Test whether MF/MS is proportional to leaf mass per area across plant functional types (as suggested by Poorter)
+\item Test whether MF/MS is proportional to leaf mass per area across plant functional types (as suggested by Poorter), once corrected for the expected strong effects of plant size.
 \item Partition the variance of leaf/stem biomass distribution into component variables with the objective to determine which aspects of plant allometry drive differences in biomass distributions between PFTs and with plant size.
-\item Partition variance of component variables into plant height, plant functional type and climate. 
+\item Partition variance of component variables into plant height, plant functional type and broad  climate variables. 
 \end{enumerate}
@@
 This conversion was necessary because half-total surface area is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}.
+
 Stem cross-sectional area was measured either at breast height and/or at the base of the plant.
-For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using a non-linear regression model, as follows.
-
-% we used basal area because many plants were <1.3m tall
-Using the subset of data where basal stem diameter ($A_S$) and diameter at breast height ($A_{Sbh}$) were measured (a total of \Sexpr{length(residuals(basalafit))} observations covering the three major PFTs), we developed a non-linear model to predict $A_S$ when only $A_{Sbh}$ and plant height ($H$) were measured.
-We fit the following equation, using non-linear regression.
+In all analyses, we used basal area because many plants (N=xxxx) were <1.3m tall.
+For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) by fitting the following equation, using non-linear regression.
 
@@
 The estimated coefficients were $c_0$ = \Sexpr{signif(coef(basalafit)[1],3)}, $c_1$ = \Sexpr{signif(coef(basalafit)[2],3)}, root-mean square error = \Sexpr{signif(summary(basalafit)$sigma,3)}.
+The relationship was fit using a total of \Sexpr{length(residuals(basalafit))} observations covering the three major PFTs.
 
@@
 
-All analyses were conducted in R v3.2.0 \citep{R-2015}.
-GAMs were fitted using \emph{mgcv} package \citep{wood2006}.
-The code to replicate this analysis are available on GitHub at http://github.com/RemkoDuursma/baadanalysis.
+All analyses were conducted in R 3.2.2 \citep{R-2015}.
+GAMs were fitted using the \emph{mgcv} package \citep{wood2006}.
+The code to replicate this analysis are available at http://github.com/RemkoDuursma/baadanalysis.
 
@@
 \item Partitioning MF/MS into three component variables, and partition each of the four into H, PFT and climate.
-\item First one is AF/AS, which is neither related to height (and, apart from large trees, AF is proportional to AS) or PFT, and weakly to climate. 
+Climate effects were very weak, with only LMA and AF/AS somewhat more dependent on climate, but not consistently.
+\item We show that AF was proportional to AS, apart from larger trees. Almost no PFT effects.
 \item Stem mass should be proportional to basal area, height, and the product of wood density and stem shape.
@@
 For larger trees, MS at a given AS*H was clearly lower for evergreen gymnosperms, reflecting differences in wood density.
-Again no climate effects
+Again no climate effects.
 \item Histograms summarize the findings, for three variables that do not, or weakly, depend on height. Large PFT effects on LMA, small ones on stem mass index, none on AF/AS. 
@@
 \item MF/MS differed strongly between PFTs and was proportional to leaf mass per area, and steeply declined with plant height as expected.
-\item This was driven by
-
+\item We found weak and inconsistent effects of climate on all variables studied, with some exception for LMA.
+However, climate affects plant height, so we do expect global patterns in biomass distribution (Klein 2015, Simard 2011).
+These results differ from Reich2014 but we don't know why.
+\item Not surprising that LMA was related to climate, but this did not drive differences in biomass distribution. 
+\item If biomass distribution is related to leaf mass per area, an important trait, perhaps this can help to place plants on a continuous scale rather than rely on fixed categories.
+This is an active area of research, but is yet to be extended to patterns in biomass distribution.
+\item Leaf area was just about proportional to AS, except for the largest trees.
+This decreases because of the formation of heartwood (and AF/sapwood may decrease a little with height, but AF/AS more because of crown rise).
+Note that we have many small plants in the dataset as well, and within individual datasets it is more likely to pick up small-scale effects like these.
+\item Stem mass index: Chave biomass estimation.
+We find support for this, with MS/AS nearly proportional to height.
+Conifers diverged, for the subset of larger trees only.
+Not sure why these differences don't show up for small plants but variance was very high!
+\item Some paragraph on future work.
+\item Consequences for GVMs.
 \end{itemize}

### 2015-11-13 manuscript.Rnw
Many captions written

@@
 \newif\ifincludefigs
-\includefigsfalse  % or \includefigstrue
+\includefigstrue  % or \includefigsfalse
 
@@
 
-%' 
-%' <<results='asis'>>=
-%' bigtable <- rbind(cbind(c("$M_F/M_T$","","",""),c("DA","EA","EG","Sum"), table_lmf),
-%'                   cbind(c("$A_F/M_T$","","",""),c("DA","EA","EG","Sum"), table_lar), 
-%'                   cbind(c("$M_F/A_S$","","",""),c("DA","EA","EG","Sum"), table_mlfastba),
-%'                   cbind(c("$A_F/A_S$","","",""),c("DA","EA","EG","Sum"), table_alfastba))
-%' rownames(bigtable) <- NULL
-%' colnames(bigtable)[1:2] <- c("Variable","PFT")
-%' 
-%' print(xtable(bigtable, align="llrlll|l",
-%'                caption="Sample sizes used in the analyses of the four variables considered in the study by plant functional type (PFT, DA = deciduous angiosperm, EA = evergreen angiosperm, EG = evergreen gymnosperm) and biome.
-Sample sizes denote number of individuals, with number of unique species in parentheses.
-",
-%'              label="table:counttable"), 
-%'       hline.after=0:16, caption.placement="top",  #c(0,4,8,12,16)
-%'       sanitize.text.function=function(x){x},
-%'       include.rownames=FALSE)
-%' @
-%' 
-%' 
-%' 
-%' <<results='asis'>>=
-%' tab1 <- table_varpart2
-%' tab2 <- table_gamr2MATMAP
-%' 
-%' tidytab <- function(x, labeltopleft=""){
-%'   names(x)[1] <- "Variable"
-%'   x[,1] <- paste0(x[,1],"(",x[,2],")")
-%'   x[,2] <- NULL
-%'   x[,2:ncol(x)] <- round(x[,2:ncol(x)], 2)  
-%'   x <- unname(rbind(names(x),as.matrix(x)))  
-%'   x <- rbind(c(labeltopleft,"","Predictors",""),x)
-%' return(x)
-%' }
-%' 
-%' tab1 <- tidytab(tab1, "Mixed-effects model")
-%' tab2 <- tidytab(tab2, "GAM")
-%' 
-%' tab <- rbind(tab1, tab2)
-%' 
-%' print(xtable(tab, align="lrrrr",
-%'              caption="Explained variance by plant functional type (PFT), plant height (H) and climate variables (either biome (B), or MAP and MAT) for five whole-plant variables.
-For the analysis with biome, each variable was added to a linear mixed-effects model, using species within dataset as a random effect.
-The $R^2$ shown is that explained by the fixed effects only.
-All fixed effects were highly significant (P \\textless 0.001).
-In addition to the fixed effects shown, all interactions were added to each of the models.
-For the analysis with continuous climate variables MAP and MAT, each variable was added as a smooth term in a generalized additive model (GAM), with the exception of PFT (a categorical variable).
-Note that explained variance for the linear mixed-effects model and GAM is not necessarily the same even with the same predictors, due to different methods for fitting and estimating the explained variance.",
-%'              label="table:statstable"),
-%'              hline.after=0:14, #c(0,1,7,8,14),
-%'              include.rownames=FALSE, 
-%'              include.colnames=FALSE, 
-%'              caption.placement="top", sanitize.text.function=function(x){x})
-%' @
-%'
+\clearpage
+\section{Figures}
 
-%' 
-%' \clearpage
-%' \section{Figures}
-%' 
-%' 
-%' \begin{figure}[h!]
-%'     \centering
-%'     \ifincludefigs 
-%'     \includegraphics[width=0.99\textwidth]{Figure1.pdf}
-%'     \fi
-%'     \caption{Overview of the data.
+\begin{figure}[h!]
+    \centering
+    \ifincludefigs 
+    \includegraphics[width=0.99\textwidth]{Figure1.pdf}
+    \fi
+    \caption{Overview of the data.
 (a) Variables were measured on up to \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species.
@@
 Colour symbols show the locations of sampled individuals for three dominant woody functional types.}
-%'     \label{fig:figoverview}
-%' \end{figure}
-%' 
-%' 
-%' \begin{figure}[h!]
-%'     \centering
-%'     \ifincludefigs 
-%'     \includegraphics[width=0.99\textwidth]{Figure2.pdf}
-%'     \fi
-%'     \caption{Raw data for leaf biomass ($M_F$) and total above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$).
+    \label{fig:figoverview}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \ifincludefigs 
+    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
+    \fi
+    \caption{Raw data for leaf biomass ($M_F$) and above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$).
 Each point is an individual plant.
-Sample sizes are listed in Table \ref{table:counttable}}
-%'     \label{fig:figrawdata}
-%' \end{figure}
-%' 
-%' \begin{figure}[h!]
-%'     \centering
-%'     \ifincludefigs 
-%'     \includegraphics[width=0.99\textwidth]{Figure3.pdf}
-%'     \fi
-%'     \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf mass per area.
-(a) Leaf mass fraction ($M_F/M_T$ = leaf mass / above-ground biomass) by PFT.
+Sample sizes are listed in Table \ref{table:counttable}.
+Lines are generalized additive model fits to visualize the highly non-linear relationships.}
+    \label{fig:figrawdata}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \ifincludefigs 
+    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
+    \fi
+    \caption{Dominant woody PFTs differ in above-ground biomass distribution due to underlying differences in leaf mass per area.
+(a) Leaf mass per unit woody above-ground biomass ($M_F/M_S$) by PFT.
 Each symbol is an individual plant.
 Lines are generalised additive model fits.
-(b) and (c) Leaf mass fraction and leaf area ratio ($A_F/M_T$) at the average plant height in the dataset, estimated from panel (a).
-(d) Average leaf mass per unit basal stem area, and (e) leaf area per unit basal stem area for the three major PFTs confirm that the between-PFT variation in leaf mass fraction is due to leaf mass per unit basal area.
+(b) and (c) $M_F/M_S$ and leaf area ratio ($A_F/M_S$) at the average plant height in the dataset, estimated from panel (a).
 Error bars are 95\% confidence intervals.
 Letters denote significant differences (at $\alpha = 0.05$).}
-%'     \label{fig:figlmf}
-%' \end{figure}
-%' 
-%' \begin{figure}[h!]
-%'     \centering
-%'     \ifincludefigs 
-%'     \includegraphics[width=0.99\textwidth]{Figure4.pdf}
-%'     \fi
-%'     \caption{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.
-Shown are histograms (as probability density functions) of leaf mass per area ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three PFTs.
+    \label{fig:figlmf}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \ifincludefigs 
+    \includegraphics[width=0.99\textwidth]{Figure4.pdf}
+    \fi
+    \caption{Total plant leaf area ($A[L]$) is approximately proportional to basal stem area ($A_S$), and this relationship is similar between PFTs.
+Lines are standardized major axis fits.
+Coefficients are shown on the figure (X), only evergreen gymnosperms showed a slightly more than proportional slope (1.05).}
+    \label{fig:alas}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \ifincludefigs 
+    \includegraphics[width=0.99\textwidth]{Figure5.pdf}
+    \fi
+    \caption{Relationships between above-ground woody biomass ($M_S$), basal stem area ($A_S$) and plant height (H).
+(a) $M_S$ as a function of $A_S$ for the entire dataset.
+Lines are generalized additive model fits.
+(b) Much additional variation in M_S can be explained by H.
+Shown is the ratio $M_S/A_S$ as a function of H, with generalized additive model fits.
+(c) Stem mass as a function of $A_S*H$ for a subset of the data in panels a and b, where the product of H and $A_S$ was larger than 0.01.
+For this subset of larger trees, stem mass was nearly proportional to $A_S*H$ (SMA fits).
+(d) From the data in panel c, the average ratio $M_S/(A_S*H$) differs between plant functional types, with a lower value for gymnosperms compared to angiosperms.}
+    \label{fig:stemmassindex}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \ifincludefigs 
+    \includegraphics[width=0.99\textwidth]{Figure6.pdf}
+    \fi
+    \caption{Histograms (as probability density functions) of leaf mass per area ($M_F/A_F$), leaf area per unit basal stem area ($A_F/A_S$) and the stem mass index ($M_S/(A_S*H)$) grouped by the three PFTs.
 Arrows indicate means by PFT.
@@
 \ref{fig:figlmf}.}
-%'     \label{fig:fighist}
-%' \end{figure}
-%' 
-%' 
-%' <<echo=FALSE>>=
-%' agg <- summaryBy(lalf_astba2 + lmlf_astba2 + llma ~ Group,
-%'                    data=dataset, FUN=mean, na.rm=TRUE, keep.names=TRUE,
-%'                    id=~pft)
-%' agg <- subset(agg, !is.na(llma))
-%' lm1 <- lm(lmlf_astba2 ~ llma * pft, data=agg)
-%' lm2 <- lm(lmlf_astba2 ~ lalf_astba2 * pft, data=agg)
-%' r2panela <- summary(lm1)$adj.r.squared
-%' r2panelb <- summary(lm2)$adj.r.squared
-%' @
-%' 
-%' \begin{figure}[h!]
-%'     \centering
-%'     \ifincludefigs 
-%'     \includegraphics[width=0.99\textwidth]{Figure5.pdf}
-%'     \fi
-%'      \caption{Leaf and stem traits capture variation in biomass partitioning across species.
-Individual plant data were averaged by species and study combinations.
-Lines are linear regressions with 95\% confidence intervals.
-Both regressions were significant ($P \textless 0.01$).
-$R^2$ for fitted relationships are \Sexpr{round(100*r2panela,1)}\% in panel (a) and \Sexpr{round(100*r2panelb,1)}\% in panel (b) (for a linear regression model including PFT as a factor variable).}
-%'   \label{fig:figlmfspecies}
-%' \end{figure}
-%' 
-%' 
-%' \begin{figure}[h!]
-%'     \centering
-%'     \ifincludefigs 
-%'     \includegraphics[width=0.99\textwidth]{Figure6.pdf}
-%'     \fi
-%'     \caption{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning.
-a) Leaf mass per unit stem cross-sectional area ($M_F/A_S$) for each PFT within three different biomes.
-DA: deciduous angiosperms, EA: evergreen angiosperms, EG: evergreen gymnosperms.
-Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect.
-Letters denote significant differences (at $\alpha = 0.05$).
-b-c) Relationships between climate and $M_F/A_S$ within each PFT.
-Solid lines are generalised additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
-%'     \label{fig:figclimate}
-%' \end{figure}
-%' 
-%'
+    \label{fig:fighist}
+\end{figure}
 

### 2015-11-13 manuscript_suppinfo.Rnw
Renumbered all figures in sequence; also thanks grid for breaking backwards compatibility in ellipseGrob; grid added to required packages
Many captions written

@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureS3.pdf}
+    \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
     \caption{Leaf area ratio ($A_F/M_T$) by plant height (H) and PFT.
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
-    \caption{Relationship between above-ground biomass and basal stem area.
-Each point is an individual plant.
-Lines are GAM fits by PFT.
-(a) Scaling of aboveground biomass ($M_T$) with basal stem area ($A_S$).
-(b) The ratio of $M_T/A_S$ as a function of plant height shows no clear differences between PFTs.
-The solid black line is a line with a slope of 1, which is expected if the aboveground biomass equation from \citet{chave2005} hold across the dataset.
-(c) Least-square means of the ratio of above-ground biomass to basal stem area, estimated from a linear mixed-effects model with basal stem area (linear and squared terms) and PFT.
-}
+    \includegraphics[width=0.99\textwidth]{FigureS3.pdf}
+    \caption{\emph{May not go in manuscript, just for notes at the moment}.
+Estimation of basal stem diameter from diameter at breast height for small plants.
+Because 36\% of the dataset is for plants with a height less than 1.3m, we used basal stem area in all analyses.
+For a subset of the data, dbh was measured but not d.ba, even though the plants were only marginally taller than breast height.
+In this figure, black points are the measured ratio of d.ba/d.bh (for plants where both were measured), red the predicted ratio from the fitted nonlinear model.
+Below 1.8m, the variance of the ratio d.ba/d.bh increases drastically, so we excluded those plants from the analysis (unless d.ba was actually measured).}
     \label{fig:figureS4}

### 2015-11-13 R/figures.R

@@ -88,8 +88,8 @@ fig.tree <- function(filename) {
   grid.text(expression(paste(M[S],", Mass of stem")),
     x = x0 + 0.03 , y = 0.18, just="left", gp=gp0)
 
-  grid.text(expression(paste(M[T],", total aboveground mass (", M[F]+M[S],")")),
-    x = x0 - 0.03 , y = -0.1, just="left", gp=gp0)
+#   grid.text(expression(paste(M[T],", total aboveground mass (", M[F]+M[S],")")),
+#     x = x0 - 0.03 , y = -0.1, just="left", gp=gp0)
 
   # leaf
   x1 <- x0 + 0.25
@@ -173,8 +173,8 @@ figure2 <- function(dataset){
                    axes=FALSE))
     with(dat, points(lh.t, lmlf, pch=16,cex=0.5,
                      col=alpha("forestgreen",0.5)))
-    smoothplot(lh.t, lmlf, data=dat, linecol="green3", add=TRUE, plotpoints=FALSE)
-    smoothplot(lh.t, lmst, data=dat, linecol="darkgoldenrod", add=TRUE, plotpoints=FALSE)
+    smoothplot(lh.t, lmlf, data=dat, linecol="green3", R="Group", add=TRUE, plotpoints=FALSE)
+    smoothplot(lh.t, lmst, data=dat, linecol="darkgoldenrod", R="Group", add=TRUE, plotpoints=FALSE)
     
     log10axes(side=1,  labels=TRUE)
     log10axes(side=2,  labels= i == 1)
@@ -271,13 +271,13 @@ figure5 <- function(dataset){
   smoothplot(log10(h.t), log10(m.st / a.stba2), pft, data=dataset, 
              pointcols=my_cols_transparent(),
              linecols=my_cols(),
-             xlab=expression(H[T]~~(m)),
+             xlab=expression(H~~(m)),
              ylab=expression(M[S]/A[S]~~(kg~m^-2)),
              pch=16)
   box()
   plotlabel("(b)", "topleft")
   with(subs, plot(log10(a.stba2 * h.t), log10(m.st), pch=16, 
-                  xlab=expression(A[S]*H[T]~~(m^3)),
+                  xlab=expression(A[S]*H~~(m^3)),
                   ylab=expression(M[S]~~(kg)),
                   axes=FALSE,
[…]

### 2015-11-15 manuscript.Rnw
ms and suppinfo including figures and captions

@@
 \item We use the recently published Biomass And Allometry Database (BAAD), a large database including observations for individual plants.
-The database covers the global climate space, allowing us to test for direct climate effects in addition to PFT.
+The database covers the global climate space, allowing us to test for direct climate effects in addition to PFT and plant height.
 \item The ratio of leaf mass to woody above-ground mass (MF/AS) varied strongly between PFTs (as defined by deciduous vs.
@@
 PFT explained much variation in MF/MS and MF/AF (but not AF/AS and only for larger plants in MS/AS).
-\item Climate variables (MAT, MAP and its interaction) consistently explained very little additional variation, demonstrating very weak control of climate on biomass distribution and leaf area.
-We argue that climate effects will largely be indirectly through plant height.
+\item Climate variables (MAT, MAP and its interaction) consistently explained very little additional variation, demonstrating very weak control of climate on biomass distribution and leaf area, at a given plant size.
+We argue that climate effects will largely be indirectly through effects of plant height.
 \item Combined, these results show consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain GVMs.
@@
 \item Partitioning MF/MS into three component variables, and partition each of the four into H, PFT and climate.
-Climate effects were very weak, with only LMA and AF/AS somewhat more dependent on climate, but not consistently.
-\item We show that AF was proportional to AS, apart from larger trees. Almost no PFT effects.
+Climate effects were very weak, with only LMA and AF/AS somewhat more dependent on climate, but not consistently (Fig S4).
+\item We show that AF was proportional to AS (with the exception of very large trees). AF/AS was nearly unrelated to PFT or plant height.
 \item Stem mass should be proportional to basal area, height, and the product of wood density and stem shape.
@@
 These results differ from Reich2014 but we don't know why.
-\item Not surprising that LMA was related to climate, but this did not drive differences in biomass distribution. 
 \item If biomass distribution is related to leaf mass per area, an important trait, perhaps this can help to place plants on a continuous scale rather than rely on fixed categories.
 This is an active area of research, but is yet to be extended to patterns in biomass distribution.
+\item Still lots and lots of variation to explain within PFTs; we do not argue for convergence, but rather that as a first step, we can understand differences between PFTs.
 \item Leaf area was just about proportional to AS, except for the largest trees.
@@
 Not sure why these differences don't show up for small plants but variance was very high!
-\item Some paragraph on future work.
+\item Some paragraph on future work, for example reconcile Poorter2015 with Reich2014 with our data.
 \item Consequences for GVMs.
@@
 Lines are generalized additive model fits.
-(b) Much additional variation in M_S can be explained by H.
+(b) Much additional variation in $M_S$ can be explained by H.
 Shown is the ratio $M_S/A_S$ as a function of H, with generalized additive model fits.
@@
 For this subset of larger trees, stem mass was nearly proportional to $A_S*H$ (SMA fits).
-(d) From the data in panel c, the average ratio $M_S/(A_S*H$) differs between plant functional types, with a lower value for gymnosperms compared to angiosperms.}
+(d) From the data in panel c, the average ratio $M_S/(A_S*H)$ differs between plant functional types, with a lower value for gymnosperms compared to angiosperms.}
     \label{fig:stemmassindex}

### 2015-11-15 manuscript_suppinfo.Rnw
ms and suppinfo including figures and captions

@@
 Shaded areas are approximate 95\% confidence intervals.}
-    \label{fig:figureS3}
+    \label{fig:figureS2}
 \end{figure}
@@
 Below 1.8m, the variance of the ratio d.ba/d.bh increases drastically, so we excluded those plants from the analysis (unless d.ba was actually measured).}
+    \label{fig:figureS3}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
+    \caption{Partial regression plots for leaf area per stem basal area (AF/AS) and leaf mass per area (MF/AF).
+These two variables showed the strongest dependency on climate of all variables studied (see Table).
+Results are for a linear mixed-effects model (using species-dataset as a random effect), with plant height (and squared height), MAT, MAP (and squared terms), PFT, and interactions between MAP, MAT and PFT.
+Non-significant terms were removed (for full model, see SI text - \emph{to be added}).
+Partial regression plots were constructed with the visreg package in R.
+Each point is a model prediction when all other terms in the model are held at their average value.
+Transparent areas are approximate 95\% confidence intervals.}
     \label{fig:figureS4}
@@
 
-\clearpage
-\subsection*{Notes S1. Modelling of biomass partitioning in global vegetation models (GVMs)}
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
+    \caption{Comparison of three methods for variance partitioning of the five studied variables (see legend) by total plant height, plant functional type and climate (MAP, MAT and interaction).
+\textbf{GAM}: variance component as estimated with a generalized additive model, as the reduction in R2 when the independent variable (i.e.
+PFT, climate or height) was excluded from a model.
+Note that a GAM does not include interactions.
+\textbf{LMM}: linear mixed model with all interactions, and squared terms for the continuous variables, where the variance component was again estimated as the reduction in R2 when the independent variable (and all its interactions) were removed from the model.
+R2 was calculated following Nakagawa and Schielzeth (2013) using the MuMIn R package (excluding random effects).
+Note that LMM and GAM give very similar ranking, and even absolute variance components, for nearly all variables.
+Both methods identified AF/AS and MF/AF as the only variables with a practically significant dependence on climate (variance component > 0.1), see Fig S4.
+\textbf{IEA}: independent effects analysis (aka hierarchical partitioning), using the algorithm by Chevan and Sutherland (1991) as implemented in the hier.part R package.
+Note that this method does not include interactions or squared terms, yet gave similar results to GAM and LMM (except for variables that explained little variation).}
+    \label{fig:figureS5}
+\end{figure}
 
-Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types (PFTs), although some algorithms make no mention of these potential differences \citep{friedlingstein1999, chen2013}.
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureS6.pdf}
+    \caption{\emph{Should come earlier}.
+A reviewer asked to see MF and AF as a function of total above-ground biomass.
+This also shows that MF at given MT diverges between PFTs (except for very small plants), but largely overlaps for AF/MT.
+Exception is for very (very) large trees, but we only have two datasets there.}
+    \label{fig:figureS6}
+\end{figure}
+
+% 
+% \clearpage
+% \subsection*{Notes S1. Modelling of biomass partitioning in global vegetation models (GVMs)}
+% 
+% 
+% Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types (PFTs), although some algorithms make no mention of these potential differences \citep{friedlingstein1999, chen2013}.
 However, there is no consensus on how these differences are parameterized, or whether they are linked to inherent differences in leaf mass per area (LMA) or leaf lifespan (LL).
@@
 Finally, we briefly summarize how current GVMs include effects of climate on biomass partitioning.
-
-\subsubsection*{Constant allocation fractions}
-
-Some models assume that a constant fraction of net primary production (NPP) is allocated to each component (leaves, stems, roots).
+% 
+% \subsubsection*{Constant allocation fractions}
+% 
+% Some models assume that a constant fraction of net primary production (NPP) is allocated to each component (leaves, stems, roots).
 CABLE, EALCO \citep{dekauwe2014} and IBIS \citep{foley1996} are examples of models that allocate a PFT-specific fraction of NPP to each plant component each timestep  (though, for both models, alternative model formulations have been used as well).
 Constant allocation fractions still give rise to differences in biomass partitioning between PFTs (see Methods in main text), since turnover of leaves (determined by LL) will always differ between at least deciduous and evergreen PFTs.
-
-\subsubsection*{Allometric constraints}
-
-In this class of models,  allometric (or other size-dependent) relationships between biomass components are used to calculate biomass allocation in each timestep.
+% 
+% 
+% \subsubsection*{Allometric constraints}
+% 
+% In this class of models,  allometric (or other size-dependent) relationships between biomass components are used to calculate biomass allocation in each timestep.
 For example, in ED2 \citep{medvigy2009}, biomass partitioning to leaves, stem sapwood and roots is constrained by a set of equations with PFT-specific parameters, estimated through model calibration.
@@
 Further differences in biomass partitioning between PFTs are then the result of differences in LMA (leaf mass at a given leaf area), as well as LL (shorter lived leaves require higher annual allocation of biomass for upkeep).
-
-\subsubsection*{Optimality}
-
-Few GVMs use a scheme in which carbon allocation is determined by assuming biomass fractions are the result of optimization, however, this scheme is widely used in smaller-scale vegetation models \citep{valentinemakela2012,franklin2012}.
+% 
+% \subsubsection*{Optimality}
+% 
+% Few GVMs use a scheme in which carbon allocation is determined by assuming biomass fractions are the result of optimization, however, this scheme is widely used in smaller-scale vegetation models \citep{valentinemakela2012,franklin2012}.
 One example of a global model is SDGVM \citep{woodward1995}, which optimizes leaf area index by considering the carbon balance of the lowest leaf layer in the canopy.
 As with LPJ described above, differences in biomass partitioning between PFTs then arise as a result of differences in LMA as well as LL.
-
-\subsubsection*{Climate effects}
-
-Models differ widely in how climate directly (or indirectly) affects biomass partitioning beyond differences between PFTs.
+% 
+% \subsubsection*{Climate effects}
+% 
+% Models differ widely in how climate directly (or indirectly) affects biomass partitioning beyond differences between PFTs.
 In one class of models, biomass partitioning is modelled based on relative resource limitations, consistent with the idea that biomass allocation proceeds to the plant component most limited by the resource it acquires \citep{friedlingstein1999}.
@@
 It is clear that these types of models predict apparent climate effects on biomass partitioning along temperature and moisture gradients.
-
-Some models do not assume effects of stress on biomass allocation directly, but differences in allocation may result from effects of stress on biomass production, and subsequent priority rules for the allocation of biomass.
+% 
+% Some models do not assume effects of stress on biomass allocation directly, but differences in allocation may result from effects of stress on biomass production, and subsequent priority rules for the allocation of biomass.
 For example, in LPJ \citep{sitchetal2003}, biomass production may be low enough in very dry years so that intended allocation to leaves and stem sapwood cannot be supported.
@@
 In dry climates, where water stress will be more common, this would gradually lead to different partitioning of biomass to leaves and stems.
-
-\bibliography{references}
+% 
+% 
+% \bibliography{references}
 

### 2015-11-15 run.R

@@ -149,7 +149,7 @@ dev.off()
 combine <- TRUE
 if(combine){
   figs <- paste(sprintf("%sFigure%i.pdf","figures/", 1:6), collapse=" ")
-  SIfigs <- paste(sprintf("%sFigureS%i.pdf", "figures/", 1:5), collapse=" ")
+  SIfigs <- paste(sprintf("%sFigureS%i.pdf", "figures/", 1:6), collapse=" ")
   cmd <- sprintf("pdftk %s %s cat output Figures%s.pdf",figs,SIfigs,
                  format(as.Date(Sys.time()),"%Y-%m-%d"))
   system(cmd)
@@ -161,7 +161,7 @@ knitr::knit("manuscript.Rnw", "manuscript.tex")
 
 # ms SuppInfo
 tex_2_pdf("manuscript.tex")
-# tex_2_pdf("manuscript_suppinfo.tex")
+tex_2_pdf("manuscript_suppinfo.tex")
 
 
 

### 2016-01-07 manuscript.Rnw
Figure 5 simplified
New Fig 7 with Reich-style climate effects on MF/MS

@@
 
-\title{Leaf traits drive differences in above-ground biomass distribution among woody plant functional types}
+\title{Leaf traits and size drive differences in above-ground biomass distribution among woody plant functional types}
 
@@
 \item Climate variables (MAT, MAP and its interaction) consistently explained very little additional variation, demonstrating very weak control of climate on biomass distribution and leaf area, at a given plant size.
-We argue that climate effects will largely be indirectly through effects of plant height.
+We argue that climate effects will largely be indirectly through effects of plant height, and climate-related shifts in plant functional types.
 \item Combined, these results show consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain GVMs.
@@
 \begin{itemize}
-\item Dataset covers climate space pretty well. 
+\item Dataset covers climate space pretty well. Evergreen angiosperms occurred at higher MAT than deciduous angiosperms and evergreen gymnosperms but there is much overlap between PFTs. 
 \item MF and MS are non-linearly related to plant height (Fig 2), as expected. Also note stem mass increases with height more quickly than leaf mass.
@@
 \item Partitioning MF/MS into three component variables, and partition each of the four into H, PFT and climate.
-Climate effects were very weak, with only LMA and AF/AS somewhat more dependent on climate, but not consistently (Fig S4).
+Climate effects, when analyzed in addition to H and PFT, were very weak, with only LMA and AF/AS somewhat more dependent on climate, but not consistently (Fig S4).
 \item We show that AF was proportional to AS (with the exception of very large trees). AF/AS was nearly unrelated to PFT or plant height.
@@
 \item Histograms summarize the findings, for three variables that do not, or weakly, depend on height. Large PFT effects on LMA, small ones on stem mass index, none on AF/AS. 
+\item We repeat an analysis similar to Reich et al 2014 to test for climate effects on biomass partitioning when only considering the angiosperm/gymnosperm divide, and not first correcting for H as we did previously (but instead binning by H).
+We find that for angiosperms, MF/MS increased with MAT for all but the smallest plants.
+But not for gymnosperms (Fig.
+7).
 \end{itemize}
@@
 Not sure why these differences don't show up for small plants but variance was very high!
+\item Our analysis of climate effects on above-ground biomass distribution between leaves and stems is largely consistent with Reich et al.
+2014.
+We found that for angiosperms, MF/MS increased with MAT (as did Reich), after splitting the dataset into five equally populated bins.
+This result is in apparent contradiction with our finding that climate (MAT, MAP and its interaction) explained consistently very little variation in MF/MS.
+In that analysis, however, the effects of climate were analyzed in addition to the potentially confounding factors plant height and plant functional type.
+The significant relationship between MF/MS and MAT for angiosperms can be understood by the consistently higher proportion of evergreen angiosperms at higher MAT (Fig.
+1), which, as a resutlt of higher LMA (Fig.
+2) also have a higher MF/MS.
+The main conclusions from the climate analysis are certainly that PFT and plant height are by far the most important factors, this is also consistent with Reich 2014 and Poorter 2015.
+
 \item Some paragraph on future work, for example reconcile Poorter2015 with Reich2014 with our data.

### 2016-01-07 R/figures.R

@@ -230,7 +230,7 @@ figure3 <- function(dataset, KGAM=4){
 }
 
 
-# new figure 4
+# LA vs. AS
 figure4 <- function(dataset){
   
   # random reorder dataset to avoid plotting artefact
@@ -248,7 +248,7 @@ figure4 <- function(dataset){
 }
 
 
-
+# MST vs. AS
 figure5 <- function(dataset){
   
   par(mfrow=c(1,2), mar=c(4,4,0.5,0.5), mgp=c(2,0.5,0), tcl=0.1)
@@ -273,12 +273,6 @@ figure5 <- function(dataset){
   plotlabel("(b)", "topleft")
   
 }
-
-
-
-  
-  
-  
 
 
 # Histograms of MF/AS, AF/AS, and MS/(AS*H)
@@ -318,8 +312,45 @@ figure6 <- function(dataset, nbin=100){
   
 }
 
+# new climate figure, cf. Reich
 
+figure7 <- function(dataset){
 
+  dataset$lhtclass <- quantcut(dataset$lh.t, 5)
+  dataset$pft2 <- ifelse(dataset$pft == "EG", "Gymnosperm", "Angiosperm")
+  
+  # Like Reich, no MAP
+  lv <- levels(dataset$lhtclass)
+  lv <- gsub("]",")",lv)
+  lv <- gsub("\\[","(",lv)
+  lv <- paste0("c",lv)
+  labs <- c()
+  for(i in 1:length(lv)){
+    x <- 10^eval(parse(text=lv[i]))
+    labs[i] <- sprintf("%.2f - %.2f", x[1], x[2])
+  }
+  labs[1] <- paste("H =",labs[1])
+  
+  par(mfrow=c(1,5), mar=c(0,0,0,0), oma=c(5,5,4,2))
+  lmes <- lapply(data_ht, function(x){
+    lme(lmlf_mst ~ pft2*MAT, random=~1|Group, data=x, na.action=na.omit)
+  })
+  for(i in 1:length(lmes)){
+    visreg(lmes[[i]], "MAT", by="pft2", ylim=c(-2,0.5), 
+           xlim=c(0,30), overlay=TRUE, legend=FALSE, axes=FALSE,
+           line.par=list(col=c("red","blue")),
+           ylim=c(-2,1.2))
+    axis(1)
+    magaxis(2, labels = i == 1, unlog=2)
+  }
[…]

### 2016-01-07 manuscript_suppinfo.Rnw
Removed fig S4; renumbered remaining figures

@@
     \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
-    \caption{Partial regression plots for leaf area per stem basal area (AF/AS) and leaf mass per area (MF/AF).
-These two variables showed the strongest dependency on climate of all variables studied (see Table).
-Results are for a linear mixed-effects model (using species-dataset as a random effect), with plant height (and squared height), MAT, MAP (and squared terms), PFT, and interactions between MAP, MAT and PFT.
-Non-significant terms were removed (for full model, see SI text - \emph{to be added}).
-Partial regression plots were constructed with the visreg package in R.
-Each point is a model prediction when all other terms in the model are held at their average value.
-Transparent areas are approximate 95\% confidence intervals.}
-    \label{fig:figureS4}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
     \caption{Comparison of three methods for variance partitioning of the five studied variables (see legend) by total plant height, plant functional type and climate (MAP, MAT and interaction).
@@
 Note that this method does not include interactions or squared terms, yet gave similar results to GAM and LMM (except for variables that explained little variation).}
-    \label{fig:figureS5}
+    \label{fig:figureS4}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureS6.pdf}
+    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
     \caption{\emph{Should come earlier}.
@@
 Exception is for very (very) large trees, but we only have two datasets there.}
-    \label{fig:figureS6}
+    \label{fig:figureS5}
 \end{figure}
 
-% 
-% \clearpage
-% \subsection*{Notes S1. Modelling of biomass partitioning in global vegetation models (GVMs)}
-% 
-% 
-% Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types (PFTs), although some algorithms make no mention of these potential differences \citep{friedlingstein1999, chen2013}.
-However, there is no consensus on how these differences are parameterized, or whether they are linked to inherent differences in leaf mass per area (LMA) or leaf lifespan (LL).
-Typically, parameters related to biomass allocation differ between PFTs (often based on little empirical evidence), which then lead to differences in biomass partitioning.
-We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them.
-We base this classification on a recent model intercomparison study \citep{dekauwe2014}.
-Note that we use 'partitioning' to indicate the actual distribution of biomass among plant components, and 'allocation' the fraction of net primary production to each component (see Methods in main text).
-Finally, we briefly summarize how current GVMs include effects of climate on biomass partitioning.
-% 
-% \subsubsection*{Constant allocation fractions}
-% 
-% Some models assume that a constant fraction of net primary production (NPP) is allocated to each component (leaves, stems, roots).
-CABLE, EALCO \citep{dekauwe2014} and IBIS \citep{foley1996} are examples of models that allocate a PFT-specific fraction of NPP to each plant component each timestep  (though, for both models, alternative model formulations have been used as well).
-Constant allocation fractions still give rise to differences in biomass partitioning between PFTs (see Methods in main text), since turnover of leaves (determined by LL) will always differ between at least deciduous and evergreen PFTs.
-% 
-% 
-% \subsubsection*{Allometric constraints}
-% 
-% In this class of models,  allometric (or other size-dependent) relationships between biomass components are used to calculate biomass allocation in each timestep.
-For example, in ED2 \citep{medvigy2009}, biomass partitioning to leaves, stem sapwood and roots is constrained by a set of equations with PFT-specific parameters, estimated through model calibration.
-Biomass allocation is then calculated at each timestep to satisfy these equations.
-Other models use leaf area as a constraint.
-For example, LPJ \citep{sitchetal2003} uses PFT-specific values of leaf area to sapwood area ratio to constrain estimates of biomass allocation.
-Further differences in biomass partitioning between PFTs are then the result of differences in LMA (leaf mass at a given leaf area), as well as LL (shorter lived leaves require higher annual allocation of biomass for upkeep).
-% 
-% \subsubsection*{Optimality}
-% 
-% Few GVMs use a scheme in which carbon allocation is determined by assuming biomass fractions are the result of optimization, however, this scheme is widely used in smaller-scale vegetation models \citep{valentinemakela2012,franklin2012}.
-One example of a global model is SDGVM \citep{woodward1995}, which optimizes leaf area index by considering the carbon balance of the lowest leaf layer in the canopy.
-As with LPJ described above, differences in biomass partitioning between PFTs then arise as a result of differences in LMA as well as LL.
-% 
-% \subsubsection*{Climate effects}
-% 
-% Models differ widely in how climate directly (or indirectly) affects biomass partitioning beyond differences between PFTs.
-In one class of models, biomass partitioning is modelled based on relative resource limitations, consistent with the idea that biomass allocation proceeds to the plant component most limited by the resource it acquires \citep{friedlingstein1999}.
-For example, in the allocation model of Arora and Boer \citep{aroraboer2005}, which is used in the ISAM model \citep{dekauwe2014}, allocation to leaves, stems and roots depends on relative limitations of light and water.
-Likewise, Friedlingstein et al.
-\citep{friedlingstein1999} developed a resource-limitation framework for the CASA model, where allocation depends on limitations of light, water and nutrients.
-It is clear that these types of models predict apparent climate effects on biomass partitioning along temperature and moisture gradients.
-% 
-% Some models do not assume effects of stress on biomass allocation directly, but differences in allocation may result from effects of stress on biomass production, and subsequent priority rules for the allocation of biomass.
-For example, in LPJ \citep{sitchetal2003}, biomass production may be low enough in very dry years so that intended allocation to leaves and stem sapwood cannot be supported.
-In this case, allocation proceeds to leaves only.
-In dry climates, where water stress will be more common, this would gradually lead to different partitioning of biomass to leaves and stems.
-% 
-% 
-% \bibliography{references}
-
 \end{document}

### 2016-01-07 R/figures.R

@@ -458,75 +458,9 @@ figureS3 <- function(dataset, basalafit){
 }
 
 
-figureS4 <- function(dataset){
-  
-  
-  m_afas <- lme(lalf_astba2 ~ pft + lh.t + MAP + MAT + I(MAP^2) + I(MAT^2) + MAP:pft,
-                random= ~1|Group,
-                data=dataset, na.action=na.omit)
-  
-  m_lma <- lme(llma ~ pft + lh.t + MAP + MAT + I(MAP^2) + I(MAT^2) + MAP:pft + MAT:pft,
-               random= ~1|Group,
-               data=dataset, na.action=na.omit)
-  
-  par(mfrow=c(2,2), mar=c(4.5,4.5,0.5,0.5), mgp=c(2.5,1,0))
-  visreg(m_afas, "MAP", by="pft", overlay=TRUE, 
-         legend=FALSE,
-         axes=FALSE,
-         xlab="MAP (mm)",
-         ylab=expression(f(MAP)~"-"~A[F]/A[S]~(m^2~m^-2)),
-         line.par=list(col=my_cols()),
-         fill.par=list(col=my_cols_transparent()),
-         points.par=list(col=my_cols()))
-  log10axes(2)
-  axis(1)
-  box()
-  plotlabel("(a)","topleft")
-  visreg(m_afas, "MAT", by="pft", overlay=TRUE, 
-         legend=FALSE,
-         axes=FALSE,
-         xlab=expression(MAT~~(degree*C)),
-         ylab=expression(f(MAT)~"-"~A[F]/A[S]~(m^2~m^-2)),
-         line.par=list(col=my_cols()),
-         fill.par=list(col=my_cols_transparent()),
-         points.par=list(col=my_cols()))
-  log10axes(2)
-  axis(1)
-  box()
-  plotlabel("(b)","topleft")
-  
-  visreg(m_lma, "MAP", by="pft", overlay=TRUE, 
-         legend=FALSE,
-         axes=FALSE,
-         xlab="MAP (mm)",
-         ylab=expression(f(MAP)~"-"~M[F]/A[F]~(kg~m^-2)),
-         line.par=list(col=my_cols()),
-         fill.par=list(col=my_cols_transparent()),
-         points.par=list(col=my_cols()))
-  log10axes(2)
-  axis(1)
-  box()
-  plotlabel("(c)","topleft")
-  visreg(m_lma, "MAT", by="pft", overlay=TRUE, 
-         legend=FALSE,
-         axes=FALSE,
[…]

### 2016-01-07 run.R

@@ -135,19 +135,14 @@ figureS3(dataset, basalafit)
 figlabel("Figure S3")
 dev.off()
 
-pdf("figures/FigureS4.pdf", width = 8L, height = 8L)
-figureS4(dataset)
+pdf("figures/FigureS4.pdf", width = 9L, height = 5L)
+figureS4(table_hierpart,table_varpart_gam,table_varpart_lmer)
 figlabel("Figure S4")
 dev.off()
 
 pdf("figures/FigureS5.pdf", width = 9L, height = 5L)
-figureS5(table_hierpart,table_varpart_gam,table_varpart_lmer)
+figureS5(dataset)
 figlabel("Figure S5")
-dev.off()
-
-pdf("figures/FigureS6.pdf", width = 9L, height = 5L)
-figureS6(dataset)
-figlabel("Figure S6")
 dev.off()
 
 
@@ -156,7 +151,7 @@ dev.off()
 combine <- TRUE
 if(combine){
   figs <- paste(sprintf("%sFigure%i.pdf","figures/", 1:7), collapse=" ")
-  SIfigs <- paste(sprintf("%sFigureS%i.pdf", "figures/", 1:6), collapse=" ")
+  SIfigs <- paste(sprintf("%sFigureS%i.pdf", "figures/", 1:5), collapse=" ")
   cmd <- sprintf("pdftk %s %s cat output Figures%s.pdf",figs,SIfigs,
                  format(as.Date(Sys.time()),"%Y-%m-%d"))
   system(cmd)

### 2016-01-08 manuscript.Rnw
New fig 7 into manuscript
caption for figure7 and paragraph in discussion

@@
 
-% The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation \citep{ise2010, dekauwe2014,friend2014}.
+The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation \citep{ise2010, dekauwe2014,friend2014}.
 Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \citep{ryanlaw2005}.
@@
 Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014,negron2015}.
-% 
+
 % % less focus on GVMs; focus on ecological/biological reasons to test these things
@@
 Sample sizes by PFT are summarised in Table \ref{table:counttable}.
+
 We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
@@
 \item MF/MS differed strongly between PFTs and was proportional to leaf mass per area, and steeply declined with plant height as expected.
-\item We found weak and inconsistent effects of climate on all variables studied, with some exception for LMA.
+\item We found that biomass distribution between leaves and stems was related to climate, but not when plant height and PFT - major confounding variables - were first accounted for.
+when PFT and height were first accounted for, we found weak and inconsistent effects of climate on all variables studied, with some exception for LMA.
 However, climate affects plant height, so we do expect global patterns in biomass distribution (Klein 2015, Simard 2011).
-These results differ from Reich2014 but we don't know why.
+Climate also affects, to some extent, distribution of major plant functional types (Fig.
+1), explaining how our results are consistent with Reich et al.
+(2014), who reported shifts in biomass distribution with MAT.
+We found that for angiosperms, MF/MS increased with MAT, after splitting the dataset into five size classes This The significant relationship between MF/MS and MAT for angiosperms can be understood by the consistently higher proportion of evergreen angiosperms at higher MAT (Fig.
+1), which, as a resutlt of higher LMA (Fig.
+2) also have a higher MF/MS.
+The main conclusions from the climate analysis are certainly that PFT and plant height are by far the most important factors, even explaining most climate effects on biomass distribution.
 \item If biomass distribution is related to leaf mass per area, an important trait, perhaps this can help to place plants on a continuous scale rather than rely on fixed categories.
@@
 Not sure why these differences don't show up for small plants but variance was very high!
-\item Our analysis of climate effects on above-ground biomass distribution between leaves and stems is largely consistent with Reich et al.
-2014.
-We found that for angiosperms, MF/MS increased with MAT (as did Reich), after splitting the dataset into five equally populated bins.
-This result is in apparent contradiction with our finding that climate (MAT, MAP and its interaction) explained consistently very little variation in MF/MS.
-In that analysis, however, the effects of climate were analyzed in addition to the potentially confounding factors plant height and plant functional type.
-The significant relationship between MF/MS and MAT for angiosperms can be understood by the consistently higher proportion of evergreen angiosperms at higher MAT (Fig.
-1), which, as a resutlt of higher LMA (Fig.
-2) also have a higher MF/MS.
-The main conclusions from the climate analysis are certainly that PFT and plant height are by far the most important factors, this is also consistent with Reich 2014 and Poorter 2015.
-
 \item Some paragraph on future work, for example reconcile Poorter2015 with Reich2014 with our data.
@@
 
+\begin{figure}[h!]
+    \centering
+    \ifincludefigs 
+    \includegraphics[width=0.99\textwidth]{Figure7.pdf}
+    \fi
+    \caption{Relationships between leaf - stem biomass distribution ($M_F/M_S$) and mean annual temperature (MAT) (degree*C).
+Similar to Reich et al.
+(2014), we divided the data into five bins of plant size (bins were chosen so that each bin had equal sample size), and used only evergreen vs.
+deciduous plant functional types.
+Lines are fits from mixed-effects models (with species within study as random effect, and only MAT and evergreen/deciduous classification as fixed effects).
+Solid lines are significant (p < 0.05), dashed lines are not (p > 0.05).
+Symbols are partial residuals as calculated with the visreg package in R.}
+    \label{fig:figreichclimate}
+\end{figure}
+
 \clearpage

### 2016-01-28 manuscript.Rnw
Introduction and results written as readable draft

@@
 
-\begin{itemize}
-\item Biomass allocation is important; need to improve GVMs and make them more consistently benchmarked against global observations. 
-\item Plant functional types; can expect differences in biomass partitioning and leaf area relationships.
-Useful from viewpoint of ecology (functional consequences of angiosperm vs.
-gymnosperm divide; evergreen vs.
-deciduous).
-Also useful for GVMs which use PFT for parameterization.
-\item Enquist and Niklas showed differences in leaf mass at given stem mass between gymnosperms and angiosperms.Poorter2015 showed X. Reich2014 showed Y. Open questions are 1,2,3. 
-\item One major gap is that these previous studies did not include estimates of whole-plant leaf area.
-Leaf area is relevant because of light interception.
-It may follow very different patterns from leaf mass.
-Poorter2015 speculated that differences in leaf mass fraction are related to PFT-differences in specific leaf area, but did not have the data to directly test this prediction.
-\item What about climate?
-We know from small-scale experiments and observational studies that biomass distribution varies with climate drivers.
-But unclear if global patterns exist.
-Reich2014 showed X, consistent with glasshouse stuff from Poorter2015/2012.
-\item We have the following objectives:
-\begin{enumerate}
-\item Test whether MF/MS is proportional to leaf mass per area across plant functional types (as suggested by Poorter), once corrected for the expected strong effects of plant size.
-\item Partition the variance of leaf/stem biomass distribution into component variables with the objective to determine which aspects of plant allometry drive differences in biomass distributions between PFTs and with plant size.
-\item Partition variance of component variables into plant height, plant functional type and broad climate variables. 
-\end{enumerate}
-\end{itemize}
-
 The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation \citep{ise2010, dekauwe2014,friend2014}.
@@
 Globally, forests store approximately 360Pg of carbon in living biomass \citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}.
-Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014,negron2015}.
+Reducing uncertainties about biomass partitioning and carbon residence times is a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014,negron2015}.
 
-% % less focus on GVMs; focus on ecological/biological reasons to test these things
-% % ... but make sense to look at PFTs though, and this is the actual reason for doing that?
-% Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation.
-While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots).
-The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010,wullschleger2014}, assumed to differ in key physiological attributes.
-While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are poorly constrained, due to limited available data.
-Moreover, there is little consensus on how biomass partitioning and allocation (see Methods for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
-These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared \citep{wolf2011}.
-% 
-% %explain why stem/leaf biomass distribution is important here.
-% In this paper we are primarily interested in the distribution of biomass between leaves and woody stems, an important component of the residence time of carbon in ecosystems \citep{friend2014}.
-Previous work based on either whole stands \citep{oneilldeangelis1981,enquistniklas2002,reich2014} and or a mix on stand-- and individual-based measurements \citep{poorter2012, poorter2015}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF).
-\citet{poorter2012} and \citet{enquistniklas2002} attributed higher LMF in gymnosperms to longer leaf lifespan (LL) compared to typical angiosperms, but this begs the question whether differences in LMF are also apparent between deciduous and evergreen functional types within angiosperms.
-It is also unknown whether PFTs with higher leaf biomass also have higher total leaf area, which is relevant because leaf area drives total light interception and thus productivity.
-Some oft-cited studies have also assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
-% 
-% Little is known about global-scale patterns in LMF and LAR in relation to climate.
-It can be expected that biomass partitioning is correlated with precipitation or mean annual temperature because smaller-scale comparisons have shown responsiveness of biomass partitioning to environmental drivers \citep{berninger1994,callaway1994,delucia2000,poyatos2007}.
-Moreover, a recent study by \citet{reich2014} demonstrated that stand-scale biomass partitioning was related to mean annual temperature across diverse forest stands.
-Indeed, many GVMs assume that climate affects biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Notes S1).
-Again, however, these models do not agree on the degree of plasticity in biomass partitioning, or which climate variables it should respond to.
-% 
-% % integrate poorter2015 and reich2014 better
-% Despite considerable advances in theory underlying allometric scaling in plants \citep{enquist2003, west1999, savage2010} we do not yet have a clear understanding of potential differences between PFTs in terms of biomass partitioning.
-Instead, most previous work has focussed on understanding size-related shifts in biomass partitioning, as governed by constraints including hydraulic supply and mechanical stability \citep{savage2010}.
-Rather than to advance a specific model of biomass partitioning or allocation, our view is that a broader evidence base is needed first to elucidate patterns between PFTs.
-This should enable those building GVMs to refine algorithms and parameter values to more closely match the observations.
-Here, we use a unique, new database \citep{falster2015} (Fig.~\ref{fig:figoverview}) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms.
-A recent compilation of plant biomass data \citep{poorter2015} speculated that differences between gymnosperms and angiosperms in distribution of biomass between leaves and stems is related to differences in leaf mass per area.
-Here we can directly test this hypothesis as our database, unlike those of \citet{enquistniklas2002}, \citet{reich2014}, and \cite{poorter2015}, includes many observations of leaf area as well as leaf mass measured on the same plants.
-% 
-% A second objective was to study relationships between biomass partitioning and climate variables or biome at a global scale.
-The dataset includes observations of biomass and size metrics for individual plants, compiled from 175 studies across nine vegetation types (Fig.
-S1), across the three major biomes (boreal, temperate and tropical).
+One puzzling aspect of the large inter-specific variation in biomass distribution between leaves and stems is the much larger leaf mass in gymnosperms compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012, poorter2015}.
+\citet{poorter2012} and \citet{enquistniklas2002} attributed higher leaf-stem biomass ratio ($M_F/M_S$) in gymnosperms to longer leaf lifespan (LL) compared to typical angiosperms.
+This is consistent with \citet{poorter2015} who reported lower $M_F/M_S$ in deciduous compared to evergreen species.
+
+It is yet unknown whether plant functional types (PFTs) (angiosperm vs.
+gymnosperm, deciduous vs.
+evergreen) with higher leaf biomass also have higher total leaf area, which is relevant because leaf area drives total light interception and thus productivity.
+Because longer LL is generally correlated with the leaf mass per area (LMA), PFTs with higher MF/MS (and thus higher LMA) do not necessarily not have larger total leaf area at a given stem mass.
+Some oft-cited studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}, but recent large compilations on biomass distribution did not include measurements of leaf area in addition to mass to test this hypothesis \citep{poorter2012,poorter2015,reich2014}.
+\citet{poorter2015} advanced the hypothesis that $MF/MS$ may be nearly proportional to LMA, arguing that total leaf area is likely less variable between PFTs.
+Here we set out to directly test this hypothesis with a database that includes observations of leaf area as well as leaf mass measured on the same plants.
+
+Little is known about global-scale patterns in biomass distribution in relation to climate.
+It can be expected that biomass distribution between leaves and stems is correlated with precipitation or mean annual temperature because smaller-scale comparisons have shown responses to environmental drivers \citep{berninger1994,callaway1994,delucia2000,poyatos2007}.
+Recently, \citet{reich2014} demonstrated that $M_F/M_S$ (for whole stands, either including or excluding below-ground components) increased with mean annual temperature, after separating the data into gymnosperm and angiosperm stands, and accounting for the large size-related variation in biomass distribution.
+Our second objective was thus to test for climate-related variation in above-ground biomass distribution, and specifically to test whether $M_F/M_S$ increases with mean annual temperature.
+
+Something about plant size; we need to account for it and it is always troublesome.
+Also it may be confounded with climate!
+Thus if we want to know direct climate effects it would make sense to study it in addition to plant size, pft.
+Also remind us that MF/MS will decrease quickly with plant size \citep{poorter2015,reich2014,enquist2002}, because of turnover differences.
+
+We use a recently compiled database \citet{falster2015} which includes observations of biomass and size metrics for individual plants (Fig.
+1), compiled from 175 studies across nine vegetation types (Fig.
+S1), across the three major climate zones (boreal, temperate and tropical).
 In this paper we focus on field-grown woody plants, spanning the entire size range of woody plants (0.01 - \textgreater 100 m height).
-One key challenge is thus to account for the very large size variation commonly found in any allometric variable \citep{niklas1994}.
-We do this by fitting a semi-parametric data-driven statistical model, i.e.
-one that does not assume a particular functional form, thereby allowing us to study PFT patterns at a common plant height.
-%
 
@@
 
-$$D_S = D_{Sbh} H^a / (H - H_{bh})^a$$
+\begin{equation}
+D_S = D_{Sbh} H^a / (H - H_{bh})^a
+\end{equation}
 
@@
 
-$$a = c_0H^{c_1}.$$
+\begin{equation}
+a = c_0H^{c_1}
+\end{equation}
 
@@
 
-We used generalised additive models (GAM) to capture  the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height.
+To better understand the PFT and climate-related variation in leaf-stem biomass distribution, it is informative to decompose it into three component variables (Eq. 1),
+
+\begin{equation}
+\frac{M_F}{M_S} = \frac{M_F}{A_F} \cdot \frac{A_F}{A_S} \cdot \frac{A_S}{M_S}
+\end{equation}
+
+where $M_F/A_F$ is the leaf mass per area (LMA), $A_F/A_S$ the ratio of plant leaf area to basal stem area, and $A_S/M_S$ the ratio of stem basal area to total stem mass (including branches).
+Across PFTs, $M_F/A_F$ will thus be proportional to LMA when $A_F/A_S$ and $A_S/M_S$ do not vary between PFTs.
+We also expect that $M_F/M_S$ will show tremendous dependence on plant size \citep{poorter2015,reich2014}.
+We expect that most of this size-related variation is due to variation in $A_S/M_S$, since the total stem mass should be proportional to the plant height, stem cross-sectional area, and a shape parameter \citet{chave2005}.
+Using variance partitioning, we quantify the contribution of plant height, plant functional type, and climate to the overall variation in $M_F/M_S$ and for all three component variables.
+
+Because, similar to \citep{poorter2015}, none of the studied relationships were linear on a log-log scale, we did not use standard allometric equations to study relationships between size and biomass variables.
+Instead, we used generalised additive models (GAM), a semi-parametric modelling approach that makes no assumption about the shape of the relationships.
+GAMs were used to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as $M_F/M_S$ at a common plant height.
 In all fitted GAMs, we used a cubic regression spline.
@@
 Variance explained by quantitative climate variables (MAP and MAT) were tested with GAMs where variables were sequentially added to the model, and the explained variance ($R^2$) calculated.
-For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect.
-We calculated the $R^2$ for linear mixed-effects models for the fixed effects only \citep{nakagawa2013}.
+We compared this approach when using linear mixed-effects models instead of GAMs (calculating the the $R^2$ for linear mixed-effects models for the fixed effects only \citep{nakagawa2013}), as well as an independent method, independent effects analysis.
+The three methods generally agreed on the ranking of variable important (SI Methods).
 
-All analyses were conducted in R 3.2.2 \citep{R-2015}.
+All analyses were conducted in R 3.2.3 \citep{R-2015}.
 GAMs were fitted using the \emph{mgcv} package \citep{wood2006}.
-The code to replicate this analysis are available at http://github.com/RemkoDuursma/baadanalysis.
+The code to replicate this analysis (and all figures) is available at http://github.com/RemkoDuursma/baadanalysis (doi: \emph{to be added}).
 
@@
 
-\begin{itemize}
-\item Dataset covers climate space pretty well. Evergreen angiosperms occurred at higher MAT than deciduous angiosperms and evergreen gymnosperms but there is much overlap between PFTs. 
-\item MF and MS are non-linearly related to plant height (Fig 2), as expected. Also note stem mass increases with height more quickly than leaf mass.
-\item Leaf/stem biomass distribution was very different between PFTs (Fig 3a), and proportional to LMA when estimated at a common height, so that leaf area ratio did not differ between PFTs.
-Consistent with variance partitioning: MF/MS mostly explained by H and PFT, but LAR not much explained by PFT.
-\item Partitioning MF/MS into three component variables, and partition each of the four into H, PFT and climate.
-Climate effects, when analyzed in addition to H and PFT, were very weak, with only LMA and AF/AS somewhat more dependent on climate, but not consistently (Fig S4).
-\item We show that AF was proportional to AS (with the exception of very large trees). AF/AS was nearly unrelated to PFT or plant height.
-\item Stem mass should be proportional to basal area, height, and the product of wood density and stem shape.
-Across all data, the relationship between MS and AS was not related to PFT, but nearly proportional to height (as expected).
-For larger trees, MS at a given AS*H was clearly lower for evergreen gymnosperms, reflecting differences in wood density.
-Again no climate effects.
-\item Histograms summarize the findings, for three variables that do not, or weakly, depend on height. Large PFT effects on LMA, small ones on stem mass index, none on AF/AS. 
-\item We repeat an analysis similar to Reich et al 2014 to test for climate effects on biomass partitioning when only considering the angiosperm/gymnosperm divide, and not first correcting for H as we did previously (but instead binning by H).
-We find that for angiosperms, MF/MS increased with MAT for all but the smallest plants.
-But not for gymnosperms (Fig.
-7).
-\end{itemize}
+%with(baad_mapmat, tapply(MAT, pft, quantile, probs=c(0.25,0.75),na.rm=T))
+The compiled dataset covered the global climate space well (defined by mean annual temperature, MAT; and mean annual precipitation, MAP).
+There was some noticeable separation between the three major PFTs in terms of climate space occupancy.
+Evergreen angiosperms occurred on average at a higher MAT (50\% of studies between 15.5 and 26.0 \textdegree C MAT) than deciduous angiosperms (6.0 - 12.6 \textdegree C) and evergreen gymnosperms (3.4 - 12.4 \textdegree C).
+Plant functional type and climate are thus at least partly confounded.
+
+As expected, both whole-plant leaf mass ($M_F$) and above-ground woody biomass ($M_S$) showed a strong depence on plant height (Fig.
+2).
+For both variables the relationship was highly non-linear on a log-log scale, demonstrating departure from simple allometric relationships.
+It is also evident that $M_S$ increases much quicker with plant height than $M_F$.
+As a result, the ratio $M_F/M_S$ quickly decreased with plant height (Fig.
+3a), from ca.
+1-2 for small saplings (< 0.1m) to less than 0.01 for large trees (>30m).
+
+We found large differences between the three major PFTs in terms of above-ground biomass distribution (Fig.
+3a).
+These differences were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm > evergreen angiosperm > deciduous angiosperm.
+Next we calculated $M_F/M_S$ at a common plant height for each of the PFTs.
+These three size-corrected $M_F/M_S$ values were highly significantly different between PFTs (p < 0.001), and were approximately proportional to the average LMA ($M_F/A_F$) across the PFTs (Fig.
+3b).
+As a result, the amount of leaf area per unit woody above-ground biomass ($A_F/M_S$) did not vary between PFTs, because the differences in $M_F/A_S$ were exactly compensated by differences in LMA (Fig.
+3c).
+
+Next, we decomposed $M_F/M_S$ into three component variables with distinct biological interpretation (Eq.
+3).
+For each of the component variables, ad well as $M_F/M_S$ and $A_F/M_S$, we estimated the variance explained by plant height, PFT and climate variables (MAT, MAP and its interaction) (Table 1).
+We did this by sequentially adding terms to a GAM, and calculating the $R^2$ for each fitted model.
+The first model included only plant height, and already explained much variation (60 - 77\%) in $M_F/M_S$, $A_F/M_S$ and $M_S/A_S$ but not LMA and $A_F/A_S$ (0 - 7 \%).
+When PFT was added to the model (as an intercept term as well as interaction with the smooth H term), the $R^2$ increased substantially for $M_F/M_S$ and $A_F/A_S$ but not $A_F/M_S$ or $M_S/A_S$ (showing PFT invariance for these two variables).
+Additional variance explained was especially large for LMA, which is consistent with large differences shown in Fig.
+3b,c.
+
+% afas
+From the variance partitioning (Table 1), it is evident that $A_F/A_S$ did not vary appreciably between PFTs.
+Indeed, when all PFTs were plotted together, similar $A_F$ was attained at a given $A_S$ across the entire plant size range (with the exception of extremely large trees with > 1 $m^2$ stem basal area, for which little data was available) (Fig.
+4).
+We fit a GAM of $A_F$ as a function of $A_S$ with or without PFT as another covariate (both as intercept and slope interaction), which showed PFT significantly improved model fit (likelihood ratio test, p < 0.001), however the $R^2$ increased only from 83.8 to 85.3 \%, and thus had little if any practical significance.
+
+Likewise, the ratio of stem basal area to stem mass ($A_S/M_S$) apparently varied little between PFTs (Table 1).
+To illustrate this further, when we plotted $M_S$ as a function of $A_S$ it revealed a remarkably tight relationship, with little apparent difference between PFTs (Fig 5a).
+The only exception was for large gymnosperm trees, which consistently showed a lower $M_S$ at a given $A_S$ (Fig.
+5b).
+We fitted a GAM of $M_S$ as a function of $A_S$ (added as a smooth term) with or without PFT as a covariate.
+Again PFT was formally significant (likelihood ratio test, p < 0.001), but the $R^2$ increased only marginally from 80.8 to 82.5\%, again showing little practical significance of PFT in explaining this variable.
+
+Taken together, these results demonstrate that 1) the majority of PFT-related variation in $M_F/M_S$ results from variation in LMA (not $A_F/A_S$ or $A_S/M_S$), and 2) the majority of height-related variation in $M_F/M_S$ results from variation in $A_S/M_S$ (not LMA or $A_F/A_S$).
+It is however important to note that there is very large variation between individuals even after accounting for PFT (Fig.
+6).
+There we expressed the stem mass ratio as $M_S/(H \cdot A_S)$ because we expect from basic geometry that M_S is proportional to H and A_S ($M_S = \phi \rho A_S H$, where $\phi$ a stem shape parameter, $\rho$ wood density).
+The histograms summarize that PFT effects are very evident in LMA (Fig.
+6a), not in $A_F/A_S$, and only a bit for $M_S/A_S$, which is entirely driven by large gymnosperm trees.
+
+Climate variables consistently added little explanatory power to the models when added to a model that included PFT and height (1 - 2.8 \%) (Table 1).
+Since this result is in contrast with recent finding of \citep{reich2014}, we also analyzed climate effects with methods similar to that study.
+When we tested for MAT effects on $M_F/M_S$ when only considering gymnosperms and angiosperms (thus grouping decidous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig.
+S6) for angiosperms for most size classes (but never for gymnosperms).
+We found, however, that this was largely due to the fact that at higher MAT, an increasing proportion of angiosperms are evergreen (Fig.
+1), which have higher $M_F/M_S$.
+When analyzed separately for deciduous and evergreen angiosperms, few and weaker significant relationships were found (Fig.
+S7).
 
@@
 These results suggest promising avenues for parameterising and simplifying biomass partitioning routines in GVMs.
-% 
-% Overall, our results establish general patterns about plant construction and thus lay an empirical base against which models can be benchmarked.
+
 A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions.
@@
 In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data.
+
+Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation.
+While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots).
+The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010,wullschleger2014}, assumed to differ in key physiological attributes.
+While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are poorly constrained, due to limited available data.
+Moreover, there is little consensus on how biomass partitioning and allocation (see Methods for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
+These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared \citep{wolf2011}.
+Rather than to advance a specific model of biomass partitioning or allocation, our view is that a broader evidence base is needed first to elucidate patterns between PFTs.
+This should enable those building GVMs to refine algorithms and parameter values to more closely match the observations.
 
@@
 
-\begin{figure}[h!]
-    \centering
-    \ifincludefigs 
-    \includegraphics[width=0.99\textwidth]{Figure7.pdf}
-    \fi
-    \caption{Relationships between leaf - stem biomass distribution ($M_F/M_S$) and mean annual temperature (MAT) (degree*C).
-Similar to Reich et al.
-(2014), we divided the data into five bins of plant size (bins were chosen so that each bin had equal sample size), and used only evergreen vs.
-deciduous plant functional types.
-Lines are fits from mixed-effects models (with species within study as random effect, and only MAT and evergreen/deciduous classification as fixed effects).
-Solid lines are significant (p < 0.05), dashed lines are not (p > 0.05).
-Symbols are partial residuals as calculated with the visreg package in R.}
-    \label{fig:figreichclimate}
-\end{figure}
-
 \clearpage

### 2016-01-28 manuscript_suppinfo.Rnw
Introduction and results written as readable draft

@@
 
+\begin{figure}[h!]
+    \centering
+    \ifincludefigs 
+    \includegraphics[width=0.99\textwidth]{Figure7.pdf}
+    \fi
+    \caption{Relationships between leaf - stem biomass distribution ($M_F/M_S$) and mean annual temperature (MAT) (degree*C).
+Similar to Reich et al.
+(2014), we divided the data into five bins of plant size (bins were chosen so that each bin had equal sample size), and used only evergreen vs.
+deciduous plant functional types.
+Lines are fits from mixed-effects models (with species within study as random effect, and only MAT and evergreen/deciduous classification as fixed effects).
+Solid lines are significant (p < 0.05), dashed lines are not (p > 0.05).
+Symbols are partial residuals as calculated with the visreg package in R.}
+    \label{fig:figreichclimate}
+\end{figure}
+
 \end{document}

### 2016-01-28 run.R

@@ -78,6 +78,10 @@ table_varpart_gam <- make_table_gamr2MATMAP(dataset)
 table_varpart_lmer <- mixedr2(dataset)
 
 
+# stats for AF as function of AS (fig 4)
+afas <- af_as_stat(dataset)
+
+
 # To check diagnostics, predictions, etc., if needed.
 #lmers <- mixedr2(dataset, returnfit=TRUE)
 figlabel <- function(txt)title(txt, outer=TRUE, line=-0.5, cex.main=0.6, font.main=3)
@@ -137,12 +141,12 @@ figlabel("Figure S5")
 dev.off()
 
 pdf("figures/FigureS5.pdf", width = 9L, height = 5L)
-figureS6_part1(dataset)
+figureS6(dataset)
 figlabel("Figure S6")
 dev.off()
 
 pdf("figures/FigureS6.pdf", width = 9L, height = 5L)
-figureS6_part2(dataset)
+figureS7(dataset)
 figlabel("Figure S7")
 dev.off()
 

### 2016-02-01 manuscript.Rnw
More fixes on stats R2 table; suppinfo fixed and working

@@
 6).
-There we expressed the stem mass ratio as $M_S/(H \cdot A_S)$ because we expect from basic geometry that M_S is proportional to H and A_S ($M_S = \phi \rho A_S H$, where $\phi$ a stem shape parameter, $\rho$ wood density).
+There we expressed the stem mass ratio as $M_S/(H \cdot A_S)$ because we expect from basic geometry that $M_S$ is proportional to H and $A_S$ ($M_S = \phi \rho A_S H$, where $\phi$ a stem shape parameter, $\rho$ wood density).
 The histograms summarize that PFT effects are very evident in LMA (Fig.
@@
 
+<<results='asis'>>=
+library(stargazer)
+m0 <- lme(lmlf_mst ~ lh.t, random=~1|Group, data=dataset, na.action=na.omit)
+m1 <- lme(lmlf_mst ~ lh.t*pft, random=~1|Group, data=dataset, na.action=na.omit)
+
+stargazer(m0,m1)
+@
+
 \clearpage

### 2016-02-01 manuscript_suppinfo.Rnw
More fixes on stats R2 table; suppinfo fixed and working

@@
     \includegraphics[width=0.99\textwidth]{FigureS3.pdf}
-    \caption{\emph{May not go in manuscript, just for notes at the moment}.
-Estimation of basal stem diameter from diameter at breast height for small plants.
-Because 36\% of the dataset is for plants with a height less than 1.3m, we used basal stem area in all analyses.
-For a subset of the data, dbh was measured but not d.ba, even though the plants were only marginally taller than breast height.
-In this figure, black points are the measured ratio of d.ba/d.bh (for plants where both were measured), red the predicted ratio from the fitted nonlinear model.
-Below 1.8m, the variance of the ratio d.ba/d.bh increases drastically, so we excluded those plants from the analysis (unless d.ba was actually measured).}
-    \label{fig:figureS3}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
     \caption{Comparison of three methods for variance partitioning of the five studied variables (see legend) by total plant height, plant functional type and climate (MAP, MAT and interaction).
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
-    \caption{\emph{Should come earlier}.
-A reviewer asked to see MF and AF as a function of total above-ground biomass.
+    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
+    \caption{A reviewer asked to see MF and AF as a function of total above-ground biomass.
 This also shows that MF at given MT diverges between PFTs (except for very small plants), but largely overlaps for AF/MT.
@@
     \centering
-    \ifincludefigs 
-    \includegraphics[width=0.99\textwidth]{Figure7.pdf}
-    \fi
+    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
     \caption{Relationships between leaf - stem biomass distribution ($M_F/M_S$) and mean annual temperature (MAT) (degree*C).
@@
 
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.99\textwidth]{FigureS6.pdf}
+    \caption{As Fig S5 but with angiosperms split up into evergreen and deciduous. Note how climate effects have disappeared.}
+    \label{fig:figreichclimate2}
+\end{figure}
+
 \end{document}

### 2016-02-01 run.R

@@ -76,7 +76,10 @@ dataset2 <- delsmallbh(dataset2)
 # new variance partitioning
 table_hierpart <- make_table_hierpart(dataset)
 
-table_varpart_gam <- make_table_gamr2MATMAP(dataset)
+
+table_varpart_gam <- make_table_gamr2MATARID(dataset)
+
+table_varpart_gam_old <- make_table_gamr2MATMAP_old(dataset)
 
 table_varpart_lmer <- mixedr2(dataset)
 
@@ -84,6 +87,8 @@ table_varpart_lmer <- mixedr2(dataset)
 # stats for AF as function of AS (fig 4)
 afas <- af_as_stat(dataset)
 
+# stats for MS as function of AS (fig 5)
+msas <- ms_as_stat(dataset)
 
 # To check diagnostics, predictions, etc., if needed.
 #lmers <- mixedr2(dataset, returnfit=TRUE)
@@ -134,7 +139,7 @@ figlabel("Figure S2")
 dev.off()
 
 pdf("figures/FigureS3.pdf", width = 9L, height = 4L)
-figureS4(table_hierpart,table_varpart_gam,table_varpart_lmer)
+figureS4(table_hierpart,table_varpart_gam_old,table_varpart_lmer)
 figlabel("Figure S4")
 dev.off()
 
@@ -166,10 +171,12 @@ if(combine){
 
 # tex
 knitr::knit("manuscript.Rnw", "manuscript.tex")
+tex_2_pdf("manuscript.tex")
+
+
 knitr::knit("manuscript_suppinfo.Rnw", "manuscript_suppinfo.tex")
 
 # pdf
-tex_2_pdf("manuscript.tex")
 tex_2_pdf("manuscript_suppinfo.tex")
 
 

### 2016-02-01 manuscript.Rnw
Moved back to remake

@@
 
-print(xtable(table_hierpart),sanitize.text.function=function(x){x},
+print(xtable(table_varpart_gam),sanitize.text.function=function(x){x},
       include.rownames=FALSE)
 
-@
-
-<<results='asis'>>=
-library(stargazer)
-m0 <- lme(lmlf_mst ~ lh.t, random=~1|Group, data=dataset, na.action=na.omit)
-m1 <- lme(lmlf_mst ~ lh.t*pft, random=~1|Group, data=dataset, na.action=na.omit)
-
-stargazer(m0,m1)
 @

### 2016-02-01 R/figures.R

@@ -388,33 +388,7 @@ figureS2 <- function(dataset, KGAM=4){
 
 
 
-
-# based on this figure, we should exclude plants with height < 1.8 that have a BH but not BA measurement.
-# test for small plants
-figureS3 <- function(dataset, basalafit){
-  test <- subset(dataset, !is.na(d.ba) & !is.na(d.bh) & h.t < 2.5 & h.bh >= 1.3 & h.bh <= 1.4)
-  
-  test$d.bapred <- predict(basalafit, test)
-  
-  # randomly reorder rows, so that colours shown in proportion to abundance
-  set.seed(1) # Set seed to ensure plot looks same each time
-  dataset <- dataset[sample(nrow(dataset)),]
- 
-  smoothplot(h.t, d.ba/d.bh, data=test, axes=FALSE, pointcols=alpha("black", 0.5), linecols="black",
-             fitoneline=TRUE,
-             ylab=expression(D[BA]/D[BH]~~("-")), xlab=expression(H~(m)),
-             ylim=c(0,15))
-  axis(1)
-  axis(2)
-  box()
-  with(test, abline(h=median(d.ba/d.bh), lty=5))
-  with(test, points(h.t, d.bapred/d.bh, pch=17, col="red"))
-}
-
-
-
-
-figureS4 <- function(table_hierpart,table_varpart_gam,table_varpart_lmer){
+figureS3 <- function(table_hierpart,table_varpart_gam,table_varpart_lmer){
   
   # compare three methods variance partitioning
   convertm <- function(tab){
@@ -460,7 +434,7 @@ figureS4 <- function(table_hierpart,table_varpart_gam,table_varpart_lmer){
 }
 
 
-figureS5 <- function(dataset){
+figureS4 <- function(dataset){
   
   # randomly reorder rows, so that colours shown in proportion to abundance
   set.seed(100) # Set seed to ensure plot looks same each time
@@ -488,7 +462,7 @@ figureS5 <- function(dataset){
 
 
 # new climate figure, cf. Reich
-figureS6 <- function(dataset){
+figureS5 <- function(dataset){
   
   dataset$lhtclass <- quantcut(dataset$lh.t, 5)
   dataset$pft2 <- ifelse(dataset$pft == "EG", "Gymnosperm", "Angiosperm")
@@ -542,7 +516,7 @@ figureS6 <- function(dataset){
 }
 
 
-figureS7 <- function(dataset){
[…]

### 2016-02-01 manuscript.Rnw
caption table 1

@@
 Figure 2a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary.
+Thanks to a small army of anonymous reviewers for suggestions that greatly improved the paper.
+Special thanks to Steve Zissou.
 
@@
 
-print(xtable(table_varpart_gam),sanitize.text.function=function(x){x},
-      include.rownames=FALSE)
+print(xtable(table_varpart_gam,
+             caption="Explained variance in four plant biomass variables by plant height (H), plant functional type (PFT) and climate.
+Shown are $R^2$ estimated with generalized additive models (GAM), with four set of predictors.
+All variables except PFT were added as smooth terms to the GAMs, avoiding the need to assume linear partial responses to the predictor variables, or search for best-fitting transformations.
+Climate variables included were either MAT or aridity index (calculated as the of PET to MAP)."),sanitize.text.function=function(x){x},
+      include.rownames=FALSE, caption.placement = "top")
 

### 2016-02-01 manuscript.Rnw
table with sample sizes added back

@@
 <<results='asis'>>=
+print(xtable(table_samplesize,
+             caption="Sample sizes used for the four studied whole-plant variables, by plant functional type.
+Numbers indicate individual plants, with number of unique species between parentheses.
+Sample sized differed between variables because not all included studied measured all variables ("),
+      sanitize.text.function=function(x){x},
+      include.rownames=TRUE, caption.placement = "top")
+@
 
+<<results='asis'>>=
+names(table_varpart_gam)[1] <- ""
 print(xtable(table_varpart_gam,

### 2016-02-01 R/tables_stats.R

@@ -176,7 +176,7 @@ make_table_gamr2MATARID <- function(dataset) {
   g0 <- gamr2(dataset, kgam=4, ranef=TRUE, climvar1="MAT", climvar2="aridity")
   
   r2table <- g0$r2table
-  r2table$Variable <- c("$M_F/M_S$","$A_F/A_S$","$M_F/A_F$","$M_S/A_S$")
+  r2table$Variable <- c("$M_F/M_S$","$A_F/A_S$","$M_F/A_F$","$A_S/M_S$")
 
   # AIC does not seem to be compatible with R2 results  
   # aicf <- function(x){
@@ -191,25 +191,32 @@ r2table
 
 
 # Count number of observations.
-tabFun <- function(x, vars, pftvar="pft", vegvar="bortemptrop"){
+tabFun <- function(x, vars, pftvar="pft"){
   
   x <- x[complete.cases(x[,vars]),]
   x$pft <- x[,pftvar]
-  x$veg <- x[,vegvar]
   
-  xt <- addmargins(xtabs( ~ pft + veg, data=x))
-  names(dimnames(xt)) <- c(pftvar,vegvar)
+  xt <- addmargins(xtabs( ~ pft, data=x))
+  names(dimnames(xt)) <- pftvar
   
-  x <- x[!duplicated(x$species,x$pft,x$veg),]
-  xs <- addmargins(xtabs( ~ pft + veg, data=x))
+  x <- x[!duplicated(x$species,x$pft),]
+  xs <- addmargins(xtabs( ~ pft, data=x))
   
-  m <- matrix(paste0(xt, " (", xs, ")"), ncol=ncol(xt))
-  dimnames(m) <- dimnames(xt)
+  m <- paste0(xt, " (", xs, ")")
   
   m[m == "0 (0)"] <- NA
   
   return(m)
 }
+
+make_samplesize_table <- function(dataset){
+  tb <- t(sapply(c("lmlf_mst","lalf_astba2","llma","lastba2_mst"),function(x)tabFun(dataset,x)))
+  rownames(tb) <- c("$M_F/M_S$","$A_F/A_S$","$M_F/A_F$","$A_S/M_S$")
+  colnames(tb) <- c("Decid. Angio.","Evergr. Angio.","Evergr. Gymno.","Total")
+  tb
+}
+
+
 
 
 # For SuppInfo Figs. S5 and S6

### 2016-02-01 manuscript.Rnw
Captions tidied

@@
 print(xtable(table_samplesize,
+             label="counttable",
              caption="Sample sizes used for the four studied whole-plant variables, by plant functional type.
@@
 print(xtable(table_varpart_gam,
+             label="varparttable",
              caption="Explained variance in four plant biomass variables by plant height (H), plant functional type (PFT) and climate.
-Shown are $R^2$ estimated with generalized additive models (GAM), with four set of predictors.
+Shown are adjusted $R^2$ estimated with generalized additive models (GAM), with four sets of predictors.
 All variables except PFT were added as smooth terms to the GAMs, avoiding the need to assume linear partial responses to the predictor variables, or search for best-fitting transformations.
-Climate variables included were either MAT or aridity index (calculated as the of PET to MAP)."),sanitize.text.function=function(x){x},
+Climate variables included were either MAT or aridity index (calculated as the of PET to MAP).
+P-values not shown because all smooth terms were always highly significant (p < 0.01), with the exception of aridity index in the model for $M_F/M_S$ (p = 0.057)."),sanitize.text.function=function(x){x},
       include.rownames=FALSE, caption.placement = "top")
 
+# p-values not added for clarity. For H and PFT, p-values always very, very small.
+# For climate variables, calculate as follows:
+# g0 <- gamr2(dataset, kgam=4, ranef=TRUE, climvar1="MAT", climvar2="aridity") # slow
+# f <- function(x)anova(x$gam)$s.pv[4]
+# t(sapply(g0$fits, function(x)sapply(x,f)))  # third column MAT, fourth column aridity.
 @
@@
     \fi
-    \caption{Overview of the data.
+    \caption{Overview of the database.
 (a) Variables were measured on up to \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species.
@@
 Sample sizes are listed in Table \ref{table:counttable}.
-Lines are generalized additive model fits to visualize the highly non-linear relationships.}
+Lines are generalized additive model fits to visualize the highly non-linear relationships.
+Note the logarithmic scale.}
     \label{fig:figrawdata}
@@
     \fi
-    \caption{Total plant leaf area ($A[L]$) is approximately proportional to basal stem area ($A_S$), and this relationship is similar between PFTs.
-Lines are standardized major axis fits.
-Coefficients are shown on the figure (X), only evergreen gymnosperms showed a slightly more than proportional slope (1.05).}
+    \caption{Total plant leaf area ($A_F$) is approximately proportional to basal stem area ($A_S$), and this relationship is similar between PFTs.
+Lines are generalised additive model fits, with grey areas approximate 95\% confidence intervals.}
     \label{fig:alas}
@@
     \fi
-    \caption{Relationships between above-ground woody biomass ($M_S$), basal stem area ($A_S$) and plant height (H).
+    \caption{Above-ground woody biomass ($M_S$) scales with basal stem area ($A_S$), similarly for the three PFTs.
 (a) $M_S$ as a function of $A_S$ for the entire dataset.
 Lines are generalized additive model fits.
-(b) Much additional variation in $M_S$ can be explained by H.
-Shown is the ratio $M_S/A_S$ as a function of H, with generalized additive model fits.
-(c) Stem mass as a function of $A_S*H$ for a subset of the data in panels a and b, where the product of H and $A_S$ was larger than 0.01.
-For this subset of larger trees, stem mass was nearly proportional to $A_S*H$ (SMA fits).
-(d) From the data in panel c, the average ratio $M_S/(A_S*H)$ differs between plant functional types, with a lower value for gymnosperms compared to angiosperms.}
+(b) A close view of a portion of the data shown in panel (a), where evergreen gymnosperms displayed significantly lower woody biomass at a given stem basal area.}
     \label{fig:stemmassindex}
@@
 
-% \textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\
-% \textbf{Fig. S2} Leaf area ratio ($A_F/M_T$) by PFT.\\
-% \textbf{Fig. S3} Relationship between above-ground biomass and basal stem area.\\
-% \textbf{Notes S1} Modelling of biomass partitioning in global vegetation models (GVMs)
+\textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\
+\textbf{Fig. S2} Leaf area ratio ($A_F/M_T$) by PFT.\\
+\textbf{Fig. S3} Comparison of three methods for variance partitioning of the four studied variables \\
+\textbf{Fig. S4} Whole-plant leaf mass ($M_F$) and leaf area ($A_F$) as a function of total aboveground biomass. \\ 
+\textbf{Fig. S5} Relationships between leaf - stem biomass distribution ($M_F/M_S$) and mean annual temperature (MAT) (\degree C). \\
+\textbf{Fig. S6} As Fig S5 but with angiosperms split up into evergreen and deciduous. \\
 

### 2016-02-01 manuscript_suppinfo.Rnw
Captions tidied

@@
     \includegraphics[width=0.99\textwidth]{FigureS3.pdf}
-    \caption{Comparison of three methods for variance partitioning of the five studied variables (see legend) by total plant height, plant functional type and climate (MAP, MAT and interaction).
+    \caption{Comparison of three methods for variance partitioning of the four studied variables (see legend) by total plant height, plant functional type and climate (MAP, MAT and interaction).
 \textbf{GAM}: variance component as estimated with a generalized additive model, as the reduction in R2 when the independent variable (i.e.
@@
     \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
-    \caption{A reviewer asked to see MF and AF as a function of total above-ground biomass.
-This also shows that MF at given MT diverges between PFTs (except for very small plants), but largely overlaps for AF/MT.
-Exception is for very (very) large trees, but we only have two datasets there.}
+    \caption{Whole-plant leaf mass ($M_F$) and leaf area ($A_F$) as a function of total aboveground biomass.
+In the main text, we have chosen to use plant height as a metric of plant size, to avoid correlations between dependent and independent variables ($M_F$ appears in both variables in panel (a)).
+The overall differences between PFTs, and the finding that leaf mass is more variable between PFTs than leaf area, still hold when we use mass as the size covariate.}
     \label{fig:figureS5}
@@
     \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
-    \caption{Relationships between leaf - stem biomass distribution ($M_F/M_S$) and mean annual temperature (MAT) (degree*C).
+    \caption{Relationships between leaf - stem biomass distribution ($M_F/M_S$) and mean annual temperature (MAT) (\degree C).
 Similar to Reich et al.
@@
     \includegraphics[width=0.99\textwidth]{FigureS6.pdf}
-    \caption{As Fig S5 but with angiosperms split up into evergreen and deciduous. Note how climate effects have disappeared.}
+    \caption{As Fig S5 but with angiosperms split up into evergreen and deciduous. Note how climate effects have all but disappeared, except for evergreen angiosperms in two size class bins.}
     \label{fig:figreichclimate2}

### 2016-02-01 manuscript.Rnw
discussion work

@@
 
-Climate variables consistently added little explanatory power to the models when added to a model that included PFT and height (1 - 2.8 \%) (Table 1).
+Climate variables consistently added little explanatory power to the models when added to a model that included PFT and height (1 - 2.8 \%) (Table~\ref{table:varparttable}).
 Since this result is in contrast with recent finding of \citep{reich2014}, we also analyzed climate effects with methods similar to that study.
@@
 \item MF/MS differed strongly between PFTs and was proportional to leaf mass per area, and steeply declined with plant height as expected.
+\item We found that, as expected, plant size strongly influences LMF (Fig.~\ref{fig:figlmf}) and LAR (Fig.
+S2) (see also \citet{poorter2012,poorter2015}).
+It is thus necessary to correct for plant size when comparing biomass partitioning parameters, as has been noted many times \citep{mcconnaughay1999}.
+We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship.
+This was useful because both LMF and LAR showed very non-linear patterns with plant height, even on a logarithmic scale, consistent with recent other results \citep{poorter2015}.
 \item We found that biomass distribution between leaves and stems was related to climate, but not when plant height and PFT - major confounding variables - were first accounted for.
 when PFT and height were first accounted for, we found weak and inconsistent effects of climate on all variables studied, with some exception for LMA.
-However, climate affects plant height, so we do expect global patterns in biomass distribution (Klein 2015, Simard 2011).
+However, climate affects plant height, so we do expect global patterns in actual biomass distribution (Klein 2015, Simard 2011).
 Climate also affects, to some extent, distribution of major plant functional types (Fig.
@@
 2) also have a higher MF/MS.
-The main conclusions from the climate analysis are certainly that PFT and plant height are by far the most important factors, even explaining most climate effects on biomass distribution.
-\item If biomass distribution is related to leaf mass per area, an important trait, perhaps this can help to place plants on a continuous scale rather than rely on fixed categories.
+However, based on a re-analysis of data in Reich et al.
+2014 (not shown), this does not explain the pattern observed by them as it still held for gymnosperms and deciduous angiosperms.
+Whatever explains that different findings between us and Reich 2014, the main conclusions from the climate analysis are certainly that PFT and plant height are by far the most important factors, even explaining most climate effects on biomass distribution.
+\item Still lots and lots of variation to explain within PFTs; we do not argue for convergence, but rather that as a first step, we can understand differences between PFTs.
+This helps parameterize PFTs, and also perhaps this can help to place plants on a continuous scale rather than rely on fixed categories.
 This is an active area of research, but is yet to be extended to patterns in biomass distribution.
-\item Still lots and lots of variation to explain within PFTs; we do not argue for convergence, but rather that as a first step, we can understand differences between PFTs.
+Our work contributes a bit to this.
 \item Leaf area was just about proportional to AS, except for the largest trees.
@@
 Not sure why these differences don't show up for small plants but variance was very high!
-\item Some paragraph on future work, for example reconcile Poorter2015 with Reich2014 with our data.
+For tropical forests, \citet{chave2005} showed that aboveground biomass can be estimated as,
+
+$$M_T = F \rho A_S H_T$$
+
+where $F$ is a form factor, and $\rho$ wood density (kg m$^3$). This explains why (for larger trees) at a given AS, conifers had lower stem mass (Fig. 5).
 \item Consequences for GVMs.
+
 \end{itemize}
-
-% We found that PFTs differed in LMF, and that LMF was proportional to LMA.
-This conclusion was particularly strong across the three PFTs studied, but it also held across species within PFT.
-The implication is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) does not differ between PFTs.
-This result seems robust, as our data include individual plants spanning the entire size range of woody plants in natural forests (0.1-100m) (Fig.~\ref{fig:figlmf}a).
-Previous studies have demonstrated a large difference in LMF between angiosperms and gymnosperms \citep{poorter2012,poorter2015}, but have not been able to explain these differences in terms of leaf traits.
-We show that LMA -- a central trait of the leaf economics spectrum (LES) \citep{wrightetal2004} -- explains differences in LMF in a consistent manner.
-% 
-% We found that, as expected, plant size strongly influences LMF (Fig.~\ref{fig:figlmf}) and LAR (Fig.
-S2) (see also \citet{poorter2012,poorter2015}).
-It is thus necessary to correct for plant size when comparing biomass partitioning parameters, as has been noted many times \citep{mcconnaughay1999}.
-We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship.
-This was useful because both LMF and LAR showed very non-linear patterns with plant height, even on a logarithmic scale, consistent with recent other results \citep{poorter2015}.
-% 
-% The finding that LMF is proportional to LMA across PFTs implies that LAR does not differ systematically between PFTs.
-One explanation for this pattern lies in the strong positive correlation between LMA and leaf lifespan (LL) \citep{wrightetal2004}.
-Both \cite{poorter2012} and \citet{enquistniklas2002} explained higher LMF in gymnosperms by higher LL; plants simply maintain more cohorts of foliage, but this explanation demands that all else is equal.
-Yet it is not self-evident that everything else is indeed equal, that is, annual production of foliage could differ between PFTs.
-Nonetheless, further supporting this explanation, \citet{reich1992} showed a positive correlation between LL and total stand leaf biomass across diverse forest stands.
-Another, not mutually exclusive, possible explanation is that plants may allocate biomass in a manner that targets a more constant LAR rather than LMF.
-Canopy size scales with $M_T$ \citep{duursma2010}, and a larger canopy means higher light interception per unit total biomass \citep{duursma2007}.
-It can thus be argued that an optimal LAR exists that balances investment in supporting woody biomass (reflected by $M_T$) and the efficiency of foliage in terms of light interception.
-% 
-% We found that patterns in $M_F/A_S$ across PFTs mirrored those of LMF (Fig.~\ref{fig:figlmf}).
-Because LMF may be decomposed into the terms $M_F/A_S$ and $M_T/A_S$ (eq.
-\ref{eq:M2}), this result suggests that $M_T/A_S$ does not vary between PFTs (see also Fig.
-S3).
-If this result holds, it means that aboveground biomass may be estimated across PFTs with a single equation that does not differ between PFTs.
-For tropical forests, \citet{chave2005} showed that aboveground biomass can be estimated as,
-% 
-% $$M_T = F \rho A_S H_T$$
-% 
-% where $F$ is a form factor, and $\rho$ wood density (kg m$^3$).
-After rearranging, this equation predicts that $M_T/A_S$ is proportional to plant height.
-We found some support for this prediction (Fig.
-S3), suggesting further improvements could be made by incorporating this dependence.
-It is timely to use database made public by \citet{falster2015} (BAAD) and \citet{poorter2015} to develop more general equations to estimate individual plant aboveground biomass across the globe.
-% 
-% We found that climate did not appreciably affect biomass partitioning between leaves and stems.
-This finding is in contrast with a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}.
-One explanation for this difference is that our measurements were taken on individual plants, whereas in  \citet{reich2014} they were on whole stands.
-This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants.
-With the caveats that there is still variation between species and studies unaccounted for (Fig.~\ref{fig:fighist}), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Table \ref{table:statstable}).
-% 
-% Our results indicate that it is possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning.
-Indeed, nearly as much variation was explained across species when using LMA instead of PFT in a statistical model.
-This is surprising because LMA only captures one aspect of functional differentiation among PFT -- leaf morphology (but including correlated effects on leaf lifespan and photosynthetic capacity via the leaf economics spectrum \citep{wrightetal2004}).
-We thus show that in a PFT-based classification, LMA is a good first estimate of biomass partitioning, however, more variation can be explained by another trait ($A_F/A_S$), which varies appreciably between species.
-These results suggest promising avenues for parameterising and simplifying biomass partitioning routines in GVMs.
 
@@
 This should enable those building GVMs to refine algorithms and parameter values to more closely match the observations.
+
+\begin{itemize}
+\item Some paragraph on future work, for example reconcile Poorter2015 with Reich2014 with our data.
+\end{itemize}
 

### 2016-02-02 manuscript.Rnw
Structuring discussion

@@
 % #' Number of tables: 2 \\
-% #' Supporting information : 1 Tables, 3 Figures and 1 Notes.
+% #' Supporting information : 6 Figures (all in colour).
 
@@
 
-\begin{itemize}
-\item MF/MS differed strongly between PFTs and was proportional to leaf mass per area, and steeply declined with plant height as expected.
-\item We found that, as expected, plant size strongly influences LMF (Fig.~\ref{fig:figlmf}) and LAR (Fig.
-S2) (see also \citet{poorter2012,poorter2015}).
-It is thus necessary to correct for plant size when comparing biomass partitioning parameters, as has been noted many times \citep{mcconnaughay1999}.
+Using the largest database of individual woody plant allometry to date, we studied patterns in above-ground biomass distribution among major woody plant functional types (PFTs), plant size, and climate.
+We found that at a common plant height, the leaf / woody biomass ratio ($M_F/M_S$) was proportional to the leaf mass per area across three PFTs, so that leaf area per unit above-ground biomass did not vary between PFTs.
+We found very weak effects of climate (mean annual temperature and aridity) on biomass distribution, the vast majority of the variation in biomass distribution was explained by plant size and PFT.
+
+As expected, $M_F/M_S$ steeply declined with plant height.
+It is well known that it is necessary to correct for plant size when comparing biomass distribution among groups \citep{mcconnaughay1999}.
 We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship.
-This was useful because both LMF and LAR showed very non-linear patterns with plant height, even on a logarithmic scale, consistent with recent other results \citep{poorter2015}.
-\item We found that biomass distribution between leaves and stems was related to climate, but not when plant height and PFT - major confounding variables - were first accounted for.
-when PFT and height were first accounted for, we found weak and inconsistent effects of climate on all variables studied, with some exception for LMA.
-However, climate affects plant height, so we do expect global patterns in actual biomass distribution (Klein 2015, Simard 2011).
-Climate also affects, to some extent, distribution of major plant functional types (Fig.
-1), explaining how our results are consistent with Reich et al.
-(2014), who reported shifts in biomass distribution with MAT.
-We found that for angiosperms, MF/MS increased with MAT, after splitting the dataset into five size classes This The significant relationship between MF/MS and MAT for angiosperms can be understood by the consistently higher proportion of evergreen angiosperms at higher MAT (Fig.
-1), which, as a resutlt of higher LMA (Fig.
-2) also have a higher MF/MS.
+This was useful because both $M_F$ and $M_S$ showed very non-linear patterns with plant height (Fig.
+2), or among biomass pools (Fig.
+S4), even on a logarithmic scale.
+This is consistent with recent results on an independently compiled large database of plant allometry \citep{poorter2015}, and further calls into question the generality of the often-assumed log-linear relationships between plant biomass pools and size metrics.
+
+After accounting for PFT and plant height, we found very weak and inconsistent effects of climate on all variables studied, with some exception for LMA (Table~\ref{table:varparttable}).
+Climate was represented as either mean annual temperature or an aridity index (PET/MAP), however, results were similarly inconclusive when studying other climate variables such as composite drought indices, growing season length, or mean annual precipitation (results not shown).
+This result appears to contradict \citet{reich2014} who reported consistent increases in foliage biomass ($M_F$) with MAT, when corrected for above-ground woody biomass ($M_S$).
+We were able to partly reconcile our findings with those of \citep{reich2014}: when we considered only angiosperms and gymnosperms (thus grouping deciduous and evergreens, as did Reich et al.) $M_F/M_S$ increased consistently with MAT for most size classes (Fig.
+S5), but not for gymnosperms.
+However, we found that for our database, this could be explained by the consistently higher proportion of evergreen angiosperms at higher MAT (Fig.
+1), which, as a result of higher LMA (Fig.
+2) also have a higher $M_F/M_S$.
 However, based on a re-analysis of data in Reich et al.
-2014 (not shown), this does not explain the pattern observed by them as it still held for gymnosperms and deciduous angiosperms.
-Whatever explains that different findings between us and Reich 2014, the main conclusions from the climate analysis are certainly that PFT and plant height are by far the most important factors, even explaining most climate effects on biomass distribution.
-\item Still lots and lots of variation to explain within PFTs; we do not argue for convergence, but rather that as a first step, we can understand differences between PFTs.
+2014 (not shown), this shift in PFT with climate does not explain their observation of an increase in $M_F/M_S$ with MAT, as this pattern still held within gymnosperms and deciduous angiosperms.
+
+Although we found no consistent effects of climate on biomass distribution after PFT and plant height were accounted for, 
+effects of climate may be mediated by plant height.
+At a global scale, average plant height is strongly related to climate, likely mediated by water availability \citet{klein2015,simard2011}.
+Since biomass distribution is strongly related to plant height (Figs.
+2, 3), we do expect correlations between climate variables and observed biomass distribution.
+
+Still lots and lots of variation to explain within PFTs; we do not argue for convergence, but rather that as a first step, we can understand differences between PFTs.
 This helps parameterize PFTs, and also perhaps this can help to place plants on a continuous scale rather than rely on fixed categories.
@@
 Our work contributes a bit to this.
-\item Leaf area was just about proportional to AS, except for the largest trees.
+
+Leaf area was just about proportional to AS, except for the largest trees.
 This decreases because of the formation of heartwood (and AF/sapwood may decrease a little with height, but AF/AS more because of crown rise).
 Note that we have many small plants in the dataset as well, and within individual datasets it is more likely to pick up small-scale effects like these.
-\item Stem mass index: Chave biomass estimation.
+
+Stem mass index: Chave biomass estimation.
 We find support for this, with MS/AS nearly proportional to height.
@@
 where $F$ is a form factor, and $\rho$ wood density (kg m$^3$). This explains why (for larger trees) at a given AS, conifers had lower stem mass (Fig. 5).
-\item Consequences for GVMs.
 
-\end{itemize}
-
+\emph{combine next two in shorter GVM paragraph}
 A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions.
@@
 This should enable those building GVMs to refine algorithms and parameter values to more closely match the observations.
-
-\begin{itemize}
-\item Some paragraph on future work, for example reconcile Poorter2015 with Reich2014 with our data.
-\end{itemize}
 

### 2016-02-03 manuscript.Rnw
polishing, fix links, finish discussion, new summary

@@
+% Notes:
+% for submission, convert to Word docx using Adobe Acrobat Professional.
+To not include line numbers here, add them in Word.
+Also do not include figures.
+Word count can appear on first page; follow instructions further below.
+
 \documentclass[a4paper]{article}
-
-%texcount manuscript.tex > wordcount.txt
-%shell("texcount manuscript.tex > wordcount.txt")
 
@@
 \usepackage{palatino}
-% \usepackage[default,osfigures,scale=0.95]{opensans}
+\usepackage[default,osfigures,scale=0.95]{opensans}
 
@@
 \usepackage{lineno}
+\usepackage{url}
+\usepackage{breakurl}
 \geometry{verbose,tmargin=2.4cm,bmargin=2.4cm,lmargin=2.4cm,rmargin=2.4cm}
@@
 
-% \linenumbers
+% Do not turn on if converting to Word later.
+%\linenumbers
 
@@
 
-\affil[1]{Hawkesbury Institute for the Environment, University of Western Sydney, Locked Bag 1797, Penrith, NSW 2751, Australia}
+\affil[1]{Hawkesbury Institute for the Environment, Western Sydney University, Locked Bag 1797, Penrith, NSW 2751, Australia}
 \affil[2]{Biological Sciences, Macquarie University, North Ryde, NSW 2109, Australia}
@@
 
-%% uncomment this after compiling the ms, then compile again. This is a recursive dependency :(
-% Also, requires Perl.
-% #' <<>>=
-% #' shell("texcount manuscript.tex > wordcount.txt")
-% #' nrw <- get_wordcount("wordcount.txt")
-% #' @
-% #' 
-% #' Total number of words: \Sexpr{nrw$Total} \\
-% #' Abstract: \Sexpr{nrw$Abstract} \\
-% #' Introduction: \Sexpr{nrw$Introduction} \\
-% #' Methods and Materials: \Sexpr{nrw$Methods} \\
-% #' Results: \Sexpr{nrw$Results} \\
-% #' Discussion: \Sexpr{nrw$Discussion} \\
-% #' 
-% #' Number of figures: 6 (all in colour) \\
-% #' Number of tables: 2 \\
-% #' Supporting information : 6 Figures (all in colour).
+% Word count
+%% Uncomment this after compiling the ms, then compile again.
+% Requires Perl to be installed.
+%' <<echo=FALSE>>=
+%' #shell("texcount manuscript.tex > wordcount.txt")
+%' nrw <- get_wordcount("wordcount.txt")
+%' @
+%' Total number of words: \Sexpr{nrw$Total} \\
+%' Abstract: \Sexpr{nrw$Abstract} \\
+%' Introduction: \Sexpr{nrw$Introduction} \\
+%' Methods and Materials: \Sexpr{nrw$Methods} \\
+%' Results: \Sexpr{nrw$Results} \\
+%' Discussion: \Sexpr{nrw$Discussion} \\
+%' Number of figures: 6 (all in colour) \\
+%' Number of tables: 2 \\
+%' Supporting information : 6 Figures (all in colour).
+
+% Alternatively, in a shell,
+% 
+%texcount manuscript.tex > wordcount.txt
+%shell("texcount manuscript.tex > wordcount.txt")
 
@@
 
-% 200 word max.
-\begin{enumerate}
-\item The partitioning of biomass into leaves and stems is an important component of global vegetation models (GVMs).
-\item We use the recently published Biomass And Allometry Database (BAAD), a large database including observations for individual plants.
-The database covers the global climate space, allowing us to test for direct climate effects in addition to PFT and plant height.
-\item The ratio of leaf mass to woody above-ground mass (MF/AS) varied strongly between PFTs (as defined by deciduous vs.
+%The Summary for research papers, which must be usable as a stand-alone document, must not exceed 200 words and should be organized using four bullet points to indicate (1) the research conducted, including the rationale, (2) methods, (3) key results and (4) the main conclusion, including the key points of discussion.
+\begin{itemize}
+\item The partitioning of biomass into leaves and stems is an important determinant of the residence time of carbon in vegetation.
+We tested for differences in biomass distribution between major woody plant functional types (PFT), and sought to understand these differences in terms of leaf traits.
+\item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species.
+We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, and climate on biomass distribution and three component variables.
+\item The ratio of leaf mass to woody above-ground mass ($M_F/A_S$) varied strongly between PFTs (as defined by deciduous vs.
 evergreen and gymnosperm vs.
 angiosperm).
-We found that MF/MS, once corrected for plant height, was proportional to leaf mass per area across PFTs.
+We found that $M_F/A_S$, once corrected for plant height, was proportional to leaf mass per area across PFTs.
 As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass.
-\item We partitioned the variance of MF/MS into three component variables (AF/AS, MF/AF and MS/AS), and each of these four variables into PFT, climate and plant height.
-Plant height explained most variation in MF/MS and MS/AS (but not H or MF/AF).
-PFT explained much variation in MF/MS and MF/AF (but not AF/AS and only for larger plants in MS/AS).
-\item Climate variables (MAT, MAP and its interaction) consistently explained very little additional variation, demonstrating very weak control of climate on biomass distribution and leaf area, at a given plant size.
-We argue that climate effects will largely be indirectly through effects of plant height, and climate-related shifts in plant functional types.
-\item Combined, these results show consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain GVMs.
-\end{enumerate}
+Climate consistently explained very little additional variation, demonstrating very weak control of climate on biomass distribution, at a given plant size.
+\item Combined, these results demonstrate consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain global vegetation models.
+\end{itemize}
 
@@
 
-Something about plant size; we need to account for it and it is always troublesome.
-Also it may be confounded with climate!
-Thus if we want to know direct climate effects it would make sense to study it in addition to plant size, pft.
-Also remind us that MF/MS will decrease quickly with plant size \citep{poorter2015,reich2014,enquist2002}, because of turnover differences.
-
-We use a recently compiled database \citet{falster2015} which includes observations of biomass and size metrics for individual plants (Fig.
+To test our hypotheses, we use a recently compiled large database \citep{falster2015} which includes observations of biomass and size metrics for individual plants (Fig.
 1), compiled from 175 studies across nine vegetation types (Fig.
@@
 
-We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total includes records for \Sexpr{nrow(baad_all$data)} individuals.
-The database has very limited overlap (n = 261, 1.7 \%) with the recent large compilation of \citet{poorter2015} and differs in that measurements are all for individual plants (where \citet{poorter2015} included many stand-based estimates converted back to invididuals).
+We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total includes records for \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species.
+The database has very limited overlap (n = 261 individuals, 1.7 \%) with the recent large compilation of \citet{poorter2015} and differs in that measurements are all for individual plants (where \citet{poorter2015} included many stand-based estimates converted back to invididuals).
 In this paper we restrict our analysis to records that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$) (n=\Sexpr{nrow(dataset)}).
@@
 We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
+Excluding plantations from the analysis did not affect the main result, but far fewer data were available for evergreen gymnosperms.
 We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms.
 We excluded  deciduous gymnosperms because few data were available.
-All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes.
-To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).
+To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature (MAT) and precipitation (MAP) from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).
+MAT and MAP were also extracted for all study locations.
+Potential evapotranspiration (PET) was extracted from the Global-PET database (http://www.cgiar-csi.org, \citet{zomer2008}), and the aridity index was calculated as PET/MAP.
 
-We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves, see \citet{falster2015} for details on the methods for each contributed study).
+We only used leaf mass per area (LMA) directly estimated for the harvested plants (typically for a subsample of leaves, see \citet{falster2015} for details on the methods for each contributed study).
 For conifers, leaf area was converted to half-total surface area using the average of a set of published conversion factors \citep{barclay2000}, with  different conversion factors applied to pines (\emph{Pinus} spp.) and non-pines.
@@
 
-Stem cross-sectional area was measured either at breast height and/or at the base of the plant.
-In all analyses, we used basal area because many plants (N=xxxx) were <1.3m tall.
+Stem cross-sectional area was measured either at breast height (1.3 - 1.34m) and/or at the base of the plant.
+In all analyses, we used basal stem area because many plants (N=\Sexpr{nrow(dataset[dataset$h.t < 1.3,])}) were \textless 1.3m tall.
 For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) by fitting the following equation, using non-linear regression.
@@
 
-To better understand the PFT and climate-related variation in leaf-stem biomass distribution, it is informative to decompose it into three component variables (Eq. 1),
+\emph{might be moved to introduction}
+To better understand the mechanisms of PFT and climate-related variation in leaf-stem biomass distribution, we decomposed it into three component variables (Eq.~\ref{eq:mfmspart}),
 
 \begin{equation}
+\label{eq:mfmspart}
 \frac{M_F}{M_S} = \frac{M_F}{A_F} \cdot \frac{A_F}{A_S} \cdot \frac{A_S}{M_S}
@@
 
-where $M_F/A_F$ is the leaf mass per area (LMA), $A_F/A_S$ the ratio of plant leaf area to basal stem area, and $A_S/M_S$ the ratio of stem basal area to total stem mass (including branches).
-Across PFTs, $M_F/A_F$ will thus be proportional to LMA when $A_F/A_S$ and $A_S/M_S$ do not vary between PFTs.
-We also expect that $M_F/M_S$ will show tremendous dependence on plant size \citep{poorter2015,reich2014}.
-We expect that most of this size-related variation is due to variation in $A_S/M_S$, since the total stem mass should be proportional to the plant height, stem cross-sectional area, and a shape parameter \citet{chave2005}.
+where $M_F/A_F$ is the leaf mass per area (LMA), $A_F/A_S$ the ratio of plant leaf area to basal stem area, and $A_S/M_S$ the ratio of stem basal area to total woody above-ground mass.
 Using variance partitioning, we quantify the contribution of plant height, plant functional type, and climate to the overall variation in $M_F/M_S$ and for all three component variables.
@@
 We compared this approach when using linear mixed-effects models instead of GAMs (calculating the the $R^2$ for linear mixed-effects models for the fixed effects only \citep{nakagawa2013}), as well as an independent method, independent effects analysis.
-The three methods generally agreed on the ranking of variable important (SI Methods).
+The three methods generally agreed on the ranking of variable importance (see Fig.
+S5 and S6).
 
@@
 GAMs were fitted using the \emph{mgcv} package \citep{wood2006}.
-The code to replicate this analysis (and all figures) is available at http://github.com/RemkoDuursma/baadanalysis (doi: \emph{to be added}).
+The code to replicate this analysis (and all figures) is available at \url{http://github.com/RemkoDuursma/baadanalysis} (doi: \emph{to be added}).
 
@@
 
-%with(baad_mapmat, tapply(MAT, pft, quantile, probs=c(0.25,0.75),na.rm=T))
-The compiled dataset covered the global climate space well (defined by mean annual temperature, MAT; and mean annual precipitation, MAP).
-There was some noticeable separation between the three major PFTs in terms of climate space occupancy.
-Evergreen angiosperms occurred on average at a higher MAT (50\% of studies between 15.5 and 26.0 \textdegree C MAT) than deciduous angiosperms (6.0 - 12.6 \textdegree C) and evergreen gymnosperms (3.4 - 12.4 \textdegree C).
+<<echo=FALSE>>=
+qp <- with(baad_mapmat, tapply(MAT, pft, function(x)round(quantile(x, probs=c(0.25,0.75),na.rm=TRUE),1)))
+@
+
+The compiled dataset covered the global climate space well (defined by mean annual temperature, MAT; and mean annual precipitation, MAP) (Fig.
+1).
+There was some separation between the three major PFTs in terms of climate space occupancy.
+Evergreen angiosperms occurred on average at a higher MAT (50\% of studies between \Sexpr{qp$EA[1]} and \Sexpr{qp$EA[2]} \textdegree C MAT) than deciduous angiosperms (\Sexpr{qp$DA[1]} - \Sexpr{qp$DA[2]} \textdegree C) and evergreen gymnosperms (\Sexpr{qp$EG[1]} - \Sexpr{qp$EG[2]} \textdegree C).
 Plant functional type and climate are thus at least partly confounded.
+
+<<echo=FALSE>>=
+mfmssmall <- round(with(dataset, 10^mean(lmlf_mst[h.t < 0.1], na.rm=TRUE)),1)
+mfmslarge <- round(with(dataset, 10^mean(lmlf_mst[h.t > 30], na.rm=TRUE)),3)
+@
 
@@
 As a result, the ratio $M_F/M_S$ quickly decreased with plant height (Fig.
-3a), from ca.
-1-2 for small saplings (< 0.1m) to less than 0.01 for large trees (>30m).
+3a), from an average of \Sexpr{mfmssmall} for very small saplings (\textless 0.1m) to an average of \Sexpr{mfmslarge} for large trees (\textgreater 30m).
 
@@
 3a).
-These differences were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm > evergreen angiosperm > deciduous angiosperm.
+These differences were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater evergreen angiosperm \textgreater deciduous angiosperm.
 Next we calculated $M_F/M_S$ at a common plant height for each of the PFTs.
-These three size-corrected $M_F/M_S$ values were highly significantly different between PFTs (p < 0.001), and were approximately proportional to the average LMA ($M_F/A_F$) across the PFTs (Fig.
+These three size-corrected $M_F/M_S$ values were highly significantly different between PFTs (p \textless 0.001), and were approximately proportional to the average LMA ($M_F/A_F$) across the PFTs (Fig.
 3b).
-As a result, the amount of leaf area per unit woody above-ground biomass ($A_F/M_S$) did not vary between PFTs, because the differences in $M_F/A_S$ were exactly compensated by differences in LMA (Fig.
-3c).
+As a result, the amount of leaf area per unit woody above-ground biomass ($A_F/M_S$) did not differ between PFTs (all pair-wise comparisons p \textgreater 0.05, see Fig.
+3c), because the differences in $M_F/A_S$ were exactly compensated by differences in LMA (see also Fig.
+S2 for $A_F/M_S$ as a function of plant height).
 
-Next, we decomposed $M_F/M_S$ into three component variables with distinct biological interpretation (Eq.
-3).
-For each of the component variables, ad well as $M_F/M_S$ and $A_F/M_S$, we estimated the variance explained by plant height, PFT and climate variables (MAT, MAP and its interaction) (Table 1).
+Next, we decomposed $M_F/M_S$ into three component variables with distinct biological interpretation (Eq.\ref{eq:mfmspart}).
+For each of the component variables and $M_F/M_S$, we estimated the variance explained by plant height, PFT and climate variables (MAT or aridity) (Table~\ref{table:varparttable}).
 We did this by sequentially adding terms to a GAM, and calculating the $R^2$ for each fitted model.
-The first model included only plant height, and already explained much variation (60 - 77\%) in $M_F/M_S$, $A_F/M_S$ and $M_S/A_S$ but not LMA and $A_F/A_S$ (0 - 7 \%).
-When PFT was added to the model (as an intercept term as well as interaction with the smooth H term), the $R^2$ increased substantially for $M_F/M_S$ and $A_F/A_S$ but not $A_F/M_S$ or $M_S/A_S$ (showing PFT invariance for these two variables).
-Additional variance explained was especially large for LMA, which is consistent with large differences shown in Fig.
+The first model included only plant height, and already explained much variation (65 - 79\%) in $M_F/M_S$ and $M_S/A_S$ but not LMA and $A_F/A_S$ (4 - 11 \%).
+When PFT was added to the model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $M_F/M_S$ and $M_F/A_F$ but not $A_F/A_S$ or $A_S/M_S$, demonstrating that PFT differences entered largely in leaf mass-based variables.
+Additional variance explained by adding PFT was especially large for LMA, which is consistent with the large differences between PFTs in LMA shown in Fig.
 3b,c.
 
-% afas
 From the variance partitioning (Table 1), it is evident that $A_F/A_S$ did not vary appreciably between PFTs.
@@
 4).
-We fit a GAM of $A_F$ as a function of $A_S$ with or without PFT as another covariate (both as intercept and slope interaction), which showed PFT significantly improved model fit (likelihood ratio test, p < 0.001), however the $R^2$ increased only from 83.8 to 85.3 \%, and thus had little if any practical significance.
+We fit a GAM of $A_F$ as a function of $A_S$ with or without PFT as another covariate (both as intercept and slope interaction), which showed PFT significantly improved model fit (likelihood ratio test, p \textless 0.001), however the $R^2$ increased only from \Sexpr{round(100*afas$R2a,2)} to \Sexpr{round(100*afas$R2b,2)}\%, and thus had little if any practical significance.
 
-Likewise, the ratio of stem basal area to stem mass ($A_S/M_S$) apparently varied little between PFTs (Table 1).
-To illustrate this further, when we plotted $M_S$ as a function of $A_S$ it revealed a remarkably tight relationship, with little apparent difference between PFTs (Fig 5a).
+Likewise, the ratio of stem basal area to stem mass ($A_S/M_S$) apparently varied little between PFTs (Table~\ref{table:varparttable}).
+To illustrate this further, a plot of $M_S$ as a function of $A_S$ revealed a tight relationship, with little apparent difference between PFTs (Fig 5a).
 The only exception was for large gymnosperm trees, which consistently showed a lower $M_S$ at a given $A_S$ (Fig.
@@
 We fitted a GAM of $M_S$ as a function of $A_S$ (added as a smooth term) with or without PFT as a covariate.
-Again PFT was formally significant (likelihood ratio test, p < 0.001), but the $R^2$ increased only marginally from 80.8 to 82.5\%, again showing little practical significance of PFT in explaining this variable.
+Again PFT was formally significant (likelihood ratio test, p \textless 0.001), but the $R^2$ increased only marginally from \Sexpr{round(100*msas$R2a,2)} to \Sexpr{round(100*msas$R2b,2)}\%, again showing little practical significance of PFT in explaining this variable.
 
@@
 6).
-There we expressed the stem mass ratio as $M_S/(H \cdot A_S)$ because we expect from basic geometry that $M_S$ is proportional to H and $A_S$ ($M_S = \phi \rho A_S H$, where $\phi$ a stem shape parameter, $\rho$ wood density).
+To correct for plant height effect on $M_S/A_S$, we expressed the stem mass ratio as $M_S/(H \cdot A_S)$ because we expect from basic geometry that $M_S$ is proportional to H and $A_S$ ($M_S = \phi \rho A_S H$, where $\phi$ a stem shape parameter, $\rho$ wood density).
 The histograms summarize that PFT effects are very evident in LMA (Fig.
-6a), not in $A_F/A_S$, and only a bit for $M_S/A_S$, which is entirely driven by large gymnosperm trees.
+6a), not in $A_F/A_S$, and only to some extent for $M_S/A_S$, which is entirely driven by large gymnosperm trees.
+However for all variables there was much unexplained variation between species (and individuals within species) within PFTs.
 
-Climate variables consistently added little explanatory power to the models when added to a model that included PFT and height (1 - 2.8 \%) (Table~\ref{table:varparttable}).
-Since this result is in contrast with recent finding of \citep{reich2014}, we also analyzed climate effects with methods similar to that study.
+Climate variables consistently added little explanatory power when added to a statistical model that included PFT and height (the adjusted R$^2$ frequently decreased, see Table~\ref{table:varparttable}).
+The only exception was LMA, for which a model that included MAT explained 53\% of the variation, compared to a model with height and PFT (45\%).
+Since the lack of climate effects on biomass distribution is in contrast with recent finding of \citep{reich2014}, we also analyzed climate effects with methods similar to that study.
 When we tested for MAT effects on $M_F/M_S$ when only considering gymnosperms and angiosperms (thus grouping decidous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig.
-S6) for angiosperms for most size classes (but never for gymnosperms).
+S5) for angiosperms for most size classes (but never for gymnosperms).
 We found, however, that this was largely due to the fact that at higher MAT, an increasing proportion of angiosperms are evergreen (Fig.
@@
 When analyzed separately for deciduous and evergreen angiosperms, few and weaker significant relationships were found (Fig.
-S7).
+S6).
 
@@
 Using the largest database of individual woody plant allometry to date, we studied patterns in above-ground biomass distribution among major woody plant functional types (PFTs), plant size, and climate.
+Our compilation is the first of its kind that also includes leaf area estimates on the same plants, allowing a direct test of the hypothesis that leaf and stem biomass distribution is related to the leaf mass per area.
 We found that at a common plant height, the leaf / woody biomass ratio ($M_F/M_S$) was proportional to the leaf mass per area across three PFTs, so that leaf area per unit above-ground biomass did not vary between PFTs.
-We found very weak effects of climate (mean annual temperature and aridity) on biomass distribution, the vast majority of the variation in biomass distribution was explained by plant size and PFT.
+We found very weak effects of climate (mean annual temperature and aridity) on biomass distribution, so that the vast majority of the variation in biomass distribution was explained by plant size and PFT.
 
@@
 This is consistent with recent results on an independently compiled large database of plant allometry \citep{poorter2015}, and further calls into question the generality of the often-assumed log-linear relationships between plant biomass pools and size metrics.
+
+We have confirmed the hypothesis that leaf area is less variable between PFTs than leaf mass.
+\citet{poorter2015} hypothesized that this is because forests should converge to a similar leaf area index (LAI), as total light interception at some point saturates with LAI.
+Although we found remarkably similar leaf area per unit above-ground woody biomass ($M_F/M_S$) between the three PFTs studied (Fig.
+S2), this does not in itself imply that LAI is equally similar.
+In a global compilation of LAI for undisturbed forest stands, \citet{iio2014} reported a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI.
+To scale our results to stand- and landscape scales requires more information on differences in stand structure between PFTs.
 
@@
 
-Although we found no consistent effects of climate on biomass distribution after PFT and plant height were accounted for, 
-effects of climate may be mediated by plant height.
-At a global scale, average plant height is strongly related to climate, likely mediated by water availability \citet{klein2015,simard2011}.
-Since biomass distribution is strongly related to plant height (Figs.
-2, 3), we do expect correlations between climate variables and observed biomass distribution.
+Although we found no consistent effects of climate on biomass distribution after PFT and plant height were accounted for, effects of climate are likely mediated via plant height.
+At a global scale, average plant height is strongly related to climate, possibly via water availability \citet{klein2015,simard2011}.
+Since biomass distribution itself is steeply related to plant height (Figs.
+2, 3), we expect correlations between climate variables and observed biomass distribution when plant height is not accounted for.
+Our database did not lend itself well to a direct test of this hypothesis, due to uneven sampling across the range in plant height along climate gradients.
 
-Still lots and lots of variation to explain within PFTs; we do not argue for convergence, but rather that as a first step, we can understand differences between PFTs.
-This helps parameterize PFTs, and also perhaps this can help to place plants on a continuous scale rather than rely on fixed categories.
-This is an active area of research, but is yet to be extended to patterns in biomass distribution.
-Our work contributes a bit to this.
+We stress that although our results show some convergence between PFTs in leaf area relationships, we do not dismiss the very large residual variation that is yet to be explained and understood (Fig.
+6).
+Our purpose was to test hypotheses related to differences between PFTs, plant height, and climate, yet clearly more work can be done to understand the variation between species and sites witin groups.
+Some of the residual variation probably has to do with site fertility (CITES), or local differences in water availability, perhaps mediated by seasonality, that were not picked up by our coarse climate variables.
 
@@
 
-Stem mass index: Chave biomass estimation.
-We find support for this, with MS/AS nearly proportional to height.
-Conifers diverged, for the subset of larger trees only.
-Not sure why these differences don't show up for small plants but variance was very high!
-For tropical forests, \citet{chave2005} showed that aboveground biomass can be estimated as,
-
-$$M_T = F \rho A_S H_T$$
-
-where $F$ is a form factor, and $\rho$ wood density (kg m$^3$). This explains why (for larger trees) at a given AS, conifers had lower stem mass (Fig. 5).
-
-\emph{combine next two in shorter GVM paragraph}
-A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions.
-Based on our findings, a first approximation within GVMs would be to assume leaf area to stem cross-sectional area a parameter that does not vary between PFTs (notwithstanding effects of plant height on this variable), and vary biomass allocation accordingly.
-LMA, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem area.
-Some models already incorporate a similar algorithm (see Notes S1), while other algorithms may be tuned to yield similar patterns between PFTs as we have presented.
-In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data.
-
-Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation.
-While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots).
-The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010,wullschleger2014}, assumed to differ in key physiological attributes.
 While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are poorly constrained, due to limited available data.
 Moreover, there is little consensus on how biomass partitioning and allocation (see Methods for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
-These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared \citep{wolf2011}.
-Rather than to advance a specific model of biomass partitioning or allocation, our view is that a broader evidence base is needed first to elucidate patterns between PFTs.
-This should enable those building GVMs to refine algorithms and parameter values to more closely match the observations.
+A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass distribution instead of using constant allocation fractions.
+Our results can be used to further constrain biomass allocation routines in GVM, in particular, the finding that $M_F/M_S$ (at a given plant height) is approximately proportional to LMA can be tested against model predictions.
+In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data.
 
@@
 print(xtable(table_samplesize,
-             label="counttable",
+             label="table:counttable",
              caption="Sample sizes used for the four studied whole-plant variables, by plant functional type.
@@
 print(xtable(table_varpart_gam,
-             label="varparttable",
+             label="table:varparttable",
              caption="Explained variance in four plant biomass variables by plant height (H), plant functional type (PFT) and climate.

### 2016-02-03 R/manuscript_functions.R

@@ -15,10 +15,9 @@ get_wordcount <- function(fn){
     # for the section. So more work to make this general
     
     l <- list()
-    l$Abstract <- count("Abstract")
+    l$Abstract <- count("Summary")
     l$Introduction <- count("Introduction")
-    l$Methods <- count("Terminology") + count("Data")  + 
-                 count("Data analysis")
+    l$Methods <- count("Data") + count("Data analysis")
     l$Results <- count("Results")
     l$Discussion <- count("Discussion")
     

### 2016-02-04 manuscript.Rnw
new short paragraph in discussion; change of title; edit key paragraph in introduction

@@
 
-\title{Leaf traits and size drive differences in above-ground biomass distribution among woody plant functional types}
+\title{Leaf mass per area and plant size drive differences in above-ground biomass distribution among woody plant functional types}
 
@@
 \item The partitioning of biomass into leaves and stems is an important determinant of the residence time of carbon in vegetation.
-We tested for differences in biomass distribution between major woody plant functional types (PFT), and sought to understand these differences in terms of leaf traits.
+We tested the hypothesis that biomass distribution in leaves and stems is related to the leaf mass per area, an important leaf trait.
 \item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species.
@@
 One puzzling aspect of the large inter-specific variation in biomass distribution between leaves and stems is the much larger leaf mass in gymnosperms compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012, poorter2015}.
-\citet{poorter2012} and \citet{enquistniklas2002} attributed higher leaf-stem biomass ratio ($M_F/M_S$) in gymnosperms to longer leaf lifespan (LL) compared to typical angiosperms.
-This is consistent with \citet{poorter2015} who reported lower $M_F/M_S$ in deciduous compared to evergreen species.
-
-It is yet unknown whether plant functional types (PFTs) (angiosperm vs.
-gymnosperm, deciduous vs.
-evergreen) with higher leaf biomass also have higher total leaf area, which is relevant because leaf area drives total light interception and thus productivity.
-Because longer LL is generally correlated with the leaf mass per area (LMA), PFTs with higher MF/MS (and thus higher LMA) do not necessarily not have larger total leaf area at a given stem mass.
+One possible explanation is that gymnosperms compensate for their low leaf mass per area (LMA) \citep{poorter2009} by maintaining a large amount of leaf biomass.
+If this is the case, total leaf area per unit above-ground biomass would be less variable between angiosperms and gymnosperms than leaf mass.
 Some oft-cited studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}, but recent large compilations on biomass distribution did not include measurements of leaf area in addition to mass to test this hypothesis \citep{poorter2012,poorter2015,reich2014}.
-\citet{poorter2015} advanced the hypothesis that $MF/MS$ may be nearly proportional to LMA, arguing that total leaf area is likely less variable between PFTs.
-Here we set out to directly test this hypothesis with a database that includes observations of leaf area as well as leaf mass measured on the same plants.
+The idea that high total leaf mass may compensate for low LMA would also explain the difference between evergreen (high LMA, high leaf mass) and deciduous (low LMA, low leaf mass) species \citep{poorter2015}.
+Here we set out to directly test the hypothesis that across PFTs, $M_F/M_S$ is related to LMA with a global database that includes observations of leaf area as well as leaf mass measured on the same plants.
 
@@
 MAT and MAP were also extracted for all study locations.
-Potential evapotranspiration (PET) was extracted from the Global-PET database (http://www.cgiar-csi.org, \citet{zomer2008}), and the aridity index was calculated as PET/MAP.
+Potential evapotranspiration (PET) was extracted from the Global-PET database (http://www.cgiar-csi.org, \citet{zomer2008}), and an aridity index was calculated as PET/MAP.
 
@@
 From the variance partitioning (Table 1), it is evident that $A_F/A_S$ did not vary appreciably between PFTs.
-Indeed, when all PFTs were plotted together, similar $A_F$ was attained at a given $A_S$ across the entire plant size range (with the exception of extremely large trees with > 1 $m^2$ stem basal area, for which little data was available) (Fig.
+Indeed, when all PFTs were plotted together, similar $A_F$ was attained at a given $A_S$ across the entire plant size range (with the exception of extremely large trees with \textgreater 1 $m^2$ stem basal area, for which comparatively little data was available) (Fig.
 4).
@@
 In a global compilation of LAI for undisturbed forest stands, \citet{iio2014} reported a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI.
-To scale our results to stand- and landscape scales requires more information on differences in stand structure between PFTs.
+To scale our results to stand- and landscape scales thus requires more information on differences in stand structure between PFTs.
+
+Previous interpretations of differences in $M_F/M_S$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002,poorter2012}.
+Indeed, if all else equal, longer LL in gymnosperms would result in higher total leaf mass.
+It is however difficult to separate cause from effect in the relationships between LL, LMA, and $M_F/M_S$, particularly because LL and LMA are correlated across species and PFTs in a nearly proportional way \citet{wright2004}.
 
@@
 Although we found no consistent effects of climate on biomass distribution after PFT and plant height were accounted for, effects of climate are likely mediated via plant height.
-At a global scale, average plant height is strongly related to climate, possibly via water availability \citet{klein2015,simard2011}.
+At a global scale, average plant height is clearly related to climate via water availability \citet{klein2015,simard2011}.
 Since biomass distribution itself is steeply related to plant height (Figs.
-2, 3), we expect correlations between climate variables and observed biomass distribution when plant height is not accounted for.
+2, 3), we expect correlations between climate variables and observed biomass distribution when plant height is not first accounted for.
 Our database did not lend itself well to a direct test of this hypothesis, due to uneven sampling across the range in plant height along climate gradients.
@@
 
-Leaf area was just about proportional to AS, except for the largest trees.
-This decreases because of the formation of heartwood (and AF/sapwood may decrease a little with height, but AF/AS more because of crown rise).
-Note that we have many small plants in the dataset as well, and within individual datasets it is more likely to pick up small-scale effects like these.
-
-While GVMs assume or predict differences in biomass partitioning between PFTs (Notes S1), these differences are poorly constrained, due to limited available data.
-Moreover, there is little consensus on how biomass partitioning and allocation (see Methods for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
+Global vegetation models assume or predict differences in biomass distribution between PFTs.
+However, these differences are currently poorly constrained, due to limited available empirical data.
+Moreover, there is little consensus on how biomass distribution (and allocation) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}.
 A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass distribution instead of using constant allocation fractions.
-Our results can be used to further constrain biomass allocation routines in GVM, in particular, the finding that $M_F/M_S$ (at a given plant height) is approximately proportional to LMA can be tested against model predictions.
-In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data.
+Our results can be used to further constrain biomass allocation routines in GVM, in particular, the finding that $M_F/M_S$ (at a given plant height) is approximately proportional to LMA can be tested against model predictions, or used directly to parameterize current algorithms.
+In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015,poorter2015,reich2014} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data.
 
@@
 Thanks also to Rich FitzJohn for advice about using the remake package.
-Figure 2a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary.
-Thanks to a small army of anonymous reviewers for suggestions that greatly improved the paper.
+Figure 2a uses an image by Lana Heydon, IAN Image Library (http://ian.umces.edu/imagelibrary).
+Thanks to anonymous reviewers for suggestions that greatly improved the paper.
 Special thanks to Steve Zissou.

### 2016-02-04 manuscript.Rnw
caption fix
More edits (intro) and references added

@@
 \begin{itemize}
-\item The partitioning of biomass into leaves and stems is an important determinant of the residence time of carbon in vegetation.
-We tested the hypothesis that biomass distribution in leaves and stems is related to the leaf mass per area, an important leaf trait.
+\item We tested the hypothesis that biomass distribution in leaves and stems is related to the leaf mass per area, an important leaf trait, across major woody plant functional types (PFTs) (as defined by deciduous vs.
+evergreen and gymnosperm vs.
+angiosperm).
 \item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species.
 We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, and climate on biomass distribution and three component variables.
-\item The ratio of leaf mass to woody above-ground mass ($M_F/A_S$) varied strongly between PFTs (as defined by deciduous vs.
-evergreen and gymnosperm vs.
-angiosperm).
-We found that $M_F/A_S$, once corrected for plant height, was proportional to leaf mass per area across PFTs.
+\item The ratio of leaf mass to woody above-ground mass ($M_F/M_S$) varied strongly between PFTs.
+We found that $M_F/M_S$, once corrected for plant height, was proportional to leaf mass per area across PFTs.
 As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass.
-Climate consistently explained very little additional variation, demonstrating very weak control of climate on biomass distribution, at a given plant size.
+Climate consistently explained very little additional variation in biomass distribution at a given plant size.
 \item Combined, these results demonstrate consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain global vegetation models.
@@
 
-The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation \citep{ise2010, dekauwe2014,friend2014}.
+The distribution of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation \citep{ise2010, dekauwe2014,friend2014}.
 Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \citep{ryanlaw2005}.
 Globally, forests store approximately 360Pg of carbon in living biomass \citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}.
-Reducing uncertainties about biomass partitioning and carbon residence times is a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014,negron2015}.
+Reducing uncertainties about biomass distributionand carbon residence times is a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014,negron2015}, in particular because projections by global vegetation models (GVMs) are particularly sensitive to this component \citep{friend2014,ise2010}.
 
-One puzzling aspect of the large inter-specific variation in biomass distribution between leaves and stems is the much larger leaf mass in gymnosperms compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012, poorter2015}.
-One possible explanation is that gymnosperms compensate for their low leaf mass per area (LMA) \citep{poorter2009} by maintaining a large amount of leaf biomass.
+Perhaps the biggest challenge is to capture the combined responses to environmental change of the more than 250,000 plant species comprising the world's vegetation.
+While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of biomass invested in different tissues (leaves, stems, roots).
+To attempt to understand the source of this functional diversity, it is customary to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010,wullschleger2014}, assumed to differ in key attributes.
+Previous work has shown consistently large differences in biomass distribution and related traits among major PFTs \citep{poorter2009,enquistniklas2002,poorter2012,poorter2015,reich2014,mccarthyetal2007}, but these differences are yet to be fully understood.
+
+One puzzling aspect of the large variation in biomass distribution among PFTs is the much larger leaf mass in gymnosperms compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012, poorter2015}.
+A possible explanation is that gymnosperms compensate for their low leaf mass per area (LMA) \citep{poorter2009} by maintaining a large amount of leaf biomass, to maintain sufficient total leaf area.
 If this is the case, total leaf area per unit above-ground biomass would be less variable between angiosperms and gymnosperms than leaf mass.
-Some oft-cited studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}, but recent large compilations on biomass distribution did not include measurements of leaf area in addition to mass to test this hypothesis \citep{poorter2012,poorter2015,reich2014}.
+Some oft-cited studies have however assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}, but recent large compilations on biomass distribution did not include measurements of leaf area in addition to mass to test this hypothesis \citep{poorter2012,poorter2015,reich2014}.
 The idea that high total leaf mass may compensate for low LMA would also explain the difference between evergreen (high LMA, high leaf mass) and deciduous (low LMA, low leaf mass) species \citep{poorter2015}.
@@
 
-We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations).
+We included only field-grown woody plants in our analysis (including natural vegetation, unmanaged and managed plantations), excluding glasshouse and common garden studies.
 Excluding plantations from the analysis did not affect the main result, but far fewer data were available for evergreen gymnosperms.
@@
 \begin{equation}
-D_S = D_{Sbh} H^a / (H - H_{bh})^a
+D_S = D_{Sbh} H^c / (H - H_{bh})^c
 \end{equation}
 
-where $D_S$ is the basal stem diameter (m), $D_{Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{Sbh}$ was measured (typically 1.3 or 1.34m), and $a$ was further expressed as a function of plant height:
+where $D_S$ is the basal stem diameter (m), $D_{Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{Sbh}$ was measured (typically 1.3 or 1.34m), and \emph{c} was further expressed as a function of plant height:
 
 \begin{equation}
-a = c_0H^{c_1}
+c = c_0H^{c_1}
 \end{equation}
@@
 In all fitted GAMs, we used species-dataset combination as a random effect.
-All variables (except MAP and MAT) were log-transformed prior to analysis.
+To homogenize variance, all variables (except MAP and MAT) were log-transformed prior to analysis.
 
@@
 Indeed, if all else equal, longer LL in gymnosperms would result in higher total leaf mass.
+Further supporting this explanation, \citet{reich1992} showed a positive correlation between LL and total stand leaf biomass across diverse forest stands.
 It is however difficult to separate cause from effect in the relationships between LL, LMA, and $M_F/M_S$, particularly because LL and LMA are correlated across species and PFTs in a nearly proportional way \citet{wright2004}.
@@
 
-We stress that although our results show some convergence between PFTs in leaf area relationships, we do not dismiss the very large residual variation that is yet to be explained and understood (Fig.
+We stress that although our results show some convergence between PFTs in leaf area relationships, we do not dismiss the very large residual variation that is yet to be explained (Fig.
 6).
-Our purpose was to test hypotheses related to differences between PFTs, plant height, and climate, yet clearly more work can be done to understand the variation between species and sites witin groups.
-Some of the residual variation probably has to do with site fertility (CITES), or local differences in water availability, perhaps mediated by seasonality, that were not picked up by our coarse climate variables.
+Our purpose was to test hypotheses related to first-order differences between PFTs, plant height, and climate, yet clearly more work can be done to understand the variation between species and sites witin groups.
+Some of the residual variation is likely explained by site fertility \citet{vanninen2005}, differences in stand density \citet{ilomaki2003}, social status within stands \citet{makela2006}, or local differences in water availability (possibly mediated by seasonality at a given MAP).
+Future work will focus on smaller-scale explanatory variables such as the ones mentioned here.
 
@@
 \clearpage
-\section{Figures}
+\section*{Figures}
 
@@
 Lines are generalised additive model fits.
-(b) and (c) $M_F/M_S$ and leaf area ratio ($A_F/M_S$) at the average plant height in the dataset, estimated from panel (a).
+(b) and (c) $M_F/M_S$ and leaf area ratio ($A_F/M_S$) at the average plant height in the dataset, estimated from fitted models in panel (a).
 Error bars are 95\% confidence intervals.

### 2016-02-24 manuscript.Rnw
bloom et al 2016
Clean trailing white space
Epic variable substitution
Define macros for common math terms

@@
 \usepackage[labelfont=bf, labelsep=quad]{caption}
-
 \renewcommand*{\thefootnote}{\fnsymbol{footnote}}
+\newcommand{\mf}{M_{\rm F}}
+\newcommand{\ms}{M_{\rm S}}
+\newcommand{\as}{A_{\rm S}}
+\newcommand{\af}{A_{\rm F}}
+\newcommand{\asb}{A_{\rm Sbh}}
+\newcommand{\mlms}{\mf/\ms}
+\newcommand{\asms}{\as/\ms}
+\newcommand{\afas}{\af/\as}
+\newcommand{\mfaf}{\mf/\af}
+\newcommand{\afms}{\af/\ms}
+\newcommand{\mfas}{\mf/\as}
+\newcommand{\msas}{\ms/\as}
+\newcommand{\mshas}{\ms/(H \cdot \as)}
 
@@
 We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, and climate on biomass distribution and three component variables.
-\item The ratio of leaf mass to woody above-ground mass ($M_F/M_S$) varied strongly between PFTs.
-We found that $M_F/M_S$, once corrected for plant height, was proportional to leaf mass per area across PFTs.
+\item The ratio of leaf mass to woody above-ground mass ($\mlms$) varied strongly between PFTs.
+We found that $\mlms$, once corrected for plant height, was proportional to leaf mass per area across PFTs.
 As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass.
@@
 
-We included only field-grown woody plants in our analysis (including natural vegetation, unmanaged and managed plantations), excluding glasshouse and common garden studies.
+We included only field-grown woody plants in our analysis (including natural vegetation, unmanaged, and managed plantations), excluding glasshouse and common garden studies.
 Excluding plantations from the analysis did not affect the main result, but far fewer data were available for evergreen gymnosperms.
 We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms.
-We excluded  deciduous gymnosperms because few data were available.
+We excluded  deciduous gymnosperms because insufficient data were available.
 To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature (MAT) and precipitation (MAP) from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).
@@
 In all analyses, we used basal stem area because many plants (N=\Sexpr{nrow(dataset[dataset$h.t < 1.3,])}) were \textless 1.3m tall.
-For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) by fitting the following equation, using non-linear regression.
+For the subset of the data where both were measured, we estimated basal stem area ($\as$) from breast height stem area ($\asb$) by fitting the following equation, using non-linear regression.
 
 \begin{equation}
-D_S = D_{Sbh} H^c / (H - H_{bh})^c
+D_{\rm S} = D_{\rm Sbh} H^c / (H - H_{\rm bh})^c
 \end{equation}
 
-where $D_S$ is the basal stem diameter (m), $D_{Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{Sbh}$ was measured (typically 1.3 or 1.34m), and \emph{c} was further expressed as a function of plant height:
+where $D_{\rm S}$ is the basal stem diameter (m), $D_{\rm Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{\rm Sbh}$ was measured (typically 1.3 or 1.34m), and \emph{c} was further expressed as a function of plant height:
 
@@
 \label{eq:mfmspart}
-\frac{M_F}{M_S} = \frac{M_F}{A_F} \cdot \frac{A_F}{A_S} \cdot \frac{A_S}{M_S}
+\frac{\mf}{\ms} = \frac{\mf}{\af} \cdot \frac{\af}{\as} \cdot \frac{\as}{\ms}
 \end{equation}
 
-where $M_F/A_F$ is the leaf mass per area (LMA), $A_F/A_S$ the ratio of plant leaf area to basal stem area, and $A_S/M_S$ the ratio of stem basal area to total woody above-ground mass.
-Using variance partitioning, we quantify the contribution of plant height, plant functional type, and climate to the overall variation in $M_F/M_S$ and for all three component variables.
+where $\mfaf$ is the leaf mass per area (LMA), $\afas$ the ratio of plant leaf area to basal stem area, and $\asms$ the ratio of stem basal area to total woody above-ground mass.
+Using variance partitioning, we quantify the contribution of plant height, plant functional type, and climate to the overall variation in $\mlms$ and for all three component variables.
 
@@
 
-As expected, both whole-plant leaf mass ($M_F$) and above-ground woody biomass ($M_S$) showed a strong depence on plant height (Fig.
+As expected, both whole-plant leaf mass ($\mf$) and above-ground woody biomass ($\ms$) showed a strong depence on plant height (Fig.
 2).
 For both variables the relationship was highly non-linear on a log-log scale, demonstrating departure from simple allometric relationships.
-It is also evident that $M_S$ increases much quicker with plant height than $M_F$.
-As a result, the ratio $M_F/M_S$ quickly decreased with plant height (Fig.
+It is also evident that $\ms$ increases much quicker with plant height than $\mf$.
+As a result, the ratio $\mlms$ quickly decreased with plant height (Fig.
 3a), from an average of \Sexpr{mfmssmall} for very small saplings (\textless 0.1m) to an average of \Sexpr{mfmslarge} for large trees (\textgreater 30m).
@@
 These differences were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater evergreen angiosperm \textgreater deciduous angiosperm.
-Next we calculated $M_F/M_S$ at a common plant height for each of the PFTs.
-These three size-corrected $M_F/M_S$ values were highly significantly different between PFTs (p \textless 0.001), and were approximately proportional to the average LMA ($M_F/A_F$) across the PFTs (Fig.
+Next we calculated $\mlms$ at a common plant height for each of the PFTs.
+These three size-corrected $\mlms$ values were highly significantly different between PFTs (p \textless 0.001), and were approximately proportional to the average LMA ($\mfaf$) across the PFTs (Fig.
 3b).
-As a result, the amount of leaf area per unit woody above-ground biomass ($A_F/M_S$) did not differ between PFTs (all pair-wise comparisons p \textgreater 0.05, see Fig.
-3c), because the differences in $M_F/A_S$ were exactly compensated by differences in LMA (see also Fig.
-S2 for $A_F/M_S$ as a function of plant height).
+As a result, the amount of leaf area per unit woody above-ground biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater 0.05, see Fig.
+3c), because the differences in $\mfas$ were exactly compensated by differences in LMA (see also Fig.
+S2 for $\afms$ as a function of plant height).
 
-Next, we decomposed $M_F/M_S$ into three component variables with distinct biological interpretation (Eq.\ref{eq:mfmspart}).
-For each of the component variables and $M_F/M_S$, we estimated the variance explained by plant height, PFT and climate variables (MAT or aridity) (Table~\ref{table:varparttable}).
+Next, we decomposed $\mlms$ into three component variables with distinct biological interpretation (Eq.\ref{eq:mfmspart}).
+For each of the component variables and $\mlms$, we estimated the variance explained by plant height, PFT, and climate variables (MAT or aridity) (Table~\ref{table:varparttable}).
 We did this by sequentially adding terms to a GAM, and calculating the $R^2$ for each fitted model.
-The first model included only plant height, and already explained much variation (65 - 79\%) in $M_F/M_S$ and $M_S/A_S$ but not LMA and $A_F/A_S$ (4 - 11 \%).
-When PFT was added to the model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $M_F/M_S$ and $M_F/A_F$ but not $A_F/A_S$ or $A_S/M_S$, demonstrating that PFT differences entered largely in leaf mass-based variables.
+The first model included only plant height, and already explained much variation (65 - 79\%) in $\mlms$ and $\msas$ but not LMA and $\afas$ (4 - 11 \%).
+When PFT was added to the model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $\mlms$ and $\mfaf$ but not $\afas$ or $\asms$, demonstrating that PFT differences entered largely in leaf mass-based variables.
 Additional variance explained by adding PFT was especially large for LMA, which is consistent with the large differences between PFTs in LMA shown in Fig.
@@
 
-From the variance partitioning (Table 1), it is evident that $A_F/A_S$ did not vary appreciably between PFTs.
-Indeed, when all PFTs were plotted together, similar $A_F$ was attained at a given $A_S$ across the entire plant size range (with the exception of extremely large trees with \textgreater 1 $m^2$ stem basal area, for which comparatively little data was available) (Fig.
+From the variance partitioning (Table 1), it is evident that $\afas$ did not vary appreciably between PFTs.
+Indeed, when all PFTs were plotted together, similar $\af$ was attained at a given $\as$ across the entire plant size range (with the exception of extremely large trees with \textgreater 1 $m^2$ stem basal area, for which comparatively little data was available) (Fig.
 4).
-We fit a GAM of $A_F$ as a function of $A_S$ with or without PFT as another covariate (both as intercept and slope interaction), which showed PFT significantly improved model fit (likelihood ratio test, p \textless 0.001), however the $R^2$ increased only from \Sexpr{round(100*afas$R2a,2)} to \Sexpr{round(100*afas$R2b,2)}\%, and thus had little if any practical significance.
+We fit a GAM of $\af$ as a function of $\as$ with or without PFT as another covariate (both as intercept and slope interaction), which showed PFT significantly improved model fit (likelihood ratio test, p \textless 0.001), however the $R^2$ increased only from \Sexpr{round(100*afas$R2a,2)} to \Sexpr{round(100*afas$R2b,2)}\%, and thus had little if any practical significance.
 
-Likewise, the ratio of stem basal area to stem mass ($A_S/M_S$) apparently varied little between PFTs (Table~\ref{table:varparttable}).
-To illustrate this further, a plot of $M_S$ as a function of $A_S$ revealed a tight relationship, with little apparent difference between PFTs (Fig 5a).
-The only exception was for large gymnosperm trees, which consistently showed a lower $M_S$ at a given $A_S$ (Fig.
+Likewise, the ratio of stem basal area to stem mass ($\asms$) apparently varied little between PFTs (Table~\ref{table:varparttable}).
+To illustrate this further, a plot of $\ms$ as a function of $\as$ revealed a tight relationship, with little apparent difference between PFTs (Fig 5a).
+The only exception was for large gymnosperm trees, which consistently showed a lower $\ms$ at a given $\as$ (Fig.
 5b).
-We fitted a GAM of $M_S$ as a function of $A_S$ (added as a smooth term) with or without PFT as a covariate.
+We fitted a GAM of $\ms$ as a function of $\as$ (added as a smooth term) with or without PFT as a covariate.
 Again PFT was formally significant (likelihood ratio test, p \textless 0.001), but the $R^2$ increased only marginally from \Sexpr{round(100*msas$R2a,2)} to \Sexpr{round(100*msas$R2b,2)}\%, again showing little practical significance of PFT in explaining this variable.
 
-Taken together, these results demonstrate that 1) the majority of PFT-related variation in $M_F/M_S$ results from variation in LMA (not $A_F/A_S$ or $A_S/M_S$), and 2) the majority of height-related variation in $M_F/M_S$ results from variation in $A_S/M_S$ (not LMA or $A_F/A_S$).
+Taken together, these results demonstrate that 1) the majority of PFT-related variation in $\mlms$ results from variation in LMA (not $\afas$ or $\asms$), and 2) the majority of height-related variation in $\mlms$ results from variation in $\asms$ (not LMA or $\afas$).
 It is however important to note that there is very large variation between individuals even after accounting for PFT (Fig.
 6).
-To correct for plant height effect on $M_S/A_S$, we expressed the stem mass ratio as $M_S/(H \cdot A_S)$ because we expect from basic geometry that $M_S$ is proportional to H and $A_S$ ($M_S = \phi \rho A_S H$, where $\phi$ a stem shape parameter, $\rho$ wood density).
+To correct for plant height effect on $\msas$, we expressed the stem mass ratio as $\mshas$ because we expect from basic geometry that $\ms$ is proportional to H and $\as$ ($\ms = \phi \rho \as H$, where $\phi$ a stem shape parameter, $\rho$ wood density).
 The histograms summarize that PFT effects are very evident in LMA (Fig.
-6a), not in $A_F/A_S$, and only to some extent for $M_S/A_S$, which is entirely driven by large gymnosperm trees.
+6a), not in $\afas$, and only to some extent for $\msas$, which is entirely driven by large gymnosperm trees.
 However for all variables there was much unexplained variation between species (and individuals within species) within PFTs.
@@
 Since the lack of climate effects on biomass distribution is in contrast with recent finding of \citep{reich2014}, we also analyzed climate effects with methods similar to that study.
-When we tested for MAT effects on $M_F/M_S$ when only considering gymnosperms and angiosperms (thus grouping decidous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig.
+When we tested for MAT effects on $\mlms$ when only considering gymnosperms and angiosperms (thus grouping decidous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig.
 S5) for angiosperms for most size classes (but never for gymnosperms).
 We found, however, that this was largely due to the fact that at higher MAT, an increasing proportion of angiosperms are evergreen (Fig.
-1), which have higher $M_F/M_S$.
+1), which have higher $\mlms$.
 When analyzed separately for deciduous and evergreen angiosperms, few and weaker significant relationships were found (Fig.
@@
 
+Read \citep{bloom2016}; global patterns in inferred allocation and biomass pools (leaf / stem).
+
 Using the largest database of individual woody plant allometry to date, we studied patterns in above-ground biomass distribution among major woody plant functional types (PFTs), plant size, and climate.
 Our compilation is the first of its kind that also includes leaf area estimates on the same plants, allowing a direct test of the hypothesis that leaf and stem biomass distribution is related to the leaf mass per area.
-We found that at a common plant height, the leaf / woody biomass ratio ($M_F/M_S$) was proportional to the leaf mass per area across three PFTs, so that leaf area per unit above-ground biomass did not vary between PFTs.
+We found that at a common plant height, the leaf / woody biomass ratio ($\mlms$) was proportional to the leaf mass per area across three PFTs, so that leaf area per unit above-ground biomass did not vary between PFTs.
 We found very weak effects of climate (mean annual temperature and aridity) on biomass distribution, so that the vast majority of the variation in biomass distribution was explained by plant size and PFT.
 
-As expected, $M_F/M_S$ steeply declined with plant height.
+As expected, $\mlms$ steeply declined with plant height.
 It is well known that it is necessary to correct for plant size when comparing biomass distribution among groups \citep{mcconnaughay1999}.
 We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship.
-This was useful because both $M_F$ and $M_S$ showed very non-linear patterns with plant height (Fig.
+This was useful because both $\mf$ and $\ms$ showed very non-linear patterns with plant height (Fig.
 2), or among biomass pools (Fig.
@@
 \citet{poorter2015} hypothesized that this is because forests should converge to a similar leaf area index (LAI), as total light interception at some point saturates with LAI.
-Although we found remarkably similar leaf area per unit above-ground woody biomass ($M_F/M_S$) between the three PFTs studied (Fig.
+Although we found remarkably similar leaf area per unit above-ground woody biomass ($\mlms$) between the three PFTs studied (Fig.
 S2), this does not in itself imply that LAI is equally similar.
@@
 
-Previous interpretations of differences in $M_F/M_S$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002,poorter2012}.
+Previous interpretations of differences in $\mlms$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002,poorter2012}.
 Indeed, if all else equal, longer LL in gymnosperms would result in higher total leaf mass.
 Further supporting this explanation, \citet{reich1992} showed a positive correlation between LL and total stand leaf biomass across diverse forest stands.
-It is however difficult to separate cause from effect in the relationships between LL, LMA, and $M_F/M_S$, particularly because LL and LMA are correlated across species and PFTs in a nearly proportional way \citet{wright2004}.
+It is however difficult to separate cause from effect in the relationships between LL, LMA, and $\mlms$, particularly because LL and LMA are correlated across species and PFTs in a nearly proportional way \citet{wright2004}.
 
@@
 Climate was represented as either mean annual temperature or an aridity index (PET/MAP), however, results were similarly inconclusive when studying other climate variables such as composite drought indices, growing season length, or mean annual precipitation (results not shown).
-This result appears to contradict \citet{reich2014} who reported consistent increases in foliage biomass ($M_F$) with MAT, when corrected for above-ground woody biomass ($M_S$).
-We were able to partly reconcile our findings with those of \citep{reich2014}: when we considered only angiosperms and gymnosperms (thus grouping deciduous and evergreens, as did Reich et al.) $M_F/M_S$ increased consistently with MAT for most size classes (Fig.
+This result appears to contradict \citet{reich2014} who reported consistent increases in foliage biomass ($\mf$) with MAT, when corrected for above-ground woody biomass ($\ms$).
+We were able to partly reconcile our findings with those of \citep{reich2014}: when we considered only angiosperms and gymnosperms (thus grouping deciduous and evergreens, as did Reich et al.) $\mlms$ increased consistently with MAT for most size classes (Fig.
 S5), but not for gymnosperms.
@@
 1), which, as a result of higher LMA (Fig.
-2) also have a higher $M_F/M_S$.
+2) also have a higher $\mlms$.
 However, based on a re-analysis of data in Reich et al.
-2014 (not shown), this shift in PFT with climate does not explain their observation of an increase in $M_F/M_S$ with MAT, as this pattern still held within gymnosperms and deciduous angiosperms.
+2014 (not shown), this shift in PFT with climate does not explain their observation of an increase in $\mlms$ with MAT, as this pattern still held within gymnosperms and deciduous angiosperms.
 
@@
 A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass distribution instead of using constant allocation fractions.
-Our results can be used to further constrain biomass allocation routines in GVM, in particular, the finding that $M_F/M_S$ (at a given plant height) is approximately proportional to LMA can be tested against model predictions, or used directly to parameterize current algorithms.
+Our results can be used to further constrain biomass allocation routines in GVM, in particular, the finding that $\mlms$ (at a given plant height) is approximately proportional to LMA can be tested against model predictions, or used directly to parameterize current algorithms.
 In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015,poorter2015,reich2014} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data.
@@
 Climate variables included were either MAT or aridity index (calculated as the of PET to MAP).
-P-values not shown because all smooth terms were always highly significant (p < 0.01), with the exception of aridity index in the model for $M_F/M_S$ (p = 0.057)."),sanitize.text.function=function(x){x},
+P-values not shown because all smooth terms were always highly significant (p < 0.01), with the exception of aridity index in the model for $\\mlms$ (p = 0.057)."),sanitize.text.function=function(x){x},
       include.rownames=FALSE, caption.placement = "top")
@@
     \fi
-    \caption{Raw data for leaf biomass ($M_F$) and above-ground woody biomass ($M_S$) for each of the PFTs, as a function of total plant height ($H$).
+    \caption{Raw data for leaf biomass ($\mf$) and above-ground woody biomass ($\ms$) for each of the PFTs, as a function of total plant height ($H$).
 Each point is an individual plant.
@@
     \caption{Dominant woody PFTs differ in above-ground biomass distribution due to underlying differences in leaf mass per area.
-(a) Leaf mass per unit woody above-ground biomass ($M_F/M_S$) by PFT.
+(a) Leaf mass per unit woody above-ground biomass ($\mlms$) by PFT.
 Each symbol is an individual plant.
 Lines are generalised additive model fits.
-(b) and (c) $M_F/M_S$ and leaf area ratio ($A_F/M_S$) at the average plant height in the dataset, estimated from fitted models in panel (a).
+(b) and (c) $\mlms$ and leaf area ratio ($\afms$) at the average plant height in the dataset, estimated from fitted models in panel (a).
 Error bars are 95\% confidence intervals.
@@
     \fi
-    \caption{Total plant leaf area ($A_F$) is approximately proportional to basal stem area ($A_S$), and this relationship is similar between PFTs.
+    \caption{Total plant leaf area ($\af$) is approximately proportional to basal stem area ($\as$), and this relationship is similar between PFTs.
 Lines are generalised additive model fits, with grey areas approximate 95\% confidence intervals.}
@@
     \fi
-    \caption{Above-ground woody biomass ($M_S$) scales with basal stem area ($A_S$), similarly for the three PFTs.
-(a) $M_S$ as a function of $A_S$ for the entire dataset.
+    \caption{Above-ground woody biomass ($\ms$) scales with basal stem area ($\as$), similarly for the three PFTs.
+(a) $\ms$ as a function of $\as$ for the entire dataset.
 Lines are generalized additive model fits.
@@
     \fi
-    \caption{Histograms (as probability density functions) of leaf mass per area ($M_F/A_F$), leaf area per unit basal stem area ($A_F/A_S$) and the stem mass index ($M_S/(A_S*H)$) grouped by the three PFTs.
+    \caption{Histograms (as probability density functions) of leaf mass per area ($\mfaf$), leaf area per unit basal stem area ($\afas$) and the stem mass index ($\mshas$) grouped by the three PFTs.
 Arrows indicate means by PFT.
@@
 \textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\
-\textbf{Fig. S2} Leaf area ratio ($A_F/M_T$) by PFT.\\
+\textbf{Fig. S2} Leaf area ratio ($\af/M_{\rm T}$) by PFT.\\
 \textbf{Fig. S3} Comparison of three methods for variance partitioning of the four studied variables \\
-\textbf{Fig. S4} Whole-plant leaf mass ($M_F$) and leaf area ($A_F$) as a function of total aboveground biomass. \\ 
-\textbf{Fig. S5} Relationships between leaf - stem biomass distribution ($M_F/M_S$) and mean annual temperature (MAT) (\degree C). \\
+\textbf{Fig. S4} Whole-plant leaf mass ($\mf$) and leaf area ($\af$) as a function of total aboveground biomass. \\
+\textbf{Fig. S5} Relationships between leaf - stem biomass distribution ($\mlms$) and mean annual temperature (MAT) (\degree C). \\
 \textbf{Fig. S6} As Fig S5 but with angiosperms split up into evergreen and deciduous. \\

### 2016-02-25 manuscript.Rnw
Comments and edits throughout
Outstanding issues have been commented in the text; see lines beginning with "%"

@@
 \usepackage{fixltx2e}
-\graphicspath{ {./figures/} }
 \usepackage{geometry}
@@
 \geometry{verbose,tmargin=2.4cm,bmargin=2.4cm,lmargin=2.4cm,rmargin=2.4cm}
-\usepackage{natbib}
 
-\bibliographystyle{newphyt}
-\renewcommand{\figurename}{Fig.}
-\usepackage[labelfont=bf, labelsep=quad]{caption}
-\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
-\newcommand{\mf}{M_{\rm F}}
-\newcommand{\ms}{M_{\rm S}}
-\newcommand{\as}{A_{\rm S}}
-\newcommand{\af}{A_{\rm F}}
-\newcommand{\asb}{A_{\rm Sbh}}
-\newcommand{\mlms}{\mf/\ms}
-\newcommand{\asms}{\as/\ms}
-\newcommand{\afas}{\af/\as}
-\newcommand{\mfaf}{\mf/\af}
-\newcommand{\afms}{\af/\ms}
-\newcommand{\mfas}{\mf/\as}
-\newcommand{\msas}{\ms/\as}
-\newcommand{\mshas}{\ms/(H \cdot \as)}
+\input{common-defs}
 
@@
 We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, and climate on biomass distribution and three component variables.
-\item The ratio of leaf mass to woody above-ground mass ($\mlms$) varied strongly between PFTs.
-We found that $\mlms$, once corrected for plant height, was proportional to leaf mass per area across PFTs.
+\item The ratio of leaf mass to woody above-ground mass ($\mfms$) varied strongly between PFTs.
+We found that $\mfms$, once corrected for plant height, was proportional to leaf mass per area across PFTs.
 As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass.
@@
 Globally, forests store approximately 360Pg of carbon in living biomass \citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}.
-Reducing uncertainties about biomass distributionand carbon residence times is a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014,negron2015}, in particular because projections by global vegetation models (GVMs) are particularly sensitive to this component \citep{friend2014,ise2010}.
+Reducing uncertainties about biomass distribution and carbon residence times is a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014,negron2015}, in particular because projections by global vegetation models (GVMs) are particularly sensitive to this component \citep{friend2014,ise2010}.
 
@@
 While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of biomass invested in different tissues (leaves, stems, roots).
-To attempt to understand the source of this functional diversity, it is customary to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010,wullschleger2014}, assumed to differ in key attributes.
+One way to capture and understand the source of this functional diversity, is to consider only a few archetypal plant functional types (PFTs), assumed to differ in key attributes \citep{harrisonetal2010,wullschleger2014},.
 Previous work has shown consistently large differences in biomass distribution and related traits among major PFTs \citep{poorter2009,enquistniklas2002,poorter2012,poorter2015,reich2014,mccarthyetal2007}, but these differences are yet to be fully understood.
@@
 One puzzling aspect of the large variation in biomass distribution among PFTs is the much larger leaf mass in gymnosperms compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012, poorter2015}.
-A possible explanation is that gymnosperms compensate for their low leaf mass per area (LMA) \citep{poorter2009} by maintaining a large amount of leaf biomass, to maintain sufficient total leaf area.
-If this is the case, total leaf area per unit above-ground biomass would be less variable between angiosperms and gymnosperms than leaf mass.
-Some oft-cited studies have however assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}, but recent large compilations on biomass distribution did not include measurements of leaf area in addition to mass to test this hypothesis \citep{poorter2012,poorter2015,reich2014}.
-The idea that high total leaf mass may compensate for low LMA would also explain the difference between evergreen (high LMA, high leaf mass) and deciduous (low LMA, low leaf mass) species \citep{poorter2015}.
-Here we set out to directly test the hypothesis that across PFTs, $M_F/M_S$ is related to LMA with a global database that includes observations of leaf area as well as leaf mass measured on the same plants.
+% Note I switched "low" and "high" in sentences below. Please verify revised versions are correct.
+One possible explanation is that gymnosperms compensate for their high leaf mass per area (LMA) \citep{poorter2009} by maintaining a large amount of leaf biomass, in order to maintain sufficient total leaf area.
+If this is the case, the total leaf area of plants at a standardised size should be less variable between angiosperms and gymnosperms than total leaf mass.
+The idea that high total leaf mass may compensate for high LMA would also explain the difference between evergreen (low LMA, low leaf mass) and deciduous (high LMA, high leaf mass) species \citep{poorter2015}.
+Some oft-cited studies make a different assumption: that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
+Although recent large compilations on biomass distribution included data on leaf mass, they did not include measurements of leaf area so were unable to evaluate these alternatives \citep{poorter2012,poorter2015,reich2014}.
+Here we set out to directly test the hypothesis that across PFTs, the ratio of leaf mass to stem mass -- $\mfms$ -- is related to LMA, but the ratio of leaf area to stem mass is invariant with respect to to LMA.
 
 Little is known about global-scale patterns in biomass distribution in relation to climate.
-It can be expected that biomass distribution between leaves and stems is correlated with precipitation or mean annual temperature because smaller-scale comparisons have shown responses to environmental drivers \citep{berninger1994,callaway1994,delucia2000,poyatos2007}.
-Recently, \citet{reich2014} demonstrated that $M_F/M_S$ (for whole stands, either including or excluding below-ground components) increased with mean annual temperature, after separating the data into gymnosperm and angiosperm stands, and accounting for the large size-related variation in biomass distribution.
-Our second objective was thus to test for climate-related variation in above-ground biomass distribution, and specifically to test whether $M_F/M_S$ increases with mean annual temperature.
+One might expect that biomass distribution between leaves and stems is correlated with precipitation or mean annual temperature, as small-scale comparisons have shown such responses \citep{berninger1994,callaway1994,delucia2000,poyatos2007}.
+Recently, \citet{reich2014} demonstrated that the ratio of leaf to total plant mass (for whole stands, either including or excluding below-ground components) increased with mean annual temperature, after separating the data into gymnosperm and angiosperm stands, and accounting for the large size-related variation in biomass distribution.
+Thus, our second objective was  to test for climate-related variation in above-ground biomass distribution, and specifically to test whether $\mfms$ increases with mean annual temperature.
 
 To test our hypotheses, we use a recently compiled large database \citep{falster2015} which includes observations of biomass and size metrics for individual plants (Fig.
-1), compiled from 175 studies across nine vegetation types (Fig.
-S1), across the three major climate zones (boreal, temperate and tropical).
-In this paper we focus on field-grown woody plants, spanning the entire size range of woody plants (0.01 - \textgreater 100 m height).
+1), compiled from 176 studies spanning nine vegetation types (Fig.
+S1) and three major climate zones (boreal, temperate, and tropical).
+In this paper we focus on field-grown woody plants, spanning the entire size range of woody plants (0.01 - \textgreater{} 100 m height).
 
@@
 We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total includes records for \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species.
-The database has very limited overlap (n = 261 individuals, 1.7 \%) with the recent large compilation of \citet{poorter2015} and differs in that measurements are all for individual plants (where \citet{poorter2015} included many stand-based estimates converted back to invididuals).
-In this paper we restrict our analysis to records that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (typically 1.3m) ($A_{Sbh}$) (n=\Sexpr{nrow(dataset)}).
-The database contains many more variables, for example root biomass for a much smaller subset of studies.
+The database has very limited overlap (n = 261 individuals, 1.7 \%) with the recent large compilation of \citet{poorter2015} and differs in that measurements are all for individual plants (where \citet{poorter2015} included many stand-based averages.
+In this paper we restrict our analysis to records that include leaf mass ($\mf$), leaf area ($\af$), above-ground woody biomass ($\ms$), plant height ($H$), and stem area measured at ground level ($\as$), or at breast height (typically 1.3m) ($\asb$) (n=\Sexpr{nrow(dataset)}).
+The database contains many more variables, for example root biomass; but usually for a much smaller subset of studies.
 Here we limit the analysis to patterns in aboveground biomass distribution.
-For each analysis, we used different subsets of the data because not all variables were measured in each study.
+Our analysis has several parts; to maximise the amount of available data for each part we we used different subsets of the data, as not all variables were measured in each study.
 Sample sizes by PFT are summarised in Table \ref{table:counttable}.
@@
 
-We only used leaf mass per area (LMA) directly estimated for the harvested plants (typically for a subsample of leaves, see \citet{falster2015} for details on the methods for each contributed study).
+We only used LMA directly estimated for the harvested plants (typically for a sub-sample of leaves, see \citet{falster2015} for details on the methods for each contributed study).
 For conifers, leaf area was converted to half-total surface area using the average of a set of published conversion factors \citep{barclay2000}, with  different conversion factors applied to pines (\emph{Pinus} spp.) and non-pines.
@@
 
-Stem cross-sectional area was measured either at breast height (1.3 - 1.34m) and/or at the base of the plant.
-In all analyses, we used basal stem area because many plants (N=\Sexpr{nrow(dataset[dataset$h.t < 1.3,])}) were \textless 1.3m tall.
-For the subset of the data where both were measured, we estimated basal stem area ($\as$) from breast height stem area ($\asb$) by fitting the following equation, using non-linear regression.
-
+% This next refers to stem xs-area, but then the equation is for diameter. Did we really use both?
+In our dataset, stem cross-sectional area was measured either at breast height (1.3 - 1.34m) and/or at the base of the plant.
+In our analyses, we used basal stem area because many plants (N=\Sexpr{nrow(dataset[dataset$h.t < 1.3,])}) were \textless{} 1.3m tall.
+We therefore converted records where area was measured at breast height only to area at base, using a relationship fitted to individuals where both measurements existed.
+For the subset of the data where both were measured, we estimated basal stem area ($\as$) from breast height stem area ($\asb$) using the equation,
 \begin{equation}
-D_{\rm S} = D_{\rm Sbh} H^c / (H - H_{\rm bh})^c
+D_{\rm S} = D_{\rm Sbh} \left(\frac{H}{H - H_{\rm bh}}\right)^c
 \end{equation}
-
-where $D_{\rm S}$ is the basal stem diameter (m), $D_{\rm Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{\rm Sbh}$ was measured (typically 1.3 or 1.34m), and \emph{c} was further expressed as a function of plant height:
-
+where $D_{\rm S}$ is the basal stem diameter (m), $D_{\rm Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{\rm Sbh}$ was measured (typically 1.3 or 1.34m).
+The parameter $c$ was further expressed as a function of plant height:
 \begin{equation}
-c = c_0H^{c_1}
+c = c_0H^{c_1}.
 \end{equation}
-
 The estimated coefficients were $c_0$ = \Sexpr{signif(coef(basalafit)[1],3)}, $c_1$ = \Sexpr{signif(coef(basalafit)[2],3)}, root-mean square error = \Sexpr{signif(summary(basalafit)$sigma,3)}.
@@
 \subsection*{Data analysis}
-
-\emph{might be moved to introduction}
-To better understand the mechanisms of PFT and climate-related variation in leaf-stem biomass distribution, we decomposed it into three component variables (Eq.~\ref{eq:mfmspart}),
-
+To better understand the mechanisms of PFT and climate-related variation in biomass distribution, we decomposed the ratio of leaf-to stem biomass it into three component variables.
+Applying the chain rule of differential calculus gives
 \begin{equation}
 \label{eq:mfmspart}
-\frac{\mf}{\ms} = \frac{\mf}{\af} \cdot \frac{\af}{\as} \cdot \frac{\as}{\ms}
+\frac{\mf}{\ms} = \frac{\mf}{\af} \cdot \frac{\af}{\as} \cdot \frac{\as}{\ms},
 \end{equation}
-
-where $\mfaf$ is the leaf mass per area (LMA), $\afas$ the ratio of plant leaf area to basal stem area, and $\asms$ the ratio of stem basal area to total woody above-ground mass.
-Using variance partitioning, we quantify the contribution of plant height, plant functional type, and climate to the overall variation in $\mlms$ and for all three component variables.
+where $\mfaf$ is the LMA, $\afas$ the ratio of plant leaf area to basal stem area, and $\asms$ the ratio of stem basal area to total woody above-ground mass.
+Using variance partitioning, we quantified the contribution of plant height, plant functional type, and climate to the overall variation in $\mfms$ and for all three component variables.
 
@@
 Instead, we used generalised additive models (GAM), a semi-parametric modelling approach that makes no assumption about the shape of the relationships.
-GAMs were used to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as $M_F/M_S$ at a common plant height.
+GAMs were used to visualise the relationships between biomass and plant size variables, and to estimate variables such as $\mfms$ at a common plant height.
 In all fitted GAMs, we used a cubic regression spline.
@@
 In all fitted GAMs, we used species-dataset combination as a random effect.
-To homogenize variance, all variables (except MAP and MAT) were log-transformed prior to analysis.
+To homogenise variance, all variables (except MAP and MAT) were log-transformed prior to analysis.
 
-Variance explained by quantitative climate variables (MAP and MAT) were tested with GAMs where variables were sequentially added to the model, and the explained variance ($R^2$) calculated.
-We compared this approach when using linear mixed-effects models instead of GAMs (calculating the the $R^2$ for linear mixed-effects models for the fixed effects only \citep{nakagawa2013}), as well as an independent method, independent effects analysis.
+The amount of variance explained by quantitative climate variables (MAP and MAT) was assessed with GAMs, with variables were sequentially added to the model and the explained variance ($R^2$) calculated.
+We compared this approach when using linear mixed-effects models instead of GAMs \citep[calculating the $R^2$ for linear mixed-effects models for the fixed effects only][]{nakagawa2013}, as well as another method, independent effects analysis.
 The three methods generally agreed on the ranking of variable importance (see Fig.
@@
 
-All analyses were conducted in R 3.2.3 \citep{R-2015}.
-GAMs were fitted using the \emph{mgcv} package \citep{wood2006}.
-The code to replicate this analysis (and all figures) is available at \url{http://github.com/RemkoDuursma/baadanalysis} (doi: \emph{to be added}).
+All analyses were conducted in R 3.2.3 \citep{R-2015}, using the packages `mgcv` \citep{wood2006}, `hier.part` \citep{walsh2013}, `MuMIn` \citep{barton2015}, `visreg` \citep{breheny2015}, and `remake` \citep{fitzjohn2015}.
+In addition, the code replicating this analysis (and all figures) is available at \url{http://github.com/RemkoDuursma/baadanalysis} (doi: \emph{to be added}).
 
@@
 
-The compiled dataset covered the global climate space well (defined by mean annual temperature, MAT; and mean annual precipitation, MAP) (Fig.
+The compiled dataset covered the global climate space well (indicated by coverage across sites classified by mean annual temperature, MAT; and mean annual precipitation, MAP; Fig.
 1).
@@
 
-As expected, both whole-plant leaf mass ($\mf$) and above-ground woody biomass ($\ms$) showed a strong depence on plant height (Fig.
+As expected, both whole-plant leaf mass ($\mf$) and above-ground woody biomass ($\ms$) showed a strong dependence on plant height (Fig.
 2).
@@
 It is also evident that $\ms$ increases much quicker with plant height than $\mf$.
-As a result, the ratio $\mlms$ quickly decreased with plant height (Fig.
-3a), from an average of \Sexpr{mfmssmall} for very small saplings (\textless 0.1m) to an average of \Sexpr{mfmslarge} for large trees (\textgreater 30m).
+As a result, the ratio $\mfms$ quickly decreased with plant height (Fig.
+3a), from an average of \Sexpr{mfmssmall} for very small saplings (\textless{} 0.1m) to an average of \Sexpr{mfmslarge} for large trees (\textgreater{} 30m).
 
-We found large differences between the three major PFTs in terms of above-ground biomass distribution (Fig.
+We found large differences between the three major PFTs in terms of above-ground biomass distribution,a s quantified by the ratio $\mfms$ (Fig.
 3a).
-These differences were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater evergreen angiosperm \textgreater deciduous angiosperm.
-Next we calculated $\mlms$ at a common plant height for each of the PFTs.
-These three size-corrected $\mlms$ values were highly significantly different between PFTs (p \textless 0.001), and were approximately proportional to the average LMA ($\mfaf$) across the PFTs (Fig.
+These differences were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater{} evergreen angiosperm \textgreater{} deciduous angiosperm.
+Next we calculated $\mfms$ at a common plant height for each of the PFTs.
+These three size-corrected $\mfms$ values were highly significantly different between PFTs (p \textless{} 0.001), and were approximately proportional to the average LMA ($\mfaf$) across the PFTs (Fig.
 3b).
-As a result, the amount of leaf area per unit woody above-ground biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater 0.05, see Fig.
+As a result, the amount of leaf area per unit woody above-ground biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater{} 0.05, see Fig.
 3c), because the differences in $\mfas$ were exactly compensated by differences in LMA (see also Fig.
@@
 
-Next, we decomposed $\mlms$ into three component variables with distinct biological interpretation (Eq.\ref{eq:mfmspart}).
-For each of the component variables and $\mlms$, we estimated the variance explained by plant height, PFT, and climate variables (MAT or aridity) (Table~\ref{table:varparttable}).
+Next, we decomposed $\mfms$ into three component variables with distinct biological interpretation (Eq.
+\ref{eq:mfmspart}).
+For each of the component variables and $\mfms$, we estimated the variance explained by plant height, PFT, and climate variables (MAT or aridity) (Table \ref{table:varparttable}).
 We did this by sequentially adding terms to a GAM, and calculating the $R^2$ for each fitted model.
-The first model included only plant height, and already explained much variation (65 - 79\%) in $\mlms$ and $\msas$ but not LMA and $\afas$ (4 - 11 \%).
-When PFT was added to the model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $\mlms$ and $\mfaf$ but not $\afas$ or $\asms$, demonstrating that PFT differences entered largely in leaf mass-based variables.
+The first model included only plant height, and already explained much variation (65 - 79\%) in $\mfms$ and $\msas$ but not LMA and $\afas$ (4 - 11 \%).
+When PFT was added to the model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $\mfms$ and $\mfaf$ but not $\afas$ or $\asms$, demonstrating that PFT differences entered largely in leaf mass-based variables.
 Additional variance explained by adding PFT was especially large for LMA, which is consistent with the large differences between PFTs in LMA shown in Fig.
@@
 
-From the variance partitioning (Table 1), it is evident that $\afas$ did not vary appreciably between PFTs.
-Indeed, when all PFTs were plotted together, similar $\af$ was attained at a given $\as$ across the entire plant size range (with the exception of extremely large trees with \textgreater 1 $m^2$ stem basal area, for which comparatively little data was available) (Fig.
+From the variance partitioning (Table \ref{table:varparttable}), it is evident that $\afas$ did not vary appreciably between PFTs.
+Indeed, when all PFTs were plotted together, similar $\af$ was attained at a given $\as$ across the entire plant size range (with the exception of extremely large trees with \textgreater{} 1m$^2$ stem basal area, for which comparatively little data was available) (Fig.
 4).
-We fit a GAM of $\af$ as a function of $\as$ with or without PFT as another covariate (both as intercept and slope interaction), which showed PFT significantly improved model fit (likelihood ratio test, p \textless 0.001), however the $R^2$ increased only from \Sexpr{round(100*afas$R2a,2)} to \Sexpr{round(100*afas$R2b,2)}\%, and thus had little if any practical significance.
+We fit a GAM to $\af$ as a function of $\as$ with and without PFT as a covariate
+% please clarify next line
+(both as intercept and slope interaction),
+which showed PFT significantly improved model fit (likelihood ratio test, p \textless{} 0.001).
+Nevertheless, the $R^2$ of the relationship only increased from \Sexpr{round(100*afas$R2a,2)} to \Sexpr{round(100*afas$R2b,2)}\%, indicating that while significant, the PFT had little additional predictive power.
 
-Likewise, the ratio of stem basal area to stem mass ($\asms$) apparently varied little between PFTs (Table~\ref{table:varparttable}).
+Likewise, the ratio of stem basal area to stem mass ($\asms$) apparently varied little between PFTs (Table \ref{table:varparttable}).
 To illustrate this further, a plot of $\ms$ as a function of $\as$ revealed a tight relationship, with little apparent difference between PFTs (Fig 5a).
@@
 We fitted a GAM of $\ms$ as a function of $\as$ (added as a smooth term) with or without PFT as a covariate.
-Again PFT was formally significant (likelihood ratio test, p \textless 0.001), but the $R^2$ increased only marginally from \Sexpr{round(100*msas$R2a,2)} to \Sexpr{round(100*msas$R2b,2)}\%, again showing little practical significance of PFT in explaining this variable.
+Again PFT was formally significant (likelihood ratio test, p \textless{} 0.001), but the $R^2$ increased only marginally from \Sexpr{round(100*msas$R2a,2)} to \Sexpr{round(100*msas$R2b,2)}\%, again showing little practical significance of PFT in explaining this variable.
 
-Taken together, these results demonstrate that 1) the majority of PFT-related variation in $\mlms$ results from variation in LMA (not $\afas$ or $\asms$), and 2) the majority of height-related variation in $\mlms$ results from variation in $\asms$ (not LMA or $\afas$).
-It is however important to note that there is very large variation between individuals even after accounting for PFT (Fig.
+Taken together, these results demonstrate that 1) the majority of PFT-related variation in $\mfms$ results from variation in LMA, not $\afas$ or $\asms$; and 2) the majority of height-related variation in $\mfms$ results from variation in $\asms$, not LMA or $\afas$.
+Nevertheless, it is important to note that there is very large variation between individuals within PFTs (Fig.
 6).
-To correct for plant height effect on $\msas$, we expressed the stem mass ratio as $\mshas$ because we expect from basic geometry that $\ms$ is proportional to H and $\as$ ($\ms = \phi \rho \as H$, where $\phi$ a stem shape parameter, $\rho$ wood density).
+To correct for plant height effect on $\msas$, we expressed the stem mass ratio as $\mshas$ because we expect from basic geometry that $\ms$ is proportional to H and $\as$, via the relationship: $\ms = \phi \rho \as H$, where $\phi$ a stem shape parameter, and $\rho$ wood density.
 The histograms summarize that PFT effects are very evident in LMA (Fig.
@@
 
-Climate variables consistently added little explanatory power when added to a statistical model that included PFT and height (the adjusted R$^2$ frequently decreased, see Table~\ref{table:varparttable}).
+Climate variables consistently added little explanatory power when added to a statistical model that included PFT and height (the adjusted R$^2$ frequently decreased, see Table \ref{table:varparttable}).
 The only exception was LMA, for which a model that included MAT explained 53\% of the variation, compared to a model with height and PFT (45\%).
 Since the lack of climate effects on biomass distribution is in contrast with recent finding of \citep{reich2014}, we also analyzed climate effects with methods similar to that study.
-When we tested for MAT effects on $\mlms$ when only considering gymnosperms and angiosperms (thus grouping decidous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig.
+When we tested for MAT effects on $\mfms$ when only considering gymnosperms and angiosperms (thus grouping decidous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig.
 S5) for angiosperms for most size classes (but never for gymnosperms).
 We found, however, that this was largely due to the fact that at higher MAT, an increasing proportion of angiosperms are evergreen (Fig.
-1), which have higher $\mlms$.
+1), which have higher $\mfms$.
 When analyzed separately for deciduous and evergreen angiosperms, few and weaker significant relationships were found (Fig.
@@
 
-Read \citep{bloom2016}; global patterns in inferred allocation and biomass pools (leaf / stem).
+% Read \citep{bloom2016}; global patterns in inferred allocation and biomass pools (leaf / stem).
 
 Using the largest database of individual woody plant allometry to date, we studied patterns in above-ground biomass distribution among major woody plant functional types (PFTs), plant size, and climate.
-Our compilation is the first of its kind that also includes leaf area estimates on the same plants, allowing a direct test of the hypothesis that leaf and stem biomass distribution is related to the leaf mass per area.
-We found that at a common plant height, the leaf / woody biomass ratio ($\mlms$) was proportional to the leaf mass per area across three PFTs, so that leaf area per unit above-ground biomass did not vary between PFTs.
+Our compilation is the first of its kind that also includes leaf area estimates on the same plants, allowing a direct test of the hypothesis that leaf and stem biomass distribution is related to the LMA.
+We found that at a common plant height, the leaf / woody biomass ratio ($\mfms$) was proportional to the LMA across three PFTs, so that leaf area per unit above-ground biomass did not vary between PFTs.
 We found very weak effects of climate (mean annual temperature and aridity) on biomass distribution, so that the vast majority of the variation in biomass distribution was explained by plant size and PFT.
 
-As expected, $\mlms$ steeply declined with plant height.
+As expected, $\mfms$ steeply declined with plant height.
 It is well known that it is necessary to correct for plant size when comparing biomass distribution among groups \citep{mcconnaughay1999}.
@@
 
-We have confirmed the hypothesis that leaf area is less variable between PFTs than leaf mass.
-\citet{poorter2015} hypothesized that this is because forests should converge to a similar leaf area index (LAI), as total light interception at some point saturates with LAI.
-Although we found remarkably similar leaf area per unit above-ground woody biomass ($\mlms$) between the three PFTs studied (Fig.
+% This paragraph has been substantially changed, please review
+We have confirmed the hypothesis that the amount of leaf area supported per plant is less variable between PFTs than the amount of leaf mass.
+\citet{poorter2015} hypothesised such a result might arise, because forests would converge to a similar leaf area index (LAI).
+Although we found remarkably similar leaf area per unit above-ground woody biomass ($\mfms$) between the three PFTs studied (Fig.
 S2), this does not in itself imply that LAI is equally similar.
-In a global compilation of LAI for undisturbed forest stands, \citet{iio2014} reported a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI.
-To scale our results to stand- and landscape scales thus requires more information on differences in stand structure between PFTs.
+Indeed, in a global compilation of LAI for undisturbed forest stands, \citet{iio2014} reported a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI.
+Moreover, we would in fact expect LAI to change with LMA, not because of changes in the amount of leaf area maintained by individual plants, but because LMA influences the shade-tolerance of seedlings \citep{poorter2006, falster2011}, and thereby the LAI of the community \citep{falster2011}.
+The ultimate reason LMA might effect shade tolerance is because is via its correlation with leaf lifespan \citep{wright2004}.
+For whole stands, we can therefore expect total stand leaf mass to be correlated with LMA, both because of of greater leaf mass per plant \citep[e.g.]{reich1992}, and the greater density of plants arising from increased leaf lifespan.
+Previous interpretations of differences in $\mfms$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002, poorter2012}.
+We argue that there are in fact two effects occurring simultaneously: differences in LMA cause differences in the $\mfms$ of individual plants, while differences in LL (and thereby LMA) cause differences in the LAI of whole stands.
 
-Previous interpretations of differences in $\mlms$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002,poorter2012}.
-Indeed, if all else equal, longer LL in gymnosperms would result in higher total leaf mass.
-Further supporting this explanation, \citet{reich1992} showed a positive correlation between LL and total stand leaf biomass across diverse forest stands.
-It is however difficult to separate cause from effect in the relationships between LL, LMA, and $\mlms$, particularly because LL and LMA are correlated across species and PFTs in a nearly proportional way \citet{wright2004}.
-
-After accounting for PFT and plant height, we found very weak and inconsistent effects of climate on all variables studied, with some exception for LMA (Table~\ref{table:varparttable}).
+After accounting for PFT and plant height, we found very weak and inconsistent effects of climate on all variables studied, with some exception for LMA (Table \ref{table:varparttable}).
 Climate was represented as either mean annual temperature or an aridity index (PET/MAP), however, results were similarly inconclusive when studying other climate variables such as composite drought indices, growing season length, or mean annual precipitation (results not shown).
 This result appears to contradict \citet{reich2014} who reported consistent increases in foliage biomass ($\mf$) with MAT, when corrected for above-ground woody biomass ($\ms$).
-We were able to partly reconcile our findings with those of \citep{reich2014}: when we considered only angiosperms and gymnosperms (thus grouping deciduous and evergreens, as did Reich et al.) $\mlms$ increased consistently with MAT for most size classes (Fig.
-S5), but not for gymnosperms.
-However, we found that for our database, this could be explained by the consistently higher proportion of evergreen angiosperms at higher MAT (Fig.
-1), which, as a result of higher LMA (Fig.
-2) also have a higher $\mlms$.
-However, based on a re-analysis of data in Reich et al.
-2014 (not shown), this shift in PFT with climate does not explain their observation of an increase in $\mlms$ with MAT, as this pattern still held within gymnosperms and deciduous angiosperms.
+We were able to partly reconcile our findings with those of \citet{reich2014}: when we categorised our species as only angiosperms vs gymnosperms (thus grouping deciduous and evergreens, as did Reich \emph{et al.}), $\mfms$ increased consistently with MAT for most size classes within angiosperms (Fig.
+S5), in line with the findings of \citet{reich2014}; on the other hadn we found no comparable relationship for gymnosperms.
+Moreover, we found that in our dataset, the significant relationships observed for angiosperms as a whole, could be explained via differences between the LMA of deciduous and evergreen species: evergreen angiosperms occurred both at higher MAT (Fig.
+1), and had a higher LMA (Fig.
+2), resulting in a correlation between $\mfms$ and MAT.
+We also re-analysed the  data in Reich et al.
+2014 (not shown) to determine whether this shift in PFT with climate could also explain their finding of an increase in $\mfms$ with MAT, but found the pattern still held within gymnosperms and deciduous angiosperms.
 
 Although we found no consistent effects of climate on biomass distribution after PFT and plant height were accounted for, effects of climate are likely mediated via plant height.
-At a global scale, average plant height is clearly related to climate via water availability \citet{klein2015,simard2011}.
+At a global scale, average plant height is clearly related to climate via water availability \citep{klein2015, simard2011}.
 Since biomass distribution itself is steeply related to plant height (Figs.
@@
 6).
-Our purpose was to test hypotheses related to first-order differences between PFTs, plant height, and climate, yet clearly more work can be done to understand the variation between species and sites witin groups.
-Some of the residual variation is likely explained by site fertility \citet{vanninen2005}, differences in stand density \citet{ilomaki2003}, social status within stands \citet{makela2006}, or local differences in water availability (possibly mediated by seasonality at a given MAP).
-Future work will focus on smaller-scale explanatory variables such as the ones mentioned here.
+Our purpose was to test hypotheses related to first-order differences between PFTs, plant height, and climate, yet clearly more work can be done to understand the variation between species and sites within groups.
+Some of the residual variation is likely explained by site fertility \citep{vanninen2005}, differences in stand density \citep{ilomaki2003}, social status within stands \citep{makela2006}, or local differences in water availability (possibly mediated by seasonality at a given MAP).
+It is also likely that the relationship between LMA and biomass distribution reported here among PFTs, also operates among species within PFTs.
+As LMA is known to vary widely within PFTs (\ref{fig:fighist}), such variation could underpin the equivalent variation in biomass distribution.
 
@@
 A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass distribution instead of using constant allocation fractions.
-Our results can be used to further constrain biomass allocation routines in GVM, in particular, the finding that $\mlms$ (at a given plant height) is approximately proportional to LMA can be tested against model predictions, or used directly to parameterize current algorithms.
-In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015,poorter2015,reich2014} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data.
+Indeed, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015,poorter2015,reich2014} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data.
+Our results also suggest a rethink in they way LMA is handled in most growth models.
+In most existing models, biomass allocation is determined first (by some means of another), and then the amount of leaf area is calculated by multiplying leaf biomass allocation with LMA.
+Our results suggest the chain of causation should be reversed: first decide the amount of leaf area, then let LMA determine the cost of building this leaf.
+Such an approach has already been implemented in at least one model of plant growth \citep{falster2016}.
+Other algorithms are of course possible, but this at least one way to ensure model predictions recover the patterns reported in this paper.
 
-\section*{Acknowledgments}
+\section*{Acknowledgements}
 
@@
 We thank Martin De Kauwe and anonymous referees for comments on an earlier version of this paper.
-Thanks also to Rich FitzJohn for advice about using the remake package.
-Figure 2a uses an image by Lana Heydon, IAN Image Library (http://ian.umces.edu/imagelibrary).
+Thanks also to Rich FitzJohn for advice about using the `remake` package.
+Figure 1a uses an image by Lana Heydon, IAN Image Library (http://ian.umces.edu/imagelibrary).
 Thanks to anonymous reviewers for suggestions that greatly improved the paper.
-Special thanks to Steve Zissou.
+Special thanks to Steve Zissou for advice on colour schemes.
 
@@
              caption="Sample sizes used for the four studied whole-plant variables, by plant functional type.
-Numbers indicate individual plants, with number of unique species between parentheses.
-Sample sized differed between variables because not all included studied measured all variables ("),
+Numbers indicate individual plants, with the number of unique species given in parentheses.
+Sample sizes differed between variables because not all included studied measured all variables"),
       sanitize.text.function=function(x){x},
@@
 
+% In the top row of table 2, it's seems surprising that H,PFT,MAT explains less variance than just H,PFT.
+% Check!!!
 <<results='asis'>>=
@@
 Climate variables included were either MAT or aridity index (calculated as the of PET to MAP).
-P-values not shown because all smooth terms were always highly significant (p < 0.01), with the exception of aridity index in the model for $\\mlms$ (p = 0.057)."),sanitize.text.function=function(x){x},
+P-values not shown because all smooth terms were always highly significant (p < 0.01), with the exception of aridity index in the model for $\\mfms$ (p = 0.057)."),sanitize.text.function=function(x){x},
       include.rownames=FALSE, caption.placement = "top")
@@
 Sample sizes are listed in Table \ref{table:counttable}.
-Lines are generalized additive model fits to visualize the highly non-linear relationships.
+Lines show generalized additive models, fitted to visualize the highly non-linear relationships.
 Note the logarithmic scale.}
@@
 
+% Could we add along the x-axis of panel a, a small arrow showing the common height where we sued to generate panels b-c?
 \begin{figure}[h!]
@@
     \caption{Dominant woody PFTs differ in above-ground biomass distribution due to underlying differences in leaf mass per area.
-(a) Leaf mass per unit woody above-ground biomass ($\mlms$) by PFT.
+(a) Leaf mass per unit woody above-ground biomass ($\mfms$) by PFT.
 Each symbol is an individual plant.
 Lines are generalised additive model fits.
-(b) and (c) $\mlms$ and leaf area ratio ($\afms$) at the average plant height in the dataset, estimated from fitted models in panel (a).
+(b) and (c) $\mfms$ and leaf area ratio ($\afms$) at the average plant height in the dataset, estimated from fitted models in panel (a).
 Error bars are 95\% confidence intervals.
@@
     \fi
+    % proportional implies slope of ~1, should we add 1:1 line in background?
     \caption{Total plant leaf area ($\af$) is approximately proportional to basal stem area ($\as$), and this relationship is similar between PFTs.
@@
 \textbf{Fig. S4} Whole-plant leaf mass ($\mf$) and leaf area ($\af$) as a function of total aboveground biomass. \\
-\textbf{Fig. S5} Relationships between leaf - stem biomass distribution ($\mlms$) and mean annual temperature (MAT) (\degree C). \\
+\textbf{Fig. S5} Relationships between leaf - stem biomass distribution ($\mfms$) and mean annual temperature (MAT) (\degree C). \\
 \textbf{Fig. S6} As Fig S5 but with angiosperms split up into evergreen and deciduous. \\

### 2016-02-25 manuscript_suppinfo.Rnw
Comments and edits throughout
Outstanding issues have been commented in the text; see lines beginning with "%"

@@
 \geometry{verbose,tmargin=2.4cm,bmargin=2.4cm,lmargin=2.4cm,rmargin=2.4cm}
-\graphicspath{ {./figures/} }
 \usepackage{parskip}
-\usepackage{natbib}
 
-\bibliographystyle{newphyt}
-\renewcommand{\figurename}{Fig.}
-\usepackage[labelfont=bf, labelsep=quad]{caption}
+\input{common-defs}
 
@@
 1 in main text.
-Symbols are colored by vegetation type as reported by the data contributors.}
+Symbols are coloured by vegetation type as reported by the data contributors.}
 \label{fig:figureS1}
@@
     \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
-    \caption{Leaf area ratio ($A_F/M_T$) by plant height (H) and PFT.
+    \caption{Leaf area ratio ($\afmt$) by plant height (H) and PFT.
 Each symbol is an individual plant.
@@
     \caption{Comparison of three methods for variance partitioning of the four studied variables (see legend) by total plant height, plant functional type and climate (MAP, MAT and interaction).
-\textbf{GAM}: variance component as estimated with a generalized additive model, as the reduction in R2 when the independent variable (i.e.
+\textbf{GAM}: variance component as estimated with a generalized additive model, as the reduction in $R^2$ when the independent variable (i.e.
 PFT, climate or height) was excluded from a model.
 Note that a GAM does not include interactions.
-\textbf{LMM}: linear mixed model with all interactions, and squared terms for the continuous variables, where the variance component was again estimated as the reduction in R2 when the independent variable (and all its interactions) were removed from the model.
-R2 was calculated following Nakagawa and Schielzeth (2013) using the MuMIn R package (excluding random effects).
+\textbf{LMM}: linear mixed model with all interactions, and squared terms for the continuous variables, where the variance component was again estimated as the reduction in $R^2$ when the independent variable (and all its interactions) were removed from the model.
+$R^2$ was calculated following \citet{nakagawa2013} using the `MuMIn` R package (excluding random effects) \citep{barton2015}.
 Note that LMM and GAM give very similar ranking, and even absolute variance components, for nearly all variables.
-Both methods identified AF/AS and MF/AF as the only variables with a practically significant dependence on climate (variance component > 0.1), see Fig S4.
-\textbf{IEA}: independent effects analysis (aka hierarchical partitioning), using the algorithm by Chevan and Sutherland (1991) as implemented in the hier.part R package.
+Both methods identified AF/AS and MF/AF as the only variables with a practically significant dependence on climate (variance component \textgreater{} 0.1), see Fig S4.
+\textbf{IEA}: independent effects analysis (aka hierarchical partitioning), using the algorithm by \citet{chevan1991}, as implemented in the `hier.part` package for R \citep{walsh2013}.
 Note that this method does not include interactions or squared terms, yet gave similar results to GAM and LMM (except for variables that explained little variation).}
@@
     \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
-    \caption{Whole-plant leaf mass ($M_F$) and leaf area ($A_F$) as a function of total aboveground biomass.
-In the main text, we have chosen to use plant height as a metric of plant size, to avoid correlations between dependent and independent variables ($M_F$ appears in both variables in panel (a)).
+    \caption{Whole-plant leaf mass ($\mf$) and leaf area ($\af$) as a function of total aboveground biomass.
+In the main text, we have chosen to use plant height as a metric of plant size, to avoid correlations between dependent and independent variables ($\mf$ appears in both variables in panel a).
 The overall differences between PFTs, and the finding that leaf mass is more variable between PFTs than leaf area, still hold when we use mass as the size covariate.}
@@
     \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
-    \caption{Relationships between leaf - stem biomass distribution ($M_F/M_S$) and mean annual temperature (MAT) (\degree C).
+    \caption{Relationships between leaf - stem biomass distribution ($\mfms$) and mean annual temperature (MAT) (\degree C).
 Similar to Reich et al.
@@
 Lines are fits from mixed-effects models (with species within study as random effect, and only MAT and evergreen/deciduous classification as fixed effects).
-Solid lines are significant (p < 0.05), dashed lines are not (p > 0.05).
-Symbols are partial residuals as calculated with the visreg package in R.}
+Solid lines are significant (p \textless{} 0.05), dashed lines are not (p \textgreater{} 0.05).
+Symbols are partial residuals as calculated with the `visreg` package in R \citep{breheny2015}.}
     \label{fig:figreichclimate}
@@
 
+\clearpage
+\bibliography{references}
+
 \end{document}

### 2016-02-25 manuscript.Rnw
Minor edits and typo fixes

@@
 One puzzling aspect of the large variation in biomass distribution among PFTs is the much larger leaf mass in gymnosperms compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012, poorter2015}.
-% Note I switched "low" and "high" in sentences below. Please verify revised versions are correct.
 One possible explanation is that gymnosperms compensate for their high leaf mass per area (LMA) \citep{poorter2009} by maintaining a large amount of leaf biomass, in order to maintain sufficient total leaf area.
@@
 We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total includes records for \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species.
-The database has very limited overlap (n = 261 individuals, 1.7 \%) with the recent large compilation of \citet{poorter2015} and differs in that measurements are all for individual plants (where \citet{poorter2015} included many stand-based averages.
+The database has very limited overlap (n = 261 individuals, 1.7 \%) with the recent large compilation of \citet{poorter2015} and differs in that measurements are all for individual plants (where \citet{poorter2015} included many stand-based averages).
 In this paper we restrict our analysis to records that include leaf mass ($\mf$), leaf area ($\af$), above-ground woody biomass ($\ms$), plant height ($H$), and stem area measured at ground level ($\as$), or at breast height (typically 1.3m) ($\asb$) (n=\Sexpr{nrow(dataset)}).
@@
 We therefore converted records where area was measured at breast height only to area at base, using a relationship fitted to individuals where both measurements existed.
-For the subset of the data where both were measured, we estimated basal stem area ($\as$) from breast height stem area ($\asb$) using the equation,
+For the subset of the data where both were measured, we estimated basal stem area ($\as$) from breast height stem area ($\asb$) from the equation,
 \begin{equation}
@@
 
-Because, similar to \citep{poorter2015}, none of the studied relationships were linear on a log-log scale, we did not use standard allometric equations to study relationships between size and biomass variables.
+Because, similar to \citet{poorter2015}, none of the studied relationships were linear on a log-log scale, we did not use standard allometric equations to study relationships between size and biomass variables.
 Instead, we used generalised additive models (GAM), a semi-parametric modelling approach that makes no assumption about the shape of the relationships.
@@
 The amount of variance explained by quantitative climate variables (MAP and MAT) was assessed with GAMs, with variables were sequentially added to the model and the explained variance ($R^2$) calculated.
-We compared this approach when using linear mixed-effects models instead of GAMs \citep[calculating the $R^2$ for linear mixed-effects models for the fixed effects only][]{nakagawa2013}, as well as another method, independent effects analysis.
+We compared this approach when using linear mixed-effects models instead of GAMs \citep[calculating the $R^2$ for linear mixed-effects models for the fixed effects only following][]{nakagawa2013}, as well as another method, independent effects analysis.
 The three methods generally agreed on the ranking of variable importance (see Fig.
@@
 The histograms summarize that PFT effects are very evident in LMA (Fig.
-6a), not in $\afas$, and only to some extent for $\msas$, which is entirely driven by large gymnosperm trees.
+6a), not in $\afas$, and only to some extent for $\mshas$, which is entirely driven by large gymnosperm trees.
 However for all variables there was much unexplained variation between species (and individuals within species) within PFTs.
@@
 The only exception was LMA, for which a model that included MAT explained 53\% of the variation, compared to a model with height and PFT (45\%).
-Since the lack of climate effects on biomass distribution is in contrast with recent finding of \citep{reich2014}, we also analyzed climate effects with methods similar to that study.
+Since the lack of climate effects on biomass distribution is in contrast with recent finding of \citet{reich2014}, we also analyzed climate effects with methods similar to that study.
 When we tested for MAT effects on $\mfms$ when only considering gymnosperms and angiosperms (thus grouping decidous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig.
@@
 We were able to partly reconcile our findings with those of \citet{reich2014}: when we categorised our species as only angiosperms vs gymnosperms (thus grouping deciduous and evergreens, as did Reich \emph{et al.}), $\mfms$ increased consistently with MAT for most size classes within angiosperms (Fig.
-S5), in line with the findings of \citet{reich2014}; on the other hadn we found no comparable relationship for gymnosperms.
+S5), in line with the findings of \citet{reich2014}; on the other hand we found no comparable relationship for gymnosperms.
 Moreover, we found that in our dataset, the significant relationships observed for angiosperms as a whole, could be explained via differences between the LMA of deciduous and evergreen species: evergreen angiosperms occurred both at higher MAT (Fig.
@@
 It is also likely that the relationship between LMA and biomass distribution reported here among PFTs, also operates among species within PFTs.
-As LMA is known to vary widely within PFTs (\ref{fig:fighist}), such variation could underpin the equivalent variation in biomass distribution.
+As LMA is known to vary widely within PFTs (Fig.~\ref{fig:fighist}), such variation could underpin the equivalent variation in biomass distribution.
 
@@
 Indeed, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015,poorter2015,reich2014} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data.
-Our results also suggest a rethink in they way LMA is handled in most growth models.
-In most existing models, biomass allocation is determined first (by some means of another), and then the amount of leaf area is calculated by multiplying leaf biomass allocation with LMA.
+Our results also suggest a rethink in the way LMA is handled in most growth models.
+In most existing models, biomass allocation is determined first (by some means or another), and then the amount of leaf area is calculated by multiplying leaf biomass allocation with LMA.
 Our results suggest the chain of causation should be reversed: first decide the amount of leaf area, then let LMA determine the cost of building this leaf.
 Such an approach has already been implemented in at least one model of plant growth \citep{falster2016}.
-Other algorithms are of course possible, but this at least one way to ensure model predictions recover the patterns reported in this paper.
+Other algorithms are of course possible, but this is at least one way to ensure model predictions recover the patterns reported in this paper.
 
@@
 Numbers indicate individual plants, with the number of unique species given in parentheses.
-Sample sizes differed between variables because not all included studied measured all variables"),
+Sample sizes differed between variables because not all included studied measured all variables."),
       sanitize.text.function=function(x){x},

### 2016-02-25 manuscript.Rnw
reference to visreg removed as we no longer use it

@@
 
-All analyses were conducted in R 3.2.3 \citep{R-2015}, using the packages `mgcv` \citep{wood2006}, `hier.part` \citep{walsh2013}, `MuMIn` \citep{barton2015}, `visreg` \citep{breheny2015}, and `remake` \citep{fitzjohn2015}.
+All analyses were conducted in R 3.2.3 \citep{R-2015}, using the packages `mgcv` \citep{wood2006}, `hier.part` \citep{walsh2013}, `MuMIn` \citep{barton2015},  and `remake` \citep{fitzjohn2015}.
 In addition, the code replicating this analysis (and all figures) is available at \url{http://github.com/RemkoDuursma/baadanalysis} (doi: \emph{to be added}).

### 2016-02-25 manuscript_suppinfo.Rnw
reference to visreg removed as we no longer use it

@@
 \usepackage{palatino}
+\usepackage[default,osfigures,scale=0.95]{opensans}
 \usepackage{geometry}
@@
     \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
-    \caption{Leaf area ratio ($\afmt$) by plant height (H) and PFT.
+    \caption{Leaf area ratio ($\afms$) by plant height (H) and PFT.
 Each symbol is an individual plant.
@@
 Note that LMM and GAM give very similar ranking, and even absolute variance components, for nearly all variables.
-Both methods identified AF/AS and MF/AF as the only variables with a practically significant dependence on climate (variance component \textgreater{} 0.1), see Fig S4.
+Both methods identified $\afas$ and LMA as the only variables with a practically significant dependence on climate (variance component \textgreater{} 0.1).
 \textbf{IEA}: independent effects analysis (aka hierarchical partitioning), using the algorithm by \citet{chevan1991}, as implemented in the `hier.part` package for R \citep{walsh2013}.
@@
 deciduous plant functional types.
-Lines are fits from mixed-effects models (with species within study as random effect, and only MAT and evergreen/deciduous classification as fixed effects).
+Lines are simple linear regression fits across averages by species within dataset, but symbols are individual plant observations.
 Solid lines are significant (p \textless{} 0.05), dashed lines are not (p \textgreater{} 0.05).
-Symbols are partial residuals as calculated with the `visreg` package in R \citep{breheny2015}.}
     \label{fig:figreichclimate}

### 2016-02-25 manuscript.Rnw
bug fix

@@
 
-All analyses were conducted in R 3.2.3 \citep{R-2015}, using the packages `mgcv` \citep{wood2006}, `hier.part` \citep{walsh2013}, `MuMIn` \citep{barton2015},  and `remake` \citep{fitzjohn2015}.
+All analyses were conducted in R 3.2.3 \citep{R-2015}, using the packages `mgcv` \citep{wood2006}, `hier.part` \citep{walsh2013}, `MuMIn` \citep{barton2015}, re and `remake` \citep{fitzjohn2015}.
 In addition, the code replicating this analysis (and all figures) is available at \url{http://github.com/RemkoDuursma/baadanalysis} (doi: \emph{to be added}).

### 2016-02-25 manuscript_suppinfo.Rnw
bug fix

@@
     \caption{Relationships between leaf - stem biomass distribution ($\mfms$) and mean annual temperature (MAT) (\degree C).
-Similar to Reich et al.
-(2014), we divided the data into five bins of plant size (bins were chosen so that each bin had equal sample size), and used only evergreen vs.
+Similar to \citet{reich2014}, we divided the data into five bins of plant size (bins were chosen so that each bin had equal sample size), and used only evergreen vs.
 deciduous plant functional types.
 Lines are simple linear regression fits across averages by species within dataset, but symbols are individual plant observations.
-Solid lines are significant (p \textless{} 0.05), dashed lines are not (p \textgreater{} 0.05).
+Solid lines are significant (p \textless{} 0.05), dashed lines are not (p \textgreater{} 0.05).}
     \label{fig:figreichclimate}

### 2016-02-25 manuscript.Rnw
Explain lower R2; 1-1 line in fig4; arrow in fig3

@@
 
-% In the top row of table 2, it's seems surprising that H,PFT,MAT explains less variance than just H,PFT.
-% Check!!!
 <<results='asis'>>=
@@
 Climate variables included were either MAT or aridity index (calculated as the of PET to MAP).
-P-values not shown because all smooth terms were always highly significant (p < 0.01), with the exception of aridity index in the model for $\\mfms$ (p = 0.057)."),sanitize.text.function=function(x){x},
+P-values not shown because all smooth terms were always  significant (p \\textless 0.01), with the exception of aridity index in the model for $\\mfms$ (p = 0.057).
+In some cases, the $R^2$ decreases after adding variables to the model, which is possible because it is the adjusted $R^2$ which discounts for the number of model parameters."),sanitize.text.function=function(x){x},
       include.rownames=FALSE, caption.placement = "top")
@@
 
-% Could we add along the x-axis of panel a, a small arrow showing the common height where we sued to generate panels b-c?
 \begin{figure}[h!]
@@
 Lines are generalised additive model fits.
+The arrow indicates the average plant height in the dataset.
 (b) and (c) $\mfms$ and leaf area ratio ($\afms$) at the average plant height in the dataset, estimated from fitted models in panel (a).
@@
     \fi
-    % proportional implies slope of ~1, should we add 1:1 line in background?
     \caption{Total plant leaf area ($\af$) is approximately proportional to basal stem area ($\as$), and this relationship is similar between PFTs.
-Lines are generalised additive model fits, with grey areas approximate 95\% confidence intervals.}
+Coloured lines are generalised additive model fits, with grey areas approximate 95\% confidence intervals.
+Black solid line is a 1:1 line.}
     \label{fig:alas}

### 2016-02-25 R/figures.R

@@ -200,6 +200,9 @@ figure3 <- function(dataset, KGAM=4){
   set.seed(42) # Set seed to ensure plot looks same each time
   dataset <- dataset[sample(nrow(dataset)),]
 
+  # Evaluate biomass distribution at mean height
+  xpred <- mean(dataset$lh.t, na.rm=TRUE)
+  
   l <- layout(matrix(c(1,2,1,3), byrow=T, ncol=2),
               widths=c(1,0.67), heights=c(1,1))
   
@@ -214,10 +217,13 @@ figure3 <- function(dataset, KGAM=4){
   box()
   plotlabel("(a)","topright")
   
+  # Arrow indicating where mean biomass distribution was estimaed
+  par(xpd=NA)
+  arrows(x0=xpred, x1=xpred, y0=-3.5, y1=-3.2, length=0.1, lwd=2)
 
   par(mar=c(0.35,4,4,1), pty="m")
   
-  xpred <- mean(dataset$lh.t, na.rm=TRUE)
+  
   plotGamPred(obj1, dataset, xpred=xpred,
               ylab=expression(M[F]/M[S]~(kg~kg^-1)),ylim=c(0,0.5),
               xlim=c(0,0.2), xlab="", xaxislabels=FALSE)
@@ -246,6 +252,7 @@ figure4 <- function(dataset){
                      pch=16, axes=FALSE,
                      pointcols=my_cols_transparent(),
                      linecols=my_cols(),
+                     panel.first=abline(0,1),
                      xlab=expression(A[S]~~(m^2)),
                      ylab=expression(A[F]~~(m^2)))
   log10axes()

### 2016-02-25 manuscript.Rnw
Clarify Eq. 1 and 2

@@
 
-% This next refers to stem xs-area, but then the equation is for diameter. Did we really use both?
 In our dataset, stem cross-sectional area was measured either at breast height (1.3 - 1.34m) and/or at the base of the plant.
@@
 where $D_{\rm S}$ is the basal stem diameter (m), $D_{\rm Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{\rm Sbh}$ was measured (typically 1.3 or 1.34m).
+Stem area and diameter were always related assuming the stem was circular.
+We chose to estimate missing stem diameter rather than stem area because a much better fit was obtained, with a more constant error variance.
+
 The parameter $c$ was further expressed as a function of plant height:
@@
 4).
-We fit a GAM to $\af$ as a function of $\as$ with and without PFT as a covariate
-% please clarify next line
-(both as intercept and slope interaction),
-which showed PFT significantly improved model fit (likelihood ratio test, p \textless{} 0.001).
+We fit a GAM to $\af$ as a function of $\as$ with and without PFT as a covariate, which showed PFT significantly improved model fit (likelihood ratio test, p \textless{} 0.001).
 Nevertheless, the $R^2$ of the relationship only increased from \Sexpr{round(100*afas$R2a,2)} to \Sexpr{round(100*afas$R2b,2)}\%, indicating that while significant, the PFT had little additional predictive power.
@@
 
-% This paragraph has been substantially changed, please review
 We have confirmed the hypothesis that the amount of leaf area supported per plant is less variable between PFTs than the amount of leaf mass.

### 2016-02-25 R/data_processing.R

@@ -265,9 +265,11 @@ prepare_dataset_1 <- function(baad, plantations=TRUE){
 BasalA_fit <- function(baad){
 
   # Predicted basal diameter. See R/predict_dba...R
-  test <- subset(baad$data, !is.na(d.ba) & !is.na(d.bh) & !is.na(h.t) & !is.na(h.bh) & h.t > h.bh)
+  test <- subset(baad$data, !is.na(a.stba) & !is.na(a.stbh) & !is.na(h.t) & !is.na(h.bh) & h.t > h.bh)
   nls(d.ba ~ d.bh * h.t^(c0*h.t^c1) /(h.t - h.bh)^(c0*h.t^c1), start=list(c0=0.9, c1=0.7),
-             data=test)
+      data=test)
+  # Fit on diameter not cross-sectional area because a much better fit is obtained this way (not sure why!),
+  # and more homogenous error variance (because large cross-sectional area outliers)
 }
 
 predictBasalA <- function(dat, baad){

### 2016-02-29 manuscript.Rnw
Minor fixes; added bloom 2016

@@
 Globally, forests store approximately 360Pg of carbon in living biomass \citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}.
-Reducing uncertainties about biomass distribution and carbon residence times is a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014,negron2015}, in particular because projections by global vegetation models (GVMs) are particularly sensitive to this component \citep{friend2014,ise2010}.
+Reducing uncertainties about biomass distribution and carbon residence times is a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014,negron2015,bloom2016}, in particular because projections by global vegetation models (GVMs) are particularly sensitive to this component \citep{friend2014,ise2010}.
 
@@
 \subsection*{Data analysis}
-To better understand the mechanisms of PFT and climate-related variation in biomass distribution, we decomposed the ratio of leaf-to stem biomass it into three component variables.
-Applying the chain rule of differential calculus gives
+To better understand the mechanisms of PFT and climate-related variation in biomass distribution, we decomposed the ratio of leaf-to stem biomass it into three component variables:
 \begin{equation}
@@
 
-% Read \citep{bloom2016}; global patterns in inferred allocation and biomass pools (leaf / stem).
-
 Using the largest database of individual woody plant allometry to date, we studied patterns in above-ground biomass distribution among major woody plant functional types (PFTs), plant size, and climate.
@@
 Indeed, in a global compilation of LAI for undisturbed forest stands, \citet{iio2014} reported a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI.
-Moreover, we would in fact expect LAI to change with LMA, not because of changes in the amount of leaf area maintained by individual plants, but because LMA influences the shade-tolerance of seedlings \citep{poorter2006, falster2011}, and thereby the LAI of the community \citep{falster2011}.
+Moreover, we would in fact expect LAI to change with LMA, not because of changes in the amount of leaf area maintained by individual plants, but because LMA correlates with the shade-tolerance of seedlings \citep{poorter2006, falster2011}, and thereby the LAI of the community \citep{falster2011}.
 The ultimate reason LMA might effect shade tolerance is because is via its correlation with leaf lifespan \citep{wright2004}.
@@
 \textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\
-\textbf{Fig. S2} Leaf area ratio ($\af/M_{\rm T}$) by PFT.\\
-\textbf{Fig. S3} Comparison of three methods for variance partitioning of the four studied variables \\
+\textbf{Fig. S2} Leaf area ratio ($\afms$) by plant height and PFT.\\
+\textbf{Fig. S3} Comparison of three methods for variance partitioning of the four studied variables. \\
 \textbf{Fig. S4} Whole-plant leaf mass ($\mf$) and leaf area ($\af$) as a function of total aboveground biomass. \\

### 2016-02-29 manuscript.Rnw
Typo fixes; added sitch2003

@@
 Although recent large compilations on biomass distribution included data on leaf mass, they did not include measurements of leaf area so were unable to evaluate these alternatives \citep{poorter2012,poorter2015,reich2014}.
-Here we set out to directly test the hypothesis that across PFTs, the ratio of leaf mass to stem mass -- $\mfms$ -- is related to LMA, but the ratio of leaf area to stem mass is invariant with respect to to LMA.
+Here we set out to directly test the hypothesis that across PFTs, the ratio of leaf mass to stem mass -- $\mfms$ -- is related to LMA, but the ratio of leaf area to stem mass is invariant with respect to LMA.
 
@@
 Here we limit the analysis to patterns in aboveground biomass distribution.
-Our analysis has several parts; to maximise the amount of available data for each part we we used different subsets of the data, as not all variables were measured in each study.
+Our analysis has several parts; to maximise the amount of available data for each part we used different subsets of the data, as not all variables were measured in each study.
 Sample sizes by PFT are summarised in Table \ref{table:counttable}.
@@
 
-All analyses were conducted in R 3.2.3 \citep{R-2015}, using the packages `mgcv` \citep{wood2006}, `hier.part` \citep{walsh2013}, `MuMIn` \citep{barton2015}, re and `remake` \citep{fitzjohn2015}.
+All analyses were conducted in R 3.2.3 \citep{R-2015}, using the packages `mgcv` \citep{wood2006}, `hier.part` \citep{walsh2013}, `MuMIn` \citep{barton2015}, and `remake` \citep{fitzjohn2015}.
 In addition, the code replicating this analysis (and all figures) is available at \url{http://github.com/RemkoDuursma/baadanalysis} (doi: \emph{to be added}).
@@
 As a result, the amount of leaf area per unit woody above-ground biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater{} 0.05, see Fig.
-3c), because the differences in $\mfas$ were exactly compensated by differences in LMA (see also Fig.
+3c), because the differences in $\mfms$ were exactly compensated by differences in LMA (see also Fig.
 S2 for $\afms$ as a function of plant height).
@@
 Moreover, we would in fact expect LAI to change with LMA, not because of changes in the amount of leaf area maintained by individual plants, but because LMA correlates with the shade-tolerance of seedlings \citep{poorter2006, falster2011}, and thereby the LAI of the community \citep{falster2011}.
-The ultimate reason LMA might effect shade tolerance is because is via its correlation with leaf lifespan \citep{wright2004}.
-For whole stands, we can therefore expect total stand leaf mass to be correlated with LMA, both because of of greater leaf mass per plant \citep[e.g.]{reich1992}, and the greater density of plants arising from increased leaf lifespan.
+The ultimate reason LMA might affect shade tolerance is because is via its correlation with leaf lifespan \citep{wright2004}.
+For whole stands, we can therefore expect total stand leaf mass to be correlated with LMA, both because of greater leaf mass per plant \citep[e.g.]{reich1992}, and the greater density of plants arising from increased leaf lifespan.
 Previous interpretations of differences in $\mfms$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002, poorter2012}.
@@
 Our results suggest the chain of causation should be reversed: first decide the amount of leaf area, then let LMA determine the cost of building this leaf.
-Such an approach has already been implemented in at least one model of plant growth \citep{falster2016}.
+Such an approach has already been implemented in at least two vegetation models \citep{falster2016,sitch2003}.
 Other algorithms are of course possible, but this is at least one way to ensure model predictions recover the patterns reported in this paper.

### 2016-04-19 manuscript.Rnw
Many revisions for 2nd round at NewPhyt

@@
 \usepackage{breakurl}
+
+% Avoid any hyphenation (for conversion to Word)
+\usepackage[none]{hyphenat}
+
 \geometry{verbose,tmargin=2.4cm,bmargin=2.4cm,lmargin=2.4cm,rmargin=2.4cm}
@@
 \item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species.
-We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, and climate on biomass distribution and three component variables.
-\item The ratio of leaf mass to woody above-ground mass ($\mfms$) varied strongly between PFTs.
+We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, and climate on biomass distribution.
+\item The ratio of leaf mass to above-ground woody mass ($\mfms$) varied strongly between PFTs.
 We found that $\mfms$, once corrected for plant height, was proportional to leaf mass per area across PFTs.
@@
 Perhaps the biggest challenge is to capture the combined responses to environmental change of the more than 250,000 plant species comprising the world's vegetation.
-While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of biomass invested in different tissues (leaves, stems, roots).
-One way to capture and understand the source of this functional diversity, is to consider only a few archetypal plant functional types (PFTs), assumed to differ in key attributes \citep{harrisonetal2010,wullschleger2014},.
+While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of biomass invested in different tissues (leaves, stems, roots) \citep{poorter2012}.
+One way to capture and understand the source of this functional diversity, is to consider only a few archetypal plant functional types (PFTs), assumed to differ in key attributes \citep{harrisonetal2010,wullschleger2014}.
 Previous work has shown consistently large differences in biomass distribution and related traits among major PFTs \citep{poorter2009,enquistniklas2002,poorter2012,poorter2015,reich2014,mccarthyetal2007}, but these differences are yet to be fully understood.
@@
 If this is the case, the total leaf area of plants at a standardised size should be less variable between angiosperms and gymnosperms than total leaf mass.
-The idea that high total leaf mass may compensate for high LMA would also explain the difference between evergreen (low LMA, low leaf mass) and deciduous (high LMA, high leaf mass) species \citep{poorter2015}.
+The idea that high total leaf mass may compensate for high LMA would also explain the difference between deciduous (low LMA, low leaf mass) and evergreen (high LMA, high leaf mass) species \citep{poorter2015}.
 Some oft-cited studies make a different assumption: that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
 Although recent large compilations on biomass distribution included data on leaf mass, they did not include measurements of leaf area so were unable to evaluate these alternatives \citep{poorter2012,poorter2015,reich2014}.
-Here we set out to directly test the hypothesis that across PFTs, the ratio of leaf mass to stem mass -- $\mfms$ -- is related to LMA, but the ratio of leaf area to stem mass is invariant with respect to LMA.
+
+To better understand the mechanisms of PFT-related variation in biomass distribution, we decomposed the ratio of leaf to stem biomass into three component variables:
+\begin{equation}
+\label{eq:mfmspart}
+\frac{\mf}{\ms} = \frac{\mf}{\af} \cdot \frac{\af}{\as} \cdot \frac{\as}{\ms},
+\end{equation}
+where $\mfaf$ is the LMA, $\afas$ the ratio of plant leaf area to basal stem area, and $\asms$ the ratio of stem basal area to above-ground woody mass. 
+\textbf{Develop hypotheses here a bit more, integrate above equation}.
+Here we set out to directly test the hypothesis that across PFTs, the ratio of leaf mass to stem mass -- $\mfms$ -- is related to LMA, whereas the ratios of stem mass to stem cross-sectional area ($\msas$) and of leaf area to stem area ($\afas$) are invariant with respect to LMA.
 
@@
 We included only field-grown woody plants in our analysis (including natural vegetation, unmanaged, and managed plantations), excluding glasshouse and common garden studies.
-Excluding plantations from the analysis did not affect the main result, but far fewer data were available for evergreen gymnosperms.
+We did not exclude plantations from the analysis because this would have removed many data points for evergreen gymnosperms, and excluding plantations did not did not affect the main results.
 We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms.
 We excluded  deciduous gymnosperms because insufficient data were available.
-To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature (MAT) and precipitation (MAP) from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).
-MAT and MAP were also extracted for all study locations.
-Potential evapotranspiration (PET) was extracted from the Global-PET database (http://www.cgiar-csi.org, \citet{zomer2008}), and an aridity index was calculated as PET/MAP.
+
+To test for climate effects on biomass distribution, we estimated mean annual temperature (MAT), precipitation (MAP) and Potential evapotranspiration (PET) for each study location.
+MAT and MAP were extracted from Worldclim \citep{worldclim}, PET from the Global-PET database (http://www.cgiar-csi.org, \citet{zomer2008}), and an aridity index was calculated as PET/MAP.
+To assess the coverage of the global climate space by the dataset, we also extracted MAT and MAP across the globe but excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).
 
@@
 \end{equation}
-The estimated coefficients were $c_0$ = \Sexpr{signif(coef(basalafit)[1],3)}, $c_1$ = \Sexpr{signif(coef(basalafit)[2],3)}, root-mean square error = \Sexpr{signif(summary(basalafit)$sigma,3)}.
+The estimated coefficients were $c_0$ = \Sexpr{signif(coef(basalafit)[1],3)}, $c_1$ = \Sexpr{signif(coef(basalafit)[2],3)}, root-mean square error = \Sexpr{signif(summary(basalafit)$sigma,3)}, R$^2$ = \Sexpr{signif(nlsr2(basalafit),3)}.
 The relationship was fit using a total of \Sexpr{length(residuals(basalafit))} observations covering the three major PFTs.
@@
 \subsection*{Data analysis}
-To better understand the mechanisms of PFT and climate-related variation in biomass distribution, we decomposed the ratio of leaf-to stem biomass it into three component variables:
-\begin{equation}
-\label{eq:mfmspart}
-\frac{\mf}{\ms} = \frac{\mf}{\af} \cdot \frac{\af}{\as} \cdot \frac{\as}{\ms},
-\end{equation}
-where $\mfaf$ is the LMA, $\afas$ the ratio of plant leaf area to basal stem area, and $\asms$ the ratio of stem basal area to total woody above-ground mass.
-Using variance partitioning, we quantified the contribution of plant height, plant functional type, and climate to the overall variation in $\mfms$ and for all three component variables.
 
-Because, similar to \citet{poorter2015}, none of the studied relationships were linear on a log-log scale, we did not use standard allometric equations to study relationships between size and biomass variables.
+Similar to \citet{poorter2015},we found that none of the studied relationships were linear on a log-log scale.
+Therefore, we did not use standard allometric equations to study relationships between size and biomass variables.
 Instead, we used generalised additive models (GAM), a semi-parametric modelling approach that makes no assumption about the shape of the relationships.
@@
 
+Using variance partitioning, we quantified the contribution of plant height, plant functional type, and climate to the overall variation in $\mfms$ and for all three component variables.
 The amount of variance explained by quantitative climate variables (MAP and MAT) was assessed with GAMs, with variables were sequentially added to the model and the explained variance ($R^2$) calculated.
@@
 
-All analyses were conducted in R 3.2.3 \citep{R-2015}, using the packages `mgcv` \citep{wood2006}, `hier.part` \citep{walsh2013}, `MuMIn` \citep{barton2015}, and `remake` \citep{fitzjohn2015}.
-In addition, the code replicating this analysis (and all figures) is available at \url{http://github.com/RemkoDuursma/baadanalysis} (doi: \emph{to be added}).
+All analyses were conducted in R 3.2.4 \citep{R-2015}, using the packages `mgcv` \citep{wood2006}, `hier.part` \citep{walsh2013}, `MuMIn` \citep{barton2015}, and `remake` \citep{fitzjohn2015}.
+In addition, the code replicating this analysis (and all figures) is available at \url{http://github.com/RemkoDuursma/baadanalysis} (doi: \emph{will be added at proof stage}).
 
@@
 These differences were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater{} evergreen angiosperm \textgreater{} deciduous angiosperm.
-Next we calculated $\mfms$ at a common plant height for each of the PFTs.
+Next we calculated $\mfms$ at a common plant height for each of the PFTs (the mean height across all data, \Sexpr{format(10^mean(dataset$lh.t, na.rm=TRUE), nsmall=1, digits=2)}m).
 These three size-corrected $\mfms$ values were highly significantly different between PFTs (p \textless{} 0.001), and were approximately proportional to the average LMA ($\mfaf$) across the PFTs (Fig.
 3b).
-As a result, the amount of leaf area per unit woody above-ground biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater{} 0.05, see Fig.
+As a result, the amount of leaf area per unit above-ground woody biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater{} 0.05, see Fig.
 3c), because the differences in $\mfms$ were exactly compensated by differences in LMA (see also Fig.
@@
 Next, we decomposed $\mfms$ into three component variables with distinct biological interpretation (Eq.
-\ref{eq:mfmspart}).
+\ref{eq:mfmspart}): LMA ($\mfaf$) is an important leaf trait, $\afas$ is a measure of hydraulic supply and mechanical support per unit leaf area, and $\asms$ depends on stem form and wood density.
 For each of the component variables and $\mfms$, we estimated the variance explained by plant height, PFT, and climate variables (MAT or aridity) (Table \ref{table:varparttable}).
 We did this by sequentially adding terms to a GAM, and calculating the $R^2$ for each fitted model.
-The first model included only plant height, and already explained much variation (65 - 79\%) in $\mfms$ and $\msas$ but not LMA and $\afas$ (4 - 11 \%).
+The first model included only plant height, and already explained much variation (65 - 79\%) in $\mfms$ and $\asms$ but not LMA and $\afas$ (4 - 11 \%).
 When PFT was added to the model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $\mfms$ and $\mfaf$ but not $\afas$ or $\asms$, demonstrating that PFT differences entered largely in leaf mass-based variables.
@@
 We fit a GAM to $\af$ as a function of $\as$ with and without PFT as a covariate, which showed PFT significantly improved model fit (likelihood ratio test, p \textless{} 0.001).
-Nevertheless, the $R^2$ of the relationship only increased from \Sexpr{round(100*afas$R2a,2)} to \Sexpr{round(100*afas$R2b,2)}\%, indicating that while significant, the PFT had little additional predictive power.
+Nevertheless, the $R^2$ of the relationship only increased from \Sexpr{round(afas$R2a,3)} to \Sexpr{round(afas$R2b,3)}, indicating that while significant, the PFT had little additional predictive power.
 
@@
 We fitted a GAM of $\ms$ as a function of $\as$ (added as a smooth term) with or without PFT as a covariate.
-Again PFT was formally significant (likelihood ratio test, p \textless{} 0.001), but the $R^2$ increased only marginally from \Sexpr{round(100*msas$R2a,2)} to \Sexpr{round(100*msas$R2b,2)}\%, again showing little practical significance of PFT in explaining this variable.
+Again PFT was formally significant (likelihood ratio test, p \textless{} 0.001), but the $R^2$ increased only marginally from \Sexpr{round(msas$R2a,3)} to \Sexpr{round(msas$R2b,3)}, again showing little practical significance of PFT in explaining this variable.
 
@@
 Our compilation is the first of its kind that also includes leaf area estimates on the same plants, allowing a direct test of the hypothesis that leaf and stem biomass distribution is related to the LMA.
-We found that at a common plant height, the leaf / woody biomass ratio ($\mfms$) was proportional to the LMA across three PFTs, so that leaf area per unit above-ground biomass did not vary between PFTs.
+We found that at a common plant height, the leaf / above-ground woody biomass ratio ($\mfms$) was proportional to the LMA across three PFTs, so that leaf area per unit above-ground woody biomass did not vary between PFTs.
 We found very weak effects of climate (mean annual temperature and aridity) on biomass distribution, so that the vast majority of the variation in biomass distribution was explained by plant size and PFT.
 
-As expected, $\mfms$ steeply declined with plant height.
+We found that $\mfms$ steeply declined with plant height.
 It is well known that it is necessary to correct for plant size when comparing biomass distribution among groups \citep{mcconnaughay1999}.
@@
     \caption{Dominant woody PFTs differ in above-ground biomass distribution due to underlying differences in leaf mass per area.
-(a) Leaf mass per unit woody above-ground biomass ($\mfms$) by PFT.
+(a) Leaf mass per unit above-ground woody biomass ($\mfms$) by PFT.
 Each symbol is an individual plant.

### 2016-04-22 manuscript.Rnw
switch to using baad.data package for access to data
Restructure introduction & results to emphasise hypotheses
- as requested by referees, move decomposition equation into introduction
- use this as basis for outlining hypotheses
- add tests in relation to af/ms as major result
- remove stem mass index from results and figures

Also, suggest

- moving Fig 6 - > Fig 4
- embedding Fig S2 in Fig 3

@@
 
-\title{Leaf mass per area and plant size drive differences in above-ground biomass distribution among woody plant functional types}
+\title{Leaf mass per area, not total leaf area, drives differences in above-ground biomass distribution among woody plant functional types}
 
@@
 
-\affil[1]{Hawkesbury Institute for the Environment, Western Sydney University, Locked Bag 1797, Penrith, NSW 2751, Australia}
-\affil[2]{Biological Sciences, Macquarie University, North Ryde, NSW 2109, Australia}
+\affil[1]{Hawkesbury Institute for the Environment, Western Sydney University, Locked Bag 1797, Penrith NSW 2751, Australia}
+\affil[2]{Biological Sciences, Macquarie University NSW 2109, Australia}
 
@@
 opts_chunk$set(echo=FALSE)
-baad <- readRDS("downloads/baad.rds")
 @
@@
 
-Perhaps the biggest challenge is to capture the combined responses to environmental change of the more than 250,000 plant species comprising the world's vegetation.
+Perhaps the biggest challenge in understanding and predicting biomass distribution is to capture the combined responses to environmental factors of the more than 250,000 plant species comprising the world's vegetation.
 While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of biomass invested in different tissues (leaves, stems, roots) \citep{poorter2012}.
-One way to capture and understand the source of this functional diversity, is to consider only a few archetypal plant functional types (PFTs), assumed to differ in key attributes \citep{harrisonetal2010,wullschleger2014}.
-Previous work has shown consistently large differences in biomass distribution and related traits among major PFTs \citep{poorter2009,enquistniklas2002,poorter2012,poorter2015,reich2014,mccarthyetal2007}, but these differences are yet to be fully understood.
+The challenge, therefore, is to identify the key traits driving differences among species.
+One way to start capturing and understanding the sources of this functional diversity, is to consider a few archetypal plant functional types (PFTs) \citep{harrisonetal2010, wullschleger2014}.
+Previous work has shown that despite widespread variation within PFTs for some traits and outcomes, others -- including biomass distribution -- vary substantially among PFTs \citep{chabothicks1982,oneilldeangelis1981,bond1989,enquistniklas2002,poorter2009,reich2014,poorter2012,poorter2015}.
 
-One puzzling aspect of the large variation in biomass distribution among PFTs is the much larger leaf mass in gymnosperms compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012, poorter2015}.
-One possible explanation is that gymnosperms compensate for their high leaf mass per area (LMA) \citep{poorter2009} by maintaining a large amount of leaf biomass, in order to maintain sufficient total leaf area.
-If this is the case, the total leaf area of plants at a standardised size should be less variable between angiosperms and gymnosperms than total leaf mass.
-The idea that high total leaf mass may compensate for high LMA would also explain the difference between deciduous (low LMA, low leaf mass) and evergreen (high LMA, high leaf mass) species \citep{poorter2015}.
-Some oft-cited studies make a different assumption: that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.
-Although recent large compilations on biomass distribution included data on leaf mass, they did not include measurements of leaf area so were unable to evaluate these alternatives \citep{poorter2012,poorter2015,reich2014}.
+The goal of the current work is to better understand differences among PFTS in their biomass distribution.
+Past work has shown that gymnosperms tend to hold a much larger leaf mass compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012,poorter2015}.
+The cause of these differences are not yet fully understood, but could arise from either differences leaf mass per area (LMA) \citep{poorter2009}, or the amount of leaf area maintained \citep{chabothicks1982,bond1989}, or both.
+It is also unknown whether there exist differences among finer-scale groupings, for example between evergreen and deciduous angiosperms.
+While past analyses have often spanned many sites and / or species, researchers have been unable to further investigate the factors underlying observed differences in biomass distribution among PFTs, because available data only included a limited number of variables.
 
-To better understand the mechanisms of PFT-related variation in biomass distribution, we decomposed the ratio of leaf to stem biomass into three component variables:
+Here, we use the recently-compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which includes observations of biomass and other size metric for \Sexpr{nrow(baad_all$data)} individual plants (Fig.
+1), to establish the underlying cause of biomass distribution among PFTs, and also to test for systematic shifts in biomass distribution with climate.
+We focus on the distribution of biomass among leaf and above-ground woody tissues in individual plants.
+To quantify this distribution we analyse two related metrics: 1) the ratio of leaf mass to stem mass, $\mfms$; and 2) the ratio of leaf area to stem mass, $\afms$.
+To understand the factors driving variation in these quantities, we decompose these ratios into component variables, each of which can be quantified in our dataset.
+The ratio of leaf mass to stem mass, can decomposed as
 \begin{equation}
 \label{eq:mfmspart}
-\frac{\mf}{\ms} = \frac{\mf}{\af} \cdot \frac{\af}{\as} \cdot \frac{\as}{\ms},
+\frac{\mf}{\ms} =  \frac{\mf}{\af} \times \frac{\af}{\as} \times \frac{\as}{\ms},
 \end{equation}
-where $\mfaf$ is the LMA, $\afas$ the ratio of plant leaf area to basal stem area, and $\asms$ the ratio of stem basal area to above-ground woody mass. 
-\textbf{Develop hypotheses here a bit more, integrate above equation}.
-Here we set out to directly test the hypothesis that across PFTs, the ratio of leaf mass to stem mass -- $\mfms$ -- is related to LMA, whereas the ratios of stem mass to stem cross-sectional area ($\msas$) and of leaf area to stem area ($\afas$) are invariant with respect to LMA.
+where $\mfaf$ is the LMA, $\afas$ the ratio of total plant leaf area to basal stem area, and $\asms$ is the ratio of stem basal area to above-ground woody mass.
+Similarly, the ratio of leaf area to stem mass can decomposed as
+\begin{equation}
+\label{eq:afmspart}
+\frac{\af}{\ms} =  \frac{\af}{\as} \times \frac{\as}{\ms}.
+\end{equation}
+The terms $\mfaf$ and $\afas$ are themselves prominent traits known to vary among species \citep{wright2004,togashi2015}, suggesting a potential to explain differences in $\mfms$ among PFTs.
+The third term is known to vary strongly with individual size, decreasing as a plant grows.
+However, \citet{chave2014} found that a single model of $\ms$ incorporating $\as$, wood density, and $H$ to hold across diverse tropical trees, irrespective of vegetation type, or environmental factors.
 
-Little is known about global-scale patterns in biomass distribution in relation to climate.
-One might expect that biomass distribution between leaves and stems is correlated with precipitation or mean annual temperature, as small-scale comparisons have shown such responses \citep{berninger1994,callaway1994,delucia2000,poyatos2007}.
-Recently, \citet{reich2014} demonstrated that the ratio of leaf to total plant mass (for whole stands, either including or excluding below-ground components) increased with mean annual temperature, after separating the data into gymnosperm and angiosperm stands, and accounting for the large size-related variation in biomass distribution.
-Thus, our second objective was  to test for climate-related variation in above-ground biomass distribution, and specifically to test whether $\mfms$ increases with mean annual temperature.
+These decompositions, combined with ideas from past studies, lead us to test the following hypotheses about biomass distribution:
 
-To test our hypotheses, we use a recently compiled large database \citep{falster2015} which includes observations of biomass and size metrics for individual plants (Fig.
-1), compiled from 176 studies spanning nine vegetation types (Fig.
-S1) and three major climate zones (boreal, temperate, and tropical).
-In this paper we focus on field-grown woody plants, spanning the entire size range of woody plants (0.01 - \textgreater{} 100 m height).
+\begin{enumerate}
+  \item Individual height captures the most majority of variation in $\mfms$ and $\afms$, reflecting the fact that height can vary several orders of magnitude within the lifespan of a single individual and thus represents the dominant form of variation in eq .
+\ref{eq:mfmspart}-\ref{eq:afmspart}.
+  \item That these height-related effects on biomass distribution are concentrated in the term $\asms$, and that while $\asms$ may vary considerably among individuals, there are no systematic differences in $\asms$ among PFTs.
+  \item That after controlling for the effects of individual height, differences in $\mfms$ among PFTs can arise due to differences in either, or both, of a) LMA, because PFTs tend to differ LMA \citep{poorter2009}, or b)$\afas$, because gymnosperms are hypothesised to maintain a greater leaf area than angiosperms \citep{chabothicks1982, bond1989}.
+  \item That any differences in $\afas$ among PFTs also leads to differences in $\afms$.
+\end{enumerate}
+
+A final aim was test for significant shifts in biomass distribution with climate.
+Little is known about global-scale patterns in biomass distribution in relation to climate, but one might expect biomass distribution to shift with precipitation or mean annual temperature, as small-scale comparisons have shown such responses \citep{berninger1994,callaway1994,delucia2000,poyatos2007}.
+\citet{reich2014} has also shown that after controlling  for the large size-related variation in biomass distribution, the ratio of leaf to total plant mass (averaged across whole stands, either including or excluding below-ground components) increases with mean annual temperature in both gymnosperms and angiosperms.
+Thus, we aimed to test for climate-related variation in $\mfms$ and $\afms$, and their component terms with PFTS.
 
@@
 
-We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total includes records for \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species.
-The database has very limited overlap (n = 261 individuals, 1.7 \%) with the recent large compilation of \citet{poorter2015} and differs in that measurements are all for individual plants (where \citet{poorter2015} included many stand-based averages).
-In this paper we restrict our analysis to records that include leaf mass ($\mf$), leaf area ($\af$), above-ground woody biomass ($\ms$), plant height ($H$), and stem area measured at ground level ($\as$), or at breast height (typically 1.3m) ($\asb$) (n=\Sexpr{nrow(dataset)}).
-The database contains many more variables, for example root biomass; but usually for a much smaller subset of studies.
-Here we limit the analysis to patterns in aboveground biomass distribution.
+We used the recently-compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total includes records for \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species.
+The database has very limited overlap (n = 261 individuals, 1.7 \%) with the recent large compilation of \citet{poorter2015} and differs in that measurements are all for individual plants (where \citet{poorter2015} included many stand-based averages) and also includes other metrics like leaf area, in additional to biomass.
+We restrict our analysis to records that include leaf mass ($\mf$), leaf area ($\af$), above-ground woody biomass ($\ms$), plant height ($H$), and stem area measured at ground level ($\as$), or at breast height (typically 1.3m) ($\asb$) (n=\Sexpr{nrow(dataset)}).
 Our analysis has several parts; to maximise the amount of available data for each part we used different subsets of the data, as not all variables were measured in each study.
@@
 We did not exclude plantations from the analysis because this would have removed many data points for evergreen gymnosperms, and excluding plantations did not did not affect the main results.
-We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms.
-We excluded  deciduous gymnosperms because insufficient data were available.
+We considered three PFTs: evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms.
+We excluded  deciduous gymnosperms because of insufficient data.
 
@@
 MAT and MAP were extracted from Worldclim \citep{worldclim}, PET from the Global-PET database (http://www.cgiar-csi.org, \citet{zomer2008}), and an aridity index was calculated as PET/MAP.
-To assess the coverage of the global climate space by the dataset, we also extracted MAT and MAP across the globe but excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).
+To assess the coverage of the global climate space by the dataset, we also extracted MAT and MAP across the globe from the global land cover database GLC-SHARE \citep{lathamglobal},but excluding areas without woody vegetation.
 
@@
 We therefore converted records where area was measured at breast height only to area at base, using a relationship fitted to individuals where both measurements existed.
-For the subset of the data where both were measured, we estimated basal stem area ($\as$) from breast height stem area ($\asb$) from the equation,
+For the subset of the data where both areas were measured, we estimated basal stem area ($\as$) from breast height stem area ($\asb$) from the equation,
 \begin{equation}
-D_{\rm S} = D_{\rm Sbh} \left(\frac{H}{H - H_{\rm bh}}\right)^c
+\label{eq:dia}
+D_{\rm S} = D_{\rm Sbh} \left(\frac{H}{H - H_{\rm bh}}\right)^c,
 \end{equation}
-where $D_{\rm S}$ is the basal stem diameter (m), $D_{\rm Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{\rm Sbh}$ was measured (typically 1.3 or 1.34m).
+where $D_{\rm S}$ is the basal stem diameter (m), $D_{\rm Sbh}$ is the stem diameter at breast height, and $H_{bh}$ is the height at which $D_{\rm Sbh}$ was measured (typically 1.3 or 1.34m).
 Stem area and diameter were always related assuming the stem was circular.
 We chose to estimate missing stem diameter rather than stem area because a much better fit was obtained, with a more constant error variance.
-
-The parameter $c$ was further expressed as a function of plant height:
+The parameter $c$ in eq.
+\ref{eq:dia} was further expressed as a function of plant height:
 \begin{equation}
@@
 
-Similar to \citet{poorter2015},we found that none of the studied relationships were linear on a log-log scale.
+Similar to \citet{poorter2015}, we found that none of the studied relationships were linear on a log-log scale.
 Therefore, we did not use standard allometric equations to study relationships between size and biomass variables.
-Instead, we used generalised additive models (GAM), a semi-parametric modelling approach that makes no assumption about the shape of the relationships.
-GAMs were used to visualise the relationships between biomass and plant size variables, and to estimate variables such as $\mfms$ at a common plant height.
+Instead, we used generalised additive models (GAM), a semi-parametric modelling approach that makes no assumption about the shape of the relationships, to visualise the relationships between biomass and plant size variables, and to estimate variables such as $\mfms$ and $\afas$ at a common plant height.
 In all fitted GAMs, we used a cubic regression spline.
@@
 Using variance partitioning, we quantified the contribution of plant height, plant functional type, and climate to the overall variation in $\mfms$ and for all three component variables.
-The amount of variance explained by quantitative climate variables (MAP and MAT) was assessed with GAMs, with variables were sequentially added to the model and the explained variance ($R^2$) calculated.
-We compared this approach when using linear mixed-effects models instead of GAMs \citep[calculating the $R^2$ for linear mixed-effects models for the fixed effects only following][]{nakagawa2013}, as well as another method, independent effects analysis.
+The amount of variance explained by quantitative climate variables (MAP and MAT) was assessed with GAMs, with variables sequentially added to the model and the explained variance ($R^2$) calculated.
+We also compared this approach to another method using linear mixed-effects models instead of GAMs \citep[calculating the $R^2$ for linear mixed-effects models for the fixed effects only following][]{nakagawa2013}, as well as third method called ``independent effects analysis''.
 The three methods generally agreed on the ranking of variable importance (see Fig.
@@
 mfmslarge <- round(with(dataset, 10^mean(lmlf_mst[h.t > 30], na.rm=TRUE)),3)
+afmssmall <- round(with(dataset, 10^mean(lalf_mst[h.t < 0.1], na.rm=TRUE)),1)
+afmslarge <- round(with(dataset, 10^mean(lalf_mst[h.t > 30], na.rm=TRUE)),3)
 @
@@
 2).
-For both variables the relationship was highly non-linear on a log-log scale, demonstrating departure from simple allometric relationships.
+For both variables the relationship was non-linear on a log-log scale, demonstrating departure from simple allometric relationships.
 It is also evident that $\ms$ increases much quicker with plant height than $\mf$.
-As a result, the ratio $\mfms$ quickly decreased with plant height (Fig.
+As a result, the ratio $\mfms$ decreased with plant height (Fig.
 3a), from an average of \Sexpr{mfmssmall} for very small saplings (\textless{} 0.1m) to an average of \Sexpr{mfmslarge} for large trees (\textgreater{} 30m).
+Similarly, $\afms$ decreased with plant height (Fig.
+S2), from an average of \Sexpr{afmssmall} for very small saplings to an average of \Sexpr{afmslarge} for large trees.
 
-We found large differences between the three major PFTs in terms of above-ground biomass distribution,a s quantified by the ratio $\mfms$ (Fig.
-3a).
-These differences were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater{} evergreen angiosperm \textgreater{} deciduous angiosperm.
-Next we calculated $\mfms$ at a common plant height for each of the PFTs (the mean height across all data, \Sexpr{format(10^mean(dataset$lh.t, na.rm=TRUE), nsmall=1, digits=2)}m).
-These three size-corrected $\mfms$ values were highly significantly different between PFTs (p \textless{} 0.001), and were approximately proportional to the average LMA ($\mfaf$) across the PFTs (Fig.
+The composition of $\mfms$ and $\afms$ into component variables confirmed our hypotheses, that size-related effects in biomass distribution are heavily concentrated in the $\asms$ term within eq.
+\ref{eq:mfmspart}-\ref{eq:afmspart}.
+For $\mfms$, $\afms$ and each of the component variables, we estimated the variance explained by individual height, PFT, and climate variables (MAT or aridity) (Table \ref{table:varparttable}).
+We did this by sequentially adding terms to a GAM, and calculating the $R^2$ for each fitted model.
+The first model included only plant height, and this variable alone explained >65\% of the variation in both $\mfms$, $\afms$, and $\asms$, but little variation (4 - 11 \%) in LMA and $\afas$.
+
+We found large differences between the three major PFTs in terms of above-ground biomass distribution, as quantified by the ratio $\mfms$ (Fig.
+3a), but not in the ratio $\afms$ (Fig.
+S2).
+Differences in $\mfms$ were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater{} evergreen angiosperm \textgreater{} deciduous angiosperm.
+Next we calculated $\mfms$ at a common plant height of 3m for each of the PFTs (the mean height across all data, \Sexpr{format(10^mean(dataset$lh.t, na.rm=TRUE), nsmall=1, digits=2)}m).
+These three size-standardised values of $\mfms$ were highly significantly different between PFTs (p \textless{} 0.001), and were approximately proportional to the average LMA ($\mfaf$) across the PFTs (Fig.
 3b).
 As a result, the amount of leaf area per unit above-ground woody biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater{} 0.05, see Fig.
-3c), because the differences in $\mfms$ were exactly compensated by differences in LMA (see also Fig.
-S2 for $\afms$ as a function of plant height).
+3c), because the differences in $\mfms$ were exactly compensated by differences in LMA.
 
-Next, we decomposed $\mfms$ into three component variables with distinct biological interpretation (Eq.
-\ref{eq:mfmspart}): LMA ($\mfaf$) is an important leaf trait, $\afas$ is a measure of hydraulic supply and mechanical support per unit leaf area, and $\asms$ depends on stem form and wood density.
-For each of the component variables and $\mfms$, we estimated the variance explained by plant height, PFT, and climate variables (MAT or aridity) (Table \ref{table:varparttable}).
-We did this by sequentially adding terms to a GAM, and calculating the $R^2$ for each fitted model.
-The first model included only plant height, and already explained much variation (65 - 79\%) in $\mfms$ and $\asms$ but not LMA and $\afas$ (4 - 11 \%).
-When PFT was added to the model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $\mfms$ and $\mfaf$ but not $\afas$ or $\asms$, demonstrating that PFT differences entered largely in leaf mass-based variables.
-Additional variance explained by adding PFT was especially large for LMA, which is consistent with the large differences between PFTs in LMA shown in Fig.
+These patterns in biomass distribution reflected differences in the distribution of components of eq.
+\ref{eq:mfmspart}-\ref{eq:afmspart}.
+In particular, LMA was found to differ substantially among PFTs, whereas $\afas$ did not (Fig.
+6), which is consistent with the large differences between PFTs in LMA shown in Fig.
 3b,c.
+When PFT was added to the statistical model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $\mfaf$ but not $\afas$  (Table \ref{table:varparttable}), demonstrating that PFT differences entered largely in leaf mass-based variables.
 
-From the variance partitioning (Table \ref{table:varparttable}), it is evident that $\afas$ did not vary appreciably between PFTs.
-Indeed, when all PFTs were plotted together, similar $\af$ was attained at a given $\as$ across the entire plant size range (with the exception of extremely large trees with \textgreater{} 1m$^2$ stem basal area, for which comparatively little data was available) (Fig.
+To confirm the finding that $\afas$ did not vary appreciably between PFTs, we further examined patterns between $\af$ and $\$as$.
+When these variables are plotted against one another, we see that similar $\af$ is attained at a given $\as$, across the entire plant size range (with the exception of extremely large trees with \textgreater{} 1m$^2$ stem basal area, for which comparatively little data was available) (Fig.
 4).
@@
 Likewise, the ratio of stem basal area to stem mass ($\asms$) apparently varied little between PFTs (Table \ref{table:varparttable}).
-To illustrate this further, a plot of $\ms$ as a function of $\as$ revealed a tight relationship, with little apparent difference between PFTs (Fig 5a).
+To illustrate this further, a plot of $\ms$ as a function of $\as$ revealed a tight relationship, with little apparent difference between PFTs (Fig.
+5a).
 The only exception was for large gymnosperm trees, which consistently showed a lower $\ms$ at a given $\as$ (Fig.
@@
 Taken together, these results demonstrate that 1) the majority of PFT-related variation in $\mfms$ results from variation in LMA, not $\afas$ or $\asms$; and 2) the majority of height-related variation in $\mfms$ results from variation in $\asms$, not LMA or $\afas$.
-Nevertheless, it is important to note that there is very large variation between individuals within PFTs (Fig.
-6).
-To correct for plant height effect on $\msas$, we expressed the stem mass ratio as $\mshas$ because we expect from basic geometry that $\ms$ is proportional to H and $\as$, via the relationship: $\ms = \phi \rho \as H$, where $\phi$ a stem shape parameter, and $\rho$ wood density.
-The histograms summarize that PFT effects are very evident in LMA (Fig.
-6a), not in $\afas$, and only to some extent for $\mshas$, which is entirely driven by large gymnosperm trees.
-However for all variables there was much unexplained variation between species (and individuals within species) within PFTs.
+Nevertheless, it is important to note that there is also very large variation between individuals within PFTs (Fig.
+6), and within all species as plant species grow form seedlings to adults.
 
@@
 The only exception was LMA, for which a model that included MAT explained 53\% of the variation, compared to a model with height and PFT (45\%).
-Since the lack of climate effects on biomass distribution is in contrast with recent finding of \citet{reich2014}, we also analyzed climate effects with methods similar to that study.
-When we tested for MAT effects on $\mfms$ when only considering gymnosperms and angiosperms (thus grouping decidous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig.
+Since the lack of climate effects on biomass distribution is in contrast with recent finding of \citet{reich2014}, we also analysed climate effects with methods similar to that study.
+When we tested for MAT effects on $\mfms$ when only considering gymnosperms and angiosperms (thus grouping deciduous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig.
 S5) for angiosperms for most size classes (but never for gymnosperms).
@@
 1), which have higher $\mfms$.
-When analyzed separately for deciduous and evergreen angiosperms, few and weaker significant relationships were found (Fig.
+When analysed separately for deciduous and evergreen angiosperms, few and weaker significant relationships were found (Fig.
 S6).
@@
 Thanks also to Rich FitzJohn for advice about using the `remake` package.
-Figure 1a uses an image by Lana Heydon, IAN Image Library (http://ian.umces.edu/imagelibrary).
+Fig.
+1a uses an image by Lana Heydon, IAN Image Library (http://ian.umces.edu/imagelibrary).
 Thanks to anonymous reviewers for suggestions that greatly improved the paper.
@@
     \fi
-    \caption{Histograms (as probability density functions) of leaf mass per area ($\mfaf$), leaf area per unit basal stem area ($\afas$) and the stem mass index ($\mshas$) grouped by the three PFTs.
+    \caption{Histograms (as probability density functions) of leaf mass per area ($\mfaf$) and leaf area per unit basal stem area ($\afas$) grouped by the three PFTs.
 Arrows indicate means by PFT.
@@
 \textbf{Fig. S5} Relationships between leaf - stem biomass distribution ($\mfms$) and mean annual temperature (MAT) (\degree C). \\
-\textbf{Fig. S6} As Fig S5 but with angiosperms split up into evergreen and deciduous. \\
+\textbf{Fig. S6} As Fig. S5 but with angiosperms split up into evergreen and deciduous. \\
 

### 2016-04-22 R/data_processing.R

@@ -229,7 +229,6 @@ prepare_dataset_1 <- function(baad, plantations=TRUE){
     lmso_astba2 <- log10(m.so / a.stba2)
     lastba2_mst <- log10(a.stba2 / m.st)
     lastba2 <- log10(a.stba2)
-    mstastbht <- log10(m.st / (a.stba2 * h.t)) # stem index
     aridity <- PET/MAP
     
     })

### 2016-04-22 R/figures.R

@@ -302,7 +302,7 @@ figure5 <- function(dataset){
 # Histograms of MF/AS, AF/AS, and MS/(AS*H)
 figure6 <- function(dataset, nbin=100){
   
-  par(mar=c(0,0,0,2), oma=c(5,5,1,1), las=1, cex.axis=0.85, mfrow=c(1,3), mgp=c(3,1.5,0))
+  par(mar=c(0,2,0,2), oma=c(5,5,1,1), las=1, cex.axis=0.85, mfrow=c(1,2), mgp=c(3,1.5,0))
   
   histbypft(llma, pft, dataset, xaxis=3,legend.cex=1,col=my_cols_transparent(),
             xlab="", overlay=TRUE,plotwhat="density",ylab="Density",
@@ -318,21 +318,10 @@ figure6 <- function(dataset, nbin=100){
             legend.text=rep("",3), meanlinecol=my_linecols())
   plotlabel("(b)","topleft")
   
-  histbypft(mstastbht, pft, dataset, xaxis=3,legend.cex=1,col=my_cols_transparent(),
-            xlab="",overlay=TRUE,plotwhat="density",ylab="Density",
-            nbin=nbin,
-            xlim=c(1,4),
-            Means=mixmean("mstastbht","pft",dataset),cicol=alpha("grey",0.65),
-            legend.text=rep("",3), meanlinecol=my_linecols())
-  plotlabel("(c)","topleft")
-  
-  
   mtext(side=1, line=3, text=expression(M[F]/A[F]~~(kg~m^-2)),
-        outer=TRUE, at=1/6, cex=0.9)
+        outer=TRUE, at=1/4, cex=0.9)
   mtext(side=1, line=3, text=expression(A[F]/A[S]~~(m^2~m^-2)),
-        outer=TRUE, at=3/6, cex=0.9)
-  mtext(side=1, line=3, text=expression(M[S]/(A[S]*H)~~(kg~m^-3)),
-        outer=TRUE, at=5/6, cex=0.9)
+        outer=TRUE, at=3/4, cex=0.9)
   
 }
 

### 2016-04-22 manuscript.Rnw
Move Fig 6 to Fig 4, to fit with new results section

@@
 In particular, LMA was found to differ substantially among PFTs, whereas $\afas$ did not (Fig.
-6), which is consistent with the large differences between PFTs in LMA shown in Fig.
+4), which is consistent with the large differences between PFTs in LMA shown in Fig.
 3b,c.
@@
 When these variables are plotted against one another, we see that similar $\af$ is attained at a given $\as$, across the entire plant size range (with the exception of extremely large trees with \textgreater{} 1m$^2$ stem basal area, for which comparatively little data was available) (Fig.
-4).
+6).
 We fit a GAM to $\af$ as a function of $\as$ with and without PFT as a covariate, which showed PFT significantly improved model fit (likelihood ratio test, p \textless{} 0.001).
@@
 To illustrate this further, a plot of $\ms$ as a function of $\as$ revealed a tight relationship, with little apparent difference between PFTs (Fig.
-5a).
+6a).
 The only exception was for large gymnosperm trees, which consistently showed a lower $\ms$ at a given $\as$ (Fig.
-5b).
+6b).
 We fitted a GAM of $\ms$ as a function of $\as$ (added as a smooth term) with or without PFT as a covariate.
@@
 Nevertheless, it is important to note that there is also very large variation between individuals within PFTs (Fig.
-6), and within all species as plant species grow form seedlings to adults.
+4), and within all species as plant species grow form seedlings to adults.
 
@@
 We stress that although our results show some convergence between PFTs in leaf area relationships, we do not dismiss the very large residual variation that is yet to be explained (Fig.
-6).
+4).
 Our purpose was to test hypotheses related to first-order differences between PFTs, plant height, and climate, yet clearly more work can be done to understand the variation between species and sites within groups.
@@
     \fi
+    \caption{Histograms (as probability density functions) of leaf mass per area ($\mfaf$) and leaf area per unit basal stem area ($\afas$) grouped by the three PFTs.
+Arrows indicate means by PFT.
+Colours as in Fig.
+\ref{fig:figlmf}.}
+    \label{fig:fighist}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \ifincludefigs
+    \includegraphics[width=0.99\textwidth]{Figure5.pdf}
+    \fi
     \caption{Total plant leaf area ($\af$) is approximately proportional to basal stem area ($\as$), and this relationship is similar between PFTs.
@@
     \ifincludefigs
-    \includegraphics[width=0.99\textwidth]{Figure5.pdf}
+    \includegraphics[width=0.99\textwidth]{Figure6.pdf}
     \fi
@@
     \label{fig:stemmassindex}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \ifincludefigs
-    \includegraphics[width=0.99\textwidth]{Figure6.pdf}
-    \fi
-    \caption{Histograms (as probability density functions) of leaf mass per area ($\mfaf$) and leaf area per unit basal stem area ($\afas$) grouped by the three PFTs.
-Arrows indicate means by PFT.
-Colours as in Fig.
-\ref{fig:figlmf}.}
-    \label{fig:fighist}
 \end{figure}

### 2016-04-22 manuscript.Rnw
Move Fig S2 into panel of Fig 3

@@
 The three methods generally agreed on the ranking of variable importance (see Fig.
-S5 and S6).
+S4 and S5).
 
@@
 Similarly, $\afms$ decreased with plant height (Fig.
-S2), from an average of \Sexpr{afmssmall} for very small saplings to an average of \Sexpr{afmslarge} for large trees.
+3c), from an average of \Sexpr{afmssmall} for very small saplings to an average of \Sexpr{afmslarge} for large trees.
 
@@
 3a), but not in the ratio $\afms$ (Fig.
-S2).
+3c).
 Differences in $\mfms$ were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater{} evergreen angiosperm \textgreater{} deciduous angiosperm.
@@
 As a result, the amount of leaf area per unit above-ground woody biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater{} 0.05, see Fig.
-3c), because the differences in $\mfms$ were exactly compensated by differences in LMA.
+3d), because the differences in $\mfms$ were exactly compensated by differences in LMA.
 
@@
 4), which is consistent with the large differences between PFTs in LMA shown in Fig.
-3b,c.
+3b.
 When PFT was added to the statistical model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $\mfaf$ but not $\afas$  (Table \ref{table:varparttable}), demonstrating that PFT differences entered largely in leaf mass-based variables.
@@
 When we tested for MAT effects on $\mfms$ when only considering gymnosperms and angiosperms (thus grouping deciduous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig.
-S5) for angiosperms for most size classes (but never for gymnosperms).
+S4) for angiosperms for most size classes (but never for gymnosperms).
 We found, however, that this was largely due to the fact that at higher MAT, an increasing proportion of angiosperms are evergreen (Fig.
@@
 When analysed separately for deciduous and evergreen angiosperms, few and weaker significant relationships were found (Fig.
-S6).
+S5).
 
@@
 2), or among biomass pools (Fig.
-S4), even on a logarithmic scale.
+S3), even on a logarithmic scale.
 This is consistent with recent results on an independently compiled large database of plant allometry \citep{poorter2015}, and further calls into question the generality of the often-assumed log-linear relationships between plant biomass pools and size metrics.
@@
 Although we found remarkably similar leaf area per unit above-ground woody biomass ($\mfms$) between the three PFTs studied (Fig.
-S2), this does not in itself imply that LAI is equally similar.
+3c), this does not in itself imply that LAI is equally similar.
 Indeed, in a global compilation of LAI for undisturbed forest stands, \citet{iio2014} reported a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI.
@@
 We were able to partly reconcile our findings with those of \citet{reich2014}: when we categorised our species as only angiosperms vs gymnosperms (thus grouping deciduous and evergreens, as did Reich \emph{et al.}), $\mfms$ increased consistently with MAT for most size classes within angiosperms (Fig.
-S5), in line with the findings of \citet{reich2014}; on the other hand we found no comparable relationship for gymnosperms.
+S4), in line with the findings of \citet{reich2014}; on the other hand we found no comparable relationship for gymnosperms.
 Moreover, we found that in our dataset, the significant relationships observed for angiosperms as a whole, could be explained via differences between the LMA of deciduous and evergreen species: evergreen angiosperms occurred both at higher MAT (Fig.
@@
 \textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\
-\textbf{Fig. S2} Leaf area ratio ($\afms$) by plant height and PFT.\\
-\textbf{Fig. S3} Comparison of three methods for variance partitioning of the four studied variables. \\
-\textbf{Fig. S4} Whole-plant leaf mass ($\mf$) and leaf area ($\af$) as a function of total aboveground biomass. \\
-\textbf{Fig. S5} Relationships between leaf - stem biomass distribution ($\mfms$) and mean annual temperature (MAT) (\degree C). \\
-\textbf{Fig. S6} As Fig. S5 but with angiosperms split up into evergreen and deciduous. \\
+\textbf{Fig. S2} Comparison of three methods for variance partitioning of the four studied variables. \\
+\textbf{Fig. S3} Whole-plant leaf mass ($\mf$) and leaf area ($\af$) as a function of total aboveground biomass. \\
+\textbf{Fig. S4} Relationships between leaf - stem biomass distribution ($\mfms$) and mean annual temperature (MAT) (\degree C). \\
+\textbf{Fig. S5} As Fig. S4 but with angiosperms split up into evergreen and deciduous. \\
 

### 2016-04-22 manuscript_suppinfo.Rnw
Restructure introduction & results to emphasise hypotheses
- as requested by referees, move decomposition equation into introduction
- use this as basis for outlining hypotheses
- add tests in relation to af/ms as major result
- remove stem mass index from results and figures

Also, suggest

- moving Fig 6 - > Fig 4
- embedding Fig S2 in Fig 3
Move Fig S2 into panel of Fig 3

@@
     \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
-    \caption{Leaf area ratio ($\afms$) by plant height (H) and PFT.
-Each symbol is an individual plant.
-Lines are generalized additive model fits using species/dataset as a random effect.
-Shaded areas are approximate 95\% confidence intervals.}
-    \label{fig:figureS2}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.99\textwidth]{FigureS3.pdf}
     \caption{Comparison of three methods for variance partitioning of the four studied variables (see legend) by total plant height, plant functional type and climate (MAP, MAT and interaction).
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
+    \includegraphics[width=0.99\textwidth]{FigureS3.pdf}
     \caption{Whole-plant leaf mass ($\mf$) and leaf area ($\af$) as a function of total aboveground biomass.
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
+    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
     \caption{Relationships between leaf - stem biomass distribution ($\mfms$) and mean annual temperature (MAT) (\degree C).
@@
     \centering
-    \includegraphics[width=0.99\textwidth]{FigureS6.pdf}
-    \caption{As Fig S5 but with angiosperms split up into evergreen and deciduous. Note how climate effects have all but disappeared, except for evergreen angiosperms in two size class bins.}
+    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
+    \caption{As Fig. S4 but with angiosperms split up into evergreen and deciduous. Note how climate effects have all but disappeared, except for evergreen angiosperms in two size class bins.}
     \label{fig:figreichclimate2}

### 2016-04-22 R/figures.R

@@ -202,39 +202,51 @@ figure3 <- function(dataset, KGAM=4){
   # Evaluate biomass distribution at mean height
   xpred <- mean(dataset$lh.t, na.rm=TRUE)
   
-  l <- layout(matrix(c(1,2,1,3), byrow=T, ncol=2),
+  l <- layout(matrix(c(1,2,3,4), byrow=F, ncol=2),
               widths=c(1,0.67), heights=c(1,1))
   
-  par(mar=c(4,4,4,1), cex.axis=0.9, cex.lab=1.2, mgp=c(2.3,0.5,0), tcl=-0.35, las=1)
+  par(mar=c(0.35,4,4,1), cex.axis=0.9, cex.lab=1.2, tcl=-0.35, las=1, mgp=c(2.3,0.5,0))
   obj1 <- smoothplot(lh.t, lmlf_mst, pft, dataset, R="Group",linecols=my_linecols(),
                      pointcols=my_cols_transparent(),axes=FALSE, ylim=c(-3,1),
-                     xlab="H (m)",kgam=KGAM,
+                     xlab="",kgam=KGAM,
                      ylab=expression(M[F]/M[S]~(kg~kg^-1)),
                      cex=0.6)
-  log10axes()
-  my_legend("bottomleft", "long")
+  log10axes(1, labels=FALSE)
+  log10axes(2)
+
   box()
   plotlabel("(a)","topright")
   
+  par(mar=c(4,4,0.35,1), cex.axis=0.9, cex.lab=1.2, tcl=-0.35, las=1, mgp=c(2.3,0.5,0))
+
+  x <- smoothplot(lh.t, lalf_mst, pft, dataset,  R="Group",
+                  linecols=my_linecols(), pointcols=my_cols_transparent(),
+                  xlab="H (m)",kgam=KGAM,axes=FALSE, cex=0.6,
+                  ylab=expression(A[F]/M[S]~~(m^2~kg^-1)))
+  plotlabel("(c)","topright")
+  log10axes()
+  box()
+  my_legend("bottomleft", labels="long")
+
   # Arrow indicating where mean biomass distribution was estimated
   par(xpd=NA)
-  arrows(x0=xpred, x1=xpred, y0=-3.5, y1=-3.2, length=0.1, lwd=2)
+  arrows(x0=xpred, x1=xpred, y0=-2.75, y1=-2.45, length=0.1, lwd=2)
 
[…]

### 2016-04-22 manuscript.Rnw
Minor editting in discussion

@@
 Using the largest database of individual woody plant allometry to date, we studied patterns in above-ground biomass distribution among major woody plant functional types (PFTs), plant size, and climate.
-Our compilation is the first of its kind that also includes leaf area estimates on the same plants, allowing a direct test of the hypothesis that leaf and stem biomass distribution is related to the LMA.
-We found that at a common plant height, the leaf / above-ground woody biomass ratio ($\mfms$) was proportional to the LMA across three PFTs, so that leaf area per unit above-ground woody biomass did not vary between PFTs.
+Our compilation is the first of its kind that also includes leaf area estimates on the same plants, allowing a direct test  hypotheses that leaf and stem biomass distribution is driven by component traits, LMA and $\afas$.
+We found that at a common plant height, the ratio of leaf to above-ground woody biomass ($\mfms$) varied due to underling differences in LMA but across three PFTs, but that there were no differences among PFTs in $\afas$.
+As a result, leaf area per unit above-ground woody biomass did not vary between PFTs.
 We found very weak effects of climate (mean annual temperature and aridity) on biomass distribution, so that the vast majority of the variation in biomass distribution was explained by plant size and PFT.
@@
 We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship.
-This was useful because both $\mf$ and $\ms$ showed very non-linear patterns with plant height (Fig.
-2), or among biomass pools (Fig.
-S3), even on a logarithmic scale.
+This was useful because both $\mf$ and $\ms$ showed non-linear patterns with plant height (Fig.
+2), and among biomass pools, even on a logarithmic scale.
 This is consistent with recent results on an independently compiled large database of plant allometry \citep{poorter2015}, and further calls into question the generality of the often-assumed log-linear relationships between plant biomass pools and size metrics.
 
-We have confirmed the hypothesis that the amount of leaf area supported per plant is less variable between PFTs than the amount of leaf mass.
-\citet{poorter2015} hypothesised such a result might arise, because forests would converge to a similar leaf area index (LAI).
-Although we found remarkably similar leaf area per unit above-ground woody biomass ($\mfms$) between the three PFTs studied (Fig.
+A significant implication of our results is that the amount of leaf area supported per plant is less variable between PFTs than the amount of leaf mass, at least at the level of individual plants.
+\citet{poorter2015} hypothesised such a result might arise at the scale of entire stands, because forests would converge to a similar leaf area index (LAI).
+Yet this idea is inconsistent with results reported by \citet{iio2014}, who found a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI.
+Although we found remarkably similar leaf area per unit above-ground woody biomass ($\afms$) (Fig.
 3c), this does not in itself imply that LAI is equally similar.
-Indeed, in a global compilation of LAI for undisturbed forest stands, \citet{iio2014} reported a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI.
-Moreover, we would in fact expect LAI to change with LMA, not because of changes in the amount of leaf area maintained by individual plants, but because LMA correlates with the shade-tolerance of seedlings \citep{poorter2006, falster2011}, and thereby the LAI of the community \citep{falster2011}.
-The ultimate reason LMA might affect shade tolerance is because is via its correlation with leaf lifespan \citep{wright2004}.
-For whole stands, we can therefore expect total stand leaf mass to be correlated with LMA, both because of greater leaf mass per plant \citep[e.g.]{reich1992}, and the greater density of plants arising from increased leaf lifespan.
+In fact, we would  expect LAI to change with LMA, not because of changes in the amount of leaf area maintained by individual plants, but because LMA correlates with the shade-tolerance of seedlings \citep{poorter2006, falster2011}, and thereby the LAI of the community \citep{falster2011}.
+The ultimate reason LMA might affect shade tolerance is most likely because is via its correlation with leaf lifespan \citep{wright2004}.
+For whole stands, we can therefore expect total stand leaf mass to be correlated with LMA \citep[e.g.]{reich1992}, both because of greater leaf mass per plant, and the greater density of plants arising from increased leaf lifespan.
 Previous interpretations of differences in $\mfms$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002, poorter2012}.
@@
 
-Although we found no consistent effects of climate on biomass distribution after PFT and plant height were accounted for, effects of climate are likely mediated via plant height.
+Although we found no consistent effects of climate on biomass distribution after PFT and plant height were accounted for, effects of climate may also arise by altering the average height of a stand.
 At a global scale, average plant height is clearly related to climate via water availability \citep{klein2015, simard2011}.
@@
 
-We stress that although our results show some convergence between PFTs in leaf area relationships, we do not dismiss the very large residual variation that is yet to be explained (Fig.
+We stress that although our results show some convergence between PFTs in leaf area relationships, we do not dismiss the very variation within PFTs (Fig.
 4).
@@
 It is also likely that the relationship between LMA and biomass distribution reported here among PFTs, also operates among species within PFTs.
-As LMA is known to vary widely within PFTs (Fig.~\ref{fig:fighist}), such variation could underpin the equivalent variation in biomass distribution.
+As LMA is known to vary widely within PFTs (Fig.
+\ref{fig:fighist}), such variation could underpin the equivalent variation in biomass distribution.
 

### 2016-04-26 manuscript.Rnw
Abstract updated; reference added; minor edits

@@
 \begin{itemize}
-\item We tested the hypothesis that biomass distribution in leaves and stems is related to the leaf mass per area, an important leaf trait, across major woody plant functional types (PFTs) (as defined by deciduous vs.
+\item Here we aim to understand differences in biomass distribution between major woody plant functional types (PFTs) (as defined by deciduous vs.
 evergreen and gymnosperm vs.
-angiosperm).
+angiosperm) in terms of underlying traits, in particular the leaf mass per area, a key leaf trait.
 \item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species.
-We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, and climate on biomass distribution.
+We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, leaf mass per area, total leaf area, stem basal area and climate on above-ground biomass distribution.
 \item The ratio of leaf mass to above-ground woody mass ($\mfms$) varied strongly between PFTs.
-We found that $\mfms$, once corrected for plant height, was proportional to leaf mass per area across PFTs.
+We found that $\mfms$, at a given plant height, was proportional to leaf mass per area across PFTs.
 As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass.
@@
 
-The goal of the current work is to better understand differences among PFTS in their biomass distribution.
+The goal of the current work is to better understand differences among PFTs in their biomass distribution.
 Past work has shown that gymnosperms tend to hold a much larger leaf mass compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012,poorter2015}.
-The cause of these differences are not yet fully understood, but could arise from either differences leaf mass per area (LMA) \citep{poorter2009}, or the amount of leaf area maintained \citep{chabothicks1982,bond1989}, or both.
-It is also unknown whether there exist differences among finer-scale groupings, for example between evergreen and deciduous angiosperms.
+The cause of this difference is not yet fully understood, but could arise from either differences leaf mass per area (LMA) \citep{poorter2009}, or the amount of leaf area maintained \citep{chabothicks1982,bond1989}, or both.
+It is also unknown whether there are differences among finer-scale groupings, for example between evergreen and deciduous angiosperms.
 While past analyses have often spanned many sites and / or species, researchers have been unable to further investigate the factors underlying observed differences in biomass distribution among PFTs, because available data only included a limited number of variables.
 
-Here, we use the recently-compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which includes observations of biomass and other size metric for \Sexpr{nrow(baad_all$data)} individual plants (Fig.
+Here, we use the recently-compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which includes observations of biomass and other size metrics for \Sexpr{nrow(baad_all$data)} individual plants (Fig.
 1), to establish the underlying cause of biomass distribution among PFTs, and also to test for systematic shifts in biomass distribution with climate.
@@
 To quantify this distribution we analyse two related metrics: 1) the ratio of leaf mass to stem mass, $\mfms$; and 2) the ratio of leaf area to stem mass, $\afms$.
-To understand the factors driving variation in these quantities, we decompose these ratios into component variables, each of which can be quantified in our dataset.
-The ratio of leaf mass to stem mass, can decomposed as
+To understand the factors driving variation in these quantities, we decompose these ratios into component variables, each of which can be quantified with our dataset.
+The ratio of leaf mass to stem mass, can be decomposed as
 \begin{equation}
@@
 \begin{enumerate}
-  \item Individual height captures the most majority of variation in $\mfms$ and $\afms$, reflecting the fact that height can vary several orders of magnitude within the lifespan of a single individual and thus represents the dominant form of variation in eq .
-\ref{eq:mfmspart}-\ref{eq:afmspart}.
-  \item That these height-related effects on biomass distribution are concentrated in the term $\asms$, and that while $\asms$ may vary considerably among individuals, there are no systematic differences in $\asms$ among PFTs.
-  \item That after controlling for the effects of individual height, differences in $\mfms$ among PFTs can arise due to differences in either, or both, of a) LMA, because PFTs tend to differ LMA \citep{poorter2009}, or b)$\afas$, because gymnosperms are hypothesised to maintain a greater leaf area than angiosperms \citep{chabothicks1982, bond1989}.
-  \item That any differences in $\afas$ among PFTs also leads to differences in $\afms$.
+  \item Individual height captures the most majority of variation in $\mfms$ and $\afms$, reflecting the fact that height can vary several orders of magnitude within the lifespan of a single individual and thus represents the dominant form of variation in Eqns \ref{eq:mfmspart}-\ref{eq:afmspart}.
+  \item Height-related effects on biomass distribution are concentrated in the term $\asms$, and that while $\asms$ may vary considerably among individuals, there are no systematic differences in $\asms$ among PFTs.
+  \item After controlling for the effects of individual height, differences in $\mfms$ among PFTs can arise due to differences in either, or both, of a) LMA, because PFTs tend to differ LMA \citep{poorter2009}, or b)$\afas$, because gymnosperms are hypothesised to maintain a greater leaf area than angiosperms \citep{chabothicks1982, bond1989}.
+  \item That any differences in $\afas$ among PFTs also lead to differences in $\afms$.
 \end{enumerate}
 
-A final aim was test for significant shifts in biomass distribution with climate.
+A final aim was to test for shifts in biomass distribution with climate.
 Little is known about global-scale patterns in biomass distribution in relation to climate, but one might expect biomass distribution to shift with precipitation or mean annual temperature, as small-scale comparisons have shown such responses \citep{berninger1994,callaway1994,delucia2000,poyatos2007}.
 \citet{reich2014} has also shown that after controlling  for the large size-related variation in biomass distribution, the ratio of leaf to total plant mass (averaged across whole stands, either including or excluding below-ground components) increases with mean annual temperature in both gymnosperms and angiosperms.
-Thus, we aimed to test for climate-related variation in $\mfms$ and $\afms$, and their component terms with PFTS.
+Thus, we aimed to test for climate-related variation in $\mfms$ and $\afms$, and their component terms with PFTs.
 
@@
 MAT and MAP were extracted from Worldclim \citep{worldclim}, PET from the Global-PET database (http://www.cgiar-csi.org, \citet{zomer2008}), and an aridity index was calculated as PET/MAP.
-To assess the coverage of the global climate space by the dataset, we also extracted MAT and MAP across the globe from the global land cover database GLC-SHARE \citep{lathamglobal},but excluding areas without woody vegetation.
+To assess the coverage of the global climate space by the dataset, we also extracted MAT and MAP from Worldclim for each 0.5\degree cell across the globe but excluding areas without woody vegetation, as assessed from the global land cover database GLC-SHARE \citep{lathamglobal}.
 
@@
 We also compared this approach to another method using linear mixed-effects models instead of GAMs \citep[calculating the $R^2$ for linear mixed-effects models for the fixed effects only following][]{nakagawa2013}, as well as third method called ``independent effects analysis''.
-The three methods generally agreed on the ranking of variable importance (see Fig.
+The three methods generally agreed on the ranking of variable importance (see Figs.
 S4 and S5).

### 2016-04-26 manuscript.Rnw
more edits

@@
 evergreen and gymnosperm vs.
-angiosperm) in terms of underlying traits, in particular the leaf mass per area, a key leaf trait.
+angiosperm) in terms of underlying traits, in particular the leaf mass per area and leaf area per unit stem basal area.
 \item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species.
@@
 We found that $\mfms$, at a given plant height, was proportional to leaf mass per area across PFTs.
-As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass.
+As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass, or per unit stem basal area.
 Climate consistently explained very little additional variation in biomass distribution at a given plant size.
@@
 
-The composition of $\mfms$ and $\afms$ into component variables confirmed our hypotheses, that size-related effects in biomass distribution are heavily concentrated in the $\asms$ term within eq.
-\ref{eq:mfmspart}-\ref{eq:afmspart}.
+The composition of $\mfms$ and $\afms$ into component variables confirmed our second hypothesis, that size-related effects in biomass distribution are heavily concentrated in the $\asms$ term within Eqn \ref{eq:mfmspart}-\ref{eq:afmspart}.
 For $\mfms$, $\afms$ and each of the component variables, we estimated the variance explained by individual height, PFT, and climate variables (MAT or aridity) (Table \ref{table:varparttable}).
@@
 Differences in $\mfms$ were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater{} evergreen angiosperm \textgreater{} deciduous angiosperm.
-Next we calculated $\mfms$ at a common plant height of 3m for each of the PFTs (the mean height across all data, \Sexpr{format(10^mean(dataset$lh.t, na.rm=TRUE), nsmall=1, digits=2)}m).
+Next we calculated $\mfms$ at a common plant height of 3m for each of the PFTs (the mean height across all data).
 These three size-standardised values of $\mfms$ were highly significantly different between PFTs (p \textless{} 0.001), and were approximately proportional to the average LMA ($\mfaf$) across the PFTs (Fig.
@@
 As a result, the amount of leaf area per unit above-ground woody biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater{} 0.05, see Fig.
-3d), because the differences in $\mfms$ were exactly compensated by differences in LMA.
+3d), because the differences in $\mfms$ were compensated by differences in LMA.
 
-These patterns in biomass distribution reflected differences in the distribution of components of eq.
-\ref{eq:mfmspart}-\ref{eq:afmspart}.
-In particular, LMA was found to differ substantially among PFTs, whereas $\afas$ did not (Fig.
+These patterns in biomass distribution reflected differences in the distribution of components of Eqn \ref{eq:mfmspart}-\ref{eq:afmspart}.
+In particular, LMA differed substantially among PFTs, whereas $\afas$ did not (Fig.
 4), which is consistent with the large differences between PFTs in LMA shown in Fig.
@@
 
-To confirm the finding that $\afas$ did not vary appreciably between PFTs, we further examined patterns between $\af$ and $\$as$.
+To confirm the finding that $\afas$ did not vary appreciably between PFTs, we further examined patterns between $\af$ and $\as$.
 When these variables are plotted against one another, we see that similar $\af$ is attained at a given $\as$, across the entire plant size range (with the exception of extremely large trees with \textgreater{} 1m$^2$ stem basal area, for which comparatively little data was available) (Fig.
@@
 Nevertheless, it is important to note that there is also very large variation between individuals within PFTs (Fig.
-4), and within all species as plant species grow form seedlings to adults.
+4), and within all species as plant species grow from seedlings to adults.
 
@@
 Using the largest database of individual woody plant allometry to date, we studied patterns in above-ground biomass distribution among major woody plant functional types (PFTs), plant size, and climate.
-Our compilation is the first of its kind that also includes leaf area estimates on the same plants, allowing a direct test  hypotheses that leaf and stem biomass distribution is driven by component traits, LMA and $\afas$.
+Our compilation is the first of its kind that also includes leaf area estimates on the same plants, allowing a direct test of the hypothesis that leaf and stem biomass distribution is driven by component traits, LMA and $\afas$.
 We found that at a common plant height, the ratio of leaf to above-ground woody biomass ($\mfms$) varied due to underling differences in LMA but across three PFTs, but that there were no differences among PFTs in $\afas$.

### 2016-04-26 manuscript.Rnw
Missing reference and minor tidying

@@
 % Notes:
-% for submission, convert to Word docx using Adobe Acrobat Professional.
-To not include line numbers here, add them in Word.
-Also do not include figures.
-Word count can appear on first page; follow instructions further below.
+% - For submission, convert to Word docx using Latex2rtf. 
+% - Do not include line numbers here, add them in Word. Also do not embed figures. 
+% - Word count can be automatically added on first page; follow instructions further below.
 
@@
 \newif\ifincludefigs
-\includefigstrue  % or \includefigsfalse
+\includefigsfalse % or \includefigstrue
 
@@
 
-% Word count
-%% Uncomment this after compiling the ms, then compile again.
-% Requires Perl to be installed.
-%' <<echo=FALSE>>=
-%' #shell("texcount manuscript.tex > wordcount.txt")
+% -- Word count
+% After compiling the ms (and no cleanup, so .tex is present), run:
+% shell("texcount manuscript.tex > wordcount.txt")
+% Then uncomment the following:
+%' <<echo=FALSE, eval=FALSE>>=
 %' nrw <- get_wordcount("wordcount.txt")
@@
 %' Supporting information : 6 Figures (all in colour).
-
-% Alternatively, in a shell,
-%
-%texcount manuscript.tex > wordcount.txt
-%shell("texcount manuscript.tex > wordcount.txt")
 
@@
 We chose to estimate missing stem diameter rather than stem area because a much better fit was obtained, with a more constant error variance.
-The parameter $c$ in eq.
-\ref{eq:dia} was further expressed as a function of plant height:
+The parameter $c$ in Eqn \ref{eq:dia} was further expressed as a function of plant height:
 \begin{equation}
@@
 All analyses were conducted in R 3.2.4 \citep{R-2015}, using the packages `mgcv` \citep{wood2006}, `hier.part` \citep{walsh2013}, `MuMIn` \citep{barton2015}, and `remake` \citep{fitzjohn2015}.
-In addition, the code replicating this analysis (and all figures) is available at \url{http://github.com/RemkoDuursma/baadanalysis} (doi: \emph{will be added at proof stage}).
+In addition, the code replicating this analysis (and all figures) is available at http://github.com/RemkoDuursma/baadanalysis (doi: \emph{will be added at proof stage}).
 
@@
 4), which is consistent with the large differences between PFTs in LMA shown in Fig.
-3b.
+3b, and the substantial variance in LMA explained by just PFT (Table \ref{table:varparttable}).
 When PFT was added to the statistical model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $\mfaf$ but not $\afas$  (Table \ref{table:varparttable}), demonstrating that PFT differences entered largely in leaf mass-based variables.
@@
 
-Climate variables consistently added little explanatory power when added to a statistical model that included PFT and height (the adjusted R$^2$ frequently decreased, see Table \ref{table:varparttable}).
-The only exception was LMA, for which a model that included MAT explained 53\% of the variation, compared to a model with height and PFT (45\%).
+Climate variables consistently added little explanatory power when added to a statistical model that included PFT and height (the adjusted R$^2$ frequently decreased), or to a model that included only PFT (see Table \ref{table:varparttable}).
+The only exception was LMA, for which a model that included MAT, PFT and height explained 53\% of the variation, compared to a model with height and PFT (45\%).
 Since the lack of climate effects on biomass distribution is in contrast with recent finding of \citet{reich2014}, we also analysed climate effects with methods similar to that study.
@@
 2, 3), we expect correlations between climate variables and observed biomass distribution when plant height is not first accounted for.
-Our database did not lend itself well to a direct test of this hypothesis, due to uneven sampling across the range in plant height along climate gradients.
+However, our database did not lend itself well to a direct test of this hypothesis, and indeed we did not find any correlations between any of the studied variables and climate when height was not first accounted for (Table 2).
 

### 2016-04-27 manuscript.Rnw
Final changes - version as submitted

@@
 % Notes:
-% - For submission, convert to Word docx using Latex2rtf. 
+% - For submission, convert to Word docx using Latex2rtf:
+%   - set clean=FALSE in call to latex_build() in remake.yml. Then run, 'latex2rt manuscript.tex' (in shell)
 % - Do not include line numbers here, add them in Word. Also do not embed figures. 
@@
 % Then uncomment the following:
-%' <<echo=FALSE, eval=FALSE>>=
-%' nrw <- get_wordcount("wordcount.txt")
-%' @
-%' Total number of words: \Sexpr{nrw$Total} \\
-%' Abstract: \Sexpr{nrw$Abstract} \\
-%' Introduction: \Sexpr{nrw$Introduction} \\
-%' Methods and Materials: \Sexpr{nrw$Methods} \\
-%' Results: \Sexpr{nrw$Results} \\
-%' Discussion: \Sexpr{nrw$Discussion} \\
-%' Number of figures: 6 (all in colour) \\
-%' Number of tables: 2 \\
-%' Supporting information : 6 Figures (all in colour).
+<<echo=FALSE, eval=TRUE>>=
+nrw <- get_wordcount("wordcount.txt")
+@
+Total number of words: \Sexpr{nrw$Total} \\
+Abstract: \Sexpr{nrw$Abstract} \\
+Introduction: \Sexpr{nrw$Introduction} \\
+Methods and Materials: \Sexpr{nrw$Methods} \\
+Results: \Sexpr{nrw$Results} \\
+Discussion: \Sexpr{nrw$Discussion} \\
+Number of figures: 6 (all in colour) \\
+Number of tables: 2 \\
+Supporting information : 6 Figures (all in colour).
 
@@
 A final aim was to test for shifts in biomass distribution with climate.
-Little is known about global-scale patterns in biomass distribution in relation to climate, but one might expect biomass distribution to shift with precipitation or mean annual temperature, as small-scale comparisons have shown such responses \citep{berninger1994,callaway1994,delucia2000,poyatos2007}.
+Little is known about global-scale patterns in biomass distribution in relation to climate, but one might expect biomass distribution to shift with precipitation or mean annual temperature, as small-scale comparisons have shown such responses \citep{berninger1995,callaway1994,delucia2000,poyatos2007}.
 \citet{reich2014} has also shown that after controlling  for the large size-related variation in biomass distribution, the ratio of leaf to total plant mass (averaged across whole stands, either including or excluding below-ground components) increases with mean annual temperature in both gymnosperms and angiosperms.
@@
 Therefore, we did not use standard allometric equations to study relationships between size and biomass variables.
-Instead, we used generalised additive models (GAM), a semi-parametric modelling approach that makes no assumption about the shape of the relationships, to visualise the relationships between biomass and plant size variables, and to estimate variables such as $\mfms$ and $\afas$ at a common plant height.
+Instead, we used generalised additive models (GAM), a semi-parametric modelling approach that makes no assumption about the shape of the relationships, to visualise the relationships between biomass and plant size variables, and to estimate variables such as $\mfms$ and $\afms$ at a common plant height.
 In all fitted GAMs, we used a cubic regression spline.
@@
 2014 (not shown) to determine whether this shift in PFT with climate could also explain their finding of an increase in $\mfms$ with MAT, but found the pattern still held within gymnosperms and deciduous angiosperms.
+
+The result that $\afas$ did not correlate with climate seems to contradict reports, from studies on one or few species, that the ratio of sapwood area to leaf area (the `Huber value`) varies across climate gradients as quantified by summer evaporative demand \citep{delucia2000,poyatos2007}, rainfall and temperature \citep{callaway1994}, and temperature \citep{berninger1995}.
+Possible reasons for the discrepancy include 1) we studied $\as$ (stem basal area) including the sapwood and heartwood, 2) small-scale climate effects may have been present in our dataset but average out on large climate gradients, and 3) climate effects may be highly species-specific; the majority of reported correlations between the Huber value and climate were for species in the genus \emph{Pinus}.
 
